## Ball Bearings and Bushings {#chap-08-bearings}

```{r setup-08, include=FALSE}
## include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
## include the required packages to make sure the page can be built.
source("R/required_packages.R")

## Set a consistent theme for all plots
theme_bearing <- theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )
```

### Learning Objectives {#sec:objectives-08}

By the end of this chapter, you will be able to:

1. **Classify** bearing types by their operating principles and load capabilities
2. **Select** appropriate ball bearing types for different applications
3. **Calculate** bearing life (L10) using the basic rating life equation
4. **Apply** load factors to determine equivalent dynamic bearing loads
5. **Determine** required dynamic load ratings for specified life requirements

::: {.callout-think}
**Before You Read:** Every rotating machine—from tiny precision instruments to massive industrial equipment—relies on bearings. What functions do you think bearings serve, and why can't rotating shafts simply rest directly in their housings?
:::

### Introduction to Bearings {#sec:bearings-intro}

Bearings are machine elements that **constrain relative motion** and **reduce friction** between moving parts. They are fundamental to virtually every rotating machine—from tiny precision instruments to massive industrial equipment.

Bearings serve three primary functions:

1. **Support loads** (radial and/or thrust)
2. **Reduce friction** between moving surfaces
3. **Guide motion** with precision and repeatability

---

#### Classification of Bearings {-}

Bearings are broadly classified into two main categories:

```{r bearing-classification, echo=FALSE, fig.width=10, fig.height=6, fig.align='center', fig.cap="Classification of bearing types."}
## Create a hierarchical diagram
ggplot() +
  ## Main header
  geom_rect(aes(xmin = 4, xmax = 8, ymin = 9, ymax = 10), fill = "steelblue", color = "black") +
  annotate("text", x = 6, y = 9.5, label = "BEARINGS", fontface = "bold", size = 5, color = "white") +

  ## Connecting lines
  geom_segment(aes(x = 6, xend = 6, y = 9, yend = 8), linewidth = 1) +
  geom_segment(aes(x = 3, xend = 9, y = 8, yend = 8), linewidth = 1) +
  geom_segment(aes(x = 3, xend = 3, y = 8, yend = 7), linewidth = 1) +
  geom_segment(aes(x = 9, xend = 9, y = 8, yend = 7), linewidth = 1) +

  ## Rolling Element Bearings
  geom_rect(aes(xmin = 1, xmax = 5, ymin = 6, ymax = 7), fill = "orange", color = "black") +
  annotate("text", x = 3, y = 6.5, label = "Rolling Element\n(Anti-Friction)", fontface = "bold", size = 3.5) +

  ## Plain Bearings
  geom_rect(aes(xmin = 7, xmax = 11, ymin = 6, ymax = 7), fill = "forestgreen", color = "black") +
  annotate("text", x = 9, y = 6.5, label = "Plain Bearings\n(Sliding/Bushings)", fontface = "bold", size = 3.5, color = "white") +

  ## Sub-types for Rolling Element
  geom_segment(aes(x = 3, xend = 3, y = 6, yend = 5), linewidth = 0.5) +
  geom_segment(aes(x = 1.5, xend = 4.5, y = 5, yend = 5), linewidth = 0.5) +

  geom_rect(aes(xmin = 0.5, xmax = 2.5, ymin = 3.5, ymax = 5), fill = "lightyellow", color = "black") +
  annotate("text", x = 1.5, y = 4.5, label = "Ball\nBearings", size = 3) +
  annotate("text", x = 1.5, y = 3.8, label = "• Deep groove\n• Angular contact\n• Thrust", size = 2.2, hjust = 0.5) +

  geom_rect(aes(xmin = 3.5, xmax = 5.5, ymin = 3.5, ymax = 5), fill = "lightyellow", color = "black") +
  annotate("text", x = 4.5, y = 4.5, label = "Roller\nBearings", size = 3) +
  annotate("text", x = 4.5, y = 3.8, label = "• Cylindrical\n• Tapered\n• Spherical", size = 2.2, hjust = 0.5) +

  ## Sub-types for Plain Bearings
  geom_segment(aes(x = 9, xend = 9, y = 6, yend = 5), linewidth = 0.5) +
  geom_segment(aes(x = 7.5, xend = 10.5, y = 5, yend = 5), linewidth = 0.5) +

  geom_rect(aes(xmin = 6.5, xmax = 8.5, ymin = 3.5, ymax = 5), fill = "lightgreen", color = "black") +
  annotate("text", x = 7.5, y = 4.5, label = "Journal\nBearings", size = 3) +
  annotate("text", x = 7.5, y = 3.8, label = "• Hydrodynamic\n• Hydrostatic", size = 2.2, hjust = 0.5) +

  geom_rect(aes(xmin = 9.5, xmax = 11.5, ymin = 3.5, ymax = 5), fill = "lightgreen", color = "black") +
  annotate("text", x = 10.5, y = 4.5, label = "Bushings", size = 3) +
  annotate("text", x = 10.5, y = 3.8, label = "• Bronze\n• Polymer\n• Sintered", size = 2.2, hjust = 0.5) +

  ## Key difference box
  geom_rect(aes(xmin = 2.5, xmax = 9.5, ymin = 1, ymax = 2.5), fill = "lavender", color = "black") +
  annotate("text", x = 6, y = 2.1, label = "KEY DIFFERENCE:", fontface = "bold", size = 3.5) +
  annotate("text", x = 6, y = 1.5, label = "Rolling elements use balls/rollers to convert sliding to rolling friction\nPlain bearings rely on a lubricant film or low-friction materials", size = 2.8) +

  coord_fixed(ratio = 1, xlim = c(0, 12), ylim = c(0.5, 10.5)) +
  theme_void()
```

---

### Ball Bearings {-}

Ball bearings are the most common type of rolling element bearing. They use spherical balls as the rolling elements, held between an inner race and an outer race.

#### Anatomy of a Ball Bearing {-}

```{r ball-bearing-anatomy, echo=FALSE, fig.width=10, fig.height=7, fig.align='center', fig.cap="Cross-section of a deep groove ball bearing showing main components."}
## Create ball bearing cross-section
theta <- seq(0, 2*pi, length.out = 100)

ggplot() +
  ## Outer race (external)
  geom_polygon(aes(x = 4 * cos(theta), y = 4 * sin(theta)), fill = "gray60", color = "black") +
  geom_polygon(aes(x = 3.3 * cos(theta), y = 3.3 * sin(theta)), fill = "white", color = "black") +

  ## Inner race
  geom_polygon(aes(x = 2.2 * cos(theta), y = 2.2 * sin(theta)), fill = "gray60", color = "black") +
  geom_polygon(aes(x = 1.5 * cos(theta), y = 1.5 * sin(theta)), fill = "white", color = "black") +

  ## Balls (8 balls)
  geom_point(aes(x = 2.75 * cos(seq(0, 2*pi - pi/4, by = pi/4)),
                 y = 2.75 * sin(seq(0, 2*pi - pi/4, by = pi/4))),
             size = 12, color = "black") +
  geom_point(aes(x = 2.75 * cos(seq(0, 2*pi - pi/4, by = pi/4)),
                 y = 2.75 * sin(seq(0, 2*pi - pi/4, by = pi/4))),
             size = 10, color = "steelblue") +

  ## Labels with lines
  geom_segment(aes(x = 3.7, xend = 5.5, y = 0, yend = 0.5), color = "red", linewidth = 0.8) +
  annotate("text", x = 6.2, y = 0.5, label = "Outer Race", color = "red", size = 3.5, hjust = 0) +

  geom_segment(aes(x = 1.85, xend = 5.5, y = 0, yend = -0.5), color = "blue", linewidth = 0.8) +
  annotate("text", x = 6.2, y = -0.5, label = "Inner Race", color = "blue", size = 3.5, hjust = 0) +

  geom_segment(aes(x = 2.75, xend = 5.5, y = 2.75, yend = 2), color = "darkgreen", linewidth = 0.8) +
  annotate("text", x = 6.2, y = 2, label = "Rolling Elements\n(Balls)", color = "darkgreen", size = 3.5, hjust = 0) +

  geom_segment(aes(x = 0, xend = -3, y = 1.5, yend = 3), color = "purple", linewidth = 0.8) +
  annotate("text", x = -3.5, y = 3.5, label = "Bore\n(Shaft fits here)", color = "purple", size = 3.5, hjust = 0.5) +

  ## Pitch diameter indicator
  geom_segment(aes(x = -2.75, xend = 2.75, y = -4.5, yend = -4.5),
               arrow = arrow(ends = "both", length = unit(0.15, "cm")), linewidth = 0.8) +
  annotate("text", x = 0, y = -5, label = "Pitch Diameter (Dp)", size = 3.5) +

  coord_fixed(ratio = 1, xlim = c(-5, 9), ylim = c(-5.5, 5)) +
  labs(title = "Deep Groove Ball Bearing Cross-Section") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

**Key Components:**

| Component | Function |
|:----------|:---------|
| **Inner Race** | Mounted on the shaft; rotates with the shaft |
| **Outer Race** | Mounted in the housing; typically stationary |
| **Rolling Elements** | Balls that roll between races, carrying the load |
| **Cage (Retainer)** | Separates and guides the balls to prevent contact |
| **Seals/Shields** | Keep lubricant in and contaminants out |

---

#### Types of Ball Bearings {-}

```{r ball-bearing-types-table, echo=FALSE, message=FALSE}
bearing_types <- tribble(
  ~Type, ~Load_Capacity, ~Applications, ~Key_Feature,
  "Deep Groove", "Radial + light axial", "Electric motors, pumps, gearboxes", "Most versatile; handles combined loads",
  "Angular Contact", "Combined radial + high axial", "Machine tool spindles, pumps", "Contact angle 15-40 deg; paired for bidirectional thrust",
  "Self-Aligning", "Radial + light axial", "Conveyors, fans, textile machinery", "Tolerates shaft deflection and misalignment",
  "Thrust Ball", "Axial only", "Automotive steering, crane hooks", "90 deg contact angle; cannot handle radial loads"
)

bearing_types %>%
  kbl(caption = "Common Ball Bearing Types and Applications",
      col.names = c("Type", "Load Capacity", "Typical Applications", "Key Feature"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white")
```

---

### Ball Bearing Life Calculations {#bearing-life}

The life of a rolling element bearing is inherently statistical—even identical bearings under identical conditions will fail at different times due to material fatigue. The **L10 life** (or **basic rating life**) is the standard measure used in bearing selection.

#### The L10 Life Concept {-}

> **Definition:** The L10 life is the number of revolutions (or hours at a given speed) that **90% of a group of identical bearings** will complete or exceed before the first evidence of fatigue failure (flaking/spalling).

In other words, 10% of bearings will fail before reaching L10 life, while 90% will survive.

#### Basic Life Equation {-}

The fundamental bearing life equation is:

$$L_{10} = \left(\frac{C}{P}\right)^p \times 10^6 \text{ revolutions}$$

**Where:**

| Symbol | Description | Units |
|:-------|:------------|:------|
| $L_{10}$ | Basic rating life | revolutions |
| $C$ | Basic dynamic load rating | N or lbf |
| $P$ | Equivalent dynamic bearing load | N or lbf |
| $p$ | Life exponent | — |

**Life Exponent Values:**

- **Ball bearings:** $p = 3$
- **Roller bearings:** $p = 10/3 \approx 3.33$

#### Converting to Hours {-}

For bearings operating at constant speed, it is more practical to express life in hours:

$$L_{10h} = \frac{L_{10}}{60 \cdot n} = \frac{1}{60 \cdot n}\left(\frac{C}{P}\right)^p \times 10^6 \text{ hours}$$

Where $n$ is the rotational speed in RPM.

This simplifies to:

$$L_{10h} = \frac{10^6}{60 \cdot n}\left(\frac{C}{P}\right)^p$$

---

#### Equivalent Dynamic Load {-}

When a bearing experiences both radial ($F_r$) and axial ($F_a$) loads simultaneously, an **equivalent dynamic load** ($P$) must be calculated:

$$P = X \cdot F_r + Y \cdot F_a$$

**Where:**

| Factor | Description |
|:-------|:------------|
| $X$ | Radial load factor (from bearing tables) |
| $Y$ | Axial load factor (from bearing tables) |
| $F_r$ | Applied radial load |
| $F_a$ | Applied axial (thrust) load |

For **pure radial loading** (most common case): $P = F_r$

```{r load-factor-table, echo=FALSE, message=FALSE}
load_factors <- tribble(
  ~Bearing_Type, ~X_Light, ~Y_Light, ~X_Heavy, ~Y_Heavy, ~e,
  "Deep Groove (single row)", 1.0, 0, 0.56, "1.0-2.3*", "0.19-0.44",
  "Angular Contact (a=15 deg)", 1.0, 0, 0.44, "1.18", "0.38",
  "Angular Contact (a=25 deg)", 1.0, 0, 0.44, "0.87", "0.68",
  "Angular Contact (a=40 deg)", 1.0, 0, 0.44, "0.55", "1.14",
  "Self-Aligning", 1.0, 0, 0.65, "3.7*cot(a)", "0.22*cot(a)"
)

load_factors %>%
  kbl(caption = "Load Factors for Ball Bearings (ISO 281)",
      col.names = c("Bearing Type", "X", "Y", "X", "Y", "e"),
      booktabs = TRUE,
      escape = FALSE) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  add_header_above(c(" " = 1, "Fa/Fr ≤ e" = 2, "Fa/Fr > e" = 2, " " = 1)) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  footnote(general = "*Values depend on bearing geometry; consult manufacturer catalog",
           general_title = "Note: ")
```

---

#### Basic Dynamic Load Rating (C) {-}

The **basic dynamic load rating** ($C$) is a catalog value provided by the manufacturer. It represents:

> The constant radial load that a bearing can theoretically endure for a basic rating life of **one million revolutions**.

Typical catalog format:

```{r catalog-table, echo=FALSE, message=FALSE}
catalog_data <- tribble(
  ~Designation, ~d_mm, ~D_mm, ~B_mm, ~C_kN, ~C0_kN, ~Speed_rpm,
  "6205", 25, 52, 15, 14.8, 7.8, 13000,
  "6206", 30, 62, 16, 20.3, 11.2, 11000,
  "6207", 35, 72, 17, 26.0, 15.0, 9500,
  "6208", 40, 80, 18, 30.7, 18.2, 8500,
  "6209", 45, 85, 19, 33.2, 20.0, 7500,
  "6210", 50, 90, 20, 35.1, 21.6, 7000
)

catalog_data %>%
  kbl(caption = "Example Bearing Catalog Data (Deep Groove Series 62xx)",
      col.names = c("Designation", "d (mm)", "D (mm)", "B (mm)", "C (kN)", "C₀ (kN)", "Limiting Speed (rpm)"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  column_spec(5, background = "#F9E7F0") %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  footnote(general = "C = Dynamic load rating; C₀ = Static load rating",
           general_title = "Legend: ")
```

---

#### Adjusted Rating Life {-}

The basic L10 equation assumes ideal operating conditions. In practice, factors such as **reliability requirements**, **lubrication quality**, and **contamination** affect bearing life. The **adjusted rating life** accounts for these factors:

$$L_{na} = a_1 \cdot a_2 \cdot a_3 \cdot L_{10}$$

Or using the modern ISO 281 formulation:

$$L_{nm} = a_1 \cdot a_{ISO} \cdot L_{10}$$

**Where:**

| Factor | Description | Typical Values |
|:-------|:------------|:---------------|
| $a_1$ | Life adjustment factor for reliability | See table below |
| $a_2$ | Material factor | 0.5 to 2.0 (modern steels ≈ 1.0) |
| $a_3$ | Operating conditions factor | 0.1 to 1.0 |
| $a_{ISO}$ | Combined lubrication-contamination factor | 0.1 to 50 |

**Reliability Factor ($a_1$):**

```{r reliability-table, echo=FALSE, message=FALSE}
reliability_data <- tribble(
  ~Reliability, ~Ln, ~a1,
  "90%", "L10", 1.00,
  "95%", "L5", 0.64,
  "96%", "L4", 0.55,
  "97%", "L3", 0.47,
  "98%", "L2", 0.37,
  "99%", "L1", 0.25
)

reliability_data %>%
  kbl(caption = "Reliability Factor a₁",
      col.names = c("Reliability", "Life Symbol", "a₁"),
      booktabs = TRUE,
      align = "c") %>%
  kable_styling(latex_options = "hold_position", full_width = FALSE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(1, background = "#E8F5E9")
```

---

#### Life Calculation Plot {-}

The following plot illustrates how bearing life varies with the C/P ratio:

```{r bearing-life-plot, echo=FALSE, fig.width=10, fig.height=6, fig.align='center', fig.cap="Bearing life as a function of load ratio (C/P)."}
## Calculate life for various C/P ratios
cp_ratio <- seq(1, 10, by = 0.1)
L10_ball <- (cp_ratio)^3 * 1e6 / 1e6  ## in millions of revolutions
L10_roller <- (cp_ratio)^(10/3) * 1e6 / 1e6

life_data <- data.frame(
  CP_Ratio = rep(cp_ratio, 2),
  Life_Millions = c(L10_ball, L10_roller),
  Type = rep(c("Ball Bearing (p=3)", "Roller Bearing (p=10/3)"), each = length(cp_ratio))
)

ggplot(life_data, aes(x = CP_Ratio, y = Life_Millions, color = Type, linetype = Type)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 3, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 3.2, y = 800, label = "C/P = 3\nDesign target", hjust = 0, size = 3) +
  geom_hline(yintercept = 27, linetype = "dotted", alpha = 0.5) +
  annotate("text", x = 8, y = 35, label = "~27M rev @ C/P=3", size = 3) +
  scale_y_log10(labels = scales::comma) +
  scale_color_manual(values = c("steelblue", "darkorange")) +
  labs(
    title = "Ball Bearing Life vs. Load Ratio",
    subtitle = "L10 = (C/P)^p × 10⁶ revolutions",
    x = "Load Ratio (C/P)",
    y = "Life L₁₀ (millions of revolutions)",
    color = "Bearing Type",
    linetype = "Bearing Type"
  ) +
  theme_bearing +
  theme(legend.position = "bottom")
```

> **Design Insight:** A small increase in bearing size (higher C) or reduction in load (lower P) dramatically increases bearing life due to the cubic relationship.

---

### Applied Example: Ball Bearing Life Calculation {#bearing-example}

::: {.question}

A deep groove ball bearing (6207) supports a shaft rotating at **1,750 RPM**. The bearing experiences a **radial load of 3,500 N** and an **axial load of 800 N**. Determine:

1. The equivalent dynamic load ($P$)
2. The basic L10 life in hours
3. The L5 life (95% reliability) in hours

**Given data from catalog:**
- Bearing: 6207
- Dynamic load rating: $C = 26.0$ kN = 26,000 N
- For Fa/Fr ratio, use: $e = 0.24$, $X = 0.56$, $Y = 1.6$
:::

---

::: {.answer}


```{r bearing-example-calc, echo=FALSE, message=FALSE}
## Given values
Fr <- 3500      ## N, radial load
Fa <- 800       ## N, axial load
C <- 26000      ## N, dynamic load rating
n <- 1750       ## RPM
p <- 3          ## ball bearing exponent

## Check Fa/Fr ratio
ratio <- Fa / Fr

## Since Fa/Fr = 0.229 < e = 0.24, use X=1, Y=0
## But let's demonstrate the general case where Fa/Fr > e
X <- 0.56
Y <- 1.6

## Equivalent dynamic load
P_light <- Fr  ## When Fa/Fr <= e
P_heavy <- X * Fr + Y * Fa  ## When Fa/Fr > e

## For this example, Fa/Fr ≈ 0.23 which is close to e
## We'll use the combined load formula for demonstration
P <- max(P_light, P_heavy)

## L10 life in revolutions
L10_rev <- (C / P)^p * 1e6

## L10 life in hours
L10_hours <- L10_rev / (60 * n)

## L5 life (95% reliability)
a1_95 <- 0.64
L5_hours <- a1_95 * L10_hours

## Create results table
results <- tribble(
  ~Step, ~Calculation, ~Result,
  "Step 1: Check load ratio", paste0("Fa/Fr = ", Fa, " / ", Fr, " = ", round(ratio, 3)),
  ifelse(ratio <= 0.24, "≤ e: Use X=1, Y=0", "> e: Use X=0.56, Y=1.6"),
  "Step 2: Equivalent load (P)", paste0("P = X·Fr + Y·Fa = ", X, "×", Fr, " + ", Y, "×", Fa),
  paste0(round(P, 0), " N"),
  "Step 3: L10 (revolutions)", paste0("L10 = (C/P)³ × 10⁶ = (", C/1000, "/", round(P/1000, 2), ")³ × 10⁶"),
  paste0(formatC(L10_rev/1e6, format = "f", digits = 1), " million rev"),
  "Step 4: L10 (hours)", paste0("L10h = L10 / (60 × n) = ", formatC(L10_rev/1e6, format = "f", digits = 1), "×10⁶ / (60 × ", n, ")"),
  paste0(formatC(L10_hours, format = "f", digits = 0, big.mark = ","), " hours"),
  "Step 5: L5 (95% reliability)", paste0("L5 = a₁ × L10 = 0.64 × ", formatC(L10_hours, format = "f", digits = 0)),
  paste0(formatC(L5_hours, format = "f", digits = 0, big.mark = ","), " hours")
)

results %>%
  kbl(caption = "Ball Bearing Life Calculation Steps",
      booktabs = TRUE,
      col.names = c("Step", "Calculation", "Result")) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(4, background = "#E8F5E9", bold = TRUE) %>%
  row_spec(5, background = "#E3F2FD", bold = TRUE)
```


###### Interpretation {-}

- The **L10 life** of `r formatC(L10_hours, format = "f", digits = 0, big.mark = ",")` hours means that 90% of these bearings will operate at least this long before fatigue failure.
- For applications requiring **95% reliability**, use the L5 life of `r formatC(L5_hours, format = "f", digits = 0, big.mark = ",")` hours.
- At 24/7 operation, L10 represents approximately `r round(L10_hours / (24*365), 1)` years of service.

:::


### Plain Bearings (Bushings) {#bushings}

Plain bearings, also called **bushings** or **sleeve bearings**, are the simplest form of bearing. They consist of a bearing surface with no rolling elements—the shaft slides directly against the bearing material.

#### Advantages of Bushings {-}

| Advantage | Explanation |
|:----------|:------------|
| **Compact design** | No need for rolling elements; minimal radial space |
| **Low cost** | Simple manufacturing; no precision balls or races |
| **Quiet operation** | No rolling elements to generate noise |
| **High load capacity** | Large contact area; excellent for shock loads |
| **Self-lubricating options** | Polymer and sintered metal bushings need no external lubrication |

#### Types of Bushings {-}

```{r bushing-types-table, echo=FALSE, message=FALSE}
bushing_types <- tribble(
  ~Type, ~Material, ~PV_Limit, ~Applications, ~Lubrication,
  "Bronze Bushings", "Cast bronze, SAE 660", "50,000 - 75,000 psi·ft/min", "Heavy machinery, pumps", "Oil/grease required",
  "Oil-Impregnated Bronze", "Sintered bronze (porous)", "25,000 - 50,000 psi·ft/min", "Fractional HP motors, fans", "Self-lubricating",
  "PTFE-Lined", "Steel + PTFE coating", "10,000 - 25,000 psi·ft/min", "Food processing, medical", "Dry running",
  "Polymer (Nylon, Acetal)", "Engineering plastics", "3,000 - 10,000 psi·ft/min", "Light duty, instrumentation", "Often dry",
  "PEEK Composite", "High-performance polymer", "15,000 - 40,000 psi·ft/min", "Aerospace, high temperature", "Dry or minimal"
)

bushing_types %>%
  kbl(caption = "Common Bushing Materials and PV Limits",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), full_width = TRUE) %>%
  column_spec(3, background = "#F9E7F0") %>%
  row_spec(0, background = "#D81B60", color = "white")
```

---

### PV Limits for Bushings {#pv-limits}

The **PV limit** is the most important design criterion for plain bearings. It represents the maximum combination of **bearing pressure** (P) and **sliding velocity** (V) that the bearing material can sustain without excessive wear or failure.

#### The PV Concept {-}

$$PV = P \times V$$

**Where:**

| Symbol | Description | Typical Units |
|:-------|:------------|:--------------|
| $P$ | Bearing pressure | psi (or MPa) |
| $V$ | Sliding velocity | ft/min (or m/s) |
| $PV$ | PV factor | psi·ft/min (or MPa·m/s) |

#### Calculating Bearing Pressure (P) {-}

For a journal bearing (shaft in a bushing):

$$P = \frac{F}{d \times L}$$

**Where:**

| Symbol | Description |
|:-------|:------------|
| $F$ | Radial load on the bearing (lbf or N) |
| $d$ | Shaft (journal) diameter (in or mm) |
| $L$ | Bearing length (in or mm) |

> **Note:** This is the **projected area** method, using $A = d \times L$, not the cylindrical surface area.

#### Calculating Sliding Velocity (V) {-}

$$V = \pi \times d \times n$$

For consistent units:

**English:**
$$V = \frac{\pi \times d \times n}{12} \quad \text{(ft/min, where } d \text{ in inches, } n \text{ in RPM)}$$

**Metric:**
$$V = \frac{\pi \times d \times n}{60,000} \quad \text{(m/s, where } d \text{ in mm, } n \text{ in RPM)}$$

---

#### PV Limit Visualization {-}

```{r pv-limit-plot, echo=FALSE, fig.width=10, fig.height=7, fig.align='center', fig.cap="PV limit diagram showing safe operating region for different bushing materials."}
## Create PV limit curves
P_values <- seq(100, 5000, by = 50)

## PV limits for different materials (psi·ft/min)
pv_bronze <- 50000
pv_sintered <- 35000
pv_ptfe <- 15000
pv_nylon <- 5000

pv_data <- data.frame(
  P = rep(P_values, 4),
  V = c(pv_bronze/P_values, pv_sintered/P_values, pv_ptfe/P_values, pv_nylon/P_values),
  Material = rep(c("Bronze (PV=50,000)", "Sintered Bronze (PV=35,000)",
                   "PTFE-Lined (PV=15,000)", "Nylon (PV=5,000)"), each = length(P_values))
)

## Limit velocity data for realistic display
pv_data$V <- pmin(pv_data$V, 1000)

ggplot(pv_data, aes(x = P, y = V, color = Material)) +
  geom_line(linewidth = 1.2) +

  ## Add safe operating point example
  geom_point(aes(x = 1000, y = 30), color = "darkgreen", size = 4) +
  annotate("text", x = 1200, y = 35, label = "Safe\noperating point", color = "darkgreen", size = 3, hjust = 0) +

  ## Add unsafe point example
  geom_point(aes(x = 2500, y = 100), color = "red", size = 4, shape = 4, stroke = 2) +
  annotate("text", x = 2700, y = 110, label = "Exceeds Nylon\nPV limit!", color = "red", size = 3, hjust = 0) +

  ## Shade regions
  annotate("rect", xmin = 100, xmax = 5000, ymin = 0, ymax = 50,
           fill = "lightgreen", alpha = 0.2) +
  annotate("text", x = 4500, y = 25, label = "Safe for all", size = 3, hjust = 1) +

  scale_color_manual(values = c("Bronze (PV=50,000)" = "sienna",
                                "Sintered Bronze (PV=35,000)" = "darkorange",
                                "PTFE-Lined (PV=15,000)" = "steelblue",
                                "Nylon (PV=5,000)" = "purple")) +
  labs(
    title = "PV Limit Curves for Common Bushing Materials",
    subtitle = "Operating point must fall BELOW the curve for the selected material",
    x = "Bearing Pressure P (psi)",
    y = "Sliding Velocity V (ft/min)",
    color = "Material"
  ) +
  coord_cartesian(ylim = c(0, 500)) +
  theme_bearing +
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold"))
```

---

#### Design Rules for PV {-}

1. **Calculate the actual PV** for your application
2. **Select a bushing material** with a PV limit **at least 2× higher** than calculated
3. **Check individual P and V limits** separately (materials have maximums for each)
4. **Consider temperature effects**—PV limits decrease at elevated temperatures

```{r individual-limits-table, echo=FALSE, message=FALSE}
limits_data <- tribble(
  ~Material, ~Max_P_psi, ~Max_V_ftmin, ~Max_PV, ~Max_Temp_F,
  "SAE 660 Bronze", 4000, 750, 75000, 450,
  "Sintered Bronze (oil-impreg.)", 2000, 400, 50000, 300,
  "PTFE/Bronze Composite", 2500, 500, 25000, 500,
  "Acetal (Delrin)", 1500, 600, 10000, 180,
  "Nylon 6/6", 1000, 400, 5000, 200,
  "PEEK", 3000, 650, 40000, 480
)

limits_data %>%
  kbl(caption = "Individual P, V, and PV Limits for Common Bushing Materials",
      col.names = c("Material", "Max P (psi)", "Max V (ft/min)", "Max PV (psi-ft/min)", "Max Temp (F)"),
      booktabs = TRUE,
      align = "lcccc") %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  column_spec(4, background = "#F9E7F0") %>%
  row_spec(0, background = "#D81B60", color = "white")
```

---

### Applied Example: Bushing PV Calculation {#bushing-example}

::: {.question}

A shaft rotating at **500 RPM** is supported by a bronze bushing. The shaft diameter is **1.5 inches** and the bushing length is **2.0 inches**. The radial load on the bushing is **800 lbf**.

Determine:
1. The bearing pressure ($P$)
2. The sliding velocity ($V$)
3. The PV value
4. Whether a SAE 660 bronze bushing is suitable
:::

::: {.answer}

```{r bushing-example-calc, echo=FALSE, message=FALSE}
## Given values
F_load <- 800    ## lbf
d_shaft <- 1.5   ## inches
L_bush <- 2.0    ## inches
n_rpm <- 500     ## RPM

## Calculations
P_bearing <- F_load / (d_shaft * L_bush)  ## psi
V_sliding <- (pi * d_shaft * n_rpm) / 12   ## ft/min
PV_actual <- P_bearing * V_sliding          ## psi·ft/min

## SAE 660 Bronze limits
PV_limit_bronze <- 75000
P_max_bronze <- 4000
V_max_bronze <- 750

## Safety factor
SF <- PV_limit_bronze / PV_actual

## Create results table
bushing_results <- tribble(
  ~Step, ~Formula, ~Calculation, ~Result,
  "Step 1: Bearing Pressure", "P = F / (d × L)",
  paste0("P = ", F_load, " / (", d_shaft, " × ", L_bush, ")"),
  paste0(round(P_bearing, 1), " psi"),

  "Step 2: Sliding Velocity", "V = π·d·n / 12",
  paste0("V = π × ", d_shaft, " × ", n_rpm, " / 12"),
  paste0(round(V_sliding, 1), " ft/min"),

  "Step 3: PV Factor", "PV = P × V",
  paste0("PV = ", round(P_bearing, 1), " × ", round(V_sliding, 1)),
  paste0(formatC(PV_actual, format = "f", digits = 0, big.mark = ","), " psi·ft/min"),

  "Step 4: Check P Limit", "P < P_max (4,000 psi)",
  paste0(round(P_bearing, 1), " < 4,000?"),
  ifelse(P_bearing < P_max_bronze, "OK ✓", "FAIL ✗"),

  "Step 5: Check V Limit", "V < V_max (750 ft/min)",
  paste0(round(V_sliding, 1), " < 750?"),
  ifelse(V_sliding < V_max_bronze, "OK ✓", "FAIL ✗"),

  "Step 6: Check PV Limit", "PV < PV_max (75,000)",
  paste0(formatC(PV_actual, format = "f", digits = 0), " < 75,000?"),
  ifelse(PV_actual < PV_limit_bronze, "OK ✓", "FAIL ✗"),

  "Step 7: Safety Factor", "SF = PV_limit / PV_actual",
  paste0("SF = 75,000 / ", formatC(PV_actual, format = "f", digits = 0)),
  paste0(round(SF, 2))
)

bushing_results %>%
  kbl(caption = "Bushing PV Calculation Steps",
      booktabs = TRUE,
      col.names = c("Step", "Formula", "Calculation", "Result")) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(3, background = "#E8F5E9", bold = TRUE) %>%
  row_spec(4:6, background = "#E3F2FD")
```

###### Conclusion {-}

The SAE 660 bronze bushing is **suitable** for this application:

- All individual limits (P, V) are satisfied
- The actual PV (`r formatC(PV_actual, format = "f", digits = 0, big.mark = ",")` psi·ft/min) is well below the material limit (75,000 psi·ft/min)
- Safety factor of `r round(SF, 2)` provides adequate margin

:::


---

### Bearing vs. Bushing Selection Guide {-}

When should you use a ball bearing versus a bushing? The following table provides guidance:

```{r selection-guide, echo=FALSE, message=FALSE}
selection_guide <- tribble(
  ~Factor, ~Ball_Bearing, ~Bushing,
  "Speed", "High speeds (>3,000 RPM)", "Low to moderate speeds",
  "Load Type", "Radial and/or axial", "Primarily radial",
  "Shock Loads", "Poor—balls can indent races", "Excellent—large contact area",
  "Space (radial)", "Requires more radial space", "Very compact",
  "Lubrication", "Sealed or periodic greasing", "Self-lubricating options available",
  "Noise", "Some noise from rolling elements", "Quieter operation",
  "Cost", "Higher initial cost", "Lower cost",
  "Maintenance", "Replace entire bearing", "Can often be replaced in-situ",
  "Life Prediction", "Well-established (L10)", "Less precise; wear-based"
)

selection_guide %>%
  kbl(caption = "Ball Bearing vs. Bushing Selection Criteria",
      col.names = c("Factor", "Ball Bearing", "Bushing (Plain Bearing)"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  column_spec(2, background = "#FFF3E0") %>%
  column_spec(3, background = "#E8F5E9")
```

---

### Summary Equations {-}

#### Ball Bearing Life {-}

| Equation | Description |
|:---------|:------------|
| $L_{10} = \left(\frac{C}{P}\right)^3 \times 10^6$ | Basic L10 life (revolutions) for ball bearings |
| $L_{10h} = \frac{10^6}{60n}\left(\frac{C}{P}\right)^3$ | L10 life in hours |
| $P = X \cdot F_r + Y \cdot F_a$ | Equivalent dynamic load |
| $L_{na} = a_1 \cdot a_{ISO} \cdot L_{10}$ | Adjusted rating life |

#### Bushing PV Limits {-}

| Equation | Description |
|:---------|:------------|
| $P = \frac{F}{d \times L}$ | Bearing pressure (projected area) |
| $V = \frac{\pi d n}{12}$ | Sliding velocity (ft/min) |
| $PV = P \times V$ | PV factor—must be less than material PV limit |

---

### Design Checklist {-}

**For Ball Bearings:**

- [ ] Determine radial and axial loads
- [ ] Calculate equivalent dynamic load ($P$)
- [ ] Select bearing with $C/P$ ratio giving adequate life
- [ ] Verify speed is below limiting speed
- [ ] Apply reliability and operating condition factors if needed
- [ ] Specify lubrication and sealing requirements

**For Bushings:**

- [ ] Calculate bearing pressure ($P$)
- [ ] Calculate sliding velocity ($V$)
- [ ] Verify $P < P_{max}$ for selected material
- [ ] Verify $V < V_{max}$ for selected material
- [ ] Verify $PV < PV_{limit}$ with safety factor ≥ 2
- [ ] Consider temperature effects on PV limits
- [ ] Specify lubrication (if required) or confirm dry-running capability
