## Ball Bearings and Bushings {#chap-08-bearings}

```{r setup-08, include=FALSE}
## include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
## include the required packages to make sure the page can be built.
source("R/required_packages.R")

## Set a consistent theme for all plots
theme_bearing <- theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )
```

### Learning Objectives {#sec:objectives-08}

By the end of this chapter, you will be able to:

1. **Classify** bearing types by their operating principles and load capabilities
2. **Select** appropriate ball bearing types for different applications
3. **Calculate** bearing life (L10) using the basic rating life equation
4. **Apply** load factors to determine equivalent dynamic bearing loads
5. **Determine** required dynamic load ratings for specified life requirements
6. **Explain** the hydrodynamic wedge effect and how a rotating shaft floats on an oil cushion
7. **Interpret** the Stribeck curve and identify the three lubrication regimes (boundary, mixed, hydrodynamic)
8. **Connect** bushing performance to sintered bearing manufacturing (see Chapter \@ref(chap-sintering))

::: {.callout-think}
**Before You Read:** Every rotating machine—from tiny precision instruments to massive industrial equipment—relies on bearings. What functions do you think bearings serve, and why can't rotating shafts simply rest directly in their housings?
:::

### Introduction to Bearings {#sec:bearings-intro}

Bearings are machine elements that **constrain relative motion** and **reduce friction** between moving parts. They are fundamental to virtually every rotating machine—from tiny precision instruments to massive industrial equipment.

Bearings serve three primary functions:

1. **Support loads** (radial and/or thrust)
2. **Reduce friction** between moving surfaces
3. **Guide motion** with precision and repeatability

---

#### Classification of Bearings {-}

Bearings are broadly classified into two main categories:

```{r bearing-classification, echo=FALSE, fig.width=10, fig.height=6, fig.align='center', fig.cap="Classification of bearing types."}
## Create a hierarchical diagram
ggplot() +
  ## Main header
  geom_rect(aes(xmin = 4, xmax = 8, ymin = 9, ymax = 10), fill = "steelblue", color = "black") +
  annotate("text", x = 6, y = 9.5, label = "BEARINGS", fontface = "bold", size = 5, color = "white") +

  ## Connecting lines
  geom_segment(aes(x = 6, xend = 6, y = 9, yend = 8), linewidth = 1) +
  geom_segment(aes(x = 3, xend = 9, y = 8, yend = 8), linewidth = 1) +
  geom_segment(aes(x = 3, xend = 3, y = 8, yend = 7), linewidth = 1) +
  geom_segment(aes(x = 9, xend = 9, y = 8, yend = 7), linewidth = 1) +

  ## Rolling Element Bearings
  geom_rect(aes(xmin = 1, xmax = 5, ymin = 6, ymax = 7), fill = "orange", color = "black") +
  annotate("text", x = 3, y = 6.5, label = "Rolling Element\n(Anti-Friction)", fontface = "bold", size = 3.5) +

  ## Plain Bearings
  geom_rect(aes(xmin = 7, xmax = 11, ymin = 6, ymax = 7), fill = "forestgreen", color = "black") +
  annotate("text", x = 9, y = 6.5, label = "Plain Bearings\n(Sliding/Bushings)", fontface = "bold", size = 3.5, color = "white") +

  ## Sub-types for Rolling Element
  geom_segment(aes(x = 3, xend = 3, y = 6, yend = 5), linewidth = 0.5) +
  geom_segment(aes(x = 1.5, xend = 4.5, y = 5, yend = 5), linewidth = 0.5) +

  geom_rect(aes(xmin = 0.5, xmax = 2.5, ymin = 3.5, ymax = 5), fill = "lightyellow", color = "black") +
  annotate("text", x = 1.5, y = 4.5, label = "Ball\nBearings", size = 3) +
  annotate("text", x = 1.5, y = 3.8, label = "• Deep groove\n• Angular contact\n• Thrust", size = 2.2, hjust = 0.5) +

  geom_rect(aes(xmin = 3.5, xmax = 5.5, ymin = 3.5, ymax = 5), fill = "lightyellow", color = "black") +
  annotate("text", x = 4.5, y = 4.5, label = "Roller\nBearings", size = 3) +
  annotate("text", x = 4.5, y = 3.8, label = "• Cylindrical\n• Tapered\n• Spherical", size = 2.2, hjust = 0.5) +

  ## Sub-types for Plain Bearings
  geom_segment(aes(x = 9, xend = 9, y = 6, yend = 5), linewidth = 0.5) +
  geom_segment(aes(x = 7.5, xend = 10.5, y = 5, yend = 5), linewidth = 0.5) +

  geom_rect(aes(xmin = 6.5, xmax = 8.5, ymin = 3.5, ymax = 5), fill = "lightgreen", color = "black") +
  annotate("text", x = 7.5, y = 4.5, label = "Journal\nBearings", size = 3) +
  annotate("text", x = 7.5, y = 3.8, label = "• Hydrodynamic\n• Hydrostatic", size = 2.2, hjust = 0.5) +

  geom_rect(aes(xmin = 9.5, xmax = 11.5, ymin = 3.5, ymax = 5), fill = "lightgreen", color = "black") +
  annotate("text", x = 10.5, y = 4.5, label = "Bushings", size = 3) +
  annotate("text", x = 10.5, y = 3.8, label = "• Bronze\n• Polymer\n• Sintered", size = 2.2, hjust = 0.5) +

  ## Key difference box
  geom_rect(aes(xmin = 2.5, xmax = 9.5, ymin = 1, ymax = 2.5), fill = "lavender", color = "black") +
  annotate("text", x = 6, y = 2.1, label = "KEY DIFFERENCE:", fontface = "bold", size = 3.5) +
  annotate("text", x = 6, y = 1.5, label = "Rolling elements use balls/rollers to convert sliding to rolling friction\nPlain bearings rely on a lubricant film or low-friction materials", size = 2.8) +

  coord_fixed(ratio = 1, xlim = c(0, 12), ylim = c(0.5, 10.5)) +
  theme_void()
```

---

### Ball Bearings {-}

Ball bearings are the most common type of rolling element bearing. They use spherical balls as the rolling elements, held between an inner race and an outer race.

#### Anatomy of a Ball Bearing {-}

```{r ball-bearing-anatomy, echo=FALSE, fig.width=10, fig.height=7, fig.align='center', fig.cap="Cross-section of a deep groove ball bearing showing main components."}
## Create ball bearing cross-section
theta <- seq(0, 2*pi, length.out = 100)

ggplot() +
  ## Outer race (external)
  geom_polygon(aes(x = 4 * cos(theta), y = 4 * sin(theta)), fill = "gray60", color = "black") +
  geom_polygon(aes(x = 3.3 * cos(theta), y = 3.3 * sin(theta)), fill = "white", color = "black") +

  ## Inner race
  geom_polygon(aes(x = 2.2 * cos(theta), y = 2.2 * sin(theta)), fill = "gray60", color = "black") +
  geom_polygon(aes(x = 1.5 * cos(theta), y = 1.5 * sin(theta)), fill = "white", color = "black") +

  ## Balls (8 balls)
  geom_point(aes(x = 2.75 * cos(seq(0, 2*pi - pi/4, by = pi/4)),
                 y = 2.75 * sin(seq(0, 2*pi - pi/4, by = pi/4))),
             size = 12, color = "black") +
  geom_point(aes(x = 2.75 * cos(seq(0, 2*pi - pi/4, by = pi/4)),
                 y = 2.75 * sin(seq(0, 2*pi - pi/4, by = pi/4))),
             size = 10, color = "steelblue") +

  ## Labels with lines
  geom_segment(aes(x = 3.7, xend = 5.5, y = 0, yend = 0.5), color = "red", linewidth = 0.8) +
  annotate("text", x = 6.2, y = 0.5, label = "Outer Race", color = "red", size = 3.5, hjust = 0) +

  geom_segment(aes(x = 1.85, xend = 5.5, y = 0, yend = -0.5), color = "blue", linewidth = 0.8) +
  annotate("text", x = 6.2, y = -0.5, label = "Inner Race", color = "blue", size = 3.5, hjust = 0) +

  geom_segment(aes(x = 2.75, xend = 5.5, y = 2.75, yend = 2), color = "darkgreen", linewidth = 0.8) +
  annotate("text", x = 6.2, y = 2, label = "Rolling Elements\n(Balls)", color = "darkgreen", size = 3.5, hjust = 0) +

  geom_segment(aes(x = 0, xend = -3, y = 1.5, yend = 3), color = "purple", linewidth = 0.8) +
  annotate("text", x = -3.5, y = 3.5, label = "Bore\n(Shaft fits here)", color = "purple", size = 3.5, hjust = 0.5) +

  ## Pitch diameter indicator
  geom_segment(aes(x = -2.75, xend = 2.75, y = -4.5, yend = -4.5),
               arrow = arrow(ends = "both", length = unit(0.15, "cm")), linewidth = 0.8) +
  annotate("text", x = 0, y = -5, label = "Pitch Diameter (Dp)", size = 3.5) +

  coord_fixed(ratio = 1, xlim = c(-5, 9), ylim = c(-5.5, 5)) +
  labs(title = "Deep Groove Ball Bearing Cross-Section") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

**Key Components:**

| Component | Function |
|:----------|:---------|
| **Inner Race** | Mounted on the shaft; rotates with the shaft |
| **Outer Race** | Mounted in the housing; typically stationary |
| **Rolling Elements** | Balls that roll between races, carrying the load |
| **Cage (Retainer)** | Separates and guides the balls to prevent contact |
| **Seals/Shields** | Keep lubricant in and contaminants out |

---

#### Types of Ball Bearings {-}

```{r ball-bearing-types-table, echo=FALSE, message=FALSE}
bearing_types <- tribble(
  ~Type, ~Load_Capacity, ~Applications, ~Key_Feature,
  "Deep Groove", "Radial + light axial", "Electric motors, pumps, gearboxes", "Most versatile; handles combined loads",
  "Angular Contact", "Combined radial + high axial", "Machine tool spindles, pumps", "Contact angle 15-40 deg; paired for bidirectional thrust",
  "Self-Aligning", "Radial + light axial", "Conveyors, fans, textile machinery", "Tolerates shaft deflection and misalignment",
  "Thrust Ball", "Axial only", "Automotive steering, crane hooks", "90 deg contact angle; cannot handle radial loads"
)

bearing_types %>%
  kbl(caption = "Common Ball Bearing Types and Applications",
      col.names = c("Type", "Load Capacity", "Typical Applications", "Key Feature"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white")
```

---

### Ball Bearing Life Calculations {#bearing-life}

The life of a rolling element bearing is inherently **statistical**—even identical bearings operating under identical conditions will fail at different times due to the random nature of material fatigue. Understanding this statistical behavior is essential for proper bearing selection and reliability engineering.

#### Understanding Statistical Bearing Life {-}

::: {.callout-think}
**Before You Read:** If you tested 100 identical bearings under identical loads and speeds, would they all fail at the same time? What factors might cause variation in their lifetimes?
:::

When bearings fail, they do so by **fatigue spalling**—small flakes of material breaking away from the rolling surfaces due to repeated stress cycles. The location where a fatigue crack initiates depends on microscopic material imperfections (inclusions, voids, grain boundaries) that vary randomly from bearing to bearing, even in high-quality manufacturing.

If we test a large group of identical bearings under identical conditions and plot the number of failures against operating time, we get a distribution of bearing lives. This distribution is **not** a normal (Gaussian) distribution—it follows a **Weibull distribution**, which is skewed to the right.

```{r bearing-life-distribution, echo=FALSE, fig.width=10, fig.height=6, fig.align='center', fig.cap="Statistical distribution of bearing failures showing L10, L50, and mean life concepts."}
## Simulate bearing life data using Weibull distribution
## Shape parameter ~1.5 is typical for rolling bearings (ISO 281)
set.seed(42)
n_bearings <- 1000
shape <- 1.5  ## Weibull shape parameter (beta)
scale <- 100  ## Weibull scale parameter (characteristic life)

## Generate Weibull-distributed life data
lives <- rweibull(n_bearings, shape = shape, scale = scale)

## Calculate key percentiles
L10 <- quantile(lives, 0.10)
L50 <- quantile(lives, 0.50)
mean_life <- mean(lives)

## Create histogram data
life_df <- data.frame(life = lives)

ggplot(life_df, aes(x = life)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = "steelblue", color = "white", alpha = 0.7) +
  geom_density(color = "darkblue", linewidth = 1.2) +
  ## Add vertical lines for L10, L50, and mean
  geom_vline(xintercept = L10, color = "red", linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = L50, color = "darkgreen", linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = mean_life, color = "purple", linewidth = 1.2, linetype = "dotted") +
  ## Add labels
  annotate("text", x = L10, y = 0.018, label = paste0("L10\n(", round(L10, 0), " hrs)"),
           color = "red", hjust = 1.1, size = 3.5, fontface = "bold") +
  annotate("text", x = L50, y = 0.016, label = paste0("L50\n(", round(L50, 0), " hrs)"),
           color = "darkgreen", hjust = -0.1, size = 3.5, fontface = "bold") +
  annotate("text", x = mean_life, y = 0.014, label = paste0("Mean\n(", round(mean_life, 0), " hrs)"),
           color = "purple", hjust = -0.1, size = 3.5, fontface = "bold") +
  ## Add shaded region for early failures (first 10%)
  annotate("rect", xmin = 0, xmax = L10, ymin = 0, ymax = 0.02,
           fill = "red", alpha = 0.2) +
  annotate("text", x = L10/2, y = 0.001, label = "10% fail\nbefore L10",
           color = "darkred", size = 3) +
  labs(
    title = "Statistical Distribution of Bearing Lives",
    subtitle = "Weibull distribution (typical for rolling element bearings)",
    x = "Bearing Life (hours)",
    y = "Probability Density"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +
  scale_x_continuous(limits = c(0, 250), expand = c(0, 0))
```

#### L10 vs. L50: Which Life Do We Use? {-}

The figure above illustrates several important life values:

| Life Designation | Definition | Typical Relationship |
|:-----------------|:-----------|:---------------------|
| **L10** | Life exceeded by 90% of bearings | **Design basis** (conservative) |
| **L50** (median) | Life exceeded by 50% of bearings | L50 ≈ 5 × L10 |
| **Mean life** | Average life of all bearings | Mean ≈ 5.4 × L10 |

::: {.callout-industry}
**Why L10, not L50?** In most engineering applications, we design for the **L10 life** because:

1. **Safety margin:** 90% of bearings will exceed this life—conservative design
2. **Replacement planning:** We know when 10% might start failing
3. **Industry standard:** All bearing catalogs use L10 ratings (ISO 281)

If we designed for L50 life, half of our bearings would fail before the "rated" life—unacceptable for most applications!
:::

The relationship between L10 and L50 also explains an important insight: if a bearing achieves its L10 life, it still has substantial remaining life. The **median bearing** (L50) lasts about 5 times longer than L10!

#### The L10 Life Concept {-}

> **Definition:** The L10 life is the number of revolutions (or hours at a given speed) that **90% of a group of identical bearings** will complete or exceed before the first evidence of fatigue failure (flaking/spalling).

In other words, 10% of bearings will fail before reaching L10 life, while 90% will survive.

#### Basic Life Equation {-}

The fundamental bearing life equation is:

$$L_{10} = \left(\frac{C}{P}\right)^p \times 10^6 \text{ revolutions}$$

**Where:**

| Symbol | Description | Units |
|:-------|:------------|:------|
| $L_{10}$ | Basic rating life | revolutions |
| $C$ | Basic dynamic load rating | N or lbf |
| $P$ | Equivalent dynamic bearing load | N or lbf |
| $p$ | Life exponent | — |

**Life Exponent Values:**

- **Ball bearings:** $p = 3$
- **Roller bearings:** $p = 10/3 \approx 3.33$

#### Converting to Hours {-}

For bearings operating at constant speed, it is more practical to express life in hours:

$$L_{10h} = \frac{L_{10}}{60 \cdot n} = \frac{1}{60 \cdot n}\left(\frac{C}{P}\right)^p \times 10^6 \text{ hours}$$

Where $n$ is the rotational speed in RPM.

This simplifies to:

$$L_{10h} = \frac{10^6}{60 \cdot n}\left(\frac{C}{P}\right)^p$$

---

#### Equivalent Dynamic Load {-}

When a bearing experiences both radial ($F_r$) and axial ($F_a$) loads simultaneously, an **equivalent dynamic load** ($P$) must be calculated:

$$P = X \cdot F_r + Y \cdot F_a$$

**Where:**

| Factor | Description |
|:-------|:------------|
| $X$ | Radial load factor (from bearing tables) |
| $Y$ | Axial load factor (from bearing tables) |
| $F_r$ | Applied radial load |
| $F_a$ | Applied axial (thrust) load |

For **pure radial loading** (most common case): $P = F_r$

```{r load-factor-table, echo=FALSE, message=FALSE}
load_factors <- tribble(
  ~Bearing_Type, ~X_Light, ~Y_Light, ~X_Heavy, ~Y_Heavy, ~e,
  "Deep Groove (single row)", 1.0, 0, 0.56, "1.0-2.3*", "0.19-0.44",
  "Angular Contact (a=15 deg)", 1.0, 0, 0.44, "1.18", "0.38",
  "Angular Contact (a=25 deg)", 1.0, 0, 0.44, "0.87", "0.68",
  "Angular Contact (a=40 deg)", 1.0, 0, 0.44, "0.55", "1.14",
  "Self-Aligning", 1.0, 0, 0.65, "3.7*cot(a)", "0.22*cot(a)"
)

load_factors %>%
  kbl(caption = "Load Factors for Ball Bearings (ISO 281)",
      col.names = c("Bearing Type", "X", "Y", "X", "Y", "e"),
      booktabs = TRUE,
      escape = FALSE) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  add_header_above(c(" " = 1, "Fa/Fr ≤ e" = 2, "Fa/Fr > e" = 2, " " = 1)) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  footnote(general = "*Values depend on bearing geometry; consult manufacturer catalog",
           general_title = "Note: ")
```

---

#### Basic Dynamic Load Rating (C) {-}

The **basic dynamic load rating** ($C$) is a catalog value provided by the manufacturer. It represents:

> The constant radial load that a bearing can theoretically endure for a basic rating life of **one million revolutions**.

Typical catalog format:

```{r catalog-table, echo=FALSE, message=FALSE}
catalog_data <- tribble(
  ~Designation, ~d_mm, ~D_mm, ~B_mm, ~C_kN, ~C0_kN, ~Speed_rpm,
  "6205", 25, 52, 15, 14.8, 7.8, 13000,
  "6206", 30, 62, 16, 20.3, 11.2, 11000,
  "6207", 35, 72, 17, 26.0, 15.0, 9500,
  "6208", 40, 80, 18, 30.7, 18.2, 8500,
  "6209", 45, 85, 19, 33.2, 20.0, 7500,
  "6210", 50, 90, 20, 35.1, 21.6, 7000
)

catalog_data %>%
  kbl(caption = "Example Bearing Catalog Data (Deep Groove Series 62xx)",
      col.names = c("Designation", "d (mm)", "D (mm)", "B (mm)", "C (kN)", "C₀ (kN)", "Limiting Speed (rpm)"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  column_spec(5, background = "#F9E7F0") %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  footnote(general = "C = Dynamic load rating; C₀ = Static load rating",
           general_title = "Legend: ")
```

---

#### Calculating the Basic Dynamic Load Rating (C) {-}

While catalog values are convenient, understanding **how** the dynamic load rating is calculated provides valuable engineering insight. The basic dynamic load rating for radial ball bearings is determined by:

$$C = f_c \cdot (i \cdot \cos\alpha)^{0.7} \cdot Z^{2/3} \cdot D^{1.8}$$

**Where:**

| Symbol | Description | Units |
|:-------|:------------|:------|
| $C$ | Basic dynamic load rating | N (metric) or lbf (inch) |
| $f_c$ | Geometry and material factor | — (from table) |
| $i$ | Number of rows of balls | — |
| $\alpha$ | Contact angle | degrees |
| $Z$ | Number of balls per row | — |
| $D$ | Ball diameter | mm (metric) or inches (inch) |

The factor $f_c$ depends on the bearing geometry and is tabulated based on the ratio $\frac{D \cos\alpha}{d_m}$, where $d_m$ is the pitch diameter of the ball set.

::: {.callout-think}
**Consider This:** Looking at the formula, which parameter has the greatest influence on load capacity? The ball diameter $D$ appears with an exponent of 1.8—nearly squared! This means **doubling the ball diameter increases load capacity by a factor of** $2^{1.8} \approx 3.5$.
:::

##### Values of $f_c$ (Geometry Factor) {-}

The following table provides $f_c$ values based on the bearing geometry ratio. Use the **Metric** column when ball diameter $D$ is in mm (result is $C$ in Newtons). Use the **Inch** column when $D$ is in inches (result is $C$ in pounds).

```{r fc-values-table, echo=FALSE, message=FALSE}
## fc values table from ANSI-AFBMA Std. 9-1978
fc_values <- tribble(
  ~Ratio, ~Single_Metric, ~Single_Inch, ~Double_Metric, ~Double_Inch, ~SelfAlign_Metric, ~SelfAlign_Inch,
  0.05, 46.7, 3550, 44.2, 3360, 17.3, 1310,
  0.06, 49.1, 3730, 46.5, 3530, 18.6, 1420,
  0.07, 51.1, 3880, 48.4, 3680, 19.9, 1510,
  0.08, 52.8, 4020, 50.0, 3810, 21.1, 1600,
  0.09, 54.3, 4130, 51.4, 3900, 22.3, 1690,
  0.10, 55.5, 4220, 52.6, 4000, 23.4, 1770,
  0.12, 57.5, 4370, 54.5, 4140, 25.6, 1940,
  0.14, 58.8, 4470, 55.7, 4230, 27.7, 2100,
  0.16, 59.6, 4530, 56.5, 4290, 29.7, 2260,
  0.18, 59.9, 4550, 56.8, 4310, 31.7, 2410,
  0.20, 59.9, 4550, 56.8, 4310, 33.5, 2550,
  0.22, 59.6, 4530, 56.5, 4290, 35.2, 2680,
  0.24, 59.0, 4480, 55.9, 4250, 36.8, 2790,
  0.26, 58.2, 4420, 55.1, 4190, 38.2, 2910,
  0.28, 57.1, 4340, 54.1, 4110, 39.4, 3000,
  0.30, 56.0, 4250, 53.0, 4030, 40.3, 3060,
  0.32, 54.6, 4160, 51.8, 3950, 40.9, 3110,
  0.34, 53.2, 4050, 50.4, 3840, 41.2, 3130,
  0.36, 51.7, 3930, 48.9, 3730, 41.3, 3140,
  0.38, 50.0, 3800, 47.4, 3610, 41.0, 3110,
  0.40, 48.4, 3670, 45.8, 3480, 40.4, 3070
)

fc_values %>%
  kbl(caption = "Values of fc for Basic Dynamic Load Rating Calculation",
      col.names = c("Dcosα/dm", "Metric", "Inch", "Metric", "Inch", "Metric", "Inch"),
      booktabs = TRUE,
      digits = c(2, 1, 0, 1, 0, 1, 0)) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), full_width = TRUE) %>%
  add_header_above(c(" " = 1,
                     "Single Row Radial;\nAngular Contact" = 2,
                     "Double Row\nRadial Contact" = 2,
                     "Self-Aligning" = 2)) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(c(6, 10, 11), background = "#E8F5E9") %>%
  footnote(general = "Source: ANSI-AFBMA Std. 9-1978. Metric fc gives C in Newtons (D in mm); Inch fc gives C in pounds (D in inches).",
           general_title = "Note: ")
```

##### Conrad-Type Single-Row Radial Bearing Dimensions {-}

The following table provides dimensions and basic load ratings for Conrad-type (deep groove) single-row radial ball bearings. These are among the most commonly used bearings in industry.

```{r conrad-bearing-table, echo=FALSE, message=FALSE}
## Conrad-type bearing dimensions and ratings
conrad_bearings <- tribble(
  ~Bearing_No, ~Bore_mm, ~Bore_in, ~OD_mm, ~OD_in, ~Width_mm, ~Width_in, ~Num_Balls, ~Ball_Dia, ~C_lbs, ~P_st,
  "102", 15, 0.5906, 32, 1.2598, 9, 0.3543, 9, "3/16", 965, 550,
  "202", 15, 0.5906, 35, 1.3780, 11, 0.4331, 7, "1/4", 1340, 760,
  "302", 15, 0.5906, 42, 1.6535, 13, 0.5118, 8, "17/64", 1660, 930,
  "103", 17, 0.6693, 35, 1.3780, 10, 0.3937, 10, "3/16", 1040, 640,
  "203", 17, 0.6693, 40, 1.5748, 12, 0.4724, 7, "5/16", 1960, 1040,
  "303", 17, 0.6693, 47, 1.8504, 14, 0.5512, 6, "3/8", 2400, 1240,
  "104", 20, 0.7874, 42, 1.6535, 12, 0.4724, 9, "1/4", 1620, 980,
  "204", 20, 0.7874, 47, 1.8504, 14, 0.5512, 8, "5/16", 2210, 1280,
  "304", 20, 0.7874, 52, 2.0472, 15, 0.5906, 7, "3/8", 2760, 1530,
  "105", 25, 0.9843, 47, 1.8504, 12, 0.4724, 10, "1/4", 1740, 1140,
  "205", 25, 0.9843, 52, 2.0472, 15, 0.5906, 9, "5/16", 2420, 1520,
  "305", 25, 0.9843, 62, 2.4409, 17, 0.6693, 8, "13/32", 3550, 2160,
  "106", 30, 1.1811, 55, 2.1654, 13, 0.5118, 11, "9/32", 2290, 1590,
  "206", 30, 1.1811, 62, 2.4409, 16, 0.6299, 9, "3/8", 3360, 2190,
  "306", 30, 1.1811, 72, 2.8346, 19, 0.7480, 8, "1/2", 5120, 3200,
  "107", 35, 1.378, 62, 2.4409, 14, 0.5512, 11, "5/16", 2760, 2010,
  "207", 35, 1.378, 72, 2.8346, 17, 0.6693, 9, "7/16", 4440, 2980,
  "307", 35, 1.378, 80, 3.1496, 21, 0.8268, 8, "17/32", 5750, 3710,
  "108", 40, 1.5748, 68, 2.6772, 15, 0.5906, 13, "5/16", 3060, 2450,
  "208", 40, 1.5748, 80, 3.1496, 18, 0.7087, 9, "1/2", 5640, 3870,
  "308", 40, 1.5748, 90, 3.5433, 23, 0.9055, 8, "5/8", 7670, 5050,
  "109", 45, 1.7717, 75, 2.9528, 16, 0.6299, 13, "11/32", 3630, 2970,
  "209", 45, 1.7717, 85, 3.3465, 19, 0.7480, 9, "1/2", 5660, 3980,
  "309", 45, 1.7717, 100, 3.9370, 25, 0.9843, 8, "11/16", 9120, 6150,
  "110", 50, 1.9685, 80, 3.1496, 16, 0.6299, 14, "11/32", 3770, 3260,
  "210", 50, 1.9685, 90, 3.5433, 20, 0.7874, 10, "1/2", 6070, 4540,
  "310", 50, 1.9685, 110, 4.3307, 27, 1.0630, 8, "3/4", 10680, 7350,
  "111", 55, 2.1654, 90, 3.5433, 18, 0.7087, 13, "13/32", 4890, 3950,
  "211", 55, 2.1654, 100, 3.9370, 21, 0.8268, 10, "9/16", 7500, 5710,
  "311", 55, 2.1654, 120, 4.7244, 29, 1.1417, 8, "13/16", 12350, 8660,
  "112", 60, 2.3622, 95, 3.7402, 18, 0.7087, 14, "13/32", 5090, 4560,
  "212", 60, 2.3622, 110, 4.3307, 22, 0.8661, 10, "5/8", 9070, 6890,
  "312", 60, 2.3622, 130, 5.1181, 31, 1.2205, 8, "7/8", 14130, 10100,
  "113", 65, 2.5591, 100, 3.9370, 18, 0.7087, 15, "13/32", 5280, 4950,
  "213", 65, 2.5591, 120, 4.7244, 23, 0.9055, 10, "11/16", 10770, 8460,
  "313", 65, 2.5591, 140, 5.5118, 33, 1.2992, 8, "15/16", 16010, 11600,
  "114", 70, 2.7559, 110, 4.3307, 20, 0.7874, 14, "15/32", 6580, 6080,
  "214", 70, 2.7559, 125, 4.9213, 24, 0.9449, 10, "11/16", 10760, 8740,
  "314", 70, 2.7559, 150, 5.9055, 35, 1.3780, 8, "1", 18000, 13260
)

conrad_bearings %>%
  select(Bearing_No, Bore_mm, Bore_in, OD_mm, OD_in, Width_mm, Width_in, Num_Balls, Ball_Dia, C_lbs, P_st) %>%
  kbl(caption = "Dimensions and Basic Load Ratings for Conrad-Type Single-Row Radial Ball Bearings",
      col.names = c("Bearing No.", "mm", "inch", "mm", "inch", "mm", "inch", "Z", "D (inch)", "C (lbf)", "P_st (lbf)"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), full_width = TRUE) %>%
  add_header_above(c(" " = 1, "Bore" = 2, "Outside Dia." = 2, "Width" = 2, "Balls" = 2, "Capacity" = 2)) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  column_spec(10, background = "#F9E7F0") %>%
  pack_rows("15mm Bore Series", 1, 3) %>%
  pack_rows("17mm Bore Series", 4, 6) %>%
  pack_rows("20mm Bore Series", 7, 9) %>%
  pack_rows("25mm Bore Series", 10, 12) %>%
  pack_rows("30mm Bore Series", 13, 15) %>%
  pack_rows("35mm Bore Series", 16, 18) %>%
  pack_rows("40mm Bore Series", 19, 21) %>%
  pack_rows("45mm Bore Series", 22, 24) %>%
  pack_rows("50mm Bore Series", 25, 27) %>%
  pack_rows("55mm Bore Series", 28, 30) %>%
  pack_rows("60mm Bore Series", 31, 33) %>%
  pack_rows("65mm Bore Series", 34, 36) %>%
  pack_rows("70mm Bore Series", 37, 39) %>%
  footnote(general = "C = Dynamic load rating; P_st = Static load rating. Source: ANSI-AFBMA standards.",
           general_title = "Note: ")
```

##### The Load-Life Relationship: Why It Matters {-}

The bearing life equation $L_{10} = \left(\frac{C}{P}\right)^3 \times 10^6$ reveals a powerful relationship: **bearing life varies with the cube of the load ratio**.

This has profound practical implications:

```{r load-life-relationship, echo=FALSE, message=FALSE}
## Create load-life relationship table
load_life <- tribble(
  ~Load_Change, ~Load_Ratio, ~Life_Change,
  "Double the load", "P → 2P", "L₁₀ → L₁₀/8 (life reduced to 1/8)",
  "Halve the load", "P → P/2", "L₁₀ → 8×L₁₀ (life increased 8×)",
  "Reduce load by 20%", "P → 0.8P", "L₁₀ → 1.95×L₁₀ (life nearly doubles)",
  "Increase load by 20%", "P → 1.2P", "L₁₀ → 0.58×L₁₀ (life nearly halves)"
)

load_life %>%
  kbl(caption = "Effect of Load Changes on Bearing Life (Cubic Relationship)",
      col.names = c("Change", "Load Ratio", "Effect on Life"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(2, background = "#E8F5E9", bold = TRUE) %>%
  row_spec(1, background = "#FFEBEE")
```

::: {.callout-industry}
**Design Insight:** If you can reduce the bearing load by just 20%, you nearly **double** the bearing life! Conversely, underestimating loads by 20% will cut your bearing life nearly in half. This is why accurate load analysis is critical in bearing selection.

In automotive applications, this relationship explains why bearing failures often cluster around vehicles that operate at higher loads (towing, heavy cargo) even when the bearings are "the same" as those in lightly-loaded vehicles.
:::

::: {.callout-check}
**Quick Check:** A bearing is rated for C = 10,000 N and operates at P = 2,500 N, giving L₁₀ = 64 million revolutions. If the load is reduced to P = 1,250 N (half), what is the new L₁₀ life?

*Answer: L₁₀ = (10,000/1,250)³ × 10⁶ = 8³ × 10⁶ = **512 million revolutions** (8× longer!)*
:::

---

#### How Much Life is Required? {-}

Before selecting a bearing, engineers must determine the **required design life** for the application. The table below provides industry guidelines for different machine types:

```{r recommended-life-table, echo=FALSE, message=FALSE}
## Recommended bearing life by application type
life_guidelines <- tribble(
  ~Machine_Type, ~Examples, ~Life_Hours,
  "Instruments and apparatus used infrequently", "Demonstration apparatus, devices for operation of sliding doors", "500",
  "Aircraft engines", "—", "500–2,000",
  "Machines for short duration or intermittent operation (service interruptions minor)", "Hand tools, lifting tackle in machinery shops, hand-driven machines, farm machinery, assembly cranes, charging machines, foundry cranes, household machines", "4,000–8,000",
  "Machines for intermittent service where dependable operation is important", "Auxiliary machines in power stations, conveying equipment in production lines, elevators, general-cargo cranes, machine tools used less frequently", "8,000–12,000",
  "Machines for 8-hour service not always fully utilized", "Machines in general mechanical industries, cranes for continuous service, blowers, jackshafts", "20,000–30,000",
  "Machines for continuous operation (24-hour service)", "Separators, compressors, pumps, mainline shafting, roller beds and conveyor rollers, mine hoists, stationary electric motors", "40,000–60,000",
  "Machines for 24-hour service where dependability is critical", "Pulp and paper machines, public power stations, mine pumps, public pumping stations, machines for continuous service aboard ships", "100,000–200,000"
)

life_guidelines %>%
  kbl(caption = "Recommended Bearing Life by Machine Type",
      col.names = c("Type of Machine", "Examples", "Life (hours)"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(c(4, 6), background = "#E8F5E9") %>%
  column_spec(3, width = "2.5cm", bold = TRUE)
```

::: {.callout-industry}
**Design Tip:** When in doubt, design for longer life. A bearing that costs 20% more but lasts twice as long often reduces total cost of ownership significantly—especially when considering downtime and labor costs for replacement.
:::

---

#### Two Approaches to Bearing Selection {-}

There are two common scenarios when working with bearing life equations:

**Scenario A: Verify a Known Bearing** — You have a specific bearing (dimensions known) and want to check if it meets application requirements.

**Scenario B: Select a Bearing** — You know the application requirements (load, speed, life) and need to find a bearing that meets them.

```{r bearing-selection-approaches, echo=FALSE, message=FALSE}
approaches <- tribble(
  ~Scenario, ~Known, ~Calculate, ~Use_Case,
  "A: Verify existing bearing", "Bearing dimensions (Z, D, dm), or catalog C value", "C from dimensions, then L₁₀", "Checking if a bearing in a machine meets requirements; validating catalog data",
  "B: Select new bearing", "Required life (L), speed (n), load (P)", "Required C, then select from catalog", "Designing a new machine; replacing a failed bearing with adequate life"
)

approaches %>%
  kbl(caption = "Two Approaches to Bearing Selection",
      col.names = c("Scenario", "What You Know", "What to Calculate", "When to Use"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(1, background = "#E3F2FD") %>%
  row_spec(2, background = "#FFF3E0")
```

The following examples demonstrate both approaches.

---

#### Applied Example: Calculate C from Bearing Dimensions {#example-calc-C-dimensions}

::: {.question}
Find the basic dynamic load rating $C$ for a **207 radial bearing** with contact angle $\alpha = 0°$.

From Table \@ref(tab:conrad-bearing-table), the 207 bearing has:

- Bore: 35 mm (1.378 in)
- Outside diameter: 72 mm (2.8346 in)
- Number of balls: Z = 9
- Ball diameter: D = 7/16 in = 0.4375 in
:::

::: {.answer}

```{r example-C-from-dimensions, echo=FALSE, message=FALSE}
## Given values for 207 bearing
bore_in <- 1.378          ## inches
OD_in <- 2.8346           ## inches
Z <- 9                    ## number of balls
D <- 7/16                 ## ball diameter in inches (0.4375)
alpha <- 0                ## contact angle (degrees)
i <- 1                    ## number of rows

## Step 1: Calculate pitch diameter dm
## dm = (bore + OD) / 2
dm <- (OD_in + bore_in) / 2

## Step 2: Calculate geometry ratio
## For α = 0°, cos(α) = 1
cos_alpha <- cos(alpha * pi / 180)
ratio <- (D * cos_alpha) / dm

## Step 3: Look up fc from table (interpolate between 0.20 and 0.22)
## At ratio ≈ 0.208, fc ≈ 4550 (from table, between 0.20=4550 and 0.22=4530)
fc <- 4550  ## Using inch units

## Step 4: Calculate (i·cos α)^0.7
term1 <- (i * cos_alpha)^0.7

## Step 5: Calculate Z^(2/3)
term2 <- Z^(2/3)

## Step 6: Calculate D^1.8
term3 <- D^1.8

## Step 7: Calculate C
C_calc <- fc * term1 * term2 * term3

## Catalog value for comparison
C_catalog <- 4440  ## lbf from Table 2

## Create step-by-step table
calc_steps <- tribble(
  ~Step, ~Formula, ~Calculation, ~Result,
  "Step 1: Pitch diameter", "dm = (OD + bore) / 2",
  paste0("dm = (", OD_in, " + ", bore_in, ") / 2"),
  paste0(round(dm, 4), " in"),

  "Step 2: Geometry ratio", "(D·cos α) / dm",
  paste0("(", D, " × 1) / ", round(dm, 4)),
  as.character(round(ratio, 3)),

  "Step 3: Look up fc", "From Table (ratio ≈ 0.208)",
  "Interpolate between 0.20 and 0.22",
  paste0("fc = ", fc),

  "Step 4: Row-angle term", "(i·cos α)^0.7",
  paste0("(", i, " × 1)^0.7"),
  as.character(round(term1, 3)),

  "Step 5: Ball count term", "Z^(2/3)",
  paste0(Z, "^(2/3)"),
  as.character(round(term2, 3)),

  "Step 6: Ball diameter term", "D^1.8",
  paste0(D, "^1.8"),
  as.character(round(term3, 4)),

  "Step 7: Dynamic load rating", "C = fc × (i·cos α)^0.7 × Z^(2/3) × D^1.8",
  paste0(fc, " × ", round(term1, 3), " × ", round(term2, 3), " × ", round(term3, 4)),
  paste0(round(C_calc, 0), " lbf")
)

calc_steps %>%
  kbl(caption = "Calculating C from Bearing Dimensions (207 Bearing)",
      booktabs = TRUE,
      col.names = c("Step", "Formula", "Calculation", "Result")) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(7, background = "#E8F5E9", bold = TRUE)
```

###### Verification {-}

$$C = f_c \cdot (i \cos\alpha)^{0.7} \cdot Z^{2/3} \cdot D^{1.8} = 4550 \times 1 \times 9^{2/3} \times 0.4375^{1.8} = \boxed{4440\ \text{lbf}}$$

This matches the catalog value of **C = 4,440 lbf** from Table \@ref(tab:conrad-bearing-table), confirming our calculation.

:::

---

#### Applied Example: Calculate Allowable Load P {#example-calc-P}

::: {.question}
For the **207 bearing** where we found C = 4,440 lbf, find the maximum radial load $P$ for a required L₁₀ life of **500 hours** at **1,500 RPM**.
:::

::: {.answer}

We need to rearrange the life equation to solve for P:

$$L_{10h} = \frac{10^6}{60 \cdot n}\left(\frac{C}{P}\right)^3$$

Solving for P:

$$P = \sqrt[3]{\frac{10^6 \cdot C^3}{60 \cdot n \cdot L_{10h}}}$$

```{r example-calc-P, echo=FALSE, message=FALSE}
## Given values
C <- 4440        ## lbf (from previous example)
L10h <- 500      ## hours
n <- 1500        ## RPM

## Calculate P
## P = (10^6 × C^3 / (60 × n × L))^(1/3)
numerator <- 1e6 * C^3
denominator <- 60 * n * L10h
P_calc <- (numerator / denominator)^(1/3)

## Verification: Calculate life with this P
L_verify <- (1e6 / (60 * n)) * (C / P_calc)^3

## Create calculation table
p_calc_steps <- tribble(
  ~Step, ~Formula, ~Calculation, ~Result,
  "Given", "C, L₁₀h, n",
  paste0("C = ", C, " lbf, L = ", L10h, " hrs, n = ", n, " rpm"),
  "—",

  "Rearranged formula", "P = ∛(10⁶·C³ / 60·n·L)",
  "Solve life equation for P",
  "—",

  "Numerator", "10⁶ × C³",
  paste0("10⁶ × ", C, "³ = ", formatC(numerator, format = "e", digits = 2)),
  formatC(numerator, format = "e", digits = 3),

  "Denominator", "60 × n × L",
  paste0("60 × ", n, " × ", L10h),
  formatC(denominator, format = "d", big.mark = ","),

  "Division", "Num / Denom",
  paste0(formatC(numerator, format = "e", digits = 2), " / ", formatC(denominator, format = "d", big.mark = ",")),
  formatC(numerator/denominator, format = "e", digits = 3),

  "Cube root", "P = ∛(result)",
  paste0("∛(", formatC(numerator/denominator, format = "e", digits = 2), ")"),
  paste0(round(P_calc, 0), " lbf")
)

p_calc_steps %>%
  kbl(caption = "Calculating Maximum Load P for Required Life",
      booktabs = TRUE,
      col.names = c("Step", "Formula", "Calculation", "Result")) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(6, background = "#E8F5E9", bold = TRUE)
```

###### Conclusion {-}

$$P = \sqrt[3]{\frac{10^6 \times 4440^3}{60 \times 1500 \times 500}} = \sqrt[3]{1.945 \times 10^9} = \boxed{1250\ \text{lbf}}$$

**Interpretation:** This bearing can support a radial load of **1,250 lbf** in an application where 90% of bearings will last at least 500 hours at 1,500 RPM.

:::

---

#### Applied Example: Select a Bearing for Required Life {#example-select-bearing}

::: {.question}
A bearing of the **X07 family** (35 mm bore) needs to run for **2,000 hours** at **5,000 RPM** and carry a load of **415 lbf**. Which bearing from the X07 family should be selected?
:::

::: {.answer}

We need to find the **required** dynamic load rating C, then select a bearing that meets or exceeds this value.

Rearranging the life equation to solve for C:

$$C = \sqrt[3]{\frac{60 \cdot n \cdot L_{10h} \cdot P^3}{10^6}}$$

```{r example-select-bearing, echo=FALSE, message=FALSE}
## Given values
L10h <- 2000     ## hours
n <- 5000        ## RPM
P <- 415         ## lbf

## Calculate required C
## C = (60 × n × L × P³ / 10⁶)^(1/3)
numerator <- 60 * n * L10h * P^3
denominator <- 1e6
C_required <- (numerator / denominator)^(1/3)

## Available X07 family bearings from catalog
bearings_X07 <- tribble(
  ~Bearing, ~C_lbf,
  "107", 2760,
  "207", 4440,
  "307", 5750
)

## Find suitable bearing
suitable <- bearings_X07 %>% filter(C_lbf >= C_required)

## Create calculation table
select_steps <- tribble(
  ~Step, ~Formula, ~Calculation, ~Result,
  "Given", "L, n, P",
  paste0("L = ", L10h, " hrs, n = ", n, " rpm, P = ", P, " lbf"),
  "—",

  "Rearranged formula", "C = ∛(60·n·L·P³ / 10⁶)",
  "Solve life equation for C",
  "—",

  "Numerator", "60 × n × L × P³",
  paste0("60 × ", n, " × ", L10h, " × ", P, "³"),
  formatC(numerator, format = "e", digits = 3),

  "Division", "Num / 10⁶",
  paste0(formatC(numerator, format = "e", digits = 2), " / 10⁶"),
  formatC(numerator/denominator, format = "e", digits = 3),

  "Cube root", "C_required = ∛(result)",
  paste0("∛(", formatC(numerator/denominator, format = "e", digits = 2), ")"),
  paste0(round(C_required, 0), " lbf")
)

select_steps %>%
  kbl(caption = "Calculating Required C for Bearing Selection",
      booktabs = TRUE,
      col.names = c("Step", "Formula", "Calculation", "Result")) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(5, background = "#E8F5E9", bold = TRUE)
```

###### Bearing Selection {-}

$$C_{required} = \sqrt[3]{\frac{60 \times 5000 \times 2000 \times 415^3}{10^6}} = \sqrt[3]{4.29 \times 10^{10}} = \boxed{3500\ \text{lbf}}$$

Now compare with available X07 family bearings from Table \@ref(tab:conrad-bearing-table):

```{r bearing-selection-table, echo=FALSE, message=FALSE}
selection_table <- tribble(
  ~Bearing, ~C_lbf, ~Meets_Requirement,
  "107", 2760, "No (2760 < 3500)",
  "207", 4440, "Yes (4440 > 3500) ✓",
  "307", 5750, "Yes (5750 > 3500)"
)

selection_table %>%
  kbl(caption = "X07 Family Bearing Comparison",
      col.names = c("Bearing", "C (lbf)", "Meets C ≥ 3500 lbf?"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position", full_width = FALSE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(2, background = "#E8F5E9", bold = TRUE)
```

###### Conclusion {-}

Select the **207 bearing** (C = 4,440 lbf). It exceeds the required C of 3,500 lbf while being more compact and economical than the 307.

The 107 bearing is **not suitable**—its C = 2,760 lbf is below the required 3,500 lbf, meaning it would fail before reaching the design life.

::: {.callout-check}
**Why not choose the 307?** While the 307 would work (C = 5,750 lbf), it's physically larger and more expensive. The 207 provides adequate life with a safety margin of 4,440/3,500 = 1.27, which is acceptable for most applications.
:::

:::

---

#### Adjusted Rating Life {-}

The basic L10 equation assumes ideal operating conditions. In practice, factors such as **reliability requirements**, **lubrication quality**, and **contamination** affect bearing life. The **adjusted rating life** accounts for these factors:

$$L_{na} = a_1 \cdot a_2 \cdot a_3 \cdot L_{10}$$

Or using the modern ISO 281 formulation:

$$L_{nm} = a_1 \cdot a_{ISO} \cdot L_{10}$$

**Where:**

| Factor | Description | Typical Values |
|:-------|:------------|:---------------|
| $a_1$ | Life adjustment factor for reliability | See table below |
| $a_2$ | Material factor | 0.5 to 2.0 (modern steels ≈ 1.0) |
| $a_3$ | Operating conditions factor | 0.1 to 1.0 |
| $a_{ISO}$ | Combined lubrication-contamination factor | 0.1 to 50 |

**Reliability Factor ($a_1$):**

```{r reliability-table, echo=FALSE, message=FALSE}
reliability_data <- tribble(
  ~Reliability, ~Ln, ~a1,
  "90%", "L10", 1.00,
  "95%", "L5", 0.64,
  "96%", "L4", 0.55,
  "97%", "L3", 0.47,
  "98%", "L2", 0.37,
  "99%", "L1", 0.25
)

reliability_data %>%
  kbl(caption = "Reliability Factor a₁",
      col.names = c("Reliability", "Life Symbol", "a₁"),
      booktabs = TRUE,
      align = "c") %>%
  kable_styling(latex_options = "hold_position", full_width = FALSE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(1, background = "#E8F5E9")
```

---

#### Life Calculation Plot {-}

The following plot illustrates how bearing life varies with the C/P ratio:

```{r bearing-life-plot, echo=FALSE, fig.width=10, fig.height=6, fig.align='center', fig.cap="Bearing life as a function of load ratio (C/P)."}
## Calculate life for various C/P ratios
cp_ratio <- seq(1, 10, by = 0.1)
L10_ball <- (cp_ratio)^3 * 1e6 / 1e6  ## in millions of revolutions
L10_roller <- (cp_ratio)^(10/3) * 1e6 / 1e6

life_data <- data.frame(
  CP_Ratio = rep(cp_ratio, 2),
  Life_Millions = c(L10_ball, L10_roller),
  Type = rep(c("Ball Bearing (p=3)", "Roller Bearing (p=10/3)"), each = length(cp_ratio))
)

ggplot(life_data, aes(x = CP_Ratio, y = Life_Millions, color = Type, linetype = Type)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = 3, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 3.2, y = 800, label = "C/P = 3\nDesign target", hjust = 0, size = 3) +
  geom_hline(yintercept = 27, linetype = "dotted", alpha = 0.5) +
  annotate("text", x = 8, y = 35, label = "~27M rev @ C/P=3", size = 3) +
  scale_y_log10(labels = scales::comma) +
  scale_color_manual(values = c("steelblue", "darkorange")) +
  labs(
    title = "Ball Bearing Life vs. Load Ratio",
    subtitle = "L10 = (C/P)^p × 10⁶ revolutions",
    x = "Load Ratio (C/P)",
    y = "Life L₁₀ (millions of revolutions)",
    color = "Bearing Type",
    linetype = "Bearing Type"
  ) +
  theme_bearing +
  theme(legend.position = "bottom")
```

> **Design Insight:** A small increase in bearing size (higher C) or reduction in load (lower P) dramatically increases bearing life due to the cubic relationship.

---

### Applied Example: Ball Bearing Life Calculation {#bearing-example}

::: {.question}

A deep groove ball bearing (6207) supports a shaft rotating at **1,750 RPM**. The bearing experiences a **radial load of 3,500 N** and an **axial load of 800 N**. Determine:

1. The equivalent dynamic load ($P$)
2. The basic L10 life in hours
3. The L5 life (95% reliability) in hours

**Given data from catalog:**
- Bearing: 6207
- Dynamic load rating: $C = 26.0$ kN = 26,000 N
- For Fa/Fr ratio, use: $e = 0.24$, $X = 0.56$, $Y = 1.6$
:::

---

::: {.answer}


```{r bearing-example-calc, echo=FALSE, message=FALSE}
## Given values
Fr <- 3500      ## N, radial load
Fa <- 800       ## N, axial load
C <- 26000      ## N, dynamic load rating
n <- 1750       ## RPM
p <- 3          ## ball bearing exponent

## Check Fa/Fr ratio
ratio <- Fa / Fr

## Since Fa/Fr = 0.229 < e = 0.24, use X=1, Y=0
## But let's demonstrate the general case where Fa/Fr > e
X <- 0.56
Y <- 1.6

## Equivalent dynamic load
P_light <- Fr  ## When Fa/Fr <= e
P_heavy <- X * Fr + Y * Fa  ## When Fa/Fr > e

## For this example, Fa/Fr ≈ 0.23 which is close to e
## We'll use the combined load formula for demonstration
P <- max(P_light, P_heavy)

## L10 life in revolutions
L10_rev <- (C / P)^p * 1e6

## L10 life in hours
L10_hours <- L10_rev / (60 * n)

## L5 life (95% reliability)
a1_95 <- 0.64
L5_hours <- a1_95 * L10_hours

## Create results table
results <- tribble(
  ~Step, ~Calculation, ~Result,
  "Step 1: Check load ratio", paste0("Fa/Fr = ", Fa, " / ", Fr, " = ", round(ratio, 3)),
  ifelse(ratio <= 0.24, "≤ e: Use X=1, Y=0", "> e: Use X=0.56, Y=1.6"),
  "Step 2: Equivalent load (P)", paste0("P = X·Fr + Y·Fa = ", X, "×", Fr, " + ", Y, "×", Fa),
  paste0(round(P, 0), " N"),
  "Step 3: L10 (revolutions)", paste0("L10 = (C/P)³ × 10⁶ = (", C/1000, "/", round(P/1000, 2), ")³ × 10⁶"),
  paste0(formatC(L10_rev/1e6, format = "f", digits = 1), " million rev"),
  "Step 4: L10 (hours)", paste0("L10h = L10 / (60 × n) = ", formatC(L10_rev/1e6, format = "f", digits = 1), "×10⁶ / (60 × ", n, ")"),
  paste0(formatC(L10_hours, format = "f", digits = 0, big.mark = ","), " hours"),
  "Step 5: L5 (95% reliability)", paste0("L5 = a₁ × L10 = 0.64 × ", formatC(L10_hours, format = "f", digits = 0)),
  paste0(formatC(L5_hours, format = "f", digits = 0, big.mark = ","), " hours")
)

results %>%
  kbl(caption = "Ball Bearing Life Calculation Steps",
      booktabs = TRUE,
      col.names = c("Step", "Calculation", "Result")) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(4, background = "#E8F5E9", bold = TRUE) %>%
  row_spec(5, background = "#E3F2FD", bold = TRUE)
```


###### Interpretation {-}

- The **L10 life** of `r formatC(L10_hours, format = "f", digits = 0, big.mark = ",")` hours means that 90% of these bearings will operate at least this long before fatigue failure.
- For applications requiring **95% reliability**, use the L5 life of `r formatC(L5_hours, format = "f", digits = 0, big.mark = ",")` hours.
- At 24/7 operation, L10 represents approximately `r round(L10_hours / (24*365), 1)` years of service.

:::

---

#### External Resources: Bearing Calculation Tools {-}

Major bearing manufacturers provide comprehensive technical documentation and online calculators for bearing selection. These resources include detailed load rating tables, application examples, and life calculation methodologies.

::: {.callout-industry}
**Manufacturer Resources:**

- **NTN Bearing Corporation** — [Technical Calculations and Bearing Data (PDF)](https://ntn.ca/wp-content/uploads/2022/02/2203E_Web.pdf)
  - Comprehensive bearing catalog with life calculations, load ratings, and application guidelines
  - Includes worked examples for various bearing types

- **NSK Ltd.** — [Rolling Bearing Catalog: Technical Section (PDF)](https://www.nsk.com/content/dam/nsk/common/catalogs/ctrgPdf/bearings/e728.pdf)
  - Detailed technical information on bearing selection and calculations
  - Includes adjusted life calculations with lubrication and contamination factors

These resources are invaluable for real-world bearing selection beyond the simplified examples in this chapter.
:::

---

### Plain Bearings (Bushings) {#bushings}

Plain bearings, also called **bushings** or **sleeve bearings**, are the simplest form of bearing. They consist of a bearing surface with no rolling elements—the shaft slides directly against the bearing material.

::: {.callout-think}
**Before You Read:** When a shaft rotates inside a bushing, what prevents the two metal surfaces from grinding together and wearing out rapidly? Is there something more sophisticated happening than just "oil between two surfaces"?
:::

#### Advantages of Bushings {-}

| Advantage | Explanation |
|:----------|:------------|
| **Compact design** | No need for rolling elements; minimal radial space |
| **Low cost** | Simple manufacturing; no precision balls or races |
| **Quiet operation** | No rolling elements to generate noise |
| **High load capacity** | Large contact area; excellent for shock loads |
| **Self-lubricating options** | Polymer and sintered metal bushings need no external lubrication |

Many of the most successful bushing designs use **sintered metals**—porous metal structures created through powder metallurgy. These porous bushings can store lubricant within their structure and release it during operation, creating truly maintenance-free bearings. For a detailed understanding of how sintered bearings are manufactured and how their porosity is controlled, see Chapter \@ref(chap-sintering).

#### Types of Bushings {-}

```{r bushing-types-table, echo=FALSE, message=FALSE}
bushing_types <- tribble(
  ~Type, ~Material, ~PV_Limit, ~Applications, ~Lubrication,
  "Bronze Bushings", "Cast bronze, SAE 660", "50,000 - 75,000 psi·ft/min", "Heavy machinery, pumps", "Oil/grease required",
  "Oil-Impregnated Bronze", "Sintered bronze (porous)*", "25,000 - 50,000 psi·ft/min", "Fractional HP motors, fans", "Self-lubricating",
  "PTFE-Lined", "Steel + PTFE coating", "10,000 - 25,000 psi·ft/min", "Food processing, medical", "Dry running",
  "Polymer (Nylon, Acetal)", "Engineering plastics", "3,000 - 10,000 psi·ft/min", "Light duty, instrumentation", "Often dry",
  "PEEK Composite", "High-performance polymer", "15,000 - 40,000 psi·ft/min", "Aerospace, high temperature", "Dry or minimal"
)

bushing_types %>%
  kbl(caption = "Common Bushing Materials and PV Limits",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), full_width = TRUE) %>%
  column_spec(3, background = "#F9E7F0") %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  footnote(symbol = c("See Chapter 7 (Powder Metallurgy and Sintering) for manufacturing details of sintered bearings"))
```

---

---

### Hydrodynamic Lubrication and the Oil Cushion {#sec-hydrodynamic-lubrication}

One of the most remarkable phenomena in bearing engineering is **hydrodynamic lubrication**—the ability of a rotating shaft to "float" on a thin cushion of oil, with complete separation between the metal surfaces. Understanding this mechanism is essential for designing reliable bushing applications.

#### The Wedge Effect {#sec-wedge-effect}

When a shaft begins to rotate inside a lubricated bushing, something remarkable happens. The shaft doesn't rotate concentrically within the bore—instead, it shifts slightly off-center, creating a **converging wedge** of oil between the shaft and bearing surfaces.

```{r fig-hydrodynamic-wedge, echo=FALSE, fig.cap="The hydrodynamic wedge: shaft rotation drags oil into the converging gap, building pressure that lifts the shaft.", fig.align='center', fig.width=10, fig.height=6}
library(ggplot2)
library(ggforce)

# Create hydrodynamic wedge illustration
ggplot() +
  # Bearing bore (outer circle)
  geom_circle(aes(x0 = 5, y0 = 5, r = 3.5), color = "gray40", fill = "tan", size = 1.5) +
  geom_circle(aes(x0 = 5, y0 = 5, r = 3), color = "gray40", fill = "white", size = 1) +

  # Shaft (offset - the key to hydrodynamic action)
  # At rest: shaft sits at bottom. Running: shaft climbs up and to the side
  geom_circle(aes(x0 = 5.3, y0 = 5.4, r = 2.7), color = "black", fill = "gray50", size = 1) +

  # Minimum film thickness point
  annotate("point", x = 6.8, y = 6.5, color = "red", size = 3) +
  annotate("segment", x = 6.8, y = 6.5, xend = 8.2, yend = 7.2, color = "red", linewidth = 0.8) +
  annotate("text", x = 8.5, y = 7.4, label = "Minimum film\nthickness (h_min)",
           color = "red", size = 3, hjust = 0) +

  # Maximum film thickness (opposite side)
  annotate("segment", x = 3.2, y = 3.8, xend = 1.8, yend = 3, color = "blue", linewidth = 0.8) +
  annotate("text", x = 1.5, y = 2.7, label = "Maximum film\nthickness",
           color = "blue", size = 3, hjust = 1) +

  # Converging wedge zone (highlighted)
  annotate("curve", x = 7.5, y = 4.5, xend = 5.5, yend = 2.3, curvature = 0.3,
           color = "darkorange", linewidth = 2, lineend = "round") +
  annotate("text", x = 7, y = 3.2, label = "Converging\nwedge zone",
           color = "darkorange", size = 3, fontface = "bold") +

  # Rotation arrow on shaft
  geom_curve(aes(x = 5.3, xend = 6.2, y = 5.4, yend = 6), curvature = -0.5,
             arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
             linewidth = 1.2, color = "black") +
  annotate("text", x = 6.5, y = 5.5, label = "ω", size = 5, fontface = "bold") +

  # Oil flow arrows (dragged by shaft rotation)
  geom_curve(aes(x = 7, y = 3.5, xend = 7.3, yend = 4.5), curvature = -0.2,
             arrow = arrow(length = unit(0.15, "cm")), color = "gold3", linewidth = 1) +
  geom_curve(aes(x = 7.4, y = 4.8, xend = 7.2, yend = 5.8), curvature = -0.2,
             arrow = arrow(length = unit(0.15, "cm")), color = "gold3", linewidth = 1) +
  annotate("text", x = 8, y = 4.2, label = "Oil dragged\ninto wedge",
           color = "gold3", size = 2.8, fontface = "italic") +

  # Pressure distribution (shown as arrows pushing shaft)
  geom_segment(aes(x = 4.2, xend = 4.7, y = 2.6, yend = 3.4),
               arrow = arrow(length = unit(0.15, "cm")), color = "purple", linewidth = 1.2) +
  geom_segment(aes(x = 5, xend = 5.3, y = 2.4, yend = 3.2),
               arrow = arrow(length = unit(0.15, "cm")), color = "purple", linewidth = 1.5) +
  geom_segment(aes(x = 5.8, xend = 5.8, y = 2.5, yend = 3.3),
               arrow = arrow(length = unit(0.15, "cm")), color = "purple", linewidth = 1.2) +
  geom_segment(aes(x = 6.5, xend = 6.2, y = 2.8, yend = 3.5),
               arrow = arrow(length = unit(0.15, "cm")), color = "purple", linewidth = 1) +

  annotate("text", x = 5, y = 1.8, label = "Pressure buildup\nlifts shaft",
           color = "purple", size = 3, fontface = "bold") +

  # Load arrow
  geom_segment(aes(x = 5.3, xend = 5.3, y = 8, yend = 7),
               arrow = arrow(length = unit(0.25, "cm")), color = "darkred", linewidth = 1.5) +
  annotate("text", x = 5.3, y = 8.3, label = "W (Load)", color = "darkred", size = 3.5, fontface = "bold") +

  # Center markers
  annotate("point", x = 5, y = 5, shape = 3, size = 3, stroke = 1.5, color = "gray40") +
  annotate("text", x = 4.5, y = 5, label = "Bearing\ncenter", size = 2.5, hjust = 1) +
  annotate("point", x = 5.3, y = 5.4, shape = 3, size = 3, stroke = 1.5, color = "black") +
  annotate("text", x = 4.8, y = 5.8, label = "Shaft\ncenter", size = 2.5, hjust = 1) +

  # Eccentricity line
  geom_segment(aes(x = 5, xend = 5.3, y = 5, yend = 5.4),
               linetype = "dashed", color = "gray30", linewidth = 0.8) +
  annotate("text", x = 5.4, y = 5.15, label = "e", size = 4, fontface = "italic") +

  coord_fixed(xlim = c(0, 10), ylim = c(1, 9)) +
  theme_void() +
  labs(title = "Hydrodynamic Wedge Effect in Journal Bearings",
       subtitle = "Shaft rotation creates pressure that lifts the shaft off the bearing surface") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 10, color = "gray40")
  )
```

The physics of this process are elegant:

1. **Shaft rotation drags oil** into the narrowing gap (viscous drag)
2. **Oil cannot escape fast enough** through the narrow exit
3. **Pressure builds up** in the converging zone
4. **The pressure "pushes back"** against the shaft, lifting it
5. **At sufficient speed**, the shaft floats entirely on oil—no metal contact

This is the **hydrodynamic cushion**: a self-generated oil film that supports the shaft load through fluid pressure alone.

#### The Pressure Distribution {#sec-pressure-distribution}

The pressure profile that develops in the oil film is not uniform. Maximum pressure occurs in the converging wedge region, just before the minimum film thickness point:

\begin{equation}
p = \frac{6 \mu U}{h^2} \frac{dh}{dx}
(\#eq:reynolds-simplified)
\end{equation}

**Where:**

| Symbol | Description |
|:-------|:------------|
| $p$ | Local pressure in the oil film |
| $\mu$ | Dynamic viscosity of the lubricant |
| $U$ | Surface velocity of the shaft |
| $h$ | Local film thickness |
| $dh/dx$ | Rate of change of film thickness |

This is a simplified form of the **Reynolds equation**, which governs hydrodynamic lubrication. The key insight is that pressure only develops where the film is **converging** ($dh/dx < 0$)—in diverging regions, the pressure drops back to ambient.

::: {.callout-industry}
**Industry Application:** Modern automotive engines rely entirely on hydrodynamic lubrication for their crankshaft main bearings and connecting rod bearings. At idle speed (~800 RPM), a typical crankshaft main bearing develops oil film pressures of 5-10 MPa, supporting loads of several kilonewtons on a film just 2-5 micrometers thick. At highway speeds, this film pressure can exceed 50 MPa.
:::

#### Eccentricity Ratio {#sec-eccentricity}

The position of the shaft within the bearing is characterized by the **eccentricity ratio** ($\varepsilon$):

\begin{equation}
\varepsilon = \frac{e}{c}
(\#eq:eccentricity-ratio)
\end{equation}

**Where:**

- $e$ = distance between bearing center and shaft center (eccentricity)
- $c$ = radial clearance = $(D - d)/2$ (bearing bore radius minus shaft radius)

| $\varepsilon$ Value | Meaning |
|:--------------------|:--------|
| 0 | Shaft is perfectly centered (no load, or very high speed) |
| 0.3 - 0.7 | Normal operating range for hydrodynamic bearings |
| → 1.0 | Shaft approaching contact with bearing (high load or low speed) |

The minimum film thickness is directly related to eccentricity:

\begin{equation}
h_{min} = c(1 - \varepsilon)
(\#eq:min-film-thickness)
\end{equation}

For safe operation, the minimum film thickness must exceed the combined surface roughness of the shaft and bearing, plus a safety margin for contaminants and vibration.

---

### Oil Circulation in Sintered Bearings {#sec-oil-circulation}

For sintered oil-impregnated bushings (see Chapter \@ref(chap-sintering) for manufacturing details), the hydrodynamic mechanism works in concert with the porous structure. The following diagram shows how oil circulates through a sintered bearing during operation:

```{r fig-oil-circulation, echo=FALSE, message=FALSE, fig.cap="Oil circulation mechanism in a sintered oil-impregnated bearing. The porous structure stores oil that is released during operation and recovered during rest periods.", fig.align='center', fig.width=10, fig.height=8}
library(ggplot2)
library(ggforce) # Essential for circles and smooth shapes

# --- Define Geometry and Colors ---
# Centers and Radii
c_bearing <- c(0, 0)
r_housing <- 5.5
r_bearing_out <- 5.0
r_bearing_in <- 3.1
# Crucial: Shaft is OFFSET upwards and left to create the wedge/cushion effect
c_shaft <- c(-0.2, 0.7) 
r_shaft <- 2.8

# Colors based on the diagram
col_housing <- "grey70"
col_porous <- "#AEEEEE" # Pale turquoise for sintered bearing
col_porous_outline <- "#5F9EA0"
col_shaft <- "#FF9933"  # Orange
col_oil_film <- "#0066CC" # Dark blue thick film
col_pressure <- "#87CEFA" # Light blue for pressure distribution blob
col_arrow <- "black"
col_text <- "#003366" # Dark blue for labels

# --- Generate Data for Irregular Shapes ---

# 1. Pressure Distribution 'Blob' (Teardrop shape)
# Using a b-spline to create a smooth, irregular pressure zone that is
# wider at the bottom-left (high pressure) and narrows around the top.
pressure_pts <- data.frame(
  x = c(0,  2.5, 3.5,  2,   -1, -4, -6, -6.5, -4, -1),
  y = c(3.5, 3,   0,   -3,  -5, -6, -5, -2,    2,  3.5)
)

# --- Plotting ---
ggplot() +
  coord_fixed() + # Ensure circles look like circles
  
  # --- LAYER 1: The Pressure "Cushion" ---
  # This goes first so it sits behind the bearing structure
  ggforce::geom_bspline_closed(data = pressure_pts, aes(x=x, y=y), 
                               fill = col_pressure, alpha = 0.6, color = NA) +
  
  # Pressure MAX dashed line and label
  annotate("segment", x = c_bearing[1], y = c_bearing[2], xend = -5.8, yend = -4.8,
           linetype = "dashed", color = "black") +
  annotate("text", x = -5.2, y = -5.2, label = "MAX", fontface = "bold", size = 3.5) +
  
  # --- LAYER 2: Bearing Structure ---
  # Housing (Outer Ring)
  geom_circle(aes(x0 = c_bearing[1], y0 = c_bearing[2], r = r_housing), 
              fill = NA, color = col_housing, size = 2) +
  
  # Sintered Porous Bearing (Middle Ring)
  # We draw the outer boundary, then fill the center with white to create the "donut"
  geom_circle(aes(x0 = c_bearing[1], y0 = c_bearing[2], r = r_bearing_out), 
              fill = col_porous, color = col_porous_outline, alpha = 0.4, size = 0.8) +
  # The "hole" in the bearing (white to obscure the pressure blob behind the shaft area)
  geom_circle(aes(x0 = c_bearing[1], y0 = c_bearing[2], r = r_bearing_in), 
              fill = "white", color = col_porous_outline, size = 0.8) +
  
  # --- LAYER 3: The Shaft and Oil Film ---
  # The thick Oil Film (Crescent right under the shaft)
  annotate("curve", 
           x = c_shaft[1] - r_shaft*0.5, y = c_shaft[2] - r_shaft*0.85,
           xend = c_shaft[1] + r_shaft*0.7, yend = c_shaft[2] - r_shaft*0.7,
           curvature = 0.3, color = col_oil_film, size = 4, lineend = "round") +
  
  # The Shaft itself (Orange circle, offset center)
  geom_circle(aes(x0 = c_shaft[1], y0 = c_shaft[2], r = r_shaft), 
              fill = col_shaft, color = "black") +
  
  # Shaft Axis Marker (+)
  annotate("segment", x = c_shaft[1]-0.3, xend = c_shaft[1]+0.3, y = c_shaft[2], yend = c_shaft[2]) +
  annotate("segment", x = c_shaft[1], xend = c_shaft[1], y = c_shaft[2]-0.3, yend = c_shaft[2]+0.3) +
  
  # --- Arrows indicating movement and flow ---
  # 1. Shaft Rotation Arrow (Clockwise)
  annotate("curve", x = c_shaft[1]+1.8, y = c_shaft[2]+0.5, 
           xend = c_shaft[1]+1.8, yend = c_shaft[2]-1.2,
           curvature = -0.5, arrow = arrow(length = unit(0.4, "cm"), type="closed"), 
           size = 1.2, color = col_arrow) +
  
  # 2. Oil Supply (Inward pump action at top)
  annotate("curve", x = -0.5, y = 4.5, xend = 0.5, yend = 3.3,
           curvature = -0.2, arrow = arrow(length = unit(0.2, "cm"), type="closed"), size=0.8) +
  annotate("curve", x = -2.5, y = 3.8, xend = -1.5, yend = 2.8,
           curvature = -0.2, arrow = arrow(length = unit(0.2, "cm"), type="closed"), size=0.8) +
  
  # 3. Oil Escape (Outward into bearing at high pressure zone)
  annotate("segment", x = -1, y = -3.3, xend = -1.8, yend = -4.2,
           arrow = arrow(length = unit(0.2, "cm"), type="closed"), size=0.8) +
  annotate("segment", x = 1, y = -3.3, xend = 1.8, yend = -4.2,
           arrow = arrow(length = unit(0.2, "cm"), type="closed"), size=0.8) +
  
  # --- Text Labels ---
  # Main Title
  ggtitle("Schematic illustration of oil circulation mechanism\nin sintered oil-impregnated bearing") +
  
  # Component Labels
  annotate("text", x=6, y=6, label="Sintered oil-impregnated\nbearing", color=col_text, hjust=0, size=4) +
  annotate("segment", x=6, y=6, xend=4.5, yend=4.5, color=col_housing) +
  
  annotate("text", x=7, y=4, label="Housing", color=col_text, hjust=0, size=4) +
  annotate("segment", x=7, y=4, xend=5.6, yend=2.5, color=col_housing) +
  
  annotate("text", x=0, y=0.5, label="Axis\n(shaft)", fontface="bold", size=3.5) +
  
  # Action Labels
  annotate("text", x=-7, y=6, label="Supply of oil from top of inner surface\nof bearing by pump-like action", 
           color=col_text, hjust=0, size=4) +
  annotate("segment", x=-4, y=5.5, xend=-1, yend=4.2, color=col_housing) +
  
  annotate("text", x=7, y=-2, label="Supply of oil to\nsliding surface by\nrotation of shaft", 
           color=col_text, hjust=0, size=4) +
  annotate("segment", x=7, y=-2, xend=3.5, yend=-1, color=col_housing) +
  
  annotate("text", x=-7, y=-2, label="Pressure distribution\nof oil", 
           color=col_text, hjust=1, size=4) +
  annotate("segment", x=-7.1, y=-2, xend=-6.2, yend=-4, color=col_housing) +
  
  annotate("text", x=3, y=-6.5, label="Escape of oil from sliding surface into interior of bearing", 
           color=col_text, hjust=0.5, size=4) +
  annotate("segment", x=0, y=-6.2, xend=0, yend=-4.5, color=col_housing) +
  
  annotate("text", x=4.5, y=-4.5, label="Oil film", color=col_text, hjust=0, size=4) +
  annotate("segment", x=4.4, y=-4.5, xend=1, yend=-3.2, color=col_housing) +
  
  # Theme Cleanup
  theme_void() + # Remove axes and grids
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold", color = "white", 
                              margin=margin(b=10), debug=FALSE),
    plot.background = element_rect(fill = "white", color = NA),
    # Simulate the dark blue title header bar
    plot.margin = margin(t=40, r=10, b=10, l=10)
  ) +
  # Add the blue title bar rectangle behind the title text
  annotation_custom(grid::rectGrob(gp=grid::gpar(fill="#003399", col=NA)), 
                    xmin=-Inf, xmax=Inf, ymin=7.2, ymax=9)

```

---

### The Stribeck Curve: Understanding Lubrication Regimes {#sec-stribeck-curve}

The transition from metal-to-metal contact to full hydrodynamic separation doesn't happen instantly. German engineer Richard Stribeck (1861-1950) systematically studied this transition and developed one of the most important diagrams in tribology—the **Stribeck curve**.

```{r fig-stribeck-curve, echo=FALSE, message=FALSE, fig.cap="The Stribeck curve showing the three lubrication regimes as a function of the bearing parameter.", fig.align='center', fig.width=10, fig.height=7}
library(ggplot2)

# Create the data for the curve using the Hersey number (dimensionless bearing parameter)
# Hersey number = (viscosity × speed) / pressure
hersey <- seq(0.001, 2, length.out = 300)

# Stribeck behavior: high at low Hersey, drops to minimum, then rises linearly
# f = a/H + b*H + c (simplified model)
friction <- 0.12 * exp(-8 * hersey) + 0.008 * hersey + 0.015

df <- data.frame(Hersey = hersey, Friction = friction)

# Find the minimum point for annotation
min_idx <- which.min(friction)
min_hersey <- hersey[min_idx]
min_friction <- friction[min_idx]

ggplot(df, aes(x = Hersey, y = Friction)) +
  # Background regions
  annotate("rect", xmin = 0, xmax = 0.15, ymin = 0, ymax = 0.14,
           fill = "coral", alpha = 0.2) +
  annotate("rect", xmin = 0.15, xmax = 0.5, ymin = 0, ymax = 0.14,
           fill = "lightyellow", alpha = 0.4) +
  annotate("rect", xmin = 0.5, xmax = 2, ymin = 0, ymax = 0.14,
           fill = "lightgreen", alpha = 0.3) +

  # The main Stribeck curve
  geom_line(color = "#D93025", linewidth = 1.5) +

  # Vertical boundary lines
  geom_vline(xintercept = c(0.15, 0.5), linetype = "dashed", color = "gray40", linewidth = 0.8) +

  # Zone labels at top
  annotate("text", x = 0.075, y = 0.13, label = "BOUNDARY", fontface = "bold", size = 4, color = "darkred") +
  annotate("text", x = 0.325, y = 0.13, label = "MIXED", fontface = "bold", size = 4, color = "darkorange4") +
  annotate("text", x = 1.25, y = 0.13, label = "HYDRODYNAMIC", fontface = "bold", size = 4, color = "darkgreen") +

  # Descriptive text in each zone
  annotate("text", x = 0.075, y = 0.08,
           label = "Metal-to-\nmetal\ncontact",
           size = 3, color = "gray30") +
  annotate("text", x = 0.325, y = 0.055,
           label = "Partial\nfilm +\nasperities",
           size = 3, color = "gray30") +
  annotate("text", x = 1.25, y = 0.06,
           label = "Full fluid film\n(no contact)",
           size = 3, color = "gray30") +

  # Minimum friction point
  geom_point(aes(x = min_hersey, y = min_friction), color = "blue", size = 4) +
  annotate("segment", x = min_hersey, xend = min_hersey + 0.2, y = min_friction, yend = min_friction + 0.02,
           color = "blue", linewidth = 0.8) +
  annotate("text", x = min_hersey + 0.25, y = min_friction + 0.025,
           label = "Optimal\noperating point",
           color = "blue", size = 3.5, hjust = 0) +

  # Axis formatting
  scale_x_continuous(expand = c(0, 0), limits = c(0, 2.1),
                     name = expression(paste("Bearing Parameter  ", eta, "N/P  (viscosity × speed / pressure)"))) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 0.145),
                     name = "Coefficient of Friction") +

  labs(title = "The Stribeck Curve",
       subtitle = "Friction coefficient vs. bearing operating parameter") +

  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 0.5, size = 10, color = "gray40"),
    axis.title = element_text(size = 11),
    panel.grid.minor = element_blank()
  )
```

#### The Three Lubrication Regimes {#sec-lubrication-regimes}

The Stribeck curve reveals three distinct operating regimes, determined by the **bearing parameter** (also called the Hersey number or Sommerfeld variable):

\begin{equation}
\text{Bearing Parameter} = \frac{\mu N}{P}
(\#eq:hersey-number)
\end{equation}

**Where:**

| Symbol | Description |
|:-------|:------------|
| $\mu$ | Dynamic viscosity of the lubricant |
| $N$ | Rotational speed (or sliding velocity) |
| $P$ | Bearing pressure (load per unit projected area) |

---

**1. Boundary Lubrication** (Low bearing parameter)

At low speeds, high loads, or low viscosity, the oil film is too thin to separate the surfaces. The shaft and bearing make direct metal-to-metal contact at their surface asperities (microscopic peaks).

- **Friction coefficient:** 0.08 - 0.15 (high)
- **Wear:** Significant—asperities deform and break off
- **Protection:** Depends on boundary additives in the oil (anti-wear, extreme pressure)
- **Typical conditions:** Startup, shutdown, slow oscillating motion

::: {.callout-check}
**Quick Check:** Why is startup the most damaging period for a hydrodynamic bearing?
:::

---

**2. Mixed Lubrication** (Intermediate bearing parameter)

As speed increases or load decreases, a partial fluid film develops. The load is shared between the oil film (hydrodynamic pressure) and the remaining asperity contacts.

- **Friction coefficient:** 0.02 - 0.08 (decreasing rapidly)
- **Wear:** Moderate—some contact remains
- **Transition zone:** Most sensitive to operating conditions
- **Typical conditions:** Acceleration/deceleration, variable loads

---

**3. Hydrodynamic Lubrication** (High bearing parameter)

At sufficient speed, the shaft "lifts off" completely and floats on the oil cushion. There is no metal-to-metal contact—all load is carried by fluid pressure.

- **Friction coefficient:** 0.001 - 0.01 (very low)
- **Wear:** Essentially zero (only contamination-related)
- **Friction source:** Viscous shearing of the oil film
- **Typical conditions:** Continuous rotation at design speed

Notice that friction actually **increases** slightly at very high bearing parameters. This is because the viscous drag in the oil film increases with speed—but this "friction" doesn't cause wear since no contact occurs.

---

#### Practical Implications for Bearing Design {#sec-stribeck-implications}

The Stribeck curve has profound implications for bearing selection and design:

| Design Goal | Implication |
|:------------|:------------|
| **Maximize bearing life** | Operate in the hydrodynamic regime; avoid boundary conditions |
| **Minimum speed requirement** | Must exceed the transition speed for hydrodynamic lubrication |
| **Oil selection** | Higher viscosity → lower bearing parameter → earlier transition to hydrodynamic |
| **Clearance design** | Affects film thickness and transition point |
| **Startup/shutdown wear** | Unavoidable boundary contact; limit start-stop cycles or use hydrostatic assist |

::: {.callout-industry}
**Industry Application:** Large turbine bearings in power plants use **hydrostatic jacking**—externally pressurized oil is pumped into the bearing at startup to lift the shaft before rotation begins. This eliminates the boundary lubrication phase entirely, preventing wear during the 10-15 seconds it takes to reach hydrodynamic speed.
:::

For sintered self-lubricating bearings (see Chapter \@ref(chap-sintering)), the oil stored in the porous structure is released by heat and pumping action during operation. However, at very low speeds or oscillating motion, hydrodynamic films cannot develop—this is why **graphite-impregnated sintered bearings** are used instead for such applications, providing solid lubrication rather than relying on fluid film formation.

---

### Oil Circulation in Sintered Bearings (continued) {#sec-oil-circulation-diagram}

The following diagram illustrates how oil circulates through a sintered oil-impregnated bearing during operation. The porous structure acts as an oil reservoir, continuously replenishing the hydrodynamic film:

```{r bushing-striebeck-curve-old, echo=FALSE, message=FALSE, include=FALSE}
# This old simple curve is replaced by the more detailed version above
# Keeping for reference but not displaying
library(ggplot2)
library(ggforce) # For the ellipses/ovals

# 1. Create the data for the curve
# Using a function to simulate the Stribeck behavior:
# Friction = a * exp(-b * speed) + c * speed + d
speed <- seq(0, 100, length.out = 300)
friction <- 0.15 * exp(-0.15 * speed) + 0.0005 * speed + 0.02

df <- data.frame(Speed = speed, Friction = friction)

# 2. Build the plot
ggplot(df, aes(x = Speed, y = Friction)) +
  # The main Stribeck curve (Red line)
  geom_line(color = "#D93025", size = 1.2) +

  # Vertical dashed boundary lines
  geom_vline(xintercept = c(10, 40), linetype = "dotted", color = "black") +

  # Text Labels for Lubrication Zones
  annotate("text", x = 5, y = 0.18, label = "Boundary Lubrication", angle = 90, size = 4) +
  annotate("text", x = 25, y = 0.18, label = "Mixed Lubrication", angle = 90, size = 4) +
  annotate("text", x = 70, y = 0.15, label = "Hydrodynamic lubrication", size = 5) +



  # Stylized Arrows for the axes
  theme_classic() +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 110)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 0.22)) +
  labs(x = "Speed [rpm]", y = "Friction Coefficient") +
  theme(
    axis.line = element_line(arrow = arrow(type = "closed", length = unit(0.15, "inches"))),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_text(size = 14, face = "plain", hjust = 1),
    plot.margin = margin(20, 20, 20, 20)
  )
```


### PV Limits for Bushings {#pv-limits}

The **PV limit** is the most important design criterion for plain bearings. It represents the maximum combination of **bearing pressure** (P) and **sliding velocity** (V) that the bearing material can sustain without excessive wear or failure.

#### The PV Concept {-}

$$PV = P \times V$$

**Where:**

| Symbol | Description | Typical Units |
|:-------|:------------|:--------------|
| $P$ | Bearing pressure | psi (or MPa) |
| $V$ | Sliding velocity | ft/min (or m/s) |
| $PV$ | PV factor | psi·ft/min (or MPa·m/s) |

#### Calculating Bearing Pressure (P) {-}

For a journal bearing (shaft in a bushing):

$$P = \frac{F}{d \times L}$$

**Where:**

| Symbol | Description |
|:-------|:------------|
| $F$ | Radial load on the bearing (lbf or N) |
| $d$ | Shaft (journal) diameter (in or mm) |
| $L$ | Bearing length (in or mm) |

> **Note:** This is the **projected area** method, using $A = d \times L$, not the cylindrical surface area.

#### Calculating Sliding Velocity (V) {-}

$$V = \pi \times d \times n$$

For consistent units:

**English:**
$$V = \frac{\pi \times d \times n}{12} \quad \text{(ft/min, where } d \text{ in inches, } n \text{ in RPM)}$$

**Metric:**
$$V = \frac{\pi \times d \times n}{60,000} \quad \text{(m/s, where } d \text{ in mm, } n \text{ in RPM)}$$

---

#### PV Limit Visualization {-}

```{r pv-limit-plot, echo=FALSE, fig.width=10, fig.height=7, fig.align='center', fig.cap="PV limit diagram showing safe operating region for different bushing materials."}
## Create PV limit curves
P_values <- seq(100, 5000, by = 50)

## PV limits for different materials (psi·ft/min)
pv_bronze <- 50000
pv_sintered <- 35000
pv_ptfe <- 15000
pv_nylon <- 5000

pv_data <- data.frame(
  P = rep(P_values, 4),
  V = c(pv_bronze/P_values, pv_sintered/P_values, pv_ptfe/P_values, pv_nylon/P_values),
  Material = rep(c("Bronze (PV=50,000)", "Sintered Bronze (PV=35,000)",
                   "PTFE-Lined (PV=15,000)", "Nylon (PV=5,000)"), each = length(P_values))
)

## Limit velocity data for realistic display
pv_data$V <- pmin(pv_data$V, 1000)

ggplot(pv_data, aes(x = P, y = V, color = Material)) +
  geom_line(linewidth = 1.2) +

  ## Add safe operating point example
  annotate("point", x = 1000, y = 30, color = "darkgreen", size = 4) +
  annotate("text", x = 1200, y = 35, label = "Safe\noperating point", color = "darkgreen", size = 3, hjust = 0) +

  ## Add unsafe point example
  annotate("point", x = 2500, y = 100, color = "red", size = 4, shape = 4, stroke = 2) +
  annotate("text", x = 2700, y = 110, label = "Exceeds Nylon\nPV limit!", color = "red", size = 3, hjust = 0) +

  ## Shade regions
  annotate("rect", xmin = 100, xmax = 5000, ymin = 0, ymax = 50,
           fill = "lightgreen", alpha = 0.2) +
  annotate("text", x = 4500, y = 25, label = "Safe for all", size = 3, hjust = 1) +

  scale_color_manual(values = c("Bronze (PV=50,000)" = "sienna",
                                "Sintered Bronze (PV=35,000)" = "darkorange",
                                "PTFE-Lined (PV=15,000)" = "steelblue",
                                "Nylon (PV=5,000)" = "purple")) +
  labs(
    title = "PV Limit Curves for Common Bushing Materials",
    subtitle = "Operating point must fall BELOW the curve for the selected material",
    x = "Bearing Pressure P (psi)",
    y = "Sliding Velocity V (ft/min)",
    color = "Material"
  ) +
  coord_cartesian(ylim = c(0, 500)) +
  theme_bearing +
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold"))
```

---

#### Design Rules for PV {-}

1. **Calculate the actual PV** for your application
2. **Select a bushing material** with a PV limit **at least 2× higher** than calculated
3. **Check individual P and V limits** separately (materials have maximums for each)
4. **Consider temperature effects**—PV limits decrease at elevated temperatures

```{r individual-limits-table, echo=FALSE, message=FALSE}
limits_data <- tribble(
  ~Material, ~Max_P_psi, ~Max_V_ftmin, ~Max_PV, ~Max_Temp_F,
  "SAE 660 Bronze", 4000, 750, 75000, 450,
  "Bronze", 2000, 1200, 60000, 300,
  "PTFE/Bronze Composite", 2500, 500, 25000, 500,
  "Acetal (Delrin)", 1500, 600, 10000, 180,
  "Nylon 6/6", 1000, 400, 5000, 200,
  "PEEK", 3000, 650, 40000, 480,
  "Iron", 3000, 400, 30000, 750,
  "Bronze-Iron", 2500, 800, 35000, 750,
  "Lead-Iron", 1000, 800, 50000, 750,
  "Aluminum", 2000, 1200, 50000, 480
)

limits_data %>%
  kbl(caption = "Individual P, V, and PV Limits for Common Bushing Materials",
      col.names = c("Material", "Max Dynamic P (psi)", "Max V (ft/min)", "Max PV (psi-ft/min)", "Max Temp (F)"),
      booktabs = TRUE,
      align = "lcccc") %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  column_spec(4, background = "#F9E7F0") %>%
  row_spec(0, background = "#D81B60", color = "white")
```

---

### Applied Example: Bushing PV Calculation {#bushing-example}

::: {.question}

A shaft rotating at **500 RPM** is supported by a bronze bushing. The shaft diameter is **1.5 inches** and the bushing length is **2.0 inches**. The radial load on the bushing is **800 lbf**.

Determine:
1. The bearing pressure ($P$)
2. The sliding velocity ($V$)
3. The PV value
4. Whether a SAE 660 bronze bushing is suitable
:::

::: {.answer}

```{r bushing-example-calc, echo=FALSE, message=FALSE}
## Given values
F_load <- 800    ## lbf
d_shaft <- 1.5   ## inches
L_bush <- 2.0    ## inches
n_rpm <- 500     ## RPM

## Calculations
P_bearing <- F_load / (d_shaft * L_bush)  ## psi
V_sliding <- (pi * d_shaft * n_rpm) / 12   ## ft/min
PV_actual <- P_bearing * V_sliding          ## psi·ft/min

## SAE 660 Bronze limits
PV_limit_bronze <- 75000
P_max_bronze <- 4000
V_max_bronze <- 750

## Safety factor
SF <- PV_limit_bronze / PV_actual

## Create results table
bushing_results <- tribble(
  ~Step, ~Formula, ~Calculation, ~Result,
  "Step 1: Bearing Pressure", "P = F / (d × L)",
  paste0("P = ", F_load, " / (", d_shaft, " × ", L_bush, ")"),
  paste0(round(P_bearing, 1), " psi"),

  "Step 2: Sliding Velocity", "V = π·d·n / 12",
  paste0("V = π × ", d_shaft, " × ", n_rpm, " / 12"),
  paste0(round(V_sliding, 1), " ft/min"),

  "Step 3: PV Factor", "PV = P × V",
  paste0("PV = ", round(P_bearing, 1), " × ", round(V_sliding, 1)),
  paste0(formatC(PV_actual, format = "f", digits = 0, big.mark = ","), " psi·ft/min"),

  "Step 4: Check P Limit", "P < P_max (4,000 psi)",
  paste0(round(P_bearing, 1), " < 4,000?"),
  ifelse(P_bearing < P_max_bronze, "OK ✓", "FAIL ✗"),

  "Step 5: Check V Limit", "V < V_max (750 ft/min)",
  paste0(round(V_sliding, 1), " < 750?"),
  ifelse(V_sliding < V_max_bronze, "OK ✓", "FAIL ✗"),

  "Step 6: Check PV Limit", "PV < PV_max (75,000)",
  paste0(formatC(PV_actual, format = "f", digits = 0), " < 75,000?"),
  ifelse(PV_actual < PV_limit_bronze, "OK ✓", "FAIL ✗"),

  "Step 7: Safety Factor", "SF = PV_limit / PV_actual",
  paste0("SF = 75,000 / ", formatC(PV_actual, format = "f", digits = 0)),
  paste0(round(SF, 2))
)

bushing_results %>%
  kbl(caption = "Bushing PV Calculation Steps",
      booktabs = TRUE,
      col.names = c("Step", "Formula", "Calculation", "Result")) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(3, background = "#E8F5E9", bold = TRUE) %>%
  row_spec(4:6, background = "#E3F2FD")
```

###### Conclusion {-}

The SAE 660 bronze bushing is **suitable** for this application:

- All individual limits (P, V) are satisfied
- The actual PV (`r formatC(PV_actual, format = "f", digits = 0, big.mark = ",")` psi·ft/min) is well below the material limit (75,000 psi·ft/min)
- Safety factor of `r round(SF, 2)` provides adequate margin

:::

---

### Applied Example: Maximum Speed Calculation {#bushing-example-speed}

::: {.question}
A bushing is loaded with **20 lbf**. The shaft has a diameter of **5/16 in**. The self-lubricating bushing is made of **Bronze-Iron**. The bushing has a length of **0.5 in**.

What is the **maximum speed** (in RPM) the motor is allowed to spin?
:::

::: {.answer}

```{r bushing-speed-calc, echo=FALSE, message=FALSE}
## Given values
F_load <- 20           ## lbf
d_shaft <- 5/16        ## inches (0.3125)
L_bush <- 0.5          ## inches

## Bronze-Iron limits (from materials table)
P_max <- 2500          ## psi
V_max <- 800           ## ft/min
PV_max <- 35000        ## psi·ft/min

## Step 1: Calculate bearing pressure
P_bearing <- F_load / (d_shaft * L_bush)  ## psi

## Step 2: Calculate max velocity from V limit
V_from_Vlimit <- V_max  ## ft/min

## Step 3: Calculate max velocity from PV limit
V_from_PVlimit <- PV_max / P_bearing  ## ft/min

## Step 4: Convert velocities to RPM
## V = π·d·n / 12  →  n = 12·V / (π·d)
N_from_Vlimit <- (12 * V_from_Vlimit) / (pi * d_shaft)
N_from_PVlimit <- (12 * V_from_PVlimit) / (pi * d_shaft)

## Step 5: The actual maximum is the LOWER of the two values
N_max <- min(N_from_Vlimit, N_from_PVlimit)
limiting_factor <- ifelse(N_from_PVlimit < N_from_Vlimit, "PV limit", "V limit")

## Create results table
speed_results <- tribble(
  ~Step, ~Formula, ~Calculation, ~Result,
  "Step 1: Bearing Pressure", "P = F / (d × L)",
  paste0("P = ", F_load, " / (", d_shaft, " × ", L_bush, ")"),
  paste0(round(P_bearing, 1), " psi"),

  "Step 2: Check P < P_max", "P < 2,500 psi?",
  paste0(round(P_bearing, 1), " < 2,500?"),
  ifelse(P_bearing < P_max, "OK ✓", "FAIL ✗"),

  "Step 3: Max V from V limit", "V_max = 800 ft/min",
  "V ≤ 800",
  paste0(V_from_Vlimit, " ft/min"),

  "Step 4: Max V from PV limit", "V = PV_max / P",
  paste0("V = 35,000 / ", round(P_bearing, 1)),
  paste0(round(V_from_PVlimit, 1), " ft/min"),

  "Step 5: N from V limit", "N = 12·V / (π·d)",
  paste0("N = 12 × 800 / (π × ", d_shaft, ")"),
  paste0(formatC(N_from_Vlimit, format = "f", digits = 0, big.mark = ","), " RPM"),

  "Step 6: N from PV limit", "N = 12·V / (π·d)",
  paste0("N = 12 × ", round(V_from_PVlimit, 1), " / (π × ", d_shaft, ")"),
  paste0(formatC(N_from_PVlimit, format = "f", digits = 0, big.mark = ","), " RPM"),

  "Step 7: Governing Limit", "N_max = min(N_V, N_PV)",
  paste0("min(", formatC(N_from_Vlimit, format = "f", digits = 0, big.mark = ","),
         ", ", formatC(N_from_PVlimit, format = "f", digits = 0, big.mark = ","), ")"),
  paste0(formatC(N_max, format = "f", digits = 0, big.mark = ","), " RPM")
)

speed_results %>%
  kbl(caption = "Maximum Speed Calculation Steps",
      booktabs = TRUE,
      col.names = c("Step", "Formula", "Calculation", "Result")) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(7, background = "#E8F5E9", bold = TRUE)
```

###### Conclusion {-}

The **PV limit governs** this design. Although the V limit alone would allow `r formatC(N_from_Vlimit, format = "f", digits = 0, big.mark = ",")` RPM, the combined PV constraint limits the speed to:

$$N_{max} = \boxed{`r formatC(N_max, format = "f", digits = 0, big.mark = ",")`\ \text{RPM}}$$

::: {.callout-think}
**Why does the PV limit govern?** The bearing pressure (P = `r round(P_bearing, 1)` psi) is low compared to the 2,500 psi maximum, but when multiplied by velocity, it quickly approaches the PV limit. This is a common scenario with small-diameter, lightly-loaded bushings operating at high speeds.
:::

:::

---

### Applied Example: Minimum Bushing Length {#bushing-example-length}

::: {.question}
A bushing is loaded with **20 lbf**. The shaft runs at **3,500 RPM** and has a diameter of **5/16 in**. The self-lubricating bushing is made of **Bronze-Iron**.

What is the **minimum length** the bushing must have?
:::

::: {.answer}

```{r bushing-length-calc, echo=FALSE, message=FALSE}
## Given values
F_load <- 20           ## lbf
d_shaft <- 5/16        ## inches (0.3125)
n_rpm <- 3500          ## RPM

## Bronze-Iron limits (from materials table)
P_max <- 2500          ## psi
V_max <- 800           ## ft/min
PV_max <- 35000        ## psi·ft/min

## Step 1: Calculate sliding velocity
V_sliding <- (pi * d_shaft * n_rpm) / 12  ## ft/min

## Step 2: Check V against V_max
V_ok <- V_sliding < V_max

## Step 3: Calculate minimum L from P limit
## P = F / (d × L) ≤ P_max  →  L ≥ F / (d × P_max)
L_from_Plimit <- F_load / (d_shaft * P_max)

## Step 4: Calculate minimum L from PV limit
## P × V ≤ PV_max  →  P ≤ PV_max / V  →  F/(d×L) ≤ PV_max/V  →  L ≥ F·V / (d × PV_max)
P_max_from_PV <- PV_max / V_sliding
L_from_PVlimit <- F_load / (d_shaft * P_max_from_PV)

## Step 5: The actual minimum is the LARGER of the two values
L_min <- max(L_from_Plimit, L_from_PVlimit)
limiting_factor <- ifelse(L_from_PVlimit > L_from_Plimit, "PV limit", "P limit")

## Verify the answer
P_actual <- F_load / (d_shaft * L_min)
PV_actual <- P_actual * V_sliding

## Create results table
length_results <- tribble(
  ~Step, ~Formula, ~Calculation, ~Result,
  "Step 1: Sliding Velocity", "V = π·d·n / 12",
  paste0("V = π × ", d_shaft, " × ", n_rpm, " / 12"),
  paste0(round(V_sliding, 1), " ft/min"),

  "Step 2: Check V < V_max", "V < 800 ft/min?",
  paste0(round(V_sliding, 1), " < 800?"),
  ifelse(V_ok, "OK ✓", "FAIL ✗"),

  "Step 3: L from P limit", "L ≥ F / (d × P_max)",
  paste0("L ≥ ", F_load, " / (", d_shaft, " × 2,500)"),
  paste0(round(L_from_Plimit, 4), " in"),

  "Step 4: Max P from PV limit", "P_max = PV_max / V",
  paste0("P_max = 35,000 / ", round(V_sliding, 1)),
  paste0(round(P_max_from_PV, 1), " psi"),

  "Step 5: L from PV limit", "L ≥ F / (d × P_max)",
  paste0("L ≥ ", F_load, " / (", d_shaft, " × ", round(P_max_from_PV, 1), ")"),
  paste0(round(L_from_PVlimit, 3), " in"),

  "Step 6: Governing Limit", "L_min = max(L_P, L_PV)",
  paste0("max(", round(L_from_Plimit, 4), ", ", round(L_from_PVlimit, 3), ")"),
  paste0(round(L_min, 3), " in")
)

length_results %>%
  kbl(caption = "Minimum Length Calculation Steps",
      booktabs = TRUE,
      col.names = c("Step", "Formula", "Calculation", "Result")) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(6, background = "#E8F5E9", bold = TRUE)
```

###### Verification {-}

Let's verify by substituting $L_{min}$ back into both constraints:

- **Bearing pressure:** $P = \frac{`r F_load`}{`r d_shaft` \times `r round(L_min, 3)`} = `r round(P_actual, 1)`$ psi < 2,500 psi ✓
- **PV factor:** $PV = `r round(P_actual, 1)` \times `r round(V_sliding, 1)` = `r formatC(PV_actual, format = "f", digits = 0, big.mark = ",")`$ psi·ft/min ≈ 35,000 psi·ft/min ✓

###### Conclusion {-}

The **PV limit governs** this design. The minimum bushing length is:

$$L_{min} = \boxed{`r round(L_min, 2)`\ \text{in}}$$

In practice, you would round up to the next standard size (e.g., **0.5625 in = 9/16 in** or **0.625 in = 5/8 in**) to provide a safety margin.

:::

---

### Applied Example: Maximum Load Capacity {#bushing-example-load}

::: {.question}
A shaft runs at **3,500 RPM** and has a diameter of **5/16 in**. The self-lubricating bushing is made of **Bronze-Iron** and has a length of **0.5 in**.

What is the **maximum load** (in lbf) the bushing can handle?
:::

::: {.answer}

```{r bushing-load-calc, echo=FALSE, message=FALSE}
## Given values
d_shaft <- 5/16        ## inches (0.3125)
L_bush <- 0.5          ## inches
n_rpm <- 3500          ## RPM

## Bronze-Iron limits (from materials table)
P_max <- 2500          ## psi
V_max <- 800           ## ft/min
PV_max <- 35000        ## psi·ft/min

## Step 1: Calculate sliding velocity
V_sliding <- (pi * d_shaft * n_rpm) / 12  ## ft/min

## Step 2: Check V against V_max
V_ok <- V_sliding < V_max

## Step 3: Calculate max F from P limit
## P = F / (d × L) ≤ P_max  →  F ≤ P_max × d × L
F_from_Plimit <- P_max * d_shaft * L_bush

## Step 4: Calculate max F from PV limit
## P × V ≤ PV_max  →  P ≤ PV_max / V  →  F/(d×L) ≤ PV_max/V  →  F ≤ (PV_max/V) × d × L
P_max_from_PV <- PV_max / V_sliding
F_from_PVlimit <- P_max_from_PV * d_shaft * L_bush

## Step 5: The actual maximum is the LOWER of the two values
F_max <- min(F_from_Plimit, F_from_PVlimit)
limiting_factor <- ifelse(F_from_PVlimit < F_from_Plimit, "PV limit", "P limit")

## Create results table
load_results <- tribble(
  ~Step, ~Formula, ~Calculation, ~Result,
  "Step 1: Sliding Velocity", "V = π·d·n / 12",
  paste0("V = π × ", d_shaft, " × ", n_rpm, " / 12"),
  paste0(round(V_sliding, 1), " ft/min"),

  "Step 2: Check V < V_max", "V < 800 ft/min?",
  paste0(round(V_sliding, 1), " < 800?"),
  ifelse(V_ok, "OK ✓", "FAIL ✗"),

  "Step 3: F from P limit", "F ≤ P_max × d × L",
  paste0("F ≤ 2,500 × ", d_shaft, " × ", L_bush),
  paste0(round(F_from_Plimit, 1), " lbf"),

  "Step 4: Max P from PV limit", "P_max = PV_max / V",
  paste0("P_max = 35,000 / ", round(V_sliding, 1)),
  paste0(round(P_max_from_PV, 1), " psi"),

  "Step 5: F from PV limit", "F ≤ P_max × d × L",
  paste0("F ≤ ", round(P_max_from_PV, 1), " × ", d_shaft, " × ", L_bush),
  paste0(round(F_from_PVlimit, 1), " lbf"),

  "Step 6: Governing Limit", "F_max = min(F_P, F_PV)",
  paste0("min(", round(F_from_Plimit, 1), ", ", round(F_from_PVlimit, 1), ")"),
  paste0(round(F_max, 1), " lbf")
)

load_results %>%
  kbl(caption = "Maximum Load Calculation Steps",
      booktabs = TRUE,
      col.names = c("Step", "Formula", "Calculation", "Result")) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  row_spec(6, background = "#E8F5E9", bold = TRUE)
```

###### Conclusion {-}

Once again, the **PV limit governs** this design. The maximum allowable load is:

$$F_{max} = \boxed{`r round(F_max, 1)`\ \text{lbf}}$$

::: {.callout-industry}
**Design Insight:** Notice that at 3,500 RPM with a 0.5 in bushing, the maximum load is only about `r round(F_max, 0)` lbf—very close to the 20 lbf used in the previous examples. This illustrates why it's critical to check both P/V limits AND the combined PV limit. The individual limits (P < 2,500 psi, V < 800 ft/min) are both easily satisfied, but the PV combination becomes the controlling factor at higher speeds.

If your application requires 20 lbf at 3,500 RPM, you would need to either:
1. Increase the bushing length (as calculated in the previous example: L ≥ 0.52 in)
2. Select a material with a higher PV limit (e.g., SAE 660 Bronze with PV = 75,000)
:::

:::

---

### Bearing vs. Bushing Selection Guide {#sec-selection-guide}

When should you use a ball bearing versus a bushing? The choice depends on operating conditions, space constraints, and maintenance requirements. For bushings, the decision between oil-impregnated and solid-lubricant types depends on the motion profile and environment—continuous rotation favors oil-impregnated bearings that can develop hydrodynamic films, while oscillating or high-temperature applications favor graphite-impregnated designs (see Chapter \@ref(chap-sintering) for details on manufacturing both types).

The following table provides guidance:

```{r selection-guide, echo=FALSE, message=FALSE}
selection_guide <- tribble(
  ~Factor, ~Ball_Bearing, ~Bushing,
  "Speed", "High speeds (>3,000 RPM)", "Low to moderate speeds",
  "Load Type", "Radial and/or axial", "Primarily radial",
  "Shock Loads", "Poor—balls can indent races", "Excellent—large contact area",
  "Space (radial)", "Requires more radial space", "Very compact",
  "Lubrication", "Sealed or periodic greasing", "Self-lubricating options available",
  "Noise", "Some noise from rolling elements", "Quieter operation",
  "Cost", "Higher initial cost", "Lower cost",
  "Maintenance", "Replace entire bearing", "Can often be replaced in-situ",
  "Life Prediction", "Well-established (L10)", "Less precise; wear-based"
)

selection_guide %>%
  kbl(caption = "Ball Bearing vs. Bushing Selection Criteria",
      col.names = c("Factor", "Ball Bearing", "Bushing (Plain Bearing)"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position", full_width = TRUE) %>%
  row_spec(0, background = "#D81B60", color = "white") %>%
  column_spec(2, background = "#FFF3E0") %>%
  column_spec(3, background = "#E8F5E9")
```

---

### Summary Equations {-}

#### Ball Bearing Life {-}

| Equation | Description |
|:---------|:------------|
| $L_{10} = \left(\frac{C}{P}\right)^3 \times 10^6$ | Basic L10 life (revolutions) for ball bearings |
| $L_{10h} = \frac{10^6}{60n}\left(\frac{C}{P}\right)^3$ | L10 life in hours |
| $P = X \cdot F_r + Y \cdot F_a$ | Equivalent dynamic load |
| $L_{na} = a_1 \cdot a_{ISO} \cdot L_{10}$ | Adjusted rating life |

#### Bushing PV Limits {-}

| Equation | Description |
|:---------|:------------|
| $P = \frac{F}{d \times L}$ | Bearing pressure (projected area) |
| $V = \frac{\pi d n}{12}$ | Sliding velocity (ft/min) |
| $PV = P \times V$ | PV factor—must be less than material PV limit |

#### Hydrodynamic Lubrication {-}

| Equation | Description |
|:---------|:------------|
| $\varepsilon = \frac{e}{c}$ | Eccentricity ratio (shaft offset / clearance) |
| $h_{min} = c(1 - \varepsilon)$ | Minimum oil film thickness |
| $\frac{\mu N}{P}$ | Bearing parameter (Stribeck curve x-axis) |

> **Note:** For sintered self-lubricating bearings, the porous structure provides an internal oil reservoir. See Chapter \@ref(chap-sintering) for details on how porosity is controlled during manufacturing to optimize oil storage capacity.

---

### Design Checklist {-}

**For Ball Bearings:**

- [ ] Determine radial and axial loads
- [ ] Calculate equivalent dynamic load ($P$)
- [ ] Select bearing with $C/P$ ratio giving adequate life
- [ ] Verify speed is below limiting speed
- [ ] Apply reliability and operating condition factors if needed
- [ ] Specify lubrication and sealing requirements

**For Bushings:**

- [ ] Calculate bearing pressure ($P$)
- [ ] Calculate sliding velocity ($V$)
- [ ] Verify $P < P_{max}$ for selected material
- [ ] Verify $V < V_{max}$ for selected material
- [ ] Verify $PV < PV_{limit}$ with safety factor ≥ 2
- [ ] Consider temperature effects on PV limits
- [ ] Specify lubrication (if required) or confirm dry-running capability

**For Hydrodynamic Lubrication (Oil-Impregnated Bushings):**

- [ ] Verify minimum operating speed exceeds transition to hydrodynamic regime
- [ ] Calculate bearing parameter ($\mu N / P$) and confirm operation in hydrodynamic zone
- [ ] Estimate minimum film thickness and compare to surface roughness
- [ ] For sintered bearings: verify porosity level (20-30%) for adequate oil storage—see Chapter \@ref(chap-sintering)
- [ ] Consider startup/shutdown wear; limit start-stop cycles or use hydrostatic jacking
- [ ] Select oil viscosity appropriate for operating temperature range

**For Solid-Lubricant Bushings (Graphite/MoS₂):**

- [ ] Confirm motion type: oscillating, reciprocating, or slow rotation (hydrodynamic not feasible)
- [ ] Verify temperature requirements (graphite to 500°C, MoS₂ to 400°C)
- [ ] Check environmental requirements (vacuum, contamination, food-safe)
- [ ] See Chapter \@ref(chap-sintering) for graphite impregnation specifications
