## Thermodynamics {#chap-15-thermo}

```{r setup-15, include=FALSE}
source("R/helpers.R")
source("R/required_packages.R")

theme_thermo <- theme_minimal() +
  theme(
    plot.title       = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle    = element_text(hjust = 0.5, size = 10),
    axis.title       = element_text(size = 11, face = "bold"),
    axis.text        = element_text(size = 9),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom",
    legend.title     = element_text(size = 10, face = "bold"),
    legend.text      = element_text(size = 9),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )
```

### Learning Objectives {#sec:objectives-15}

By the end of this chapter, you will be able to:

1. **Identify** open, closed, and isolated thermodynamic systems and classify processes as isothermal, isobaric, isochoric, or adiabatic
2. **Convert** between temperature scales (Celsius, Kelvin, Fahrenheit, Rankine) and between absolute and gauge pressure
3. **Apply** the First Law of Thermodynamics to calculate heat transfer, work, internal energy, and enthalpy changes for ideal gases
4. **Evaluate** boundary work for each standard process type using the appropriate integrated formula
5. **Apply** the Steady-State Flow Energy Equation (SFEE) to turbines, compressors, nozzles, and heat exchangers
6. **Calculate** thermal efficiency of heat engines and the Carnot efficiency limit for given source and sink temperatures
7. **Determine** the Coefficient of Performance for refrigeration and heat pump cycles
8. **Analyse** the Otto, Diesel, Rankine, and Brayton cycles, computing efficiency from cycle parameters
9. **Apply** thermodynamic principles to manufacturing and automotive engineering problems

::: {.callout-think}
**Before You Read:** Your car engine burns fuel and turns a crankshaft. A refrigerator runs on electricity and keeps your food cold. A steam turbine at a power plant generates electricity from burning coal or nuclear heat. These all seem completely different — yet they all obey exactly the same two laws. What are those laws, and why can't any engine ever be 100% efficient?
:::

***

### Introduction {#sec:thermo-intro-15}

Thermodynamics is the science of energy — how it is stored, transferred, and converted from one form to another. Every manufacturing process, every engine, every heating and cooling system operates according to thermodynamic principles. Understanding these principles allows engineers to:

- **Design** energy-efficient processes and machines
- **Predict** how systems will behave under different operating conditions
- **Identify** the theoretical limits of performance (so you know when you've done as well as physics allows)
- **Troubleshoot** systems that are underperforming

The word itself comes from the Greek *therme* (heat) and *dynamis* (power). Two foundational laws govern all of thermodynamics:

- **First Law:** Energy cannot be created or destroyed — it can only change form. The total energy of a system and its surroundings is always conserved.
- **Second Law:** Not all energy conversions are equally useful. Heat naturally flows from hot to cold. You can never convert heat completely to work. Every real process increases the total entropy (disorder) of the universe.

These are not suggestions or approximations — they are the most thoroughly tested laws in all of science.

***

### Part 1: Fundamentals {#sec:thermo-fundamentals-15}

#### Systems and Surroundings {#sec:systems-15}

Every thermodynamic analysis starts by defining what you are studying.

| Term | Definition | Example |
|:-----|:-----------|:--------|
| **System** | The region or matter being analyzed | Gas inside a cylinder |
| **Surroundings** | Everything outside the system boundary | The piston, walls, atmosphere |
| **Boundary** | The real or imaginary surface separating system from surroundings | The cylinder walls and piston face |

Systems are classified by what can cross their boundary:

| Type | Mass crosses boundary? | Energy crosses boundary? | Example |
|:-----|:-----:|:-----:|:--------|
| **Open (Control Volume)** | Yes | Yes | Turbine, pump, compressor, nozzle |
| **Closed (Control Mass)** | No | Yes | Gas in a sealed piston-cylinder |
| **Isolated** | No | No | Perfectly insulated sealed tank (idealized) |

::: {.callout-industry}
**Industry Application:** When analyzing a turbocharger in an automotive engine, you treat it as an **open system** — exhaust gas enters, compressed air exits, and heat is exchanged with the engine compartment. When analyzing the compression stroke of a single cylinder (valves closed), you treat it as a **closed system** — no mass crosses the boundary, but the piston does work on the gas.
:::

#### State, Properties, and Processes {#sec:state-props-15}

The **state** of a system is its condition at a given instant, fully described by a set of **intensive properties** (independent of mass, e.g., temperature, pressure, density) and **extensive properties** (proportional to mass, e.g., volume, internal energy, enthalpy).

A **process** is a path from one state to another. The four standard process types for a gas are:

| Process | Constant Quantity | Defining Equation | Boundary Work |
|:--------|:-----------------|:-----------------|:-------------|
| **Isothermal** | Temperature ($T$) | $PV = \text{const}$ | $W = P_1 V_1 \ln\!\left(\frac{V_2}{V_1}\right)$ |
| **Isobaric** | Pressure ($P$) | $V/T = \text{const}$ | $W = P(V_2 - V_1)$ |
| **Isochoric** | Volume ($V$) | $P/T = \text{const}$ | $W = 0$ |
| **Adiabatic** | Heat transfer ($Q = 0$) | $PV^{\gamma} = \text{const}$ | $W = \frac{P_1 V_1 - P_2 V_2}{\gamma - 1}$ |

A process is **reversible** if it can be returned to its original state with no net effect on the surroundings. Real processes are always **irreversible** due to friction, heat transfer across finite temperature differences, and mixing.

#### Temperature Scales {#sec:temperature-15}

Four temperature scales are in common use in engineering. The two **absolute** scales (Kelvin and Rankine) have their zero at absolute zero — the point of minimum possible thermal energy.

\begin{equation}
T(\text{K}) = T(°\text{C}) + 273.15
(\#eq:celsius-kelvin-15)
\end{equation}

\begin{equation}
T(°\text{F}) = \frac{9}{5}\,T(°\text{C}) + 32
(\#eq:celsius-fahrenheit-15)
\end{equation}

\begin{equation}
T(\text{R}) = T(°\text{F}) + 459.67
(\#eq:fahrenheit-rankine-15)
\end{equation}

\begin{equation}
T(\text{R}) = 1.8 \times T(\text{K})
(\#eq:kelvin-rankine-15)
\end{equation}

::: {.callout-think}
**Think About It:** Why must all thermodynamic efficiency calculations use absolute temperatures (Kelvin or Rankine), not Celsius or Fahrenheit?
:::

**Worked Example 1: Temperature Conversion**

A steel billet leaves a heat treatment furnace at **350°C**. Convert this to Kelvin, Fahrenheit, and Rankine.

*Solution:*

$$T(\text{K}) = 350 + 273.15 = \mathbf{623.15 \text{ K}}$$

$$T(°\text{F}) = \frac{9}{5}(350) + 32 = 630 + 32 = \mathbf{662\,°\text{F}}$$

$$T(\text{R}) = 662 + 459.67 = \mathbf{1{,}121.67 \text{ R}}$$

#### Pressure {#sec:pressure-15}

**Absolute pressure** ($P_{abs}$) is measured relative to a perfect vacuum. **Gauge pressure** ($P_{gauge}$) is measured relative to the local atmospheric pressure. A **vacuum pressure** is a gauge pressure below atmospheric.

\begin{equation}
P_{abs} = P_{gauge} + P_{atm}
(\#eq:pressure-abs-15)
\end{equation}

\begin{equation}
P_{vacuum} = P_{atm} - P_{abs}
(\#eq:pressure-vacuum-15)
\end{equation}

Standard atmospheric pressure: $P_{atm} = 101.325 \text{ kPa} = 1 \text{ atm} = 14.696 \text{ psi}$

**Worked Example 2: Gauge vs. Absolute Pressure**

A car tyre is inflated to a gauge pressure of **220 kPa**. What is the absolute pressure inside the tyre?

*Solution:*

$$P_{abs} = P_{gauge} + P_{atm} = 220 + 101.325 = \mathbf{321.3 \text{ kPa}}$$

Note that if the tyre were completely flat, the pressure inside would equal atmospheric (101.3 kPa), not zero.

#### The Zeroth Law of Thermodynamics {#sec:zeroth-law-15}

> *If two bodies are each in thermal equilibrium with a third body, they are in thermal equilibrium with each other.*

This law establishes that **temperature** is a meaningful, measurable property — and it is the basis for all thermometry. When a thermometer reads 350°C, it means the thermometer and the object have reached thermal equilibrium.

#### Phases of Matter and the Phase Diagram {#sec:phases-15}

Matter exists in three principal phases: **solid**, **liquid**, and **vapor (gas)**. The phase of a pure substance depends on its temperature and pressure.

```{r phase-diagram-15, fig.cap="Pressure–temperature (P–T) phase diagram for a pure substance. The triple point is the unique state where all three phases coexist. The critical point marks the end of the liquid–vapor boundary; above it, the fluid becomes supercritical and no phase transition occurs.", fig.align='center', echo=FALSE}
library(ggplot2)

# Approximate shape curves for a generic pure substance (e.g., water-like)
# Sublimation curve (solid-vapor): P rises steeply as T rises to triple point
tp_T <- 0.01    # Triple point temperature (normalized)
tp_P <- 0.006   # Triple point pressure (normalized)
cp_T <- 1.0     # Critical point temperature (normalized)
cp_P <- 1.0     # Critical point pressure (normalized)

# Sublimation curve: from very low T up to triple point
sub_T <- seq(-1.0, tp_T, length.out = 80)
sub_P <- tp_P * exp(3.5 * (sub_T - tp_T))

# Fusion (melting) curve: nearly vertical, slight positive slope
fus_T <- seq(tp_T, 1.4, length.out = 40)
fus_P <- tp_P + 8 * (fus_T - tp_T)^0.9

# Vaporization curve: from triple point to critical point
vap_T <- seq(tp_T, cp_T, length.out = 100)
vap_P <- tp_P + (cp_P - tp_P) * ((vap_T - tp_T) / (cp_T - tp_T))^2.5

df_sub <- data.frame(T = sub_T, P = sub_P, curve = "Sublimation")
df_fus <- data.frame(T = fus_T, P = fus_P, curve = "Fusion (Melting)")
df_vap <- data.frame(T = vap_T, P = vap_P, curve = "Vaporization")
df_all <- rbind(df_sub, df_fus, df_vap)

ggplot(df_all, aes(x = T, y = P, colour = curve)) +
  geom_line(linewidth = 1.2) +
  # Region labels
  annotate("text", x = -0.55, y = 0.6,  label = "SOLID",  size = 5, fontface = "bold", colour = "#264653") +
  annotate("text", x =  0.55, y = 0.65, label = "LIQUID", size = 5, fontface = "bold", colour = "#2A9D8F") +
  annotate("text", x =  0.55, y = 0.08, label = "VAPOR",  size = 5, fontface = "bold", colour = "#E76F51") +
  # Triple point
  annotate("point", x = tp_T, y = tp_P, size = 4, colour = "black") +
  annotate("text",  x = tp_T + 0.08, y = tp_P + 0.04,
           label = "Triple\nPoint", size = 3.2, hjust = 0, colour = "black") +
  # Critical point
  annotate("point", x = cp_T, y = cp_P, size = 4, colour = "#E63946") +
  annotate("text",  x = cp_T - 0.08, y = cp_P + 0.06,
           label = "Critical\nPoint", size = 3.2, hjust = 1, colour = "#E63946") +
  # Supercritical region label
  annotate("text", x = 1.15, y = 0.7, label = "SUPER-\nCRITICAL", size = 3.5,
           fontface = "italic", colour = "grey40") +
  scale_colour_manual(values = c("Sublimation" = "#E76F51",
                                  "Fusion (Melting)" = "#264653",
                                  "Vaporization" = "#2A9D8F")) +
  scale_x_continuous(limits = c(-1.1, 1.5)) +
  scale_y_continuous(limits = c(-0.01, 1.15)) +
  labs(
    title    = "Phase Diagram of a Pure Substance",
    x        = "Temperature (normalized)",
    y        = "Pressure (normalized)",
    colour   = "Phase Boundary"
  ) +
  theme_thermo
```

**Key features of the phase diagram:**

- **Triple point:** The unique combination of $T$ and $P$ at which solid, liquid, and vapor coexist in equilibrium. For water: $T_{tp} = 0.01°\text{C}$, $P_{tp} = 0.6117 \text{ kPa}$.
- **Critical point:** Above this temperature and pressure, the distinction between liquid and vapor disappears — the fluid becomes **supercritical**. For water: $T_c = 374.1°\text{C}$, $P_c = 22.09 \text{ MPa}$.
- **Sublimation:** Direct transition from solid to vapor (bypassing liquid), occurring below the triple point pressure. Dry ice (CO$_2$) sublimates at atmospheric pressure.

#### Quality of a Wet Steam Mixture {#sec:quality-15}

When a pure substance exists as a **mixture** of saturated liquid and saturated vapor (the wet region), its state is defined by the **quality** $x$:

\begin{equation}
x = \frac{m_{vapor}}{m_{total}} = \frac{m_{vapor}}{m_{liquid} + m_{vapor}}
(\#eq:quality-15)
\end{equation}

Quality ranges from 0 (saturated liquid) to 1 (saturated vapor, also called dry saturated steam). Any specific property $y$ of the mixture is:

\begin{equation}
y = y_f + x \cdot y_{fg}
(\#eq:mixture-property-15)
\end{equation}

where $y_f$ is the value at saturated liquid, $y_g$ is the value at saturated vapor, and $y_{fg} = y_g - y_f$.

::: {.callout-industry}
**Industry Application:** In food processing plants, steam quality is critical. Low-quality steam (high moisture content) can cause temperature non-uniformity in retort sterilization of canned goods. Processors specify **steam quality of 0.97 or higher** at the point of use. Steam traps and separators are installed to remove condensate and ensure high-quality dry steam reaches cooking vessels.
:::

***

### Part 2: The First Law and Energy {#sec:first-law-15}

#### The First Law of Thermodynamics {#sec:first-law-statement-15}

> *Energy cannot be created or destroyed. The total energy of the universe is constant.*

For a **closed system** undergoing a process, the First Law states:

\begin{equation}
Q - W = \Delta U = U_2 - U_1
(\#eq:first-law-closed-15)
\end{equation}

where:
- $Q$ = heat transferred **into** the system (positive when heat enters)
- $W$ = work done **by** the system (positive when the system does work on surroundings)
- $\Delta U$ = change in internal energy of the system

This sign convention is the **physics convention**. (Some engineering texts use $Q - W = \Delta U$ consistently; always check which sign convention a textbook uses before applying formulas.)

#### Internal Energy and Enthalpy {#sec:internal-energy-enthalpy-15}

**Internal energy** ($U$, or specific internal energy $u$ in J/kg) represents the microscopic energy of a substance — the kinetic and potential energy of its molecules.

**Enthalpy** ($H$) is defined as:

\begin{equation}
H = U + PV \quad \Leftrightarrow \quad h = u + Pv
(\#eq:enthalpy-def-15)
\end{equation}

Enthalpy is a convenient property for **open systems** and **constant-pressure processes**. At constant pressure, the heat transferred equals the change in enthalpy:

$$Q_P = \Delta H = m \cdot c_p \cdot \Delta T$$

::: {.callout-think}
**Think About It:** Why is enthalpy more useful than internal energy for analyzing a steam turbine? What crosses the system boundary in that case?
:::

#### Heat and Work {#sec:heat-work-15}

**Heat** ($Q$) is energy transfer driven by a temperature difference. **Work** ($W$) is energy transfer due to a force acting through a distance. Both are **path functions** — their values depend on the process taken, not just the initial and final states.

The most common form of work in closed systems is **boundary work** (also called $P$-$V$ work):

\begin{equation}
W_{boundary} = \int_{V_1}^{V_2} P \, dV
(\#eq:boundary-work-15)
\end{equation}

For each process type, this integral evaluates to:

\begin{equation}
W_{isochoric} = 0 \quad (V = \text{const})
(\#eq:work-isochoric-15)
\end{equation}

\begin{equation}
W_{isobaric} = P(V_2 - V_1) = mR(T_2 - T_1)
(\#eq:work-isobaric-15)
\end{equation}

\begin{equation}
W_{isothermal} = P_1 V_1 \ln\!\left(\frac{V_2}{V_1}\right) = mRT \ln\!\left(\frac{V_2}{V_1}\right)
(\#eq:work-isothermal-15)
\end{equation}

\begin{equation}
W_{adiabatic} = \frac{P_1 V_1 - P_2 V_2}{\gamma - 1} = \frac{mR(T_1 - T_2)}{\gamma - 1}
(\#eq:work-adiabatic-15)
\end{equation}

```{r pv-diagram-15, fig.cap="P–V diagram showing the four standard thermodynamic processes for an ideal gas starting from the same initial state (State 1). The isothermal and adiabatic curves are both hyperbolic but the adiabatic is steeper (higher pressure drop for the same volume increase) because no heat is added to compensate. The area under each curve equals the boundary work for that process.", fig.align='center', echo=FALSE}
library(ggplot2)

V1 <- 1.0; P1 <- 100.0; T1 <- P1 * V1   # PV = RT (normalized, m=1, R=1)
gamma <- 1.4

V <- seq(V1, 3.5, length.out = 200)

P_iso  <- P1 * V1 / V                          # Isothermal: PV = const
P_ibar <- rep(P1, length(V))                   # Isobaric:   P = const
P_ivol <- NULL                                  # Isochoric:  vertical line
P_adia <- P1 * (V1 / V)^gamma                  # Adiabatic:  PV^γ = const

df_iso  <- data.frame(V = V, P = P_iso,  process = "Isothermal (T = const)")
df_ibar <- data.frame(V = V, P = P_ibar, process = "Isobaric (P = const)")
df_adia <- data.frame(V = V, P = P_adia, process = "Adiabatic (Q = 0)")

# Isochoric: vertical line at V1
df_ivol <- data.frame(V = c(V1, V1), P = c(P1, P1 * 0.3), process = "Isochoric (V = const)")

df_all <- rbind(df_iso, df_ibar, df_adia, df_ivol)

ggplot(df_all, aes(x = V, y = P, colour = process)) +
  geom_line(linewidth = 1.3) +
  geom_point(aes(x = V1, y = P1), colour = "black", size = 3.5) +
  annotate("text", x = V1 - 0.08, y = P1 + 4, label = "State 1", size = 3.5,
           fontface = "bold", hjust = 1) +
  scale_colour_manual(values = c(
    "Isothermal (T = const)"  = "#2A9D8F",
    "Isobaric (P = const)"    = "#E9C46A",
    "Adiabatic (Q = 0)"       = "#E63946",
    "Isochoric (V = const)"   = "#264653"
  )) +
  labs(
    title   = "P–V Diagram: Four Standard Processes",
    x       = expression(Volume ~ (m^3)),
    y       = "Pressure (kPa)",
    colour  = "Process"
  ) +
  coord_cartesian(ylim = c(0, 115)) +
  theme_thermo
```

#### Specific Heats {#sec:specific-heats-15}

The **specific heat** of a substance is the energy required to raise the temperature of 1 kg by 1°C (or 1 K).

- $c_v$ = specific heat at **constant volume** (J/kg·K)
- $c_p$ = specific heat at **constant pressure** (J/kg·K)

For an ideal gas, these are related to the specific gas constant $R$:

\begin{equation}
c_p - c_v = R
(\#eq:cp-cv-R-15)
\end{equation}

\begin{equation}
\gamma = \frac{c_p}{c_v}
(\#eq:gamma-15)
\end{equation}

Changes in internal energy and enthalpy for an ideal gas:

\begin{equation}
\Delta u = c_v \Delta T \quad \Rightarrow \quad \Delta U = m \, c_v \Delta T
(\#eq:delta-u-15)
\end{equation}

\begin{equation}
\Delta h = c_p \Delta T \quad \Rightarrow \quad \Delta H = m \, c_p \Delta T
(\#eq:delta-h-15)
\end{equation}

Typical values for air: $c_p = 1.005 \text{ kJ/kg·K}$, $c_v = 0.718 \text{ kJ/kg·K}$, $R = 0.287 \text{ kJ/kg·K}$, $\gamma = 1.4$.

#### The Ideal Gas Law {#sec:ideal-gas-15}

An **ideal gas** is a model in which molecules have negligible volume and no intermolecular forces. This is a good approximation for gases at low-to-moderate pressures and temperatures well above the boiling point.

\begin{equation}
PV = nRT_{universal} \quad \text{(molar form, } R_u = 8.314 \text{ J/mol·K)}
(\#eq:ideal-gas-molar-15)
\end{equation}

\begin{equation}
Pv = RT \quad \text{(specific form, } v = V/m, R = R_u / M_w\text{)}
(\#eq:ideal-gas-specific-15)
\end{equation}

\begin{equation}
\frac{P_1 V_1}{T_1} = \frac{P_2 V_2}{T_2} \quad \text{(combined gas law)}
(\#eq:combined-gas-law-15)
\end{equation}

For air: $R_{air} = 0.287 \text{ kJ/kg·K}$.

**Worked Example 3: Isothermal Compression**

Air at $P_1 = 100 \text{ kPa}$, $T = 300 \text{ K}$ is compressed isothermally in a piston-cylinder from $V_1 = 0.5 \text{ m}^3$ to $V_2 = 0.1 \text{ m}^3$. Find the boundary work done **on** the air.

*Solution:*

Since the process is isothermal: $W = P_1 V_1 \ln\!\left(\dfrac{V_2}{V_1}\right)$

$$W = (100 \times 10^3)(0.5) \ln\!\left(\frac{0.1}{0.5}\right) = 50{,}000 \times \ln(0.2)$$

$$W = 50{,}000 \times (-1.609) = -80{,}472 \text{ J} = \mathbf{-80.5 \text{ kJ}}$$

The negative sign means work is done **on** the system (compression). The surroundings (piston) do 80.5 kJ of work on the air.

**Worked Example 4: Adiabatic Compression**

Air is compressed adiabatically from $P_1 = 100 \text{ kPa}$, $T_1 = 300 \text{ K}$ to $P_2 = 600 \text{ kPa}$. Find the final temperature and the specific work input.

*Solution:*

For an adiabatic process with an ideal gas: $T_1 P_1^{-\gamma/(\gamma-1)} = T_2 P_2^{-\gamma/(\gamma-1)}$, equivalently:

$$\frac{T_2}{T_1} = \left(\frac{P_2}{P_1}\right)^{(\gamma-1)/\gamma}$$

$$T_2 = 300 \times \left(\frac{600}{100}\right)^{(1.4-1)/1.4} = 300 \times 6^{0.2857}$$

$$T_2 = 300 \times 1.6685 = \mathbf{500.6 \text{ K}} \quad (227.4\,°\text{C})$$

Specific work: $w = c_v(T_1 - T_2) = 0.718(300 - 500.6) = \mathbf{-144 \text{ kJ/kg}}$ (work in)

::: {.callout-industry}
**Industry Application:** This is exactly what happens in a turbocharger compressor. Air enters at near-ambient conditions and is compressed (approximately adiabatically at high flow rates) to 2–3 times atmospheric pressure, raising the temperature by 100–200°C. An **intercooler** is placed after the compressor to cool the air before it enters the engine — cooler air is denser and carries more oxygen per litre, increasing power.
:::

**Worked Example 5: Heating in a Rigid Tank**

A 2 kg sample of air in a rigid sealed tank ($c_v = 0.718 \text{ kJ/kg·K}$) is heated from $T_1 = 25°\text{C}$ to $T_2 = 200°\text{C}$. Find the heat added and the change in enthalpy.

*Solution:*

Rigid tank → isochoric process → $W = 0$. From the First Law: $Q = \Delta U$.

$$\Delta U = m c_v \Delta T = 2 \times 0.718 \times (200 - 25) = 2 \times 0.718 \times 175$$

$$\Delta U = \mathbf{251.3 \text{ kJ}} \quad \Rightarrow \quad Q = 251.3 \text{ kJ}$$

$$\Delta H = m c_p \Delta T = 2 \times 1.005 \times 175 = \mathbf{351.75 \text{ kJ}}$$

Note: $\Delta H \neq Q$ here because the process is not at constant pressure. The enthalpy change is larger than the heat added because enthalpy also accounts for the $P\Delta V$ term (which is zero in the rigid tank, but is computed from the temperature change alone for an ideal gas).

#### Steady-State Flow Energy Equation (SFEE) {#sec:sfee-15}

For an **open system** at steady state (mass flow rate $\dot{m}$ = constant, no accumulation), the energy balance per unit mass is:

\begin{equation}
q - w_s = (h_2 - h_1) + \frac{V_2^2 - V_1^2}{2} + g(z_2 - z_1)
(\#eq:sfee-15)
\end{equation}

where:
- $q$ = specific heat transfer (J/kg), positive into system
- $w_s$ = specific shaft work (J/kg), positive out of system
- $h$ = specific enthalpy (J/kg)
- $V$ = fluid velocity (m/s)
- $z$ = elevation (m)

For most turbomachinery (turbines, compressors, pumps), kinetic and potential energy changes are negligible, simplifying to:

$$q - w_s = h_2 - h_1$$

**Power** output (or input): $\dot{W}_s = \dot{m} \cdot w_s$

**Worked Example 6: Steam Turbine Power**

Steam enters a turbine at $h_1 = 3{,}450 \text{ kJ/kg}$ and exits at $h_2 = 2{,}300 \text{ kJ/kg}$. The mass flow rate is $\dot{m} = 10 \text{ kg/s}$ and the turbine is well-insulated ($q = 0$). Find the power output.

*Solution:*

SFEE (no KE or PE changes, adiabatic): $-w_s = h_2 - h_1$

$$w_s = h_1 - h_2 = 3{,}450 - 2{,}300 = 1{,}150 \text{ kJ/kg}$$

$$\dot{W}_s = \dot{m} \cdot w_s = 10 \times 1{,}150 = \mathbf{11{,}500 \text{ kW}} = \mathbf{11.5 \text{ MW}}$$

::: {.callout-think}
**Think About It:** A compressor is the reverse of a turbine — it receives shaft work to raise the fluid pressure. How would you modify the SFEE equation to find the power input to a compressor?
:::

**Worked Example 13: Manufacturing — Heating a Steel Billet**

A steel forging plant needs to heat a **450 kg steel billet** from room temperature (20°C) to forging temperature (1,200°C). The specific heat of steel is approximately $c_p = 0.490 \text{ kJ/kg·K}$. How much energy is required?

*Solution:*

$$Q = m c_p \Delta T = 450 \times 0.490 \times (1{,}200 - 20)$$

$$Q = 450 \times 0.490 \times 1{,}180 = \mathbf{260{,}190 \text{ kJ}} \approx \mathbf{260 \text{ MJ}}$$

At an electric furnace efficiency of 85%, the electrical energy drawn from the grid is:

$$E_{electric} = \frac{260}{0.85} = \mathbf{306 \text{ MJ}} \approx \mathbf{85 \text{ kWh}}$$

This is roughly equivalent to running a typical home for 3 full days — just to heat one billet.

::: {.callout-industry}
**Industry Application:** In automotive stamping plants (such as those producing structural body panels), large gas-fired furnaces operate at 950–1,100°C. Thermal efficiency of the furnace is a major cost driver. Recuperators (heat exchangers that pre-heat the incoming combustion air using hot exhaust gas) can recover 30–50% of the exhaust heat, significantly reducing natural gas consumption and CO₂ emissions.
:::

***

### Part 3: The Second Law and Entropy {#sec:second-law-15}

#### The Second Law of Thermodynamics {#sec:second-law-statement-15}

The First Law tells us *how much* energy is conserved. The Second Law tells us *in which direction* processes naturally occur and places limits on what is achievable.

There are two classical statements:

**Kelvin–Planck Statement:** It is impossible to construct a device that operates in a cycle and produces a net amount of work while exchanging heat with only one thermal reservoir (i.e., you cannot convert heat completely to work in a cyclic device).

**Clausius Statement:** It is impossible to construct a device that operates in a cycle and transfers heat from a cooler body to a warmer body without consuming work.

Both statements are equivalent — if one is violated, so is the other.

::: {.callout-think}
**Think About It:** A wind turbine converts wind kinetic energy directly to electricity with no heat rejection needed. Does this violate the Kelvin–Planck statement? Why or why not?
:::

#### Thermal Efficiency of Heat Engines {#sec:thermal-efficiency-15}

A **heat engine** is a cyclic device that receives heat from a high-temperature source ($Q_H$), converts some to work ($W_{net}$), and rejects the rest to a low-temperature sink ($Q_L$).

From energy conservation: $W_{net} = Q_H - Q_L$

\begin{equation}
\eta_{th} = \frac{W_{net}}{Q_H} = 1 - \frac{Q_L}{Q_H}
(\#eq:thermal-efficiency-15)
\end{equation}

**Worked Example 7: Heat Engine Efficiency**

A heat engine receives $Q_H = 1{,}200 \text{ kJ}$ per cycle from a furnace and rejects $Q_L = 750 \text{ kJ}$ to the atmosphere. Find the thermal efficiency and net work output.

*Solution:*

$$W_{net} = Q_H - Q_L = 1{,}200 - 750 = \mathbf{450 \text{ kJ}}$$

$$\eta_{th} = 1 - \frac{750}{1{,}200} = 1 - 0.625 = \mathbf{0.375 = 37.5\%}$$

Only 37.5% of the fuel energy becomes useful work; 62.5% is rejected as waste heat.

**Worked Example 14: Automotive Engine Efficiency**

A gasoline engine produces **100 kW of shaft power** while burning fuel at a rate that releases **280 kW of thermal energy**. What is the engine's thermal efficiency? How much power is rejected as heat?

*Solution:*

$$\eta_{th} = \frac{\dot{W}_{net}}{\dot{Q}_H} = \frac{100}{280} = \mathbf{35.7\%}$$

$$\dot{Q}_L = \dot{Q}_H - \dot{W}_{net} = 280 - 100 = \mathbf{180 \text{ kW}}$$

This 180 kW of rejected heat exits through the exhaust (roughly 120 kW) and the radiator (roughly 60 kW) — which is why both get very hot. Turbocharged and direct-injection engines achieve 38–42% thermal efficiency; the best automotive spark-ignition engines reach approximately 45%.

#### The Carnot Cycle and Maximum Efficiency {#sec:carnot-15}

The **Carnot cycle** is a theoretical reversible cycle operating between two fixed temperatures. It is the most efficient possible cycle between those temperatures. Its efficiency is:

\begin{equation}
\eta_{Carnot} = 1 - \frac{T_L}{T_H}
(\#eq:carnot-efficiency-15)
\end{equation}

where $T_H$ and $T_L$ are the **absolute** (Kelvin) temperatures of the hot source and cold sink.

The Carnot cycle consists of four reversible processes:
1. **Isothermal expansion** at $T_H$ (heat addition $Q_H$)
2. **Adiabatic expansion** from $T_H$ to $T_L$
3. **Isothermal compression** at $T_L$ (heat rejection $Q_L$)
4. **Adiabatic compression** from $T_L$ to $T_H$

No real cycle can achieve Carnot efficiency because real processes are irreversible.

**Worked Example 8: Carnot Efficiency**

Find the maximum possible efficiency for a heat engine operating between a furnace at $T_H = 800 \text{ K}$ and the atmosphere at $T_L = 300 \text{ K}$.

*Solution:*

$$\eta_{Carnot} = 1 - \frac{T_L}{T_H} = 1 - \frac{300}{800} = 1 - 0.375 = \mathbf{0.625 = 62.5\%}$$

No engine operating between these two temperatures can exceed 62.5% efficiency, regardless of its design. Real engines achieve 30–45% — well below this limit — due to friction, heat losses, and irreversibilities.

```{r carnot-ts-15, fig.cap="T–s diagram for the Carnot cycle. The cycle traces a perfect rectangle: isothermal heat addition at T_H (horizontal line, left to right), adiabatic expansion to T_L (vertical drop), isothermal heat rejection at T_L (right to left), and adiabatic compression back to T_H (vertical rise). The enclosed area equals the net work output. The ratio of the shaded rectangle to the total rectangle equals the Carnot efficiency.", fig.align='center', echo=FALSE}
library(ggplot2)

T_H <- 800; T_L <- 300
s1 <- 1.0; s2 <- 3.0

df_carnot <- data.frame(
  s = c(s1, s2, s2, s1, s1),
  T = c(T_H, T_H, T_L, T_L, T_H)
)

# Shaded interior
df_shade <- data.frame(
  s = c(s1, s2, s2, s1),
  T = c(T_H, T_H, T_L, T_L)
)

ggplot() +
  geom_polygon(data = df_shade, aes(x = s, y = T), fill = "#2A9D8F", alpha = 0.25) +
  geom_path(data = df_carnot, aes(x = s, y = T), colour = "#264653", linewidth = 1.4) +
  # Process labels
  annotate("text", x = (s1+s2)/2, y = T_H + 20, label = "① Isothermal Expansion (Q_H in)",
           hjust = 0.5, size = 3.2, colour = "#264653") +
  annotate("text", x = s2 + 0.08, y = (T_H+T_L)/2, label = "② Adiabatic\nExpansion",
           hjust = 0, size = 3.0, colour = "#264653") +
  annotate("text", x = (s1+s2)/2, y = T_L - 20, label = "③ Isothermal Compression (Q_L out)",
           hjust = 0.5, size = 3.2, colour = "#264653") +
  annotate("text", x = s1 - 0.08, y = (T_H+T_L)/2, label = "④ Adiabatic\nCompression",
           hjust = 1, size = 3.0, colour = "#264653") +
  # State points
  annotate("point", x = c(s1, s2, s2, s1), y = c(T_H, T_H, T_L, T_L),
           size = 3, colour = "#E63946") +
  annotate("text",  x = c(s1-0.05, s2+0.05, s2+0.05, s1-0.05),
                    y = c(T_H+15, T_H+15, T_L-15, T_L-15),
           label = c("1","2","3","4"), size = 3.5, colour = "#E63946", fontface = "bold") +
  # Reference lines
  geom_hline(yintercept = T_H, linetype = "dashed", colour = "#E9C46A", linewidth = 0.7) +
  geom_hline(yintercept = T_L, linetype = "dashed", colour = "#E9C46A", linewidth = 0.7) +
  annotate("text", x = 3.4, y = T_H, label = paste("T_H =", T_H, "K"), hjust = 0, size = 3) +
  annotate("text", x = 3.4, y = T_L, label = paste("T_L =", T_L, "K"), hjust = 0, size = 3) +
  # W_net annotation
  annotate("text", x = (s1+s2)/2, y = (T_H+T_L)/2,
           label = "W_net\n(enclosed area)", hjust = 0.5, size = 3.5,
           colour = "#2A9D8F", fontface = "italic") +
  scale_x_continuous(limits = c(0.5, 4.0)) +
  scale_y_continuous(limits = c(200, 900)) +
  labs(
    title = "T–s Diagram: Carnot Cycle",
    x     = "Specific Entropy, s (kJ/kg·K)",
    y     = "Temperature, T (K)"
  ) +
  theme_thermo
```

#### Entropy {#sec:entropy-15}

**Entropy** ($S$, specific entropy $s$ in J/kg·K) is a measure of the microscopic disorder of a system. For a reversible heat transfer at temperature $T$:

\begin{equation}
dS = \frac{\delta Q_{rev}}{T} \quad \Rightarrow \quad \Delta S = \int_1^2 \frac{\delta Q_{rev}}{T}
(\#eq:entropy-def-15)
\end{equation}

For an **irreversible** process, the entropy generated is always positive (Clausius inequality):

\begin{equation}
\Delta S_{universe} = \Delta S_{system} + \Delta S_{surroundings} \geq 0
(\#eq:entropy-universe-15)
\end{equation}

Equality holds for reversible processes. For **isentropic** (adiabatic + reversible) processes: $\Delta s = 0$.

Entropy change of an ideal gas between two states:

\begin{equation}
\Delta s = c_p \ln\!\left(\frac{T_2}{T_1}\right) - R \ln\!\left(\frac{P_2}{P_1}\right)
(\#eq:entropy-ideal-gas-15)
\end{equation}

\begin{equation}
\Delta s = c_v \ln\!\left(\frac{T_2}{T_1}\right) + R \ln\!\left(\frac{v_2}{v_1}\right)
(\#eq:entropy-ideal-gas-v-15)
\end{equation}

#### Coefficient of Performance — Refrigerators and Heat Pumps {#sec:cop-15}

A **refrigerator** uses work input to transfer heat from a cold space ($Q_L$) to a warm space. A **heat pump** uses work input to transfer heat into a warm space ($Q_H$) from a cold space. In both cases:

\begin{equation}
\text{COP}_{refrigerator} = \frac{Q_L}{W_{net,in}} = \frac{Q_L}{Q_H - Q_L}
(\#eq:cop-refrig-15)
\end{equation}

\begin{equation}
\text{COP}_{heat pump} = \frac{Q_H}{W_{net,in}} = \frac{Q_H}{Q_H - Q_L}
(\#eq:cop-hp-15)
\end{equation}

\begin{equation}
\text{COP}_{HP} = \text{COP}_{refrig} + 1
(\#eq:cop-relationship-15)
\end{equation}

The **Carnot COP** (maximum possible) for a refrigerator:

\begin{equation}
\text{COP}_{refrig,Carnot} = \frac{T_L}{T_H - T_L}
(\#eq:cop-carnot-refrig-15)
\end{equation}

**Worked Example 9: Refrigerator COP**

A refrigerator removes $Q_L = 800 \text{ kJ/h}$ from a freezer compartment while drawing $W_{in} = 250 \text{ kJ/h}$ of electrical power. Find the COP and the heat rejected to the kitchen.

*Solution:*

$$\text{COP}_{refrig} = \frac{Q_L}{W_{in}} = \frac{800}{250} = \mathbf{3.2}$$

$$Q_H = Q_L + W_{in} = 800 + 250 = \mathbf{1{,}050 \text{ kJ/h}}$$

A COP of 3.2 means the refrigerator moves 3.2 J of heat for every 1 J of electrical energy consumed — much better than 1.0, which is why heat pumps are more efficient heaters than electric resistance heaters.

::: {.callout-industry}
**Industry Application:** Food processing facilities use large-scale industrial refrigeration systems (ammonia or CO₂ refrigerant) to maintain storage warehouses at −25°C. Optimizing the COP of these systems — by reducing the temperature lift (difference between $T_H$ and $T_L$) and minimizing compressor irreversibilities — directly reduces the facility's electricity bill. A 10% improvement in COP on a 500 kW refrigeration system saves approximately 400,000 kWh/year.
:::

***

### Part 4: Power Cycles and Applications {#sec:cycles-15}

#### The Rankine Cycle (Steam Power) {#sec:rankine-15}

The **Rankine cycle** is the standard cycle for steam power plants — coal, nuclear, and biomass electricity generation all use variations of it. It operates with water as the working fluid.

The four processes are:

| Process | Device | Description |
|:--------|:-------|:------------|
| 1 → 2 | **Pump** | Adiabatic compression of saturated liquid (low work) |
| 2 → 3 | **Boiler** | Constant-pressure heat addition — liquid heats, boils, superheats |
| 3 → 4 | **Turbine** | Adiabatic expansion — produces work output |
| 4 → 1 | **Condenser** | Constant-pressure heat rejection — vapor condenses to liquid |

\begin{equation}
\eta_{Rankine} = \frac{w_{turbine} - w_{pump}}{q_{boiler}} = \frac{(h_3 - h_4) - (h_2 - h_1)}{h_3 - h_2}
(\#eq:rankine-efficiency-15)
\end{equation}

```{r rankine-ts-15, fig.cap="T–s diagram for the ideal Rankine cycle showing the saturation dome (liquid–vapor boundary). Process 1→2 is liquid pumping (nearly vertical on the left of the dome), 2→3 is boiler heat addition (horizontal then up and right for superheating), 3→4 is turbine expansion (vertical drop), and 4→1 is condensation (horizontal return along the bottom of the dome). The enclosed area represents the net work output.", fig.align='center', echo=FALSE}
library(ggplot2)

# Approximate saturation dome shape for water (schematic, not exact thermodynamic data)
s_f <- seq(0.1, 1.3, length.out = 60)
T_f <- 50 + 270 * (s_f / 1.3)^0.6   # Saturated liquid curve

s_g <- seq(1.3, 8.0, length.out = 60)
T_g <- 374 - 320 * ((s_g - 1.3) / 6.7)^0.55  # Saturated vapor curve

dome_T <- c(T_f, rev(T_g))
dome_s <- c(s_f, rev(s_g))
df_dome <- data.frame(s = dome_s, T = dome_T)

# Rankine cycle state points (approximate, schematic values)
# State 1: saturated liquid after condenser (T=45°C, s_f ≈ 0.64)
s1 <- 0.64; T1 <- 45
# State 2: compressed liquid after pump (barely changes s)
s2 <- 0.65; T2 <- 46
# State 3: superheated steam after boiler (T=500°C, s≈6.8)
s3 <- 6.80; T3 <- 500
# State 4: wet steam after turbine (s=s3=6.8, T_cond=45°C)
s4 <- 6.80; T4 <- 45

cycle_s <- c(s1, s2, s3, s4, s1)
cycle_T <- c(T1, T2, T3, T4, T1)
df_cycle <- data.frame(s = cycle_s, T = cycle_T)

# Shade the cycle interior (approximate)
df_shade <- data.frame(s = cycle_s[-5], T = cycle_T[-5])

ggplot() +
  geom_polygon(data = df_shade, aes(x = s, y = T), fill = "#457B9D", alpha = 0.2) +
  geom_path(data = df_dome, aes(x = s, y = T), colour = "#264653",
            linewidth = 1.0, linetype = "dashed") +
  geom_path(data = df_cycle, aes(x = s, y = T), colour = "#E63946", linewidth = 1.4) +
  # State points
  geom_point(data = data.frame(s = c(s1,s2,s3,s4), T = c(T1,T2,T3,T4)),
             aes(x = s, y = T), size = 3.5, colour = "#E63946") +
  annotate("text", x = s1-0.2, y = T1,    label = "1\nCondenser out", hjust=1, size=3.0) +
  annotate("text", x = s2-0.2, y = T2+15, label = "2\nPump out",      hjust=1, size=3.0) +
  annotate("text", x = s3+0.1, y = T3,    label = "3\nTurbine in",    hjust=0, size=3.0) +
  annotate("text", x = s4+0.1, y = T4+15, label = "4\nTurbine out",   hjust=0, size=3.0) +
  # Process annotations
  annotate("text", x = 3.7,  y = 420, label = "Boiler (2→3)", size = 3.2,
           colour = "#2A9D8F", fontface = "italic") +
  annotate("text", x = 6.9,  y = 270, label = "Turbine\n(3→4)", size = 3.2,
           colour = "#2A9D8F", fontface = "italic") +
  annotate("text", x = 3.7,  y = 30,  label = "Condenser (4→1)", size = 3.2,
           colour = "#2A9D8F", fontface = "italic") +
  annotate("text", x = 0.55, y = 200, label = "Pump\n(1→2)", size = 3.0,
           colour = "#2A9D8F", fontface = "italic") +
  # Critical point marker
  annotate("point", x = 1.3,  y = 374, size = 4, colour = "#E63946") +
  annotate("text",  x = 1.5,  y = 374, label = "Critical\nPoint", hjust = 0, size = 3.0) +
  scale_x_continuous(limits = c(-0.2, 9.0)) +
  scale_y_continuous(limits = c(-10, 550)) +
  labs(
    title = "T–s Diagram: Ideal Rankine Cycle",
    x     = "Specific Entropy, s (kJ/kg·K)",
    y     = "Temperature, T (°C)"
  ) +
  theme_thermo
```

**Worked Example 12: Rankine Cycle Turbine Power**

Steam enters a turbine (State 3) at $h_3 = 3{,}450 \text{ kJ/kg}$ and exits (State 4) at $h_4 = 2{,}150 \text{ kJ/kg}$. The pump requires $w_p = 4 \text{ kJ/kg}$. The boiler adds $q_{boiler} = h_3 - h_2 = 3{,}200 \text{ kJ/kg}$. Mass flow rate $\dot{m} = 15 \text{ kg/s}$.

*Solution:*

$$w_{turbine} = h_3 - h_4 = 3{,}450 - 2{,}150 = 1{,}300 \text{ kJ/kg}$$

$$\dot{W}_{turbine} = 15 \times 1{,}300 = \mathbf{19{,}500 \text{ kW} = 19.5 \text{ MW}}$$

$$\eta_{Rankine} = \frac{1{,}300 - 4}{3{,}200} = \frac{1{,}296}{3{,}200} = \mathbf{40.5\%}$$

#### The Otto Cycle (Spark-Ignition Gasoline Engine) {#sec:otto-15}

The **Otto cycle** is the ideal cycle for spark-ignition internal combustion engines (gasoline/petrol engines). It consists of four internally reversible processes:

| Process | Description |
|:--------|:------------|
| 1 → 2 | Isentropic compression |
| 2 → 3 | Constant-volume heat addition (combustion) |
| 3 → 4 | Isentropic expansion (power stroke) |
| 4 → 1 | Constant-volume heat rejection (exhaust blowdown) |

The **compression ratio** is:

\begin{equation}
r = \frac{V_1}{V_2} = \frac{V_{BDC}}{V_{TDC}}
(\#eq:compression-ratio-15)
\end{equation}

The thermal efficiency of the ideal Otto cycle depends only on the compression ratio:

\begin{equation}
\eta_{Otto} = 1 - \frac{1}{r^{\gamma - 1}}
(\#eq:otto-efficiency-15)
\end{equation}

**Worked Example 10: Otto Cycle Efficiency**

A gasoline engine has a compression ratio of $r = 8$ and uses air as the working fluid ($\gamma = 1.4$). Find the thermal efficiency.

*Solution:*

$$\eta_{Otto} = 1 - \frac{1}{8^{1.4-1}} = 1 - \frac{1}{8^{0.4}}$$

$$8^{0.4} = e^{0.4 \ln 8} = e^{0.4 \times 2.079} = e^{0.832} = 2.298$$

$$\eta_{Otto} = 1 - \frac{1}{2.298} = 1 - 0.435 = \mathbf{0.565 = 56.5\%}$$

This is the **ideal** (air-standard) efficiency. Real gasoline engines achieve 30–40% due to heat losses, friction, incomplete combustion, and valve timing limitations.

```{r otto-ts-15, fig.cap="T–s diagram for the ideal Otto cycle. Process 1→2 is isentropic compression (vertical line, no entropy change, temperature rises). Process 2→3 is constant-volume combustion (entropy increases, temperature spikes). Process 3→4 is isentropic expansion (vertical drop). Process 4→1 is constant-volume heat rejection (entropy decreases). The enclosed area equals the net work output per cycle.", fig.align='center', echo=FALSE}
library(ggplot2)

# Otto cycle on T-s diagram (schematic, not exact air data)
# State 1: BDC before compression
T1 <- 300; s1 <- 1.0
# State 2: TDC after isentropic compression (T rises, s constant)
T2 <- 689; s2 <- 1.0   # T2 = T1 * r^(gamma-1) = 300 * 8^0.4 ≈ 689 K
# State 3: TDC after combustion (constant volume, s rises)
T3 <- 2500; s3 <- 3.2
# State 4: BDC after isentropic expansion (T drops, s constant)
T4 <- 1088; s4 <- 3.2  # T4/T3 = (V3/V4)^(gamma-1) = (1/r)^(gamma-1)

cycle_s <- c(s1, s2, s3, s4, s1)
cycle_T <- c(T1, T2, T3, T4, T1)
df_cycle <- data.frame(s = cycle_s, T = cycle_T)
df_shade <- data.frame(s = cycle_s[-5], T = cycle_T[-5])

ggplot() +
  geom_polygon(data = df_shade, aes(x = s, y = T), fill = "#E9C46A", alpha = 0.3) +
  geom_path(data = df_cycle, aes(x = s, y = T), colour = "#264653", linewidth = 1.4) +
  geom_point(data = data.frame(s = c(s1,s2,s3,s4), T = c(T1,T2,T3,T4)),
             aes(x = s, y = T), size = 3.5, colour = "#E63946") +
  # State labels
  annotate("text", x = s1-0.1, y = T1-30, label = "1\n(BDC)", hjust=1, size=3.2, colour="#264653") +
  annotate("text", x = s2-0.1, y = T2+30, label = "2\n(TDC)", hjust=1, size=3.2, colour="#264653") +
  annotate("text", x = s3+0.1, y = T3,    label = "3\n(TDC\nPost-combustion)", hjust=0, size=3.0, colour="#264653") +
  annotate("text", x = s4+0.1, y = T4,    label = "4\n(BDC\nPost-expansion)", hjust=0, size=3.0, colour="#264653") +
  # Process annotations
  annotate("text", x = (s1+s2)/2, y = (T1+T2)/2 + 60,
           label = "1→2\nIsentropic\nCompression", hjust=0.5, size=2.8, colour="#457B9D") +
  annotate("text", x = (s2+s3)/2 - 0.1, y = (T2+T3)/2 + 100,
           label = "2→3\nConst. Vol.\nHeat Addition", hjust=1, size=2.8, colour="#E63946") +
  annotate("text", x = (s3+s4)/2, y = (T3+T4)/2 - 60,
           label = "3→4\nIsentropic\nExpansion", hjust=0.5, size=2.8, colour="#457B9D") +
  annotate("text", x = (s4+s1)/2 + 0.1, y = (T4+T1)/2 - 100,
           label = "4→1\nConst. Vol.\nHeat Rejection", hjust=0, size=2.8, colour="#2A9D8F") +
  scale_x_continuous(limits = c(0.3, 4.2)) +
  scale_y_continuous(limits = c(150, 2800)) +
  labs(
    title = "T–s Diagram: Ideal Otto Cycle",
    x     = "Specific Entropy, s (kJ/kg·K)",
    y     = "Temperature, T (K)"
  ) +
  theme_thermo
```

#### The Diesel Cycle (Compression-Ignition Engine) {#sec:diesel-15}

The **Diesel cycle** models compression-ignition engines (diesel engines). The key difference from the Otto cycle: combustion occurs at **constant pressure** (not constant volume), allowing higher compression ratios without knock.

The **cutoff ratio** defines how far the piston travels during combustion:

\begin{equation}
r_c = \frac{V_3}{V_2} \quad \text{(volume ratio during constant-pressure combustion)}
(\#eq:cutoff-ratio-15)
\end{equation}

\begin{equation}
\eta_{Diesel} = 1 - \frac{1}{r^{\gamma-1}} \cdot \frac{r_c^{\gamma} - 1}{\gamma(r_c - 1)}
(\#eq:diesel-efficiency-15)
\end{equation}

**Worked Example 11: Diesel Cycle Efficiency**

A diesel engine has compression ratio $r = 18$ and cutoff ratio $r_c = 2.5$, with $\gamma = 1.4$.

*Solution:*

$$r^{\gamma-1} = 18^{0.4} = e^{0.4 \ln 18} = e^{0.4 \times 2.890} = e^{1.156} = 3.177$$

$$\frac{r_c^{\gamma} - 1}{\gamma(r_c - 1)} = \frac{2.5^{1.4} - 1}{1.4 \times (2.5 - 1)} = \frac{3.285 - 1}{1.4 \times 1.5} = \frac{2.285}{2.1} = 1.088$$

$$\eta_{Diesel} = 1 - \frac{1.088}{3.177} = 1 - 0.342 = \mathbf{0.658 = 65.8\%}$$

Note: The Diesel cycle efficiency (ideal) is lower than the Otto cycle at the same compression ratio, but diesel engines run at much higher compression ratios (16–22 vs. 8–12 for gasoline), so real diesel engines typically achieve higher thermal efficiencies (40–50%) than real gasoline engines.

#### The Brayton Cycle (Gas Turbine) {#sec:brayton-15}

The **Brayton cycle** is the ideal cycle for gas turbine engines — used in aircraft jet engines, helicopter turbines, and industrial power generation.

| Process | Device | Description |
|:--------|:-------|:------------|
| 1 → 2 | **Compressor** | Isentropic compression of air |
| 2 → 3 | **Combustor** | Constant-pressure heat addition (fuel combustion) |
| 3 → 4 | **Turbine** | Isentropic expansion |
| 4 → 1 | **Exhaust** | Constant-pressure heat rejection (atmosphere) |

The **pressure ratio** $r_p = P_2/P_1$ governs efficiency:

\begin{equation}
\eta_{Brayton} = 1 - \frac{1}{r_p^{(\gamma-1)/\gamma}}
(\#eq:brayton-efficiency-15)
\end{equation}

For $r_p = 10$ and air ($\gamma = 1.4$):

$$\eta_{Brayton} = 1 - \frac{1}{10^{0.4/1.4}} = 1 - \frac{1}{10^{0.2857}} = 1 - \frac{1}{1.931} = 1 - 0.518 = \mathbf{48.2\%}$$

::: {.callout-industry}
**Industry Application:** The General Electric CF34 turbofan engine (used on many regional jets) achieves a pressure ratio of approximately 28:1 in the core compressor. Modern high-bypass turbofan engines have overall pressure ratios exceeding 50:1. Defense applications such as the Rolls-Royce Adour (used in the BAE Hawk trainer) use similar thermodynamic principles but are optimized for high thrust-to-weight ratio rather than fuel efficiency.
:::

#### Vapor-Compression Refrigeration Cycle {#sec:refrigeration-cycle-15}

The **vapor-compression refrigeration cycle** is used in virtually every modern refrigerator, air conditioner, and heat pump. It is the reverse of the Rankine power cycle.

| Process | Device | Description |
|:--------|:-------|:------------|
| 1 → 2 | **Compressor** | Isentropic compression of low-pressure vapor to high-pressure superheated vapor |
| 2 → 3 | **Condenser** | Constant-pressure heat rejection — superheated vapor condenses to liquid |
| 3 → 4 | **Expansion valve** | Isenthalpic throttling — liquid drops in pressure and partially flashes |
| 4 → 1 | **Evaporator** | Constant-pressure heat absorption — wet mixture evaporates (this is the refrigeration effect) |

The refrigeration effect: $q_L = h_1 - h_4$

Compressor work: $w_{comp} = h_2 - h_1$

$$\text{COP}_{refrig} = \frac{h_1 - h_4}{h_2 - h_1}$$

Common refrigerants include R-134a (automotive AC), R-410A (residential AC), R-744 (CO$_2$, food retail), and ammonia (R-717, industrial).

::: {.callout-industry}
**Industry Application:** Modern electric vehicles use a **heat pump** system (reversible refrigeration cycle) for cabin heating and cooling. Rather than an inefficient resistance heater that converts 1 kWh of electricity to 1 kWh of heat, a heat pump can deliver 2.5–4 kWh of heat per kWh of electricity consumed (COP = 2.5–4.0). This significantly extends EV range in cold climates, where heating is the dominant energy consumer at low speeds.
:::

#### Heat Exchangers — LMTD Method {#sec:heat-exchangers-15}

A **heat exchanger** transfers heat between two fluid streams without mixing them. The **Log Mean Temperature Difference (LMTD)** method calculates the required heat transfer area:

\begin{equation}
\dot{Q} = U A \cdot \text{LMTD}
(\#eq:lmtd-q-15)
\end{equation}

\begin{equation}
\text{LMTD} = \frac{\Delta T_1 - \Delta T_2}{\ln(\Delta T_1 / \Delta T_2)}
(\#eq:lmtd-15)
\end{equation}

where $\Delta T_1$ and $\Delta T_2$ are the hot–cold temperature differences at each end of the exchanger, $U$ is the overall heat transfer coefficient (W/m²·K), and $A$ is the heat transfer surface area.

For a **counterflow** exchanger (most efficient): the two fluids flow in opposite directions, maximizing the temperature driving force throughout.

***

### Cycle Efficiency Comparison {#sec:efficiency-comparison-15}

```{r efficiency-comparison-15, fig.cap="Comparison of ideal thermal efficiencies for major thermodynamic power cycles at representative operating conditions. The Carnot efficiency (at T_H=800 K, T_L=300 K) sets the theoretical maximum. Real engines always fall below their ideal cycle efficiency due to irreversibilities, heat losses, and friction.", fig.align='center', echo=FALSE}
library(ggplot2)

cycles <- data.frame(
  cycle = c("Carnot\n(T_H=800K, T_L=300K)",
            "Diesel\n(r=18, r_c=2.5)",
            "Brayton\n(r_p=10)",
            "Otto\n(r=8)",
            "Rankine\n(Superheated)"),
  efficiency = c(62.5, 65.8, 48.2, 56.5, 40.5),
  type = c("Theoretical Limit", "Compression-Ignition", "Gas Turbine",
           "Spark-Ignition", "Steam Power")
)

# Reorder for display
cycles$cycle <- factor(cycles$cycle, levels = cycles$cycle[order(cycles$efficiency)])

cols <- c("Theoretical Limit"    = "#E63946",
          "Compression-Ignition" = "#264653",
          "Gas Turbine"          = "#2A9D8F",
          "Spark-Ignition"       = "#E9C46A",
          "Steam Power"          = "#457B9D")

ggplot(cycles, aes(x = cycle, y = efficiency, fill = type)) +
  geom_col(width = 0.6, colour = "white") +
  geom_text(aes(label = paste0(efficiency, "%")), vjust = -0.4, size = 4, fontface = "bold") +
  scale_fill_manual(values = cols) +
  scale_y_continuous(limits = c(0, 80), breaks = seq(0, 80, 10)) +
  labs(
    title    = "Ideal Thermal Efficiency — Major Power Cycles",
    subtitle = "Note: Real efficiencies are 60–70% of ideal values shown",
    x        = NULL,
    y        = "Thermal Efficiency (%)",
    fill     = "Cycle Type"
  ) +
  theme_thermo +
  theme(legend.position = "right")
```

```{r cop-comparison-15, fig.cap="Coefficient of Performance (COP) for refrigerators and heat pumps at the same Carnot conditions (T_L = 250 K, T_H = 300 K). A heat pump always has a COP exactly 1.0 higher than the corresponding refrigerator. Both exceed a COP of 1.0, which is why heat pumps are far more energy-efficient than resistance heaters.", fig.align='center', echo=FALSE}
library(ggplot2)

# COP vs temperature lift (T_H - T_L), keeping T_L = 250 K
T_L <- 250
T_H_vals <- seq(260, 340, by = 5)
delta_T   <- T_H_vals - T_L

COP_R <- T_L / delta_T          # Carnot COP refrigerator
COP_HP <- T_H_vals / delta_T    # Carnot COP heat pump

df_cop <- data.frame(
  delta_T = rep(delta_T, 2),
  COP     = c(COP_R, COP_HP),
  Device  = rep(c("Refrigerator (COP_R)", "Heat Pump (COP_HP)"), each = length(delta_T))
)

ggplot(df_cop, aes(x = delta_T, y = COP, colour = Device)) +
  geom_line(linewidth = 1.4) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 75, y = 1.15, label = "COP = 1.0\n(electric resistance heater baseline)",
           hjust = 1, size = 3.0, colour = "grey40") +
  scale_colour_manual(values = c("Refrigerator (COP_R)" = "#2A9D8F",
                                  "Heat Pump (COP_HP)"   = "#E63946")) +
  scale_x_continuous(breaks = seq(10, 90, 10)) +
  labs(
    title    = "Carnot COP vs. Temperature Lift (T_L = 250 K)",
    subtitle = "COP_HP = COP_R + 1 at all conditions",
    x        = "Temperature Lift, T_H − T_L (K)",
    y        = "Coefficient of Performance (COP)",
    colour   = NULL
  ) +
  theme_thermo
```

***

### Key Takeaways {#sec:thermo-takeaways-15}

1. **Systems and processes:** Define your system boundary first. Open systems exchange mass and energy; closed systems exchange energy only. Process type (isothermal, isobaric, isochoric, adiabatic) determines which boundary work formula applies.

2. **First Law:** Energy is always conserved: $Q - W = \Delta U$. For open systems at steady state, use the SFEE: $q - w_s = \Delta h + \Delta KE + \Delta PE$.

3. **Ideal Gas Law:** $PV = nRT$ or $Pv = RT$. Use absolute temperature (Kelvin) and absolute pressure always.

4. **Second Law:** No cyclic heat engine can be 100% efficient. The Carnot efficiency $\eta = 1 - T_L/T_H$ sets the upper limit. Every real process generates entropy.

5. **Efficiency and COP:** Heat engines have $\eta < 1$. Refrigerators and heat pumps have $\text{COP} > 1$. COP$_{HP}$ = COP$_{refrig}$ + 1.

6. **Cycle performance:** Otto efficiency depends on compression ratio; Diesel adds the cutoff ratio correction; Rankine and Brayton are analyzed using enthalpies from steam tables and air properties respectively.

***

### Practice Problems {#sec:thermo-practice-15}

1. A gas at 150 kPa gauge pressure is at 85°C. Convert both values to SI absolute units (kPa abs, Kelvin) and to US customary units (psia, °F).

2. Air is compressed isothermally from 100 kPa, 2.0 m³ to 500 kPa. Calculate the final volume, the boundary work done on the air, and the heat transfer (if any).

3. A rigid tank contains 3 kg of air at 27°C. The tank is heated until the temperature reaches 127°C. Find the heat added and the final pressure if the initial pressure was 200 kPa.

4. A Carnot engine operates between 900 K and 350 K and produces 500 kW of power. Find: (a) thermal efficiency, (b) heat input rate $\dot{Q}_H$, and (c) heat rejection rate $\dot{Q}_L$.

5. A diesel engine has a compression ratio of 20 and a cutoff ratio of 2.0 ($\gamma = 1.4$). Compare its efficiency with an Otto cycle at the same compression ratio.

6. A refrigeration system maintains a cold room at −10°C in a factory where the ambient temperature is 30°C. If the actual COP is 60% of the Carnot COP, and the refrigeration load is 50 kW, what is the electrical power consumed?

7. Steam enters a turbine at $h_1 = 3{,}500 \text{ kJ/kg}$ and exits with a quality of $x = 0.90$ at $T_{sat} = 40°\text{C}$ where $h_f = 167.5 \text{ kJ/kg}$ and $h_{fg} = 2{,}432 \text{ kJ/kg}$. If the mass flow rate is 8 kg/s, find the power output (assume adiabatic).

8. A counter-flow heat exchanger heats oil from 30°C to 90°C using water that enters at 120°C and exits at 70°C. If $U = 500 \text{ W/m}^2\text{·K}$ and the heat duty is 200 kW, find the required surface area $A$.

***

### Formula Summary {#sec:thermo-formulas-15}

| Formula Name | Equation | Variables | Conditions / Notes |
|:-------------|:---------|:----------|:------------------|
| **Temperature: °C to K** | $T_K = T_C + 273.15$ | $T_K$ = Kelvin, $T_C$ = Celsius | Always use K in thermo equations |
| **Temperature: °C to °F** | $T_F = \tfrac{9}{5}T_C + 32$ | $T_F$ = Fahrenheit | Conversion only |
| **Temperature: °F to R** | $T_R = T_F + 459.67$ | $T_R$ = Rankine | Absolute scale, US customary |
| **Temperature: K to R** | $T_R = 1.8 \, T_K$ | — | Direct conversion |
| **Absolute Pressure** | $P_{abs} = P_{gauge} + P_{atm}$ | $P_{atm} = 101.325 \text{ kPa}$ | Always convert to absolute for gas laws |
| **Vacuum Pressure** | $P_{vac} = P_{atm} - P_{abs}$ | — | For pressures below atmospheric |
| **Steam Quality** | $x = m_{vap} / m_{total}$ | $0 \leq x \leq 1$ | Wet mixture region only |
| **Mixture Property** | $y = y_f + x \, y_{fg}$ | $y_{fg} = y_g - y_f$ | Any specific property (h, u, s, v) |
| **First Law (Closed System)** | $Q - W = \Delta U$ | $Q>0$ in; $W>0$ out | Per-cycle or per-process |
| **Enthalpy** | $H = U + PV$ ; $h = u + Pv$ | $h$ = specific enthalpy (kJ/kg) | Useful for open systems and const-P |
| **Internal Energy Change (Ideal Gas)** | $\Delta U = m c_v \Delta T$ | $c_v$ = specific heat at const vol | Ideal gas only |
| **Enthalpy Change (Ideal Gas)** | $\Delta H = m c_p \Delta T$ | $c_p$ = specific heat at const press | Ideal gas only |
| **Specific Heat Relation** | $c_p - c_v = R$ | $R$ = specific gas constant (kJ/kg·K) | Ideal gas only |
| **Ratio of Specific Heats** | $\gamma = c_p / c_v$ | Air: $\gamma = 1.4$ | Used in adiabatic/isentropic relations |
| **Ideal Gas Law (molar)** | $PV = nR_uT$ | $R_u = 8.314 \text{ J/mol·K}$ | $n$ = moles |
| **Ideal Gas Law (specific)** | $Pv = RT$ | $R = R_u / M_w$ (kJ/kg·K) | $v$ = specific volume |
| **Combined Gas Law** | $P_1V_1/T_1 = P_2V_2/T_2$ | Absolute T and P | Fixed mass of ideal gas |
| **Boundary Work (General)** | $W = \int P \, dV$ | — | Closed system only |
| **Work: Isochoric** | $W = 0$ | $V = \text{const}$ | No volume change |
| **Work: Isobaric** | $W = P(V_2 - V_1) = mR\Delta T$ | $P = \text{const}$ | — |
| **Work: Isothermal** | $W = P_1V_1 \ln(V_2/V_1)$ | $T = \text{const}$, ideal gas | $= mRT\ln(V_2/V_1)$ |
| **Work: Adiabatic** | $W = (P_1V_1 - P_2V_2)/(\gamma-1)$ | $Q=0$, ideal gas | $= mR(T_1-T_2)/(\gamma-1)$ |
| **Adiabatic Temperature Ratio** | $T_2/T_1 = (P_2/P_1)^{(\gamma-1)/\gamma}$ | Isentropic process | Ideal gas |
| **SFEE (per unit mass)** | $q - w_s = \Delta h + \Delta KE + \Delta PE$ | $w_s$ = shaft work out | Open system, steady state |
| **Thermal Efficiency** | $\eta_{th} = W_{net}/Q_H = 1 - Q_L/Q_H$ | $W_{net} = Q_H - Q_L$ | Any heat engine |
| **Carnot Efficiency** | $\eta_{Carnot} = 1 - T_L/T_H$ | $T$ in Kelvin | Maximum possible; reversible cycle |
| **Entropy Change (Reversible)** | $dS = \delta Q_{rev}/T$ | $S$ in J/K; $s$ in J/kg·K | Reversible heat transfer only |
| **Entropy Change (Ideal Gas, P)** | $\Delta s = c_p \ln(T_2/T_1) - R\ln(P_2/P_1)$ | — | Between any two states |
| **Entropy Change (Ideal Gas, v)** | $\Delta s = c_v \ln(T_2/T_1) + R\ln(v_2/v_1)$ | — | Between any two states |
| **COP — Refrigerator** | $\text{COP}_R = Q_L/W_{in} = Q_L/(Q_H-Q_L)$ | $Q_L$ = cooling load | Can exceed 1.0 |
| **COP — Heat Pump** | $\text{COP}_{HP} = Q_H/W_{in}$ | — | $= \text{COP}_R + 1$ always |
| **COP — Carnot Refrigerator** | $\text{COP}_{R,C} = T_L/(T_H-T_L)$ | $T$ in Kelvin | Maximum possible |
| **Otto Cycle Efficiency** | $\eta_{Otto} = 1 - 1/r^{\gamma-1}$ | $r = V_1/V_2$ (compression ratio) | Ideal, air-standard |
| **Diesel Cycle Efficiency** | $\eta_{Diesel} = 1 - \frac{1}{r^{\gamma-1}} \cdot \frac{r_c^\gamma - 1}{\gamma(r_c-1)}$ | $r_c = V_3/V_2$ (cutoff ratio) | Ideal, air-standard |
| **Brayton Cycle Efficiency** | $\eta_{Brayton} = 1 - 1/r_p^{(\gamma-1)/\gamma}$ | $r_p = P_2/P_1$ (pressure ratio) | Ideal, air-standard |
| **Rankine Cycle Efficiency** | $\eta_{Rankine} = (w_t - w_p)/q_b = [(h_3-h_4)-(h_2-h_1)]/(h_3-h_2)$ | $h$ from steam tables | Ideal Rankine |
| **Heat Exchanger (LMTD)** | $\dot{Q} = U A \cdot \text{LMTD}$ | $U$ = overall coeff. (W/m²·K) | — |
| **LMTD** | $\text{LMTD} = (\Delta T_1 - \Delta T_2)/\ln(\Delta T_1/\Delta T_2)$ | $\Delta T$ = hot–cold temp diff at each end | Counterflow or parallel flow |
