## Powder Metallurgy and Sintering {#chap-sintering}

```{r setup-sintering, include=FALSE}
## include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
## include the required packages to make sure the page can be built.
source("R/required_packages.R")
```

### Learning Objectives {#sec:objectives-sintering}

By the end of this chapter, you will be able to:

1. **Explain** the fundamental principles of powder metallurgy and the sintering process
2. **Identify** the key stages of powder metallurgy: powder production, compaction, and sintering
3. **Compare** different density levels achievable through sintering and their applications
4. **Design** sintered components for specific applications including self-lubricating bushings
5. **Evaluate** the advantages and limitations of powder metallurgy compared to other manufacturing processes

::: {.callout-think}
**Before You Read:** Have you ever wondered how bushings in electric motors or automotive components can run for years without external lubrication? What manufacturing process could create a part that stores its own lubricant?
:::

### Introduction to Powder Metallurgy {#sec:pm-intro}

Powder metallurgy (PM) is a manufacturing process that creates metal parts from powdered materials through compaction and heat treatment. Unlike casting, which starts with molten metal, or machining, which removes material from a solid block, powder metallurgy builds parts by pressing metal powder into a die and then heating the compact to bond the particles together.

The key distinction of powder metallurgy is that the metal never fully melts. Instead, during sintering, the particles bond at their contact points through solid-state diffusion—atoms migrate across particle boundaries, creating metallurgical bonds while maintaining the overall shape established during compaction.

This unique approach enables several capabilities impossible with other manufacturing methods:

- **Controlled porosity:** Parts can be made with precise levels of interconnected voids
- **Near-net-shape production:** Parts require minimal or no machining
- **Material combinations:** Metals that cannot be alloyed conventionally can be combined
- **High material utilization:** Over 97% of the starting powder becomes the final part

---

### The Powder Metallurgy Process {#sec:pm-process}

The powder metallurgy process consists of four main stages: powder production, blending, compaction, and sintering.

```{r fig-pm-process-flow, echo=FALSE, fig.cap="The four main stages of the powder metallurgy process.", fig.align='center', fig.width=12, fig.height=5}
library(ggplot2)

ggplot() +
  # Stage 1: Powder Production
  annotate("text", x = 1.5, y = 5.2, label = "1. POWDER\nPRODUCTION", fontface = "bold", size = 3.5) +
  geom_rect(aes(xmin = 0.5, xmax = 2.5, ymin = 2.5, ymax = 4.8), fill = "lightgray", color = "black", linewidth = 1) +
  # Powder particles
  geom_point(aes(x = c(1, 1.3, 1.6, 1.9, 1.2, 1.5, 1.8, 1.1, 1.4, 1.7, 2, 1.25, 1.55, 1.85),
                 y = c(3, 3.2, 3.1, 3.3, 3.5, 3.7, 3.6, 4, 4.2, 4.1, 4.3, 4.5, 3.9, 3.4)),
             size = 3, color = "darkgray") +
  annotate("text", x = 1.5, y = 2, label = "Atomization,\nreduction, or\nelectrolysis", size = 2.5, fontface = "italic") +

  # Stage 2: Blending
  annotate("text", x = 4.5, y = 5.2, label = "2. BLENDING", fontface = "bold", size = 3.5) +
  geom_rect(aes(xmin = 3.5, xmax = 5.5, ymin = 2.5, ymax = 4.8), fill = "lightyellow", color = "black", linewidth = 1) +
  # Mixed powder particles
  geom_point(aes(x = c(4, 4.3, 4.6, 4.9, 4.2, 4.5, 4.8, 4.1, 4.4, 4.7, 5, 4.25, 4.55, 4.85),
                 y = c(3, 3.2, 3.1, 3.3, 3.5, 3.7, 3.6, 4, 4.2, 4.1, 4.3, 4.5, 3.9, 3.4)),
             size = 3, color = c("darkgray", "steelblue", "darkgray", "steelblue", "darkgray",
                                  "steelblue", "darkgray", "steelblue", "darkgray", "steelblue",
                                  "darkgray", "steelblue", "darkgray", "steelblue")) +
  annotate("text", x = 4.5, y = 2, label = "Mix with lubricants\nand additives", size = 2.5, fontface = "italic") +

  # Stage 3: Compaction
  annotate("text", x = 7.5, y = 5.2, label = "3. COMPACTION", fontface = "bold", size = 3.5) +
  geom_rect(aes(xmin = 6.5, xmax = 8.5, ymin = 2.5, ymax = 4.8), fill = "lightblue", color = "black", linewidth = 1) +
  # Die walls
  geom_rect(aes(xmin = 6.8, xmax = 7.1, ymin = 2.7, ymax = 4.6), fill = "gray40", color = "black") +
  geom_rect(aes(xmin = 7.9, xmax = 8.2, ymin = 2.7, ymax = 4.6), fill = "gray40", color = "black") +
  # Compacted part
  geom_rect(aes(xmin = 7.1, xmax = 7.9, ymin = 3.2, ymax = 4.2), fill = "tan", color = "black") +
  # Upper punch
  geom_rect(aes(xmin = 7.1, xmax = 7.9, ymin = 4.2, ymax = 4.6), fill = "gray50", color = "black") +
  # Lower punch
  geom_rect(aes(xmin = 7.1, xmax = 7.9, ymin = 2.7, ymax = 3.2), fill = "gray50", color = "black") +
  # Arrows showing pressure
  geom_segment(aes(x = 7.5, xend = 7.5, y = 5, yend = 4.7),
               arrow = arrow(length = unit(0.2, "cm")), linewidth = 1.2, color = "red") +
  geom_segment(aes(x = 7.5, xend = 7.5, y = 2.3, yend = 2.6),
               arrow = arrow(length = unit(0.2, "cm")), linewidth = 1.2, color = "red") +
  annotate("text", x = 7.5, y = 2, label = "Press at\n400-700 MPa", size = 2.5, fontface = "italic") +

  # Stage 4: Sintering
  annotate("text", x = 10.5, y = 5.2, label = "4. SINTERING", fontface = "bold", size = 3.5) +
  geom_rect(aes(xmin = 9.5, xmax = 11.5, ymin = 2.5, ymax = 4.8), fill = "coral", color = "black", linewidth = 1) +
  # Sintered part
  geom_rect(aes(xmin = 10, xmax = 11, ymin = 3.2, ymax = 4.2), fill = "sienna", color = "black", linewidth = 1.5) +
  # Heat waves
  geom_curve(aes(x = 9.7, xend = 9.9, y = 3.5, yend = 3.7), curvature = 0.3, color = "orange", linewidth = 0.8) +
  geom_curve(aes(x = 9.7, xend = 9.9, y = 3.8, yend = 4), curvature = 0.3, color = "orange", linewidth = 0.8) +
  geom_curve(aes(x = 11.1, xend = 11.3, y = 3.5, yend = 3.7), curvature = -0.3, color = "orange", linewidth = 0.8) +
  geom_curve(aes(x = 11.1, xend = 11.3, y = 3.8, yend = 4), curvature = -0.3, color = "orange", linewidth = 0.8) +
  annotate("text", x = 10.5, y = 2, label = "Heat to 70-90%\nof melting point", size = 2.5, fontface = "italic") +

  # Arrows between stages
  geom_segment(aes(x = 2.7, xend = 3.3, y = 3.7, yend = 3.7),
               arrow = arrow(length = unit(0.2, "cm")), linewidth = 1) +
  geom_segment(aes(x = 5.7, xend = 6.3, y = 3.7, yend = 3.7),
               arrow = arrow(length = unit(0.2, "cm")), linewidth = 1) +
  geom_segment(aes(x = 8.7, xend = 9.3, y = 3.7, yend = 3.7),
               arrow = arrow(length = unit(0.2, "cm")), linewidth = 1) +

  coord_fixed(ratio = 1, xlim = c(0, 12), ylim = c(1.5, 5.8)) +
  theme_void() +
  labs(title = "Powder Metallurgy Process Flow") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

#### Stage 1: Powder Production {#sec:powder-production}

Metal powders are produced through several methods, each yielding particles with different characteristics:

**Atomization:** Molten metal is dispersed by high-pressure gas or water jets, producing spherical or irregular particles. This is the most common method for iron, steel, copper, and aluminum powders.

**Chemical Reduction:** Metal oxides are reduced using hydrogen or carbon at elevated temperatures. This method is widely used for iron and tungsten powders.

**Electrolysis:** Metal is deposited at a cathode from an electrolyte solution, then ground to powder. This produces very pure, dendritic (tree-like) particles.

**Mechanical Milling:** Solid metal is ground in ball mills, often used to produce alloy powders or modify particle shapes.

Particle size typically ranges from 1 to 200 micrometers, with most PM parts using powders in the 40-150 micrometer range.

#### Stage 2: Blending {#sec:blending}

Raw powders are rarely used directly. Instead, they are blended with:

- **Lubricants** (0.5-1.5%): Such as zinc stearate or lithium stearate, which reduce friction during compaction and ejection from the die
- **Alloying elements:** To achieve desired mechanical properties
- **Graphite:** For carbon content in steel parts (0.5-1.0%)

Thorough blending ensures uniform distribution of all constituents, which is critical for consistent final properties.

#### Stage 3: Compaction {#sec:compaction}

The blended powder is pressed in a rigid die at pressures typically between 400 and 700 MPa (60,000-100,000 psi). During compaction:

1. Particles rearrange and pack more densely
2. Particles deform plastically at contact points
3. Cold welding occurs between particles
4. The lubricant is displaced to particle surfaces

The resulting "green compact" has sufficient strength to be handled but is still relatively fragile. Green density typically reaches 75-90% of the theoretical (fully dense) material.

```{r fig-compaction-pressure, echo=FALSE, fig.cap="Effect of compaction pressure on green density for iron powder.", fig.align='center', fig.width=8, fig.height=5}
library(ggplot2)

# Generate compaction curve data
pressure <- seq(100, 800, by = 25)
# Typical compaction behavior follows a logarithmic relationship
set.seed(42)
density <- 5.2 + 1.6 * log(pressure/100) + rnorm(length(pressure), 0, 0.05)
density <- pmin(density, 7.2)  # Cap at ~92% of theoretical density (7.87)

compaction_data <- data.frame(
  pressure = pressure,
  density = density,
  theoretical_pct = density / 7.87 * 100
)

ggplot(compaction_data, aes(x = pressure, y = density)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_hline(yintercept = 7.87, linetype = "dashed", color = "red") +
  annotate("text", x = 700, y = 7.95, label = "Theoretical Density (7.87 g/cm³)",
           color = "red", size = 3.5) +
  geom_hline(yintercept = 6.3, linetype = "dotted", color = "darkgreen") +
  annotate("text", x = 700, y = 6.4, label = "80% - Structural Parts",
           color = "darkgreen", size = 3) +
  geom_hline(yintercept = 5.5, linetype = "dotted", color = "purple") +
  annotate("text", x = 700, y = 5.6, label = "70% - Self-Lubricating Bearings",
           color = "purple", size = 3) +
  labs(
    title = "Compaction Curve for Iron Powder",
    x = "Compaction Pressure (MPa)",
    y = "Green Density (g/cm³)"
  ) +
  scale_y_continuous(
    sec.axis = sec_axis(~ . / 7.87 * 100, name = "Percent Theoretical Density (%)")
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

#### Stage 4: Sintering {#sec:sintering-process}

Sintering is the heart of the powder metallurgy process. The green compact is heated in a controlled atmosphere furnace to a temperature typically 70-90% of the material's absolute melting point. During sintering:

1. **Lubricant burnoff:** At 150-500°C, organic lubricants decompose and are removed
2. **Oxide reduction:** Reducing atmospheres (hydrogen, dissociated ammonia) remove surface oxides
3. **Neck formation:** Atoms diffuse across particle boundaries, forming metallurgical bonds
4. **Pore rounding:** Angular pores become spherical, reducing stress concentrations
5. **Densification:** Pores shrink as material moves from particle interiors to the necks

```{r fig-sintering-stages, echo=FALSE, fig.cap="Particle bonding evolution during sintering: from point contacts to neck growth and pore rounding.", fig.align='center', fig.width=10, fig.height=4}
library(ggplot2)

ggplot() +
  # Stage 1: Initial Contact
  annotate("text", x = 1.5, y = 3.8, label = "Initial\nContact", fontface = "bold", size = 3.5) +
  # Two particles touching at a point
  geom_point(aes(x = 1.2, y = 2), size = 25, color = "gray60") +
  geom_point(aes(x = 1.8, y = 2), size = 25, color = "gray60") +
  annotate("text", x = 1.5, y = 0.8, label = "Point contact\nbetween particles", size = 2.5, fontface = "italic") +

  # Stage 2: Early Sintering
  annotate("text", x = 4.5, y = 3.8, label = "Early\nSintering", fontface = "bold", size = 3.5) +
  geom_point(aes(x = 4.25, y = 2), size = 25, color = "gray50") +
  geom_point(aes(x = 4.75, y = 2), size = 25, color = "gray50") +
  # Small neck
  geom_rect(aes(xmin = 4.4, xmax = 4.6, ymin = 1.8, ymax = 2.2), fill = "gray50") +
  annotate("text", x = 4.5, y = 0.8, label = "Neck begins\nto form", size = 2.5, fontface = "italic") +

  # Stage 3: Intermediate Sintering
  annotate("text", x = 7.5, y = 3.8, label = "Intermediate\nSintering", fontface = "bold", size = 3.5) +
  geom_point(aes(x = 7.3, y = 2), size = 25, color = "gray40") +
  geom_point(aes(x = 7.7, y = 2), size = 25, color = "gray40") +
  # Larger neck
  geom_rect(aes(xmin = 7.35, xmax = 7.65, ymin = 1.7, ymax = 2.3), fill = "gray40") +
  annotate("text", x = 7.5, y = 0.8, label = "Neck growth,\npore rounding", size = 2.5, fontface = "italic") +

  # Stage 4: Final Sintering
  annotate("text", x = 10.5, y = 3.8, label = "Final\nSintering", fontface = "bold", size = 3.5) +
  # Nearly merged particles
  geom_point(aes(x = 10.35, y = 2), size = 25, color = "gray30") +
  geom_point(aes(x = 10.65, y = 2), size = 25, color = "gray30") +
  geom_rect(aes(xmin = 10.25, xmax = 10.75, ymin = 1.6, ymax = 2.4), fill = "gray30") +
  # Small remaining pore
  geom_point(aes(x = 10.5, y = 2.5), size = 4, color = "white") +
  annotate("text", x = 10.5, y = 0.8, label = "Strong bonds,\nisolated pores", size = 2.5, fontface = "italic") +

  # Arrows between stages
  geom_segment(aes(x = 2.3, xend = 3.5, y = 2, yend = 2),
               arrow = arrow(length = unit(0.2, "cm")), linewidth = 0.8) +
  geom_segment(aes(x = 5.3, xend = 6.5, y = 2, yend = 2),
               arrow = arrow(length = unit(0.2, "cm")), linewidth = 0.8) +
  geom_segment(aes(x = 8.3, xend = 9.5, y = 2, yend = 2),
               arrow = arrow(length = unit(0.2, "cm")), linewidth = 0.8) +

  coord_fixed(ratio = 1, xlim = c(0, 12), ylim = c(0.3, 4.2)) +
  theme_void() +
  labs(title = "Evolution of Particle Bonding During Sintering") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

Typical sintering conditions:

| Material | Sintering Temperature (°C) | Time (min) | Atmosphere |
|:---------|:--------------------------:|:----------:|:-----------|
| Iron/Steel | 1100-1300 | 20-45 | H₂, N₂/H₂, endothermic |
| Copper | 750-900 | 15-30 | H₂, N₂/H₂ |
| Bronze | 760-870 | 15-30 | H₂, N₂/H₂ |
| Stainless Steel | 1150-1350 | 30-60 | H₂, vacuum |
| Aluminum | 590-620 | 15-30 | N₂ |

> **Note:** The atmosphere during sintering is critical. Oxygen must be excluded to prevent oxidation, and reducing gases help remove existing oxides from particle surfaces, enabling metallurgical bonding.

---

### Density Control in Sintered Parts {#sec:density-control}

One of the unique advantages of powder metallurgy is the ability to precisely control the final density of the part. Unlike casting or forging, which produce fully dense material, sintered parts can be manufactured with intentional porosity levels tailored to specific applications.

#### Understanding Density Levels {#sec:density-levels}

Sintered density is typically expressed as a percentage of the theoretical density (the density of the fully dense, wrought material). For iron, the theoretical density is 7.87 g/cm³.

```{r fig-density-applications, echo=FALSE, fig.cap="Relationship between sintered density and typical applications.", fig.align='center', fig.width=10, fig.height=6}
library(ggplot2)
library(dplyr)

# Create data for density ranges and applications
density_data <- data.frame(
  category = c("Filters", "Self-Lubricating\nBearings", "Low-Strength\nStructural",
               "Medium-Strength\nStructural", "High-Performance\nStructural", "Fully Dense"),
  min_density = c(40, 70, 80, 85, 90, 98),
  max_density = c(50, 80, 85, 90, 95, 100),
  porosity = c("50-60%", "20-30%", "15-20%", "10-15%", "5-10%", "< 2%"),
  color = c("lightblue", "purple", "darkgreen", "forestgreen", "orange", "red")
)

density_data$category <- factor(density_data$category,
                                 levels = rev(density_data$category))

ggplot(density_data, aes(y = category)) +
  geom_segment(aes(x = min_density, xend = max_density, yend = category, color = category),
               linewidth = 15, lineend = "round") +
  geom_text(aes(x = (min_density + max_density)/2, label = porosity),
            color = "white", fontface = "bold", size = 3.5) +
  scale_color_manual(values = c("Filters" = "lightblue",
                                 "Self-Lubricating\nBearings" = "purple",
                                 "Low-Strength\nStructural" = "darkgreen",
                                 "Medium-Strength\nStructural" = "forestgreen",
                                 "High-Performance\nStructural" = "orange",
                                 "Fully Dense" = "red")) +
  scale_x_continuous(breaks = seq(40, 100, 10), limits = c(35, 105)) +
  labs(
    title = "Sintered Density Ranges and Applications",
    x = "Density (% of Theoretical)",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    legend.position = "none",
    axis.text.y = element_text(size = 10, face = "bold"),
    panel.grid.major.y = element_blank()
  )
```

#### Factors Controlling Final Density {#sec:density-factors}

The final density of a sintered part depends on several controllable parameters:

| Factor | Effect on Density |
|:-------|:------------------|
| Compaction pressure | Higher pressure → higher green and final density |
| Powder characteristics | Finer, spherical particles pack more densely |
| Sintering temperature | Higher temperature → more densification |
| Sintering time | Longer time → more densification |
| Lubricant content | More lubricant → lower green density |
| Powder composition | Alloying elements affect sintering kinetics |

**Double pressing and double sintering (DPDS):** For applications requiring densities above 90%, parts may be compacted, sintered to soften them, recompacted at higher pressure, and sintered again. This process can achieve densities up to 95-97%.

**Hot isostatic pressing (HIP):** Parts are subjected to high temperature and high gas pressure simultaneously, collapsing pores and achieving near-theoretical density (>99%).

::: {.callout-check}
**Quick Check:** Why can't conventional single-press, single-sinter PM achieve 100% theoretical density? What physical limitation prevents complete densification?
:::

---

### Self-Lubricating Sintered Bearings {#sec:sintered-bearings}

One of the most successful applications of powder metallurgy is the manufacture of self-lubricating bearings—also called porous metal bearings, oilite bearings, or sintered bronze bearings. These components exploit the inherent porosity of sintered materials to store and deliver lubricant throughout their service life.

#### The Self-Lubrication Mechanism {#sec:self-lub-mechanism}

Self-lubricating bearings work on a simple but elegant principle: interconnected pores within the bearing structure act as microscopic reservoirs that hold lubricant. During operation:

1. **Startup:** Heat generated by shaft rotation causes thermal expansion of the lubricant
2. **Oil release:** The expanding lubricant is forced out of the pores onto the bearing surface
3. **Film formation:** A thin lubricant film develops between the shaft and bearing
4. **Shutdown:** When rotation stops, capillary action draws lubricant back into the pores

This self-regulating system means the bearing can operate for extended periods—often the entire life of the equipment—without external lubrication.

```{r fig-self-lub-mechanism, echo=FALSE, fig.cap="Cross-section of a self-lubricating bearing showing oil release during operation.", fig.align='center', fig.width=10, fig.height=5}
library(ggplot2)

ggplot() +
  # Bearing body (porous)
  geom_rect(aes(xmin = 1, xmax = 5, ymin = 1, ymax = 4), fill = "tan", color = "black", linewidth = 1) +

  # Pores with oil (small circles)
  geom_point(aes(x = c(1.5, 2, 2.5, 3, 3.5, 4, 4.5, 1.7, 2.3, 2.8, 3.3, 3.8, 4.3,
                       1.4, 1.9, 2.4, 2.9, 3.4, 3.9, 4.4),
                 y = c(1.5, 1.6, 1.4, 1.7, 1.5, 1.6, 1.4, 2.2, 2.3, 2.1, 2.4, 2.2, 2.3,
                       3, 3.1, 2.9, 3.2, 3, 3.1, 2.9)),
             size = 4, color = "gold3", alpha = 0.8) +

  # Shaft bore
  geom_rect(aes(xmin = 2, xmax = 4, ymin = 1.8, ymax = 3.2), fill = "white", color = "black", linewidth = 0.5) +

  # Shaft
  geom_rect(aes(xmin = 2.1, xmax = 3.9, ymin = 1.9, ymax = 3.1), fill = "gray50", color = "black", linewidth = 1) +

  # Oil film
  annotate("segment", x = 2.05, xend = 2.05, y = 1.9, yend = 3.1, color = "gold", linewidth = 2) +
  annotate("segment", x = 3.95, xend = 3.95, y = 1.9, yend = 3.1, color = "gold", linewidth = 2) +

  # Rotation arrow on shaft
  geom_curve(aes(x = 3, xend = 3.3, y = 2.5, yend = 2.8), curvature = -0.3,
             arrow = arrow(length = unit(0.15, "cm")), linewidth = 1, color = "black") +
  annotate("text", x = 3.5, y = 2.65, label = "Rotation", size = 3) +

  # Right side - Operating condition
  geom_rect(aes(xmin = 7, xmax = 11, ymin = 1, ymax = 4), fill = "tan", color = "black", linewidth = 1) +

  # Fewer pores (oil released)
  geom_point(aes(x = c(7.5, 8, 8.5, 9, 9.5, 10, 10.5, 7.7, 8.3, 8.8, 9.3, 9.8, 10.3),
                 y = c(1.5, 1.6, 1.4, 1.7, 1.5, 1.6, 1.4, 3, 3.1, 2.9, 3.2, 3, 3.1)),
             size = 3, color = "gold3", alpha = 0.5) +

  # Shaft bore
  geom_rect(aes(xmin = 8, xmax = 10, ymin = 1.8, ymax = 3.2), fill = "white", color = "black", linewidth = 0.5) +

  # Shaft (slightly offset - hydrodynamic)
  geom_rect(aes(xmin = 8.05, xmax = 9.85, ymin = 1.85, ymax = 3.05), fill = "gray50", color = "black", linewidth = 1) +

  # Thick oil film (hydrodynamic)
  annotate("segment", x = 8.02, xend = 8.02, y = 1.85, yend = 3.05, color = "gold", linewidth = 4) +
  annotate("segment", x = 9.88, xend = 9.88, y = 1.85, yend = 3.05, color = "gold", linewidth = 1) +

  # Fast rotation
  geom_curve(aes(x = 9, xend = 9.4, y = 2.45, yend = 2.9), curvature = -0.4,
             arrow = arrow(length = unit(0.2, "cm")), linewidth = 1.5, color = "black") +
  annotate("text", x = 9.6, y = 2.7, label = "Fast\nRotation", size = 2.5) +

  # Labels
  annotate("text", x = 3, y = 4.5, label = "AT REST", fontface = "bold", size = 4) +
  annotate("text", x = 9, y = 4.5, label = "OPERATING", fontface = "bold", size = 4) +
  annotate("text", x = 3, y = 0.5, label = "Oil stored in pores", size = 3, fontface = "italic") +
  annotate("text", x = 9, y = 0.5, label = "Oil released to form\nhydrodynamic film", size = 3, fontface = "italic") +

  # Arrow between states
  geom_segment(aes(x = 5.5, xend = 6.5, y = 2.5, yend = 2.5),
               arrow = arrow(length = unit(0.25, "cm")), linewidth = 1.2) +
  annotate("text", x = 6, y = 3, label = "Heat &\nRotation", size = 2.5) +

  coord_fixed(ratio = 1, xlim = c(0, 12), ylim = c(0, 5)) +
  theme_void() +
  labs(title = "Self-Lubricating Bearing Operation") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

#### Bearing Density and Porosity Requirements {#sec:bearing-density}

For effective self-lubrication, bearings require interconnected porosity in the range of 20-30%, corresponding to a sintered density of 70-80% of theoretical. At lower densities, the structure becomes too weak; at higher densities, insufficient lubricant can be stored.

```{r tab-bearing-specs, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)

bearing_specs <- data.frame(
  Property = c("Density Range", "Porosity Range", "Oil Content by Volume",
               "Maximum Surface Speed", "Maximum PV Value"),
  Bronze = c("6.0-6.8 g/cm³", "20-30%", "18-25%",
             "6 m/s (1200 ft/min)", "1.75 MPa·m/s"),
  Iron = c("5.5-6.4 g/cm³", "18-30%", "15-25%",
           "3 m/s (600 ft/min)", "1.05 MPa·m/s"),
  `Iron-Copper` = c("5.8-6.6 g/cm³", "16-26%", "16-22%",
                    "4 m/s (800 ft/min)", "1.40 MPa·m/s")
)

kable(bearing_specs,
      caption = "Typical Specifications for Self-Lubricating Sintered Bearings",
      col.names = c("Property", "Bronze", "Iron", "Iron-Copper"),
      align = "lrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                latex_options = c("hold_position"),
                full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Material" = 3))
```

> **PV Value:** The pressure-velocity (PV) product is a key parameter for bearing selection. It represents the intensity of the tribological conditions. Exceeding the maximum PV rating will cause overheating, lubricant degradation, and bearing failure.

---

### Low-Speed and Reciprocating Applications: Solid Lubricant Bearings {#sec:solid-lubricant-bearings}

Not all bearing applications can use oil-impregnated bushings effectively. In certain conditions, solid lubricant bearings provide superior performance:

- **Low-speed or oscillating motion:** Oil films cannot develop hydrodynamically
- **High temperatures:** Oils degrade or evaporate
- **Contaminated environments:** Oils can attract abrasive particles
- **Food and pharmaceutical applications:** Oil contamination is unacceptable
- **Vacuum or space applications:** Oils outgas or evaporate

#### Graphite-Impregnated Bearings {#sec:graphite-bearings}

For these challenging applications, sintered bearings are impregnated with solid lubricants, most commonly **graphite**. Graphite is a form of carbon with a layered crystal structure—the layers slide easily over one another, providing lubrication through mechanical shearing rather than fluid film formation.

**How Graphite Impregnation Works:**

1. The bearing is sintered to the desired density (typically 70-80%)
2. The pores are vacuum-impregnated with a graphite suspension or graphite-resin mixture
3. The bearing is heat-treated to cure any binders and lock the graphite in place
4. During operation, the graphite transfers from the bearing to the shaft, forming a solid lubricant film

```{r fig-graphite-structure, echo=FALSE, fig.cap="Graphite's layered structure enables easy shearing between planes, providing solid lubrication.", fig.align='center', fig.width=8, fig.height=5}
library(ggplot2)

# Create graphite layer representation
ggplot() +
  # Multiple layers
  # Layer 1
  geom_segment(aes(x = seq(1, 9, by = 0.6), xend = seq(1, 9, by = 0.6) + 0.3,
                   y = rep(4, 14), yend = rep(4, 14)),
               linewidth = 3, color = "gray30", lineend = "round") +
  # Layer 2 (offset)
  geom_segment(aes(x = seq(1.3, 9, by = 0.6), xend = seq(1.3, 9, by = 0.6) + 0.3,
                   y = rep(3.5, 13), yend = rep(3.5, 13)),
               linewidth = 3, color = "gray40", lineend = "round") +
  # Layer 3
  geom_segment(aes(x = seq(1, 9, by = 0.6), xend = seq(1, 9, by = 0.6) + 0.3,
                   y = rep(3, 14), yend = rep(3, 14)),
               linewidth = 3, color = "gray30", lineend = "round") +
  # Layer 4 (offset)
  geom_segment(aes(x = seq(1.3, 9, by = 0.6), xend = seq(1.3, 9, by = 0.6) + 0.3,
                   y = rep(2.5, 13), yend = rep(2.5, 13)),
               linewidth = 3, color = "gray40", lineend = "round") +
  # Layer 5
  geom_segment(aes(x = seq(1, 9, by = 0.6), xend = seq(1, 9, by = 0.6) + 0.3,
                   y = rep(2, 14), yend = rep(2, 14)),
               linewidth = 3, color = "gray30", lineend = "round") +

  # Shearing arrow
  geom_segment(aes(x = 0.3, xend = 0.3, y = 2.5, yend = 3.5),
               arrow = arrow(length = unit(0.15, "cm"), ends = "both"),
               linewidth = 1, color = "red") +
  annotate("text", x = 0.3, y = 1.7, label = "Weak\nbonds", color = "red", size = 3) +

  # Movement arrow
  geom_segment(aes(x = 4, xend = 6, y = 4.3, yend = 4.3),
               arrow = arrow(length = unit(0.2, "cm")), linewidth = 1.2, color = "blue") +
  annotate("text", x = 5, y = 4.6, label = "Easy sliding direction", color = "blue", size = 3.5) +

  # Labels
  annotate("text", x = 5, y = 1.3,
           label = "Strong covalent bonds within layers\nWeak van der Waals bonds between layers",
           size = 3.5, fontface = "italic") +

  coord_fixed(ratio = 1, xlim = c(0, 10), ylim = c(0.8, 5)) +
  theme_void() +
  labs(title = "Graphite Crystal Structure - Basis for Solid Lubrication") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

#### Other Solid Lubricants {#sec-other-solid-lubes}

While graphite is most common, other solid lubricants are used for specific conditions:

| Lubricant | Temperature Range | Best Applications |
|:----------|:-----------------:|:------------------|
| Graphite | -200°C to +500°C | General purpose, humid environments |
| Molybdenum Disulfide (MoS₂) | -200°C to +400°C | Vacuum, dry environments, high loads |
| PTFE (Teflon) | -200°C to +260°C | Low friction, chemical resistance |
| Lead | -40°C to +150°C | Conformability, embedded in bronze |

::: {.callout-industry}
**Industry Application:** Food processing equipment extensively uses graphite-impregnated bronze bushings in conveyor systems, ovens, and packaging machines. The bearings operate at elevated temperatures where oil would contaminate products, and the graphite provides reliable lubrication without any risk of food contamination.
:::

---

### High-Speed Applications: Oil-Impregnated Bearings {#sec:oil-impregnated-bearings}

For applications involving continuous, high-speed rotation, oil-impregnated sintered bearings are the preferred solution. At sufficient speeds, these bearings develop a hydrodynamic oil film that completely separates the shaft from the bearing surface, virtually eliminating wear.

#### The Hydrodynamic Lubrication Principle {#sec:hydrodynamic-principle}

When a shaft rotates within an oil-lubricated bearing, it creates a converging wedge of lubricant. At sufficient speed:

1. Oil is dragged into the narrowing gap by shaft rotation
2. Pressure builds up in the converging zone
3. The pressure "lifts" the shaft, creating complete separation
4. The shaft essentially floats on a thin oil cushion

This hydrodynamic action requires a minimum speed to develop—below this speed, the bearing operates in boundary lubrication mode with metal-to-metal contact.

```{r fig-hydrodynamic, echo=FALSE, fig.cap="Hydrodynamic film formation in an oil-lubricated bearing at speed.", fig.align='center', fig.width=9, fig.height=5}
library(ggplot2)

ggplot() +
  # Bearing (outer circle representation)
  geom_rect(aes(xmin = 1, xmax = 9, ymin = 1, ymax = 7), fill = "tan", color = "black", linewidth = 1) +

  # Bore (circular cutout representation)
  geom_point(aes(x = 5, y = 4), size = 70, color = "white") +
  geom_point(aes(x = 5, y = 4), size = 68, color = "black", shape = 1, stroke = 1) +

  # Shaft (offset to right/up for hydrodynamic illustration)
  geom_point(aes(x = 5.15, y = 4.1), size = 55, color = "gray50") +
  geom_point(aes(x = 5.15, y = 4.1), size = 53, color = "gray60") +

  # Oil film - varying thickness
  # Thick side (left/bottom)
  annotate("segment", x = 3.2, xend = 3.5, y = 4, yend = 4,
           color = "gold", linewidth = 8) +
  annotate("segment", x = 4.5, xend = 4.5, y = 2.4, yend = 2.7,
           color = "gold", linewidth = 6) +
  # Thin side (right/top)
  annotate("segment", x = 6.5, xend = 6.7, y = 4, yend = 4,
           color = "gold", linewidth = 2) +
  annotate("segment", x = 5.5, xend = 5.5, y = 5.4, yend = 5.5,
           color = "gold", linewidth = 2) +

  # Rotation direction
  geom_curve(aes(x = 5.15, xend = 5.8, y = 4.1, yend = 4.7), curvature = -0.4,
             arrow = arrow(length = unit(0.2, "cm")), linewidth = 1.2) +
  annotate("text", x = 6.3, y = 5, label = "Rotation", size = 3.5) +

  # Pressure distribution arrows
  geom_segment(aes(x = 3.3, xend = 2.8, y = 4, yend = 4),
               arrow = arrow(length = unit(0.15, "cm")), color = "red", linewidth = 1) +
  geom_segment(aes(x = 3.3, xend = 2.5, y = 3.5, yend = 3.2),
               arrow = arrow(length = unit(0.15, "cm")), color = "red", linewidth = 1) +
  geom_segment(aes(x = 4.5, xend = 4.5, y = 2.5, yend = 1.8),
               arrow = arrow(length = unit(0.15, "cm")), color = "red", linewidth = 1) +

  annotate("text", x = 2, y = 3.5, label = "High\nPressure\nZone", color = "red", size = 3, fontface = "bold") +

  # Labels
  annotate("text", x = 5, y = 0.3,
           label = "Shaft 'floats' on oil film - no metal contact",
           size = 3.5, fontface = "italic") +
  annotate("text", x = 7.5, y = 4, label = "Thin film\n(convergent\nwedge)", size = 2.5, color = "darkgoldenrod") +
  annotate("text", x = 2.8, y = 4.5, label = "Thick\nfilm", size = 2.5, color = "darkgoldenrod") +

  coord_fixed(ratio = 1, xlim = c(0, 10), ylim = c(0, 8)) +
  theme_void() +
  labs(title = "Hydrodynamic Lubrication - Shaft Floats on Oil Film") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

#### The Oil Impregnation Process {#sec:oil-impregnation}

Oil impregnation is performed after sintering and any secondary operations. The process ensures complete filling of the interconnected pore network:

**Vacuum Impregnation Method:**

1. **Heating:** Bearings are heated to 80-90°C to reduce oil viscosity and expand the pore openings
2. **Vacuum application:** Bearings are placed in a chamber and evacuated to less than 50 mbar (0.05 atm)
3. **Oil introduction:** While under vacuum, heated oil is introduced to cover the bearings
4. **Pressure application:** The vacuum is released and pressure (3-5 bar) is applied to force oil into any remaining unfilled pores
5. **Draining:** Excess surface oil is drained and bearings are allowed to stabilize

This process typically achieves oil fill levels of 90-98% of the available pore volume.

```{r fig-impregnation-process, echo=FALSE, fig.cap="Vacuum-pressure oil impregnation process for sintered bearings.", fig.align='center', fig.width=12, fig.height=4}
library(ggplot2)

ggplot() +
  # Stage 1: Heating
  annotate("text", x = 1.5, y = 4.2, label = "1. HEAT\nBEARINGS", fontface = "bold", size = 3) +
  geom_rect(aes(xmin = 0.5, xmax = 2.5, ymin = 1.5, ymax = 3.5), fill = "coral", color = "black", linewidth = 1) +
  # Bearings
  geom_point(aes(x = c(1, 1.5, 2, 1.25, 1.75), y = c(2, 2.5, 2, 3, 3)),
             size = 8, color = "tan", shape = 21, fill = "tan", stroke = 1) +
  annotate("text", x = 1.5, y = 1, label = "80-90°C", size = 3, fontface = "italic") +

  # Stage 2: Vacuum
  annotate("text", x = 4, y = 4.2, label = "2. APPLY\nVACUUM", fontface = "bold", size = 3) +
  geom_rect(aes(xmin = 3, xmax = 5, ymin = 1.5, ymax = 3.5), fill = "lightblue", color = "black", linewidth = 1) +
  geom_point(aes(x = c(3.5, 4, 4.5, 3.75, 4.25), y = c(2, 2.5, 2, 3, 3)),
             size = 8, color = "tan", shape = 21, fill = "tan", stroke = 1) +
  # Air escaping
  geom_segment(aes(x = c(3.5, 4, 4.5), xend = c(3.3, 4, 4.7), y = c(2.2, 2.7, 2.2), yend = c(2.5, 3, 2.5)),
               arrow = arrow(length = unit(0.1, "cm")), color = "gray50", linewidth = 0.5) +
  annotate("text", x = 4, y = 1, label = "< 50 mbar", size = 3, fontface = "italic") +

  # Stage 3: Introduce Oil
  annotate("text", x = 6.5, y = 4.2, label = "3. INTRODUCE\nOIL", fontface = "bold", size = 3) +
  geom_rect(aes(xmin = 5.5, xmax = 7.5, ymin = 1.5, ymax = 3.5), fill = "gold", color = "black", linewidth = 1) +
  geom_point(aes(x = c(6, 6.5, 7, 6.25, 6.75), y = c(2, 2.5, 2, 3, 3)),
             size = 8, color = "tan", shape = 21, fill = "sienna", stroke = 1) +
  # Oil level line
  geom_segment(aes(x = 5.5, xend = 7.5, y = 2.8, yend = 2.8), color = "darkgoldenrod", linewidth = 1, linetype = "dashed") +
  annotate("text", x = 6.5, y = 1, label = "Oil covers\nbearings", size = 3, fontface = "italic") +

  # Stage 4: Pressure
  annotate("text", x = 9, y = 4.2, label = "4. APPLY\nPRESSURE", fontface = "bold", size = 3) +
  geom_rect(aes(xmin = 8, xmax = 10, ymin = 1.5, ymax = 3.5), fill = "gold", color = "black", linewidth = 1.5) +
  geom_point(aes(x = c(8.5, 9, 9.5, 8.75, 9.25), y = c(2, 2.5, 2, 3, 3)),
             size = 8, color = "sienna", shape = 21, fill = "sienna", stroke = 1) +
  # Pressure arrows
  geom_segment(aes(x = 9, xend = 9, y = 4, yend = 3.6),
               arrow = arrow(length = unit(0.15, "cm")), color = "red", linewidth = 1.2) +
  annotate("text", x = 9, y = 1, label = "3-5 bar", size = 3, fontface = "italic") +

  # Stage 5: Complete
  annotate("text", x = 11.5, y = 4.2, label = "5. DRAIN &\nSTABILIZE", fontface = "bold", size = 3) +
  geom_rect(aes(xmin = 10.5, xmax = 12.5, ymin = 1.5, ymax = 3.5), fill = "white", color = "black", linewidth = 1) +
  geom_point(aes(x = c(11, 11.5, 12, 11.25, 11.75), y = c(2, 2.5, 2, 3, 3)),
             size = 8, color = "sienna", shape = 21, fill = "sienna", stroke = 1.5) +
  annotate("text", x = 11.5, y = 1, label = "90-98% oil\nfill achieved", size = 3, fontface = "italic") +

  # Arrows between stages
  geom_segment(aes(x = 2.6, xend = 2.9, y = 2.5, yend = 2.5),
               arrow = arrow(length = unit(0.15, "cm")), linewidth = 0.8) +
  geom_segment(aes(x = 5.1, xend = 5.4, y = 2.5, yend = 2.5),
               arrow = arrow(length = unit(0.15, "cm")), linewidth = 0.8) +
  geom_segment(aes(x = 7.6, xend = 7.9, y = 2.5, yend = 2.5),
               arrow = arrow(length = unit(0.15, "cm")), linewidth = 0.8) +
  geom_segment(aes(x = 10.1, xend = 10.4, y = 2.5, yend = 2.5),
               arrow = arrow(length = unit(0.15, "cm")), linewidth = 0.8) +

  coord_fixed(ratio = 1, xlim = c(0, 13), ylim = c(0.5, 4.8)) +
  theme_void() +
  labs(title = "Vacuum-Pressure Oil Impregnation Process") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

#### Oil Selection for Impregnation {#sec:oil-selection}

The lubricating oil must be carefully matched to the application:

| Application | Oil Viscosity (cSt @ 40°C) | Considerations |
|:------------|:--------------------------:|:---------------|
| Small motors, fans | 10-22 | Low viscosity for easy release |
| Automotive applications | 32-68 | Moderate viscosity, wide temp range |
| Industrial equipment | 68-150 | Higher viscosity for heavy loads |
| High temperature | 150-320 | High viscosity index, thermal stability |

> **Note:** Some applications use synthetic oils for improved temperature range and longer life. Ester-based and PAO (polyalphaolefin) synthetics are common choices for demanding applications.

::: {.callout-industry}
**Industry Application:** Electric motor manufacturers rely heavily on oil-impregnated sintered bronze bearings for fractional horsepower motors. A typical ceiling fan motor contains two such bearings that operate continuously for 10+ years without maintenance. The hydrodynamic oil film at operating speed means the shaft never contacts the bearing surface, enabling this exceptional longevity.
:::

---

### Comparing Bearing Types for Different Applications {#sec:bearing-comparison}

The choice between oil-impregnated and solid-lubricant bearings depends on the specific operating conditions:

```{r fig-bearing-selection, echo=FALSE, fig.cap="Selection guide for self-lubricating bearing types based on operating conditions.", fig.align='center', fig.width=10, fig.height=7}
library(ggplot2)

ggplot() +
  # Create quadrant structure
  geom_hline(yintercept = 5, linewidth = 1) +
  geom_vline(xintercept = 5, linewidth = 1) +

  # Axis labels
  annotate("text", x = 5, y = 0.3, label = "SPEED", fontface = "bold", size = 4) +
  annotate("text", x = 0.3, y = 5, label = "LOAD", fontface = "bold", size = 4, angle = 90) +

  # Low speed, low load - Graphite
  geom_rect(aes(xmin = 0.5, xmax = 4.8, ymin = 0.5, ymax = 4.8),
            fill = "lightgreen", alpha = 0.5) +
  annotate("text", x = 2.65, y = 3.5, label = "SOLID LUBRICANT\n(Graphite/MoS2)",
           fontface = "bold", size = 3.5, color = "darkgreen") +
  annotate("text", x = 2.65, y = 2,
           label = "- Oscillating motion\n- Reciprocating\n- Start-stop cycles\n- High temperature",
           size = 2.5) +

  # High speed, low load - Oil
  geom_rect(aes(xmin = 5.2, xmax = 9.5, ymin = 0.5, ymax = 4.8),
            fill = "gold", alpha = 0.5) +
  annotate("text", x = 7.35, y = 3.5, label = "OIL IMPREGNATED\n(Hydrodynamic)",
           fontface = "bold", size = 3.5, color = "darkorange4") +
  annotate("text", x = 7.35, y = 2,
           label = "- Continuous rotation\n- Electric motors\n- Fans, pumps\n- Long life required",
           size = 2.5) +

  # Low speed, high load - Graphite + Design considerations
  geom_rect(aes(xmin = 0.5, xmax = 4.8, ymin = 5.2, ymax = 9.5),
            fill = "lightblue", alpha = 0.5) +
  annotate("text", x = 2.65, y = 8, label = "SOLID LUBRICANT\n+ Heavy Duty Design",
           fontface = "bold", size = 3.5, color = "steelblue") +
  annotate("text", x = 2.65, y = 6.5,
           label = "- Heavy machinery\n- Construction equip.\n- Slow pivots\n- Shock loading",
           size = 2.5) +

  # High speed, high load - Special considerations
  geom_rect(aes(xmin = 5.2, xmax = 9.5, ymin = 5.2, ymax = 9.5),
            fill = "coral", alpha = 0.5) +
  annotate("text", x = 7.35, y = 8, label = "CONSIDER\nROLLING ELEMENT",
           fontface = "bold", size = 3.5, color = "darkred") +
  annotate("text", x = 7.35, y = 6.5,
           label = "- Ball bearings\n- Roller bearings\n- Force-fed lubrication\n- PM may not suffice",
           size = 2.5) +

  # Speed/Load indicators
  annotate("text", x = 2.65, y = 0.1, label = "LOW", size = 3) +
  annotate("text", x = 7.35, y = 0.1, label = "HIGH", size = 3) +
  annotate("text", x = 0.1, y = 2.65, label = "LOW", size = 3, angle = 90) +
  annotate("text", x = 0.1, y = 7.35, label = "HIGH", size = 3, angle = 90) +

  coord_fixed(ratio = 1, xlim = c(0, 10), ylim = c(0, 10)) +
  theme_void() +
  labs(title = "Self-Lubricating Bearing Selection Guide") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

#### Summary Comparison Table {#sec-comparison-table}

```{r tab-bearing-comparison, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)

comparison <- data.frame(
  Factor = c("Speed capability", "Load capacity", "Temperature range",
             "Maintenance", "Cost", "Typical life", "Best motion type",
             "Environmental tolerance"),
  Oil_Impregnated = c("High (develops hydrodynamic film)",
                       "Moderate",
                       "-40°C to +120°C (limited by oil)",
                       "None required",
                       "Low",
                       "10,000+ hours at rated conditions",
                       "Continuous rotation",
                       "Clean environments preferred"),
  Graphite_Solid = c("Low to moderate",
                      "High (direct contact acceptable)",
                      "-200°C to +500°C",
                      "None required",
                      "Moderate",
                      "Depends on wear rate",
                      "Oscillating, reciprocating, slow rotation",
                      "Contaminated, wet, vacuum OK")
)

kable(comparison,
      caption = "Comparison of Oil-Impregnated vs. Graphite (Solid Lubricant) Bearings",
      col.names = c("Factor", "Oil-Impregnated", "Graphite/Solid Lubricant"),
      align = "lll") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                latex_options = c("hold_position", "scale_down"),
                full_width = TRUE)
```

::: {.callout-check}
**Quick Check:** A food processing company needs bearings for a conveyor system operating at 150°C in a flour-dusty environment with oscillating motion. Would you recommend oil-impregnated or graphite-impregnated bearings? Why?
:::

---

### Advantages and Limitations of Powder Metallurgy {#sec:pm-advantages}

#### Advantages {#sec-pm-pros}

1. **Near-net-shape capability:** Parts require minimal or no machining, reducing waste and cost
2. **Controlled porosity:** Enables unique products like self-lubricating bearings and filters
3. **High material utilization:** >97% of powder becomes final product (vs. 40-50% for machining)
4. **Complex geometries:** Can produce shapes difficult to machine or cast
5. **Composition flexibility:** Can combine immiscible metals (e.g., copper-tungsten contacts)
6. **Excellent repeatability:** High-volume production with consistent quality
7. **Energy efficiency:** Lower processing temperatures than casting

#### Limitations {#sec-pm-cons}

1. **Size constraints:** Parts typically limited to <25 kg due to press capacity
2. **Shape constraints:** Difficult to produce re-entrant features, threads, or horizontal holes
3. **Tooling cost:** Dies are expensive, requiring high volumes to amortize
4. **Density limitations:** Conventional PM rarely exceeds 95% density
5. **Mechanical properties:** Often lower than wrought materials at same density
6. **Secondary operations:** May require sizing, machining, or heat treatment

```{r fig-pm-vs-alternatives, echo=FALSE, fig.cap="Comparison of powder metallurgy with alternative manufacturing processes.", fig.align='center', fig.width=10, fig.height=6}
library(ggplot2)
library(dplyr)

# Create comparison data
process_data <- data.frame(
  process = rep(c("Powder Metallurgy", "Die Casting", "Investment Casting",
                   "CNC Machining", "Metal Injection Molding"), each = 4),
  metric = rep(c("Material Utilization", "Geometric Complexity",
                  "Production Rate", "Unit Cost (High Vol.)"), 5),
  score = c(
    # PM
    95, 70, 85, 90,
    # Die Casting
    90, 60, 95, 85,
    # Investment Casting
    85, 95, 50, 60,
    # CNC Machining
    45, 80, 40, 50,
    # MIM
    95, 90, 70, 75
  )
)

ggplot(process_data, aes(x = metric, y = score, fill = process)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = c("Powder Metallurgy" = "steelblue",
                                "Die Casting" = "coral",
                                "Investment Casting" = "darkgreen",
                                "CNC Machining" = "purple",
                                "Metal Injection Molding" = "orange")) +
  labs(
    title = "Manufacturing Process Comparison",
    subtitle = "Relative scores (higher is better)",
    x = NULL,
    y = "Score",
    fill = "Process"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 15, hjust = 1),
    legend.position = "right"
  ) +
  coord_cartesian(ylim = c(0, 100))
```

---

### Design Guidelines for Sintered Parts {#sec:pm-design-guidelines}

Designing for powder metallurgy requires understanding the unique capabilities and constraints of the process.

#### Geometric Considerations {#sec-pm-geometry}

| Feature | Guideline |
|:--------|:----------|
| Wall thickness | Minimum 1.5 mm; uniform preferred |
| Holes | Parallel to pressing direction only; min. diameter 1.5 mm |
| Chamfers | Required on all edges (min. 0.3 mm × 45°) for die filling |
| Fillets | Use generous radii to aid powder flow |
| Draft angles | Generally not required (parts eject axially) |
| Undercuts | Not possible without secondary operations |
| Threads | Cannot be pressed; must be machined after sintering |

```{r fig-pm-design, echo=FALSE, fig.cap="Design features for powder metallurgy parts.", fig.align='center', fig.width=10, fig.height=5}
library(ggplot2)

ggplot() +
  # Good design example
  annotate("text", x = 2.5, y = 5.5, label = "GOOD DESIGN", fontface = "bold", size = 4, color = "darkgreen") +

  # Part outline - good
  geom_rect(aes(xmin = 1, xmax = 4, ymin = 2, ymax = 4.5), fill = "lightgray", color = "black", linewidth = 1) +
  # Chamfered edges
  geom_polygon(aes(x = c(1, 1.3, 1.3, 1), y = c(4.5, 4.5, 4.2, 4.5)), fill = "white", color = "black") +
  geom_polygon(aes(x = c(4, 3.7, 3.7, 4), y = c(4.5, 4.5, 4.2, 4.5)), fill = "white", color = "black") +
  geom_polygon(aes(x = c(1, 1.3, 1.3, 1), y = c(2, 2, 2.3, 2)), fill = "white", color = "black") +
  geom_polygon(aes(x = c(4, 3.7, 3.7, 4), y = c(2, 2, 2.3, 2)), fill = "white", color = "black") +
  # Vertical hole (OK)
  geom_rect(aes(xmin = 2.3, xmax = 2.7, ymin = 2, ymax = 4.5), fill = "white", color = "black") +

  # Annotations - good
  annotate("segment", x = 1.15, xend = 0.7, y = 4.35, yend = 4.7,
           arrow = arrow(length = unit(0.1, "cm"))) +
  annotate("text", x = 0.5, y = 4.9, label = "Chamfer", size = 2.5) +
  annotate("segment", x = 2.5, xend = 2.5, y = 4.7, yend = 5.1,
           arrow = arrow(length = unit(0.1, "cm"))) +
  annotate("text", x = 2.5, y = 5.3, label = "Hole parallel\nto press", size = 2.5) +
  annotate("text", x = 2.5, y = 1.5, label = "Uniform wall thickness", size = 2.5, fontface = "italic") +

  # Bad design example
  annotate("text", x = 7.5, y = 5.5, label = "POOR DESIGN", fontface = "bold", size = 4, color = "red") +

  # Part outline - bad
  geom_rect(aes(xmin = 6, xmax = 9, ymin = 2, ymax = 4.5), fill = "lightgray", color = "black", linewidth = 1) +
  # Sharp corners (no chamfers)
  # Horizontal hole (NOT OK)
  geom_rect(aes(xmin = 6, xmax = 7.5, ymin = 3, ymax = 3.5), fill = "white", color = "red", linewidth = 1.5) +
  # Thin wall section
  geom_rect(aes(xmin = 8.2, xmax = 8.4, ymin = 2, ymax = 4.5), fill = "darkgray", color = "black") +
  # Undercut
  geom_polygon(aes(x = c(6, 6, 6.4, 6.4), y = c(2.3, 2, 2, 2.3)), fill = "white", color = "red", linewidth = 1.5) +

  # Annotations - bad
  annotate("segment", x = 6.8, xend = 6.2, y = 3.25, yend = 3.8,
           arrow = arrow(length = unit(0.1, "cm")), color = "red") +
  annotate("text", x = 6, y = 4.1, label = "Horizontal\nhole - NO", size = 2.5, color = "red") +
  annotate("segment", x = 8.3, xend = 8.8, y = 3.25, yend = 3.25,
           arrow = arrow(length = unit(0.1, "cm")), color = "red") +
  annotate("text", x = 9.2, y = 3.25, label = "Thin\nwall", size = 2.5, color = "red") +
  annotate("segment", x = 6.2, xend = 5.7, y = 2.15, yend = 1.8,
           arrow = arrow(length = unit(0.1, "cm")), color = "red") +
  annotate("text", x = 5.5, y = 1.5, label = "Undercut\n- NO", size = 2.5, color = "red") +
  annotate("text", x = 9, y = 4.5, label = "Sharp\ncorners", size = 2.5, color = "red") +

  # Press direction arrows
  geom_segment(aes(x = 2.5, xend = 2.5, y = 0.8, yend = 1.3),
               arrow = arrow(length = unit(0.2, "cm")), linewidth = 1.2, color = "blue") +
  geom_segment(aes(x = 2.5, xend = 2.5, y = 5.7, yend = 5.2),
               arrow = arrow(length = unit(0.2, "cm")), linewidth = 1.2, color = "blue") +
  annotate("text", x = 3.5, y = 0.8, label = "Press direction", size = 3, color = "blue") +

  coord_fixed(ratio = 1, xlim = c(0, 10), ylim = c(0.5, 6)) +
  theme_void() +
  labs(title = "Design for Powder Metallurgy") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

::: {.callout-discussion}
**Group Discussion:** Your team is designing a small gear (25 mm pitch diameter, 20 teeth) for a consumer appliance. Compare the suitability of powder metallurgy vs. machining vs. die casting. Consider: production volume (100,000 units/year), required precision (AGMA 8), and cost constraints.
:::

---

### Practice Problems {#sec:sintering-practice}

::: {.question}
**Problem 1:** A sintered bronze bushing has a density of 6.4 g/cm³. Bronze has a theoretical density of 8.8 g/cm³. Calculate:
a) The percentage of theoretical density
b) The approximate porosity percentage
c) The expected oil content by volume after vacuum impregnation (assuming 95% pore filling)
:::

::: {.answer}
**Solution:**

a) Percentage of theoretical density:
$$\text{% TD} = \frac{6.4}{8.8} \times 100 = 72.7\%$$

b) Porosity percentage:
$$\text{Porosity} = 100\% - 72.7\% = 27.3\%$$

c) Oil content by volume:
$$\text{Oil content} = 27.3\% \times 0.95 = 25.9\% \text{ by volume}$$

This porosity level is appropriate for a self-lubricating bearing application.
:::

::: {.question}
**Problem 2:** An engineer must select bearings for two applications:
- Application A: Electric fan motor, 1800 RPM, 25°C ambient, clean environment
- Application B: Oven conveyor pivot, oscillating ±15°, 200°C, flour dust present

Recommend the appropriate bearing type for each and justify your selection.
:::

::: {.answer}
**Solution:**

**Application A: Oil-impregnated sintered bronze bearing**
- Continuous rotation at high speed will develop hydrodynamic lubrication
- Clean environment won't contaminate the oil
- Moderate temperature is within oil operating range
- Long maintenance-free life expected

**Application B: Graphite-impregnated sintered bronze bearing**
- Oscillating motion cannot develop hydrodynamic film—solid lubricant needed
- 200°C exceeds the operating range of most lubricating oils
- Flour dust would contaminate oil and accelerate wear
- Graphite functions in contaminated environments and high temperatures
:::

::: {.question}
**Problem 3:** Calculate the green density of an iron powder compact if 150 g of powder is pressed into a cylindrical shape with diameter 25 mm and height 15 mm. Express the result as g/cm³ and as a percentage of theoretical density (7.87 g/cm³).
:::

::: {.answer}
**Solution:**

Volume of the compact:
$$V = \frac{\pi d^2}{4} \times h = \frac{\pi (2.5)^2}{4} \times 1.5 = 7.36 \text{ cm}^3$$

Green density:
$$\rho_{green} = \frac{m}{V} = \frac{150}{7.36} = 6.52 \text{ g/cm}^3$$

Percentage of theoretical density:
$$\text{% TD} = \frac{6.52}{7.87} \times 100 = 82.8\%$$

This is typical green density for a structural PM part.
:::

---

### Key Takeaways {#sec:sintering-takeaways}

1. **Powder metallurgy** produces parts by compacting metal powder and heating below the melting point—the sintering process bonds particles through solid-state diffusion.

2. **Controlled porosity** is a unique capability of PM, enabling products like self-lubricating bearings and filters that cannot be made by other processes.

3. **Sintered density** (typically 70-95% of theoretical) is controlled through compaction pressure, powder characteristics, and sintering parameters.

4. **Oil-impregnated bearings** work best for continuous, high-speed rotation where hydrodynamic lubrication can develop.

5. **Graphite-impregnated bearings** are preferred for oscillating motion, high temperatures, contaminated environments, or food-contact applications.

6. **Vacuum impregnation** is the standard process for filling sintered bearing pores with oil, achieving 90-98% pore filling.

7. **Design for PM** requires understanding constraints: holes parallel to pressing, chamfered edges, no undercuts, and uniform wall thickness.

---

### Further Resources {#sec:sintering-resources}

**Videos:**

```{r video-sintering-intro, echo=FALSE, results='asis'}
embed_youtube("MgukjCT9o80", "Introduction to Powder Metallurgy and Sintering")
```

```{r video-sintering-walk-through, echo=FALSE, results='asis'}
embed_youtube("dnxYjXZO9u0", "Walk through the Powdered Metal Sintering Cycle")
```

```{r video-self-lub-bearings, echo=FALSE, results='asis'}
embed_youtube("olo_vK13WuU", "Self-Lubricating Bearing Manufacturing")
```

```{r video-sintering-process-animation, echo=FALSE, results='asis'}
embed_youtube("4zXkn9iyC0c", "Sintering Process Animation")
```

**Reading:**

- [Metal Powder Industries Federation (MPIF)](https://www.mpif.org/) - Industry standards and design guides
- [GKN Powder Metallurgy](https://www.gknpm.com/en/our-businesses/gkn-sinter-metals/) - Application examples
- [Oiles America](https://www.oiles.com/) - Self-lubricating bearing specifications
