## Design for 3D Printing {#chap-3d-printing}

```{r setup-3d-printing, include=FALSE}
## include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
## include the required packages to make sure the page can be built.
source("R/required_packages.R")
```

### Learning Objectives {#sec:objectives-3d-printing}

By the end of this chapter, you will be able to:

1. **Explain** the principles of additive manufacturing and how 3D printing differs from traditional manufacturing
2. **Compare** different 3D printing technologies including SLA, SLS, and FDM/FFF
3. **Apply** design guidelines for wall thickness, feature size, and overhangs
4. **Identify** strategies for creating functional features such as living hinges, thread inserts, and embedded parts
5. **Evaluate** designs for printability and select appropriate post-processing techniques

::: {.callout-think}
**Before You Read:** Consider a part you use daily. How might it be designed differently if manufactured using 3D printing instead of traditional methods? What geometric features would be easier or harder to achieve?
:::

### Introduction to 3D Printing {#sec:3d-printing-intro}

3D printing or additive manufacturing (AM) technologies create three-dimensional parts from computer-aided design (CAD) models by successively adding material layer by layer until the physical part is created.

While 3D printing technologies have been around since the 1980s, recent advances in machinery, materials, and software have made 3D printing accessible to a wider range of businesses, enabling more and more companies to use tools previously limited to a few high-tech industries. Today, professional, low-cost desktop and benchtop 3D printers accelerate innovation and support businesses in various industries including engineering, manufacturing, dentistry, healthcare, education, entertainment, jewelry, and audiology.

For an introduction to 3D printing, watch: [3D Printing - An Introduction](https://youtu.be/a42znmU6BLE)

```{r video-3d-printing-intro, echo=FALSE, results='asis'}
embed_youtube("a42znmU6BLE", "3D Printing - An Introduction")
```

### The 3D Printing Process {#sec:3d-printing-process}

All 3D printing processes start with a CAD model that is sent to software to prepare the design. Depending on the technology, the 3D printer might produce the part layer by layer by solidifying resin or sintering powder. The parts are then removed from the printer and post-processed for the specific application.

For more detail on how 3D printing works, watch: [How Does 3D Printing Work?](https://youtu.be/Hr4DKHX9Hsg)

```{r video-how-does-3d-printing-work, echo=FALSE, results='asis'}
embed_youtube("Hr4DKHX9Hsg", "How Does 3D Printing Work?")
```

#### Design

3D printers create parts from three-dimensional models, the mathematical representations of any three-dimensional surface created using computer-aided design (CAD) software or developed from 3D scan data. The design is then exported as an STL or OBJ file readable by print preparation software.

3D printers include software to specify print settings and slice the digital model into layers that represent horizontal cross-sections of the part. Adjustable printing settings include orientation, support structures (if needed), layer height, and material. Once setup is complete, the software sends the instructions to the printer via a wireless or cable connection.

#### 3D Print

Some 3D printers use a laser to cure liquid resin into hardened plastic, others fuse small particles of polymer powder at high temperatures to build parts. Most 3D printers can run unattended until the print is complete, and modern systems automatically refill the material required for the parts from cartridges.

#### Post-Process

Depending on the technology and the material, the printed parts may require rinsing in isopropyl alcohol (IPA) to remove any uncured resin from their surface, post-curing to stabilize mechanical properties, manual work to remove support structures, or cleaning with compressed air or a media blaster to remove excess powder. Some of these processes can be automated with accessories.

### Types of 3D Printers {#sec:3d-printer-types}

#### SLA (Stereolithography) {#sec:sla}

Stereolithography was the world's first 3D printing technology, invented in the 1980s, and is still one of the most popular technologies for professionals. SLA 3D printers use a laser to cure liquid resin into hardened plastic in a process called photopolymerization.

SLA resin 3D printers have become vastly popular for their ability to produce high-accuracy, isotropic, and watertight prototypes and parts in a range of advanced materials with fine features and smooth surface finish. SLA resin formulations offer a wide range of optical, mechanical, and thermal properties to match those of standard, engineering, and industrial thermoplastics.

[Resin 3D printing](https://formlabs.com/blog/resin-3d-printer-comparison-sla-vs-dlp/) is a great option for highly detailed prototypes requiring tight tolerances and smooth surfaces, such as molds, patterns, and functional parts. SLA 3D printers are widely used in a range of industries from engineering and product design to manufacturing, dentistry, jewelry, model making, and education.

#### SLS (Selective Laser Sintering) {#sec:sls}

Selective laser sintering (SLS) 3D printers use a high-power laser to sinter small particles of polymer powder into a solid structure. The unfused powder supports the part during printing and eliminates the need for dedicated support structures. This makes SLS ideal for complex geometries, including interior features, undercuts, thin walls, and negative features. Parts produced with SLS printing have excellent mechanical characteristics, with strength resembling that of injection-molded parts.

The most common material for selective laser sintering is nylon, a popular engineering thermoplastic with excellent mechanical properties. Nylon is lightweight, strong, and flexible, as well as stable against impact, chemicals, heat, UV light, water, and dirt.

The combination of low cost per part, high productivity, and established materials make SLS a popular choice among engineers for functional prototyping, and a cost-effective alternative to injection molding for limited-run or bridge manufacturing.

#### FDM (Fused Deposition Modelling) or FFF (Fused Filament Fabrication) {#sec:fdm}

Fused deposition modeling (FDM), also known as fused filament fabrication (FFF), is the most widely used type of 3D printing at the consumer level. FDM 3D printers work by extruding thermoplastic filaments, such as ABS (Acrylonitrile Butadiene Styrene) or PLA (Polylactic Acid), through a heated nozzle, melting the material and applying the plastic layer by layer to a build platform. Each layer is laid down one at a time until the part is complete.

For more information, see the [FDM Print Process Introduction](https://www.hubs.com/knowledge-base/what-is-fdm-3d-printing/).

### Design Considerations {#sec:design-considerations-3d-print}

A good summary on design considerations for 3D printing can be found in the [Ultimate Guide: How to Design for 3D Printing](https://wikifactory.com/+wikifactory/stories/ultimate-guide-how-to-design-for-3d-printing).

#### Wall Thickness and Feature Size {#sec:wall-thickness}

The minimum wall thickness is 2x the nozzle size or 0.8mm for Prusa MK4 which has a 0.4mm nozzle. If you desire stronger parts, increase to 5x nozzle width. If you design with smaller wall thickness, the slicer may ignore the walls and not attempt to print them.

The minimum feature size is 2mm.

Thin vertical pins are the most challenging features for a FDM 3D printer. Excessive heights of them are difficult to print.

#### Bridges, Overhangs and Support Material {#sec:bridges-overhangs}

Bridges are possible up to about 10mm. However, you will have to account for unknown droop between the risers. This will result in smaller dimensions in the height of the gap.

Overhangs up to 45° are possible. For steeper overhangs, support material is required. Consider orienting the part differently on the printer bed first.

#### Print in Place {#sec:print-in-place}

This means that the finished part has moving parts that are not assembled, but printed together at the same time (in place). When the parts are removed from the print bed, the parts stay together and are freely movable. This is used for hinges.

#### Living Hinge {#sec:living-hinge}

A living hinge connects two parts together with some flexible material. 3D print material is solid; in order to make it bend, thin wall thickness and extra length has to be given. The idea is to distribute the stress over more material so that each material sees only a little stress.

A good guide on how to design living hinges with minimum length formula can be found here: [How to Design 3D Printed Hinges](https://www.protolabs.com/resources/design-tips/how-to-design-3d-printed-hinges/)

#### Thread Inserts {#sec:thread-inserts}

These inserts are heated with the tip of a soldering iron and pressed into the undersized printed hole that is prepared for them. Use the soldering iron for that process to keep the insert hot. Because you are melting the hole wall and this is a manual process, dimensional accuracy is limited.

More detailed description with pull out strength comparison is published at [CNC Kitchen - Threaded Inserts for 3D Prints](https://www.cnckitchen.com/blog/threaded-inserts-for-3d-prints-cheap-vs-expensive).

#### Embedding Parts {#sec:embedding-parts}

In this method, the print is paused when a cavity for the part to be inserted is finished. The part is inserted and then the cavity is closed by printing over top of it. Take care of tolerances and clearances with this method! You cannot remove the part from the print bed to insert the part.

A step by step process can be found at [Captive Nuts and More in 3D Printing](https://www.instructables.com/CAPTIVE-NUTS-AND-MORE-IN-3D-PRINTING/). Examples of parts to be embedded: Nuts, Magnets, Bearings, Bushings, Carbon Fibre struts, and many more.

#### Watertight Parts {#sec:watertight-parts}

Printing parts that are watertight is not easy because the layers are built up of individual beads of material. Typical ways to improve the watertightness include:

- Increase nozzle temperature
- Increase material flow
- Slow the nozzle speed
- Add extra exterior walls

All of this is done in an effort to have very good re-melting of the lower layers when the next one is added on top.

#### Stronger Parts {#sec:stronger-parts}

How to increase the strength of the part?

- Ensure the printed part is stressed in the XY direction (remember, XY is twice as strong as the Z direction)
- Increase the number of external walls
- Design holes into the part instead of relying on infill. This forces the slicer to add extra walls giving it more strength than increasing the infill percentage

#### Holes {#sec:holes}

##### Vertical Holes

Vertical holes (diameter in XY plane) suffer from being undersized. This is because the export file of the CAD model approximates the hole as a polygon. The dimension across the hole between the flat parts is less than the diameter. This can be limited with slicer settings, but not completely removed. Add 0.2mm to the hole diameter.

##### Horizontal Holes and Arches

Horizontal holes require the slicer to bridge the hole at the top layer. There will be droop and this reduces the diameter of the holes to an unknown dimension. This can be desirable if a bolt has to cut its own thread. This can also mean that the pin cannot be assembled because the hole is too small. In this case, change the top geometry of the hole to avoid bridging and rely on the rest of the hole/arch for the fit you desire.

#### Splitting and Connecting 3D Prints {#sec:splitting-connecting}

3D printers will never be big enough for the biggest parts you would like to print. But the slicing software can help you by cutting the print into smaller parts. It also adds dowels and holes for aligning the printed parts back together. Use Crazy glue or equivalent CA glue. Read these [instructions on how to cut your model in Prusa Slicer](https://help.prusa3d.com/article/cut-tool_1779).

#### Vase Mode {#sec:vase-mode}

Vase Mode is a special mode the slicer offers where the CAD model is printed as a single wall in a spiral pattern. No infill and no overhangs will be generated. There won't be any seams where the slicer steps Z height up a step. Instead, the slicer builds up the part in a single bead of material following a spiral representing the outer contour only.

### Print Defects {#sec:print-defects}

There are many print defects. If you encounter one, check out this [excellent summary](https://www.simplify3d.com/resources/print-quality-troubleshooting/) for solution suggestions.

### Special and Upcoming Techniques {#sec:special-techniques}

**Hairy Models**: [The Hairy Lion](https://www.printables.com/model/119009-the-hairy-lion)

**Printable Velcro**:
- [Printable Velcro](https://www.printables.com/model/543802-printable-velcro)
- [Hook & Loop Prints Contest](https://www.printables.com/contest/397-hook-loop-prints)

**Arc Overhangs**: [Arc Overhangs in PrusaSlicer](https://www.cnckitchen.com/blog/arc-overhangs-in-prusaslicer)

**Brick Layers**:
- [Brick Layers Make 3D Prints Stronger](https://www.cnckitchen.com/blog/brick-layers-make-3d-prints-stronger)
- [Bricklayers GitHub Repository](https://github.com/TengerTechnologies/Bricklayers)

### Post Processing {#sec:post-processing}

#### Annealing {#sec:annealing}

Annealing is a heat treatment process that can improve the mechanical properties of 3D printed parts. By heating the part to a temperature below its melting point and then slowly cooling it, internal stresses are relieved and the crystalline structure of the material can be improved. This is particularly useful for PLA and other semi-crystalline polymers, resulting in increased strength, heat resistance, and dimensional stability.

### Check Your Understanding

1. What is the minimum recommended wall thickness for a printer with a 0.4mm nozzle?
   a) 0.2mm
   b) 0.4mm
   c) 0.8mm
   d) 1.2mm
   **Answer: c**

2. What is the maximum overhang angle that can typically be printed without support material?
   a) 30°
   b) 45°
   c) 60°
   d) 90°
   **Answer: b**

3. Which 3D printing technology uses a laser to cure liquid resin?
   a) FDM
   b) FFF
   c) SLS
   d) SLA
   **Answer: d**

4. True or False: Parts printed with FDM have equal strength in all directions.
   **Answer: False** (Parts are stronger in the XY direction than the Z direction)

5. How much should you add to vertical hole diameters to compensate for undersizing?
   a) 0.1mm
   b) 0.2mm
   c) 0.5mm
   d) 1.0mm
   **Answer: b**
