# Design for Stamping

```{r stamping-setup, include=FALSE}
library(ggplot2)
library(grid)

# Set a consistent theme for all plots
theme_stamping <- theme_minimal() +
theme(
 plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
 plot.subtitle = element_text(hjust = 0.5, size = 10),
 axis.text = element_blank(),
 axis.title = element_blank(),
 axis.ticks = element_blank(),
 panel.grid = element_blank(),
 plot.background = element_rect(fill = "white", color = NA),
 panel.background = element_rect(fill = "white", color = NA)
)
```

## 💡 Introduction to the Stamping Process {-}

Metal stamping is one of the most versatile and widely used manufacturing processes in modern industry. It involves placing flat sheet metal—either in coil or blank form—into a stamping press where a tool and die surface forms the metal into the desired shape. Stamping encompasses a variety of operations including **blanking, punching, bending, drawing, embossing, and coining**.

The significance of stamping lies in its ability to produce high volumes of identical parts with excellent dimensional accuracy, tight tolerances, and superior surface finish—all at relatively low cost per part. This makes stamping essential for industries ranging from automotive and aerospace to electronics and consumer appliances.

---

> **🤔 Think About It:** Look around you right now. How many stamped metal parts can you identify? Consider items like your phone's SIM tray, car body panels, kitchen utensils, or even the clips holding papers together.

---

## 🧱 Essential Components of Stamping {-}

Every stamping operation requires several basic elements working in harmony:
1. **The Sheet Metal (Workpiece):** This is the raw material, typically supplied as coils or flat sheets. Common materials include steel, aluminum, brass, copper, and stainless steel.

2. **The Die:** A specialized tool that contains the negative (female) form of the part. The die is mounted on the press bed.

3. **The Punch:** The male counterpart that presses the sheet metal into or through the die cavity.

4. **The Press:** The machine that provides the force to drive the punch into the die. Presses can be mechanical, hydraulic, or servo-driven.

5. **The Blank Holder (in drawing operations):** A component that controls material flow and prevents wrinkling during deep drawing operations.

```{r basic-stamping-setup, echo=FALSE, fig.width=8, fig.height=6, fig.align='center', fig.cap="Basic stamping setup showing punch, die, and workpiece arrangement."}
ggplot() +
 # Press ram (top)
 geom_rect(aes(xmin = 2, xmax = 8, ymin = 8, ymax = 9), fill = "gray40", color = "black", linewidth = 1) +
 annotate("text", x = 5, y = 8.5, label = "Press Ram", color = "white", fontface = "bold", size = 4) +

 # Punch
 geom_rect(aes(xmin = 4, xmax = 6, ymin = 5.5, ymax = 8), fill = "steelblue", color = "black", linewidth = 1) +
 annotate("text", x = 5, y = 6.75, label = "Punch", color = "white", fontface = "bold", size = 4) +

 # Sheet metal (workpiece)
 geom_rect(aes(xmin = 1, xmax = 9, ymin = 5, ymax = 5.5), fill = "orange", color = "black", linewidth = 1) +
 annotate("text", x = 1.8, y = 5.25, label = "Sheet Metal", color = "black", fontface = "bold", size = 3) +

 # Die (with cavity)
 geom_rect(aes(xmin = 1, xmax = 3.8, ymin = 3, ymax = 5), fill = "gray60", color = "black", linewidth = 1) +
 geom_rect(aes(xmin = 6.2, xmax = 9, ymin = 3, ymax = 5), fill = "gray60", color = "black", linewidth = 1) +
 geom_rect(aes(xmin = 3.8, xmax = 6.2, ymin = 2, ymax = 5), fill = "white", color = "black", linewidth = 1) +
 annotate("text", x = 2.4, y = 4, label = "Die", color = "white", fontface = "bold", size = 4) +
 annotate("text", x = 5, y = 3.5, label = "Die\nCavity", color = "gray40", fontface = "bold", size = 3) +

 # Press bed (bottom)
 geom_rect(aes(xmin = 0, xmax = 10, ymin = 1, ymax = 2), fill = "gray40", color = "black", linewidth = 1) +
 annotate("text", x = 5, y = 1.5, label = "Press Bed", color = "white", fontface = "bold", size = 4) +

 # Force arrow
 geom_segment(aes(x = 5, xend = 5, y = 10, yend = 9.2),
             arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
             linewidth = 1.5, color = "red") +
 annotate("text", x = 5, y = 10.3, label = "Force (F)", color = "red", fontface = "bold", size = 4) +

 # Clearance annotation
 geom_segment(aes(x = 6, xend = 6.2, y = 5.25, yend = 5.25), linewidth = 0.5) +
 geom_segment(aes(x = 6, xend = 6, y = 5.1, yend = 5.4), linewidth = 0.5) +
 geom_segment(aes(x = 6.2, xend = 6.2, y = 5.1, yend = 5.4), linewidth = 0.5) +
 annotate("text", x = 6.7, y = 5.25, label = "Clearance", size = 3) +

 coord_fixed(ratio = 1, xlim = c(0, 10), ylim = c(0, 11)) +
 labs(title = "Basic Stamping Setup") +
 theme_stamping
```

---

## ⚙️ Types of Stamping Operations {-}

Stamping is not a single process but a family of related operations. Understanding each operation is crucial for designing parts that can be efficiently manufactured.

### Blanking and Punching

**Blanking** and **punching** are shearing operations that cut sheet metal using a punch and die.

* **Blanking:** The cut-out piece (the blank) is the desired part, and the remaining sheet is scrap.
* **Punching (Piercing):** The cut-out piece is scrap (the slug), and the remaining sheet with the hole is the desired part.

```{r blanking-vs-punching, echo=FALSE, fig.width=10, fig.height=5, fig.align='center', fig.cap="Blanking vs. Punching: In blanking, the cut piece is the product. In punching, the sheet with the hole is the product."}
# Create data for circles
circle_pts <- function(cx, cy, r, n = 100) {
 t <- seq(0, 2*pi, length.out = n)
 data.frame(x = cx + r * cos(t), y = cy + r * sin(t))
}

blanking_circle <- circle_pts(2.5, 2, 1)
punching_circle <- circle_pts(7.5, 2, 1)

ggplot() +
 # BLANKING SIDE
 # Sheet with hole (scrap)
 geom_rect(aes(xmin = 0.5, xmax = 4.5, ymin = 0.5, ymax = 3.5),
           fill = "gray70", color = "black", linewidth = 1) +
 geom_polygon(data = blanking_circle, aes(x = x, y = y), fill = "white", color = "black", linewidth = 1) +
 annotate("text", x = 2.5, y = 4.2, label = "BLANKING", fontface = "bold", size = 5) +
 annotate("text", x = 1.2, y = 3.2, label = "Scrap", color = "gray40", size = 3.5) +

 # Blank (product) - shown below
 geom_polygon(data = circle_pts(2.5, -1, 1), aes(x = x, y = y),
             fill = "forestgreen", color = "black", linewidth = 1) +
 annotate("text", x = 2.5, y = -1, label = "Product\n(Blank)", color = "white", fontface = "bold", size = 3.5) +

 # Arrow
 geom_segment(aes(x = 2.5, xend = 2.5, y = 0.3, yend = 0.1),
             arrow = arrow(length = unit(0.2, "cm")), linewidth = 1) +

 # PUNCHING SIDE
 # Sheet with hole (product)
 geom_rect(aes(xmin = 5.5, xmax = 9.5, ymin = 0.5, ymax = 3.5),
           fill = "forestgreen", color = "black", linewidth = 1) +
 geom_polygon(data = punching_circle, aes(x = x, y = y), fill = "white", color = "black", linewidth = 1) +
 annotate("text", x = 7.5, y = 4.2, label = "PUNCHING", fontface = "bold", size = 5) +
 annotate("text", x = 6.2, y = 3.2, label = "Product", color = "white", fontface = "bold", size = 3.5) +

 # Slug (scrap) - shown below
 geom_polygon(data = circle_pts(7.5, -1, 1), aes(x = x, y = y),
             fill = "gray70", color = "black", linewidth = 1) +
 annotate("text", x = 7.5, y = -1, label = "Scrap\n(Slug)", color = "gray40", size = 3.5) +

 # Arrow
 geom_segment(aes(x = 7.5, xend = 7.5, y = 0.3, yend = 0.1),
             arrow = arrow(length = unit(0.2, "cm")), linewidth = 1) +

 # Dividing line
 geom_vline(xintercept = 5, linetype = "dashed", color = "gray50") +

 coord_fixed(ratio = 1, xlim = c(0, 10), ylim = c(-2.5, 5)) +
 theme_stamping
```

> **❓ Question for Reflection:** A manufacturer needs to create both circular washers and sheets with circular holes for ventilation. Which operation would they use for each product?

<details>
<summary>Click to reveal the answer</summary>

* **Washers:** Blanking—the circular cut-out is the desired product.
* **Ventilation sheets:** Punching—the sheet with holes is the desired product, and the circular slugs are scrap.

</details>

---

### Bending

**Bending** is a forming operation where the sheet metal is deformed along a straight axis to create an angled shape. The metal on the outside of the bend is stretched (in tension), while the metal on the inside is compressed.

```{r bending-operation, echo=FALSE, fig.width=8, fig.height=6, fig.align='center', fig.cap="V-bending operation showing the punch, die, and bent workpiece."}
ggplot() +
 # V-Die (bottom)
 geom_polygon(aes(x = c(0, 3, 5, 7, 10, 10, 0),
                  y = c(3, 3, 0.5, 3, 3, 0, 0)),
             fill = "gray60", color = "black", linewidth = 1) +
 annotate("text", x = 1.5, y = 1.5, label = "Die", color = "white", fontface = "bold", size = 4) +

 # V-Punch (top)
 geom_polygon(aes(x = c(3, 5, 7, 7, 3),
                  y = c(7, 4, 7, 8, 8)),
             fill = "steelblue", color = "black", linewidth = 1) +
 annotate("text", x = 5, y = 7.2, label = "Punch", color = "white", fontface = "bold", size = 4) +

 # Bent sheet metal
 geom_polygon(aes(x = c(1.5, 5, 8.5, 8.3, 5, 1.7),
                  y = c(4.2, 2, 4.2, 4.5, 2.3, 4.5)),
             fill = "orange", color = "black", linewidth = 1) +
 annotate("text", x = 7.5, y = 5, label = "Workpiece", color = "darkorange", fontface = "bold", size = 3.5) +

 # Bend angle annotation
 geom_curve(aes(x = 5, y = 3.5, xend = 6.5, yend = 4.2),
           curvature = -0.3, linewidth = 0.5, color = "red") +
 annotate("text", x = 6.8, y = 3.5, label = "Bend\nAngle", color = "red", size = 3) +

 # Bend radius annotation
 geom_point(aes(x = 5, y = 2.15), size = 2, color = "red") +
 annotate("text", x = 5, y = 1.3, label = "Bend Radius (R)", color = "red", size = 3) +

 # Tension/Compression labels
 annotate("text", x = 3, y = 3.8, label = "Tension\n(outside)", color = "blue", size = 2.5, fontface = "italic") +
 annotate("text", x = 3.2, y = 2.8, label = "Compression\n(inside)", color = "purple", size = 2.5, fontface = "italic") +

 # Force arrow
 geom_segment(aes(x = 5, xend = 5, y = 9, yend = 8.2),
             arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
             linewidth = 1.5, color = "red") +
 annotate("text", x = 5, y = 9.3, label = "F", color = "red", fontface = "bold", size = 5) +

 coord_fixed(ratio = 1, xlim = c(-0.5, 10.5), ylim = c(-0.5, 10)) +
 labs(title = "V-Bending Operation") +
 theme_stamping
```

Key bending terminology:

| Term | Definition |
|:-----|:-----------|
| **Bend Radius** | The inside radius of the bend |
| **Bend Angle** | The angle of the bend measured from the original flat position |
| **Springback** | The elastic recovery of the material after the bending force is removed |
| **Bend Allowance** | The arc length of the neutral axis in the bend region |
| **K-Factor** | The ratio of the neutral axis location to the material thickness |

> **🤔 Critical Thinking:** Why do you think the neutral axis (where there is neither tension nor compression) is NOT exactly at the center of the sheet thickness during bending?

---

### Drawing

**Drawing** is a process that transforms a flat sheet metal blank into a hollow, three-dimensional shape (like a cup or box) by pulling the material into a die cavity with a punch.

```{r deep-drawing-process, echo=FALSE, fig.width=10, fig.height=5, fig.align='center', fig.cap="Deep drawing process showing blank transformation into a cup shape."}
ggplot() +
 # BEFORE (left side)
 # Blank holder top
 geom_rect(aes(xmin = 0.3, xmax = 1.5, ymin = 3, ymax = 3.5), fill = "gray50", color = "black") +
 geom_rect(aes(xmin = 3.5, xmax = 4.7, ymin = 3, ymax = 3.5), fill = "gray50", color = "black") +
 annotate("text", x = 0.9, y = 3.8, label = "Blank\nHolder", size = 2.5) +

 # Flat blank
 geom_rect(aes(xmin = 0.3, xmax = 4.7, ymin = 2.7, ymax = 3), fill = "orange", color = "black", linewidth = 1) +

 # Die
 geom_rect(aes(xmin = 0, xmax = 1.5, ymin = 0, ymax = 2.7), fill = "gray60", color = "black") +
 geom_rect(aes(xmin = 3.5, xmax = 5, ymin = 0, ymax = 2.7), fill = "gray60", color = "black") +
 annotate("text", x = 0.75, y = 1.3, label = "Die", color = "white", fontface = "bold", size = 4) +

 # Punch (above)
 geom_rect(aes(xmin = 1.7, xmax = 3.3, ymin = 3.5, ymax = 5), fill = "steelblue", color = "black") +
 annotate("text", x = 2.5, y = 4.25, label = "Punch", color = "white", fontface = "bold", size = 3.5) +

 # Labels
 annotate("text", x = 2.5, y = 5.8, label = "BEFORE", fontface = "bold", size = 5) +
 annotate("text", x = 2.5, y = 2.3, label = expression(D[b]), size = 4) +
 geom_segment(aes(x = 0.3, xend = 4.7, y = 2.4, yend = 2.4),
             arrow = arrow(length = unit(0.15, "cm"), ends = "both"), linewidth = 0.5) +

 # Arrow between stages
 geom_segment(aes(x = 5.5, xend = 6.5, y = 3, yend = 3),
             arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
             linewidth = 1.5, color = "darkgreen") +

 # AFTER (right side)
 # Blank holder
 geom_rect(aes(xmin = 7.3, xmax = 8.3, ymin = 3, ymax = 3.5), fill = "gray50", color = "black") +
 geom_rect(aes(xmin = 10.7, xmax = 11.7, ymin = 3, ymax = 3.5), fill = "gray50", color = "black") +

 # Die
 geom_rect(aes(xmin = 7, xmax = 8.5, ymin = 0, ymax = 2.7), fill = "gray60", color = "black") +
 geom_rect(aes(xmin = 10.5, xmax = 12, ymin = 0, ymax = 2.7), fill = "gray60", color = "black") +

 # Drawn cup shape
 geom_polygon(aes(x = c(7.3, 8.3, 8.5, 8.7, 8.7, 10.3, 10.3, 10.5, 10.7, 11.7, 11.7, 7.3),
                  y = c(3, 3, 2.7, 2.5, 0.5, 0.5, 2.5, 2.7, 3, 3, 2.7, 2.7)),
             fill = "orange", color = "black", linewidth = 1) +

 # Punch (lowered)
 geom_rect(aes(xmin = 8.7, xmax = 10.3, ymin = 0.5, ymax = 5), fill = "steelblue", color = "black") +
 annotate("text", x = 9.5, y = 4, label = "Punch", color = "white", fontface = "bold", size = 3.5) +

 # Labels
 annotate("text", x = 9.5, y = 5.8, label = "AFTER", fontface = "bold", size = 5) +
 annotate("text", x = 9.5, y = -0.3, label = expression(D[p]), size = 4) +
 geom_segment(aes(x = 8.7, xend = 10.3, y = -0.1, yend = -0.1),
             arrow = arrow(length = unit(0.15, "cm"), ends = "both"), linewidth = 0.5) +

 coord_fixed(ratio = 1, xlim = c(-0.5, 12.5), ylim = c(-1, 6.5)) +
 labs(title = "Deep Drawing Process") +
 theme_stamping
```

The distinction between **shallow drawing** and **deep drawing** is based on the ratio of drawing depth to diameter:

$$\text{Drawing Ratio} = \frac{D_b}{D_p}$$

Where:

* $D_b$ = Diameter of the original blank
* $D_p$ = Diameter of the punch (final part)

When the drawing ratio exceeds approximately **2.0**, the operation typically requires multiple drawing stages (redrawing) to avoid failure.

> **❓ Calculation Challenge:** A beverage can manufacturer wants to draw a can body with a 66 mm diameter from a circular blank. If the maximum single-draw ratio is 2.0, what is the minimum blank diameter that can be used in a single drawing operation?

<details>
<summary>Click to reveal the solution</summary>

Using the drawing ratio formula:

$$D_b = D_p \times \text{Drawing Ratio} = 66 \text{ mm} \times 2.0 = 132 \text{ mm}$$

The maximum blank diameter for a single draw operation is **132 mm**. Any larger blank would require multiple drawing stages.

</details>

---

### Embossing and Coining

* **Embossing:** Creates raised or recessed designs in sheet metal without changing the material thickness significantly. One side has a raised pattern while the other has a corresponding depression.

* **Coining:** A severe squeezing operation that creates fine detail by forcing metal to flow into die cavities under very high pressure. Unlike embossing, coining can significantly change local thickness.

```{r embossing-coining, echo=FALSE, fig.width=10, fig.height=5, fig.align='center', fig.cap="Comparison of embossing (left) and coining (right) operations."}
ggplot() +
 # EMBOSSING (left)
 annotate("text", x = 2.5, y = 5.5, label = "EMBOSSING", fontface = "bold", size = 5) +

 # Embossing punch
 geom_polygon(aes(x = c(0.5, 0.5, 2, 2.5, 3, 4.5, 4.5),
                  y = c(4.5, 3.5, 3.5, 3, 3.5, 3.5, 4.5)),
             fill = "steelblue", color = "black", linewidth = 1) +

 # Embossed sheet (wavy)
 geom_polygon(aes(x = c(0, 0, 2, 2.5, 3, 5, 5, 3, 2.5, 2, 0),
                  y = c(2.7, 3, 3, 2.5, 3, 3, 2.7, 2.7, 2.2, 2.7, 2.7)),
             fill = "orange", color = "black", linewidth = 1) +

 # Embossing die
 geom_polygon(aes(x = c(0, 0, 2, 2.5, 3, 5, 5),
                  y = c(0, 2.2, 2.2, 1.7, 2.2, 2.2, 0)),
             fill = "gray60", color = "black", linewidth = 1) +

 # Thickness annotation
 annotate("text", x = 0.8, y = 2.85, label = "t", size = 3, color = "red") +
 annotate("text", x = 2.5, y = 2.35, label = "≈ t", size = 3, color = "red") +
 annotate("text", x = 2.5, y = 0.8, label = "Thickness\nnearly constant", size = 3, fontface = "italic") +

 # COINING (right)
 annotate("text", x = 8.5, y = 5.5, label = "COINING", fontface = "bold", size = 5) +

 # Coining punch (flat with detail)
 geom_rect(aes(xmin = 6.5, xmax = 10.5, ymin = 3.2, ymax = 4.5), fill = "steelblue", color = "black", linewidth = 1) +
 # Detail pattern
 geom_rect(aes(xmin = 7.5, xmax = 8, ymin = 3.2, ymax = 3.4), fill = "steelblue4", color = "black") +
 geom_rect(aes(xmin = 8.3, xmax = 9, ymin = 3.2, ymax = 3.5), fill = "steelblue4", color = "black") +
 geom_rect(aes(xmin = 9.3, xmax = 9.6, ymin = 3.2, ymax = 3.35), fill = "steelblue4", color = "black") +

 # Coined sheet (with impressed detail, variable thickness)
 geom_polygon(aes(x = c(6, 6, 7.5, 7.5, 8, 8, 8.3, 8.3, 9, 9, 9.3, 9.3, 9.6, 9.6, 11, 11,
                       9.6, 9.6, 9.3, 9.3, 9, 9, 8.3, 8.3, 8, 8, 7.5, 7.5, 6),
                  y = c(2.7, 3.2, 3.2, 3, 3, 3.2, 3.2, 2.9, 2.9, 3.2, 3.2, 3.05, 3.05, 3.2, 3.2, 2.7,
                       2.7, 2.7, 2.7, 2.7, 2.7, 2.7, 2.7, 2.7, 2.7, 2.7, 2.7, 2.7, 2.7)),
             fill = "orange", color = "black", linewidth = 1) +

 # Coining die
 geom_rect(aes(xmin = 6, xmax = 11, ymin = 0, ymax = 2.7), fill = "gray60", color = "black", linewidth = 1) +

 # Thickness annotation
 annotate("text", x = 6.7, y = 2.95, label = "t", size = 3, color = "red") +
 annotate("text", x = 8.65, y = 2.8, label = "< t", size = 3, color = "red") +
 annotate("text", x = 8.5, y = 0.8, label = "Metal flows;\nthickness changes", size = 3, fontface = "italic") +

 # High pressure arrows
 geom_segment(aes(x = 8.5, xend = 8.5, y = 5.2, yend = 4.7),
             arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
             linewidth = 1.2, color = "red") +
 annotate("text", x = 8.5, y = 5.4, label = "High F", color = "red", size = 3) +

 # Dividing line
 geom_vline(xintercept = 5.5, linetype = "dashed", color = "gray50") +

 coord_fixed(ratio = 1, xlim = c(-0.5, 11.5), ylim = c(-0.5, 6)) +
 theme_stamping
```

> **🤔 Think About It:** Why are coins made using the coining process rather than embossing? Consider factors like detail, durability, and counterfeit prevention.

---

## 🏭 Stamping Press Types and Progressive Dies {-}

### Types of Stamping Presses

| Press Type | Force Mechanism | Characteristics | Best Applications |
|:-----------|:----------------|:----------------|:------------------|
| **Mechanical** | Flywheel and crankshaft | High speed, consistent stroke | High-volume blanking and forming |
| **Hydraulic** | Fluid pressure | Variable speed, full force throughout stroke | Deep drawing, complex forming |
| **Servo** | Electric servo motors | Programmable motion profiles, energy efficient | Precision parts, variable operations |

### Progressive Die Stamping

Progressive die stamping is a highly efficient method where a coil of sheet metal is fed through a series of stations, each performing a different operation. The part remains attached to a carrier strip until the final station, where it is cut free.

```{r progressive-die, echo=FALSE, fig.width=12, fig.height=5, fig.align='center', fig.cap="Progressive die stamping: multiple operations performed sequentially as the strip advances through the die."}
ggplot() +
 # Strip (carrier)
 geom_rect(aes(xmin = 0, xmax = 14, ymin = 2, ymax = 2.8), fill = "orange", color = "black", linewidth = 1) +

 # Pilot holes (for alignment)
 geom_point(aes(x = c(1, 3.5, 6, 8.5, 11), y = rep(2.4, 5)), size = 3, color = "white") +
 geom_point(aes(x = c(1, 3.5, 6, 8.5, 11), y = rep(2.4, 5)), size = 3, shape = 1, color = "black") +

 # Station 1: Pilot holes + blanking
 geom_rect(aes(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5), fill = "steelblue", color = "black") +
 annotate("text", x = 1, y = 5, label = "Station 1\nPilot Holes", size = 3, fontface = "bold") +

 # Station 2: Piercing
 geom_rect(aes(xmin = 3, xmax = 4, ymin = 3.5, ymax = 4.5), fill = "steelblue", color = "black") +
 geom_point(aes(x = 3.5, y = 2.4), size = 5, color = "white") +
 annotate("text", x = 3.5, y = 5, label = "Station 2\nPiercing", size = 3, fontface = "bold") +

 # Station 3: Forming/Bending
 geom_polygon(aes(x = c(5.5, 5.5, 6, 6.5, 6.5), y = c(4.5, 3.5, 3.2, 3.5, 4.5)),
             fill = "steelblue", color = "black") +
 # Show bent tab
 geom_polygon(aes(x = c(5.7, 6, 6.3, 6.3, 6, 5.7), y = c(2.8, 2.5, 2.8, 2.6, 2.3, 2.6)),
             fill = "darkorange", color = "black") +
 annotate("text", x = 6, y = 5, label = "Station 3\nForming", size = 3, fontface = "bold") +

 # Station 4: Additional forming
 geom_rect(aes(xmin = 8, xmax = 9, ymin = 3.5, ymax = 4.5), fill = "steelblue", color = "black") +
 annotate("text", x = 8.5, y = 5, label = "Station 4\nCoining", size = 3, fontface = "bold") +

 # Station 5: Cutoff
 geom_polygon(aes(x = c(10.3, 10.5, 11.5, 11.7, 11.7, 10.3),
                  y = c(4.5, 3.5, 3.5, 4.5, 4.5, 4.5)), fill = "steelblue", color = "black") +
 annotate("text", x = 11, y = 5, label = "Station 5\nCutoff", size = 3, fontface = "bold") +

 # Finished part (falling)
 geom_polygon(aes(x = c(11.5, 12.5, 12.8, 12.8, 11.2, 11.2),
                  y = c(0.8, 0.8, 1, 1.5, 1.5, 1)),
             fill = "forestgreen", color = "black", linewidth = 1) +
 annotate("text", x = 12, y = 0.3, label = "Finished\nPart", size = 3, color = "darkgreen", fontface = "bold") +

 # Feed direction arrow
 geom_segment(aes(x = 0.5, xend = 3, y = 1.3, yend = 1.3),
             arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
             linewidth = 1.5, color = "darkgreen") +
 annotate("text", x = 1.75, y = 0.8, label = "Feed Direction", size = 3.5, fontface = "bold") +

 # Die block (bottom)
 geom_rect(aes(xmin = 0, xmax = 14, ymin = -0.2, ymax = 2), fill = "gray60", color = "black") +
 annotate("text", x = 7, y = 0.9, label = "DIE BLOCK", color = "white", fontface = "bold", size = 5) +

 # Scrap label
 annotate("text", x = 13.2, y = 2.4, label = "Scrap\n(carrier)", size = 2.5, color = "gray40") +

 coord_fixed(ratio = 1, xlim = c(-0.5, 14.5), ylim = c(-0.5, 5.5)) +
 labs(title = "Progressive Die Stamping - Multiple Stations") +
 theme_stamping
```

**Advantages of Progressive Dies:**

* Extremely high production rates (hundreds of parts per minute)
* Multiple operations in a single press stroke
* Reduced handling and labor costs
* Consistent part quality

> **❓ Design Question:** If you were designing a simple bracket that requires three holes and two 90° bends, sketch how you might arrange the stations in a progressive die. What would be the sequence of operations?

---

```{=html}
<div style="position: relative; padding-bottom: 56.25%; height: 0; overflow: hidden; max-width: 100%;">
  <iframe
    style="position: absolute; top: 0; left: 0; width: 100%; height: 100%;"
    src="https://www.youtube.com/embed/RPzv3WJ3mqM"
    title="Advantages of Progressive Die Stamping"
    frameborder="0"
    allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share"
    referrerpolicy="strict-origin-when-cross-origin"
    allowfullscreen>
  </iframe>
</div>
```

```{=html}
<div style="position: relative; padding-bottom: 56.25%; height: 0; overflow: hidden; max-width: 100%;">
  <iframe
    style="position: absolute; top: 0; left: 0; width: 100%; height: 100%;"
    src="https://www.youtube.com/embed/NuFpzJLMnFs"
    title="Progressive Die Stamping"
    frameborder="0"
    allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share"
    referrerpolicy="strict-origin-when-cross-origin"
    allowfullscreen>
  </iframe>
</div>
```


---

## 📊 Advantages and Limitations of Stamping {-}

| Aspect | Advantages | Limitations |
|:-------|:-----------|:------------|
| **Production Volume** | Extremely economical for high volumes; can produce thousands of parts per hour | High tooling costs make it uneconomical for low volumes (typically need >10,000 parts) |
| **Precision** | Excellent repeatability and tight tolerances (±0.1 mm achievable) | Complex 3D shapes may require multiple operations or secondary processes |
| **Material Utilization** | Can optimize blank layouts to minimize scrap | Some operations inherently produce scrap (e.g., punching slugs) |
| **Surface Finish** | Good surface finish directly from the process | Die wear can affect surface quality over time |
| **Part Strength** | Work hardening can increase strength in formed areas | Excessive forming can lead to cracking or tearing |
| **Lead Time** | Fast production once tooling is ready | Long lead time for die design and fabrication (weeks to months) |

---

## 🔬 Material Considerations for Stamping {-}

Not all materials are equally suitable for stamping. The key material properties that influence stampability include:

### Formability

**Formability** describes how much a material can be deformed before it cracks or tears. It is influenced by:

* **Ductility:** The ability to undergo plastic deformation without fracture
* **Work Hardening Rate (n-value):** How quickly the material strengthens as it deforms
* **Anisotropy (r-value):** Directional differences in properties due to rolling direction

```{r forming-limit-diagram, echo=FALSE, fig.width=8, fig.height=6, fig.align='center', fig.cap="Forming Limit Diagram (FLD) showing safe and failure zones for different strain combinations."}
# Create FLD curve data
minor_strain <- seq(-0.3, 0.4, length.out = 100)
# FLD curve (simplified parabolic shape)
major_strain_limit <- ifelse(minor_strain < 0,
                             0.4 + 0.8 * minor_strain,
                             0.4 + 0.3 * minor_strain - 0.5 * minor_strain^2)

fld_curve <- data.frame(minor = minor_strain, major = major_strain_limit)

ggplot() +
 # Failure zone (above curve)
 geom_ribbon(data = fld_curve, aes(x = minor, ymin = major, ymax = 0.8),
            fill = "red", alpha = 0.3) +

 # Safe zone (below curve)
 geom_ribbon(data = fld_curve, aes(x = minor, ymin = -0.1, ymax = major),
            fill = "green", alpha = 0.3) +

 # FLD curve
 geom_line(data = fld_curve, aes(x = minor, y = major), linewidth = 2, color = "darkred") +

 # Marginal zone
 geom_line(data = fld_curve, aes(x = minor, y = major - 0.05),
          linewidth = 1, color = "orange", linetype = "dashed") +

 # Zone labels
 annotate("text", x = 0, y = 0.65, label = "FAILURE\nZONE", color = "darkred",
         fontface = "bold", size = 5) +
 annotate("text", x = 0, y = 0.15, label = "SAFE\nZONE", color = "darkgreen",
         fontface = "bold", size = 5) +
 annotate("text", x = 0.25, y = 0.42, label = "Forming Limit\nCurve (FLC)",
         color = "darkred", size = 3.5, fontface = "italic") +

 # Strain state examples
 geom_point(aes(x = -0.15, y = 0.25), size = 4, color = "blue") +
 annotate("text", x = -0.15, y = 0.20, label = "Drawing", color = "blue", size = 3) +

 geom_point(aes(x = 0, y = 0.30), size = 4, color = "purple") +
 annotate("text", x = 0.08, y = 0.28, label = "Plane\nStrain", color = "purple", size = 3) +

 geom_point(aes(x = 0.15, y = 0.25), size = 4, color = "darkorange") +
 annotate("text", x = 0.15, y = 0.20, label = "Stretching", color = "darkorange", size = 3) +

 # Axes
 geom_hline(yintercept = 0, linewidth = 0.5) +
 geom_vline(xintercept = 0, linewidth = 0.5) +

 scale_x_continuous(breaks = seq(-0.3, 0.4, 0.1),
                   labels = scales::percent_format(accuracy = 1)) +
 scale_y_continuous(breaks = seq(0, 0.7, 0.1),
                   labels = scales::percent_format(accuracy = 1)) +

 labs(title = "Forming Limit Diagram (FLD)",
     x = "Minor Strain (ε₂)",
     y = "Major Strain (ε₁)") +
 theme_minimal() +
 theme(
   plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
   axis.title = element_text(size = 11),
   panel.grid.minor = element_blank()
 ) +
 coord_cartesian(xlim = c(-0.35, 0.45), ylim = c(-0.05, 0.75))
```

The **Forming Limit Diagram (FLD)** is a critical tool that shows the combination of major and minor strains a material can withstand before failure.

### Common Stamping Materials

| Material | Formability | Typical Applications | Special Considerations |
|:---------|:------------|:---------------------|:-----------------------|
| **Low Carbon Steel** | Excellent | Automotive body panels, appliances | Most economical; good for deep drawing |
| **Stainless Steel** | Good to Moderate | Kitchen equipment, medical devices | Work hardens rapidly; requires more force |
| **Aluminum Alloys** | Good | Aerospace, beverage cans | Springback issues; galling tendency |
| **Brass** | Excellent | Electrical connectors, decorative items | Expensive but forms easily |
| **Copper** | Excellent | Electrical components | Very ductile but soft |

> **🤔 Material Selection Challenge:** You are designing a stamped enclosure for outdoor electrical equipment. The part requires moderate drawing depth and must resist corrosion. Rank these materials from most to least suitable: Low carbon steel, 304 stainless steel, Aluminum 5052, Brass. Explain your reasoning.

<details>
<summary>Click to reveal analysis</summary>

**Recommended ranking:**

1. **304 Stainless Steel** - Excellent corrosion resistance, good formability, strong and durable
2. **Aluminum 5052** - Good corrosion resistance, lighter weight, but may require thicker gauge for strength
3. **Brass** - Good corrosion resistance but expensive and may be overkill
4. **Low Carbon Steel** - Poor corrosion resistance unless coated/painted; would require additional finishing

The final choice depends on budget constraints and specific environmental conditions.

</details>

---

## 📐 Design Guidelines for Stamping {-}

Designing parts for efficient stamping requires understanding the capabilities and limitations of the process. Following these guidelines will result in lower costs, better quality, and faster production.

### Hole Design

#### Minimum Hole Size

The minimum hole diameter should be at least equal to the material thickness to prevent punch breakage.

$$d_{min} \geq t$$

Where:

* $d_{min}$ = Minimum hole diameter
* $t$ = Material thickness

For non-circular holes, the minimum dimension should follow the same rule.

```{r minimum-hole-size, echo=FALSE, fig.width=8, fig.height=4, fig.align='center', fig.cap="Minimum hole size relative to material thickness."}
ggplot() +
 # Sheet cross-section (left - bad)
 geom_rect(aes(xmin = 0.5, xmax = 3.5, ymin = 1.5, ymax = 2.5), fill = "orange", color = "black", linewidth = 1) +
 # Small hole
 geom_rect(aes(xmin = 1.7, xmax = 2.3, ymin = 1.5, ymax = 2.5), fill = "white", color = "black") +
 # Thickness dimension
 geom_segment(aes(x = 0.3, xend = 0.3, y = 1.5, yend = 2.5), linewidth = 0.5) +
 geom_segment(aes(x = 0.2, xend = 0.4, y = 1.5, yend = 1.5), linewidth = 0.5) +
 geom_segment(aes(x = 0.2, xend = 0.4, y = 2.5, yend = 2.5), linewidth = 0.5) +
 annotate("text", x = 0.3, y = 2, label = "t", size = 4, fontface = "bold") +
 # Hole dimension
 geom_segment(aes(x = 1.7, xend = 2.3, y = 2.8, yend = 2.8),
             arrow = arrow(length = unit(0.1, "cm"), ends = "both"), linewidth = 0.5) +
 annotate("text", x = 2, y = 3.1, label = "d < t", size = 4, color = "red") +
 # Bad label
 annotate("text", x = 2, y = 0.8, label = "TOO SMALL", color = "red", fontface = "bold", size = 4) +
 geom_point(aes(x = 2, y = 1.1), shape = 4, size = 8, color = "red", stroke = 2) +

 # Sheet cross-section (right - good)
 geom_rect(aes(xmin = 5, xmax = 8, ymin = 1.5, ymax = 2.5), fill = "orange", color = "black", linewidth = 1) +
 # Proper hole
 geom_rect(aes(xmin = 5.8, xmax = 7.2, ymin = 1.5, ymax = 2.5), fill = "white", color = "black") +
 # Thickness dimension
 geom_segment(aes(x = 4.8, xend = 4.8, y = 1.5, yend = 2.5), linewidth = 0.5) +
 geom_segment(aes(x = 4.7, xend = 4.9, y = 1.5, yend = 1.5), linewidth = 0.5) +
 geom_segment(aes(x = 4.7, xend = 4.9, y = 2.5, yend = 2.5), linewidth = 0.5) +
 annotate("text", x = 4.8, y = 2, label = "t", size = 4, fontface = "bold") +
 # Hole dimension
 geom_segment(aes(x = 5.8, xend = 7.2, y = 2.8, yend = 2.8),
             arrow = arrow(length = unit(0.1, "cm"), ends = "both"), linewidth = 0.5) +
 annotate("text", x = 6.5, y = 3.1, label = "d ≥ t", size = 4, color = "darkgreen") +
 # Good label
 annotate("text", x = 6.5, y = 0.8, label = "ACCEPTABLE", color = "darkgreen", fontface = "bold", size = 4) +
 geom_point(aes(x = 6.5, y = 1.1), shape = 10, size = 8, color = "darkgreen", stroke = 2) +

 coord_fixed(ratio = 1, xlim = c(0, 8.5), ylim = c(0.3, 3.5)) +
 labs(title = "Minimum Hole Diameter Rule: d ≥ t") +
 theme_stamping
```

#### Hole Spacing

Holes should be spaced adequately apart to prevent distortion and ensure die strength:

* **Hole to hole:** Minimum spacing = 2× material thickness
* **Hole to edge:** Minimum distance = 1× material thickness (preferably 1.5×)

```{r hole-spacing, echo=FALSE, fig.width=9, fig.height=5, fig.align='center', fig.cap="Recommended hole spacing guidelines."}
# Helper function for circles
circle_df <- function(cx, cy, r, id, n = 50) {
 t <- seq(0, 2*pi, length.out = n)
 data.frame(x = cx + r * cos(t), y = cy + r * sin(t), id = id)
}

# Create circles
circles <- rbind(
 circle_df(2.5, 3, 0.5, 1),
 circle_df(5, 3, 0.5, 2),
 circle_df(7.5, 3, 0.5, 3)
)

ggplot() +
 # Sheet metal outline
 geom_rect(aes(xmin = 0.5, xmax = 9.5, ymin = 1, ymax = 5),
           fill = "orange", color = "black", linewidth = 1) +

 # Holes
 geom_polygon(data = circles, aes(x = x, y = y, group = id), fill = "white", color = "black") +

 # Hole-to-hole spacing (between first two holes)
 geom_segment(aes(x = 3, xend = 4.5, y = 3, yend = 3),
             arrow = arrow(length = unit(0.15, "cm"), ends = "both"),
             linewidth = 0.8, color = "blue") +
 annotate("text", x = 3.75, y = 3.5, label = "≥ 2t", color = "blue", fontface = "bold", size = 4) +
 annotate("text", x = 3.75, y = 2.5, label = "(hole-to-hole)", color = "blue", size = 3) +

 # Hole-to-edge spacing (left hole to edge)
 geom_segment(aes(x = 0.5, xend = 2, y = 3, yend = 3),
             arrow = arrow(length = unit(0.15, "cm"), ends = "both"),
             linewidth = 0.8, color = "red") +
 annotate("text", x = 1.25, y = 3.5, label = "≥ t", color = "red", fontface = "bold", size = 4) +
 annotate("text", x = 1.25, y = 2.5, label = "(edge)", color = "red", size = 3) +

 # Hole-to-edge spacing (right hole to edge)
 geom_segment(aes(x = 8, xend = 9.5, y = 3, yend = 3),
             arrow = arrow(length = unit(0.15, "cm"), ends = "both"),
             linewidth = 0.8, color = "red") +
 annotate("text", x = 8.75, y = 3.5, label = "≥ t", color = "red", fontface = "bold", size = 4) +

 # Hole-to-edge spacing (top)
 geom_segment(aes(x = 5, xend = 5, y = 3.5, yend = 5),
             arrow = arrow(length = unit(0.15, "cm"), ends = "both"),
             linewidth = 0.8, color = "red") +
 annotate("text", x = 5.5, y = 4.25, label = "≥ t", color = "red", fontface = "bold", size = 4) +

 # Legend box
 geom_rect(aes(xmin = 0.6, xmax = 3.8, ymin = 5.3, ymax = 6.3), fill = "white", color = "black") +
 annotate("text", x = 2.2, y = 6, label = "t = material thickness", size = 3.5) +
 annotate("text", x = 2.2, y = 5.6, label = "d = hole diameter", size = 3.5) +

 coord_fixed(ratio = 1, xlim = c(0, 10), ylim = c(0.5, 6.5)) +
 labs(title = "Hole Spacing Requirements") +
 theme_stamping
```

> **❓ Quick Check:** A designer specifies 3 mm diameter holes in 2 mm thick steel sheet, with 3 mm spacing between holes. Is this design acceptable? Why or why not?

<details>
<summary>Click to reveal answer</summary>

**Partially acceptable:**

* Hole diameter (3 mm) ≥ thickness (2 mm) ✓
* Hole spacing (3 mm) should be ≥ 2 × 2 mm = 4 mm ✗

The hole spacing is too close. The designer should increase the spacing to at least 4 mm to prevent distortion and die damage.

</details>

---

### Bend Design

#### Minimum Bend Radius

The minimum inside bend radius depends on material type, thickness, and grain direction. Bending perpendicular to the grain direction allows tighter radii.

| Material | Minimum Bend Radius (× thickness) |
|:---------|:----------------------------------|
| Soft aluminum | 0 - 0.5t |
| Low carbon steel | 0.5 - 1.0t |
| Stainless steel | 0.5 - 1.0t |
| Brass (soft) | 0 - 0.5t |
| Spring steel | 2.0 - 4.0t |

```{r bend-radius-cracking, echo=FALSE, fig.width=10, fig.height=5, fig.align='center', fig.cap="Bend radius and its relationship to material cracking."}
ggplot() +
 # LEFT: Good bend (large radius)
 annotate("text", x = 2.5, y = 5.5, label = "GOOD: Adequate Bend Radius",
         fontface = "bold", size = 4, color = "darkgreen") +

 # Bent sheet - good
 geom_polygon(aes(x = c(0.5, 0.5, 2, 2.5, 2.5, 2.3, 2, 0.8),
                  y = c(1, 3, 3, 4.5, 4.8, 4.8, 3.3, 3.3)),
             fill = "orange", color = "black", linewidth = 1) +

 # Radius arc indicator
 geom_curve(aes(x = 1.5, y = 3, xend = 2.3, yend = 4), curvature = -0.5,
           linewidth = 0.5, linetype = "dashed", color = "blue") +
 annotate("text", x = 1.2, y = 3.8, label = "R ≥ t", color = "blue", size = 4, fontface = "bold") +

 # Smooth surface indication
 annotate("text", x = 3.5, y = 4.5, label = "Smooth\nouter surface",
         color = "darkgreen", size = 3, fontface = "italic") +

 # Check mark
 geom_point(aes(x = 2.5, y = 1.5), shape = 10, size = 10, color = "darkgreen", stroke = 2) +

 # RIGHT: Bad bend (small radius)
 annotate("text", x = 8, y = 5.5, label = "BAD: Radius Too Small",
         fontface = "bold", size = 4, color = "red") +

 # Bent sheet - bad (sharp corner)
 geom_polygon(aes(x = c(6, 6, 7.5, 7.5, 7.8, 7.8, 7.2, 6.3),
                  y = c(1, 3, 3, 4.8, 4.8, 4.5, 3, 3)),
             fill = "orange", color = "black", linewidth = 1) +

 # Crack lines at bend
 geom_segment(aes(x = 7.5, xend = 7.65, y = 3.3, yend = 3.5), color = "red", linewidth = 1) +
 geom_segment(aes(x = 7.52, xend = 7.7, y = 3.5, yend = 3.75), color = "red", linewidth = 1) +
 geom_segment(aes(x = 7.55, xend = 7.75, y = 3.7, yend = 4), color = "red", linewidth = 1) +
 geom_segment(aes(x = 7.58, xend = 7.78, y = 3.9, yend = 4.2), color = "red", linewidth = 1) +

 # Sharp radius indicator
 annotate("text", x = 6.5, y = 3.5, label = "R ≈ 0", color = "red", size = 4, fontface = "bold") +

 # Crack annotation
 annotate("text", x = 9, y = 3.8, label = "CRACKS\non outer\nsurface!",
         color = "red", size = 3, fontface = "bold") +
 geom_curve(aes(x = 8.5, xend = 7.8, y = 3.8, yend = 3.7),
           arrow = arrow(length = unit(0.15, "cm")), curvature = 0.3, color = "red") +

 # X mark
 geom_point(aes(x = 8, y = 1.5), shape = 4, size = 10, color = "red", stroke = 2) +

 # Dividing line
 geom_vline(xintercept = 5, linetype = "dashed", color = "gray50") +

 coord_fixed(ratio = 1, xlim = c(0, 10), ylim = c(0.5, 6)) +
 labs(title = "Effect of Bend Radius on Material Integrity") +
 theme_stamping
```

> **⚠️ Warning:** Specifying a bend radius that is too small will result in cracking on the outer surface of the bend. Always check material specifications!

#### Springback Compensation

All materials exhibit **springback**—the tendency to partially return to their original shape after bending. Designers must account for this:

* Typical springback angles: 2° to 10° depending on material
* Solutions: Over-bend the part, use bottoming dies, or specify springback allowance

```{r springback, echo=FALSE, fig.width=10, fig.height=5, fig.align='center', fig.cap="Springback phenomenon: the part opens up slightly after the bending force is removed."}
ggplot() +
 # DURING BENDING (left)
 annotate("text", x = 2.5, y = 5.5, label = "During Bending", fontface = "bold", size = 4) +

 # Punch
 geom_polygon(aes(x = c(1.5, 2.5, 3.5, 3.5, 1.5),
                  y = c(4.5, 2.5, 4.5, 5, 5)),
             fill = "steelblue", color = "black", linewidth = 1) +

 # Bent part (in die)
 geom_polygon(aes(x = c(0.5, 2.5, 4.5, 4.3, 2.5, 0.7),
                  y = c(3.8, 2, 3.8, 4.1, 2.3, 4.1)),
             fill = "orange", color = "black", linewidth = 1) +

 # Die
 geom_polygon(aes(x = c(0, 2.5, 5, 5, 0),
                  y = c(3.5, 1, 3.5, 0, 0)),
             fill = "gray60", color = "black", linewidth = 1) +

 # Angle annotation
 geom_curve(aes(x = 2.5, y = 3, xend = 3.8, yend = 3.8), curvature = -0.3,
           linewidth = 0.5, color = "blue") +
 annotate("text", x = 3.5, y = 3, label = "θ₁", color = "blue", size = 4, fontface = "bold") +

 # Force arrow
 geom_segment(aes(x = 2.5, xend = 2.5, y = 5.8, yend = 5.2),
             arrow = arrow(length = unit(0.2, "cm")), linewidth = 1.2, color = "red") +
 annotate("text", x = 2.5, y = 6, label = "F", color = "red", fontface = "bold", size = 4) +

 # Arrow to next state
 geom_segment(aes(x = 5.3, xend = 6.2, y = 3, yend = 3),
             arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
             linewidth = 1.5, color = "darkgreen") +
 annotate("text", x = 5.75, y = 3.5, label = "Release\nforce", size = 3) +

 # AFTER RELEASE (right)
 annotate("text", x = 8.5, y = 5.5, label = "After Release (Springback)", fontface = "bold", size = 4) +

 # Part with springback (more open angle)
 geom_polygon(aes(x = c(6.3, 8.5, 10.7, 10.5, 8.5, 6.5),
                  y = c(4.2, 2, 4.2, 4.5, 2.3, 4.5)),
             fill = "orange", color = "black", linewidth = 1) +

 # Original position (dashed)
 geom_segment(aes(x = 8.5, xend = 6.5, y = 2, yend = 3.8),
             linetype = "dashed", color = "gray50", linewidth = 0.8) +
 geom_segment(aes(x = 8.5, xend = 10.5, y = 2, yend = 3.8),
             linetype = "dashed", color = "gray50", linewidth = 0.8) +

 # Springback angle
 geom_curve(aes(x = 8.5, y = 3.2, xend = 10, yend = 4.1), curvature = -0.3,
           linewidth = 0.5, color = "red") +
 annotate("text", x = 9.8, y = 3.5, label = "θ₂", color = "red", size = 4, fontface = "bold") +

 # Springback annotation
 geom_curve(aes(x = 10, y = 3.8, xend = 10.4, yend = 4.1), curvature = 0.5,
           linewidth = 1, color = "red",
           arrow = arrow(length = unit(0.15, "cm"), ends = "both")) +
 annotate("text", x = 11, y = 4, label = "Δθ\n(springback)", color = "red", size = 3) +

 # Formula
 annotate("text", x = 8.5, y = 0.8, label = "θ₂ = θ₁ + Δθ", size = 4, fontface = "bold") +
 annotate("text", x = 8.5, y = 0.3, label = "(final angle > die angle)", size = 3, fontface = "italic") +

 coord_fixed(ratio = 1, xlim = c(-0.5, 12), ylim = c(-0.2, 6.2)) +
 theme_stamping
```

> **🤔 Engineering Challenge:** A bracket requires a precise 90° bend. If the material has 5° of springback, what angle should the die produce to achieve the final 90° angle?

<details>
<summary>Click to reveal solution</summary>

The die should over-bend the material by the springback amount:

$$\text{Die angle} = 90° + 5° = 95°$$

The die should bend the part to **95°** so that after springback, the final angle is 90°.

</details>

#### Bend Relief

When a bend line intersects or comes close to an edge, **bend relief** cuts or notches are required to prevent tearing.

```{r bend-relief, echo=FALSE, fig.width=10, fig.height=5, fig.align='center', fig.cap="Bend relief prevents tearing at the intersection of bends and edges."}
ggplot() +
 # WITHOUT BEND RELIEF (left) - Bad
 annotate("text", x = 2, y = 5.5, label = "WITHOUT Bend Relief", fontface = "bold", size = 4, color = "red") +

 # Flat pattern
 geom_polygon(aes(x = c(0.5, 0.5, 2.5, 3.5, 3.5, 2.5, 2.5, 0.5),
                  y = c(1, 4, 4, 3, 1, 1, 3, 3)),
             fill = "orange", color = "black", linewidth = 1) +

 # Bend line
 geom_segment(aes(x = 0.5, xend = 3.5, y = 3, yend = 3),
             linetype = "dashed", color = "blue", linewidth = 1) +
 annotate("text", x = 0.2, y = 3, label = "Bend\nLine", size = 2.5, color = "blue") +

 # Tear indication
 geom_segment(aes(x = 2.5, xend = 2.7, y = 3, yend = 3.3), color = "red", linewidth = 1.5) +
 geom_segment(aes(x = 2.5, xend = 2.3, y = 3, yend = 2.7), color = "red", linewidth = 1.5) +
 annotate("text", x = 3.2, y = 2.5, label = "TEAR!", color = "red", fontface = "bold", size = 4) +

 # X mark
 geom_point(aes(x = 2, y = 0.5), shape = 4, size = 8, color = "red", stroke = 2) +

 # WITH BEND RELIEF (right) - Good
 annotate("text", x = 7.5, y = 5.5, label = "WITH Bend Relief", fontface = "bold", size = 4, color = "darkgreen") +

 # Flat pattern with relief
 geom_polygon(aes(x = c(5.5, 5.5, 7.5, 7.5, 7.8, 8.5, 8.5, 7.8, 7.5, 7.5, 5.5),
                  y = c(1, 4, 4, 3.3, 3.3, 3.3, 2.7, 2.7, 2.7, 1, 1)),
             fill = "orange", color = "black", linewidth = 1) +

 # Bend line
 geom_segment(aes(x = 5.5, xend = 7.5, y = 3, yend = 3),
             linetype = "dashed", color = "blue", linewidth = 1) +

 # Relief dimensions
 geom_segment(aes(x = 7.5, xend = 7.8, y = 3.5, yend = 3.5),
             arrow = arrow(length = unit(0.1, "cm"), ends = "both"), linewidth = 0.5) +
 annotate("text", x = 7.65, y = 3.7, label = "≥ t", size = 3) +

 geom_segment(aes(x = 8.7, xend = 8.7, y = 2.7, yend = 3.3),
             arrow = arrow(length = unit(0.1, "cm"), ends = "both"), linewidth = 0.5) +
 annotate("text", x = 9.2, y = 3, label = "≥ t + R", size = 3) +

 # Relief label
 geom_curve(aes(x = 9, xend = 8, y = 2.2, yend = 2.9),
           arrow = arrow(length = unit(0.15, "cm")), curvature = -0.3, color = "darkgreen") +
 annotate("text", x = 9.2, y = 2, label = "Bend\nRelief", color = "darkgreen", fontface = "bold", size = 3) +

 # Check mark
 geom_point(aes(x = 7.5, y = 0.5), shape = 10, size = 8, color = "darkgreen", stroke = 2) +

 # Dividing line
 geom_vline(xintercept = 4.5, linetype = "dashed", color = "gray50") +

 coord_fixed(ratio = 1, xlim = c(-0.5, 10), ylim = c(0, 6)) +
 labs(title = "Bend Relief Requirements") +
 theme_stamping
```

**Bend relief guidelines:**

* Width of relief ≥ material thickness
* Length of relief ≥ material thickness + bend radius

---

### Drawing Design

#### Draw Depth Limits

The maximum depth of a single draw operation is limited by the drawing ratio. Exceeding this limit causes wrinkling, tearing, or both.

**Rule of thumb:** Single draw depth should not exceed 40-50% of the blank diameter for cylindrical parts.

```{r draw-depth-limit, echo=FALSE, fig.width=10, fig.height=5, fig.align='center', fig.cap="Draw ratio limits for single-stage drawing operations."}
ggplot() +
 # ACCEPTABLE (left)
 annotate("text", x = 2.5, y = 5.8, label = "ACCEPTABLE", fontface = "bold", size = 4, color = "darkgreen") +
 annotate("text", x = 2.5, y = 5.4, label = "Draw Ratio ≤ 2.0", size = 3.5) +

 # Cup shape - good
 geom_polygon(aes(x = c(1, 1, 1.3, 1.3, 3.7, 3.7, 4, 4, 3.7, 1.3, 1),
                  y = c(3.5, 4, 4, 1.5, 1.5, 4, 4, 3.5, 3.5, 3.5, 3.5)),
             fill = "orange", color = "black", linewidth = 1) +

 # Depth dimension
 geom_segment(aes(x = 0.5, xend = 0.5, y = 1.5, yend = 4),
             arrow = arrow(length = unit(0.1, "cm"), ends = "both"), linewidth = 0.5) +
 annotate("text", x = 0.2, y = 2.75, label = "h", size = 4, fontface = "bold") +

 # Diameter dimension
 geom_segment(aes(x = 1.3, xend = 3.7, y = 1.2, yend = 1.2),
             arrow = arrow(length = unit(0.1, "cm"), ends = "both"), linewidth = 0.5) +
 annotate("text", x = 2.5, y = 0.9, label = expression(D[p]), size = 4, fontface = "bold") +

 # Check mark
 geom_point(aes(x = 2.5, y = 0.3), shape = 10, size = 8, color = "darkgreen", stroke = 2) +

 # TOO DEEP (right)
 annotate("text", x = 8, y = 5.8, label = "TOO DEEP", fontface = "bold", size = 4, color = "red") +
 annotate("text", x = 8, y = 5.4, label = "Draw Ratio > 2.0", size = 3.5) +

 # Cup shape - bad (with defects)
 geom_polygon(aes(x = c(6.5, 6.5, 6.8, 6.8, 9.2, 9.2, 9.5, 9.5, 9.2, 6.8, 6.5),
                  y = c(4, 4.5, 4.5, 0.5, 0.5, 4.5, 4.5, 4, 4, 4, 4)),
             fill = "orange", color = "black", linewidth = 1) +

 # Tearing indication at bottom corner
 geom_segment(aes(x = 6.8, xend = 6.5, y = 0.7, yend = 0.4), color = "red", linewidth = 1.2) +
 geom_segment(aes(x = 6.8, xend = 6.6, y = 0.9, yend = 0.5), color = "red", linewidth = 1.2) +
 annotate("text", x = 5.8, y = 0.6, label = "Tear!", color = "red", size = 3, fontface = "bold") +

 # Wrinkling indication at top
 geom_curve(aes(x = 6.6, xend = 6.7, y = 3.8, yend = 3.5), curvature = 0.5, color = "red", linewidth = 0.8) +
 geom_curve(aes(x = 6.65, xend = 6.75, y = 3.5, yend = 3.2), curvature = -0.5, color = "red", linewidth = 0.8) +
 annotate("text", x = 5.7, y = 3.5, label = "Wrinkle!", color = "red", size = 3, fontface = "bold") +

 # Depth dimension
 geom_segment(aes(x = 6, xend = 6, y = 0.5, yend = 4.5),
             arrow = arrow(length = unit(0.1, "cm"), ends = "both"), linewidth = 0.5) +
 annotate("text", x = 5.7, y = 2.5, label = "h", size = 4, fontface = "bold") +

 # X mark
 geom_point(aes(x = 8, y = 0.3), shape = 4, size = 8, color = "red", stroke = 2) +

 # Dividing line
 geom_vline(xintercept = 5.2, linetype = "dashed", color = "gray50") +

 coord_fixed(ratio = 1, xlim = c(-0.5, 10.5), ylim = c(-0.2, 6.2)) +
 labs(title = "Drawing Depth Limitations") +
 theme_stamping
```

#### Corner Radii in Drawn Parts

Sharp corners in drawn parts concentrate stress and cause tearing. Minimum corner radii:

* **Bottom corner radius:** ≥ 4× material thickness
* **Top (die entry) radius:** ≥ 4× material thickness
* **Wall corner radius (rectangular parts):** ≥ 3× material thickness

```{r draw-corner-radii, echo=FALSE, fig.width=8, fig.height=6, fig.align='center', fig.cap="Critical corner radii in drawn parts."}
ggplot() +
 # Cross-section of drawn cup
 # Die
 geom_polygon(aes(x = c(0, 0, 2.5, 2.5, 2.8, 2.8, 6.2, 6.2, 6.5, 6.5, 9, 9),
                  y = c(0, 4.5, 4.5, 4.2, 4.2, 1, 1, 4.2, 4.2, 4.5, 4.5, 0)),
             fill = "gray60", color = "black", linewidth = 1) +

 # Drawn cup
 geom_polygon(aes(x = c(2.5, 2.5, 3, 3, 6, 6, 6.5, 6.5, 6.3, 6.3, 2.7, 2.7),
                  y = c(5.5, 4.5, 4.2, 1.3, 1.3, 4.2, 4.5, 5.5, 5.5, 4.5, 4.5, 5.5)),
             fill = "orange", color = "black", linewidth = 1) +

 # Top radius annotation (die entry)
 geom_curve(aes(x = 2.5, xend = 2.8, y = 4.5, yend = 4.2), curvature = 0.4,
           linewidth = 1.5, color = "blue") +
 annotate("text", x = 1.5, y = 4.7, label = expression(R[die] >= 4*t),
         color = "blue", size = 3.5, fontface = "bold") +
 geom_segment(aes(x = 1.8, xend = 2.4, y = 4.5, yend = 4.4),
             arrow = arrow(length = unit(0.15, "cm")), color = "blue") +

 # Bottom corner radius annotation
 geom_curve(aes(x = 3, xend = 3.3, y = 1.3, yend = 1), curvature = -0.5,
           linewidth = 1.5, color = "red") +
 annotate("text", x = 4, y = 0.5, label = expression(R[bottom] >= 4*t),
         color = "red", size = 3.5, fontface = "bold") +
 geom_segment(aes(x = 3.8, xend = 3.2, y = 0.7, yend = 1.1),
             arrow = arrow(length = unit(0.15, "cm")), color = "red") +

 # Punch
 geom_polygon(aes(x = c(3.2, 3.2, 3.5, 3.5, 5.5, 5.5, 5.8, 5.8),
                  y = c(7, 1.5, 1.2, 1, 1, 1.2, 1.5, 7)),
             fill = "steelblue", color = "black", linewidth = 1) +
 annotate("text", x = 4.5, y = 6, label = "Punch", color = "white", fontface = "bold", size = 4) +

 # Punch radius annotation
 geom_curve(aes(x = 3.2, xend = 3.5, y = 1.5, yend = 1.2), curvature = 0.4,
           linewidth = 1.5, color = "purple") +
 annotate("text", x = 2, y = 1.8, label = expression(R[punch] >= 4*t),
         color = "purple", size = 3.5, fontface = "bold") +
 geom_segment(aes(x = 2.3, xend = 3.1, y = 1.6, yend = 1.4),
             arrow = arrow(length = unit(0.15, "cm")), color = "purple") +

 # Material thickness annotation
 geom_segment(aes(x = 6.5, xend = 6.3, y = 3, yend = 3),
             arrow = arrow(length = unit(0.1, "cm"), ends = "both"), linewidth = 0.5) +
 annotate("text", x = 6.8, y = 3, label = "t", size = 4, fontface = "bold") +

 # Legend
 geom_rect(aes(xmin = 6.8, xmax = 9.2, ymin = 5.5, ymax = 7), fill = "white", color = "black") +
 annotate("text", x = 8, y = 6.7, label = "t = thickness", size = 3) +
 annotate("text", x = 8, y = 6.3, label = expression(R[die]*" = die entry"), size = 3) +
 annotate("text", x = 8, y = 5.9, label = expression(R[punch]*" = punch"), size = 3) +

 coord_fixed(ratio = 1, xlim = c(-0.5, 9.5), ylim = c(-0.5, 7.5)) +
 labs(title = "Critical Radii in Deep Drawing") +
 theme_stamping
```

> **❓ Design Review:** Examine the part specifications below. A rectangular box is being drawn from 1.5 mm thick aluminum with the following specifications: depth = 50 mm, bottom corner radius = 3 mm, wall corner radius = 6 mm. Identify any design violations.

<details>
<summary>Click to reveal analysis</summary>

**Analysis:**

* Bottom corner radius: 3 mm should be ≥ 4 × 1.5 mm = 6 mm ✗ **TOO SMALL**
* Wall corner radius: 6 mm should be ≥ 3 × 1.5 mm = 4.5 mm ✓ OK

**Recommendation:** Increase the bottom corner radius to at least 6 mm to prevent tearing.

</details>

---

### General Design Best Practices

#### Uniform Wall Thickness

Maintain uniform wall thickness throughout the part when possible. Significant thickness variations cause uneven material flow and stress concentrations.

#### Avoid Sharp Corners

Replace all sharp corners with radii or chamfers:

* Reduces stress concentration
* Extends die life
* Improves material flow

#### Design for Nesting

Consider how blanks will be nested on the sheet to minimize material waste.

```{r blank-nesting, echo=FALSE, fig.width=10, fig.height=5, fig.align='center', fig.cap="Good nesting design minimizes scrap and material cost."}
# Helper function for circles
circle_df2 <- function(cx, cy, r, id, n = 50) {
 t <- seq(0, 2*pi, length.out = n)
 data.frame(x = cx + r * cos(t), y = cy + r * sin(t), id = id)
}

# Poor nesting - circles in grid
poor_circles <- rbind(
 circle_df2(1, 1, 0.4, 1), circle_df2(2, 1, 0.4, 2), circle_df2(3, 1, 0.4, 3),
 circle_df2(1, 2, 0.4, 4), circle_df2(2, 2, 0.4, 5), circle_df2(3, 2, 0.4, 6),
 circle_df2(1, 3, 0.4, 7), circle_df2(2, 3, 0.4, 8), circle_df2(3, 3, 0.4, 9)
)

# Good nesting - staggered circles
good_circles <- rbind(
 circle_df2(6, 1, 0.4, 1), circle_df2(6.85, 1, 0.4, 2), circle_df2(7.7, 1, 0.4, 3), circle_df2(8.55, 1, 0.4, 4),
 circle_df2(6.425, 1.75, 0.4, 5), circle_df2(7.275, 1.75, 0.4, 6), circle_df2(8.125, 1.75, 0.4, 7),
 circle_df2(6, 2.5, 0.4, 8), circle_df2(6.85, 2.5, 0.4, 9), circle_df2(7.7, 2.5, 0.4, 10), circle_df2(8.55, 2.5, 0.4, 11),
 circle_df2(6.425, 3.25, 0.4, 12), circle_df2(7.275, 3.25, 0.4, 13), circle_df2(8.125, 3.25, 0.4, 14)
)

ggplot() +
 # POOR NESTING (left)
 annotate("text", x = 2, y = 4.3, label = "POOR NESTING", fontface = "bold", size = 4, color = "red") +
 geom_rect(aes(xmin = 0.3, xmax = 3.7, ymin = 0.3, ymax = 3.7),
           fill = "lightyellow", color = "black", linewidth = 1) +
 geom_polygon(data = poor_circles, aes(x = x, y = y, group = id),
             fill = "orange", color = "black") +
 annotate("text", x = 2, y = -0.2, label = "9 parts\n~64% utilization", size = 3) +

 # GOOD NESTING (right)
 annotate("text", x = 7.3, y = 4.3, label = "GOOD NESTING", fontface = "bold", size = 4, color = "darkgreen") +
 geom_rect(aes(xmin = 5.5, xmax = 9.1, ymin = 0.3, ymax = 3.7),
           fill = "lightyellow", color = "black", linewidth = 1) +
 geom_polygon(data = good_circles, aes(x = x, y = y, group = id),
             fill = "orange", color = "black") +
 annotate("text", x = 7.3, y = -0.2, label = "14 parts\n~78% utilization", size = 3) +

 # Arrow between
 geom_segment(aes(x = 4.2, xend = 5, y = 2, yend = 2),
             arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
             linewidth = 1, color = "darkgreen") +

 coord_fixed(ratio = 1, xlim = c(-0.5, 10), ylim = c(-0.8, 4.8)) +
 labs(title = "Blank Nesting Comparison") +
 theme_stamping
```

> **💰 Cost Impact:** Material costs can represent 40-60% of total stamped part cost. A well-nested design that improves material utilization from 70% to 85% can significantly reduce part cost!

#### Symmetry

Design symmetrical parts whenever possible:

* Simplifies die design
* Improves material flow during forming
* Reduces the chance of assembly errors

---

## 🛠️ Common Stamping Defects and Prevention {-}

Understanding potential defects helps designers avoid them through proper design.

| Defect | Description | Causes | Prevention |
|:-------|:------------|:-------|:-----------|
| **Wrinkling** | Wavy, folded material, typically in flange areas | Excessive compressive stress; insufficient blank holder force | Increase blank holder pressure; optimize blank size; add draw beads |
| **Tearing** | Cracks or splits in the material | Excessive tensile stress; sharp corners; insufficient material | Increase radii; use more ductile material; add draw stages |
| **Springback** | Part doesn't hold intended shape | Elastic recovery of material | Over-bend; use bottoming dies; adjust tooling |
| **Burrs** | Sharp, rough edges on cut surfaces | Excessive die clearance; worn tooling | Maintain proper clearance (5-10% of thickness); replace worn tools |
| **Galling** | Material pickup on die surfaces | Metal-to-metal adhesion | Use lubrication; apply die coatings; select appropriate die material |
| **Orange Peel** | Rough, textured surface | Excessive grain growth; over-forming | Use fine-grain material; reduce strain |

```{r stamping-defects, echo=FALSE, fig.width=12, fig.height=4, fig.align='center', fig.cap="Common stamping defects: wrinkling in flanges, tearing at corners, and burrs on cut edges."}
ggplot() +
 # WRINKLING
 annotate("text", x = 2, y = 3.8, label = "WRINKLING", fontface = "bold", size = 4, color = "red") +

 # Cup with wrinkled flange
 geom_polygon(aes(x = c(0.5, 0.5, 1, 1, 3, 3, 3.5, 3.5),
                  y = c(3, 3.3, 3.3, 1, 1, 3.3, 3.3, 3)),
             fill = "orange", color = "black", linewidth = 1) +
 # Wrinkle waves
 geom_curve(aes(x = 0.5, xend = 0.8, y = 3.15, yend = 3.3), curvature = -0.5, color = "darkorange", linewidth = 1.5) +
 geom_curve(aes(x = 0.8, xend = 1.1, y = 3.3, yend = 3.15), curvature = -0.5, color = "darkorange", linewidth = 1.5) +
 geom_curve(aes(x = 3.5, xend = 3.2, y = 3.15, yend = 3.3), curvature = 0.5, color = "darkorange", linewidth = 1.5) +
 geom_curve(aes(x = 3.2, xend = 2.9, y = 3.3, yend = 3.15), curvature = 0.5, color = "darkorange", linewidth = 1.5) +

 annotate("text", x = 2, y = 0.4, label = "Wavy flange", size = 3, fontface = "italic") +

 # TEARING
 annotate("text", x = 6.5, y = 3.8, label = "TEARING", fontface = "bold", size = 4, color = "red") +

 # Cup with tear
 geom_polygon(aes(x = c(5, 5, 5.5, 5.5, 7.5, 7.5, 8, 8),
                  y = c(3, 3.3, 3.3, 1, 1, 3.3, 3.3, 3)),
             fill = "orange", color = "black", linewidth = 1) +
 # Tear crack
 geom_segment(aes(x = 5.5, xend = 5.3, y = 1.2, yend = 0.8), color = "red", linewidth = 2) +
 geom_segment(aes(x = 5.3, xend = 5.1, y = 0.8, yend = 0.5), color = "red", linewidth = 2) +
 geom_segment(aes(x = 5.5, xend = 5.7, y = 1.2, yend = 1.5), color = "red", linewidth = 1.5) +

 annotate("text", x = 6.5, y = 0.4, label = "Crack at corner", size = 3, fontface = "italic") +

 # BURRS
 annotate("text", x = 11, y = 3.8, label = "BURRS", fontface = "bold", size = 4, color = "red") +

 # Sheet with punched hole and burrs
 geom_rect(aes(xmin = 9.5, xmax = 12.5, ymin = 1.5, ymax = 2.5),
           fill = "orange", color = "black", linewidth = 1) +
 # Hole
 geom_rect(aes(xmin = 10.5, xmax = 11.5, ymin = 1.5, ymax = 2.5),
           fill = "white", color = "black") +
 # Burrs (jagged edges)
 geom_segment(aes(x = 10.5, xend = 10.45, y = 1.5, yend = 1.35), color = "darkorange", linewidth = 1) +
 geom_segment(aes(x = 10.7, xend = 10.68, y = 1.5, yend = 1.3), color = "darkorange", linewidth = 1) +
 geom_segment(aes(x = 10.9, xend = 10.88, y = 1.5, yend = 1.33), color = "darkorange", linewidth = 1) +
 geom_segment(aes(x = 11.1, xend = 11.12, y = 1.5, yend = 1.28), color = "darkorange", linewidth = 1) +
 geom_segment(aes(x = 11.3, xend = 11.32, y = 1.5, yend = 1.35), color = "darkorange", linewidth = 1) +
 geom_segment(aes(x = 11.5, xend = 11.55, y = 1.5, yend = 1.32), color = "darkorange", linewidth = 1) +

 annotate("text", x = 11, y = 1, label = "Rough edge", size = 3, fontface = "italic") +
 geom_curve(aes(x = 11, xend = 11, y = 1.15, yend = 1.4),
           arrow = arrow(length = unit(0.1, "cm")), curvature = 0, color = "red") +

 # Dividing lines
 geom_vline(xintercept = c(4.25, 8.75), linetype = "dashed", color = "gray50") +

 coord_fixed(ratio = 1, xlim = c(0, 13), ylim = c(0, 4.2)) +
 labs(title = "Common Stamping Defects") +
 theme_stamping
```

> **🔍 Defect Diagnosis:** A production run of drawn cups is showing wrinkling in the flange area. List three possible corrective actions the manufacturing team could take.

<details>
<summary>Click to reveal solutions</summary>

**Possible corrective actions for flange wrinkling:**

1. **Increase blank holder force** - More pressure prevents the material from buckling
2. **Reduce blank diameter** - Less excess material to wrinkle
3. **Add draw beads** - Control material flow and increase restraining force
4. **Use better lubrication** - Reduce friction and improve material flow
5. **Anneal the material** - Softer material flows more easily

</details>

---

## 📋 Design Checklist for Stamping {-}

Before finalizing a stamped part design, verify the following:

### Holes and Cutouts
- [ ] Hole diameter ≥ material thickness
- [ ] Hole-to-hole spacing ≥ 2× material thickness
- [ ] Hole-to-edge distance ≥ 1× material thickness
- [ ] No holes in bend zones (or proper relief provided)

### Bends
- [ ] Bend radius ≥ minimum for material (check data sheet)
- [ ] Bend relief provided where bend meets edge
- [ ] Springback compensation accounted for
- [ ] Grain direction considered for tight bends

### Draws
- [ ] Draw ratio within single-stage limits
- [ ] Bottom radius ≥ 4× material thickness
- [ ] Wall radii ≥ 3× material thickness
- [ ] No sharp transitions between surfaces

### General
- [ ] All corners filleted or chamfered
- [ ] Uniform wall thickness (or gradual transitions)
- [ ] Part can be nested efficiently on sheet
- [ ] Tolerances achievable by stamping process
- [ ] Part can be ejected from die without interference

---

## 🎯 Case Study: Automotive Bracket Design {-}

Let's apply our knowledge to a real-world example.

**Design Requirements:**

* Material: Low carbon steel, 2.0 mm thick
* Overall dimensions: 80 mm × 50 mm × 25 mm height
* Two mounting holes: 8 mm diameter
* Two flanges bent at 90°
* Must be produced at 50,000 units/year

### Initial Design Evaluation

```{r bracket-case-study, echo=FALSE, fig.width=10, fig.height=6, fig.align='center', fig.cap="Initial bracket design with potential issues highlighted."}
ggplot() +
 # Main bracket body (flat view representation)
 # Base
 geom_rect(aes(xmin = 1, xmax = 9, ymin = 2, ymax = 4),
           fill = "orange", color = "black", linewidth = 1) +

 # Left flange (shown as extension)
 geom_rect(aes(xmin = 1, xmax = 2.5, ymin = 4, ymax = 6),
           fill = "lightyellow", color = "black", linewidth = 1, linetype = "dashed") +
 annotate("text", x = 1.75, y = 5, label = "Flange\n(bent 90°)", size = 3) +

 # Right flange
 geom_rect(aes(xmin = 7.5, xmax = 9, ymin = 4, ymax = 6),
           fill = "lightyellow", color = "black", linewidth = 1, linetype = "dashed") +
 annotate("text", x = 8.25, y = 5, label = "Flange\n(bent 90°)", size = 3) +

 # Bend lines
 geom_segment(aes(x = 1, xend = 2.5, y = 4, yend = 4),
             color = "blue", linewidth = 1, linetype = "dashed") +
 geom_segment(aes(x = 7.5, xend = 9, y = 4, yend = 4),
             color = "blue", linewidth = 1, linetype = "dashed") +

 # Holes (too close to bend line - problem!)
 geom_point(aes(x = 2, y = 3), size = 8, color = "white") +
 geom_point(aes(x = 2, y = 3), size = 8, shape = 1, color = "black", stroke = 1) +
 geom_point(aes(x = 8, y = 3), size = 8, color = "white") +
 geom_point(aes(x = 8, y = 3), size = 8, shape = 1, color = "black", stroke = 1) +

 # Problem indicators
 # Problem 1: Hole too close to bend
 geom_segment(aes(x = 2, xend = 2, y = 3.5, yend = 4),
             arrow = arrow(length = unit(0.1, "cm"), ends = "both"),
             color = "red", linewidth = 1) +
 annotate("text", x = 2.8, y = 3.75, label = "Too close!\n(< 4mm)",
         color = "red", size = 3, fontface = "bold") +

 # Problem 2: No bend relief
 geom_rect(aes(xmin = 0.8, xmax = 1.2, ymin = 3.8, ymax = 4.2),
           fill = NA, color = "red", linewidth = 1.5, linetype = "solid") +
 geom_segment(aes(x = 0.5, xend = 0.9, y = 4.5, yend = 4.1),
             arrow = arrow(length = unit(0.15, "cm")), color = "red") +
 annotate("text", x = 0.3, y = 4.8, label = "No bend\nrelief!",
         color = "red", size = 3, fontface = "bold") +

 # Problem 3: Sharp corner
 geom_rect(aes(xmin = 8.8, xmax = 9.2, ymin = 1.8, ymax = 2.2),
           fill = NA, color = "red", linewidth = 1.5) +
 geom_segment(aes(x = 9.5, xend = 9.1, y = 1.5, yend = 1.9),
             arrow = arrow(length = unit(0.15, "cm")), color = "red") +
 annotate("text", x = 9.7, y = 1.2, label = "Sharp\ncorner!",
         color = "red", size = 3, fontface = "bold") +

 # Dimensions
 geom_segment(aes(x = 1, xend = 9, y = 1.5, yend = 1.5),
             arrow = arrow(length = unit(0.1, "cm"), ends = "both"), linewidth = 0.5) +
 annotate("text", x = 5, y = 1.2, label = "80 mm", size = 3) +

 geom_segment(aes(x = 0.5, xend = 0.5, y = 2, yend = 4),
             arrow = arrow(length = unit(0.1, "cm"), ends = "both"), linewidth = 0.5) +
 annotate("text", x = 0.2, y = 3, label = "50 mm", size = 3, angle = 90) +

 # Material spec
 annotate("text", x = 5, y = 6.5, label = "Material: Low Carbon Steel, t = 2.0 mm",
         size = 4, fontface = "bold") +

 # Legend
 geom_rect(aes(xmin = 3.5, xmax = 6.5, ymin = 0.2, ymax = 0.8), fill = "white", color = "black") +
 annotate("text", x = 5, y = 0.5, label = "Red = Design Issues to Address",
         color = "red", size = 3) +

 coord_fixed(ratio = 1, xlim = c(-0.5, 10.5), ylim = c(0, 7)) +
 labs(title = "Automotive Bracket - Initial Design Review") +
 theme_stamping
```

> **🤔 Your Turn:** Based on the guidelines covered, identify at least three potential issues with this initial design and suggest improvements.

<details>
<summary>Click to reveal analysis</summary>

**Potential Issues and Solutions:**

1. **Hole too close to bend line**
   - Issue: Hole may distort during bending
   - Solution: Move hole at least 2× thickness (4 mm) + bend radius away from bend line

2. **Sharp inside corners on bent flanges**
   - Issue: Stress concentration and potential cracking
   - Solution: Add fillet radius of at least 1× thickness (2 mm)

3. **No bend relief at flange edges**
   - Issue: Tearing at the bend-to-edge intersection
   - Solution: Add bend relief notches

4. **Hole spacing check needed**
   - Verify holes are at least 4 mm apart (2× thickness)

</details>

---

## 📚 "Summary: Key Design Rules for Stamping" {-}

| Design Feature | Minimum Requirement |
|:---------------|:--------------------|
| Hole diameter | ≥ Material thickness |
| Hole-to-hole spacing | ≥ 2× Material thickness |
| Hole-to-edge distance | ≥ 1× Material thickness |
| Inside bend radius | 0.5-4× Material thickness (varies by material) |
| Draw bottom radius | ≥ 4× Material thickness |
| Draw wall radius | ≥ 3× Material thickness |
| Maximum single draw ratio | ~2.0 |
| Bend relief width | ≥ Material thickness |

---

## ✍️ Practice Problems {-}

Test your understanding with these problems:

**Problem 1:** A part requires a blanked shape with 50 holes, 4 mm in diameter, arranged in a grid pattern. The material is 1.5 mm thick mild steel. What is the minimum allowable spacing between hole centers?

<details>
<summary>Click for solution</summary>

Minimum hole-to-hole spacing = 2 × thickness = 2 × 1.5 mm = **3.0 mm**

Minimum center-to-center distance = hole spacing + hole diameter = 3.0 mm + 4.0 mm = **7.0 mm**

</details>

---

**Problem 2:** An engineer wants to draw a cylindrical cup with a 40 mm inner diameter and 60 mm depth from a single blank. Is this feasible in a single drawing operation? If not, how would you approach this?

<details>
<summary>Click for solution</summary>

For a single draw, depth typically should not exceed 40-50% of blank diameter.

Assuming 50% maximum: Maximum depth = 0.5 × blank diameter

The blank diameter for a 40 mm cup (using drawing ratio of 2.0) = 40 × 2 = 80 mm

Maximum single-draw depth ≈ 0.5 × 80 = **40 mm**

Since the required depth (60 mm) > maximum single-draw depth (40 mm), **this cannot be done in a single operation.**

**Solution:** Use a redrawing operation (two or more drawing stages) to achieve the full 60 mm depth.

</details>

---

**Problem 3:** A bracket made from 2.5 mm aluminum sheet requires a 90° bend. The material specifications indicate 6° of springback. What should the die angle be set to?

<details>
<summary>Click for solution</summary>

Die angle = Desired angle + Springback = 90° + 6° = **96°**

The die should over-bend the part to 96° so that after springback, the final angle is 90°.

</details>

---

## 🔗 Further Reading {-}

[Stamping Design Guidelines - Xometry](https://www.xometry.com/resources/sheet/metal-stamping-design-guidelines/)

[Metal Stamping Basics - Thomasnet](https://www.thomasnet.com/articles/custom-manufacturing-fabricating/metal-stamping/)

[Progressive Die Stamping - Manor Tool](https://www.manortool.com/progressive-die-stamping/)

[Sheet Metal Forming - CustomPart.Net](https://www.custompartnet.com/wu/sheet-metal-forming)

---

## 🎬 Supplemental Videos {-}

### How Stamping Dies Work

```{=html}
<div style="position: relative; padding-bottom: 56.25%; height: 0; overflow: hidden; max-width: 100%;">
  <iframe
    style="position: absolute; top: 0; left: 0; width: 100%; height: 100%;"
    src="https://www.youtube.com/embed/ipMnAe0xNwM"
    title="How Stamping Dies Work"
    frameborder="0"
    allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share"
    referrerpolicy="strict-origin-when-cross-origin"
    allowfullscreen>
  </iframe>
</div>
```

### Deep Drawing Process Explained

```{=html}
<div style="position: relative; padding-bottom: 56.25%; height: 0; overflow: hidden; max-width: 100%;">
  <iframe
    style="position: absolute; top: 0; left: 0; width: 100%; height: 100%;"
    src="https://www.youtube.com/embed/2KZRQSI8GkQ"
    title="Deep Drawing Process"
    frameborder="0"
    allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share"
    referrerpolicy="strict-origin-when-cross-origin"
    allowfullscreen>
  </iframe>
</div>
```

### Stamping by 'Smarter Every Day'

```{=html}
<div style="position: relative; padding-bottom: 56.25%; height: 0; overflow: hidden; max-width: 100%;">
  <iframe
    style="position: absolute; top: 0; left: 0; width: 100%; height: 100%;"
    src="https://www.youtube.com/embed/GDzBE6vz5r0"
    title="Stamping by Smarter Every Day"
    frameborder="0"
    allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share"
    referrerpolicy="strict-origin-when-cross-origin"
    allowfullscreen>
  </iframe>
</div>
```

---

## 📖 References {-}

1. Groover, M. P. (2020). *Fundamentals of Modern Manufacturing: Materials, Processes, and Systems* (7th ed.). Wiley.

2. Boljanovic, V. (2004). *Sheet Metal Forming Processes and Die Design*. Industrial Press.

3. Hosford, W. F., & Caddell, R. M. (2011). *Metal Forming: Mechanics and Metallurgy* (4th ed.). Cambridge University Press.

4. Suchy, I. (2006). *Handbook of Die Design* (2nd ed.). McGraw-Hill.
