---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Wire, Cable, and Circuit Protection

```{r setup-06-04, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter, you will be able to:

1. **Select** an appropriately sized conductor (AWG) for a given continuous load current, using the ampacity-table concept represented by CEC Table 2 / NEC Table 310.16.
2. **Apply** ambient temperature correction and conductor bundling/grouping adjustment factors to derate a base conductor ampacity, and **resize** a conductor when a derated circuit no longer meets the required ampacity.
3. **Calculate** voltage drop for DC/single-phase and three-phase circuits, and **evaluate** a design against the 3% rule of thumb, including judging when a tighter or looser limit is actually appropriate.
4. **Select** an appropriate cable construction (TECK90, AC90, tray cable, VFD cable) for a given installation environment, using a documented decision framework.
5. **Evaluate** the trade-offs between fuses and circuit breakers for a given overcurrent protection application, including let-through energy, resettability, cost, and selectivity.
6. **Explain** the concept of selective coordination between upstream and downstream overcurrent devices, and **evaluate** whether a given device pair is coordinated at a specified fault current.
7. **Integrate** conductor sizing, derating, voltage drop, cable selection, and overcurrent protection decisions into a single, defensible electrical design package for a plant circuit.

## Introduction

A Tier 1 automotive supplier in Cambridge, Ontario is expanding its stamping and sub-assembly plant: a new pretreatment and e-coat line, a dozen new VFD-driven conveyor and coolant-pump drives on the machining floor, and a new distribution panel to feed all of it. Every one of those additions eventually lands on the desk of an electrical designer who has to answer a deceptively simple question, over and over, for every single circuit: *what wire, and what protects it?*

That question breaks into several smaller ones, and getting any of them wrong has real consequences. Pick a conductor too small for the actual load and it runs hot, ages its insulation faster than the plant expects, and can become a fire risk long before anyone notices a problem — pick it appropriately and it runs cool and lasts the life of the installation. Forget that a cable tray full of other circuits, or an unusually warm mezzanine near a process oven, changes how much current a conductor can actually carry, and a properly-sized-on-paper conductor overheats anyway. Run a long, small-gauge 24 VDC cable out to a sensor at the far end of a conveyor line without checking voltage drop, and the sensor either misreads or stops working entirely — even though every splice, connector, and power supply along the way is functioning exactly as designed. Choose the wrong cable construction for the environment — indoor-only cable in a washdown zone, standard power cable on a VFD's output — and the failure shows up months later as insulation breakdown or bearing damage. Choose the wrong overcurrent device — or the right device type sized without checking how it interacts with everything upstream and downstream of it — and a single motor fault can black out an entire production line instead of just the one cell that actually failed.

This chapter works through all of these decisions as six design tasks: selecting a base conductor ampacity, derating that conductor for real installed conditions, calculating voltage drop, selecting the right cable construction for the environment, choosing between a fuse and a breaker, and checking whether a pair of overcurrent devices is selectively coordinated. The worked examples follow the automotive plant's pretreatment line and VFD fleet, with a stop along the way in a food and beverage plant's washdown zone, where cable jacketing and chemical resistance become the deciding factor rather than an afterthought. As in every chapter in this part, **Canadian code and standards are cited first** — the Canadian Electrical Code (CEC) as adopted and amended by the Ontario Electrical Safety Code (OESC), and the relevant CSA standards — with the NEC/UL/NEMA equivalents noted afterward, including a genuinely consequential difference between Canadian and US practice that shows up directly in this chapter's voltage-drop math: Canada's 600 V motor voltage class versus the US 480 V convention.

## Design Task 1: Conductor Ampacity Selection

### Scenario

The pretreatment line's new immersion heater bank draws a steady 58 A, three-phase, from the plant's 600 V distribution system, running continuously through most of a shift to hold the phosphate bath at temperature. The electrical designer needs to select the smallest copper conductor that can legally and safely carry that load, before the circuit even gets to the question of the actual installed environment (that comes in Design Task 2).

### Design Guidelines and Rules of Thumb

An **ampacity table** is the source of record for how much continuous current a given conductor size can carry without exceeding its insulation's rated temperature. In Canada, that source is **CEC Table 2** (Ampacities of copper conductors, rated 0-2000 V, based on a 30°C ambient, for not more than three current-carrying conductors in a raceway or cable) as adopted by the OESC; the widely-used US equivalent is **NEC Table 310.16**. Both tables are organized the same conceptual way: conductor size (AWG or kcmil) down one axis, insulation temperature rating (60°C, 75°C, 90°C are the common columns) across the other, with the ampacity for each combination read directly out of the cell — no calculation is involved in this base lookup, only correct table entry. This chapter does not reproduce either table's actual numeric entries; the worked examples below use realistic, representative ampacity values in the spirit of those tables, and any real design must go to the current edition of CEC Table 2 (or NEC Table 310.16) directly.

Two things determine which column of the table applies before a designer even looks at conductor size: the conductor's **insulation temperature rating** (marked on the cable jacket — common industrial types like RW90 or THHN/THWN-2 are 90°C-rated, but many designers conservatively use the 75°C column even for 90°C-rated conductors, because the *terminations* — breakers, lugs, motor terminals — are very often only rated for 75°C, and the ampacity has to respect the lowest-rated point in the circuit, not just the cable itself); and the conductor **material** (copper vs. aluminum have different ampacities at the same size, and this chapter uses copper throughout, the default for most industrial control and branch power wiring in this size range).

Before comparing the load current to the table, the load current itself often needs an adjustment. A load that runs continuously (conventionally, for three hours or more at a stretch) is treated more conservatively than an intermittent one, because sustained full-current operation is exactly the condition the ampacity table's thermal limit is protecting against. In the CEC, this shows up through **Rule 8-104** (conductor ampacity selection) together with **Rule 8-106** (maximum circuit loading), which together limit how much of a conductor's and an overcurrent device's rating can be continuously loaded; the NEC equivalent is more explicit about the arithmetic, requiring a minimum conductor ampacity of **125% of the continuous load** plus 100% of any noncontinuous portion (NEC 210.19(A) for branch circuits, 215.2(A) for feeders). Both approaches land in the same place in practice for a purely continuous load like this heater bank: **size the conductor for 125% of the continuous current**, not the bare nameplate current.

**Rule of thumb:** for a genuinely continuous industrial load (an oven, a heater bank, a pump that runs all shift), size the conductor to 125% of the actual running current before consulting the ampacity table. This rule of thumb breaks down for loads that are clearly *not* continuous by the three-hour convention — a hoist, a press, a robot weld gun — where sizing at 125% needlessly oversizes the conductor for a duty the circuit will never actually see; for those, the applicable rule shifts toward the equipment's own duty-cycle rating, which is a different (and often more forgiving) calculation than this straight continuous-load factor.

### Governing Calculation

$$I_{required} = I_{load} \times CLF$$

Select the smallest standard conductor size from the applicable ampacity table (CEC Table 2 / NEC Table 310.16), at the correct insulation temperature column, such that:

$$I_{table}(\text{AWG}) \geq I_{required}$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $I_{load}$ | Actual running current of the load | A |
| $CLF$ | Continuous load factor (1.25 for a continuous load per Rule 8-106 / NEC 210.19(A); 1.0 for a non-continuous load) | dimensionless |
| $I_{required}$ | Minimum ampacity the selected conductor must meet or exceed | A |
| $I_{table}(\text{AWG})$ | Table ampacity for a given AWG size, at the applicable insulation temperature rating and base (30°C, ≤3 current-carrying conductors) conditions | A |

```{r fig-06-04-ampacity-bar, echo=FALSE, fig.cap="Representative copper conductor ampacity by AWG size (75C insulation, base 30C ambient, three or fewer current-carrying conductors) — illustrative values in the spirit of CEC Table 2 / NEC Table 310.16, not reproduced from either table. The dashed line marks this circuit's required ampacity after the 125% continuous-load factor.", fig.align='center', fig.width=10, fig.height=6}
awg_levels <- c("8", "6", "4", "3", "2", "1", "1/0")
amp_df <- data.frame(
  awg = factor(awg_levels, levels = awg_levels),
  ampacity = c(50, 65, 85, 100, 115, 130, 150)
)
I_required_fig <- 72.5
amp_df$status <- ifelse(amp_df$ampacity >= I_required_fig,
                         "Meets or exceeds required ampacity", "Below required ampacity")

ggplot(amp_df, aes(x = awg, y = ampacity, fill = status)) +
  geom_col() +
  annotate("segment", x = 0.5, xend = 7.5, y = I_required_fig, yend = I_required_fig,
           linetype = "dashed", color = "firebrick", linewidth = 1) +
  annotate("text", x = 1.1, y = I_required_fig + 6, label = "Required ampacity (72.5 A)",
           color = "firebrick", size = 3.2 * diagram_text_scale, hjust = 0) +
  scale_fill_manual(values = c("Meets or exceeds required ampacity" = "#2ECC71",
                                "Below required ampacity" = "#BDC3C7")) +
  labs(title = "Representative Copper Conductor Ampacity by AWG Size",
       subtitle = "75C insulation, illustrative values in the spirit of CEC Table 2 / NEC Table 310.16",
       x = "Conductor size (AWG)", y = "Base ampacity (A)", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 9.5 * diagram_text_scale),
        axis.text = element_text(size = 10 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale),
        legend.text = element_text(size = 9 * diagram_text_scale))
```

### Worked Example

```{r ampacity-selection-06-04, echo=TRUE}
## Representative base ampacity table: copper, 75C insulation, 30C ambient,
## <=3 current-carrying conductors -- illustrative values in the spirit of
## CEC Table 2 / NEC Table 310.16, NOT reproduced from either table.
awg_table <- data.frame(
  awg = c("8", "6", "4", "3", "2", "1", "1/0"),
  ampacity_75C = c(50, 65, 85, 100, 115, 130, 150)
)

I_load <- 58              # A, immersion heater bank, continuous duty
continuous_factor <- 1.25 # 125% sizing factor for a continuous load
I_required <- I_load * continuous_factor

awg_table$passes <- awg_table$ampacity_75C >= I_required

cat("Load current:", I_load, "A (continuous)\n")
cat("Required ampacity (125% of continuous load):", I_required, "A\n\n")
print(awg_table)

selected <- awg_table[awg_table$passes, ][1, ]
cat("\nSmallest conductor meeting the requirement: AWG", selected$awg,
    "( table ampacity =", selected$ampacity_75C, "A )\n")
```

**Sanity check:** AWG 6 (65 A) falls just short of the 72.5 A requirement, and AWG 4 (85 A) clears it with about 17% margin — a sensible result, since jumping straight to a conductor that barely squeaks by would leave zero room for anything discovered in Design Task 2, where the same circuit gets derated for its actual installed ambient temperature and tray fill. This base selection, at ideal 30°C/uncrowded conditions, is only the starting point, not the final answer.

## Design Task 2: Derating for Ambient Temperature and Bundling

### Scenario

The AWG 4 conductor selected in Design Task 1 was sized under the ampacity table's *base* conditions: 30°C ambient, three or fewer current-carrying conductors. The actual installed circuit doesn't live in those conditions. The heater bank's feeder runs through a mezzanine cable tray directly above the phosphate bath, where ambient temperature regularly reaches 40°C, and that same tray section carries eight other current-carrying conductors from adjacent circuits — nine total, including this one. The designer has to check whether AWG 4 still holds up once both real-world effects are applied, not just the one it was originally sized against.

### Design Guidelines and Rules of Thumb

A base ampacity table value applies only under the specific reference conditions it was built around. Two separate, independent effects reduce a conductor's *actual* allowable current below that base value, and both need to be applied, not just the more obvious one:

**Ambient temperature correction.** A conductor's insulation has a fixed maximum operating temperature (60°C, 75°C, or 90°C, matching its rating). The base ampacity table already assumes the conductor is heating up from 30°C ambient to that maximum under full load; if the actual ambient is hotter than 30°C, there's less thermal headroom left before the insulation limit is reached, so the allowable current must come down. The CEC's ambient correction factors live in **Table 5A**; the NEC equivalent is **Table 310.15(B)(1)**. Both are structured as a multiplier, less than 1.0, that gets smaller (a bigger derating) as ambient climbs.

**Bundling / grouping adjustment.** Conductors carrying current generate heat, and when several current-carrying conductors are bundled together in the same tray, conduit, or cable — with three or fewer being the table's baseline assumption — they can't dissipate that heat as freely as an isolated conductor, since each one is now partly insulated from ambient air by the conductors around it. The more current-carrying conductors grouped together, the larger this second derating. The CEC's grouping factors live in **Table 5C**; the NEC equivalent is **Table 310.15(C)(1)**. Note that only *current-carrying* conductors count toward the group size in most cases — a properly identified neutral on a balanced circuit, or an equipment grounding conductor, is typically excluded from the count, which is worth double-checking on a mixed-circuit tray rather than assuming every conductor in the bundle counts.

These two derating factors are **independent and multiplicative** — apply both to the same base ampacity, not just whichever one looks worse in isolation. It's a common and costly mistake to check ambient temperature, confirm the conductor still passes, and stop there without ever checking the grouping factor (or vice versa) — a conductor can pass either check individually and still fail once both are applied together, which is exactly the situation this design task's scenario is built to catch.

**Rule of thumb:** as a first-pass sanity check, expect the *combined* ambient-and-bundling derating factor on a busy industrial tray to land somewhere in the 0.55-0.75 range — if a calculation comes out well outside that band, it's worth double-checking which table row/column was actually used. This rule of thumb breaks down for lightly-loaded trays with only a handful of circuits and a genuinely cool ambient (close to 1.0, since neither factor has much to bite on) and for extreme cases — a tray running through a process oven's enclosure, or a tray genuinely packed with 20+ current-carrying conductors — where the combined factor can fall well below 0.55 and a bare rule-of-thumb sanity check stops being a reliable guide; the actual table values always govern.

### Governing Calculation

$$I_{allowable} = I_{base} \times CF_{ambient} \times CF_{bundling}$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $I_{base}$ | Base table ampacity at 30°C ambient, ≤3 current-carrying conductors (CEC Table 2 / NEC Table 310.16) | A |
| $CF_{ambient}$ | Ambient temperature correction factor for the actual installed ambient (CEC Table 5A / NEC Table 310.15(B)(1)) | dimensionless (0-1) |
| $CF_{bundling}$ | Grouping/bundling adjustment factor for the actual number of current-carrying conductors (CEC Table 5C / NEC Table 310.15(C)(1)) | dimensionless (0-1) |
| $I_{allowable}$ | Actual allowable ampacity of the conductor under real installed conditions | A |

A design passes when $I_{allowable} \geq I_{required}$, using the same $I_{required}$ computed in Design Task 1.

```{r fig-06-04-derating, echo=FALSE, fig.cap="Derated ampacity by conductor size at the actual installed conditions (40C ambient, nine current-carrying conductors bundled in one tray). AWG 4, selected under base conditions in Design Task 1, no longer clears the required ampacity once both derating factors are applied.", fig.align='center', fig.width=10, fig.height=6}
derate_df <- data.frame(
  awg = factor(c("4", "3", "2", "1", "1/0"), levels = c("4", "3", "2", "1", "1/0")),
  derated = c(52.36, 61.60, 70.84, 80.08, 92.40)
)
I_required_fig <- 72.5

ggplot(derate_df, aes(x = awg, y = derated)) +
  geom_col(fill = "#3498DB") +
  annotate("segment", x = 0.5, xend = 5.5, y = I_required_fig, yend = I_required_fig,
           linetype = "dashed", color = "firebrick", linewidth = 1) +
  annotate("text", x = 1.1, y = I_required_fig + 4, label = "Required ampacity (72.5 A)",
           color = "firebrick", size = 3.2 * diagram_text_scale, hjust = 0) +
  labs(title = "Derated Ampacity vs. Conductor Size",
       subtitle = "40C ambient (CF = 0.88) x 9 bundled conductors (CF = 0.70), applied to Task 1's base table values",
       x = "Conductor size (AWG)", y = "Derated allowable ampacity (A)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 9 * diagram_text_scale),
        axis.text = element_text(size = 10 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale))
```

### Worked Example

Live cell — try a different ambient temperature correction factor, bundling adjustment factor, or number of conductors and see which AWG size passes.

```{webr-r}
#| label: derating-06-04
## Continue Task 1's circuit: base conductor selected there was AWG 4
## (85 A @ 75C, 30C ambient, single circuit)
awg_candidates <- data.frame(
  awg = c("4", "3", "2", "1", "1/0"),
  base_ampacity_75C = c(85, 100, 115, 130, 150)
)

## Ambient temperature correction factor (illustrative, 75C column) --
## representative of the shape of CEC Table 5A / NEC Table 310.15(B)(1)
ambient_C <- 40
temp_corr <- 0.88   # representative factor at ~40C for 75C insulation

## Bundling/grouping adjustment factor (illustrative) -- representative of
## the shape of CEC Table 5C / NEC Table 310.15(C)(1)
n_conductors <- 9
bundle_adj <- 0.70  # representative factor for 7-9 current-carrying conductors

I_required <- 72.5  # A, from Task 1 (58 A x 1.25 continuous load factor)

awg_candidates$derated_ampacity <- awg_candidates$base_ampacity_75C * temp_corr * bundle_adj
awg_candidates$passes <- awg_candidates$derated_ampacity >= I_required

cat("Ambient:", ambient_C, "C -> correction factor =", temp_corr, "\n")
cat("Bundled current-carrying conductors:", n_conductors, "-> adjustment factor =", bundle_adj, "\n")
cat("Combined derating factor:", round(temp_corr * bundle_adj, 3), "\n")
cat("Required ampacity:", I_required, "A\n\n")
print(awg_candidates)

selected <- awg_candidates[awg_candidates$passes, ][1, ]
cat("\nSmallest conductor that still clears the requirement after derating: AWG", selected$awg,
    "( derated ampacity =", round(selected$derated_ampacity, 1), "A )\n")
```

**Sanity check:** the combined derating factor here is 0.88 × 0.70 ≈ 0.62 — squarely inside the 0.55-0.75 rule-of-thumb band for a busy industrial tray, a useful confirmation that nothing was mistyped. AWG 4, which comfortably passed Task 1's base-condition check with 17% margin, now derates to just 52.4 A — barely 72% of the 72.5 A requirement — and the circuit has to jump all the way to AWG 1 to pass. This is precisely the trap the rule of thumb above warns against: checking only ambient (AWG 4 at 0.88 alone gives 74.8 A, which still *looks* like a pass) or only bundling (AWG 4 at 0.70 alone gives 59.5 A, an obvious fail) each tells an incomplete story; only the combined check catches the real answer.

## Design Task 3: Voltage Drop Calculations

### Scenario

Two very different circuits on the same expansion project both need a voltage drop check, and they fail — or pass — for very different reasons. First, a 24 VDC solenoid valve at the far end of the new conveyor system, 120 m one-way from its control panel, needs its power run checked before the cable is pulled. Second, the pretreatment line's new 75 HP, 600 V three-phase agitator motor needs its feeder from the MCC checked over a 70 m run. Both circuits are "voltage drop problems," but the physics, the wire sizes, and the tolerance for how much drop is acceptable are not the same problem wearing two hats.

### Design Guidelines and Rules of Thumb

Any current flowing through a conductor's resistance (and, for AC, its reactance) produces a voltage drop along the conductor's length — the load sees less voltage than the source is actually putting out, and the difference is dissipated as heat in the wire itself rather than doing useful work at the load. For a DC or single-phase circuit, current makes a round trip (out on one conductor, back on the other), so the resistive drop has to be counted over *twice* the one-way run length. For a three-phase circuit, only the resistance (and reactance) in the phase conductors matters, and the geometry of a balanced three-phase system introduces a factor of $\sqrt{3}$ rather than 2.

**Rule of thumb: keep voltage drop at or under 3%** of the nominal system voltage, measured from the source to the farthest point on the circuit, for general power and lighting branch circuits and feeders. This is a widely used industry guideline (echoed informally in CEC and NEC guidance notes, though it is a rule of thumb rather than a hard code-mandated numeric limit in most general circuits) rather than a strict pass/fail code rule, and it breaks down in both directions:

- It can be **too loose** for long, low-voltage instrumentation and control runs. A 24 VDC circuit feeding a sensitive analog device — a 0-10 V analog input, a device whose output scaling depends on a stable, well-regulated supply voltage — can have its actual measurement or output accuracy degraded by a voltage drop that would be entirely harmless to a simple on/off digital sensor or solenoid running on the exact same nominal voltage. For circuits like that, a tighter limit (often 5% is still workable for many devices, but the real number to check is the specific device's documented minimum operating voltage and its sensitivity to supply variation) is the safer target, not the generic 3%.
- It can be **too tight** for short branch circuits where 3% would demand an unreasonably large conductor for a trivial length. A receptacle branch circuit only a few meters long, feeding equipment that tolerates a wide input voltage range, can often run comfortably at 4-5% drop without any real operational consequence — the 3% figure is a reasonable *default* starting point, not a number that has to be defended to the decimal point on every circuit in the plant.

The right practice is to treat 3% as the default first check, then ask two follow-up questions before accepting or rejecting a result on that basis alone: what is the *specific device's* actual minimum operating voltage (not just a generic percentage), and is signal integrity — not just delivered power — riding on the same conductors?

### Governing Calculation

For a DC or single-phase (two-conductor) circuit, accounting for the round trip:

$$V_{drop} = 2\,I\,L\,R$$

For a balanced three-phase circuit:

$$V_{drop} = \sqrt{3}\,I\,L\,\left(R\cos\theta + X\sin\theta\right)$$

$$\%V_{drop} = \frac{V_{drop}}{V_{system}} \times 100\%$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $I$ | Load current | A |
| $L$ | One-way conductor length | km (or 1000 ft, matching whichever $R$/$X$ table is used) |
| $R$ | Conductor AC resistance per unit length, from a conductor properties table (CEC Table 4 / NEC Chapter 9, Table 8/9) | Ω/km |
| $X$ | Conductor reactance per unit length, from the same table (three-phase circuits only) | Ω/km |
| $\theta$ | Load power factor angle ($\cos\theta$ = power factor) | — |
| $V_{system}$ | Nominal system voltage | V |
| $V_{drop}$, $\%V_{drop}$ | Voltage drop, in volts and as a percentage of system voltage | V, % |

### Worked Example (a): 24 VDC Sensor/Actuator Run

```{r fig-06-04-24vdc-drop, echo=FALSE, fig.cap="Percent voltage drop vs. one-way cable run length for a 0.35 A, 24 VDC solenoid circuit, at three conductor gauges. The 3% rule-of-thumb line is shown for reference; the marked point is the 120 m scenario evaluated in the worked example.", fig.align='center', fig.width=10, fig.height=6}
L_seq <- seq(10, 200, by = 5)   # m, one-way
I_load3 <- 0.35                  # A
gauge_R <- c("18 AWG" = 25.7, "14 AWG" = 10.2, "12 AWG" = 6.42)  # ohm/km

df_list <- lapply(names(gauge_R), function(g) {
  R <- gauge_R[[g]]
  data.frame(L = L_seq, gauge = g, pct = 2 * I_load3 * (L_seq / 1000) * R / 24 * 100)
})
vdrop_df <- do.call(rbind, df_list)
vdrop_df$gauge <- factor(vdrop_df$gauge, levels = names(gauge_R))

mark_L <- 120
mark_pct_18 <- 2 * I_load3 * (mark_L / 1000) * gauge_R[["18 AWG"]] / 24 * 100

ggplot(vdrop_df, aes(x = L, y = pct, color = gauge)) +
  geom_line(linewidth = 1.2) +
  annotate("segment", x = 10, xend = 200, y = 3, yend = 3,
           linetype = "dashed", color = "grey30", linewidth = 0.9) +
  annotate("text", x = 135, y = 3.7, label = "3% rule-of-thumb limit",
           color = "grey30", size = 3 * diagram_text_scale) +
  annotate("point", x = mark_L, y = mark_pct_18, color = "black", size = 3, shape = 4, stroke = 1.5) +
  annotate("text", x = mark_L + 5, y = mark_pct_18 + 0.6, label = "18 AWG @ 120 m",
           size = 2.8 * diagram_text_scale, hjust = 0) +
  labs(title = "24 VDC Circuit Voltage Drop vs. Run Length",
       subtitle = "0.35 A solenoid load, round-trip resistive drop",
       x = "One-way cable run length (m)", y = "Voltage drop (% of 24 V)", color = "Conductor size") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 9.5 * diagram_text_scale),
        axis.text = element_text(size = 10 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale),
        legend.text = element_text(size = 9 * diagram_text_scale))
```

Live cell — try a different cable length, gauge, or current draw and see how the delivered voltage and percent drop respond.

```{webr-r}
#| label: dc-voltage-drop-06-04
## 24 VDC solenoid valve at the far end of the new conveyor system
V_supply <- 24        # VDC nominal
I_load <- 0.35        # A, solenoid coil current
L_one_way_m <- 120    # m, one-way cable run length

## Representative copper conductor DC resistance values (illustrative,
## consistent with the order of magnitude in CEC Table 4 / NEC Chapter 9
## Table 8 -- consult the actual table for precise figures)
R_18awg <- 25.7   # ohm/km
R_14awg <- 10.2   # ohm/km
R_12awg <- 6.42   # ohm/km

calc_drop <- function(R_ohm_km, L_m, I) {
  L_km <- L_m / 1000
  2 * I * L_km * R_ohm_km   # round-trip resistive drop
}

Vd_18 <- calc_drop(R_18awg, L_one_way_m, I_load)
Vd_14 <- calc_drop(R_14awg, L_one_way_m, I_load)
Vd_12 <- calc_drop(R_12awg, L_one_way_m, I_load)

min_operating_V <- 20.4  # V, representative -15% IEC 61131-2 tolerance floor
                          # for a nominal 24 VDC industrial I/O device

for (nm in c("R_18awg", "R_14awg", "R_12awg")) {
  R <- get(nm)
  Vd <- calc_drop(R, L_one_way_m, I_load)
  pct <- Vd / V_supply * 100
  delivered <- V_supply - Vd
  cat(sprintf("%s (%.2f ohm/km): drop = %.3f V (%.2f%%), delivered = %.2f V -> %s (3%% rule), %s (functional floor)\n",
              nm, R, Vd, pct, delivered,
              ifelse(pct <= 3, "PASS", "FAIL"),
              ifelse(delivered >= min_operating_V, "PASS", "FAIL")))
}
```

**Sanity check:** 18 AWG, a common default for sensor/actuator wiring, fails the 3% rule of thumb badly here (about 9%) — yet the delivered voltage (21.8 V) still clears the solenoid's functional floor (20.4 V) with room to spare, which is exactly the "too loose vs. too tight" tension the guidance above describes: a simple on/off solenoid tolerates this drop just fine, but the same 18 AWG run feeding a 0-10 V analog level sensor at the same distance would corrupt that sensor's reading well before the voltage itself became a problem. Stepping up to 14 AWG brings the drop to about 3.6% — still marginal — and 12 AWG clears the 3% line comfortably at 2.25%. For this specific solenoid, 18 AWG is functionally acceptable; for an analog instrument on the same run, 12 AWG is the safer specification.

### Worked Example (b): 600 V-Class Motor Feeder

A word on voltage class first: Canadian industrial practice, under the CEC, commonly runs three-phase motor and distribution circuits at a nominal **600 V**, where equivalent US industrial installations, under the NEC, commonly standardize on **480 V**. This isn't a minor labeling difference — for the same delivered motor horsepower, the 600 V circuit draws meaningfully less current than its 480 V counterpart (roughly in the inverse ratio of the two voltages), which directly reduces both the $I^2R$ heating that drives conductor sizing and the voltage drop calculated below. A Canadian-trained designer moving between the two conventions needs to recognize that a "480 V motor feeder" spec sheet from a US-sourced project is not a drop-in match for 600 V practice — the conductor and protective device sizing will come out different even for numerically identical horsepower.

```{r fig-06-04-voltage-class-compare, echo=FALSE, fig.cap="Illustrative percent voltage drop for the same motor, feeder length, and conductor, delivered at the Canadian 600 V motor voltage class versus the US 480 V convention. The lower current draw at 600 V produces a meaningfully smaller percent drop for the same physical installation.", fig.align='center', fig.width=9, fig.height=5.5}
compare_df <- data.frame(
  system = factor(c("600 V (CEC / Canadian convention)", "480 V (NEC / US convention)"),
                   levels = c("600 V (CEC / Canadian convention)", "480 V (NEC / US convention)")),
  pct = c(0.825, 1.289)
)

ggplot(compare_df, aes(x = system, y = pct, fill = system)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = sprintf("%.2f%%", pct)), vjust = -0.6, size = 3.5 * diagram_text_scale) +
  scale_fill_manual(values = c("600 V (CEC / Canadian convention)" = "#3498DB",
                                "480 V (NEC / US convention)" = "#F39C12")) +
  ylim(0, 1.6) +
  labs(title = "Same Motor, Same Feeder: Voltage Class Effect on Percent Drop",
       x = NULL, y = "Voltage drop (% of system voltage)") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(size = 13 * diagram_text_scale, face = "bold"),
        axis.text = element_text(size = 10 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale))
```

Live cell — try a different feeder length, motor FLA, power factor, or system voltage and see how the drop responds.

```{webr-r}
#| label: motor-feeder-vdrop-06-04
## Pretreatment line agitator motor: 75 HP, 600 V, three-phase, fed from the
## MCC 70 m away. Nameplate FLA comes from the motor selection covered in
## the Motor, Drive, and Gearbox Sizing chapter -- treated as a given input here.
V_system <- 600   # V, Canadian 600V motor voltage class
I_FLA <- 78        # A, motor nameplate full-load amps
L_m <- 70           # m, one-way feeder length, MCC to motor
pf <- 0.87          # motor power factor at rated load

## Representative conductor resistance/reactance (illustrative, consistent
## with the order of magnitude in CEC Table 4 / NEC Chapter 9 Table 9 for a
## 2/0 AWG copper conductor in steel conduit)
R_ohm_km <- 0.525
X_ohm_km <- 0.135

L_km <- L_m / 1000
costheta <- pf
sintheta <- sqrt(1 - pf^2)

Vdrop_3ph <- sqrt(3) * I_FLA * L_km * (R_ohm_km * costheta + X_ohm_km * sintheta)
pct_drop <- Vdrop_3ph / V_system * 100

cat(sprintf("Feeder: %.0f m, FLA = %.0f A, pf = %.2f\n", L_m, I_FLA, pf))
cat(sprintf("Voltage drop: %.3f V (%.3f%% of %d V) -> %s (3%% rule)\n\n",
            Vdrop_3ph, pct_drop, V_system, ifelse(pct_drop <= 3, "PASS", "FAIL")))

## Illustrative comparison: same motor power, same physical feeder, at the
## US 480V convention instead -- current scales up roughly in inverse
## proportion to voltage for the same delivered power
V_us <- 480
I_FLA_480 <- I_FLA * (V_system / V_us)
Vdrop_480 <- sqrt(3) * I_FLA_480 * L_km * (R_ohm_km * costheta + X_ohm_km * sintheta)
pct_drop_480 <- Vdrop_480 / V_us * 100

cat(sprintf("Same motor at 480V convention: I_FLA ~= %.1f A, drop = %.3f V (%.3f%%)\n",
            I_FLA_480, Vdrop_480, pct_drop_480))
cat(sprintf("Drop is %.1fx larger (as a percentage) at 480V for the identical feeder run.\n",
            pct_drop_480 / pct_drop))
```

**Sanity check:** at well under 1% drop, the 600 V feeder passes the 3% rule of thumb with enormous margin — for a properly-derated feeder conductor at typical industrial run lengths, three-phase motor feeders rarely get anywhere near the 3% line unless the run is unusually long or the conductor unusually small, which is itself a useful sanity check on the result. The side-by-side comparison makes the voltage-class point concrete: the identical feeder, at the identical length, produces roughly 1.6 times the percent drop when the same motor is served at 480 V instead of 600 V — a real, quantifiable consequence of the Canadian convention, not just a difference in nameplate labeling.

## Design Task 4: Cable Type Selection

### Scenario

Three very different runs are going into the same expansion project, and no single cable construction is right for all of them. A new yard feeder has to run underground from a new outdoor transformer pad to the plant's main electrical room. New VFD-driven pump and conveyor drives on the machining floor need output cabling from each drive to its motor. And in a separate part of the business, a food and beverage plant's packaging line washdown zone — where the equipment gets a caustic chemical wash and full hose-down every shift — needs new instrumentation and branch power cable that has to survive that environment for years, not months.

### Design Guidelines and Rules of Thumb

Four cable constructions cover the large majority of industrial power and control wiring situations this chapter's scenarios touch on, and each is built for a specific mechanical and environmental problem.

**TECK90** is an interlocked-armor cable (traditionally aluminum armor) with an overall PVC (or, for tougher chemical/UV service, CPE or CSPE) outer jacket over 90°C-rated conductors, standardized in Canada under **CSA C22.2 No. 131**. The armor gives it real mechanical protection — it can be run exposed in a cable tray, strapped to structure, or direct-buried (when the specific product is rated for it) without additional conduit, which is exactly why it's the default choice for a plant's heavier, harder-to-protect runs: yard feeders, direct burial to remote equipment, and any tray run exposed to real mechanical risk (forklift traffic, falling material, outdoor weather).

**AC90** is a related interlocked-armor cable, standardized under **CSA C22.2 No. 51**, but built for a gentler duty: it has no overall outer jacket (the armor itself is the outer surface), which keeps it lighter and less expensive but restricts it to indoor, dry locations — concealed or exposed wiring in commercial and light industrial buildings. It is not rated for wet locations or direct burial, and using it as a cost-saving substitute for TECK90 anywhere with moisture, washdown, or outdoor exposure is a straightforward code violation, not a judgment call.

**Tray cable** is a multiconductor cable (typically with an overall jacket, no metallic armor) specifically rated to be supported in a cable tray system without conduit. It's a common, cost-effective choice for control and instrumentation wiring runs in a process plant that already has tray infrastructure in place — lighter and cheaper than TECK90 per meter, since it isn't carrying armor's mechanical-protection cost, but it also isn't offering that protection: physical protection against impact comes only from the tray's own routing and covers, not from the cable itself, and some jurisdictions restrict tray cable for power circuits above certain voltage or conductor-size thresholds without additional engineering justification.

**VFD cable** is built to solve a problem none of the other three constructions address at all: a variable frequency drive's PWM output waveform has very fast voltage rise times (high dv/dt), and a standard three-conductor power cable running that waveform to a motor can produce reflected-wave voltage spikes at the motor terminals, common-mode electrical noise that couples into nearby signal wiring, and elevated shaft currents that erode motor bearings over time. VFD cable addresses this with a distinctive **symmetrical construction**: three phase conductors plus three equally-sized, symmetrically-placed ground conductors (rather than one undersized ground), typically under an overall shield or braid, all under a common jacket. The symmetry and shielding give the return current a low-impedance, geometrically balanced path back to the drive, which is what actually controls the reflected-wave and common-mode effects — a standard TC or AC90 cable with a single small ground conductor does not provide this, no matter how correctly it's otherwise sized for the motor's current.

```{r fig-06-04-cable-cross-section, echo=FALSE, fig.cap="Schematic cross-sections: TECK90 (interlocked armor, PVC jacket, standard ground) vs. VFD cable (shielded, symmetrical ground conductors interleaved with the phase conductors). Schematic, not to scale.", fig.align='center', fig.width=11, fig.height=6}
teck_cx <- -4; teck_cy <- 0
vfd_cx <- 4; vfd_cy <- 0

trigonal_pos <- function(cx, cy, R, n = 3, start = 90) {
  ang <- start + seq(0, 360, length.out = n + 1)[1:n]
  data.frame(x0 = cx + R * cos(ang * pi / 180), y0 = cy + R * sin(ang * pi / 180))
}

teck_jacket <- data.frame(x0 = teck_cx, y0 = teck_cy, r = 2.6)
teck_armor  <- data.frame(x0 = teck_cx, y0 = teck_cy, r = 2.3)
teck_cond <- trigonal_pos(teck_cx, teck_cy, 1.1)
teck_cond$r <- 0.75
teck_cond$label <- c("L1", "L2", "L3")

vfd_jacket <- data.frame(x0 = vfd_cx, y0 = vfd_cy, r = 2.6)
vfd_shield <- data.frame(x0 = vfd_cx, y0 = vfd_cy, r = 2.3)
vfd_phase <- trigonal_pos(vfd_cx, vfd_cy, 1.1, start = 90)
vfd_phase$r <- 0.7
vfd_phase$label <- c("L1", "L2", "L3")
vfd_gnd <- trigonal_pos(vfd_cx, vfd_cy, 1.1, start = 150)
vfd_gnd$r <- 0.35
vfd_gnd$label <- c("G", "G", "G")

ggplot() +
  geom_circle(data = teck_jacket, aes(x0 = x0, y0 = y0, r = r), color = "grey30", linewidth = 1.1) +
  geom_circle(data = teck_armor, aes(x0 = x0, y0 = y0, r = r), color = "grey50", linetype = "dashed", linewidth = 0.8) +
  geom_circle(data = teck_cond, aes(x0 = x0, y0 = y0, r = r), fill = "#F39C12", color = "black") +
  geom_text(data = teck_cond, aes(x = x0, y = y0, label = label), size = 3 * diagram_text_scale) +
  annotate("text", x = teck_cx, y = -3.3, label = "TECK90\n(interlocked armor + PVC jacket)",
           size = 3.2 * diagram_text_scale, fontface = "bold") +
  geom_circle(data = vfd_jacket, aes(x0 = x0, y0 = y0, r = r), color = "grey30", linewidth = 1.1) +
  geom_circle(data = vfd_shield, aes(x0 = x0, y0 = y0, r = r), color = "#3498DB", linetype = "dashed", linewidth = 0.8) +
  geom_circle(data = vfd_phase, aes(x0 = x0, y0 = y0, r = r), fill = "#F39C12", color = "black") +
  geom_text(data = vfd_phase, aes(x = x0, y = y0, label = label), size = 3 * diagram_text_scale) +
  geom_circle(data = vfd_gnd, aes(x0 = x0, y0 = y0, r = r), fill = "#95A5A6", color = "black") +
  geom_text(data = vfd_gnd, aes(x = x0, y = y0, label = label), size = 2.4 * diagram_text_scale) +
  annotate("text", x = vfd_cx, y = -3.3, label = "VFD Cable\n(shielded, symmetrical ground conductors)",
           size = 3.2 * diagram_text_scale, fontface = "bold") +
  coord_fixed(xlim = c(-7.2, 7.2), ylim = c(-4, 3.5)) +
  theme_void()
```

The US/NEC equivalents track closely: TECK90 corresponds most closely to Type MC (metal-clad) cable (NEC Article 330, product-listed to UL 1569); AC90 corresponds to Type AC (armored) cable (NEC Article 320); tray cable corresponds to Type TC cable (NEC Article 336, product-listed to UL 1277); VFD cable has no single dedicated NEC article, but is commonly product-listed to UL 2277 and specified against NEMA motor-drive application guidance.

<details>
<summary>Industry connection: washdown-rated jacketing in a food plant</summary>

Standard TECK90 and tray cable jackets (plain PVC) are not automatically the right answer in a food and beverage plant's washdown zone: the daily caustic-chemical wash and hot-water hose-down that a CFIA-regulated sanitary environment expects will degrade an ordinary PVC jacket over a few years, well before the cable's conductors themselves are anywhere near end of life. The same cable constructions are available with chemical- and UV-resistant jacket compounds (commonly CPE or CSPE) built for exactly this duty, and terminations in these areas need IP69K-rated cable glands to actually deliver on the cable's washdown rating — specifying the right conductor and armor but the wrong jacket compound, or the right cable but a standard (non-sanitary) gland, is a common and expensive oversight on a food-plant electrical spec.

</details>

```{r tbl-06-04-cable-selection, echo=FALSE, message=FALSE}
cable_options <- tribble(
  ~Cable_Type, ~Construction, ~Typical_Permitted_Use, ~Key_Limitation,
  "TECK90", "Interlocked armor (aluminum) + PVC/CPE/CSPE overall jacket, 90C conductors",
  "Direct burial (jacket-rated products), cable tray, exposed runs without conduit, wet/damp and outdoor locations, mechanically hazardous areas",
  "Heaviest and most expensive of the four; larger bend radius; overkill for benign indoor dry runs",
  "AC90", "Interlocked armor over individually insulated conductors; no overall outer jacket",
  "Indoor, dry locations only; concealed and exposed wiring in commercial/light industrial buildings",
  "No outer jacket -- no real moisture, chemical, or UV resistance; not permitted for wet locations or direct burial",
  "Tray cable (TC)", "Multiconductor cable, overall jacket, no metallic armor, listed for tray support without conduit",
  "Control/instrumentation and (within limits) power runs in plants with existing cable tray systems",
  "No mechanical armor -- physical protection depends entirely on the tray, not the cable; some jurisdictions restrict TC for larger power circuits",
  "VFD cable", "Three symmetrical phase conductors + three symmetrical ground conductors + overall shield/braid + jacket",
  "VFD-to-motor output runs, especially longer runs, where reflected-wave voltage spikes, common-mode EMI, and bearing current are a concern",
  "Meaningfully more expensive per meter; the symmetrical/shielded construction is wasted cost on a non-VFD circuit"
)

cable_options %>%
  kbl(caption = "Cable type selection guide: construction and permitted use",
      col.names = c("Cable Type", "Construction", "Typical Permitted Use", "Key Limitation")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, background = "#2C3E50", color = "white")
```

### Governing Calculation

There's no formula here — cable type selection is a documented decision framework, matching each installation's environment and electrical characteristics against the four constructions' permitted uses and limitations above:

1. Does the run need mechanical protection without conduit (direct burial, exposed to impact, outdoor)? -> **TECK90**.
2. Is it strictly indoor and dry, with no mechanical-protection need beyond what the armor itself provides? -> **AC90** is acceptable and typically the lower-cost choice.
3. Is it a control/instrumentation or moderate power run in an existing, well-protected tray system? -> **Tray cable**.
4. Is it a VFD's motor output? -> **VFD cable**, regardless of how any of the first three questions would otherwise answer, because the symmetrical/shielded construction addresses a problem (reflected wave, common-mode EMI, bearing current) that only exists on a PWM drive output.
5. Regardless of the above, does the environment involve washdown, caustic chemicals, or sustained UV/outdoor exposure? -> add a chemical/UV-resistant jacket option (and IP69K-rated glanding at every termination) on top of whichever base construction the first four questions selected.

### Worked Example

Live cell — try changing an application's environment or electrical characteristics and see the recommendation change.

```{webr-r}
#| label: cable-selection-06-04
## Four runs from the two scenarios in this design task
applications <- data.frame(
  run = c("New outdoor yard feeder, transformer pad to electrical room",
          "VFD output, machining floor drive to conveyor gearmotor",
          "Overhead control wiring, dry electrical room, existing tray",
          "Washdown-zone instrumentation power, food plant packaging line"),
  direct_buried_or_exposed_outdoor = c(TRUE, FALSE, FALSE, FALSE),
  is_vfd_output = c(FALSE, TRUE, FALSE, FALSE),
  dry_indoor_tray_only = c(FALSE, FALSE, TRUE, FALSE),
  washdown_or_chemical = c(FALSE, FALSE, FALSE, TRUE),
  stringsAsFactors = FALSE
)

for (i in seq_len(nrow(applications))) {
  a <- applications[i, ]
  cat("Run:", a$run, "\n")

  if (a$is_vfd_output) {
    choice <- "VFD cable"
    reason <- "the reflected-wave and common-mode effects of a PWM drive output override every other consideration."
  } else if (a$direct_buried_or_exposed_outdoor) {
    choice <- "TECK90"
    reason <- "the run needs full mechanical protection without conduit, including direct burial and outdoor exposure."
  } else if (a$dry_indoor_tray_only) {
    choice <- "Tray cable (or AC90, if not tray-supported)"
    reason <- "the environment is dry, indoor, and already served by a tray system -- armor is not needed."
  } else {
    choice <- "Base construction per the other criteria"
    reason <- "no distinguishing factor beyond the base decision framework."
  }

  if (a$washdown_or_chemical) {
    choice <- paste(choice, "+ CPE/CSPE chemical-resistant jacket + IP69K glanding")
    reason <- paste(reason, "The washdown/caustic-chemical environment also requires a",
                     "chemical-resistant jacket compound and sanitary-rated (IP69K) glands",
                     "at every termination, regardless of which base cable is otherwise selected.")
  }

  cat("  Recommended:", choice, "\n")
  cat("  Reasoning:", reason, "\n\n")
}
```

**Sanity check:** each recommendation traces back to a single dominant environmental or electrical driver — burial/outdoor exposure for TECK90, PWM output characteristics for VFD cable, an existing tray system for tray cable — exactly the traceability a documented decision framework is supposed to provide. The washdown zone is a useful reminder that jacket selection is a *layer on top of* the base cable-type decision, not a fifth cable type competing with the other four.

## Design Task 5: Fuse vs. Breaker Tradeoffs

### Scenario

The new machining-floor VFD lineup needs overcurrent protection specified for each drive's branch circuit, and the panel feeding them sits close enough to the plant's main service transformer that a recent short-circuit study puts the available fault current at that bus at 42 kA. The electrical designer has two fundamentally different device families to choose from — fuses and circuit breakers — and the choice has real consequences for both safety and plant uptime that go well beyond which one is cheaper to buy.

### Design Guidelines and Rules of Thumb

Fuses and breakers both exist to open a circuit on overcurrent, but they interrupt a fault in genuinely different ways, and that difference drives every practical tradeoff between them.

A **current-limiting fuse** physically melts a metallic element and interrupts the fault within a fraction of a cycle for a high enough fault current — fast enough, for large faults, that it actually cuts off the fault current's rise before it reaches its full available peak. This is the fuse's defining advantage: low **let-through energy** ($I^2t$) and low peak **let-through current**, both of which determine how much stress downstream equipment (busbars, conductors, the VFD itself) actually experiences during a fault, independent of how large the *available* fault current at that point in the system happens to be. A **circuit breaker**, unless it's specifically built and listed as current-limiting, generally does not cut the fault current off before it reaches its natural peak — it senses the overcurrent and then mechanically opens contacts, a process that (even at breaker speed) lets substantially more energy through for the same available fault current.

That let-through difference matters directly for a piece of equipment's **short-circuit current rating (SCCR)** — the maximum fault current the equipment (a VFD, a motor starter, an entire panel) is rated to survive without becoming a hazard. Manufacturers frequently publish a much higher *tested* SCCR for their equipment when protected by a specific current-limiting fuse than when protected by a standard breaker, because the fuse genuinely reduces what the equipment has to survive.

Resettability runs the opposite direction: a breaker, once it trips on a fault, can typically be reset and returned to service in seconds (after the fault itself is found and cleared) — a fuse has to be physically replaced, which means carrying the correct spare in inventory and dispatching someone to swap it. On a production line where downtime is measured in dollars per minute, that difference between "flip a switch" and "find a spare, get a technician, replace a part" is not a minor convenience question.

Cost runs in the breaker's favor at the individual-device level for moderate interrupting ratings — but a fair cost comparison has to include the recurring cost of fuse replacement over the equipment's life, and, if a lower-cost breaker option doesn't clear the required SCCR at all, the "cheaper" option isn't actually a valid option.

Selectivity — covered in full in Design Task 6 — is generally easier to achieve reliably between two same-class fuses of sufficiently different ampere ratings than between two standard (non-current-limiting) breakers of the same family, because fuse melting time is driven by ampere rating and energy in a way that scales predictably, where many standard breakers share a fixed, non-adjustable instantaneous trip region that doesn't separate cleanly between an upstream and downstream device at high fault currents.

**Rule of thumb:** as a first-pass selectivity check between two fuses of the same class, an ampere-rating ratio of **at least 2:1** (upstream to downstream) is usually enough to guarantee full selective coordination across the available fault-current range, without needing to overlay the manufacturer's actual time-current curves. This rule of thumb breaks down between two *breakers*, especially from the same product family — ampere rating alone says very little about whether their instantaneous trip regions actually separate, and it also breaks down for fuses of *different* classes or manufacturers, where the 2:1 heuristic isn't guaranteed to hold and an actual curve overlay is needed.

Governing device standards, cited by name and number rather than reproduced: in Canada, low-voltage fuses fall under **CSA C22.2 No. 248** (series) and molded-case circuit breakers under **CSA C22.2 No. 5**; the corresponding US/UL listings are **UL 248** (series) for fuses and **UL 489** for molded-case breakers.

```{r fig-06-04-letthrough, echo=FALSE, fig.cap="Illustrative peak let-through current vs. available fault current for a current-limiting fuse compared with a standard (non-current-limiting) molded-case breaker. The marked point is the 42 kA available fault current in this design task's scenario. Representative shape only, not manufacturer let-through data.", fig.align='center', fig.width=10, fig.height=6}
I_avail_kA <- seq(1, 100, by = 1)
breaker_let_through <- 2.3 * I_avail_kA
fuse_threshold <- 6   # kA, illustrative current-limiting threshold for this fuse rating
fuse_let_through <- ifelse(I_avail_kA <= fuse_threshold,
                            2.3 * I_avail_kA,
                            2.3 * fuse_threshold + (I_avail_kA - fuse_threshold) * 0.15)

lt_df <- data.frame(
  I = rep(I_avail_kA, 2),
  peak = c(breaker_let_through, fuse_let_through),
  type = rep(c("Standard molded-case breaker (not current-limiting)", "Current-limiting fuse"),
             each = length(I_avail_kA))
)

mark_I <- 42
mark_breaker <- 2.3 * mark_I
mark_fuse <- 2.3 * fuse_threshold + (mark_I - fuse_threshold) * 0.15

ggplot(lt_df, aes(x = I, y = peak, color = type)) +
  geom_line(linewidth = 1.2) +
  annotate("point", x = mark_I, y = mark_breaker, color = "black", size = 3) +
  annotate("point", x = mark_I, y = mark_fuse, color = "black", size = 3) +
  annotate("segment", x = mark_I, xend = mark_I, y = 0, yend = mark_breaker,
           linetype = "dotted", color = "grey40") +
  annotate("text", x = mark_I + 2, y = mark_breaker + 5, label = "42 kA available\n(this panel's fault study)",
           size = 2.8 * diagram_text_scale, hjust = 0) +
  scale_color_manual(values = c("Standard molded-case breaker (not current-limiting)" = "#E74C3C",
                                 "Current-limiting fuse" = "#2ECC71")) +
  labs(title = "Peak Let-Through Current vs. Available Fault Current (Illustrative)",
       x = "Available fault current, RMS symmetrical (kA)", y = "Peak let-through current (kA)", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 13 * diagram_text_scale, face = "bold"),
        axis.text = element_text(size = 10 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale),
        legend.text = element_text(size = 9 * diagram_text_scale))
```

### Governing Calculation

There's no arithmetic formula for "fuse or breaker" — it's a decision framework, anchored by one hard pass/fail gate and a set of softer trade-offs:

$$SCCR_{tested} \geq I_{fault,available} \quad \text{(hard requirement -- equipment must survive the actual available fault current)}$$

$$\text{Fuse selectivity check (rule of thumb):} \quad \frac{I_{upstream}}{I_{downstream}} \geq 2$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $SCCR_{tested}$ | Equipment's tested/listed short-circuit current rating with the specific protective device combination | A (often expressed in kA) |
| $I_{fault,available}$ | Available fault current at the point of installation, from a facility short-circuit study | A (kA) |
| $I_{upstream}$, $I_{downstream}$ | Ampere ratings of the upstream and downstream fuses in a series pair | A |

### Worked Example

Live cell — try a different available fault current, or different tested SCCR values, and see which device(s) still pass.

```{webr-r}
#| label: fuse-breaker-06-04
## Available fault current at the branch panel bus, from a facility
## short-circuit study
I_fault_available <- 42000   # A (42 kA) at 600 V, this panel's bus

## Two protection options for the VFD branch circuit -- tested SCCR values
## as published by the VFD manufacturer for each protective device combination
fuse_tested_SCCR <- 100000     # A, VFD's listed SCCR with the manufacturer's
                                # specified Class J current-limiting fuse
breaker_tested_SCCR <- 25000   # A, VFD's listed SCCR with a standard
                                # molded-case breaker (no specific tested combo)

fuse_passes <- fuse_tested_SCCR >= I_fault_available
breaker_passes <- breaker_tested_SCCR >= I_fault_available

cat(sprintf("Available fault current: %s A (%.0f kA)\n",
            format(I_fault_available, big.mark = " ", scientific = FALSE), I_fault_available / 1000))
cat(sprintf("Fuse-protected tested SCCR: %s A -> %s\n",
            format(fuse_tested_SCCR, big.mark = " ", scientific = FALSE), ifelse(fuse_passes, "PASSES", "FAILS")))
cat(sprintf("Breaker-protected tested SCCR: %s A -> %s\n\n",
            format(breaker_tested_SCCR, big.mark = " ", scientific = FALSE), ifelse(breaker_passes, "PASSES", "FAILS")))

if (fuse_passes && !breaker_passes) {
  cat("Conclusion: only the current-limiting fuse option clears the actual\n")
  cat("available fault current at this bus -- the standard breaker option is\n")
  cat("not a valid choice here regardless of its lower unit cost or resettability.\n\n")
}

## Fuse-to-fuse selectivity ratio check for the branch/main fuse pair
I_upstream_fuse <- 400    # A, main/feeder fuse
I_downstream_fuse <- 60   # A, VFD branch fuse
ratio <- I_upstream_fuse / I_downstream_fuse
selectivity_ok <- ratio >= 2

cat(sprintf("Fuse ampere-rating ratio (upstream/downstream): %.2f:1 -> %s (2:1 rule of thumb)\n",
            ratio, ifelse(selectivity_ok, "PASSES", "FAILS")))
```

**Sanity check:** the standard breaker option fails outright here, not on a soft trade-off basis but on the hard SCCR gate — its tested rating (25 kA) is well below the panel's actual 42 kA available fault current, which makes it an unsafe choice regardless of how attractive its resettability or lower unit cost might otherwise be. This is the right order of operations for this decision: check the hard SCCR requirement first, and only weigh cost, resettability, and selectivity among the options that actually pass it.

## Design Task 6: Selective Coordination

### Scenario

The machining floor's new VFD lineup shares its 400 A main feeder breaker (or fuse, depending on the Design Task 5 outcome) with several other cells' drives, all fed from branch devices downstream of that shared main. If a single drive's power module faults, the plant needs only that one drive's branch protection to open — not the shared main, which would black out every other cell on that feeder for the sake of one failure. Whether that actually happens is a question with a specific, checkable answer: **selective coordination**.

### Design Guidelines and Rules of Thumb

Selective coordination means that, for any fault current the system can actually produce at a given point, the **downstream** (nearest-to-the-fault) overcurrent device clears the fault completely before the **upstream** device has any chance to begin operating. When that's true across the full range of fault currents the circuit could see, a fault anywhere downstream trips only the one device closest to it — every other branch on the shared upstream device keeps running.

The way to check this is to compare each device's **time-current characteristic** — how long the device takes to clear a fault, as a function of the fault current's magnitude — over the full range of fault currents the installation could actually produce. If the downstream device's clearing time curve sits entirely below (faster than) the upstream device's minimum operating time curve across that whole range, the pair is coordinated. If the two curves ever touch or cross — most commonly at high fault currents, where many devices' behavior changes character — coordination is not guaranteed at that current, and a fault there could trip both devices.

This is exactly where fuses and breakers tend to diverge in practice, building on Design Task 5's distinction. Two properly ampere-rated fuses of the same class, particularly with the 2:1 (or greater) ratio discussed in Design Task 5, tend to stay well separated across their *entire* operating range, because fuse clearing time is governed by the element's ampere rating and thermal energy in a way that scales predictably with size. Many standard, non-current-limiting circuit breakers, by contrast, include an **instantaneous trip** element with no intentional time delay — a design feature meant to clear a severe fault as fast as mechanically possible — and that instantaneous element typically has a fixed pickup multiplier (commonly around 8-10 times the breaker's rated current) that doesn't scale down for the upstream device relative to the downstream one. Once a fault current is large enough to exceed *both* devices' instantaneous pickup thresholds, both breakers can begin opening at essentially the same, near-instant speed — a race condition, not a designed hierarchy, and one where either device (not reliably the downstream one) might open first.

Why this matters beyond a single line's uptime: in a multi-cell automotive plant, the shared feeder or MCC bus behind an uncoordinated pair is very often serving several other cells that had nothing to do with the fault. Losing that shared main over one drive's internal fault turns a single-cell repair into a whole-line stoppage, with every other cell's operators standing idle and every other cell's WIP now sitting mid-process — a dramatically larger cost than the original fault, entirely attributable to a coordination gap that could have been caught at the design stage.

**Rule of thumb:** as a first-pass check without overlaying full manufacturer time-current curves, look for a clearing-time **margin of at least 2:1** (upstream time divided by downstream time) at the highest fault current the pair could realistically see. This breaks down at the exact boundary of an instantaneous trip region — a margin that looks comfortable at a moderate fault current can collapse to nearly 1:1 (or worse) once the fault current climbs high enough to push both devices into their instantaneous bands, which is exactly the failure mode illustrated below; a genuine coordination study checks the margin across the *entire* current range, not just at one assumed fault level.

CEC Section 14 (Overcurrent Protection) expects protective devices to be applied so that a fault is cleared by the nearest upstream device — general good coordination practice rather than a single explicit numeric mandate for most circuits — and Canadian emergency-power system design under **CSA C282** expects reliable, selective operation for life-safety loads specifically. The NEC is more explicit: **NEC 240.12** formally defines "selective coordination," and **Articles 700, 701, and 708** (emergency systems, legally required standby systems, and critical operations power systems, respectively) make full selective coordination a mandatory requirement, not just good practice, for those specific system types.

```{r fig-06-04-coordination, echo=FALSE, fig.cap="Illustrative time-current comparison: a coordinated fuse pair stays clearly separated across the whole plotted range, while an uncoordinated breaker pair's curves converge and nearly overlap once both devices reach their instantaneous trip regions. Representative shape only, not manufacturer curve data.", fig.align='center', fig.width=11, fig.height=6}
fuse_downstream <- data.frame(I = c(100, 200, 500, 1000, 3500, 6000),
                               t = c(8, 1.2, 0.09, 0.015, 0.004, 0.004),
                               device = "Downstream 60A fuse", pair = "Coordinated fuse pair")
fuse_upstream <- data.frame(I = c(500, 1000, 2000, 3500, 6000, 10000),
                             t = c(25, 3.5, 0.7, 0.28, 0.10, 0.03),
                             device = "Upstream 400A fuse", pair = "Coordinated fuse pair")
brk_downstream <- data.frame(I = c(300, 600, 800, 1000, 3500, 6000),
                              t = c(8, 1.0, 0.025, 0.025, 0.025, 0.025),
                              device = "Downstream 100A breaker", pair = "Uncoordinated breaker pair")
brk_upstream <- data.frame(I = c(500, 1000, 2000, 3200, 3500, 6000),
                            t = c(30, 6, 1.0, 0.03, 0.03, 0.03),
                            device = "Upstream 400A breaker", pair = "Uncoordinated breaker pair")

curve_df <- rbind(fuse_downstream, fuse_upstream, brk_downstream, brk_upstream)
curve_df$device <- factor(curve_df$device,
                           levels = c("Downstream 60A fuse", "Upstream 400A fuse",
                                      "Downstream 100A breaker", "Upstream 400A breaker"))
curve_df$pair <- factor(curve_df$pair, levels = c("Coordinated fuse pair", "Uncoordinated breaker pair"))

ggplot(curve_df, aes(x = I, y = t, color = device, linetype = pair)) +
  geom_line(linewidth = 1.1) +
  annotate("segment", x = 3500, xend = 3500, y = 0.002, yend = 40,
           linetype = "dotted", color = "grey40") +
  annotate("text", x = 3500, y = 45, label = "Fault current of interest (3500 A)",
           size = 2.7 * diagram_text_scale, color = "grey40") +
  scale_x_log10(labels = scales::label_comma(big.mark = " ")) +
  scale_y_log10() +
  guides(linetype = "none") +
  facet_wrap(~pair) +
  labs(title = "Selective Coordination: Time-Current Comparison (Illustrative)",
       x = "Fault current (A, log scale)", y = "Clearing time (s, log scale)", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(size = 13 * diagram_text_scale, face = "bold"),
        axis.text = element_text(size = 9 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale),
        legend.text = element_text(size = 8.5 * diagram_text_scale),
        strip.text = element_text(size = 10 * diagram_text_scale))
```

### Governing Calculation

$$\text{Coordinated at } I \iff t_{downstream}(I) < t_{upstream}(I), \quad \frac{t_{upstream}(I)}{t_{downstream}(I)} \geq M$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $I$ | Fault current at the point of interest | A |
| $t_{downstream}(I)$ | Downstream device's clearing time at fault current $I$ | s |
| $t_{upstream}(I)$ | Upstream device's minimum operating time at fault current $I$ | s |
| $M$ | Margin threshold applied as a rule-of-thumb confidence check (2, as above) | dimensionless |

### Worked Example

Live cell — try a different fault current, or different curve points, and see whether each pair stays coordinated.

```{webr-r}
#| label: selective-coordination-06-04
## Illustrative (not manufacturer) time-current data points for two
## upstream/downstream device pairs, read from the figure above
fuse_downstream <- data.frame(I = c(100, 200, 500, 1000, 3500, 6000),
                               t = c(8, 1.2, 0.09, 0.015, 0.004, 0.004))
fuse_upstream <- data.frame(I = c(500, 1000, 2000, 3500, 6000, 10000),
                             t = c(25, 3.5, 0.7, 0.28, 0.10, 0.03))
brk_downstream <- data.frame(I = c(300, 600, 800, 1000, 3500, 6000),
                              t = c(8, 1.0, 0.025, 0.025, 0.025, 0.025))
brk_upstream <- data.frame(I = c(500, 1000, 2000, 3200, 3500, 6000),
                            t = c(30, 6, 1.0, 0.03, 0.03, 0.03))

interp_t <- function(curve, I_fault) approx(curve$I, curve$t, xout = I_fault, rule = 2)$y

I_fault <- 3500   # A, available fault current at the branch panel, from a
                   # facility short-circuit study

t_fuse_down <- interp_t(fuse_downstream, I_fault)
t_fuse_up   <- interp_t(fuse_upstream, I_fault)
t_brk_down  <- interp_t(brk_downstream, I_fault)
t_brk_up    <- interp_t(brk_upstream, I_fault)

margin_fuse <- t_fuse_up / t_fuse_down
margin_brk  <- t_brk_up / t_brk_down

cat(sprintf("At I = %s A:\n\n", format(I_fault, big.mark = " ", scientific = FALSE)))
cat(sprintf("Fuse pair:    downstream = %.4f s, upstream = %.4f s, margin = %.1fx -> %s\n",
            t_fuse_down, t_fuse_up, margin_fuse, ifelse(margin_fuse >= 2, "COORDINATED", "NOT COORDINATED")))
cat(sprintf("Breaker pair: downstream = %.4f s, upstream = %.4f s, margin = %.2fx -> %s\n\n",
            t_brk_down, t_brk_up, margin_brk, ifelse(margin_brk >= 2, "COORDINATED", "NOT COORDINATED")))

if (margin_brk < 2) {
  cat("Both breakers have entered their instantaneous trip region at this fault\n")
  cat("current (downstream pickup 800 A, upstream pickup 3200 A, both well below\n")
  cat("3500 A), so their clearing times converge and coordination cannot be\n")
  cat("relied upon -- a fault here could trip the shared upstream breaker along\n")
  cat("with the intended downstream device, blacking out every other cell fed\n")
  cat("from that main.\n")
}
```

**Sanity check:** the fuse pair's 70x margin at the fault current of interest isn't a close call in either direction — it stays coordinated with room to spare, consistent with the 2:1 selectivity ratio rule of thumb from Design Task 5 translating into an even larger margin once actual clearing times are compared. The breaker pair's 1.2x margin is the story this design task exists to teach: both devices are individually correctly rated and individually would clear this fault safely, but *together*, at this specific fault current, neither one is reliably faster than the other — exactly the failure mode a plant only discovers the hard way, during an actual fault, if it isn't checked at the design stage.

<details>
<summary>Industry connection: multi-cell line uptime in automotive assembly</summary>

On a multi-cell automotive line where a dozen or more robot cells, conveyor drives, and vision stations share a small number of upstream MCC feeders, an uncoordinated overcurrent pair turns a single cell's routine drive fault into a full-line stoppage — every other cell on that shared feeder goes dark along with the one that actually failed, and restart after an unplanned trip (re-homing robots, clearing WIP, re-establishing PLC/network state across every affected cell) can cost far more line time than the original fault itself would have, which is exactly why a plant's electrical standard should require a documented coordination study for any new MCC feeder addition, not just a code-minimum overcurrent device selection.

</details>

## Summary

```{r tbl-06-04-summary, echo=FALSE}
summary_df <- data.frame(
  `Design Task` = c(
    "1. Conductor ampacity selection",
    "2. Derating (ambient + bundling)",
    "3. Voltage drop (DC/1-phase & 3-phase)",
    "4. Cable type selection",
    "5. Fuse vs. breaker",
    "6. Selective coordination"
  ),
  `Governing Relationship` = c(
    "$I_{required}=I_{load}\\times CLF$; select AWG where $I_{table}\\geq I_{required}$",
    "$I_{allowable}=I_{base}\\times CF_{ambient}\\times CF_{bundling}$",
    "$V_{drop}=2IL R$ (DC/1$\\phi$); $V_{drop}=\\sqrt{3}IL(R\\cos\\theta+X\\sin\\theta)$ (3$\\phi$)",
    "Decision framework: burial/mechanical exposure, VFD output, tray-only, washdown jacket",
    "$SCCR_{tested}\\geq I_{fault,available}$; fuse ratio $I_{up}/I_{down}\\geq 2$",
    "$t_{downstream}(I) < t_{upstream}(I)$, margin $\\geq 2\\times$ across the fault-current range"
  ),
  `Key Rule of Thumb` = c(
    "Size a continuous load at 125% before consulting the ampacity table",
    "Combined ambient x bundling factor is typically 0.55-0.75 on a busy tray -- apply both, never just one",
    "3% is a default, not a law -- tighten for sensitive analog 24VDC runs, loosen for short branch circuits",
    "Match construction to the dominant environmental/electrical driver; add washdown jacket as a separate layer",
    "Check the hard SCCR gate first; only then weigh cost, resettability, and selectivity",
    "A 2:1 ampere-ratio rule of thumb works for fuses; breaker instantaneous trip regions can defeat it entirely"
  ),
  check.names = FALSE
)
summary_df[] <- lapply(summary_df, tex_safe)

kable(summary_df, caption = "Summary: Wire, Cable, and Circuit Protection design tasks", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = knitr::is_html_output()) %>%
  row_spec(0, bold = TRUE, background = "#4a86c7", color = "white") %>%
  column_spec(1, width = "1in")
```

## Practice Problems

<details>
<summary>Design Task 1, Problem 1 (plug-in): sizing a conductor for a 92 A continuous oven bank</summary>

Using the same representative ampacity table as the worked example (8 = 50 A, 6 = 65 A, 4 = 85 A, 3 = 100 A, 2 = 115 A, 1 = 130 A, 1/0 = 150 A), select the smallest AWG conductor for a 92 A continuous load.

**Solution:**

$$I_{required} = 92 \times 1.25 = 115\ \text{A}$$

AWG 2 (115 A) is the smallest size that meets or exceeds 115 A — it passes with zero margin to spare, which is worth flagging for a closer look rather than accepting automatically (see Problem 2 below).

</details>

<details>
<summary>Design Task 1, Problem 2 (design decision): does the 125% factor apply to an intermittent compactor motor?</summary>

A scrap compactor motor draws 78 A, but only runs for about 90 seconds at a time, a few times per hour, never continuously for anything close to three hours. A junior designer wants to size the conductor at 78 × 1.25 = 97.5 A "to be safe." Is that the right call?

**Answer:** Not necessarily, and blindly applying the continuous-load factor to every circuit is itself a design mistake, not a conservative one — it can lead to habitually oversized (and more expensive) conductors on circuits that never needed the margin, while giving false confidence that "125%" is a universal safety factor rather than a specific response to a specific duty pattern. This motor's duty cycle (90 seconds on, then idle) does not meet the conventional three-hour continuous-duty threshold, so the 125% continuous-load factor is not automatically required here; the conductor can likely be sized directly to the motor's actual running current (with any applicable motor-circuit-specific code provisions still applied — motor branch circuits have their own sizing rules, separate from this chapter's general continuous-load treatment). The right response isn't to apply 125% out of habit, nor to skip it out of habit in the other direction — it's to check the actual documented duty cycle against the code's continuous-load definition before deciding which factor, if any, applies.

</details>

<details>
<summary>Design Task 2, Problem 1 (plug-in): derating an AWG 3 conductor at 45C with 5 bundled conductors</summary>

An AWG 3 conductor has a base ampacity of 100 A. At 45°C ambient (correction factor 0.82) with 5 current-carrying conductors bundled together (adjustment factor 0.80), and a required ampacity of 70 A, does it still pass?

**Solution:**

$$I_{allowable} = 100 \times 0.82 \times 0.80 = 65.6\ \text{A}$$

65.6 A < 70 A required — **FAILS**. The next standard size up would need to be checked against the same combined 0.656 derating factor before being accepted.

</details>

<details>
<summary>Design Task 2, Problem 2 (design decision): upsize the conductor, or split the tray?</summary>

A derating calculation like the one above fails. Two fixes are on the table: upsize the conductor to the next standard AWG size, or reroute a portion of the bundled circuits into a second, separate tray to reduce the bundling count (and its adjustment-factor penalty) on the original tray. Which would you recommend, and what would you check first?

**Answer:** Both fixes work, and the right answer depends on what's actually driving the number. If the bundling adjustment factor is the dominant penalty (many current-carrying conductors crowded into one tray) and there's a nearby second tray with spare capacity, splitting the circuits often solves the problem for every circuit in that tray at once, not just the one being checked — a broader fix for the cost of running a jumper or re-strapping cable, versus upsizing copper on potentially several circuits individually. If the ambient temperature correction is the dominant penalty (a genuinely hot area, not a crowding problem), splitting the tray does nothing — the conductor still has to be upsized (or the routing moved away from the heat source entirely, if that's practical). The right first step is exactly the sanity check built into this design task: separate the two derating factors and see which one is actually doing the damage before picking a fix that only addresses one of them.

</details>

<details>
<summary>Design Task 3, Problem 1 (plug-in): a shorter 24VDC run at higher current</summary>

A 24 VDC photoelectric sensor draws 0.12 A over a 45 m one-way run of 18 AWG cable (R = 25.7 Ω/km). Find the voltage drop and percent drop.

**Solution:**

$$V_{drop} = 2 \times 0.12 \times 0.045 \times 25.7 \approx 0.278\ \text{V}$$

$$\%V_{drop} = \frac{0.278}{24} \times 100 \approx 1.16\%$$

Comfortably under the 3% rule of thumb — a short run at low current, unlike the 120 m solenoid example in the chapter body.

</details>

<details>
<summary>Design Task 3, Problem 2 (design decision): upsize the feeder, or add a local voltage boost?</summary>

A 600 V three-phase feeder to a remote outdoor pump motor comes in at 4.2% voltage drop — over the 3% rule of thumb, but the motor's own tolerance (per its nameplate) allows up to 10% under-voltage before performance is meaningfully affected. Two fixes are available: upsize the feeder conductor by two sizes, or accept the 4.2% drop since it's within the motor's actual tolerance. Which would you choose, and what else matters besides the number?

**Answer:** The motor's own tolerance being wide enough to absorb 4.2% is a legitimate data point, not something to dismiss — this is exactly the "3% rule of thumb can be too tight" case the chapter describes. But before accepting the as-built drop, it's worth checking two things the bare percentage doesn't capture: whether other loads share this same feeder or transformer secondary (a voltage sag from a large motor start, on top of an already-marginal 4.2% steady-state drop, could push a shared circuit below acceptable levels for something else on it), and whether the plant plans to add load to this same run later (a 4.2% drop with no headroom today becomes a real problem the day a second pump gets added downstream). If this feeder is dedicated to this one motor with no expansion plans, accepting the 4.2% drop is a defensible, lower-cost decision. If either of those conditions doesn't hold, upsizing the conductor now is cheaper than a second mobilization to fix it later.

</details>

<details>
<summary>Design Task 4, Problem 1 (plug-in): cable type for a dry indoor control panel wiring run</summary>

New overhead wiring is being run in an existing cable tray, entirely indoors, in a climate-controlled electrical room with no washdown, no direct burial, and no VFD involved. Which cable type from the decision framework applies, and why?

**Solution:** Working through the framework: not direct-buried or exposed outdoors (rules out needing TECK90's full mechanical protection), not a VFD output (rules out VFD cable), dry indoor and tray-supported (points to **tray cable**, or AC90 if it weren't specifically tray-routed), and no washdown/chemical exposure (no added jacket requirement). **Tray cable** is the appropriate, most cost-effective choice.

</details>

<details>
<summary>Design Task 4, Problem 2 (design decision): TECK90 or tray cable near forklift traffic?</summary>

A new tray route for control wiring runs along a wall at a height where it will occasionally be within reach of forklift traffic turning a tight corner in the aisle below — not a certainty of impact, but a real, non-trivial risk over the installation's life. Tray cable is the lower-cost default for an indoor tray run; TECK90 costs more but adds real mechanical protection. Which would you specify?

**Answer:** This is a case where the decision framework's default answer (tray cable, since it's an indoor tray run) needs to be overridden by the specific mechanical-exposure risk the framework's first question is actually asking about. A "non-trivial risk over the installation's life" from forklift traffic is exactly the kind of mechanical hazard TECK90's armor exists to handle, and tray cable offers essentially no protection against a physical strike — a single impact could take down every circuit in that section of tray at once, on a run that's expensive and disruptive to re-pull. The safer and, over the life of the installation, likely cheaper specification is **TECK90** for this specific exposed section (even if the rest of the tray run elsewhere in the room uses tray cable), or alternatively physically relocating/protecting the tray route out of the forklift's swing radius if that's feasible — but if the route can't be moved, don't let "it's an indoor tray run" default logic override an obvious, specific physical hazard the framework's first question was designed to catch.

</details>

<details>
<summary>Design Task 5, Problem 1 (plug-in): SCCR check at a different fault current</summary>

A panel's available fault current is 18 kA. A VFD is listed for 65 kA SCCR when protected by its specified fuse, or 22 kA SCCR when protected by a standard breaker. Do both options pass?

**Solution:**

Fuse: 65 kA ≥ 18 kA -> **PASSES**. Breaker: 22 kA ≥ 18 kA -> **PASSES**. Unlike the chapter's 42 kA scenario, both options clear the SCCR gate here, which means the choice between them can now legitimately be made on cost, resettability, and selectivity grounds rather than being forced by a hard safety requirement.

</details>

<details>
<summary>Design Task 5, Problem 2 (design decision): fuse or breaker for a remote, rarely-visited pump station?</summary>

An unmanned outdoor pump station, visited by plant staff roughly once a week, has both fuse and breaker protection options that clear its SCCR requirement. Which would you specify, and why does the remote, rarely-visited nature of the site matter to the decision?

**Answer:** Favor the **breaker** here, and the site's remoteness is the deciding factor, not a side note. A fuse that opens on a nuisance trip or a real fault leaves the pump down until someone notices (on a site visited weekly, that could mean days of unplanned downtime) and then makes a special trip with the correct spare fuse in hand. A breaker lets a technician — or, with the right remote monitoring, even an operator dispatched without special parts — reset it and confirm the pump is back online in minutes once the underlying issue is addressed. This doesn't mean breakers are always the right call at remote sites (a location with genuinely high available fault current might still force a fuse purely on the SCCR gate from Problem 1), but when both options clear that gate, resettability without a special parts run is exactly the criterion that should tip the decision at a low-traffic, hard-to-reach location.

</details>

<details>
<summary>Design Task 6, Problem 1 (plug-in): checking coordination at a lower fault current</summary>

Using the chapter's breaker pair data (downstream instantaneous pickup 800 A at 0.025 s, upstream instantaneous pickup 3200 A at 0.03 s, with longer inverse-time behavior below those pickups), is the pair coordinated at a fault current of 1500 A?

**Solution:** At 1500 A, the downstream 100 A breaker (pickup 800 A) has already entered its instantaneous region and clears in about 0.025 s. The upstream 400 A breaker (pickup 3200 A) has *not* reached its instantaneous region yet at 1500 A, so it's still on its slower inverse-time curve — from the chapter's illustrative data, roughly 1 s at 2000 A, so noticeably longer than 0.025 s at 1500 A. The margin here is large — **COORDINATED** at this current — which is the opposite conclusion from the chapter's 3500 A example. This is exactly why a real coordination study checks the *entire* fault-current range rather than a single assumed value: this same breaker pair is coordinated at 1500 A and uncoordinated at 3500 A.

</details>

<details>
<summary>Design Task 6, Problem 2 (design decision): upgrade the breaker's trip unit, or accept the risk?</summary>

The uncoordinated breaker pair from the chapter's worked example could be fixed by replacing the upstream 400 A breaker with a unit that has an adjustable, time-delayed short-time trip setting (an electronic trip unit) instead of a fixed instantaneous element — at a meaningfully higher purchase cost per breaker. Given that the specific fault current that causes the coordination failure (3500 A) is a realistic but not certain fault magnitude for this bus, is the upgrade worth it?

**Answer:** For a shared main feeding multiple production cells — exactly this chapter's automotive multi-cell scenario — the upgrade is very likely worth it. The cost comparison isn't "breaker A vs. breaker B" in isolation; it's the incremental cost of the electronic trip unit against the expected cost of a full-line stoppage (every cell on that shared main going down, plus the re-homing and restart time described in the Industry Connection above) each time a downstream fault happens to land in the uncoordinated current range. Given that 3500 A sits well within a plausible range for a downstream fault on this size of system — not an exotic edge case — treating the coordination gap as an acceptable residual risk is a harder position to defend than the incremental cost of the upgrade, especially once the upgrade is amortized across every cell the shared main protects, not charged against just one of them.

</details>

## Open-Ended Design Scenario

<details>
<summary>Capstone scenario: CIP skid and recirculation pump for a food and beverage plant</summary>

You are the electrical designer responsible for a new clean-in-place (CIP) skid and its recirculation pump at a food and beverage plant's pasteurizing line. Pull together everything from this chapter into one design package:

- A new electric heater bank on the CIP skid draws 45 A continuous, three-phase, 600 V.
- The heater bank's feeder runs through an equipment room that regularly reaches 42°C ambient (the room also houses the pasteurizer's own heat-generating equipment), sharing a tray with 5 other current-carrying conductors from adjacent skids.
- The recirculation pump is driven by a new VFD, mounted 95 m from the pump motor across the plant floor, at 600 V three-phase, with a nameplate FLA of 34 A and a rated power factor of 0.85.
- Three distinct cable runs are needed: (1) the heater bank's feeder, indoors, dry, tray-routed; (2) the VFD's output run to the pump motor; (3) instrumentation and small power wiring inside the CIP skid's own washdown-exposed enclosure area, which gets hosed down along with the rest of the skid every shift.
- A recent facility short-circuit study puts the available fault current at the branch panel feeding both the heater bank and the VFD at 28 kA.
- The branch panel's main is a 400 A device on a feeder shared with two other CIP skids elsewhere in the plant; the recirculation pump's VFD branch device is rated 70 A.

Work through this chapter's six design tasks for this installation:

1. Select a base conductor size for the heater bank feeder (Design Task 1), then derate it for the actual 42°C/6-conductor tray conditions (Design Task 2). Does the base selection still hold up?
2. Check voltage drop on the VFD's output feeder to the pump motor (Design Task 3b) — does it clear the 3% rule of thumb, and if the pump's VFD keyway/analog speed reference signal ran on a separate 24 VDC pair alongside it, would you apply the same tolerance to that signal circuit?
3. Select the appropriate cable construction for each of the three runs described above (Design Task 4), including any jacket/glanding requirements the washdown-exposed run needs.
4. Using the 28 kA available fault current, evaluate whether a fuse or a breaker is the right choice for the VFD branch circuit's overcurrent protection (Design Task 5) — state what additional information (equipment SCCR ratings) you would need from the VFD manufacturer to finish this check.
5. Given the 400 A shared main and the 70 A VFD branch device, describe what you would need to verify to confirm selective coordination (Design Task 6), and explain in your own words what happens to the other two CIP skids on that shared main if coordination fails.

There's no single correct numeric answer for every input here — different reasonable assumptions about ambient conditions, conductor tables, and device ratings will lead to different (defensible) specifications. The goal is a documented design package, in the style of this chapter's worked examples, that could actually be handed to an electrical contractor for installation and to a plant's engineering department for sign-off.

</details>
