---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Linear Motion Design

```{r setup-06-08, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter, you will be able to:

1. **Calculate** a ball screw's critical (whirling) speed as a function of root diameter, unsupported span, and end-fixity, and **verify** a candidate screw against a required operating speed with margin.
2. **Calculate** a ball screw's Euler column buckling load for its loaded span and **verify** a candidate screw's buckling capacity against the actual axial thrust load with a safety factor.
3. **Calculate** a ball screw's L10 dynamic life from its dynamic load rating, mean applied axial load, and duty cycle, and **evaluate** the result against a target service life.
4. **Calculate** a linear guide carriage's equivalent dynamic load, including how pitch, yaw, and roll moment loads combine with direct force, and **verify** a candidate rail/carriage against an L10 life target.
5. **Select** among lead screw, ball screw, belt drive, and rack-and-pinion actuation using a documented decision framework spanning speed, precision, cost, and self-locking behaviour.

## Introduction

At a Tier 1 supplier's EV battery-pack assembly line in Ingersoll, Ontario, a horizontal gantry axis carries a structural-adhesive dispensing head back and forth across the top of every battery tray before the cover goes on. The axis looks simple from the outside — a motor, a screw, a carriage riding on two rails — but every one of those components was sized against a specific number, and getting any one of those numbers wrong shows up on the plant floor in a very specific way. Size the screw's unsupported span wrong and the shaft starts to whip at operating speed, throwing the dispensing head off its programmed path in the middle of a bead. Undersize the screw's column strength and a momentary snag in the adhesive feed sends the shaft into a buckled bow. Underestimate the load the screw or the rails actually see — especially the moment loads that a cantilevered dispensing head puts on the carriage, which don't show up if you only check the head's static weight — and the axis wears out its ball tracks years before the plant intended to touch it again.

This chapter builds a single gantry axis from the ground up, one governing calculation at a time. Design Task 1 asks how fast the screw can safely turn before it starts to whip. Design Task 2 asks how much axial thrust the same screw can carry before it buckles like a column. Design Task 3 asks how long the screw will actually last, in kilometres of travel, under the axis's real duty cycle. Design Task 4 moves from the screw to the linear rails the carriage rides on, and asks the same life question — but now the load isn't just the carriage's weight, it's the weight *plus* the overturning moments a cantilevered dispensing head creates. Design Task 5 steps back from this specific axis and asks a broader question every linear-motion designer eventually faces: when does a ball screw stop being the right answer at all, and what do you reach for instead — a lead screw, a belt drive, or a rack-and-pinion?

The same four governing calculations in Design Tasks 1 through 4 apply just as directly to a very different kind of linear axis: a precision inspection stage on a defense manufacturing line, where a heavy optical sensor head has to hold position to within a few microns and the documentation behind every sizing calculation has to survive an audit. Design Task 5 picks that scenario up directly.

## Design Task 1: Ball Screw Critical Speed

### Scenario

The mechanical designer responsible for the battery-tray dispensing gantry has a first-pass candidate screw in hand: a 32 mm nominal diameter ball screw with a 10 mm lead, root (minor) diameter 27.4 mm, mounted with the motor-coupling end held in a preloaded angular-contact bearing pair (a **fixed** support — it resists both radial load and the tendency of the shaft to tip) and the far end carried in a single radial bearing (a **supported**, or simple, mounting — it resists radial load only, not tipping). The distance between those two bearing supports, measured along the shaft, is 1.35 m. The process engineer needs the axis to traverse the dispensing head at 0.4 m/s during the bead-laydown move. Before anyone commits to this screw on a purchase order, the first question is simple: can it physically turn fast enough without the shaft itself becoming the problem?

```{r fig-06-08-gantry-layout, echo=FALSE, fig.cap="Side-view schematic of the dispensing gantry axis: a fixed bearing block at the motor end, a supported bearing block at the far end, the ball screw spanning between them, and the carriage (with cantilevered dispensing head) riding the ball nut and the parallel linear rail below it.", fig.align='center', fig.width=11, fig.height=5.5}
## Fixed and supported bearing block positions (schematic, arbitrary units)
x_fixed <- 0.5
x_support <- 9.5
y_shaft <- 2.5

## Carriage position along the span (roughly mid-travel for the illustration)
x_carriage <- 5.5

ggplot() +
  ## Rail (parallel to screw, offset below)
  annotate("segment", x = 0, xend = 10, y = 0.6, yend = 0.6, linewidth = 2, color = "grey55") +
  annotate("text", x = 0.4, y = 0.25, label = "Linear rail", size = 2.8 * diagram_text_scale,
           color = "grey45", hjust = 0) +

  ## Ball screw shaft
  annotate("segment", x = x_fixed, xend = x_support, y = y_shaft, yend = y_shaft,
           linewidth = 2.2, color = "steelblue") +

  ## Fixed bearing block (motor end)
  annotate("rect", xmin = x_fixed - 0.45, xmax = x_fixed + 0.15, ymin = y_shaft - 0.9,
           ymax = y_shaft + 0.9, fill = "#2C3E50") +
  annotate("text", x = x_fixed - 0.15, y = y_shaft + 1.55, label = "Fixed bearing\n(motor end)",
           size = 2.9 * diagram_text_scale, hjust = 0.5) +

  ## Supported bearing block (far end)
  annotate("rect", xmin = x_support - 0.15, xmax = x_support + 0.45, ymin = y_shaft - 0.7,
           ymax = y_shaft + 0.7, fill = "#7F8C8D") +
  annotate("text", x = x_support + 0.15, y = y_shaft + 1.35, label = "Supported bearing\n(far end)",
           size = 2.9 * diagram_text_scale, hjust = 0.5) +

  ## Ball nut / carriage plate
  annotate("rect", xmin = x_carriage - 0.35, xmax = x_carriage + 0.35, ymin = y_shaft - 0.35,
           ymax = y_shaft + 0.35, fill = "#E67E22") +

  ## Cantilevered dispensing head, hanging below and ahead of the carriage
  annotate("segment", x = x_carriage, xend = x_carriage + 0.55, y = y_shaft - 0.35,
           yend = 0.95, linewidth = 2.5, color = "#C0392B") +
  annotate("rect", xmin = x_carriage + 0.3, xmax = x_carriage + 0.85, ymin = 0.75, ymax = 0.95,
           fill = "#C0392B") +
  geom_circle(data = data.frame(x0 = x_carriage + 0.85, y0 = 0.85, r = 0.06),
              aes(x0 = x0, y0 = y0, r = r), fill = "black", color = "black") +
  annotate("text", x = x_carriage + 0.2, y = 0.35, label = "Dispensing head\n(cantilevered)",
           size = 2.7 * diagram_text_scale) +

  ## Unsupported span dimension
  annotate("segment", x = x_fixed, xend = x_support, y = 4.35, yend = 4.35,
           arrow = arrow(length = unit(0.15, "cm"), ends = "both", type = "closed"), linewidth = 0.8) +
  annotate("text", x = (x_fixed + x_support) / 2, y = 4.7,
           label = "L = 1.35 m (bearing-to-bearing span)", size = 3 * diagram_text_scale) +

  coord_fixed(xlim = c(-0.5, 10.5), ylim = c(-0.2, 5.2)) +
  theme_void() +
  labs(title = "Dispensing Gantry Axis — Fixed-Supported Ball Screw Layout") +
  theme(plot.title = element_text(size = 14 * diagram_text_scale, face = "bold", hjust = 0.5))
```

### Design Guidelines and Rules of Thumb

A ball screw shaft spinning between its bearing supports behaves like any other slender rotating shaft: it has a natural lateral (bending) frequency, and if the shaft is driven at or near that frequency, small manufacturing eccentricities get amplified into a growing lateral whip — the same resonance phenomenon that makes an unbalanced drive shaft shudder at a particular speed. The rotational speed at which this resonance becomes a problem is the shaft's **critical speed**. Below it, the screw runs smoothly; as the operating speed approaches it, lateral vibration grows sharply, degrading positioning accuracy and, if left unchecked, damaging the ball nut, the bearings, or the shaft itself.

Two geometric facts about a ball screw matter here, and both cut against the intuition of just "buying a bigger screw." First, the relevant diameter for a stiffness calculation is the shaft's **root (minor) diameter** — the diameter at the base of the ball thread groove — not the nominal outside diameter used to name the screw. The thread groove removes material, and the shaft's bending stiffness (and therefore its critical speed) is governed by the smaller cross-section that's actually left. Second, critical speed falls off with the *square* of the unsupported span: doubling the distance between bearing supports doesn't halve the critical speed, it cuts it to a quarter. A long, slender screw whips at a speed that a short screw of the same diameter would never approach.

End fixity — how rigidly each end of the shaft is held — changes the shape (and therefore the frequency) of the shaft's first bending mode. A **fixed** support (a preloaded angular-contact bearing pair resisting both radial load and moment) holds that end of the shaft rigid against rotation of the shaft's own bending slope; a **supported** (simple) mounting resists radial deflection but lets the shaft's slope rotate freely at that point, the way a shaft resting in a single bearing does; a **free** end is unrestrained entirely (rare for a driven ball screw, but relevant for a stub shaft or an idler configuration). A stiffer combination of end conditions raises the critical speed for the same diameter and span.

**Rule of thumb:** operate a ball screw at no more than 80% of its calculated critical speed, not just below it. This margin exists because the calculated critical speed assumes idealized, perfectly rigid bearing supports and a perfectly straight shaft — real bearing preload degrades slowly with wear, real housings have some compliance, and manufacturing straightness tolerances mean the actual first bending mode is never quite where the idealized calculation puts it. This margin can be tightened toward the calculated value on a short, stiff, well-supported axis where those uncertainties are small, but it should never be loosened past 80% on a long-span axis, where the same percentage error in end-fixity stiffness translates into a much larger absolute error in the calculated critical speed.

No CSA or Ontario-specific standard governs ball screw critical speed rating; it is addressed through the ball screw manufacturer's published engineering data, built on the same Euler-Bernoulli beam theory used in any mechanics-of-materials or mechanical vibrations reference (e.g., Hibbeler, Rao). The international reference for ball screw load ratings and general nomenclature is the ISO 3408 series.

### Governing Calculation

Treating the ball screw shaft as a uniform beam of circular cross-section (using the root diameter as the effective stiffness diameter), the first lateral natural frequency depends on the shaft's bending stiffness, its mass per unit length, its span, and its end-fixity:

$$
\omega_n = K_1 \, \frac{d_r}{L^2} \sqrt{\frac{E}{\rho}}
\qquad\qquad
N_{cr} = \frac{60}{2\pi}\,\omega_n
$$

where $K_1 = \beta_1^2/4$ and $\beta_1^2$ is the standard Euler-Bernoulli first-mode eigenvalue for the shaft's end conditions:

| End condition | $\beta_1^2$ | $K_1 = \beta_1^2/4$ |
|:---|:---:|:---:|
| Fixed – Free (cantilever) | 3.52 | 0.879 |
| Fixed – Supported (this chapter's default) | 15.42 | 3.855 |
| Fixed – Fixed | 22.37 | 5.593 |
| Supported – Supported | 9.87 | 2.467 |

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $d_r$ | Screw root (minor) diameter | m |
| $L$ | Unsupported span, bearing support to bearing support | m |
| $E$ | Elastic modulus of the screw material (bearing-grade steel, $\approx 205$ GPa) | Pa |
| $\rho$ | Density of the screw material ($\approx 7850$ kg/m³ for steel) | kg/m³ |
| $\omega_n$ | First lateral natural (angular) frequency | rad/s |
| $N_{cr}$ | Critical rotational speed | rpm |
| $\beta_1^2$ | End-fixity eigenvalue (first bending mode) | dimensionless |

### Worked Example

```{r critical-speed-06-08, echo=TRUE}
## Material properties (bearing-grade steel)
E <- 205e9    # Pa
rho <- 7850   # kg/m^3

## Candidate screw: 32 mm nominal, 10 mm lead
d_r <- 0.0274      # m, root diameter
L_span <- 1.35     # m, bearing-to-bearing span (fixed-supported)
beta1_sq <- 15.42  # fixed-supported, first bending mode
K1 <- beta1_sq / 4

omega_n <- K1 * (d_r / L_span^2) * sqrt(E / rho)
N_cr <- omega_n * 60 / (2 * pi)

## Required operating speed from the process
v_traverse <- 0.4  # m/s, dispensing traverse speed
lead <- 10          # mm/rev
N_req <- v_traverse * 1000 / lead * 60  # rpm

margin_threshold <- 0.8 * N_cr

cat("Root diameter d_r =", d_r * 1000, "mm\n")
cat("Bearing-to-bearing span L =", L_span, "m\n")
cat("K1 (fixed-supported) =", round(K1, 4), "\n")
cat("Critical speed N_cr =", round(N_cr, 1), "rpm\n")
cat("Required operating speed N_req =", N_req, "rpm\n")
cat("80% margin threshold =", round(margin_threshold, 1), "rpm\n\n")

if (N_req <= N_cr) {
  cat("N_req <= N_cr: raw check PASSES\n")
} else {
  cat("N_req > N_cr: raw check FAILS\n")
}

if (N_req <= margin_threshold) {
  cat("N_req <= 80% of N_cr: margin rule PASSES\n")
} else {
  cat("N_req > 80% of N_cr: margin rule FAILS -- even though the raw check passed\n\n")

  ## Corrective action: step up to the next screw family (40 mm nominal),
  ## keeping the same root-to-nominal diameter ratio (~0.855) and lead
  d_r2 <- 0.0342  # m, 40 mm nominal screw
  omega_n2 <- K1 * (d_r2 / L_span^2) * sqrt(E / rho)
  N_cr2 <- omega_n2 * 60 / (2 * pi)
  margin2 <- 0.8 * N_cr2

  cat("Corrective action: upsize to the 40 mm nominal screw family (d_r =",
      d_r2 * 1000, "mm)\n")
  cat("New critical speed N_cr =", round(N_cr2, 1), "rpm\n")
  cat("New 80% threshold =", round(margin2, 1), "rpm\n")
  if (N_req <= margin2) {
    cat("N_req <=  new 80% threshold: PASSES with margin\n")
  }
}
```

**Sanity check:** the 32 mm candidate's raw check looks fine — 2400 rpm required against a 2828 rpm critical speed is a comfortable-looking 15% margin. It's only the 80% rule of thumb that catches the real problem: 2400 rpm is *above* 80% of 2828 rpm (2262 rpm), meaning this screw would be run closer to resonance than good practice allows, with no cushion for bearing wear or housing compliance over the life of the machine. Stepping up to the 40 mm screw family raises the critical speed enough that the required 2400 rpm now sits comfortably below the 80% threshold. This chapter continues to check the *original* 32 mm candidate in Design Tasks 2 and 3 as well — a real design review runs every check against the candidate part before deciding what changes, rather than stopping at the first failure — but keep in mind that the final selected screw is the 40 mm family, and every other check below only gets easier to pass on the larger shaft.

```{r fig-06-08-critical-speed-curve, echo=FALSE, fig.cap="Critical speed falls off with the square of the unsupported span. The fixed-supported curve (this axis's end condition) is compared against fixed-free and fixed-fixed alternatives; the operating point sits above the 80% margin line at L = 1.35 m for the 32 mm candidate.", fig.align='center', fig.width=10, fig.height=6}
E_fig <- 205e9
rho_fig <- 7850
d_r_fig <- 0.0274

L_seq <- seq(0.6, 2.2, by = 0.02)
compute_ncr <- function(L, K1) {
  omega <- K1 * (d_r_fig / L^2) * sqrt(E_fig / rho_fig)
  omega * 60 / (2 * pi)
}

curve_df <- data.frame(
  L = rep(L_seq, 3),
  N_cr = c(compute_ncr(L_seq, 3.855), compute_ncr(L_seq, 0.879), compute_ncr(L_seq, 5.593)),
  condition = rep(c("Fixed-Supported (this axis)", "Fixed-Free", "Fixed-Fixed"), each = length(L_seq))
)

req_speed <- 2400

ggplot(curve_df, aes(x = L, y = N_cr, color = condition)) +
  geom_line(linewidth = 1.2) +
  annotate("segment", x = 0.6, xend = 2.2, y = req_speed, yend = req_speed,
           linetype = "dashed", color = "grey30", linewidth = 0.8) +
  annotate("text", x = 1.85, y = req_speed + 250, label = "Required speed (2400 rpm)",
           size = 3 * diagram_text_scale, color = "grey30") +
  annotate("point", x = 1.35, y = compute_ncr(1.35, 3.855), size = 4, shape = 4,
           stroke = 1.5, color = "black") +
  annotate("text", x = 1.35, y = compute_ncr(1.35, 3.855) + 350,
           label = "Operating point\n(L = 1.35 m)", size = 2.9 * diagram_text_scale, hjust = 0.4) +
  scale_color_manual(values = c("Fixed-Supported (this axis)" = "steelblue",
                                 "Fixed-Free" = "firebrick", "Fixed-Fixed" = "#2ECC71")) +
  labs(title = "Critical Speed vs. Unsupported Span", x = "Span, L (m)",
       y = "Critical speed (rpm)", color = "End condition") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale),
    legend.text = element_text(size = 10 * diagram_text_scale),
    legend.title = element_text(size = 10.5 * diagram_text_scale)
  )
```

## Design Task 2: Ball Screw Column Buckling

### Scenario

The same candidate screw also has to survive the compressive axial thrust the dispensing carriage puts on it — not just while cruising at constant speed, but during the harder acceleration ramps at each end of a pass. Axial thrust turns the screw into a loaded column over the distance between the fixed (thrust-carrying) bearing and wherever the ball nut sits when the load is applied — in the worst case, when the nut is at its farthest travel position from the fixed end. If that compressive load gets high enough relative to the shaft's slenderness, the screw doesn't yield in compression the way a short, stubby part would — it buckles, bowing sideways well before the material's compressive strength is ever approached.

### Design Guidelines and Rules of Thumb

Column buckling is a stability failure, not a material-strength failure — a slender ball screw can buckle at an axial load far below what the steel itself could carry in simple compression, because the shaft doesn't fail by crushing, it fails by suddenly bowing sideways once the compressive load exceeds a critical threshold set by the shaft's bending stiffness and its **effective** (not physical) unsupported length. That effective length depends on end fixity in exactly the way it did for critical speed, but through a different, related factor: a fixed end genuinely restrains the shaft's rotation at that point, which resists the buckled shape from forming, and shortens the effective length the buckling calculation "sees" relative to the physical span.

The buckling load rises with the *fourth* power of the root diameter (through the shaft's bending stiffness) but falls with the *square* of the effective length — meaning a screw that comfortably resists buckling at a short span can fail well before its rated thrust load if the same shaft is stretched across a much longer axis without any other change. This is exactly the length-sensitivity this chapter returns to in Design Task 5, when a much longer transfer axis rules a single-span ball screw out entirely.

**Rule of thumb:** require a buckling safety factor of at least 3 (calculated buckling load divided by actual maximum axial thrust) for a ball screw column, somewhat lower than the 4-6x factors often used for structural steel columns, because a ball screw's actual operating loads are usually better characterized (known process forces, known acceleration profile) than a structural member's design loads, which must cover a wider range of uncertain load cases. This margin should be raised well above 3 for any axis carrying shock loads — a snag, a jam, an impact — since a sudden load spike sees the column's buckling resistance instantaneously, with none of the gradual warning a slower-developing overload might give.

No CSA or Ontario-specific standard governs ball screw column strength specifically; Euler column theory itself is standard mechanics of materials (any reference such as Hibbeler or Beer & Johnston), and manufacturer engineering catalogs apply it directly to the screw's root diameter and mounting.

### Governing Calculation

$$
P_{cr} = \frac{\pi^2 E I}{(K_e L)^2}
\qquad\qquad
I = \frac{\pi d_r^4}{64}
\qquad\qquad
SF = \frac{P_{cr}}{F_a}
$$

| End condition | $K_e$ |
|:---|:---:|
| Fixed – Free | 2.0 |
| Fixed – Supported (this chapter's default) | 0.699 |
| Fixed – Fixed | 0.5 |
| Supported – Supported | 1.0 |

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $P_{cr}$ | Euler critical buckling load | N |
| $I$ | Second moment of area of the shaft cross-section (root diameter) | m⁴ |
| $K_e$ | Effective-length factor for the end condition | dimensionless |
| $L$ | Loaded (buckling) span — fixed bearing to the farthest nut position | m |
| $F_a$ | Actual maximum axial thrust load | N |
| $SF$ | Buckling safety factor | dimensionless |

### Worked Example

Live cell — try changing `mass_carriage`, `accel`, or `L_buck` and re-run to see the safety factor update.

```{webr-r}
#| label: buckling-06-08
## Material and candidate screw (same 32 mm candidate as Design Task 1)
E <- 205e9    # Pa
d_r <- 0.0274 # m, root diameter

## End fixity: fixed (thrust bearing, motor end) - supported (far end)
Ke <- 0.699

## Loaded (buckling) span: fixed bearing to the farthest ball-nut position,
## slightly less than the full bearing-to-bearing span since the far support
## bearing sits a short distance beyond the nut's extreme travel position
L_buck <- 1.15  # m

I <- pi * d_r^4 / 64
Pcr <- pi^2 * E * I / (Ke * L_buck)^2

## Actual axial thrust: acceleration force + preload/friction drag + process
## reaction, with a 15% pad for real-world uncertainty
mass_carriage <- 42   # kg, carriage plate + nut flange + dispensing head + hoses
accel <- 4.5          # m/s^2, commanded acceleration
F_accel <- mass_carriage * accel
F_friction <- 180      # N, ball nut preload + wiper seal + guide drag
F_process <- 50        # N, dispensing head reaction force
Fa_raw <- F_accel + F_friction + F_process
pad <- 1.15
Fa <- Fa_raw * pad

SF <- Pcr / Fa

cat("Second moment of area I =", format(I, scientific = TRUE), "m^4\n")
cat("Euler critical buckling load P_cr =", format(round(Pcr, 0), big.mark = " "), "N\n\n")
cat("Acceleration force =", F_accel, "N\n")
cat("Friction/preload drag =", F_friction, "N\n")
cat("Process reaction force =", F_process, "N\n")
cat("Raw axial thrust =", Fa_raw, "N ; padded axial thrust F_a =", round(Fa, 1), "N\n\n")
cat("Safety factor SF = P_cr / F_a =", round(SF, 1), "\n\n")

if (SF >= 3) {
  cat("SF >= 3: PASSES the buckling rule of thumb, with large margin\n\n")
} else {
  cat("SF < 3: FAILS -- shorten the loaded span, increase root diameter, or reduce thrust\n\n")
}

## How much longer could this span get before SF erodes to the minimum of 3?
Pcr_target <- 3 * Fa
L_at_SF3 <- sqrt(pi^2 * E * I / Pcr_target) / Ke
cat("Loaded span at which SF would fall to 3:", round(L_at_SF3, 2), "m\n")
```

**Sanity check:** a safety factor around 180 is not a rounding artifact — it is the honest result of a stocky, short-span screw carrying a modest thrust load. Buckling essentially never governs this axis; the length at which the safety factor would erode to the 3x minimum is nearly 9 metres, and long before a single-span ball screw ever reaches that length, critical speed and cost (Design Tasks 1 and 5) have already ruled it out. The lesson is not "skip the buckling check" — it is that buckling only becomes the *controlling* failure mode on long, slender, or heavily thrust-loaded screws, such as a tall vertical Z-axis or a screw carrying a pressing/clamping force, and the only way to know which situation you're in is to run the check.

```{r fig-06-08-buckling-curve, echo=FALSE, fig.cap="Euler buckling load falls off with the square of the loaded span. Even at the operating span of 1.15 m, the buckling load (86.6 kN) dwarfs the actual thrust load (482 N); the safety factor only erodes to the 3x minimum near an 8.9 m span.", fig.align='center', fig.width=10, fig.height=6}
E_fig <- 205e9
d_r_fig <- 0.0274
Ke_fig <- 0.699
I_fig <- pi * d_r_fig^4 / 64
Fa_fig <- 481.85

L_seq <- seq(0.5, 10, by = 0.05)
Pcr_seq <- pi^2 * E_fig * I_fig / (Ke_fig * L_seq)^2
buckling_df <- data.frame(L = L_seq, Pcr = Pcr_seq / 1000)  # kN

ggplot(buckling_df, aes(x = L, y = Pcr)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  annotate("segment", x = 0.5, xend = 10, y = 3 * Fa_fig / 1000, yend = 3 * Fa_fig / 1000,
           linetype = "dashed", color = "firebrick", linewidth = 0.8) +
  annotate("text", x = 7, y = 3 * Fa_fig / 1000 + 6, label = "3x actual thrust (minimum SF line)",
           size = 2.9 * diagram_text_scale, color = "firebrick") +
  annotate("point", x = 1.15, y = pi^2 * E_fig * I_fig / (Ke_fig * 1.15)^2 / 1000,
           size = 4, shape = 4, stroke = 1.5, color = "black") +
  annotate("segment", x = 1.15, xend = 3.0,
           y = pi^2 * E_fig * I_fig / (Ke_fig * 1.15)^2 / 1000,
           yend = 220, linewidth = 0.5, color = "grey40", linetype = "dotted") +
  annotate("text", x = 3.1, y = 220,
           label = "Operating point\n(1.15 m, 86.6 kN)", size = 2.9 * diagram_text_scale, hjust = 0) +
  scale_y_continuous(labels = scales::label_comma(big.mark = " ")) +
  labs(title = "Euler Buckling Load vs. Loaded Span", x = "Loaded span, L (m)",
       y = "Critical buckling load (kN)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

## Design Task 3: Ball Screw L10 Life From Dynamic Load Rating

### Scenario

The screw passes (or, in the 32 mm candidate's case, was already flagged for an upsize in) the speed and buckling checks. The plant's tooling durability specification also requires a minimum service life before the screw or its ball nut assembly needs replacement: at least 25,000 km of cumulative travel, chosen because a screw/nut swap is a manageable few-hour maintenance event the plant schedules around, not a structural teardown. The dispensing axis runs a two-segment duty cycle every cycle — a controlled forward pass while the adhesive bead is laid down, and a faster, unloaded-feeling return pass — and each segment puts a different mean axial load on the screw.

### Design Guidelines and Rules of Thumb

A ball screw's ball-to-raceway contact fatigues under repeated rolling contact stress, exactly like a rolling-element bearing, and the same statistical **L10 life** concept applies: the distance (or number of revolutions) 90% of a population of identical screws is expected to reach or exceed before the first sign of fatigue spalling, under a given constant load. Because it's a rolling-contact fatigue phenomenon, life scales with the *cube* of the ratio between the screw's rated dynamic load capacity and the actual applied load — a small reduction in load capacity relative to the load buys a disproportionately large increase in life, and vice versa.

Most real duty cycles don't apply one constant load; they apply a sequence of different loads over different portions of the travel, the way this axis's forward (loaded) and return (unloaded) passes do. The correct way to combine a variable duty cycle into a single equivalent constant load for the life formula is a distance-weighted **cubic mean load** — the same idea as an RMS average, but with a cube instead of a square, matching the cube-law life relationship.

**Rule of thumb:** treat the calculated L10 life as a starting point, not a design deadline, particularly when the result comes out very large (hundreds of years at the axis's duty cycle, as this worked example will show). The cube-law life formula rewards even a modest load margin so strongly that lightly loaded precision positioning axes routinely calculate out to service lives far beyond anything a plant would ever plan around. When that happens, it means classical ball fatigue is *not* the constraint that will determine the screw's practical replacement interval — preload loss, backlash growth, and contamination ingress from the shop floor environment will replace the screw first, and those are addressed by an inspection and maintenance interval, not by this formula. This rule of thumb inverts on a heavily loaded or shock-loaded axis, where the calculated L10 life can come out uncomfortably close to (or below) the target service life — in that case, the number absolutely is the design deadline, and it should drive the screw selection directly.

On a defense-sector inspection fixture or precision machining application, this calculation (and the assumptions and duty cycle behind it) is exactly the kind of record a program's quality system expects to be able to produce on demand during a customer or Controlled Goods Program audit — the traceable, documented justification for why a specific screw was selected, not just the part number that ended up on the drawing.

No CSA or Ontario-specific standard governs ball screw dynamic load rating and life; the international reference is the ISO 3408 series, which defines the dynamic load rating convention this chapter uses.

### Governing Calculation

$$
F_m = \left(\frac{\sum_i F_i^3\, d_i}{\sum_i d_i}\right)^{1/3}
\qquad\qquad
L_{10} = \left(\frac{C_a}{F_m}\right)^3 \times 10^6 \ \text{revolutions}
$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $F_i$ | Mean axial load during duty-cycle segment $i$ | N |
| $d_i$ | Distance travelled during duty-cycle segment $i$ | mm |
| $F_m$ | Distance-weighted cubic mean axial load | N |
| $C_a$ | Screw's rated dynamic axial load capacity (from datasheet) | N |
| $L_{10}$ | L10 life, in revolutions ($\times$ lead = distance) | rev |

```{r tbl-06-08-duty-cycle, echo=FALSE, message=FALSE}
duty_cycle <- tribble(
  ~Segment, ~Distance_mm, ~Mean_Axial_Load_N, ~Notes,
  "Forward (dispense) pass", 900, 850, "Controlled speed; includes adhesive back-pressure reaction",
  "Return (rapid) pass", 900, 400, "Faster move; carriage inertia and friction only, no process load"
)

duty_cycle %>%
  kbl(caption = "Dispensing axis duty cycle, per cycle (one round trip)",
      col.names = c("Segment", "Distance (mm)", "Mean Axial Load (N)", "Notes")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, background = "#2C3E50", color = "white")
```

### Worked Example

Live cell — try changing `Ca`, the segment loads, or the production rate and re-run to see the life estimate update.

```{webr-r}
#| label: ballscrew-life-06-08
## Candidate screw's rated dynamic axial load capacity (from datasheet)
Ca <- 24500   # N
lead <- 10    # mm/rev

## Duty cycle: forward (loaded) pass and return (unloaded) pass
F1 <- 850; d1 <- 900   # N, mm - forward dispense pass
F2 <- 400; d2 <- 900   # N, mm - return pass

## Distance-weighted cubic mean load
Fm <- ((F1^3 * d1 + F2^3 * d2) / (d1 + d2))^(1/3)

## L10 life in revolutions, then converted to distance
L10_rev <- (Ca / Fm)^3 * 1e6
L10_mm <- L10_rev * lead
L10_km <- L10_mm / 1e6

cat("Cubic mean axial load F_m =", round(Fm, 1), "N\n")
cat("L10 life =", format(L10_rev, scientific = TRUE, digits = 4), "revolutions\n")
cat("L10 life =", format(round(L10_km, 0), big.mark = " "), "km of travel\n\n")

## Production rate and annual travel distance
jph <- 40          # jobs/hour
hrs_day <- 16      # 2-shift operation
days_wk <- 5
wks_yr <- 50
cycles_yr <- jph * hrs_day * days_wk * wks_yr
dist_per_cycle_mm <- d1 + d2
annual_km <- cycles_yr * dist_per_cycle_mm / 1e6

cat("Cycles per year:", format(cycles_yr, big.mark = " "), "\n")
cat("Annual travel distance:", annual_km, "km/year\n\n")

target_km <- 25000
ratio <- L10_km / target_km
cat("Target L10 life:", format(target_km, big.mark = " "), "km\n")
cat("Calculated L10 / target ratio:", round(ratio, 1), "x\n")

if (L10_km >= target_km) {
  cat("Calculated L10 life meets the target: PASSES\n\n")
} else {
  cat("Calculated L10 life falls short of the target: FAILS\n\n")
}

implied_years <- L10_km / annual_km
cat("At this duty cycle, L10 life corresponds to about", round(implied_years, 0),
    "years of continuous operation\n")
```

**Sanity check:** the calculated life clears the 25,000 km target by more than 17x, and translated into years at the plant's actual duty cycle, it works out to roughly 1,500 years. That number is not a mistake — it is the direct, honest consequence of a cube-law life formula applied to a screw whose rated capacity (24.5 kN) is more than 30 times the mean applied load (about 700 N). No one is designing this screw to actually run for 1,500 years; the number's real value is confirming that classical ball fatigue is definitively *not* the constraint on this axis's service life. The plant's actual preventive maintenance plan for this screw should be built around inspection intervals for backlash growth, preload loss, and seal/wiper condition — not around a calculated fatigue date that will never be reached.

## Design Task 4: Linear Guide L10 Life and Moment Loading

### Scenario

The dispensing carriage rides on two parallel linear rails, each carrying two guide blocks — four blocks total, arranged in a rectangle 420 mm apart in the direction of travel and 340 mm apart between the two rails. The cantilevered dispensing head hangs below and ahead of that block pattern's centroid, which means its weight doesn't load the four blocks equally: it creates pitch, roll, and yaw moments on top of the assembly's direct weight, and those moments redistribute load unevenly across the four blocks. The plant's linear guide durability specification calls for a minimum L10 life of 250,000 km — a stricter target than the ball screw's, because swapping a guide rail means realigning the entire gantry beam, a multi-day event the plant wants to avoid for the platform's full production life.

### Design Guidelines and Rules of Thumb

A linear guide carriage's rated dynamic load capacity, like a ball screw's, describes how much *direct* load the carriage can carry for a given L10 life. But a real carriage rarely sees pure direct load — a cantilevered tool, an off-center gripper, or a payload mounted above or ahead of the carriage's rolling-element centroid creates overturning moments about all three axes: **pitch** (rotation about the axis across the direction of travel, from a fore-aft offset), **yaw** (rotation about the vertical axis, from a side-to-side offset in the direction of travel), and **roll** (rotation about the direction of travel itself, from a side-to-side offset perpendicular to travel).

Because a guide block pattern behaves statically like any other rigid body supported on distributed fasteners — the exact same logic used to analyze an eccentrically loaded bolt group — an overturning moment redistributes load between the blocks on either side of the pattern's centroid, adding to the direct load on the most heavily loaded block and subtracting from the least loaded one. The additional force on the worst-loaded block from a given moment is approximately that moment divided by the effective lever arm between the load-sharing pair of blocks resisting it: the front-to-rear block spacing for pitch and yaw, the rail-to-rail spacing for roll. Manufacturer catalogs often publish this same calculation pre-packaged as an "equivalent load factor" specific to a carriage model's internal geometry — the derivation here is what that published factor is actually doing.

**Rule of thumb:** never accept a direct-load-only check ("the carriage only weighs X, well under the rated capacity") as proof that a linear guide selection is adequate. Exactly like the robot end-effector payload check that CG offset degrades, a cantilevered payload on a linear guide can push the *equivalent* load on the worst-loaded block to double the direct load or more, and it's entirely possible to fail an L10 life target at a fraction of the carriage's nameplate capacity once the moments are included. This rule of thumb weakens for compact, well-centered payloads with minimal offset from the block pattern's centroid — for those, direct load really is close to the whole story.

**Rule of thumb:** apply a service (duty) factor $f_w$ of roughly 1.0-1.2 to the equivalent load before computing L10 life, to cover minor shock, vibration, and speed effects the static equivalent-load calculation doesn't otherwise capture. This factor should be pushed toward 1.5-2.0 or higher for guides on axes with genuine shock loading (a mechanical stop, an impact process) — a flat 1.2 badly understates risk on an axis that isn't smoothly, continuously loaded the way this dispensing gantry is.

No CSA or Ontario-specific standard governs linear guide load ratings; the international reference is the ISO 14728 series, which defines the same 50 km reference-distance L10 life convention used across essentially every linear guide manufacturer's catalog (THK, Hiwin, NSK, Bosch Rexroth, Schneeberger, and others).

### Governing Calculation

$$
F_{eq} = \frac{W_{total}}{n} + \frac{M_{pitch}}{2\,l_x} + \frac{M_{roll}}{2\,l_y} + \frac{M_{yaw}}{2\,l_x}
\qquad\qquad
P = f_w\,F_{eq}
\qquad\qquad
L_{10} = \left(\frac{C}{P}\right)^3 \times 50 \ \text{km}
$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $W_{total}$ | Total moving weight (carriage + nut flange + tooling) | N |
| $n$ | Number of guide blocks sharing the load (4 in this layout) | dimensionless |
| $M_{pitch}, M_{roll}, M_{yaw}$ | Applied moments about the block pattern's centroid, from CAD mass properties | N·m |
| $l_x$ | Front-to-rear block spacing (pitch/yaw lever arm) | m |
| $l_y$ | Rail-to-rail spacing (roll lever arm) | m |
| $F_{eq}$ | Equivalent load on the worst-loaded block, before service factor | N |
| $f_w$ | Service (duty) factor | dimensionless |
| $P$ | Equivalent dynamic load used in the life calculation | N |
| $C$ | Guide block's rated dynamic load capacity (from datasheet) | N |
| $L_{10}$ | L10 life, in km of travel | km |

```{r fig-06-08-guide-pattern, echo=FALSE, fig.cap="Top-view schematic of the four-block guide pattern on the dispensing carriage. The dispensing head's weight acts off-centroid, creating pitch and roll moments that redistribute load toward the front-right block (marked as the worst-loaded case).", fig.align='center', fig.width=9, fig.height=7}
lx <- 0.42
ly <- 0.34

## Each block's label is nudged outward, away from the pattern centroid --
## a per-row offset computed from that row's own x/y sign, not a literal
## constant, so this stays a genuine data-driven geom_text() call.
blocks <- data.frame(
  x = c(-lx/2, lx/2, -lx/2, lx/2),
  y = c(ly/2, ly/2, -ly/2, -ly/2),
  label = c("Rear-left", "Front-left", "Rear-right", "Front-right\n(worst-loaded)")
)
blocks$label_y <- blocks$y + ifelse(blocks$y > 0, 0.13, -0.13)
blocks$hj <- ifelse(blocks$x > 0, 0, 1)

ggplot() +
  ## Rails
  annotate("segment", x = -0.95, xend = 0.95, y = ly/2, yend = ly/2, linewidth = 2, color = "grey55") +
  annotate("segment", x = -0.95, xend = 0.95, y = -ly/2, yend = -ly/2, linewidth = 2, color = "grey55") +
  annotate("text", x = -0.9, y = ly/2 + 0.09, label = "Rail 1", size = 2.7 * diagram_text_scale,
           color = "grey45", hjust = 0) +
  annotate("text", x = -0.9, y = -ly/2 - 0.11, label = "Rail 2", size = 2.7 * diagram_text_scale,
           color = "grey45", hjust = 0) +

  ## Carriage plate outline
  annotate("rect", xmin = -lx/2 - 0.08, xmax = lx/2 + 0.08, ymin = -ly/2 - 0.08, ymax = ly/2 + 0.08,
           fill = NA, color = "#2C3E50", linewidth = 1, linetype = "dashed") +

  ## Blocks, each labeled outward from the pattern (away from the centroid)
  geom_point(data = blocks, aes(x = x, y = y), size = 6, shape = 15, color = "#2C3E50") +
  geom_text(data = blocks, aes(x = x, y = label_y, label = label, hjust = hj),
            size = 2.6 * diagram_text_scale) +

  ## Centroid, offset to the left of the "+" mark so it doesn't collide with blocks
  annotate("point", x = 0, y = 0, size = 3, shape = 3, color = "grey40") +
  annotate("text", x = -0.05, y = 0, label = "Block pattern centroid", size = 2.5 * diagram_text_scale,
           color = "grey40", hjust = 1) +

  ## Offset payload CG -- between the centroid and the front-right block, so it
  ## reads as "offset toward that corner" without sitting on top of either mark
  annotate("point", x = lx * 0.3, y = -ly * 0.28, size = 4, color = "#C0392B") +
  annotate("segment", x = lx * 0.3, xend = 0.62, y = -ly * 0.28, yend = -0.55,
           linewidth = 0.5, color = "#C0392B", linetype = "dotted") +
  annotate("text", x = 0.65, y = -0.55, label = "Dispensing head CG\n(offset forward + right)",
           size = 2.6 * diagram_text_scale, color = "#C0392B", hjust = 0) +

  ## Lever arm dimensions
  annotate("segment", x = -lx/2, xend = lx/2, y = -ly/2 - 0.28, yend = -ly/2 - 0.28,
           arrow = arrow(length = unit(0.12, "cm"), ends = "both", type = "closed"), linewidth = 0.6) +
  annotate("text", x = 0, y = -ly/2 - 0.35, label = "l_x = 0.42 m", size = 2.7 * diagram_text_scale) +
  annotate("segment", x = lx/2 + 0.22, xend = lx/2 + 0.22, y = ly/2, yend = -ly/2,
           arrow = arrow(length = unit(0.12, "cm"), ends = "both", type = "closed"), linewidth = 0.6) +
  annotate("text", x = lx/2 + 0.38, y = 0, label = "l_y =\n0.34 m", size = 2.7 * diagram_text_scale) +

  coord_fixed(xlim = c(-1.1, 1.1), ylim = c(-0.75, 0.6)) +
  theme_void() +
  labs(title = "Four-Block Guide Pattern — Top View") +
  theme(plot.title = element_text(size = 14 * diagram_text_scale, face = "bold", hjust = 0.5))
```

### Worked Example

Live cell — try changing `C_candidate`, the moment values, or `fw` and re-run to see the life result update.

```{webr-r}
#| label: guide-life-06-08
## Block pattern geometry (carriage plate design, not a vendor-specific number)
lx <- 0.42   # m, front-to-rear block spacing
ly <- 0.34   # m, rail-to-rail spacing
n_blocks <- 4

## Total moving weight: carriage plate + nut flange + dispensing head + hoses
mass_total <- 15  # kg
W_total <- mass_total * 9.81
F_direct <- W_total / n_blocks

## Applied moments about the block pattern centroid, from CAD mass properties
M_pitch <- 22  # N.m
M_roll <- 9    # N.m
M_yaw <- 4     # N.m

d_pitch <- M_pitch / (2 * lx)
d_roll <- M_roll / (2 * ly)
d_yaw <- M_yaw / (2 * lx)

F_eq <- F_direct + d_pitch + d_roll + d_yaw

## Service factor
fw <- 1.2
P_eq <- fw * F_eq
P_naive <- fw * F_direct   # what a direct-load-only (wrong) check would use

cat("Direct load share per block:", round(F_direct, 2), "N\n")
cat("Pitch contribution:", round(d_pitch, 2), "N\n")
cat("Roll contribution:", round(d_roll, 2), "N\n")
cat("Yaw contribution:", round(d_yaw, 2), "N\n")
cat("Equivalent load on worst-loaded block, F_eq =", round(F_eq, 1), "N\n")
cat("(", round(F_eq / F_direct, 2), "x the direct load alone)\n\n")
cat("With service factor f_w =", fw, ": P =", round(P_eq, 1), "N\n\n")

## First-pass candidate guide block
C_candidate <- 1450  # N, rated dynamic load capacity
L10_km <- (C_candidate / P_eq)^3 * 50
L10_naive_km <- (C_candidate / P_naive)^3 * 50

target_km <- 250000

cat("Candidate guide dynamic capacity C =", C_candidate, "N\n")
cat("Naive (direct-load-only) L10 =", format(round(L10_naive_km, 0), big.mark = " "), "km",
    "-- looks like it clears the target\n")
cat("Real L10 (moments included) =", format(round(L10_km, 0), big.mark = " "), "km\n")
cat("Target L10 life =", format(target_km, big.mark = " "), "km\n\n")

if (L10_km >= target_km) {
  cat("Real L10 meets the target: PASSES\n")
} else {
  cat("Real L10 falls short of the target: FAILS -- even though the naive check looked fine\n\n")

  ## Corrective action: upsize to the next guide block size
  C_upsized <- 2600  # N
  L10_upsized_km <- (C_upsized / P_eq)^3 * 50
  cat("Corrective action: upsize to a", C_upsized, "N-rated guide block\n")
  cat("New L10 =", format(round(L10_upsized_km, 0), big.mark = " "), "km\n")
  if (L10_upsized_km >= target_km) {
    cat("New L10 meets the target: PASSES\n")
  }
}
```

**Sanity check:** the moment contributions more than double the equivalent load on the worst-loaded block (from about 37 N direct to about 81 N once pitch, roll, and yaw are included) — and that difference is exactly what separates a first-pass candidate that looks like it clears the target by 7x (the naive check) from one that actually falls short by a third (the real check). This is precisely the same trap as the robot end-effector CG-offset payload check: the nameplate/direct-load number understates the true demand, and the moment loading is what closes the gap. Upsizing to the next guide block size restores a comfortable margin above the 250,000 km target.

## Design Task 5: Lead Screw vs. Ball Screw vs. Belt Drive vs. Rack-and-Pinion Selection

### Scenario

Three different linear axes are on the table across two facilities this quarter, and each one needs an actuation technology decision before detailed design can start:

1. **The dispensing gantry's horizontal traverse** — this chapter's own axis, already validated in Design Tasks 1 through 4.
2. **A 3.5 m horizontal transfer axis** in a different area of the same automotive plant, moving large stamped sheet-metal panels between two press-line stations. Positioning tolerance is loose (±1 mm is fine), but cycle time is tight and the payload is substantial.
3. **A precision Z-axis on a defense-sector optical inspection fixture**, positioning a heavy sensor head to verify machined feature locations on a structural aerospace component. The specification calls for sub-10-micron repeatability, and — because the fixture sits in a manned inspection cell — the axis must hold its position if electrical power is lost, without relying on the sensor head's own weight to hold it (a fail-safe requirement).

Not every axis should use the same actuation technology, and picking the right one for each requires weighing the same four criteria against very different priorities.

### Design Guidelines and Rules of Thumb

Four actuation technologies cover the overwhelming majority of industrial linear motion applications, and each one earns its place through a different combination of speed, precision, cost, and back-drivability.

```{r tbl-06-08-selection-matrix, echo=FALSE, message=FALSE}
selection_matrix <- tribble(
  ~Technology, ~Typical_Speed, ~Precision_Repeatability, ~Relative_Cost, ~Self_Locking,
  "Lead screw (Acme/trapezoidal)", "Low (sliding friction limits speed and generates heat)",
  "Moderate (tens of microns with an anti-backlash nut; wear degrades this over time)",
  "Lowest -- simple thread form, low tooling cost",
  "Often yes, if lead angle < friction angle (efficiency roughly < 50%)",
  "Ball screw", "High (rolling contact; limited mainly by critical speed and DN value)",
  "High (single-digit to tens of microns, preloaded, per ISO 3408 accuracy grades)",
  "Moderate-to-high -- cost rises sharply with unsupported length",
  "No -- rolling contact is efficient (~90%+) and back-drives freely",
  "Timing belt drive", "Highest -- minimal reciprocating mass, well suited to long travel",
  "Moderate -- limited by belt stretch/compliance without closed-loop correction",
  "Low-to-moderate -- cost scales gently with travel length, unlike a screw",
  "No -- low-friction system, back-drives freely",
  "Rack-and-pinion", "High", "Moderate-to-high -- helical, preloaded dual-pinion setups improve this substantially",
  "Moderate -- rack + gearbox cost scales with length, less steeply than a screw",
  "Generally no, unless a high-ratio worm-gear pinion stage is used"
)

selection_matrix %>%
  kbl(caption = "Linear actuation technology comparison across speed, precision, cost, and self-locking behaviour",
      col.names = c("Technology", "Typical Speed", "Precision / Repeatability", "Relative Cost", "Self-Locking / Back-Drivable")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, background = "#2C3E50", color = "white")
```

Two structural facts drive most of these differences. First, a ball screw's cost and practical span both degrade sharply with unsupported length — Design Tasks 1 and 2 showed exactly why: critical speed collapses with the square of the span, and while buckling capacity usually stays comfortable, the screw's diameter (and therefore its cost and reflected inertia) has to grow to keep critical speed acceptable on a long axis. A belt or rack-and-pinion drive doesn't have this problem in the same way — a belt's "shaft" is a flexible loop with no whirling mode to speak of, and a rack's pinion only ever spans the length of one small gear, no matter how long the rack itself runs. **Rule of thumb:** treat roughly 3-4 m of unsupported span as the practical upper limit for a single ball screw axis before a belt or rack-and-pinion becomes both cheaper and mechanically simpler. This threshold moves lower on a fast axis (critical speed bites sooner) and can push somewhat higher on a slow, short-duty-cycle axis where critical speed margin is easy to maintain even at length.

Second, self-locking is a property of sliding friction, not rolling contact — which is exactly why a lead screw can hold a vertical load with the motor de-energized while a ball screw, a belt, and (usually) a rack-and-pinion cannot. A power screw's thread is self-locking when the friction angle at the thread interface exceeds the thread's lead angle; below that threshold, an applied axial load can back-drive the screw and turn the motor shaft in reverse, exactly the way an efficient (low-friction) gear train back-drives easily. **Rule of thumb:** never treat self-locking friction alone as a certified safety holding mechanism for a load that could injure someone if it moved — friction coefficients change with lubrication condition, wear, and contamination, and a self-locking screw that held reliably when new can back-drive years later as the thread surface wears. Where a genuine fail-safe holding requirement exists, pair it with a positively engaged mechanical or electrical brake, not friction alone.

### Governing Calculation

For axes 1 and 2, the decision comes down to reading the table above against each application's dominant constraint (precision for a tight process, cost and speed for a long, loose-tolerance transfer, and so on) — a documented decision framework, not a single formula. For the self-locking behaviour specifically, though, there is a real governing calculation:

$$
\lambda = \arctan\left(\frac{L}{\pi d_m}\right)
\qquad\qquad
\eta = \frac{\tan\lambda}{\tan(\lambda + \phi)}
\qquad\qquad
\phi = \arctan(\mu)
$$

Self-locking occurs when $\mu \geq \tan\lambda$ (equivalently, when $\eta < 50\%$, a useful quick cross-check).

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $L$ | Screw lead | mm |
| $d_m$ | Thread mean (pitch) diameter | mm |
| $\lambda$ | Thread lead angle | degrees |
| $\mu$ | Coefficient of friction at the thread interface | dimensionless |
| $\phi$ | Friction angle | degrees |
| $\eta$ | Power screw efficiency | dimensionless |

### Worked Example

Live cell — try changing `L_lead`, `d_mean`, or `mu` and re-run to see whether the thread stays self-locking.

```{webr-r}
#| label: selflock-06-08
## Candidate lead screw for a self-locking application
L_lead <- 5     # mm, lead
d_mean <- 19    # mm, mean (pitch) diameter of the Acme thread
mu <- 0.15      # coefficient of friction, lightly lubricated Acme thread

lambda_rad <- atan(L_lead / (pi * d_mean))
lambda_deg <- lambda_rad * 180 / pi
tan_lambda <- tan(lambda_rad)

phi_rad <- atan(mu)
eta <- tan_lambda / tan(lambda_rad + phi_rad)

cat("Lead angle, lambda =", round(lambda_deg, 2), "degrees (tan =", round(tan_lambda, 4), ")\n")
cat("Coefficient of friction, mu =", mu, "\n")
cat("Efficiency, eta =", round(eta * 100, 1), "%\n\n")

if (mu >= tan_lambda) {
  cat("mu >= tan(lambda): thread is SELF-LOCKING\n")
} else {
  cat("mu < tan(lambda): thread is NOT self-locking -- will back-drive under axial load\n")
}

## Ball screw comparison (typical rolling-contact efficiency, not derived)
eta_ballscrew <- 0.90
cat("\nFor comparison, a preloaded ball screw typically runs at about",
    eta_ballscrew * 100, "% efficiency -- far too efficient to be self-locking.\n")
```

**Sanity check:** at a 5 mm lead on a 19 mm mean diameter, the thread's lead angle is under 5 degrees, well inside the friction angle for a lightly lubricated Acme thread (efficiency comes out to about 35%, comfortably under the 50% self-locking rule-of-thumb threshold). The ball screw's ~90% efficiency, by contrast, isn't a design flaw — it's the entire reason ball screws dominate wherever speed and precision matter — but it means a ball screw can never provide fail-safe holding on its own.

```{r tbl-06-08-scenario-recommendations, echo=FALSE, message=FALSE}
recommendations <- tribble(
  ~Axis, ~Dominant_Requirement, ~Recommendation, ~Reasoning,
  "1. Dispensing gantry traverse", "Precision dispensing path, 900 mm span, validated in Tasks 1-4",
  "Ball screw (as designed)",
  "Span is well inside the practical ball screw range; precision and speed both favour rolling contact over sliding friction",
  "2. 3.5 m panel transfer axis", "Long span, heavy payload, loose tolerance, fast cycle",
  "Rack-and-pinion",
  "Span exceeds the practical single ball screw limit (Design Tasks 1-2); rack-and-pinion carries the heavier payload more stiffly than a belt would at this length, at a cost that scales gently with span",
  "3. Defense inspection Z-axis", "Sub-10-micron repeatability AND fail-safe hold on power loss",
  "Ball screw + fail-safe motor brake",
  "Precision requirement rules out a lead screw; since the ball screw itself is not self-locking, the fail-safe holding requirement must be met with an added spring-applied, power-released brake rather than by the screw's friction alone"
)

recommendations %>%
  kbl(caption = "Selection decisions for three linear axes",
      col.names = c("Axis", "Dominant Requirement", "Recommendation", "Reasoning")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = knitr::is_html_output()) %>%
  row_spec(0, background = "#2C3E50", color = "white") %>%
  column_spec(1, width = "1.2in") %>%
  column_spec(2, width = "1.6in") %>%
  column_spec(3, width = "1.3in") %>%
  column_spec(4, width = "2.2in")
```

**Sanity check:** notice that Axis 3's answer isn't a single row lookup in the comparison table — no single technology satisfies both "sub-10-micron precision" and "self-locking" at once, so the decision has to split the requirement across two components (a precise-but-not-self-locking screw, plus an added brake) rather than force-fitting one technology to do both jobs. That's the real skill this design task is teaching: the table narrows the field, but the final call is still a judgment call.

## Summary

```{r tbl-06-08-summary, echo=FALSE, message=FALSE}
summary_table <- tribble(
  ~Design_Task, ~Governing_Relationship, ~Key_Rule_of_Thumb,
  "1. Ball screw critical speed", "$N_{cr} = \\frac{60}{2\\pi}K_1\\frac{d_r}{L^2}\\sqrt{E/\\rho}$",
  "Operate at or below 80% of N_cr, not just below it",
  "2. Ball screw column buckling", "$P_{cr} = \\pi^2EI/(K_eL)^2$",
  "Require a buckling safety factor of at least 3x actual thrust",
  "3. Ball screw L10 life", "$L_{10} = (C_a/F_m)^3 \\times 10^6$ rev",
  "Treat an enormous calculated L10 as proof fatigue isn't the limiting factor, not a literal service date",
  "4. Linear guide L10 with moments", "$L_{10} = (C/(f_w F_{eq}))^3 \\times 50$ km",
  "Never trust a direct-load-only check; moments can double the equivalent load or more",
  "5. Actuation technology selection", "Decision framework (speed / precision / cost / self-locking)",
  "Single ball screw spans beyond ~3-4 m favour belt or rack-and-pinion instead"
)

summary_table[] <- lapply(summary_table, tex_safe)

summary_table %>%
  kbl(caption = "Design task summary: governing relationships and key rules of thumb", escape = FALSE,
      col.names = c("Design Task", "Governing Relationship", "Key Rule of Thumb")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, background = "#2C3E50", color = "white")
```

## Practice Problems

<details>
<summary>Task 1 (plug-in): A candidate ball screw has root diameter 21.5 mm and a fixed-supported span of 0.95 m. Using the same steel properties (E = 205 GPa, rho = 7850 kg/m^3) and K1 = 3.855, what is its critical speed? If the required operating speed is 2600 rpm, does it pass the 80% margin rule?</summary>

$\omega_n = 3.855 \times (0.0215 / 0.95^2) \times \sqrt{205\text{e}9/7850} = 3.855 \times 0.02382 \times 5110.25 \approx 469.3$ rad/s

$N_{cr} = 60/(2\pi) \times 469.3 \approx 4483$ rpm

80% of $N_{cr}$ = 3586 rpm. Since 2600 rpm < 3586 rpm, this **passes** the margin rule comfortably (2600 rpm is only about 58% of the critical speed).

</details>

<details>
<summary>Task 1 (design decision): A layout engineer proposes solving a critical-speed failure by simply commanding a lower traverse speed instead of changing the screw or its mounting. The process engineer objects that this will blow the station's cycle time. Whose position is more defensible, and what would you check before picking a side?</summary>

Both positions have a legitimate basis, and the right answer depends on numbers neither has stated yet. Slowing the traverse speed is the cheapest possible fix — no new hardware, no re-qualification of a different screw — but it only works if the resulting cycle time still fits the station's rate requirement; if it doesn't, the "fix" simply trades one failure (critical speed) for another (throughput), and the axis is no better off. Before picking a side, check whether the required speed reduction (from Design Task 1's numbers, roughly 2400 rpm down to under 2262 rpm, less than a 6% cut) actually threatens the line rate at all — a small reduction might be well within the cycle time's existing slack, in which case the layout engineer's fix is not just cheaper but essentially free. If the station has no slack, the process engineer is right that a hardware change (a shorter span via an intermediate support, or a larger-diameter screw as in the worked example) is the only way to keep both the required speed and the margin.

</details>

<details>
<summary>Task 2 (plug-in): A candidate screw has P_cr = 42,000 N and sees an actual axial thrust of 9,500 N. Does it pass the SF >= 3 rule of thumb, and what is the actual safety factor?</summary>

$SF = 42000 / 9500 \approx 4.42$

Since 4.42 > 3, this **passes** the rule of thumb, though with considerably less margin than the chapter's worked example.

</details>

<details>
<summary>Task 2 (design decision): A vertical Z-axis (gravity acting along the screw's own axis) is being sized for a heavy tooling head, and the buckling safety factor comes out to 2.1 -- below the 3x rule of thumb. Three fixes are proposed: (1) increase the root diameter, (2) add an intermediate shaft-support bearing to shorten the loaded span, (3) simply accept SF = 2.1 since "it still technically holds the load." Which would you recommend, and why is option 3 riskier on this axis than it might be on a horizontal one?</summary>

Option 3 should be rejected outright, and it's riskier here specifically because this is a *vertical* axis: on a horizontal axis, a buckling failure bends the screw sideways within roughly its own weight and the process loads already at play, but on a vertical axis carrying a heavy tooling head, a buckling event can let that head drop under gravity the moment the screw loses its column stability -- a safety hazard, not just an equipment failure. Between options 1 and 2, increasing root diameter (option 1) raises buckling capacity with the fourth power of diameter, a strong lever, but it also changes the ball nut and possibly the bearing bore, a bigger hardware and cost change. Adding an intermediate support (option 2) shortens the loaded span with a $1/L^2$ effect and often needs only a support bracket, not a new screw -- typically the lower-cost, lower-disruption fix, provided the support hardware doesn't block the ball nut's travel path (a real support-rail product, not an improvised solution). The recommended path is to evaluate option 2 first for exactly that reason, escalating to option 1 only if geometry doesn't allow an intermediate support.

</details>

<details>
<summary>Task 3 (plug-in): A screw has Ca = 18,000 N and a duty cycle of F1 = 600 N over 500 mm and F2 = 250 N over 500 mm. What is the cubic mean load, and the L10 life in revolutions?</summary>

$F_m = \left(\frac{600^3 \times 500 + 250^3 \times 500}{1000}\right)^{1/3} = \left(\frac{108{,}000{,}000{,}000 + 7{,}812{,}500{,}000}{1000}\right)^{1/3} \approx 487.7 \text{ N}$

Working the arithmetic: $600^3 = 216{,}000{,}000$; $250^3 = 15{,}625{,}000$. Weighted equally (500/500): mean of cubes $= (216{,}000{,}000 + 15{,}625{,}000)/2 = 115{,}812{,}500$. $F_m = 115{,}812{,}500^{1/3} \approx 487.7$ N.

$L_{10} = (18000/487.7)^3 \times 10^6 = (36.91)^3 \times 10^6 \approx 50{,}280 \times 10^6 \approx 5.03 \times 10^{10}$ revolutions.

</details>

<details>
<summary>Task 3 (design decision): A calculated L10 life comes out to only 8,000 km against a 25,000 km target -- a genuine, meaningful shortfall, unlike this chapter's worked example. Three options are on the table: (1) select a screw with a larger dynamic load rating, (2) reduce the mean load by lightening the carriage or reducing acceleration, (3) reduce the target itself by shortening the planned replacement interval. Which would you pursue first, and what would make you reconsider?</summary>

Because life scales with the cube of the capacity-to-load ratio, both options 1 and 2 are genuinely powerful levers -- a modest increase in either direction produces an outsized life improvement. Option 2 (reducing the mean load) is usually worth investigating first precisely because of that cube sensitivity: if the carriage can be lightened by even 15-20% (removing unnecessary structure, relocating a heavy component closer to the drive axis) or acceleration can be trimmed slightly without hurting cycle time, the life gain can be large enough to close an 8,000-to-25,000 km gap without changing the screw at all -- the lowest-cost, no-new-part-number fix. Option 1 (a bigger screw) is the reliable fallback if the load can't be reduced enough, but it's a bigger design change (potentially different nut, different mounting envelope). Option 3 (shrinking the target) should be the last resort, not the first move -- it doesn't fix anything, it just relabels the shortfall as acceptable, and should only be considered if the plant's maintenance team confirms a shorter replacement interval is genuinely tolerable from a cost and downtime standpoint, not as an easy way out of a failing calculation.

</details>

<details>
<summary>Task 4 (plug-in): A guide block sees a direct load share of 45 N, plus moment contributions of 30 N (pitch), 18 N (roll), and 6 N (yaw). With a service factor of 1.15 and a candidate block rated at C = 2,100 N, what is the L10 life in km?</summary>

$F_{eq} = 45 + 30 + 18 + 6 = 99$ N

$P = 1.15 \times 99 = 113.85$ N

$L_{10} = (2100/113.85)^3 \times 50 = (18.45)^3 \times 50 \approx 6{,}280 \times 50 \approx 314{,}000$ km

</details>

<details>
<summary>Task 4 (design decision): Two fixes are proposed for a guide block that fails its L10 target once moments are included: (1) upsize to a larger guide block (higher C), or (2) redesign the mounting bracket to pull the payload's CG closer to the block pattern centroid, reducing the moments directly. Which fixes the problem more fundamentally, and when might the "easy" fix (option 1) actually be the wrong long-term choice?</summary>

Option 2 fixes the problem more fundamentally, because it attacks the actual cause -- the moment arms driving up the equivalent load -- rather than just buying more capacity to absorb the same oversized load. Reducing the CG offset (a shorter or better-balanced bracket, or repositioning a heavy component like a valve or transformer closer to the carriage) shrinks pitch and roll contributions directly, and because the equivalent load feeds into a cube-law life formula, even a modest offset reduction can restore a large life margin. Option 1 (a bigger guide block) is the faster fix and sometimes the only practical one if the payload's CG genuinely can't move for process reasons (the tool has to reach where it has to reach), but it's the wrong long-term choice when the same underlying design (an unnecessarily long cantilever, an avoidably heavy bracket) gets reused on a future, still-heavier variant of the same platform -- at that point the moment problem returns and demands an even bigger guide, chasing the symptom instead of the cause. The general principle: fix the geometry when the application allows it; upsize the hardware when the geometry is fixed by the process.

</details>

<details>
<summary>Task 5 (plug-in): A lead screw candidate has a 4 mm lead and a 16 mm mean diameter, with mu = 0.12 (well-lubricated Acme thread). Is it self-locking?</summary>

$\lambda = \arctan(4 / (\pi \times 16)) = \arctan(0.0796) \approx 4.55$ degrees

$\tan(\lambda) \approx 0.0796$

Since $\mu = 0.12 \geq \tan(\lambda) = 0.0796$, this thread **is self-locking**.

</details>

<details>
<summary>Task 5 (design decision): The 3.5 m panel transfer axis (Axis 2 from the worked example) is re-evaluated after a design change: the payload drops from a full stamped panel (heavy) to a lightweight plastic tray, and the plant now wants sub-0.1 mm precision for a new vision-guided pick step added to that station. Does the original rack-and-pinion recommendation still hold, and if not, what changes?</summary>

The rack-and-pinion recommendation was driven specifically by *heavy payload plus loose tolerance* at a long span -- both of those conditions just changed. With a lightweight payload, a belt drive's main weakness at this length (higher compliance under heavy load) is far less of a concern, and belts are typically the lower-cost, higher-speed option once payload stiffness stops being the deciding factor. The new sub-0.1 mm precision requirement, however, is tighter than an open-loop timing belt reliably delivers on its own (belt stretch and compliance typically limit repeatability to several tenths of a millimetre without closed-loop correction) -- so the real answer depends on whether the vision-guided pick step tolerates a closed-loop linear-encoder-corrected belt axis (which can reach the required precision at low cost) or genuinely needs the open-loop repeatability of a rack-and-pinion or ball screw. This is exactly the kind of question the decision framework is built to raise, not resolve by table lookup alone -- the recommendation should be revisited with the automation team once it's clear whether closed-loop position correction is already part of the vision system's plan for this axis.

</details>

## Open-Ended Design Scenario

<details>
<summary>Capstone scenario: complete linear motion design review for the dispensing gantry axis</summary>

You are the mechanical designer responsible for releasing the battery-tray dispensing gantry axis from this chapter to production. Pull together everything from this chapter into one design review, using the exact values from the worked examples unless a task states otherwise:

- **Candidate screw:** 32 mm nominal, 10 mm lead, root diameter 27.4 mm, fixed-supported span 1.35 m, required traverse speed 0.4 m/s.
- **Buckling:** loaded span 1.15 m, carriage mass 42 kg, acceleration 4.5 m/s², friction/preload drag 180 N, process reaction 50 N, 15% uncertainty pad.
- **Ball screw life:** Ca = 24,500 N, forward pass 850 N over 900 mm, return pass 400 N over 900 mm, 40 JPH, 2-shift/5-day/50-week operation, 25,000 km target.
- **Linear guide life:** block pattern 420 mm x 340 mm, 15 kg moving mass, moments M_pitch = 22 N·m, M_roll = 9 N·m, M_yaw = 4 N·m, service factor 1.2, 250,000 km target, first-pass candidate C = 1,450 N.
- **A cost-cutting proposal:** the program manager, trying to hold the tooling budget, asks whether the plant could skip the 40 mm screw upsize from Design Task 1 and instead just run the traverse move slower, and separately asks whether the linear guide upsize from Design Task 4 is really necessary given how comfortably the ball screw life and buckling checks passed.

Work through all five design tasks and reach a clear conclusion for each: does the candidate hardware pass, and if not, what specific change do you make? Then answer the program manager's two questions directly, using the numbers from your own analysis rather than intuition -- does slowing the traverse move actually solve the critical speed problem without a hardware change, and does the ball screw's enormous life margin have any bearing on whether the linear guide needs its own upsize? Finally, write two or three sentences on why the two structural findings in this chapter -- critical speed collapsing with span, and buckling capacity remaining enormous at the same span -- point toward *different* upper limits for how long a single ball screw axis can practically be, and how Design Task 5's ~3-4 m rule of thumb reconciles them.

</details>
