---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Control Panel Design

```{r setup-06-03, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter, you will be able to:

1. **Calculate** wire duct fill percentage from conductor cross-sectional area and verify it against an industry fill limit, and **apply** minimum bend radius rules of thumb to control panel wiring.
2. **Select** an appropriate NEMA/IP enclosure rating for a given operating environment, using a documented decision framework and a NEMA-to-IP cross-reference.
3. **Calculate** a control panel's internal temperature rise from summed component heat loss, enclosure surface area, and material heat-transfer coefficient, and **decide** between passive, forced-air, and air-conditioned cooling.
4. **Size** motor branch circuit conductors, overload protection, and time-delay fuse or breaker protection per CEC Section 28 (NEC Article 430 equivalent), including the 125% continuous-load rule.
5. **Verify** an assembled panel's short-circuit current rating (SCCR) using the weakest-link principle, and **select** components (or protective combinations) to meet a target SCCR.
6. **Evaluate** whether a control panel requires UL 508A / CSA C22.2 No. 286 panel shop certification or an Ontario Electrical Safety Authority (ESA) field evaluation before it can be legally energized.

## Introduction

A controls integrator in Cambridge, Ontario has just been awarded the electrical package for a new press-line transfer system at a Tier 1 automotive stamping plant. The bill of materials is already on the desk: three variable frequency drives, a dozen contactors, a safety PLC, a control power transformer, terminal blocks, and a stack of I/O modules. The customer's electrical specification is four pages long, and buried in it are the sentences that actually drive the design: *"Panel shall be rated 65 kA SCCR minimum. Enclosure shall be UL 508A listed or CSA C22.2 No. 286 certified. Branch circuit protection per NEC Article 430 / CEC Section 28."* None of that tells the designer what size enclosure to buy, how to route the wiring inside it, or how to keep it from cooking itself alive once it's closed up and running eight VFDs' worth of heat into a sealed steel box.

This is the part of the project that looks like paperwork until it isn't. A panel that's the wrong physical size for its wire duct fails an inspection before it ever gets powered up. A panel with the wrong enclosure rating lets water into an IP54 box in a washdown zone and takes down a production line six months later. A panel that runs 15°C hotter than its design margin allows shortens the life of every VFD inside it, quietly, for years, until one fails during a shift and nobody can say why. A panel assembled from a 65 kA main breaker and a handful of 5 kA contactors, with no current-limiting protection tying them together, is *not* a 65 kA panel — no matter what's printed on the breaker's own nameplate. And a panel built without the right shop certification, or without an ESA field evaluation when it needs one, simply isn't allowed to be connected to the building's electrical system in Ontario, full stop, regardless of how good the engineering inside it is.

This chapter works through six design tasks that a controls engineer or panel designer actually has to close out, in roughly the order a real project encounters them: physical layout and wire duct sizing, enclosure environmental rating, internal heat management, motor branch circuit protection, panel-level short-circuit current rating, and the certification pathway that determines whether the finished panel is even legal to plug in. Several of these tasks lean on numbers that come from other chapters in this part — motor losses feeding the heat calculation come from the kind of drive sizing covered in the [Motor, Drive, and Gearbox Sizing](06-06-Motor-Drive-Gearbox-Sizing.qmd) chapter, and the wire segregation practice here connects directly to the bonding, grounding, and EMC discipline covered elsewhere in this part. Here, those numbers are treated as known inputs, the way a panel designer receives them from a drive sizing sheet and has to turn them into a physical box, a duct layout, and a certification file.

## Design Task 1: Enclosure Sizing, Wire Duct Fill, and Bend Radius

### Scenario

The transfer system's control panel needs to house eighteen 16 AWG I/O conductors, eight 14 AWG relay coil and pilot device conductors, and four 12 AWG 24 VDC power distribution taps, all routed through a single wire duct run between the terminal strip and the I/O rack. The panel shop's layout drawing needs a duct size specified before the enclosure can be released for fabrication — pick a duct too small, and the shop either can't physically get the wires in, or does anyway and creates a service nightmare the first time a technician has to add a wire during commissioning.

### Design Guidelines and Rules of Thumb

Wire duct (also called wireway or raceway, depending on the region and the product family) exists to keep field wiring organized, protected, and serviceable — but a duct that's packed full defeats all three purposes at once. A technician troubleshooting a fault needs to be able to physically get a hand or a meter probe into the duct without wrestling a solid brick of wire; a duct with room to spare lets that happen, and a duct at its limit does not. There's also a real thermal reason to avoid overfilling: conductors bundled tightly together can't shed heat to the surrounding air as effectively as conductors with clearance around them, so a badly overfilled duct can run hotter than a lightly filled one carrying the same current.

**Rule of thumb:** keep wire duct fill at or below 40% of the duct's usable cross-sectional area, and prefer something closer to 25-30% if the panel is expected to grow (a common expectation on a first-of-its-kind cell, where the customer will add I/O points during commissioning). This 40% figure mirrors the philosophy behind conduit fill limits used elsewhere in the electrical trade, though it is fundamentally a panel-shop workmanship practice — driven by UL 508A's expectations for wireway construction and general industry norms — rather than a specific numeric limit written into the CEC or NEC themselves. This rule of thumb breaks down for ducts carrying only a handful of large power conductors rather than many small ones: a duct at 35% fill with three heavy motor leads has plenty of physical slack even though the percentage looks similar to a duct at 35% fill with sixty small control wires, because the failure mode being guarded against (an impossible-to-service tangle of many wires) doesn't really apply to three large ones.

Conductors and cables also have a minimum bend radius, below which the insulation can crack, a shield can be damaged, or — for multi-conductor cable — the individual conductors inside can shift and stress their terminations. **Rule of thumb:** allow a minimum bend radius of about 4-6 times the cable's outside diameter (OD) for standard unshielded multi-conductor control cable, and 8-10 times OD for shielded cable, such as VFD output (motor) cable, where crushing or kinking the shield at a tight bend can compromise the very EMI shielding the cable was specified for in the first place. These are rules of thumb, not a substitute for the cable manufacturer's data sheet, which states the actual minimum bend radius for that specific product — and they break down for cable specifically designed for tight-radius applications (some drag-chain-rated cable is built to tolerate much tighter bends than a standard cable of the same OD), where following the generic multiplier would be needlessly conservative.

Physical segregation matters as much as duct sizing: power wiring (motor leads, VFD output cable) and control/signal wiring (I/O, communication network cable, safety circuit wiring) should run in separate ducts or maintain a documented minimum separation where they must cross, because a VFD's switching-frequency output carries enough high-frequency noise to induce errors on a nearby low-level signal conductor. This segregation practice is covered in more depth in the bonding, grounding, and EMC design task elsewhere in this part; the layout habit starts here, at the duct-sizing stage. CEC Section 12 addresses conductor installation practices including bending, and CSA C22.2 No. 18.1 covers wireway construction; NEC Article 366 and UL 508A's wireway/wiring space requirements are the equivalent US references.

### Governing Calculation

For a duct carrying $n_i$ conductors of type $i$, each with insulated outside diameter $d_i$:

$$
A_{cond,i} = \pi \left(\frac{d_i}{2}\right)^2
$$

$$
A_{total} = \sum_i n_i \, A_{cond,i}
$$

$$
Fill\% = \frac{A_{total}}{A_{duct}} \times 100\%
$$

$$
R_{bend,min} = k_{bend} \times OD
$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $d_i$ | Insulated outside diameter of one conductor of type $i$ | mm |
| $A_{cond,i}$ | Cross-sectional area of one conductor of type $i$ | mm² |
| $n_i$ | Number of conductors of type $i$ in the duct run | -- |
| $A_{total}$ | Total cross-sectional area of all conductors in the duct | mm² |
| $A_{duct}$ | Usable (fingered/partitioned) cross-sectional area of the duct, not the raw nominal width x height | mm² |
| $Fill\%$ | Wire duct fill percentage | % |
| $OD$ | Outside diameter of a cable | mm |
| $k_{bend}$ | Bend radius multiplier: ~4-6 for unshielded multi-conductor cable, ~8-10 for shielded cable | -- |
| $R_{bend,min}$ | Minimum allowable bend radius for that cable | mm |

```{r fig-06-03-duct-fill, echo=FALSE, fig.cap="Wire duct fill percentage for the same conductor bundle routed through three candidate duct sizes, against the 40% rule-of-thumb fill limit.", fig.align='center', fig.width=9, fig.height=6}
## Same total conductor area as the worked example below, compared across
## three standard duct sizes to show why the duct choice matters
total_conductor_area_mm2 <- 134.1  # mm^2

duct_options <- data.frame(
  duct_size = c("25 x 25 mm\n(1 x 1 in)", "40 x 25 mm\n(1.5 x 1 in)", "40 x 40 mm\n(1.5 x 1.5 in)"),
  nominal_area_mm2 = c(25 * 25, 40 * 25, 40 * 40)
)
duct_options$duct_size <- factor(duct_options$duct_size, levels = duct_options$duct_size)

## Usable area is less than the nominal width x height because of the duct's
## finger walls/dividers -- a representative fraction, not a specific
## manufacturer's published spec
usable_fraction <- 0.48
duct_options$usable_area_mm2 <- duct_options$nominal_area_mm2 * usable_fraction
duct_options$fill_pct <- total_conductor_area_mm2 / duct_options$usable_area_mm2 * 100

ggplot(duct_options, aes(x = duct_size, y = fill_pct)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_text(aes(label = paste0(round(fill_pct, 1), "%")), vjust = -0.6, size = 3.6 * diagram_text_scale) +
  annotate("segment", x = 0.5, xend = 3.5, y = 40, yend = 40,
           linetype = "dashed", color = "firebrick", linewidth = 0.9) +
  annotate("text", x = 3.5, y = 44, label = "40% fill limit", color = "firebrick",
           hjust = 1, size = 3 * diagram_text_scale) +
  labs(title = "Wire Duct Fill by Candidate Duct Size",
       x = "Duct size (nominal external dimensions)", y = "Fill (%)") +
  ylim(0, 55) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

### Worked Example

```{r wire-duct-fill-06-03, echo=TRUE}
## Conductors sharing one duct run between the terminal strip and I/O rack
n_16awg <- 18   # I/O wiring
n_14awg <- 8    # relay coil / pilot device wiring
n_12awg <- 4    # 24VDC power distribution taps

## Representative insulated outside diameters for stranded MTW/TEW-rated
## machine tool wire (consult the specific wire manufacturer's data sheet
## for exact figures on a real job)
od_16awg <- 2.2  # mm
od_14awg <- 2.5  # mm
od_12awg <- 2.9  # mm

conductor_area <- function(od) pi * (od / 2)^2

a16 <- conductor_area(od_16awg)
a14 <- conductor_area(od_14awg)
a12 <- conductor_area(od_12awg)

total_conductor_area <- n_16awg * a16 + n_14awg * a14 + n_12awg * a12

cat("Conductor area (mm^2): 16 AWG =", round(a16, 2),
    " 14 AWG =", round(a14, 2), " 12 AWG =", round(a12, 2), "\n")
cat("Total conductor area =", round(total_conductor_area, 1), "mm^2\n\n")

## Usable duct cross-sectional area accounts for the duct's finger
## walls/dividers, not just the raw nominal width x height
usable_fraction <- 0.48

check_duct <- function(name, w_mm, h_mm) {
  nominal_area <- w_mm * h_mm
  usable_area <- nominal_area * usable_fraction
  fill_pct <- total_conductor_area / usable_area * 100
  cat(name, ": nominal", w_mm, "x", h_mm, "mm, usable area =",
      round(usable_area, 0), "mm^2, fill =", sprintf("%.1f", fill_pct), "%")
  if (fill_pct <= 40) cat("  -- PASS\n") else cat("  -- FAILS 40% limit\n")
  invisible(fill_pct)
}

cat("--- Candidate duct sizes ---\n")
fill_25x25 <- check_duct("25 x 25 mm duct", 25, 25)
fill_40x25 <- check_duct("40 x 25 mm duct", 40, 25)

cat("\nSelected duct: 40 x 25 mm, fill =", sprintf("%.1f", fill_40x25),
    "% -- comfortably under 40%, with room for the panel to grow.\n\n")

## Minimum bend radius check for the VFD output (motor) cable at the duct's
## 90-degree corner
vfd_cable_od <- 12       # mm, shielded VFD output cable
k_bend_shielded <- 8     # rule-of-thumb multiplier for shielded cable
required_bend_radius <- k_bend_shielded * vfd_cable_od

planned_corner_radius <- 60  # mm, the duct fitting specified on the layout

cat("VFD output cable OD =", vfd_cable_od, "mm, shielded\n")
cat("Required minimum bend radius (", k_bend_shielded, "x OD ) =",
    required_bend_radius, "mm\n")
cat("Planned duct corner fitting radius =", planned_corner_radius, "mm\n")
if (planned_corner_radius >= required_bend_radius) {
  cat("PASS -- corner radius meets the minimum bend radius\n")
} else {
  cat("FAILS -- corner radius is too tight for this shielded cable\n")
  cat("Corrective action: route the VFD output cable through a wider-sweep\n")
  cat("elbow or a separate raceway sized for its bend radius, rather than\n")
  cat("forcing it through the standard 60 mm duct corner fitting.\n")
}
```

**Sanity check:** the 40 x 25 mm duct's fill percentage is well under 40% while still being the second-smallest standard size available — jumping straight to a much larger duct "to be safe" would waste panel space without buying much practical benefit. The bend radius check catches a real, easy-to-miss failure mode: a duct that passes its fill check with room to spare can still fail if a corner fitting is too tight for a specific shielded cable, since fill percentage and bend radius are independent checks governed by different physics (crowding versus insulation/shield stress).

<details>
<summary>Industry connection: why segregation matters more on a VFD-heavy panel</summary>

On an automotive body-shop panel running a dozen VFDs in one enclosure, keeping the power wireway (VFD output cable to the motors) physically separated from the control wireway (safety network cable, low-level I/O) isn't a nice-to-have — drive switching noise coupled onto a safety network cable can produce intermittent diagnostic faults that are brutal to troubleshoot on a running line, because they don't reproduce reliably during a quiet maintenance window.

</details>

## Design Task 2: NEMA vs. IP Enclosure Rating Selection

### Scenario

The same integrator is quoting three different panels for three different customers: a general assembly area panel for the automotive stamping plant, an outdoor-adjacent panel for a tank farm, and a control panel for a dairy processing line that sits directly in a washdown zone where the sanitation crew hoses down the equipment with hot water and detergent at the end of every shift. One enclosure rating does not fit all three.

### Design Guidelines and Rules of Thumb

NEMA enclosure ratings (from NEMA 250, the US National Electrical Manufacturers Association standard) and IP (Ingress Protection) ratings (from IEC 60529, the international standard) both describe how well an enclosure keeps things out — but they are not the same test, and treating them as directly interchangeable is a common and costly mistake. IP ratings describe *only* ingress protection against solids (first digit) and liquids (second digit) under specific, narrowly defined test conditions. NEMA ratings describe ingress protection too, but also fold in requirements that IP doesn't test at all — corrosion resistance, gasket aging, ice formation resistance (for outdoor types), and oil/coolant resistance for some types. The cross-reference table below is the industry-conventional approximate mapping, not a certified equivalence — a panel genuinely required to meet a specific NEMA type needs to be tested and rated to that NEMA type, not just to an IP number that looks similar on paper.

```{r tbl-06-03-nema-ip-crossref, echo=FALSE, message=FALSE}
nema_ip <- tribble(
  ~NEMA_Type, ~Approx_IP_Equivalent, ~Typical_Environment, ~Example_Use,
  "NEMA 1", "IP20", "Indoor, general purpose, no dust or water exposure expected", "Office-adjacent equipment rooms",
  "NEMA 12", "IP54", "Indoor, dust, dirt, and light dripping non-corrosive liquids, no washdown", "General industrial floor panels (the automotive plant's main assembly area)",
  "NEMA 3R", "~IP24", "Outdoor, rain and sleet, not dust-tight", "Outdoor disconnect switches, yard lighting panels",
  "NEMA 4", "IP66", "Indoor/outdoor, hosedown-rated, watertight, painted/mild steel", "Wash-down-adjacent equipment where corrosion resistance isn't the primary concern",
  "NEMA 4X", "IP66 + corrosion resistance", "Indoor/outdoor, hosedown-rated, corrosion-resistant material (stainless steel, fiberglass, or polycarbonate)", "Food/beverage plants, chemical-adjacent areas, marine/coastal sites",
  "NEMA 12/IP54 (typical general panel default)", "IP54", "Baseline for most indoor industrial panels with no special exposure", "Machine control panels in a dry manufacturing area",
  "No direct NEMA equivalent", "IP69K", "High-pressure, high-temperature washdown (per DIN 40050-9 test conditions)", "Direct food-contact-adjacent equipment in dairy, meat, and beverage processing"
)

nema_ip %>%
  kbl(caption = "Approximate NEMA-to-IP enclosure rating cross-reference and typical applications",
      col.names = c("NEMA Type", "Approx. IP Equivalent", "Typical Environment", "Example Use")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, background = "#2C3E50", color = "white")
```

IP69K deserves special attention because it has **no direct NEMA equivalent at all** — it comes from a German standard (DIN 40050-9) originally written for the automotive industry's own high-pressure washdown testing, and it has since become the de facto requirement for any enclosure that sits in a high-pressure, high-temperature washdown zone: think hot water at high pressure, sprayed at close range, several times a day, for the life of the equipment. A food plant that specifies "NEMA 4X" alone, without also specifying IP69K, may receive a stainless enclosure that survives a garden-hose-style washdown (the NEMA 4/4X hosedown test) but degrades under the much more aggressive high-pressure, high-temperature spray a sanitation crew actually uses.

**Rule of thumb:** for washdown areas, specify **both** a NEMA type (for material/corrosion resistance) and an IP rating (for the specific ingress severity), rather than relying on one system alone — e.g., "NEMA 4X, IP69K" for a direct food-contact-adjacent panel, versus "NEMA 4X, IP66" for a food-plant panel that's corrosion-resistant but sits somewhere the high-pressure wand never directly reaches. This rule of thumb breaks down for indoor, dry, general-purpose panels, where specifying both systems is needless duplication — NEMA 12 alone (or IP54 alone) is the accepted shorthand almost everywhere in general industrial work, and asking a supplier to certify both for a panel that will never see water is just added cost for no benefit.

CSA C22.2 No. 94.1 and No. 94.2 are the Canadian enclosure standards (harmonized with, but not textually identical to, NEMA 250 and IEC 60529); a Canadian-market panel should be specified and marked against the CSA standard, with NEMA/IP cited as the equivalent US/international reference where the customer's specification uses that language, which is extremely common on automotive OEM specifications written centrally in the US.

### Governing Calculation (Decision Framework)

There is no single formula for enclosure rating selection — it is a documented decision framework built from the answers to a short set of environmental questions:

1. **Is the panel indoors or outdoors?** Outdoor exposure alone (with no washdown) points toward NEMA 3R/4 minimum, for rain, sleet, and temperature cycling.
2. **Is there any washdown, hosedown, or high-pressure spray exposure?** None -> NEMA 12/IP54 is typically sufficient. Occasional hosedown -> NEMA 4/IP66. Routine high-pressure, high-temperature washdown (food/beverage direct-contact-adjacent) -> NEMA 4X **and** IP69K together.
3. **Is the environment corrosive** (washdown chemicals, salt air, acidic/caustic cleaning agents, coastal or marine air)? If yes, the "X" suffix (corrosion-resistant construction: stainless steel, fiberglass, or polycarbonate) is required regardless of the ingress answer above.
4. **Is there conductive or combustible dust, or a hazardous (classified) location present?** This chapter does not cover hazardous-location (explosion-proof) enclosures, but a "yes" here means NEMA 250's Type 7/9 (or the IEC/CSA hazardous-location equivalents) apply instead of the ordinary-location types discussed here, and a specialist review is required.

### Worked Example

Live cell — try changing the environment inputs for a fourth panel and re-run to see the recommended rating change.

```{webr-r}
#| label: nema-ip-select-06-03
## Three candidate panels for three different sites
panels <- data.frame(
  panel = c("Automotive plant - general assembly area",
            "Tank farm - outdoor, no routine washdown",
            "Dairy processing line - direct washdown zone"),
  outdoor = c(FALSE, TRUE, FALSE),
  routine_washdown = c(FALSE, FALSE, TRUE),
  high_pressure_hot_washdown = c(FALSE, FALSE, TRUE),
  corrosive_or_food_environment = c(FALSE, FALSE, TRUE),
  stringsAsFactors = FALSE
)

for (i in seq_len(nrow(panels))) {
  p <- panels[i, ]
  cat("Panel:", p$panel, "\n")

  if (p$high_pressure_hot_washdown) {
    rating <- "NEMA 4X, IP69K"
    reason <- paste("routine high-pressure, high-temperature washdown demands the",
                     "DIN 40050-9 IP69K test condition, which NEMA 4X alone does not cover;",
                     "corrosion-resistant (stainless or polycarbonate) construction is required.")
  } else if (p$routine_washdown || p$corrosive_or_food_environment) {
    rating <- "NEMA 4X, IP66"
    reason <- "occasional hosedown and/or a corrosive/food-adjacent environment justify a corrosion-resistant, hosedown-rated enclosure."
  } else if (p$outdoor) {
    rating <- "NEMA 4 (or 3R if dust-tightness is not required), IP66"
    reason <- "outdoor exposure to rain and temperature cycling requires at least a hosedown-rated or rain-tight enclosure."
  } else {
    rating <- "NEMA 12, IP54"
    reason <- "indoor, dry, general industrial environment -- the standard baseline rating."
  }

  cat("  Recommended rating:", rating, "\n")
  cat("  Reasoning:", reason, "\n\n")
}
```

**Sanity check:** the dairy line panel and the general assembly panel land on very different ratings from the same decision framework because the framework is driven by the environment, not by "how important" the panel is — a mission-critical automotive panel in a dry area still only needs NEMA 12, while a comparatively simple dairy-line control station needs NEMA 4X/IP69K because of where it physically sits, which is exactly the point of separating the decision from intuition.

<details>
<summary>Industry connection: the cost of getting this wrong in a food plant</summary>

A CFIA-regulated dairy plant that installs a NEMA 12-rated (IP54) panel in a washdown zone isn't just risking equipment failure — a compromised enclosure becomes a moisture trap behind a control panel door, exactly the kind of hidden bacterial growth site a sanitary design audit is built to catch, and a finding like that can hold up a plant's food safety certification renewal, not just its production schedule.

</details>

## Design Task 3: Internal Heat Dissipation and Cooling Method Selection

### Scenario

The same transfer system panel is nearly laid out: three 7.5 kW VFDs, a control power transformer, eight contactors, and a PLC rack are all going into one free-standing 800 mm (W) x 1000 mm (H) x 300 mm (D) painted mild steel enclosure. The plant floor runs at 35°C ambient in summer, and the drives are only rated to operate reliably up to 50°C internal ambient. Before the panel is released for fabrication, someone has to confirm the box won't simply cook its own contents once the door is closed and the drives are running.

### Design Guidelines and Rules of Thumb

Every component inside a control panel that draws current dissipates some fraction of that power as heat — a VFD's semiconductor switching losses, a transformer's core and winding losses, a contactor's coil holding power, a power supply's conversion inefficiency. None of that heat disappears; it has to leave the enclosure through the enclosure's own surface, by conduction through the metal and convection to the surrounding air (or, with active cooling, by moving air directly). A sealed enclosure with no forced airflow only has its own outer surface area to work with, which is why enclosure heat dissipation is fundamentally a surface-area problem, not just a "how much heat is inside" problem — the same heat load in a small, tightly packed enclosure runs far hotter than in a large, roomy one.

**Rule of thumb:** for a first-pass estimate, only count the enclosure's front, back, two sides, and top as effective heat-dissipating surface (a free-standing enclosure), or drop the back face entirely for a wall-mounted enclosure (it isn't exposed to moving air). Never count the bottom face — it typically faces down into still air near the floor or a support structure and contributes little. This rule of thumb breaks down for enclosures mounted tightly between other equipment or inside a machine base, where even the "free" faces don't see much air movement; a cramped installation needs a more conservative (lower) effective heat-transfer coefficient than the open-floor assumption used here.

Once the passive (natural convection) capacity is exceeded, the next step up is a **filtered fan** — a fan draws in outside air, forces it through the enclosure past the hot components, and exhausts it, dramatically increasing the effective heat transfer compared to natural convection alone. Fans only help, though, if the incoming air is actually cooler than the desired internal temperature and clean enough not to foul the components; a filtered fan also breaches the enclosure's ingress-protection seal, which rules it out entirely for a washdown-rated (IP66/IP69K) enclosure like the dairy-line panel from the previous design task. Where the ambient air is too hot, too dirty, or the enclosure must remain sealed, the next step up is an **air conditioner** (or, for a lighter load, a sealed air-to-air heat exchanger) — a closed-loop device that rejects heat to the outside air via a refrigeration cycle without ever exchanging enclosure air with the (unsuitable) ambient air.

**Rule of thumb:** a passive (unventilated) enclosure is usually only viable for heat loads on the order of a few hundred watts in a mid-sized panel; loads climbing into the low kilowatts, like this VFD-heavy panel, routinely need forced ventilation, and loads combined with a sealed or extreme-ambient requirement push toward air conditioning. This is a rough sizing intuition, not a substitute for the calculation below — always run the numbers for the specific enclosure and heat load rather than guessing from load magnitude alone.

### Governing Calculation

$$
Q_{total} = \sum_j P_{loss,j}
$$

$$
\Delta T = \frac{Q_{total}}{k \times A_{eff}}
$$

$$
V_{flow} = \frac{Q_{total}}{\rho \, c_p \, \Delta T_{allow}}
$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $P_{loss,j}$ | Heat loss of component $j$ (VFD, transformer, contactor, power supply, etc.) | W |
| $Q_{total}$ | Total internal heat load | W |
| $A_{eff}$ | Effective heat-dissipating surface area (see rule of thumb above) | m² |
| $k$ | Heat-transfer coefficient for the enclosure material under natural convection (~5.5 W/(m²·°C) is a representative value for painted mild steel; stainless and fiberglass differ) | W/(m²·°C) |
| $\Delta T$ | Calculated internal temperature rise above ambient | °C |
| $\Delta T_{allow}$ | Allowable temperature rise (max. component-rated internal ambient minus site ambient) | °C |
| $\rho$ | Air density (~1.2 kg/m³ at typical plant conditions) | kg/m³ |
| $c_p$ | Specific heat of air (~1005 J/(kg·K)) | J/(kg·K) |
| $V_{flow}$ | Required forced-air volumetric flow rate to hold the enclosure at $\Delta T_{allow}$ | m³/s |

```{r fig-06-03-heat-load, echo=FALSE, fig.cap="Heat load contribution by component category for the transfer system panel. VFD switching losses dominate the total, which is typical for a drive-heavy panel.", fig.align='center', fig.width=9, fig.height=6}
heat_components <- data.frame(
  component = c("VFDs (x3)", "Transformer", "Contactors", "PLC supply", "Misc./terminals"),
  loss_w = c(675, 50, 24, 40, 20)
)
heat_components$component <- factor(heat_components$component,
                                     levels = heat_components$component[order(heat_components$loss_w)])

ggplot(heat_components, aes(x = component, y = loss_w)) +
  geom_col(fill = "darkorange", width = 0.6) +
  geom_text(aes(label = paste0(loss_w, " W")), hjust = -0.15, size = 3.4 * diagram_text_scale) +
  coord_flip(ylim = c(0, 780)) +
  labs(title = "Heat Load by Component Category",
       x = NULL, y = "Heat loss (W)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

### Worked Example

Live cell — try changing the enclosure dimensions or `k`, and re-run to see how the passive/fan/AC decision changes.

```{webr-r}
#| label: heat-load-06-03
## Component heat losses (W)
vfd_count <- 3
vfd_rated_kw <- 7.5
vfd_loss_fraction <- 0.03   # ~3% of rated output as switching/conduction loss
vfd_loss_each <- vfd_rated_kw * 1000 * vfd_loss_fraction
vfd_loss_total <- vfd_count * vfd_loss_each

transformer_loss <- 50   # W, 1 kVA control transformer at ~5% loss
contactor_loss <- 8 * 3  # W, 8 contactors x 3 W coil holding power each
plc_supply_loss <- 40    # W
misc_loss <- 20          # W

Q_total <- vfd_loss_total + transformer_loss + contactor_loss + plc_supply_loss + misc_loss

cat("VFD loss (3 x", vfd_loss_each, "W ) =", vfd_loss_total, "W\n")
cat("Transformer loss =", transformer_loss, "W\n")
cat("Contactor coil loss =", contactor_loss, "W\n")
cat("PLC power supply loss =", plc_supply_loss, "W\n")
cat("Misc./terminal loss =", misc_loss, "W\n")
cat("Total heat load, Q_total =", Q_total, "W\n\n")

## Enclosure: free-standing painted mild steel, W x H x D
W <- 0.8  # m
H <- 1.0  # m
D <- 0.3  # m

A_eff <- 2 * (H * W) + 2 * (H * D) + (W * D)  # front+back, 2 sides, top (no bottom)
k <- 5.5  # W/(m^2.C), representative for painted mild steel, natural convection

delta_T <- Q_total / (k * A_eff)

cat("Effective dissipating surface area, A_eff =", round(A_eff, 2), "m^2\n")
cat("Heat-transfer coefficient, k =", k, "W/(m^2.C)\n")
cat("Calculated passive temperature rise, delta_T =", round(delta_T, 1), "C\n\n")

ambient_C <- 35
max_internal_C <- 50
delta_T_allow <- max_internal_C - ambient_C

cat("Site ambient =", ambient_C, "C, max. allowable internal ambient =", max_internal_C, "C\n")
cat("Allowable temperature rise, delta_T_allow =", delta_T_allow, "C\n\n")

if (delta_T <= delta_T_allow) {
  cat("Passive (natural convection) cooling: PASS -- no fan or AC required.\n")
} else {
  cat("Passive cooling FAILS (", round(delta_T, 1), "C >", delta_T_allow, "C allowed ) --\n")
  cat("evaluating forced-air ventilation next.\n\n")

  rho <- 1.2     # kg/m^3, air density
  cp <- 1005     # J/(kg.K), specific heat of air

  V_flow_m3s <- Q_total / (rho * cp * delta_T_allow)
  V_flow_m3h <- V_flow_m3s * 3600
  V_flow_cfm <- V_flow_m3h / 1.699

  cat("Required forced-air flow to hold delta_T_allow =", round(V_flow_m3h, 0),
      "m^3/h (", round(V_flow_cfm, 0), "CFM )\n")

  ## Rule of thumb: oversize the fan selection to cover filter loading and
  ## elevation derating over the equipment's service life
  fan_oversize_factor <- 1.25
  V_flow_selected <- V_flow_m3h * fan_oversize_factor
  cat("With a 25% oversize allowance for filter loading/derating:",
      round(V_flow_selected, 0), "m^3/h required from the selected fan\n\n")

  cat("Decision: this is a general-area panel (NEMA 12/IP54, not washdown-rated),\n")
  cat("so a filtered intake/exhaust fan pair is a viable option -- select a\n")
  cat("standard fan module rated at or above the oversized airflow figure.\n")
  cat("(Note: if this were the dairy-line IP69K panel from the previous design\n")
  cat("task, a filter fan would breach the enclosure seal and would not be an\n")
  cat("option -- a sealed air-to-air heat exchanger or air conditioner would be\n")
  cat("required instead, regardless of how modest the airflow number is.)\n")
}
```

**Sanity check:** a calculated 60°C passive rise against a 15°C allowance is a large, unambiguous fail — a useful reminder that a VFD-heavy panel almost never survives on natural convection alone, matching the rule of thumb above. The required forced-air flow (roughly 160-200 m³/h once oversized) is well within the range of an off-the-shelf filtered fan module, so the fix here is a straightforward part selection, not a redesign — but that fix only works because this particular panel doesn't need a sealed enclosure; a washdown-rated version of the same panel would need a fundamentally different (and more expensive) cooling approach for the exact same heat load.

## Design Task 4: Motor Branch Circuit Protection

### Scenario

One of the transfer system's servo-driven axes is backed up by a conventional 30 hp, 600 V, three-phase induction motor on the material-handling conveyor feeding the press line. The branch circuit feeding that motor — conductors, overload protection, and short-circuit protection — needs to be sized before the panel's power distribution can be finalized.

### Design Guidelines and Rules of Thumb

Motor branch circuits are protected in **two separate stages**, and conflating them is one of the most common mistakes a student (or a rushed designer) makes. The **overload relay** (or an electronic overload function built into a motor starter or VFD) protects the motor itself from sustained overload — running hot for too long — and is set close to the motor's actual full-load amps (FLA), typically 115-125% of FLA depending on the motor's service factor and rated temperature rise. The **branch-circuit protective device** (a fuse or breaker) protects the conductors and equipment from short-circuit and ground-fault currents, and is deliberately sized much larger than FLA — often 175% or more — because it has to ride through the motor's own starting inrush without opening, something the overload relay is not asked to do (the overload relay's thermal memory is sized specifically to tolerate that same starting surge without tripping on a normal start).

A motor's starting (locked-rotor) current is far higher than its running FLA — a NEMA Design B motor typically draws somewhere in the range of 5-7 times FLA for the first several seconds of a start, decaying as the motor comes up to speed. A standard (non-time-delay) fuse sized close to FLA would blow on every single start; a **time-delay (dual-element) fuse** is built specifically to tolerate that inrush for the duration of a normal start while still opening fast enough to protect the circuit under an actual fault.

**Rule of thumb:** for a fuse-protected motor branch circuit, a time-delay fuse rated at roughly 150-175% of FLA (rounded up to the next standard fuse size) is the usual starting point, sized larger only if the specific motor's starting characteristics (a high-inertia load, a long acceleration time) demand it. This rule of thumb breaks down for motors with unusually long or heavy starts — a large flywheel load or a motor started under full load rather than no-load can need a larger multiplier or a different protection scheme (soft starter or VFD current limiting) entirely, and simply oversizing the fuse further is not a safe substitute for addressing the starting profile itself.

CEC Section 28 (Motors and Generators) governs motor branch circuit conductor sizing, overload protection, and short-circuit protection in Canada, as adopted and amended by the OESC; NEC Article 430 is the direct US equivalent, with broadly similar percentage-based rules though the specific tables and some terminology differ. Note the Canadian convention of a **600 V** motor voltage class for many industrial three-phase applications, versus the more common **480 V** convention south of the border — a designer moving between US and Canadian projects needs to watch for this, since it changes both the FLA (for a given horsepower) and the available conductor/component ratings.

### Governing Calculation

$$
I_{cond} \ge 1.25 \times I_{FLA}
$$

$$
I_{OL} = m_{OL} \times I_{FLA}
$$

$$
I_{OCPD,calc} = m_{fuse} \times I_{FLA}
$$

$$
I_{OCPD} = \text{next standard device size} \ge I_{OCPD,calc}
$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $I_{FLA}$ | Motor full-load amps, from the nameplate or applicable ampacity table | A |
| $I_{cond}$ | Minimum required branch circuit conductor ampacity | A |
| $m_{OL}$ | Overload relay setting multiplier (typically 1.15-1.25, depending on service factor/temperature rise) | -- |
| $I_{OL}$ | Overload relay trip setting | A |
| $m_{fuse}$ | Time-delay fuse sizing multiplier (typically ~1.5-1.75x FLA as a starting point) | -- |
| $I_{OCPD,calc}$ | Calculated maximum branch-circuit protective device rating | A |
| $I_{OCPD}$ | Selected (standard size) branch-circuit protective device rating | A |

```{r fig-06-03-tcc-schematic, echo=FALSE, fig.cap="Schematic time-current coordination: a properly selected time-delay fuse curve sits below the conductor's damage (withstand) curve at every current, and above the motor's starting-current point, so the fuse rides through a normal start but still protects the conductor. Illustrative shape only, not actual manufacturer trip data.", fig.align='center', fig.width=9, fig.height=6}
mult <- seq(1.2, 12, length.out = 200)  # multiples of the fuse's ampere rating
fuse_t <- 200 / mult^2.2
conductor_t <- 2000 / mult^2.2

curve_df <- data.frame(
  mult = rep(mult, 2),
  t = c(fuse_t, conductor_t),
  curve = rep(c("Time-delay fuse melting curve", "Conductor damage (I2t) withstand"), each = length(mult))
)

start_point <- data.frame(mult = 3.47, t = 8)

ggplot(curve_df, aes(x = mult, y = t, color = curve)) +
  geom_line(linewidth = 1.2) +
  annotate("point", x = start_point$mult, y = start_point$t, size = 3.5,
           shape = 4, stroke = 1.6, color = "black") +
  annotate("text", x = start_point$mult, y = start_point$t * 2.4,
           label = "Motor starting:\nLRA ~ 3.5x fuse rating\nfor ~8 s",
           size = 2.9 * diagram_text_scale, hjust = 0.5) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = c("Time-delay fuse melting curve" = "steelblue",
                                 "Conductor damage (I2t) withstand" = "firebrick")) +
  labs(title = "Time-Current Coordination (Schematic)",
       x = "Current (multiples of fuse ampere rating, log scale)",
       y = "Time (s, log scale)", color = NULL) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale),
    legend.position = "bottom",
    legend.text = element_text(size = 9 * diagram_text_scale)
  )
```

### Worked Example

Live cell — try changing `FLA`, `service_factor`, or `m_fuse` and re-run to see the conductor and fuse selection update.

```{webr-r}
#| label: branch-protection-06-03
## 30 hp, 600 V, three-phase motor -- representative tabulated/nameplate FLA
## (as would be read from the motor nameplate or the applicable CEC ampacity
## table for this horsepower/voltage combination)
FLA <- 32  # A

## Step 1: branch-circuit conductor ampacity (125% continuous-duty rule)
conductor_ampacity_required <- 1.25 * FLA
cat("Motor FLA =", FLA, "A\n")
cat("Required conductor ampacity (1.25 x FLA) =", conductor_ampacity_required, "A\n")

conductor_selected <- "8 AWG copper, 90C insulation"
conductor_ampacity_actual <- 44  # A, representative ampacity for this conductor/insulation
cat("Selected conductor:", conductor_selected, "-- ampacity =", conductor_ampacity_actual, "A")
if (conductor_ampacity_actual >= conductor_ampacity_required) cat("  -- PASS\n") else cat("  -- FAILS\n")

## Step 2: overload relay setting (protects the motor, separate from the
## branch OCPD below)
service_factor <- 1.15
m_OL <- ifelse(service_factor >= 1.15, 1.25, 1.15)
I_OL <- m_OL * FLA
cat("\nOverload relay multiplier (service factor =", service_factor, ") =", m_OL, "\n")
cat("Overload relay setting =", round(I_OL, 1), "A\n")

## Step 3: time-delay fuse sizing (protects conductors/equipment; rides
## through starting inrush)
m_fuse <- 1.75
I_OCPD_calc <- m_fuse * FLA
standard_fuse_sizes <- c(15, 20, 25, 30, 35, 40, 45, 50, 60, 70, 80, 90, 100, 110, 125)
I_OCPD <- min(standard_fuse_sizes[standard_fuse_sizes >= I_OCPD_calc])

cat("\nCalculated max. time-delay fuse rating (", m_fuse, "x FLA ) =",
    round(I_OCPD_calc, 1), "A\n")
cat("Selected standard fuse size =", I_OCPD, "A (",
    round(I_OCPD / FLA * 100, 0), "% of FLA )\n\n")

## Check against a locked-rotor current estimate to confirm the fuse rides
## through a normal start
LRA_multiplier <- 6.5  # typical for a NEMA Design B motor, rule of thumb
LRA <- LRA_multiplier * FLA
cat("Estimated locked-rotor current (", LRA_multiplier, "x FLA ) =", round(LRA, 0), "A\n")
cat("Fuse rating as a multiple of estimated LRA =",
    sprintf("%.2f", I_OCPD / LRA), "x -- a time-delay fuse tolerates a brief",
    "multi-second overload well above its own rating, so this is expected\n")

cat("\n--- Branch circuit summary ---\n")
cat("Conductor:", conductor_selected, "(", conductor_ampacity_actual, "A ) -- PASS\n")
cat("Overload relay setting:", round(I_OL, 1), "A\n")
cat("Time-delay fuse:", I_OCPD, "A\n")
```

**Sanity check:** the selected fuse (60 A) sits at roughly 1.9x FLA, comfortably inside the range this chapter's rule of thumb predicts, and the overload relay (40 A) sits between FLA (32 A) and the conductor's ampacity (44 A) exactly where it should — protecting the motor from sustained overload well before the conductor itself is threatened, while the fuse stands by purely for short-circuit and ground-fault events. If the fuse rating had landed close to or above the conductor's ampacity, that would be a red flag worth double-checking, since it would mean the "short-circuit protection" device was no longer meaningfully protecting the conductor it shares a circuit with.

## Design Task 5: Panel Short-Circuit Current Rating (SCCR)

### Scenario

The automotive OEM's electrical specification requires this panel to carry a 65 kA SCCR marking on its nameplate. The panel's main incoming circuit breaker is already spec'd at 65 kA interrupting rating — job done, right? The engineer reviewing the bill of materials isn't so sure, and pulls the individual SCCR ratings for every downstream component to check.

### Design Guidelines and Rules of Thumb

**Short-circuit current rating (SCCR)** is the maximum prospective fault current a piece of equipment (or an assembled panel) can safely withstand without creating a hazard, even though it may be damaged in the process. It is not the same thing as a breaker's interrupting rating, though the two are related: interrupting rating describes what a single overcurrent device can safely open into, while SCCR describes what the *entire assembly downstream of that device* can survive while the fault is being cleared.

This is where the **weakest-link principle** becomes the whole story: a panel's overall SCCR is the *lowest* SCCR of any single component in the fault current path, unless a documented, tested combination changes that for a specific pairing. A 65 kA main breaker feeding a bank of ordinary 5 kA-rated IEC contactors does not produce a 65 kA panel — it produces, at best, a 5 kA panel, because a fault severe enough to approach 65 kA would destroy those contactors (and potentially create an arc-flash hazard) long before the main breaker's own rating becomes the limiting factor.

The fix is not automatically "buy every component individually rated at 65 kA," which is often prohibitively expensive or simply unavailable for certain device types (many standard IEC contactors, for instance, top out at modest standalone SCCR ratings). Instead, panel builders commonly use a **current-limiting device** — typically a specific current-limiting fuse — installed immediately upstream of the low-rated component. Fuse and component manufacturers publish **tested combination ratings**: a specific fuse type and amperage, paired with a specific contactor (or other device) model, tested together and certified to a much higher combination SCCR than the contactor's own standalone rating, because the fuse's current-limiting action clips the let-through current before it ever reaches the contactor's damage threshold. UL 508A Supplement SB documents this method and lists many manufacturer-published component SCCR and combination-rating tables that a panel builder can reference directly rather than testing each combination independently.

**Rule of thumb:** for any panel with a customer-specified SCCR requirement, don't just check the main breaker — check *every* component in the fault current path (starters, contactors, terminal blocks, disconnects, VFDs, power supplies), because the weakest link is very often not the main breaker at all. This rule of thumb has essentially no failure mode worth noting other than the discipline to actually do the check on every device, since skipping even one low-rated component silently caps the entire panel's real SCCR regardless of what the nameplate says.

**Rule of thumb:** automotive OEMs commonly specify 65 kA SCCR as a blanket requirement across an entire plant, even where the actual available fault current at a given panel's point of connection is considerably lower. This is a deliberate standardization choice, not a reflection of every panel's actual exposure — a plant with a single specification simplifies procurement, spare-parts stocking, and future relocations of equipment between service points with different available fault currents, at the cost of some components being rated well above what any individual panel strictly needs.

### Governing Calculation

$$
SCCR_{panel} = \min\left(SCCR_1,\ SCCR_2,\ \ldots,\ SCCR_n\right)
$$

where, for each component $i$:

$$
SCCR_i =
\begin{cases}
SCCR_{combination} & \text{if protected by a documented current-limiting combination} \\
SCCR_{standalone} & \text{otherwise}
\end{cases}
$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $SCCR_{standalone}$ | A component's own individually tested/marked SCCR | kA |
| $SCCR_{combination}$ | The higher SCCR achieved when a specific component is protected by a specific, documented current-limiting device (per the manufacturer's published tested combination) | kA |
| $SCCR_i$ | The effective SCCR of component $i$ as actually installed in the panel | kA |
| $SCCR_{panel}$ | The panel's overall marked SCCR: the minimum across every component in the fault current path | kA |

```{r fig-06-03-weakest-link, echo=FALSE, fig.cap="The weakest-link principle: a 65 kA main breaker does not make a 65 kA panel if a downstream contactor is only rated 5 kA standalone. A documented current-limiting fuse/contactor combination raises the effective rating of that branch to match.", fig.align='center', fig.width=10, fig.height=5}
nodes <- data.frame(
  id = 1:6,
  label = c("Main breaker\n65 kA", "Contactor\n(standalone)\n5 kA", "Panel SCCR\n= 5 kA",
            "Main breaker\n65 kA", "CL fuse +\ncontactor combo\n65 kA", "Panel SCCR\n= 65 kA"),
  x = c(1, 3.2, 5.4, 1, 3.2, 5.4),
  y = c(2, 2, 2, 0.6, 0.6, 0.6),
  status = c("ok", "weak", "fail", "ok", "ok", "pass")
)

edges <- data.frame(
  x = c(1.55, 3.75, 1.55, 3.75),
  y = c(2, 2, 0.6, 0.6),
  xend = c(2.65, 4.85, 2.65, 4.85),
  yend = c(2, 2, 0.6, 0.6)
)

ggplot() +
  geom_segment(data = edges, aes(x = x, y = y, xend = xend, yend = yend),
               arrow = arrow(length = unit(0.2, "cm")), color = "grey50", linewidth = 0.7) +
  geom_label(data = nodes, aes(x = x, y = y, label = label, fill = status),
             size = 3.1 * diagram_text_scale, label.padding = unit(0.3, "lines"), show.legend = FALSE) +
  scale_fill_manual(values = c(ok = "#D6EAF8", weak = "#FDEBD0", fail = "#F5B7B1", pass = "#ABEBC6")) +
  annotate("text", x = -0.15, y = 2, label = "Before:", fontface = "bold",
           hjust = 0, size = 3.2 * diagram_text_scale) +
  annotate("text", x = -0.15, y = 0.6, label = "After:", fontface = "bold",
           hjust = 0, size = 3.2 * diagram_text_scale) +
  xlim(-0.6, 6.2) + ylim(-0.1, 2.9) +
  theme_void() +
  labs(title = "Weakest-Link Principle for Panel SCCR") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14 * diagram_text_scale))
```

### Worked Example

Live cell — try changing `standalone_sccr_kA` or `cl_fuse_combination_sccr_kA` for a component and re-run to see the panel's overall SCCR update.

```{webr-r}
#| label: sccr-check-06-03
required_sccr <- 65  # kA, automotive OEM panel specification

components <- data.frame(
  component = c("Main incoming circuit breaker",
                "Weld-pump motor contactor (Branch 1)",
                "Coolant-pump motor contactor (Branch 2)",
                "24VDC control power supply"),
  standalone_sccr_kA = c(65, 5, 5, 5),
  cl_fuse_combination_sccr_kA = c(NA, 65, 65, 65),
  stringsAsFactors = FALSE
)

## Pass 1: as originally specified, before any current-limiting fuse is
## deliberately paired with the low-SCCR branch devices
components$effective_sccr_before <- components$standalone_sccr_kA
panel_sccr_before <- min(components$effective_sccr_before)

cat("--- Before: standalone component ratings only ---\n")
for (i in seq_len(nrow(components))) {
  cat(sprintf("%-38s %5.0f kA\n", components$component[i], components$effective_sccr_before[i]))
}
cat("Panel SCCR (weakest link) =", panel_sccr_before, "kA\n")
if (panel_sccr_before >= required_sccr) {
  cat("PASS -- meets the 65 kA requirement\n\n")
} else {
  cat("FAILS -- far short of the 65 kA requirement\n\n")
}

## Pass 2: add a documented current-limiting fuse ahead of each low-rated
## branch device, using the manufacturer's published tested combination
## rating for that specific fuse/device pairing
components$effective_sccr_after <- ifelse(
  is.na(components$cl_fuse_combination_sccr_kA),
  components$standalone_sccr_kA,
  components$cl_fuse_combination_sccr_kA
)
panel_sccr_after <- min(components$effective_sccr_after)

cat("--- After: current-limiting fuse combinations applied ---\n")
for (i in seq_len(nrow(components))) {
  cat(sprintf("%-38s %5.0f kA\n", components$component[i], components$effective_sccr_after[i]))
}
cat("Panel SCCR (weakest link) =", panel_sccr_after, "kA\n")
if (panel_sccr_after >= required_sccr) {
  cat("PASS -- panel now meets the 65 kA requirement\n")
} else {
  cat("Still FAILS -- further component or fusing changes are required\n")
}
```

**Sanity check:** the "before" case shows a panel with a genuinely 65 kA-rated main breaker still only achieving a 5 kA overall SCCR — the exact trap the rule of thumb above warns about. The fix doesn't touch the main breaker at all; it adds current-limiting fuses ahead of the three low-rated branch devices, each backed by a manufacturer-published tested combination rating, which is the documentation a UL 508A panel shop's build file needs to justify the 65 kA nameplate marking, not just an engineer's assertion that it should work.

<details>
<summary>Industry connection: why automotive plants standardize on 65 kA</summary>

A Tier 1 automotive stamping plant typically sits on a stiff utility feed shared across dozens of press lines, weld cells, and conveyor systems — as more equipment is added over the plant's life, available fault current at any given point tends to creep up, not down. Specifying every new panel to 65 kA up front means a panel installed today doesn't become an SCCR liability five years from now when the plant's electrical infrastructure grows around it.

</details>

## Design Task 6: Panel Certification Pathway — UL 508A, CSA C22.2 No. 286, or ESA Field Evaluation

### Scenario

Two panels are headed for the same Ontario plant in the same month. Panel A is being built new, in-house, by the integrator's own panel shop. Panel B is a control panel that came bundled with a used stamping press, purchased secondhand from an overseas equipment reseller, with no certification marks the receiving electrician can find anywhere on it. Both panels need to be connected to the plant's electrical system — but they are not going to get there the same way.

### Design Guidelines and Rules of Thumb

In Ontario, the *Electrical Safety Authority* (ESA), administering the OESC's adoption of the CEC, generally will not approve connection of electrical equipment that doesn't bear a recognized certification mark from an accredited certification body — a CSA mark, a cULus/cUL mark, an ETL mark, or an equivalent. For an industrial control panel specifically, that mark is most often earned one of two ways:

**UL 508A Listed Panel Shop:** UL (Underwriters Laboratories) operates a program under which a panel-building company itself — not just individual components — is audited, certified, and enrolled in an ongoing Follow-Up Service. A shop holding this status is authorized to build panels according to its own filed procedures and **self-apply** the UL Listing mark directly to each panel it builds, without sending every individual panel out for third-party testing. This is why so many automotive OEM specifications simply require "UL 508A listed panel shop" — it's a fast, well-understood, US-centric certification path that most major machine builders and integrators already hold, and OEM specifications are frequently written centrally in the US and applied across all of a company's plants, including Canadian ones.

**CSA C22.2 No. 286 Certified Panel Shop:** this is the direct Canadian equivalent — CSA's own standard for industrial control panels, with a parallel shop-certification and self-application program. A Canadian-market panel shop can hold a CSA C22.2 No. 286 certification, a UL 508A listing with Canadian ("C") coverage, or both, depending on which markets it serves; many shops serving cross-border automotive supply chains maintain dual coverage so the same panel design can ship to either country without a separate certification exercise.

**ESA Field Evaluation:** not every panel a plant needs comes from a certified shop — a panel bundled with used equipment, a one-off panel built in-house by plant maintenance without a certified shop's program, or an older panel whose original certification can no longer be verified all fall outside the listed-shop path. For these cases, Ontario's Field Evaluation program (administered through the ESA in conjunction with accredited field evaluation agencies) provides an alternative: an accredited agency inspects the **specific physical unit** against the applicable standard (commonly CSA C22.2 No. 286 or an equivalent), and — once any deficiencies are corrected — applies a Special Inspection / field evaluation label directly to that unit. That label covers only the inspected unit itself, not the model in general or any future copies of it; a second, identical used panel bought from the same reseller would need its own separate field evaluation.

**Rule of thumb:** a plant should default to assuming any incoming panel needs a field evaluation unless a recognized, currently valid certification mark is physically visible on it and the panel has not been field-modified since it was listed (adding a component, changing a circuit, or altering the SCCR-relevant makeup of the panel after listing can itself invalidate the original mark and trigger the need for a field evaluation of just the modification). This rule of thumb rarely causes real harm when it "over-triggers" a review of a panel that turns out to be fine — the cost is a short delay, not a hazard — whereas skipping a needed field evaluation because "it's probably fine" risks connecting genuinely unverified equipment to the plant's electrical system.

```{r tbl-06-03-cert-comparison, echo=FALSE, message=FALSE}
cert_paths <- tribble(
  ~Path, ~Who_Certifies, ~What_It_Covers, ~When_It_Applies,
  "UL 508A Listed Panel Shop", "UL (or another accredited NRTL), auditing the shop's own procedures under ongoing Follow-Up Service", "Every panel the shop builds under its filed program; shop self-applies the mark", "New panels from a shop holding this status -- the most common path for US-originated OEM specs",
  "CSA C22.2 No. 286 Certified Panel Shop", "CSA Group (or another accredited certification body), auditing the shop similarly to the UL program", "Every panel the shop builds under its filed program; shop self-applies the mark", "New panels built specifically for the Canadian market, or shops holding dual CSA/UL coverage",
  "ESA Field Evaluation (Special Inspection)", "An accredited field evaluation agency, engaged case-by-case", "The one specific physical unit inspected -- not the model, not future copies", "Used/reused equipment, in-house builds without a certified shop, or panels with no verifiable/valid existing mark"
)

cert_paths %>%
  kbl(caption = "Panel certification pathways: UL 508A, CSA C22.2 No. 286, and ESA field evaluation") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, background = "#2C3E50", color = "white")
```

### Governing Calculation (Decision Framework)

1. **Does the panel bear a recognized, currently valid certification mark** (CSA, cULus/cUL, ETL, or equivalent) from an accredited body? If no, a field evaluation is required before ESA will approve connection.
2. **If a mark is present, has the panel been field-modified since it was listed** in any way that could affect the characteristic the listing certified (added components, changed circuits, altered SCCR-relevant makeup)? If yes, the modification itself needs evaluation, even though the original listing remains valid for the unmodified portions.
3. **If no mark is present, engage an accredited field evaluation agency**, inspect the specific unit against the applicable standard, correct any deficiencies identified, and have the agency apply a Special Inspection / field evaluation label before the plant connects and energizes the equipment.

### Worked Example

Live cell — try setting `recognized_mark_present` to `TRUE` and re-run to see the conclusion change.

```{webr-r}
#| label: cert-pathway-06-03
## Panel B: used stamping press control panel from an overseas reseller
recognized_mark_present <- FALSE  # no CSA, cULus, ETL, or other accredited mark visible

cat("Recognized certification mark present:", recognized_mark_present, "\n\n")

if (!recognized_mark_present) {
  cat("Conclusion: the ESA will not approve this panel for connection as-is.\n")
  cat("Required path: ESA field evaluation (Special Inspection)\n\n")
  cat("Steps:\n")
  cat(" 1. Engage an accredited field evaluation agency\n")
  cat(" 2. Agency inspects THIS specific panel against the applicable standard\n")
  cat("    (e.g., CSA C22.2 No. 286)\n")
  cat(" 3. Correct any deficiencies identified during inspection\n")
  cat(" 4. Agency applies a Special Inspection / field evaluation label to\n")
  cat("    this physical unit (the label does not extend to any other unit,\n")
  cat("    even an identical one)\n")
  cat(" 5. File the documentation with the ESA; the plant may then connect\n")
  cat("    and energize the panel\n")
} else {
  cat("Conclusion: a recognized mark is present -- the ESA will generally\n")
  cat("accept the panel without a separate field evaluation, provided it has\n")
  cat("not been field-modified in a way that affects the original listing.\n")
}

## Panel A, for comparison: new build from a UL 508A listed panel shop,
## self-applied mark, 65 kA SCCR on the nameplate
recognized_mark_present_2 <- TRUE
field_modified_since_listing <- FALSE

cat("\n--- Comparison: Panel A (new build, UL 508A listed shop) ---\n")
cat("Recognized certification mark present:", recognized_mark_present_2, "\n")
cat("Field-modified since listing:", field_modified_since_listing, "\n")
if (recognized_mark_present_2 && !field_modified_since_listing) {
  cat("Conclusion: ESA field evaluation is NOT required for this panel.\n")
}
```

**Sanity check:** the two panels headed for the same plant, in the same month, reach opposite conclusions purely because of documentation history, not because either panel is necessarily better or worse engineering — which is exactly the point of a certification pathway that runs on verifiable marks and inspection records rather than on someone's confidence that "it looks fine."

<details>
<summary>Industry connection: build records on a controlled-goods defense project</summary>

On a defense-manufacturing project subject to Canada's Controlled Goods Program, the certification discipline in this design task extends further than getting a panel connected: the shop's UL 508A or CSA C22.2 No. 286 build file, the specific component certificates used, and the field evaluation records (where applicable) become part of the traceable documentation package an auditor can request years later — the same habit of "prove it, don't just assert it" that governs SCCR combination ratings applies just as directly to demonstrating a controlled-goods build followed its approved process from procurement through final inspection.

</details>

## Summary

```{r tbl-06-03-summary, echo=FALSE, message=FALSE}
summary_table <- tribble(
  ~Design_Task, ~Governing_Relationship, ~Key_Rule_of_Thumb,
  "1. Enclosure layout / wire duct fill", "Fill% = (sum of conductor areas) / (duct usable area) x 100%; R_bend_min = k x OD",
  "Keep fill <= 40% (25-30% preferred for growth); shielded cable needs 8-10x OD bend radius",
  "2. NEMA vs. IP rating selection", "Decision framework: outdoor? washdown severity? corrosive/food environment?",
  "Washdown zones need BOTH a NEMA type and an IP rating -- IP69K has no NEMA equivalent at all",
  "3. Internal heat dissipation", "delta_T = Q_total / (k x A_eff); V_flow = Q_total / (rho x cp x delta_T_allow)",
  "A drive-heavy panel rarely survives on passive cooling alone -- check forced air before assuming AC is needed",
  "4. Branch circuit protection", "I_cond >= 1.25 x FLA; I_OCPD = next standard size >= m_fuse x FLA",
  "Time-delay fuse ~150-175% of FLA rides through starting inrush that a standard fuse would trip on",
  "5. Panel SCCR", "SCCR_panel = min(SCCR_1, ..., SCCR_n) across every component in the fault path",
  "Check every component, not just the main breaker -- the weakest link is rarely the breaker itself",
  "6. Certification pathway", "Checklist: recognized mark present? field-modified since listing?",
  "Default to assuming a field evaluation is needed until a valid, unmodified certification mark says otherwise"
)

summary_table %>%
  kbl(caption = "Design task summary: governing relationship and key rule of thumb",
      col.names = c("Design Task", "Governing Relationship", "Key Rule of Thumb")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, background = "#2C3E50", color = "white")
```

## Practice Problems

<details>
<summary>Task 1 (plug-in): A different duct run carries 10 conductors of 16 AWG (OD 2.2 mm) and 6 conductors of 12 AWG (OD 2.9 mm). Using the same 48% usable-area fraction, does this bundle fit in a 25 x 25 mm duct (usable area = 300 mm²) within the 40% fill limit?</summary>

Conductor area: 16 AWG = π(1.1)² ≈ 3.80 mm²; 12 AWG = π(1.45)² ≈ 6.60 mm²

Total area = 10 × 3.80 + 6 × 6.60 = 38.0 + 39.6 = 77.6 mm²

Fill% = 77.6 / 300 × 100 ≈ 25.9%

25.9% < 40% — **PASS**, with room to spare (well inside the 25-30% growth-friendly range too).

</details>

<details>
<summary>Task 1 (design decision): A technician needs to add six more 16 AWG conductors to the panel from the worked example (already at 27.9% fill in the 40 x 25 mm duct) during commissioning. The panel shop offers two options: re-pull all the wiring into a larger 40 x 40 mm duct, or add a second parallel 25 x 25 mm duct just for the new conductors. Which would you recommend, and what would you check first?</summary>

Adding the six conductors to the existing duct first: six more 16 AWG conductors add 6 × 3.80 ≈ 22.8 mm² to the existing 134.1 mm², giving 156.9 mm². At 40 x 25 mm (usable area 480 mm²), fill becomes 156.9/480 ≈ 32.7% — still under 40%, so re-pulling into the *same* duct size isn't even necessary; the existing 40 x 25 mm duct still has room. If the question is instead whether to size up preemptively for anticipated further growth beyond these six conductors, upsizing to 40 x 40 mm (usable area 768 mm², fill ≈ 20.4% with the new total) buys more headroom in one duct rather than splitting the run. A second parallel duct is the least attractive option here — it adds a second physical raceway to install, label, and maintain for a fill increase that doesn't actually threaten the 40% limit; it would only make sense if physical routing space, not fill percentage, were the actual constraint (e.g., no room to swap the existing duct for a larger one along its run). The recommended check first is simply recomputing fill% with the added conductors before assuming a resize is needed at all — in this case it isn't.

</details>

<details>
<summary>Task 2 (plug-in): A beverage bottling line has a control panel mounted directly above a filler head, in a zone that gets an occasional light rinse but never a direct high-pressure washdown, and is not exposed to significant corrosive chemicals. What rating does the decision framework recommend?</summary>

Working through the framework: indoor (yes, inside the plant), no routine high-pressure hot washdown (only an occasional light rinse), and not significantly corrosive. This falls into the "occasional hosedown" branch rather than the full IP69K branch — the framework recommends **NEMA 4X, IP66**: corrosion-resistant construction as a sensible precaution in a food/beverage environment, with a hosedown-rated (but not IP69K) ingress rating, since the exposure described doesn't reach the high-pressure, high-temperature severity that specifically demands IP69K.

</details>

<details>
<summary>Task 2 (design decision): The dairy-line panel from the worked example (NEMA 4X, IP69K) costs roughly $3,200 more in stainless steel and certified gasket hardware than an equivalent NEMA 12/IP54 painted steel panel would cost. The plant manager asks whether the cheaper option could be used "since the panel door stays closed most of the time anyway." How would you respond?</summary>

"The door stays closed most of the time" addresses the wrong exposure — NEMA/IP ratings for a panel in a washdown zone aren't primarily about protecting against someone reaching inside; they're about protecting against water and moisture intrusion through door seals, conduit entries, and enclosure seams during the routine, repeated high-pressure washdown the sanitation crew performs at the end of every shift, whether or not anyone opens the door. A NEMA 12/IP54 panel's gaskets and construction are not rated to survive that exposure even once, let alone daily for years, and moisture intrusion into a "closed" panel creates exactly the hidden corrosion and bacterial-growth-site risk described earlier in this chapter — a risk that a food safety or CFIA audit is specifically designed to catch. The right response is to hold the line on NEMA 4X/IP69K for this location: the $3,200 premium is cheap compared to the cost of a failed panel, an unplanned production stoppage from a moisture-related fault, or a food safety audit finding tied directly to enclosure condition in a washdown zone.

</details>

<details>
<summary>Task 3 (plug-in): Using the same enclosure (A_eff = 2.44 m², k = 5.5 W/(m^2*C)) and the same allowable rise (15C), what forced-air flow rate (m^3/h) is required if the total heat load is only 400 W instead of 809 W?</summary>

Passive check first: delta_T = 400 / (5.5 × 2.44) = 400 / 13.42 ≈ 29.8°C, still above the 15°C allowance, so forced air is still needed.

V_flow = Q / (rho × cp × delta_T_allow) = 400 / (1.2 × 1005 × 15) = 400 / 18 090 ≈ 0.0221 m³/s

Converting to m³/h: 0.0221 × 3600 ≈ 79.6 m³/h — roughly half the airflow required in the original 809 W worked example, which makes sense since airflow requirement scales directly with heat load for a fixed allowable temperature rise.

</details>

<details>
<summary>Task 3 (design decision): The plant later decides this panel must be upgraded to NEMA 4X/IP66 (occasional washdown-adjacent exposure) after it's already been laid out with a filtered fan sized per the original worked example. What has to change in the cooling design, and why?</summary>

A filtered intake/exhaust fan draws in ambient air and exhausts panel air directly through the enclosure wall, which is exactly the kind of opening a NEMA 4X/IP66 ingress rating cannot tolerate — the fan and its filter openings would breach the enclosure's seal, defeating the washdown rating regardless of how good the fan's own filtration is. The cooling approach has to change to something that rejects heat without exchanging enclosure air with the outside: a sealed air-to-air heat exchanger (which moves heat across a sealed barrier without air exchange, suitable for a more modest heat load) or a compressor-based air conditioner (for a heat load like this panel's 809 W, which is more than a typical passive heat exchanger comfortably handles). This is also a cost and footprint increase over the simple filtered fan, and it needs to be caught during layout — discovering it after the enclosure and fan cutouts are already fabricated means scrapping or reworking sheet metal, not just swapping a part number.

</details>

<details>
<summary>Task 4 (plug-in): A 20 hp, 600 V, three-phase motor has a representative FLA of 22 A. Using the same rules as the worked example (1.25x for conductor ampacity, 1.75x for the time-delay fuse calculation, standard fuse sizes from the same list), what conductor ampacity is required and what standard fuse size would be selected?</summary>

Required conductor ampacity = 1.25 × 22 = 27.5 A (select a conductor with ampacity at or above this, e.g., a 10 AWG conductor with a representative 90°C ampacity around 30-35 A, which would pass).

Calculated max. fuse rating = 1.75 × 22 = 38.5 A

From the standard sizes (15, 20, 25, 30, 35, 40, 45, 50, 60, ...), the next size at or above 38.5 A is **40 A**.

</details>

<details>
<summary>Task 4 (design decision): The 40 A time-delay fuse selected in the previous problem trips nuisance-fashion on two of the motor's first five starts after commissioning, even though the motor and conductors both check out fine electrically. The maintenance team's first instinct is to jump straight to a 60 A fuse. What would you check before agreeing to that change, and why?</summary>

Jumping straight to a larger fuse treats the symptom, not necessarily the cause, and a larger fuse provides less short-circuit protection margin for the conductor without first confirming that's actually the right fix. Before resizing, it's worth checking: is the load being started unusually heavy (a conveyor started fully loaded rather than empty, a longer-than-expected acceleration ramp) that pushes the real starting duration beyond what a 175%-sized fuse is meant to tolerate? Is the fuse genuinely a time-delay (dual-element) type and not a standard fast-acting fuse installed by mistake, which would nuisance-trip on any normal motor start? Is the starting current itself higher than the 6.5x-FLA rule-of-thumb estimate used in the sizing calculation, perhaps because of a low-inertia assumption that doesn't match the actual load? If the fuse type and rating both check out as correctly selected and specified, and the starting profile is confirmed to be a normal, expected start, then moving up to the next standard size (still within the 225% upper bound Section 28/Article 430 generally allow for cases where 175% proves insufficient) is a reasonable next step — but it should be a documented decision based on the actual starting current profile, not a default first reaction to a nuisance trip.

</details>

<details>
<summary>Task 5 (plug-in): A panel has a main breaker rated 65 kA, a disconnect switch rated 25 kA standalone (protected by an upstream fuse with a published combination rating of 65 kA), and a terminal block rated 100 kA standalone. What is this panel's overall SCCR?</summary>

Effective SCCR of each component: main breaker = 65 kA; disconnect switch = 65 kA (using its combination rating, since it's protected by the documented current-limiting fuse pairing); terminal block = 100 kA (standalone, no protection needed since it already exceeds the target).

Panel SCCR = min(65, 65, 100) = **65 kA**.

</details>

<details>
<summary>Task 5 (design decision): A cost-conscious purchasing agent proposes buying the cheaper, non-combination-rated version of the branch contactors used in the worked example (standalone 5 kA, with no published current-limiting fuse combination available from that particular manufacturer) to save $40 per contactor across twelve contactors ($480 total). What's the actual tradeoff here, and how would you respond?</summary>

Without a published, documented current-limiting combination for that specific contactor, there is no way to justify anything higher than its 5 kA standalone rating on the panel's SCCR nameplate — the panel would revert to a 5 kA SCCR overall, regardless of what any other component in the panel is rated. That fails the customer's 65 kA specification outright, which isn't a minor deviation; it's a nonconformance that could hold up final acceptance, require costly rework after the panel is already built and wired, or (worse) go unnoticed and leave a genuinely under-rated panel in production. The $480 saved on contactors is trivial next to the cost of reworking an assembled panel, and trivial next to the safety exposure of a panel that can't actually survive the fault current it's nameplated for. The right response is to decline the substitution unless purchasing can find an equivalent contactor from a different manufacturer that *does* have a published 65 kA combination rating with an available current-limiting fuse — the cost comparison has to include that documentation requirement, not just the unit price of the contactor itself.

</details>

<details>
<summary>Task 6 (plug-in): A panel bears a valid CSA mark from when it was originally listed three years ago, and no components, circuits, or wiring have been added or changed since. Does this panel need an ESA field evaluation before being reconnected after a plant relocation?</summary>

Working through the checklist: is a recognized, currently valid certification mark present? Yes (CSA mark, three years old but not expired or revoked). Has the panel been field-modified since listing in a way that could affect the certified characteristics? No — no components, circuits, or wiring changes since original listing. Per the decision framework, this panel **does not require a field evaluation** — the existing CSA mark remains valid, and a physical relocation alone (with no modification to the panel itself) doesn't trigger a new evaluation. It's still good practice to visually inspect the panel and its mark for damage after the move and to document that no modifications occurred, but a formal field evaluation is not required by this scenario alone.

</details>

<details>
<summary>Task 6 (design decision): A plant's maintenance team wants to add a new VFD to an existing UL 508A listed panel to control a recently added conveyor motor, tapping into spare space inside the enclosure. The panel's original SCCR marking was justified using the exact component list and current-limiting combinations documented at the time of listing. What should happen before this modification is made, not after?</summary>

Adding a new VFD changes the panel's fault-current path and heat load — both of which were part of what the original UL 508A listing certified. Doing the physical work first and figuring out the paperwork later is backwards: the modification should be evaluated *before* it's made, ideally by confirming with the original panel shop (or a qualified engineer familiar with UL 508A/CSA C22.2 No. 286 requirements) whether the new VFD, as specifically selected and protected, keeps the panel's SCCR rating and heat dissipation within what the original listing covers. If it does, the addition can likely be documented as within the existing listing's scope. If it doesn't — say, the new VFD's own SCCR requirement, or its added heat load, pushes the panel outside what was originally certified — the modification itself may need a field evaluation, or the panel's listing basis needs to be formally updated, before the VFD is energized in service. Making the physical change first and discovering the mismatch during an audit or incident investigation is exactly the scenario this chapter's certification discipline exists to prevent.

</details>

## Open-Ended Design Scenario

<details>
<summary>Capstone scenario: certify the transfer system panel for production</summary>

You are the panel designer responsible for releasing the transfer system control panel from this chapter for fabrication and, eventually, for connection at the Cambridge automotive plant. Pull together everything from this chapter into one design review:

- The panel's I/O and control wiring duct run carries 18 conductors of 16 AWG (OD 2.2 mm), 8 of 14 AWG (OD 2.5 mm), and 4 of 12 AWG (OD 2.9 mm). A separate run carries one shielded VFD output cable, OD 12 mm, through a duct corner fitting with a 60 mm radius.
- The panel sits in the plant's general assembly area — indoor, dry, no washdown, not corrosive.
- The enclosure is 800 x 1000 x 300 mm, free-standing, painted mild steel, housing three 7.5 kW VFDs, a 1 kVA control transformer, eight contactors, a PLC rack, and miscellaneous terminal hardware, in a 35°C ambient with a 50°C maximum internal ambient rating on the drives.
- One branch feeds a 30 hp, 600 V, three-phase motor with a representative FLA of 32 A.
- The customer's specification requires a 65 kA SCCR marking. The main breaker is rated 65 kA; two branch motor contactors and the 24VDC control power supply are each rated only 5 kA standalone, but current-limiting fuse combinations rated 65 kA are available and documented for each of them.
- The panel is being built by your company's own UL 508A listed panel shop, and no components will be added after the shop applies its Listing mark.

For each of the six design tasks in this chapter, work through the calculation or decision framework and reach a clear conclusion: does the panel pass, and if not, what specific change would you make? Then write two or three sentences tying your conclusions together — does a change you made for one design task (say, upsizing the wire duct) create or resolve any tension with your conclusions for another (say, the enclosure's available surface area for cooling, or the physical space available for a current-limiting fuse ahead of a contactor)? A strong answer treats this as one connected panel design, not six independent homework problems.

</details>
