---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Robot Cell Design

```{r setup-06-01, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter, you will be able to:

1. **Verify** that a robot's rated reach envelope covers all required work points with adequate margin, accounting for tooling length and mounting offset.
2. **Calculate** a robot's allowable payload at a given end-effector center-of-gravity (CG) offset and compare it to the actual carried load.
3. **Estimate** a robot cell's cycle time from a sequence of point-to-point moves, using a trapezoidal velocity profile plus process dwell time.
4. **Calculate** the minimum safe separation distance for a guarding device per ISO 13855, given the system's overall stopping time.
5. **Select** an appropriate guarding technology — fixed fencing, light curtain, or area scanner — for a given access point, using a documented decision framework.
6. **Evaluate** whether a "collaborative" robot application still requires a full risk assessment under ISO/TS 15066.
7. **Determine** whether a new or modified robot cell triggers an Ontario Pre-Start Health and Safety Review under O. Reg. 851, and identify the documentation a professional engineer will expect.

## Introduction

A Tier 1 automotive supplier in Cambridge, Ontario is bringing up a new robot cell on its door subassembly line. A six-axis robot, mounted on a short pedestal at the edge of the fixture, needs to place five spot welds on a door inner panel before the part indexes forward to the next station. On paper this looks like a single purchase order: pick a robot, pick a gun, bolt it down, program it. In practice, before that cell is allowed to run production, someone has to answer seven separate engineering questions, and getting any one of them wrong either stops the line or puts someone in the hospital.

Can the robot actually reach every weld point, with the gun attached, from where the pedestal sits — not just barely, but with enough margin that a small fixture tolerance doesn't turn a reachable point into an unreachable one? Can the wrist carry the gun's weight at the distance it sits off the flange, or does that offset quietly cut the usable payload to a fraction of the nameplate rating? Can the whole weld sequence — five moves, five welds, a return to home — finish inside the time the line allows it, cycle after cycle, all shift long? If a person opens the gate into that cell while the robot is moving, how far back does the light curtain have to sit before the robot can stop moving before that person's hand reaches the weld gun? Given the different kinds of access the cell needs — routine operator access at one point, rare maintenance access at another, no access at all along most of the perimeter — what guarding technology belongs at each one? If someone suggests swapping the robot for a "collaborative" model to avoid the guarding altogether, does that actually remove the need for a risk assessment? And once the cell is designed, built, and ready to run, does Ontario law require a professional engineer to review it before it's allowed to operate?

This chapter works through all seven of these design tasks in the order a project team actually encounters them: reach, payload, cycle time, minimum safe distance, guarding selection, collaborative-operation limits, and the regulatory sign-off. The fourth of these — the ISO 13855 minimum safe distance calculation — is the anchor calculation of the chapter, and it depends on one number this chapter does not calculate: the robot's stopping time. That number comes out of the safety circuit design itself — the category of stop, the response time of the safety relay or safety PLC, the braking behavior of the servo drives — which is covered in the [Machine Safety Circuits](06-02-Machine-Safety-Circuits.qmd) chapter. Here, the stopping time is treated as a given input, the way it would be for a mechanical or industrial engineer who receives it from the controls team and has to turn it into a fence line on a layout drawing.

## Design Task 1: Work Envelope and Reach Verification

### Scenario

The door subassembly cell's robot sits on a 300 mm steel pedestal at the edge of the fixture, offset from the plant's cell-layout reference origin. Five weld points are specified on the door inner panel, taken directly from the CAD model in cell coordinates. Before the integrator finalizes the pedestal location and orders the robot, the layout engineer needs to confirm that every one of those five points is reachable — not marginally reachable, but reachable with enough margin that the robot isn't fighting a wrist singularity at the exact point where precise gun placement matters most.

### Design Guidelines and Rules of Thumb

A robot manufacturer's "reach" specification — the number on the datasheet, and the radius used to draw the reach envelope in ISO 9946's presentation format — is measured from the centerline of axis 1 to the **wrist center point (WCP)**, not to the tip of whatever tool is bolted to the flange. Any end-effector — a spot-welding gun, a servo gripper, a dispensing nozzle — adds a further offset that has to be accounted for separately. Early in a project, before the exact approach orientation for every point has been finalized in CAD, the conservative approach is to simply add the tool length to the required reach rather than try to subtract it based on an orientation that might still change. This overstates the reach requirement for orientations where the tool happens to point back toward the robot, but it avoids finding out late in commissioning that a "reachable" point actually isn't, once the real tool orientation is locked in.

The rated maximum reach is also not a hard wall. Near the outer edge of the envelope, the wrist loses degrees of freedom — the arm runs out of usable orientation flexibility exactly where full flexibility is often needed most, to aim a welding gun or gripper at an odd angle. The reach envelope is also not a full sphere: most robots have a minimum-reach dead zone close to the base where the arm folds against itself.

**Rule of thumb:** design to a reach margin of at least 10% below the manufacturer's rated maximum reach for any point that requires a specific tool orientation, not just a bare touch. This margin degrades quickly for tasks that need a wide range of approach angles at the same point (e.g., a gun that has to swing through several orientations to clear a flange) — for those, a full CAD reach study with the actual tool model is worth doing even when the straight-line distance check passes comfortably.

CSA Z434 (the Canadian standard for industrial robots and robot systems) expects the work envelope to be documented as part of the cell's layout and risk assessment; ISO 10218 and ANSI RIA R15.06 are the international and US equivalents.

### Governing Calculation

For a target point $P$ and a robot mounting position (the axis-1 centerline) at $B$, both expressed in the same cell coordinate system:

$$
D = \sqrt{(x_P - x_B)^2 + (y_P - y_B)^2 + (z_P - z_B)^2}
$$

$$
\text{Required reach} = D + L_{tool}
$$

$$
\text{Margin} = \frac{R_{max} - \text{Required reach}}{R_{max}} \times 100\%
$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $P$ | Target point (weld point, pick point, etc.) in cell coordinates | mm |
| $B$ | Robot mounting position (axis-1 centerline), in cell coordinates | mm |
| $D$ | Straight-line distance from base to target | mm |
| $L_{tool}$ | Tool length, added conservatively regardless of orientation | mm |
| $R_{max}$ | Manufacturer-rated maximum reach (to WCP) | mm |
| Margin | Fraction of rated reach left unused after the tool-length allowance | % |

### Worked Example

```{r reach-check-06-01, echo=TRUE}
## Robot mounting position (axis-1 centerline), cell reference coordinates, mm
base <- c(x = 1200, y = 0, z = 300)   # pedestal riser adds 300 mm in z

## Farthest required weld point on the door subassembly, mm
p5 <- c(x = 3100, y = 850, z = 650)

## Spot-welding gun (tool) length from wrist flange to electrode tip, mm
L_tool <- 380

## Manufacturer-rated maximum reach (to wrist center point), mm
R_max <- 2655

## Straight-line distance from base to the target point
d_vec <- p5 - base
D <- sqrt(sum(d_vec^2))

## Conservative worst-case required reach: add the tool length rather than
## subtract it, since the exact approach orientation isn't finalized yet
required_reach <- D + L_tool
margin_pct <- (R_max - required_reach) / R_max * 100

cat("Vector from base to P5 (mm): dx =", d_vec["x"], " dy =", d_vec["y"], " dz =", d_vec["z"], "\n")
cat("Straight-line distance D =", format(round(D, 1), big.mark = " "), "mm\n")
cat("Required reach (D + tool length) =", format(round(required_reach, 1), big.mark = " "), "mm\n")
cat("Rated maximum reach R_max =", format(R_max, big.mark = " "), "mm\n")
cat("Margin =", sprintf("%.1f", margin_pct), "%\n\n")

if (margin_pct >= 10) {
  cat("Margin >= 10% rule-of-thumb threshold: PASS\n")
} else {
  cat("Margin < 10% rule-of-thumb threshold: FAILS the margin check\n")
  cat("(even though the required reach is still less than R_max)\n\n")

  ## Corrective action: move the pedestal 300 mm closer in x
  base2 <- base
  base2["x"] <- base2["x"] + 300
  d_vec2 <- p5 - base2
  D2 <- sqrt(sum(d_vec2^2))
  required2 <- D2 + L_tool
  margin2_pct <- (R_max - required2) / R_max * 100

  cat("Corrective action: relocate the pedestal +300 mm in x\n")
  cat("New required reach =", format(round(required2, 1), big.mark = " "), "mm\n")
  cat("New margin =", sprintf("%.1f", margin2_pct), "%\n")
  if (margin2_pct >= 10) cat("New margin >= 10%: PASS\n")
}
```

**Sanity check:** the required reach (about 2.49 m) is still comfortably less than the rated maximum reach (2.655 m) — a naive check that just compares required reach to $R_{max}$ would say this cell passes. It's only the 10% margin rule of thumb that catches the problem: this point sits right at the edge of the dexterous envelope, where wrist orientation flexibility is limited. Moving the pedestal 300 mm closer restores a comfortable margin without changing the robot model or the fixture design.

```{r fig-06-01-cell-layout, echo=FALSE, fig.cap="Plan view of a spot-welding robot cell: rated reach envelope (dashed), a 10% margin band (dotted), and the five required weld points, with the farthest point (P5) driving the reach check. Schematic, not to scale.", fig.align='center', fig.width=9, fig.height=7}
base_x <- 0
base_y <- 0
r_max  <- 4.2

targets <- data.frame(
  x = c(3.6, 2.0, -1.6, 0.6, 3.35),
  y = c(1.55, -2.35, 2.55, -3.05, -0.55),
  label = c("P1", "P2", "P3", "P4", "P5 (farthest)")
)
worst <- targets[5, ]

## Envelope and margin-band circles: geom_circle() (not annotate()) because
## ggforce's circle geom relies on a stat transform (x0/y0/r -> x/y path)
## that annotate()'s forced stat="identity" bypasses
envelope_circles <- data.frame(
  x0 = c(base_x, base_x),
  y0 = c(base_y, base_y),
  r = c(r_max, r_max * 0.9),
  kind = c("envelope", "margin")
)

ggplot() +
  geom_circle(data = envelope_circles, aes(x0 = x0, y0 = y0, r = r, linetype = kind, color = kind),
              linewidth = 1, fill = NA, show.legend = FALSE) +
  scale_linetype_manual(values = c(envelope = "dashed", margin = "dotted")) +
  scale_color_manual(values = c(envelope = "steelblue", margin = "grey55")) +
  annotate("rect", xmin = base_x - 0.35, xmax = base_x + 0.35,
           ymin = base_y - 0.35, ymax = base_y + 0.35, fill = "grey30") +
  annotate("text", x = base_x, y = base_y - 0.75, label = "Robot base\n(pedestal mount)",
           size = 3.2 * diagram_text_scale) +
  annotate("segment", x = base_x, xend = worst$x, y = base_y, yend = worst$y,
           linewidth = 0.9, color = "firebrick", linetype = "dashed") +
  geom_point(data = targets, aes(x = x, y = y), size = 3, color = "black") +
  geom_text(data = targets, aes(x = x, y = y, label = label), vjust = -1.1,
            size = 3.2 * diagram_text_scale) +
  annotate("text", x = r_max * 0.62, y = r_max * 0.62, label = "R_max (rated reach)",
           color = "steelblue", size = 3 * diagram_text_scale) +
  annotate("text", x = -r_max * 0.55, y = r_max * 0.78, label = "10% margin band",
           color = "grey45", size = 2.7 * diagram_text_scale) +
  coord_fixed() +
  theme_void() +
  labs(title = "Reach Envelope and Weld Points") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14 * diagram_text_scale))
```

## Design Task 2: Payload Check Including End-Effector CG Offset

### Scenario

At a beverage co-packer's distribution centre outside London, Ontario, a palletizing robot picks finished cases from an infeed conveyor and stacks them onto outbound pallets. The gripper is a case-turning vacuum head — it has to rotate a case 90 degrees mid-pick to match the pallet pattern, which means the mechanism that does the turning extends well off the wrist flange. The robot's nameplate payload rating looks generous next to the actual weight being carried. The question is whether that nameplate number is even the right number to compare against.

### Design Guidelines and Rules of Thumb

A robot's rated payload is only fully available when the load's center of gravity sits at (or very close to) the wrist flange centerline. As soon as the combined center of gravity of the end-effector and workpiece moves away from that centerline — horizontally, vertically, or both — the wrist has to resist a growing overturning moment, and the manufacturer's payload-vs-CG-offset chart derates the allowable payload accordingly. These charts decline sharply, not gently: doubling the offset can cut the allowable payload by half or more, because the wrist bearings and gearbox are the limiting factor, not the servo motor's torque.

This chapter uses a simplified but physically grounded model of that relationship: the wrist can sustain a roughly constant **mass-moment capacity** (nameplate payload multiplied by a small reference offset) once the offset exceeds a flat region near the flange. Treat this as the correct *shape* of the real curve, not the real numbers — always pull the actual payload-vs-CG-offset chart from the specific robot's datasheet for a final design; the simplified curve here is for building intuition about how sharply these curves fall off, not for signing off a design.

**Rule of thumb:** never accept "the total mass is well under the rated payload" as proof that a design passes. Nameplate payload only applies at (near) zero offset — any real gripper has some offset, and it's entirely possible to fail a payload check at a fraction of the nameplate rating once that offset is accounted for. This rule of thumb breaks down at very small offsets (under roughly the flat-region boundary on the chart, typically tens of millimeters) where the derating essentially doesn't apply yet — for compact, centered grippers, the nameplate number really is the number.

Real datasheets also derate for vertical offset (how far below or above the flange face the CG sits) and for dynamic effects — the additional moment generated by acceleration during a fast move, on top of the static moment from gravity alone. Both matter in a complete design; this chapter's worked example checks the static horizontal-offset case only, which is the effect that most often surprises a first-pass design.

### Governing Calculation

$$
M_{cap} = W_{max0} \times L_0
$$

$$
Payload_{max}(L) =
\begin{cases}
W_{max0} & L \le L_0 \\[4pt]
\dfrac{M_{cap}}{L} & L > L_0
\end{cases}
$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $W_{max0}$ | Rated payload at (near) zero CG offset, from the datasheet | kg |
| $L_0$ | Offset below which the full rated payload still applies | mm |
| $M_{cap}$ | Constant allowable mass-moment beyond $L_0$ (illustrative model) | kg·mm |
| $L$ | Actual combined CG offset of end-effector + workpiece from the flange centerline | mm |
| $Payload_{max}(L)$ | Maximum allowable total carried mass at offset $L$ | kg |

```{r fig-06-01-cg-schematic, echo=FALSE, fig.cap="Side-view schematic of an offset end-effector: the combined center of gravity of gripper and workpiece sits a horizontal distance L from the wrist flange centerline, creating an overturning moment about the wrist. Schematic, not to scale.", fig.align='center', fig.width=9, fig.height=5}
ggplot() +
  annotate("rect", xmin = -0.3, xmax = 0.3, ymin = 3.6, ymax = 4.2, fill = "grey30") +
  annotate("text", x = 0, y = 4.55, label = "Wrist flange", size = 3.2 * diagram_text_scale) +
  annotate("segment", x = 0, xend = 0, y = 3.6, yend = 0.2, linetype = "dashed",
           color = "grey50", linewidth = 0.6) +
  annotate("segment", x = 0, xend = 2.4, y = 3.6, yend = 3.0, linewidth = 3, color = "steelblue") +
  annotate("segment", x = 2.4, xend = 2.4, y = 3.0, yend = 1.6, linewidth = 3, color = "steelblue") +
  annotate("rect", xmin = 1.9, xmax = 2.9, ymin = 0.9, ymax = 1.6, fill = "tan", color = "black") +
  annotate("text", x = 2.4, y = 1.25, label = "Gripper +\ncase", size = 2.8 * diagram_text_scale) +
  annotate("segment", x = 2.4, xend = 2.4, y = 0.9, yend = 0.2,
           arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
           linewidth = 1, color = "firebrick") +
  annotate("text", x = 2.75, y = 0.45, label = "W", color = "firebrick", size = 3.5 * diagram_text_scale) +
  annotate("segment", x = 0, xend = 2.4, y = -0.3, yend = -0.3,
           arrow = arrow(length = unit(0.15, "cm"), ends = "both", type = "closed"), linewidth = 0.8) +
  annotate("text", x = 1.2, y = -0.65, label = "L (CG offset)", size = 3.2 * diagram_text_scale) +
  annotate("curve", x = -0.6, xend = -0.4, y = 3.0, yend = 3.5, curvature = -0.5,
           color = "darkorange", linewidth = 1) +
  annotate("text", x = -1.05, y = 3.2, label = "Overturning\nmoment", color = "darkorange",
           size = 2.8 * diagram_text_scale) +
  coord_fixed(xlim = c(-1.7, 3.4), ylim = c(-1, 5)) +
  theme_void()
```

```{r fig-06-01-payload-curve, echo=FALSE, fig.cap="Illustrative payload-vs-CG-offset derating curve: rated payload holds at W_max0 out to L0, then falls off sharply. The marked point is the gripper's actual combined mass and CG offset, which lies above (outside) the allowable curve.", fig.align='center', fig.width=9, fig.height=6}
W_max0 <- 130   # kg, rated payload for offsets at/below L0
L0     <- 60    # mm, flat region boundary
M_cap  <- W_max0 * L0  # kg*mm, constant allowable mass-moment beyond L0

L_seq <- seq(5, 400, by = 5)
payload_max <- ifelse(L_seq <= L0, W_max0, M_cap / L_seq)
curve_df <- data.frame(L = L_seq, payload = payload_max)

actual_L <- 210
actual_W <- 44

ggplot(curve_df, aes(x = L, y = payload)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  annotate("point", x = actual_L, y = actual_W, color = "firebrick", size = 4, shape = 4, stroke = 1.5) +
  annotate("text", x = actual_L + 15, y = actual_W + 10,
           label = "Actual: 44 kg at\nL = 210 mm (FAILS)", color = "firebrick",
           size = 3 * diagram_text_scale, hjust = 0) +
  labs(title = "Payload vs. CG Offset (Illustrative)",
       x = "Horizontal CG offset, L (mm)", y = "Maximum allowable payload (kg)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

### Worked Example

Live cell — try changing `gripper_mass`, `case_mass`, or `L_offset` and re-run to see when the check flips from fail to pass.

```{webr-r}
#| label: payload-check-06-01
## Rated payload at near-zero offset (kg) and the flat-region boundary (mm)
W_max0 <- 130
L0 <- 60
M_cap <- W_max0 * L0  # kg*mm, constant allowable mass-moment beyond L0

## End-effector and workpiece
gripper_mass <- 28   # kg, case-turning vacuum gripper
case_mass <- 16       # kg, one case of canned product
total_mass <- gripper_mass + case_mass

## Combined CG offset from the wrist flange centerline (mm) -- simplified by
## assuming the case's CG sits close enough to the gripper's own CG that a
## single combined offset is a reasonable estimate for this stage of design
L_offset <- 210

payload_max <- if (L_offset <= L0) W_max0 else M_cap / L_offset

cat("Total carried mass:", total_mass, "kg\n")
cat("CG offset from flange:", L_offset, "mm\n")
cat("Allowable mass-moment, M_cap =", W_max0, "kg x", L0, "mm =", M_cap, "kg-mm\n")
cat("Allowable payload at this offset =", round(payload_max, 1), "kg\n\n")

if (total_mass <= payload_max) {
  cat("Total mass <= allowable payload: PASS\n")
} else {
  cat("Total mass exceeds allowable payload at this offset: FAILS\n")
  cat("(Note: 44 kg is well under the", W_max0, "kg flange rating --\n")
  cat(" the failure is entirely due to CG offset.)\n\n")

  max_allowable_offset <- M_cap / total_mass
  reduction_needed <- L_offset - max_allowable_offset
  cat("Maximum allowable offset at", total_mass, "kg =", round(max_allowable_offset, 1), "mm\n")
  cat("Offset must be reduced by at least", round(reduction_needed, 1), "mm (",
      round(reduction_needed / L_offset * 100, 1), "% )\n")
}
```

**Sanity check:** 44 kg is barely a third of the robot's 130 kg nameplate rating, which is exactly why this failure is easy to miss on a first pass — the fix isn't a bigger robot, it's a more compact gripper, a counterweight to pull the combined CG back toward the flange, or a robot family with a flatter payload-vs-offset curve.

## Design Task 3: Cycle Time Estimation from Move Profiles

### Scenario

The same door-panel welding robot has to complete its full weld sequence — home to P1, P1 to P2, and so on through P5, then back to home — inside a fixed slice of the station's overall cycle time, because five other operations share this multi-robot station and the whole station has to hit a 60 jobs-per-hour (JPH) line rate. The process engineer needs a cycle time estimate before the robot program even exists, to confirm the motion plan is feasible before committing to the layout.

### Design Guidelines and Rules of Thumb

Robot point-to-point moves don't travel at a constant speed — they accelerate up to a commanded maximum speed, hold it if the move is long enough, then decelerate into the target point. This is the same **trapezoidal velocity profile** used in servo motion generally: a ramp up, a plateau, a ramp down. If the move is too short to ever reach the commanded speed, the profile collapses into a triangle — the robot accelerates for half the move and decelerates for the other half, never touching the top speed at all.

Cycle time is the sum of every move's time, plus every dwell (process) time along the way — weld time, glue-dispense time, part-present sensor wait, I/O handshakes with the cell PLC. It's easy to model the moves carefully and then forget that the welds themselves take real time too; on a short-cycle cell, the dwell time often dominates the total, not the motion.

**Rule of thumb:** pad a first-pass kinematic cycle time estimate by 10-15% before comparing it to a hard rate requirement, to cover real-world variability the model doesn't capture — PLC scan time, minor part-to-part variation, network jitter on the safety and I/O buses. This rule of thumb breaks down on cells with very tight rate requirements (sub-second margins), where a flat percentage pad isn't precise enough and the team needs actual timed cycles from a simulation tool or a physical run-off instead.

### Governing Calculation

For a single point-to-point move of distance $d$, with commanded maximum speed $v_{max}$ and acceleration/deceleration $a$:

$$
d_{accel} = \frac{v_{max}^2}{2a}
$$

$$
t_{move} =
\begin{cases}
\dfrac{2v_{max}}{a} + \dfrac{d - 2d_{accel}}{v_{max}} & d \ge 2d_{accel} \ \text{(trapezoidal)} \\[8pt]
2\sqrt{\dfrac{d}{a}} & d < 2d_{accel} \ \text{(triangular)}
\end{cases}
$$

$$
t_{cycle} = \sum_{i} t_{move,i} + \sum_{j} t_{dwell,j}
$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $d$ | Distance for one point-to-point move | mm |
| $v_{max}$ | Commanded maximum TCP speed | mm/s |
| $a$ | Commanded acceleration and deceleration (assumed equal) | mm/s² |
| $d_{accel}$ | Distance covered during the acceleration phase alone | mm |
| $t_{move}$ | Time for one move | s |
| $t_{dwell}$ | Process/dwell time at a point (weld, dispense, handshake, etc.) | s |
| $t_{cycle}$ | Total cell cycle time | s |

```{r fig-06-01-velocity-profile, echo=FALSE, fig.cap="Trapezoidal velocity profile for a single point-to-point robot move: acceleration, constant-velocity, and deceleration phases. Move time is simply the sum of the three phase durations.", fig.align='center', fig.width=9, fig.height=5}
d_ex <- 800      # mm
v_max_ex <- 1200 # mm/s
a_ex <- 4000     # mm/s^2

t_accel_ex <- v_max_ex / a_ex
d_accel_ex <- v_max_ex^2 / (2 * a_ex)
t_const_ex <- (d_ex - 2 * d_accel_ex) / v_max_ex
t_total_ex <- 2 * t_accel_ex + t_const_ex

t_seq <- seq(0, t_total_ex, length.out = 300)
v_seq <- ifelse(
  t_seq <= t_accel_ex, a_ex * t_seq,
  ifelse(t_seq <= t_accel_ex + t_const_ex, v_max_ex,
         v_max_ex - a_ex * (t_seq - t_accel_ex - t_const_ex))
)
profile_df <- data.frame(t = t_seq, v = v_seq)

ggplot(profile_df, aes(x = t, y = v)) +
  geom_area(fill = "steelblue", alpha = 0.25) +
  geom_line(color = "steelblue", linewidth = 1.3) +
  annotate("segment", x = t_accel_ex, xend = t_accel_ex, y = 0, yend = v_max_ex,
           linetype = "dotted", color = "grey40") +
  annotate("segment", x = t_accel_ex + t_const_ex, xend = t_accel_ex + t_const_ex,
           y = 0, yend = v_max_ex, linetype = "dotted", color = "grey40") +
  annotate("text", x = t_accel_ex / 2, y = v_max_ex * 0.4, label = "Accel",
           size = 3.2 * diagram_text_scale) +
  annotate("text", x = t_accel_ex + t_const_ex / 2, y = v_max_ex * 1.06,
           label = "Constant velocity", size = 3.2 * diagram_text_scale) +
  annotate("text", x = t_accel_ex + t_const_ex + (t_total_ex - t_accel_ex - t_const_ex) / 2,
           y = v_max_ex * 0.4, label = "Decel", size = 3.2 * diagram_text_scale) +
  labs(title = "Trapezoidal Point-to-Point Velocity Profile",
       x = "Time (s)", y = "TCP speed (mm/s)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14 * diagram_text_scale, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale)
  )
```

### Worked Example

Live cell — try changing `v_max`, `a`, or the `distances` vector and re-run to see the cycle time and margin update.

```{webr-r}
#| label: cycle-time-calc-06-01
## Move distances between successive weld points, mm (home -> P1 -> ... -> P5 -> home)
distances <- c(450, 620, 380, 710, 500, 900)

v_max <- 1500   # mm/s, commanded TCP speed
a <- 5000       # mm/s^2, commanded acceleration/deceleration

move_time <- function(d, v_max, a) {
  d_accel <- v_max^2 / (2 * a)
  if (d >= 2 * d_accel) {
    ## Trapezoidal: reaches v_max
    t_accel <- v_max / a
    t_const <- (d - 2 * d_accel) / v_max
    return(2 * t_accel + t_const)
  } else {
    ## Triangular: never reaches v_max
    v_peak <- sqrt(a * d)
    return(2 * v_peak / a)
  }
}

move_times <- sapply(distances, move_time, v_max = v_max, a = a)
total_move_time <- sum(move_times)

weld_dwell_each <- 0.35   # s, squeeze + weld + hold per point
n_welds <- 5
total_dwell <- weld_dwell_each * n_welds

handshake_time <- 0.30    # s, I/O handshake with the cell PLC

total_cycle <- total_move_time + total_dwell + handshake_time

cat("Move times (s):", paste(round(move_times, 3), collapse = ", "), "\n")
cat("Total move time:", round(total_move_time, 2), "s\n")
cat("Total weld dwell:", round(total_dwell, 2), "s (", n_welds, "welds x", weld_dwell_each, "s )\n")
cat("PLC handshake time:", handshake_time, "s\n")
cat("Total cycle time:", round(total_cycle, 2), "s\n\n")

allotment <- 9.5  # s, this robot's share of the multi-robot station's 60 JPH target
margin_s <- allotment - total_cycle
margin_pct <- margin_s / allotment * 100

cat("Cycle time allotment:", allotment, "s\n")
cat("Margin:", round(margin_s, 2), "s (", round(margin_pct, 1), "% )\n")
if (total_cycle <= allotment) {
  cat("Cycle time fits within the allotment: PASS\n")
} else {
  cat("Cycle time exceeds the allotment: FAILS -- rebalance the station or raise speed/accel\n")
}
```

**Sanity check:** roughly 6.2 seconds of robot motion and dwell time against a 9.5 second allotment leaves a healthy margin — comfortable enough to absorb the 10-15% real-world pad from the rule of thumb above without threatening the line rate.

## Design Task 4: Minimum Safe Separation Distance (ISO 13855)

### Scenario

The door-panel cell's operator access point needs a safety light curtain, and the layout engineer has to decide exactly how far back from the robot's reach envelope that curtain can be mounted. Mount it too close, and a person can cross the detection plane and reach the hazard before the robot finishes stopping. Mount it unnecessarily far back, and the cell eats floor space the plant doesn't have to spare. This is the calculation that settles the question with a number instead of a guess, and it is the anchor calculation of this chapter.

### Design Guidelines and Rules of Thumb

The logic behind the minimum safe distance is simple even though the standard's presentation can look intimidating: a guard has to be far enough from the hazard that, by the time a person crosses the detection plane, the machine has *completely* stopped before that person's body (or the fastest-moving part of it) can reach the hazard. The distance a person can cover during the response comes from an approach speed, and the standard specifies that speed rather than leaving it to the designer to guess — this chapter uses ISO 13855's specified value directly rather than treating it as a rule of thumb.

The stopping time input, $T$, is not something this chapter calculates — it comes out of the safety circuit design covered in the Machine Safety Circuits chapter: the robot's own braking/stop-category response time, plus the response time of the detection device itself (a light curtain's own signal processing time, for instance), plus any additional safety relay or safety PLC processing time in the chain. Every one of those pieces adds directly to $T$, and $T$ is multiplied by the approach speed, so shaving milliseconds off the safety circuit's response time has an outsized effect on how much floor space the guarding needs.

The additional distance $C$ accounts for something the pure speed-times-time term doesn't: a coarse-resolution detection device (an area scanner tuned to detect a torso-sized object, say) can miss a hand or arm reaching through its detection field before triggering a stop. ISO 13855 defines this intrusion-depth allowance based on the detection capability (resolution) of the specific guarding device — a fine-resolution light curtain typically requires a smaller $C$ than a coarse-resolution area scanner, because it catches a smaller intrusion closer to the detection plane. The exact value is read from a table in the standard, keyed to the device's detection resolution; this chapter does not reproduce that table, only the relationship it expresses.

**Rule of thumb:** as a rough sanity check, expect $S$ to land somewhere in the 1.5-3 m range for a typical light-curtain-guarded cell with a sub-second overall stopping time. If a calculated $S$ comes out under a meter, double-check units before trusting it — the most common error here is mixing millimeters and meters somewhere in the calculation, not an error in the physics.

**Rule of thumb:** if the calculated $S$ doesn't fit in the available floor space, fix it by shortening $T$ (a faster stop category, or a faster-responding detection device) or by choosing a finer-resolution device to shrink $C$ — not by trying to argue down $K$, which is fixed by the standard and not a design variable.

### Governing Calculation

$$
S = K \times T + C
$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $S$ | Minimum safe separation distance from the detection plane to the hazard | mm |
| $K$ | Approach speed constant specified by ISO 13855 (1600 mm/s for whole-body approach) | mm/s |
| $T$ | Overall system stopping time: robot stop time + detection device response time + any additional safety-logic processing time | s |
| $C$ | Additional intrusion-depth distance, dependent on the detection capability (resolution) of the guarding device, per ISO 13855 | mm |

```{r fig-06-01-s-kt-c-schematic, echo=FALSE, fig.cap="ISO 13855 minimum separation distance: the guard must stand back far enough that a person can be detected and the robot fully stopped before reach reaches the hazard, plus an added intrusion-depth allowance C. Schematic, not to scale.", fig.align='center', fig.width=10, fig.height=4.5}
ggplot() +
  annotate("rect", xmin = 7.5, xmax = 9.5, ymin = 0.3, ymax = 2.3, fill = "mistyrose", color = "firebrick") +
  annotate("text", x = 8.5, y = 2.6, label = "Hazard zone\n(robot reach envelope)",
           size = 3 * diagram_text_scale, color = "firebrick") +
  annotate("segment", x = 2.0, xend = 2.0, y = 0.1, yend = 2.5, linewidth = 2, color = "steelblue") +
  annotate("text", x = 2.0, y = 2.8, label = "Light curtain\ndetection plane",
           size = 3 * diagram_text_scale, color = "steelblue") +
  annotate("segment", x = 2.0, xend = 7.5, y = -0.4, yend = -0.4,
           arrow = arrow(length = unit(0.15, "cm"), ends = "both", type = "closed"), linewidth = 0.8) +
  annotate("text", x = 4.75, y = -0.8, label = "S = K x T + C",
           size = 3.4 * diagram_text_scale, fontface = "bold") +
  annotate("segment", x = 2.0, xend = 5.6, y = 0.0, yend = 0.0,
           arrow = arrow(length = unit(0.12, "cm"), ends = "both", type = "closed"),
           linewidth = 0.6, color = "grey40") +
  annotate("text", x = 3.8, y = 0.35, label = "K x T\n(distance covered while\nstopping + detecting)",
           size = 2.6 * diagram_text_scale, color = "grey30") +
  annotate("segment", x = 5.6, xend = 7.5, y = 0.0, yend = 0.0,
           arrow = arrow(length = unit(0.12, "cm"), ends = "both", type = "closed"),
           linewidth = 0.6, color = "darkorange") +
  annotate("text", x = 6.55, y = 0.35, label = "C\n(intrusion\nallowance)",
           size = 2.6 * diagram_text_scale, color = "darkorange") +
  annotate("point", x = 0.8, y = 1.2, size = 5, color = "grey20") +
  annotate("segment", x = 0.8, xend = 1.9, y = 1.2, yend = 1.2,
           arrow = arrow(length = unit(0.15, "cm"), type = "closed"),
           linewidth = 0.8, color = "grey20", linetype = "dashed") +
  annotate("text", x = 1.3, y = 1.55, label = "approach", size = 2.6 * diagram_text_scale) +
  coord_fixed(xlim = c(0, 10), ylim = c(-1.1, 3.2)) +
  theme_void()
```

### Worked Example

Live cell — try changing `T_robot`, `C`, or `planned_distance` and re-run to see whether the design passes.

```{webr-r}
#| label: separation-distance-06-01
## Overall system stopping time T = robot stop time + detection device response time
## (robot stop time comes from the safety circuit design -- see the Machine Safety
## Circuits chapter -- and is treated as a given input here)
T_robot <- 0.65   # s, safety-rated monitored stop time for this cell
T_device <- 0.04  # s, light curtain response time (typical Type 4 device)
T_total <- T_robot + T_device

## ISO 13855 approach speed constant (standard-specified, not a rule of thumb)
K <- 1600  # mm/s

## Intrusion-depth allowance for a light curtain with 30 mm detection
## resolution (illustrative value -- consult the current edition of
## ISO 13855's intrusion-distance table for the exact figure for a given device)
C <- 850   # mm

S_required <- K * T_total + C

cat("Robot stopping time, T_robot =", T_robot, "s\n")
cat("Light curtain response time, T_device =", T_device, "s\n")
cat("Total stopping time, T =", T_total, "s\n")
cat("Approach speed, K =", K, "mm/s\n")
cat("Intrusion allowance, C =", C, "mm\n")
cat("Required minimum separation, S = K x T + C =",
    format(round(S_required, 0), big.mark = " "), "mm\n\n")

planned_distance <- 1800  # mm, distance the layout currently plans to mount the curtain

if (planned_distance >= S_required) {
  cat("Planned mounting distance (", planned_distance, "mm ) >= S: PASS\n")
} else {
  shortfall <- S_required - planned_distance
  cat("Planned mounting distance (", planned_distance, "mm ) < S: FAILS by",
      round(shortfall, 0), "mm\n\n")

  cat("Option A -- move the light curtain out to at least",
      format(ceiling(S_required / 50) * 50, big.mark = " "), "mm\n\n")

  ## Option B: switch to a finer-resolution light curtain with a smaller C
  C_fine <- 650  # mm, illustrative value for a finer-resolution device
  S_fine <- K * T_total + C_fine
  cat("Option B -- use a finer-resolution light curtain (C =", C_fine, "mm ):\n")
  cat("  New required S =", format(round(S_fine, 0), big.mark = " "), "mm\n")
  if (planned_distance >= S_fine) {
    cat("  Planned distance now passes without moving the curtain.\n")
  }
}
```

**Sanity check:** the required distance (about 1.95 m) falls squarely in the 1.5-3 m range the rule of thumb above predicts for a sub-second stopping time — a useful check that no units got mixed up. The planned 1.8 m mounting distance falls just short, by less than 10%, which is exactly the kind of near-miss that a rushed layout review can wave through if the calculation isn't done explicitly.

## Design Task 5: Guarding Technology Selection

### Scenario

The same door-panel cell has three distinct access points around its perimeter: a far side with no routine access at all, the operator load/unload station where a person reaches in every cycle to place a part, and an aisle where an AGV delivers empty door-panel racks and occasionally a person walks through on foot. One guarding technology does not fit all three.

### Design Guidelines and Rules of Thumb

Fixed fencing, safety light curtains, and safety-rated area scanners are the three guarding technologies that show up over and over in robot cell design, and each one solves a different problem well. Fixed fencing is the default for any boundary that doesn't need a routine opening — it's the cheapest option per linear meter of perimeter and the easiest to specify, but it is genuinely fixed: moving an access point later means cutting fence and re-installing it. A light curtain earns its keep at a point that needs frequent, fast, hands-free access — an operator loading and unloading a fixture — but it's a simple plane of detection, vulnerable to misalignment and to anything that can block a beam (dust, spatter, or in a food plant, steam or wash-down spray). An area scanner trades a higher unit cost for the ability to define multiple zones — a slow-down warning zone and a further-in hard-stop zone — from a single sensor, covering an irregular perimeter shape that fencing or a curtain can't follow economically; this makes it the natural fit for an AGV aisle that needs a graduated response rather than a single trip point.

Fixed fencing still needs interlocked access doors wherever people cross it routinely, and any opening or mesh aperture in that fencing has to respect an opening-to-standoff-distance relationship so that a hand or arm can't reach through to the hazard — CSA Z434 and ISO 13857 define that relationship; this chapter does not reproduce the table.

**Rule of thumb:** default to fixed fencing for any boundary segment without a routine access need, and only reach for a light curtain or area scanner where fencing genuinely can't do the job — access frequency, aisle geometry, or a need for graduated (warn-then-stop) response. This default breaks down where floor space is tight enough that fencing's larger footprint (posts, base plates, swing clearance on gates) is itself the constraint — in a cramped cell, a scanner's small footprint can be worth its higher cost even at a rarely-used boundary.

CSA Z434 is the primary Canadian reference for guarding selection on industrial robot systems; ISO 10218 and ANSI RIA R15.06 are the international and US equivalents.

```{r tbl-06-01-guarding-tradeoff, echo=FALSE, message=FALSE}
guarding_options <- tribble(
  ~Guarding_Type, ~Typical_Cost, ~Footprint, ~Flexibility, ~Best_Fit_Use, ~Key_Limitation,
  "Fixed fencing", "Low ($/m of perimeter)", "Fixed, often the largest footprint",
  "None -- relocating it means cutting and re-installing fence",
  "Rarely-accessed perimeter with a stable, high-hazard boundary",
  "No convenient pass-through for material or people",
  "Safety light curtain", "Moderate", "Small (a detection plane at the opening)",
  "Good -- can be relocated if the opening moves",
  "Frequent operator access at a fixed load/unload point",
  "Sensitive to misalignment, dust, and blocked beams from debris or steam",
  "Safety-rated area scanner", "High", "Small physical footprint, protects an irregular shape",
  "Best -- multiple software-defined zones (warn/slow/stop) from one sensor",
  "Irregular perimeters, AGV/forklift aisles, cells needing a reduced-speed zone before full stop",
  "Highest unit cost; performance depends on floor reflectivity and mounting height"
)

guarding_options %>%
  kbl(caption = "Guarding technology trade-offs for robot cell perimeters",
      col.names = c("Guarding Type", "Typical Installed Cost", "Footprint",
                     "Flexibility / Reconfigurability", "Best-Fit Use Case", "Key Limitation")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, background = "#2C3E50", color = "white")
```

### Worked Example

Live cell — try changing the `access_frequency` or `needs_reconfig_zones` values and re-run to see the recommendation change.

```{webr-r}
#| label: guarding-selection-06-01
## Three access points around the cell, with the attributes that drive the
## guarding decision
access_points <- data.frame(
  point = c("Far perimeter (no routine access)",
            "Operator load/unload station",
            "AGV pallet rack aisle"),
  access_frequency = c("Never", "Every cycle (high)", "Continuous AGV traffic, rare person"),
  needs_reconfig_zones = c(FALSE, FALSE, TRUE),
  stringsAsFactors = FALSE
)

for (i in seq_len(nrow(access_points))) {
  ap <- access_points[i, ]
  cat("Access point:", ap$point, "\n")
  if (ap$access_frequency == "Never") {
    choice <- "Fixed fencing"
    reason <- "no routine opening is needed, so the lowest-cost option satisfies the hazard."
  } else if (ap$needs_reconfig_zones) {
    choice <- "Safety-rated area scanner"
    reason <- paste("the aisle needs a reduced-speed warning zone for the AGV plus a",
                     "hard-stop zone for a person, which only a software-configurable",
                     "scanner provides economically.")
  } else {
    choice <- "Safety light curtain"
    reason <- paste("the point is accessed every cycle by an operator, and the",
                     "environment is clean enough that beam blockage from debris",
                     "is not a major risk.")
  }
  cat("  Recommended guarding:", choice, "\n")
  cat("  Reasoning:", reason, "\n\n")
}
```

**Sanity check:** each recommendation follows directly from a single dominant criterion — access frequency for the fence-vs-curtain split, need for graduated zones for the scanner — which is exactly the point of a decision framework: it should be traceable back to *why*, not just produce an answer.

## Design Task 6: Collaborative Operation Limits (ISO/TS 15066)

### Scenario

A plant manager, trying to avoid the cost of fencing and light curtains altogether, proposes replacing a guarded fastening station with a "collaborative" robot arm — one rated by its manufacturer as power- and force-limiting, marketed as safe to work alongside without a cage. The task is screwdriving: the robot carries a powered driver with a pointed bit that spins during fastening, working close to an operator's hands. Does specifying a collaborative-rated arm remove the need for a risk assessment on this station?

### Design Guidelines and Rules of Thumb

ISO/TS 15066 recognizes four distinct techniques for collaborative robot operation, and a given application typically relies on one or a combination of them — not just "the robot is collaborative" as a blanket property.

```{r tbl-06-01-collab-techniques, echo=FALSE, message=FALSE}
collab_modes <- tribble(
  ~Technique, ~How_It_Works, ~Typical_Use,
  "Safety-rated monitored stop", "Robot halts completely whenever a person is in the collaborative workspace; resumes once clear", "Tasks where the robot only needs to move when no one is present",
  "Hand guiding", "Operator physically leads the robot through motions via a hand-held pendant/handle at the tool", "Manual teaching, low-volume flexible assembly",
  "Speed and separation monitoring", "Robot speed is continuously reduced as a person gets closer, tracked by vision or area scanners", "Cells where the robot and person share space but rarely touch",
  "Power and force limiting", "The robot (or a compliant joint/skin) is inherently limited so contact forces stay under injury thresholds", "Direct human-robot contact tasks with a smooth, blunt end-effector"
)

collab_modes %>%
  kbl(caption = "The four collaborative operation techniques recognized in ISO 10218 / ISO/TS 15066") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, background = "#2C3E50", color = "white")
```

Here is the misconception that this design task exists to correct: a robot being marketed as "collaborative" describes the *bare arm's* certification, tested and rated in isolation, usually with no tooling attached. It says nothing about the end-effector the arm is carrying, the workpiece it's manipulating, or the specific motion the application demands. A power/force-limiting arm carrying a sharp gripper, a hot glue nozzle, or — as in this scenario — a spinning pointed driver bit introduces its own contact hazards that the robot's own certification does not cover. ISO/TS 15066 sets biomechanical limits (transient and quasi-static contact pressure and force thresholds by body region) that the *complete application* has to satisfy, not just the robot alone.

**Rule of thumb:** treat a vendor's "collaborative" or "cobot" label as necessary, never sufficient. It tells you the arm passed a bare-arm certification test; it does not tell you the application-specific risk assessment has been done, or that it would pass. This holds regardless of price point or reputation — the label describes a test condition, not a guarantee about your specific tooling and task.

CSA Z434 incorporates collaborative robot requirements in the Canadian context by reference to ISO 10218-1/-2 and ISO/TS 15066; there is no separate Canadian-specific collaborative-robot standard.

### Worked Example

Live cell — try setting `tool_has_sharp_or_pinch_hazard` to `FALSE` (e.g., swap the driver for a blunt-tip capacitive screwdriver) and re-run to see the conclusion change.

```{webr-r}
#| label: collab-check-06-01
## A "collaborative" robot arm is being considered for a screwdriving task,
## carrying a pointed driver bit that spins during fastening.

robot_is_power_force_limited <- TRUE  # vendor-rated collaborative arm
tool_has_sharp_or_pinch_hazard <- TRUE  # spinning pointed bit
biomechanical_limit_exceeded <- TRUE    # estimated bit-tip contact pressure exceeds
                                         # ISO/TS 15066 Annex A transient contact
                                         # limits for the fingertip region

cat("Robot arm is vendor-rated power/force-limiting collaborative:", robot_is_power_force_limited, "\n")
cat("End-effector introduces its own contact hazard (sharp/spinning bit):", tool_has_sharp_or_pinch_hazard, "\n")
cat("Estimated bit-tip contact exceeds ISO/TS 15066 biomechanical limit:", biomechanical_limit_exceeded, "\n\n")

if (robot_is_power_force_limited && !tool_has_sharp_or_pinch_hazard && !biomechanical_limit_exceeded) {
  cat("Conclusion: the application-level risk assessment supports unguarded\n")
  cat("collaborative operation as configured.\n")
} else {
  cat("Conclusion: the robot's own collaborative rating does NOT clear this\n")
  cat("application. The end-effector's sharp, spinning bit is a separate\n")
  cat("hazard that the robot's power/force limiting does not cover.\n\n")
  cat("Required mitigations (at least one):\n")
  cat(" - Slow or stop bit rotation whenever a person is within reach\n")
  cat("   (speed and separation monitoring around the fastening motion)\n")
  cat(" - Add a physical guard or light curtain around the fastening zone\n")
  cat("   so the bit is never exposed while a person is present\n")
  cat(" - Redesign the end-effector to remove the exposed point/pinch hazard\n\n")
  cat("A vendor 'collaborative' marketing claim is not, by itself, a\n")
  cat("substitute for this application-specific risk assessment.\n")
}
```

**Sanity check:** the conclusion doesn't hinge on the robot at all — every input that flips the outcome is a property of the *tool and task*, which is exactly the point: the risk assessment for a collaborative application lives mostly in the end-effector and the specific motion, not in the robot's datasheet.

## Design Task 7: Ontario Pre-Start Health and Safety Review (PSR) Applicability

### Scenario

The door-panel cell is fully designed, built, and ready to commission at the Cambridge plant. Before it's released to run production, someone on the project has to answer a question that has nothing to do with reach, payload, or cycle time: does Ontario law require a Pre-Start Health and Safety Review before this cell can operate?

### Design Guidelines and Rules of Thumb

Ontario Regulation 851 (Industrial Establishments), made under the Occupational Health and Safety Act, requires a Pre-Start Health and Safety Review for certain new or modified equipment involving guards, protective devices, or structural changes that affect worker safety — a brand-new robot cell, with new fencing, a new light curtain, and a new safety PLC program, is a textbook case. A PSR is generally **not** triggered by routine, like-for-like maintenance (swapping a failed sensor for an identical replacement) or by a change that doesn't touch guarding or safety controls at all (a robot path program tweak that doesn't change speed, force, or the safety circuit).

Where a PSR is required, a **Professional Engineer (P.Eng.)** licensed to practice in Ontario reviews the installation against the applicable standards and prepares (or reviews and seals) the report that certifies the design meets the requirements of the OHSA and its regulations. This is not a rubber stamp: the P.Eng. is putting their professional seal — and their personal liability under the Professional Engineers Act — behind the conclusion that the design is safe. Expected documentation typically includes general arrangement and safety circuit drawings, a hazard/risk assessment, the specific standards referenced in the design (CSA Z434, ISO 10218, ISO 13849-1 for the safety circuit's performance level, and others as applicable), and commissioning/test records demonstrating the safety devices actually work as designed.

**Rule of thumb:** if you're asking whether a given change needs a PSR, the safe working answer is *yes, assume it does* until a P.Eng. reviews the specific change and says otherwise. This rule of thumb never really "breaks down" in the sense of becoming wrong to follow — the honest failure mode is the opposite direction, skipping a PSR that was actually required, and the cost of an unnecessary review is far smaller than the liability (and, more importantly, the safety exposure) of skipping one that wasn't.

PSR is a distinctly Ontario mechanism; other Canadian provinces and US jurisdictions have their own pre-startup or acceptance-inspection requirements (some US states, for instance, require documented machine-guarding compliance reviews under their own OSHA state-plan rules), but there is no single ANSI or federal OSHA equivalent to a mandatory, professional-engineer-led PSR — a project team working across the border should not assume a US sister plant's sign-off process satisfies Ontario's requirement, or vice versa.

```{r fig-06-01-psr-flowchart, echo=FALSE, fig.cap="Simplified decision flow for whether a new robot cell triggers an Ontario Pre-Start Health and Safety Review under O. Reg. 851. Consult a P.Eng. for the definitive answer on a specific project.", fig.align='center', fig.width=10, fig.height=6.5}
nodes <- data.frame(
  id = 1:5,
  label = c("New or modified\nrobot cell installed?",
            "Guards, barriers, or\nprotective devices\nare part of the change?",
            "Is it identical\n(like-for-like)\nreplacement?",
            "PSR required:\nengage a P.Eng.,\nprepare documentation",
            "PSR not required\n(document the\ndecision anyway)"),
  x = c(5, 5, 5, 8, 2),
  y = c(9, 6.5, 4, 1.5, 1.5),
  kind = c("Question", "Question", "Question", "Required", "NotRequired")
)

edges <- data.frame(
  x = c(5, 5, 5, 5),
  y = c(8.4, 5.9, 3.4, 3.4),
  xend = c(5, 5, 8, 2),
  yend = c(7.1, 4.6, 1.9, 1.9)
)

edge_labels <- data.frame(
  x = c(5.6, 5.6, 6.5, 3.5),
  y = c(7.7, 5.2, 2.5, 2.5),
  label = c("Yes", "Yes", "No", "Yes")
)

ggplot() +
  geom_segment(data = edges, aes(x = x, y = y, xend = xend, yend = yend),
               arrow = arrow(length = unit(0.2, "cm")), color = "grey50", linewidth = 0.7) +
  geom_label(data = nodes, aes(x = x, y = y, label = label, fill = kind),
             size = 3 * diagram_text_scale, label.padding = unit(0.3, "lines")) +
  geom_text(data = edge_labels, aes(x = x, y = y, label = label),
            size = 2.7 * diagram_text_scale, fontface = "italic") +
  annotate("text", x = 1.9, y = 8.6, label = "No -> PSR not\nrequired for this\nchange alone",
           size = 2.4 * diagram_text_scale, color = "grey40", fontface = "italic") +
  annotate("text", x = 1.9, y = 6.2, label = "No -> PSR not\nrequired for this\nchange alone",
           size = 2.4 * diagram_text_scale, color = "grey40", fontface = "italic") +
  scale_fill_manual(values = c("Question" = "#FADBD8", "Required" = "#D5F5E3",
                                "NotRequired" = "#FCF3CF"), guide = "none") +
  xlim(0, 10) + ylim(0.5, 9.8) +
  theme_void() +
  labs(title = "Does This Robot Cell Need a Pre-Start Health and Safety Review?") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13 * diagram_text_scale))
```

```{r tbl-06-01-psr-triggers, echo=FALSE, message=FALSE}
psr_triggers <- tribble(
  ~Situation, ~Likely_PSR_Status,
  "New robot cell installed for the first time, with new guarding and safety controls", "PSR required",
  "Existing cell relocated to a new area of the plant", "PSR required (new installation context)",
  "Guarding, light curtain, or safety PLC program materially changed on an existing cell", "PSR required",
  "Robot software/path program updated, no change to guards or safety circuit", "PSR generally not required",
  "Like-for-like replacement of a failed component with an identical part", "PSR generally not required",
  "Routine preventive maintenance per the manufacturer's manual", "PSR not required"
)

psr_triggers %>%
  kbl(caption = "Situations that typically do -- and do not -- trigger a Pre-Start Health and Safety Review under O. Reg. 851") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, background = "#2C3E50", color = "white")
```

### Worked Example

Live cell — try setting `like_for_like_replacement` to `TRUE` (imagine this is a repair, not a new install) and re-run to see the conclusion change.

```{webr-r}
#| label: psr-check-06-01
## New robot cell scenario at an Ontario auto-parts plant
new_apparatus <- TRUE          # brand-new robot cell, never operated before
new_guards_or_devices <- TRUE  # perimeter fencing, light curtain, and safety
                                 # PLC program are all new
like_for_like_replacement <- FALSE

cat("New or modified apparatus installed:", new_apparatus, "\n")
cat("Guards or protective devices are part of the change:", new_guards_or_devices, "\n")
cat("Is this a like-for-like replacement:", like_for_like_replacement, "\n\n")

psr_required <- new_apparatus && new_guards_or_devices && !like_for_like_replacement

if (psr_required) {
  cat("Conclusion: a Pre-Start Health and Safety Review IS required before\n")
  cat("this cell is put into production.\n\n")
  cat("Next steps:\n")
  cat(" - Engage a Professional Engineer (P.Eng.) licensed in Ontario to\n")
  cat("   perform or certify the review\n")
  cat(" - Assemble documentation: general arrangement and safety circuit\n")
  cat("   drawings, hazard/risk assessment, applicable standards referenced\n")
  cat("   (CSA Z434, ISO 10218, ISO 13849-1 for the safety circuit),\n")
  cat("   commissioning/test records for the safety devices\n")
  cat(" - Obtain the P.Eng.'s signed and sealed PSR report before the\n")
  cat("   cell is released for production use\n")
} else {
  cat("Conclusion: based on these inputs, a PSR is not clearly triggered --\n")
  cat("but document the reasoning; when in doubt, consult a P.Eng. or the\n")
  cat("plant's health and safety representative.\n")
}
```

**Sanity check:** every input in this scenario points the same direction (new apparatus, new guarding, not a replacement), so the conclusion is unambiguous — a genuinely borderline case (say, a significant reprogram that changes robot speed near a guard opening but touches no hardware) is exactly where a rule-based checklist like this one runs out of authority and a P.Eng.'s judgment takes over.

## Summary

```{r tbl-06-01-summary, echo=FALSE, message=FALSE}
summary_table <- tribble(
  ~Design_Task, ~Governing_Relationship, ~Key_Rule_of_Thumb,
  "1. Reach verification", "Required reach = D + L_tool, compared to R_max with margin",
  "Keep >= 10% margin below rated max reach for orientation-sensitive points",
  "2. Payload / CG offset", "Payload_max(L) = M_cap / L beyond the flat region L0",
  "Nameplate payload alone never proves a design passes -- always check the CG-offset derating",
  "3. Cycle time estimation", "t_move (trapezoidal or triangular) summed with all dwell times",
  "Pad a first-pass kinematic estimate 10-15% for real-world variability",
  "4. Minimum safe distance (ISO 13855)", "S = K x T + C",
  "Fix a marginal S by shortening T or shrinking C -- never by disputing K",
  "5. Guarding technology selection", "Decision framework: access frequency, footprint, zone flexibility",
  "Default to fixed fencing unless frequent access or an irregular perimeter says otherwise",
  "6. Collaborative operation limits", "Checklist: robot rating + tooling hazard + biomechanical limits (ISO/TS 15066)",
  "\"Collaborative\" on the nameplate is necessary, never sufficient, on its own",
  "7. PSR applicability (O. Reg. 851)", "Checklist: new/modified apparatus + guards/devices + not like-for-like",
  "If guarding or safety controls changed at all, assume a PSR is needed until a P.Eng. says otherwise"
)

summary_table %>%
  kbl(caption = "Design task summary: governing relationship and key rule of thumb",
      col.names = c("Design Task", "Governing Relationship", "Key Rule of Thumb")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, background = "#2C3E50", color = "white")
```

## Practice Problems

<details>
<summary>Task 1 (plug-in): A new weld point P6 is added at cell coordinates (2900, -700, 500) mm. Using the same robot base (1200, 0, 300), tool length (380 mm), and R_max (2655 mm) as the worked example, does this point pass the reach check with a 10% margin?</summary>

Vector from base to P6: dx = 1700, dy = -700, dz = 200.

$D = \sqrt{1700^2 + 700^2 + 200^2} = \sqrt{2\,890\,000 + 490\,000 + 40\,000} = \sqrt{3\,420\,000} \approx 1849.3$ mm

Required reach = 1849.3 + 380 = 2229.3 mm

Margin = (2655 - 2229.3) / 2655 × 100 ≈ 16.0%

This clears the 10% rule-of-thumb threshold comfortably — **PASS**.

</details>

<details>
<summary>Task 1 (design decision): The integrator offers two robot options for this cell: Robot A has R_max = 2655 mm and costs $48,000; Robot B has R_max = 2200 mm and costs $39,000. Both pass the straight reach check for the farthest point (required reach 2490.7 mm from the original worked example), but which would you choose, and why?</summary>

Robot B fails outright — its rated reach (2200 mm) is less than the required reach (2490.7 mm) even before applying the margin rule of thumb, so it cannot physically complete the task from this pedestal location without relocating it or accepting an unreachable point. Robot A passes the raw reach check but, as the worked example showed, only with a 6.2% margin — below the 10% threshold this chapter recommends. The right choice is neither robot as originally proposed: either select Robot A and relocate the pedestal (as the worked example did, restoring margin to over 16%), or specify a robot with reach comfortably above 2490.7 mm from the current pedestal location. Choosing on cost alone (Robot B) without checking reach first would produce a cell that cannot do its job.

</details>

<details>
<summary>Task 2 (plug-in): A gripper with mass 22 kg and a case of mass 12 kg operates at a combined CG offset of 140 mm, on a robot with W_max0 = 130 kg and L0 = 60 mm (M_cap = 7800 kg·mm). Does this pass the payload check?</summary>

Total mass = 22 + 12 = 34 kg

Since L = 140 mm > L0 = 60 mm: Payload_max = 7800 / 140 = 55.7 kg

34 kg < 55.7 kg — **PASS**, with margin (55.7 - 34) / 55.7 ≈ 39%.

</details>

<details>
<summary>Task 2 (design decision): The palletizing cell's gripper fails the payload check by a wide margin (as in the worked example: 44 kg carried vs. 37.1 kg allowed at 210 mm offset). Three fixes are on the table: (1) add a 15 kg counterweight positioned to pull the combined CG back toward the flange, (2) redesign the gripper frame to reduce the offset from 210 mm to 160 mm, or (3) upgrade to a robot with a higher payload rating at the same offset. Which would you recommend, and what would you check before committing?</summary>

A counterweight (option 1) sounds appealing but adds mass to a system that's already failing a moment-based check — unless it's positioned close enough to the flange (on the *opposite* side of the offset) to meaningfully reduce the *combined* CG offset rather than just adding weight at the existing offset, it can make the moment worse, not better. It would need its own offset/mass calculation before being trusted. A gripper redesign (option 2) directly attacks the variable that's actually failing the check (L), and the new allowable payload at 160 mm would be 7800/160 = 48.75 kg — comfortably above the 44 kg load — without touching the robot at all, which is usually the lowest-cost fix if the frame redesign is mechanically feasible. Upgrading the robot (option 3) solves the problem but is the most expensive option and may not be necessary if a modest frame redesign closes the gap. The recommended path is to price out the gripper redesign first, since it fixes the root cause (offset) directly; only escalate to a robot upgrade if the offset can't be reduced enough for mechanical reasons (e.g., the case-turning mechanism physically needs that reach).

</details>

<details>
<summary>Task 3 (plug-in): Using v_max = 1200 mm/s and a = 4000 mm/s^2, what is the move time for a single 900 mm move?</summary>

$d_{accel} = v_{max}^2 / (2a) = 1200^2 / 8000 = 180$ mm, so $2 d_{accel} = 360$ mm.

Since 900 mm ≥ 360 mm, this is trapezoidal:

$t_{accel} = v_{max}/a = 1200/4000 = 0.3$ s

$t_{const} = (900 - 360)/1200 = 0.45$ s

$t_{move} = 2(0.3) + 0.45 = 1.05$ s

</details>

<details>
<summary>Task 3 (design decision): A revised weld sequence pushes total cycle time to 11.2 seconds against a 9.5 second allotment. The controls engineer proposes three fixes: raise v_max by 25%, raise acceleration by 25%, or drop one weld point by combining two adjacent welds into a single larger weld. Which would you try first, and why?</summary>

For short moves that never reach v_max (triangular profile), raising acceleration shortens the move time directly ($t_{move} = 2\sqrt{d/a}$, which shrinks with the square root of $a$); raising v_max does nothing at all for those moves since they never reach the cap. For long moves that do reach v_max (trapezoidal), raising v_max shortens the constant-velocity portion, while raising acceleration only shortens the (usually smaller) accel/decel portions. Since this cell's moves are a mix (some triangular, some trapezoidal, per the original worked example), raising acceleration tends to help more of the sequence at once — and it's often cheaper in practice, since many servo drives have acceleration headroom that isn't being used, whereas raising v_max may run into vibration or overshoot limits on the fixture. Dropping a weld point (combining two welds) is the most disruptive option: it changes the part's structural design, not just the robot program, and would need sign-off from product engineering, not just the controls team. The recommended order is to try raising acceleration first (cheapest, no part impact), check v_max next if that's insufficient, and treat combining welds as a last resort requiring design approval.

</details>

<details>
<summary>Task 4 (plug-in): A cell has T_robot = 0.40 s, T_device = 0.05 s, and uses an area scanner with C = 1200 mm. What is the required minimum separation distance S?</summary>

T = 0.40 + 0.05 = 0.45 s

S = K × T + C = 1600 × 0.45 + 1200 = 720 + 1200 = 1920 mm

</details>

<details>
<summary>Task 4 (design decision): A calculated S comes out to 3.4 m for a cell where the plant only has 2.1 m of floor space available at that boundary. The safety engineer is considering either a faster-stopping safety circuit (reducing T_robot from 0.65 s to 0.30 s) or switching from an area scanner (C = 1200 mm) to a finer-resolution light curtain (C = 700 mm). Which change has the bigger effect here, and is either one enough on its own?</summary>

Reducing T_robot from 0.65 s to 0.30 s (assuming T_device stays around 0.04-0.05 s, so T drops from about 0.69 s to about 0.35 s) reduces the K×T term from roughly 1104 mm to about 552 mm — a reduction of about 552 mm. Switching from C = 1200 mm to C = 700 mm reduces the C term by 500 mm directly. The two changes are similar in magnitude here, but neither one alone closes a gap of roughly 1.3 m (3.4 m required vs. 2.1 m available assuming the original inputs implied something close to that). This is a case where both changes are likely needed together: a faster stop category *and* a finer-resolution device, and even then the design should be rechecked with the actual updated numbers rather than assumed to pass — floor space this tight is also a signal to reconsider the cell's overall layout, not just the guarding technology.

</details>

<details>
<summary>Task 5 (plug-in): A fourth access point is added to the cell: a maintenance panel accessed roughly twice a year by a technician, in a location with no AGV or other automated traffic. Which guarding type does the decision framework recommend, and why?</summary>

Access frequency is effectively "never" from a routine-operations standpoint (twice a year is maintenance access, not production access), and there's no need for graduated warn/stop zones since there's no automated traffic to warn. The framework recommends **fixed fencing** with a lockable, interlocked access door for the rare maintenance entries — the lowest-cost option, appropriate because the access pattern doesn't demand the fast, hands-free entry a light curtain is built for, nor the zone flexibility an area scanner provides.

</details>

<details>
<summary>Task 5 (design decision): The plant's capital budget can afford either an area scanner for the AGV aisle ($18,000 installed) or a light curtain plus a physical AGV-height barrier rail ($9,000 installed), but not both at full spec. The light curtain option would require the AGV to come to a complete stop and wait for a "clear" signal every time it enters the aisle, adding roughly 8 seconds per AGV pass; the scanner option allows the AGV to slow but not fully stop unless a person is detected. With the AGV making about 40 passes per shift, which option would you recommend, and what else would you want to know before deciding?</summary>

The light curtain option costs $9,000 upfront but adds a recurring throughput penalty: 40 passes × 8 seconds ≈ 320 seconds (about 5.3 minutes) of added AGV wait time per shift, every shift, indefinitely. Whether that's acceptable depends on how tightly the AGV's delivery schedule is loaded — if the AGV already has slack time in its cycle, 5.3 minutes a shift may be a non-issue; if the AGV is a bottleneck feeding multiple cells, that recurring cost could easily exceed the $9,000 price difference within months in lost throughput. Before deciding, it's worth getting the AGV's current utilization/slack time from the automation team and running the throughput cost forward over the equipment's expected life, not just comparing sticker prices — a cheaper guarding solution that quietly taxes a bottleneck process every single cycle is often the more expensive choice once you look past the installation invoice.

</details>

<details>
<summary>Task 6 (plug-in): A collaborative-rated robot arm is fitted with a smooth, rounded, blunt plastic pusher (no sharp edges, no pinch points, low approach speed) to nudge parts into a fixture. Does this application clear the collaborative-operation checklist as described in this chapter?</summary>

Checking the three conditions from the worked example: the robot is vendor-rated power/force-limiting collaborative (yes); does the end-effector introduce its own separate contact hazard (no — smooth, blunt, no pinch points); does the estimated contact force/pressure exceed ISO/TS 15066's biomechanical limits for the relevant body region at this speed (assume no, given the low approach speed and blunt geometry, but this still needs to be actually calculated/verified, not assumed). With the first condition true and the other two false, this application is a reasonable candidate for unguarded collaborative operation — but "reasonable candidate" still means the application-specific risk assessment needs to be documented and the contact-force limit genuinely checked against ISO/TS 15066's tables, not waved through because the tool "looks safe."

</details>

<details>
<summary>Task 6 (design decision): Testing shows the pusher from the previous problem exceeds ISO/TS 15066's transient contact force limit, but only when the robot is running at its maximum programmed speed — at 60% of that speed, it passes comfortably. Production wants full speed for cycle time reasons. What are the options, and which would you recommend?</summary>

Options include: (1) permanently cap the robot's speed at or below the passing threshold, sacrificing some cycle time; (2) implement speed and separation monitoring so the robot only runs at full speed when no person is within reach, dropping to the safe speed (or stopping) when a person approaches — preserving cycle time during unattended running while still meeting the biomechanical limit whenever contact is possible; (3) redesign the pusher (softer material, larger contact area to spread the force, added compliance/spring mounting) to reduce peak contact pressure enough to pass at full speed. Option 2 is usually the strongest choice where the cell has genuine unattended run time, since it recovers most of the cycle-time benefit of full speed without accepting the contact-force risk — but it adds the cost and complexity of a presence-sensing system (vision or area scanner) and needs its own validation that the speed transition is fast and reliable enough to actually be safe. Option 3 is worth evaluating in parallel since a successful redesign removes the constraint entirely at lower ongoing complexity than a monitored system. Option 1 is the simplest and cheapest but accepts the full cycle-time cost permanently — a reasonable interim measure while 2 or 3 is developed, not necessarily the final answer.

</details>

<details>
<summary>Task 7 (plug-in): A plant technician swaps a failed light curtain for an identical replacement unit from the same manufacturer, same model number, mounted in the same location, using the same wiring. Does this trigger a PSR?</summary>

Working through the checklist: is this a new or modified apparatus? Not really — it's a repair, not an installation or modification. Is it a like-for-like replacement? Yes — same model, same location, same wiring. Per the decision framework, this is a case where a PSR is **generally not required**, since it falls squarely in the like-for-like maintenance category. It's still good practice to document the replacement and re-verify the device's function as part of routine maintenance records, even without a formal PSR.

</details>

<details>
<summary>Task 7 (design decision): A plant manager, facing a tight production launch deadline, suggests skipping the PSR on a new robot cell because "it's basically the same as the cell we installed last year, just with a different gripper." As the engineer responsible for the installation, how would you respond?</summary>

"Basically the same" is doing a lot of work in that sentence, and it's exactly the kind of informal judgment call the PSR process exists to remove from an individual's shoulders. A different gripper can change payload, reach, and CG offset (Design Tasks 1 and 2 of this chapter), which can change the robot's motion envelope and therefore the guarding layout and the ISO 13855 separation distance — none of which is "the same" even if the robot model and most of the cell structure are unchanged. The correct response is to treat this as a new installation (which it is — a new cell, even if similar to a previous one) and assume a PSR is required until a P.Eng. reviews the specific change and says otherwise, per the rule of thumb in this chapter. Skipping it to save time on a launch schedule trades a schedule risk for a safety and legal liability risk — if something goes wrong on an un-reviewed cell, "it was basically the same as last time" is not a defence under the OHSA, and the individuals who signed off on skipping the review (not just the company) can be held personally accountable. The right move is to flag the schedule conflict early, get a P.Eng. engaged as soon as the cell design is stable enough to review (it doesn't have to wait until the very end), and treat the PSR as a fixed part of the launch schedule, not an optional step to cut when time gets tight.

</details>

## Open-Ended Design Scenario

<details>
<summary>Capstone scenario: finish the door-panel cell</summary>

You are the layout engineer responsible for releasing the door subassembly welding cell from this chapter to production. Pull together everything from this chapter into one design review:

- The robot is mounted on a pedestal at (1200, 0, 300) mm in cell coordinates. Five weld points are specified on the door inner panel, with the farthest at (3100, 850, 650) mm. The spot-welding gun is 380 mm long. The candidate robot has a rated reach of 2655 mm.
- The gun assembly (including the transformer pack) has a mass of 92 kg and a combined CG offset of 240 mm from the wrist flange. The candidate robot's rated payload at zero offset is 165 kg, with a flat region out to 70 mm offset.
- The weld sequence requires six moves (home to P1 through P5, then back to home) with distances of 480, 550, 620, 400, 710, and 850 mm, a commanded speed of 1400 mm/s, and acceleration of 4500 mm/s². Each weld takes 0.4 s, and the cell PLC handshake adds 0.25 s. The station's overall target is 58 jobs per hour, and this robot's operations are allotted 40% of that cycle time.
- The safety circuit design (from the Machine Safety Circuits chapter) specifies a robot stopping time of 0.55 s. The planned light curtain has a response time of 0.03 s and a resolution that corresponds to a 750 mm intrusion allowance (C).
- The cell perimeter has three access types: the operator load station (every cycle), a rarely-used maintenance panel, and a side aisle shared with a tugger cart delivering panel racks a few times per shift.
- A supplier has pitched a "collaborative" version of this robot as an option to eliminate the guarding budget entirely.
- This is the plant's first robot cell of this specific design, though the plant has other robot cells of different designs already running.

For each of the seven design tasks in this chapter, work through the calculation or decision framework and reach a clear conclusion: does the cell pass, and if not, what specific change would you make? Then write two or three sentences tying your conclusions together — does fixing one issue (say, relocating the pedestal for the reach check) create or resolve a conflict with any of your other conclusions (say, the payload check, which is also sensitive to geometry)? A strong answer treats this as one connected design, not seven independent homework problems.

</details>
