---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Pneumatic System Design

```{r setup-06-07, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
library(tidyr)
```

## Learning Objectives

By the end of this chapter, you will be able to:

1. **Calculate** the required cylinder bore from a specified force, correctly accounting for the reduced piston area on the rod side of the cylinder, and **apply** the 25-50% oversize rule of thumb to select a standard catalog bore.
2. **Distinguish** theoretical force (P x A) from actual delivered force, and **calculate** actual force using a realistic combined efficiency factor for seal friction and pressure drop.
3. **Calculate** a cylinder's air consumption in SCFM and NL/min for a specified cycle rate, converting compressed (line-pressure) volume to free-air (atmospheric) volume.
4. **Select** a directional control valve by calculating its required flow coefficient (Cv) from a target cylinder fill time, using the standard compressible-flow Cv equation for air.
5. **Specify** an FRL (filter-regulator-lubricator) unit for a given flow demand, and **decide**, using seal type and process environment, whether line lubrication is appropriate.
6. **Explain** the industry rule-of-thumb cost of generating compressed air, and **evaluate** the annual cost impact of an oversized cylinder or an air leak versus a right-sized system.
7. **Evaluate** the trade-offs between pneumatic and electric linear actuators and **select** the appropriate technology for a given positioning and duty-cycle requirement.

## Introduction

Walk the floor of almost any automotive stamping or sub-assembly plant and the sound you notice first isn't a motor — it's air. A rhythmic *chuff-chuff* of cylinders extending and retracting, punctuated by the occasional hiss of a fitting nobody has gotten around to fixing. Pneumatics are everywhere on that floor: pick-and-place units transferring stamped parts off a press line, clamp cylinders holding a weldment square while a robot fires its guns, blow-off nozzles clearing chips, diverter gates sorting parts on a conveyor. Compressed air is popular for a reason — a pneumatic cylinder is simple, fast, tolerant of a dirty environment, and cheap compared to an equivalent electric axis. It's also easy to get wrong in ways that don't show up until the line is running: a cylinder that looked fine on paper but can't actually deliver enough force once real losses are accounted for, a valve that's too small to fill the cylinder fast enough to hit cycle time, or a system so casually oversized that nobody notices it's quietly costing the plant real money every single year.

This chapter builds a complete, checkable method for specifying a pneumatic actuation system, using one worked example as the throughline: a vertical pick-and-place lift cylinder on an automotive stamping line, transferring a stamped bracket from a press discharge conveyor onto a rack. Design Tasks 1 through 4 size that cylinder and its valve, piece by piece — bore, then real-world force, then air consumption, then valve selection — the same way an actual project would move from mechanical concept to a bill of materials. Design Task 5 shifts to a food-processing washdown environment to work through how to specify the filter-regulator-lubricator unit feeding the system, including when *not* to lubricate the air. Design Task 6 makes the economic case for right-sizing pneumatic equipment, using both this chapter's own cylinder and a compressed-air leak to put a real dollar figure on what "it's just air" actually costs a plant over a year. Design Task 7 closes with the question every project should ask before defaulting to pneumatics at all: when does the job actually call for an electric actuator instead?

A note on scope: the guarding and interlocking of a pneumatically actuated clamp or press cylinder — pinch-point risk assessment, safety-rated valve circuits, presence sensing — is developed in the [Machine Safety Circuits](06-02-Machine-Safety-Circuits.qmd) chapter. Here, the focus is strictly the fluid power sizing problem: getting the right amount of force, speed, and flow to the right place.

## Design Task 1: Cylinder Bore Sizing from Required Force

### Scenario

A Tier-1 automotive stamping supplier producing structural door-hinge reinforcement brackets needs a pneumatic pick-and-place unit at the press discharge conveyor. A vertical lift cylinder, mounted rod-down above the conveyor, does two things each cycle: on the **extend** stroke, it lowers a vacuum-cup gripper down onto the bracket sitting on the conveyor and lightly seats the gripper's locating pins into the part before the vacuum engages — the mechanical designer has specified this needs only about 180 N, mostly to overcome cylinder and guide-rail friction. On the **retract** stroke, the cylinder lifts the gripper, its mounting plate, and the bracket itself — a combined moving mass of roughly 14.5 kg — clear of the conveyor fast enough to keep the line's cycle time, requiring an estimated acceleration around 25 m/s². Between the inertial force, the assembly's weight, and friction, the process engineer has specified the retract stroke must deliver at least **550 N** at the rod.

The plant runs its pneumatic distribution at a standard shop pressure of 6 bar gauge (600 kPa, about 87 psi) — a typical line pressure for North American manufacturing plants, chosen to balance actuator force density against compressor energy cost (Design Task 6 returns to why that trade-off matters). The mechanical designer's job now: pick a bore size.

```{r fig-06-07-pnp-schematic, echo=FALSE, fig.cap="Vertical pick-and-place lift cylinder over a press discharge conveyor. The extend stroke lowers the gripper to pick the part; the retract stroke lifts the gripper and part clear.", fig.align='center', fig.width=10, fig.height=7}
library(ggplot2)

ggplot() +
  # conveyor
  annotate("rect", xmin = -3, xmax = 7, ymin = 0, ymax = 0.6, fill = "grey55", color = "black") +
  annotate("text", x = 5.3, y = 0.3, label = "Discharge conveyor", size = 3 * diagram_text_scale, color = "white") +
  # part on conveyor
  annotate("rect", xmin = 1.6, xmax = 3.0, ymin = 0.6, ymax = 1.0, fill = "#7F8C8D", color = "black") +
  annotate("text", x = 2.3, y = 0.8, label = "Bracket", size = 2.6 * diagram_text_scale, color = "white") +
  # cylinder body
  annotate("rect", xmin = 1.3, xmax = 3.3, ymin = 4.2, ymax = 6.6, fill = "#3498DB", color = "black") +
  annotate("text", x = 2.3, y = 5.4, label = "Lift cylinder\n(50 mm bore)", size = 3 * diagram_text_scale, color = "white", fontface = "bold") +
  # rod
  annotate("segment", x = 2.3, xend = 2.3, y = 4.2, yend = 1.6, linewidth = 3, color = "grey30") +
  # gripper
  annotate("rect", xmin = 1.7, xmax = 2.9, ymin = 1.0, ymax = 1.6, fill = "#E67E22", color = "black") +
  annotate("text", x = 2.3, y = 1.3, label = "Gripper", size = 2.6 * diagram_text_scale, color = "white") +
  # extend arrow (down)
  annotate("segment", x = 3.7, xend = 3.7, y = 4.0, yend = 1.8,
           arrow = arrow(length = unit(0.25, "cm"), type = "closed"), color = "#2ECC71", linewidth = 1.1) +
  annotate("text", x = 4.5, y = 2.9, label = "Extend\n(lower)\n180 N req.", size = 2.7 * diagram_text_scale, color = "#2ECC71") +
  # retract arrow (up)
  annotate("segment", x = 0.9, xend = 0.9, y = 1.8, yend = 4.0,
           arrow = arrow(length = unit(0.25, "cm"), type = "closed"), color = "#E74C3C", linewidth = 1.1) +
  annotate("text", x = 0.1, y = 2.9, label = "Retract\n(lift)\n550 N req.", size = 2.7 * diagram_text_scale, color = "#E74C3C") +
  # FRL + valve block
  annotate("rect", xmin = -3, xmax = -0.8, ymin = 4.4, ymax = 6.4, fill = "#95A5A6", color = "black") +
  annotate("text", x = -1.9, y = 5.4, label = "FRL +\ndirectional\nvalve", size = 2.6 * diagram_text_scale, color = "white") +
  annotate("segment", x = -0.8, xend = 1.3, y = 5.5, yend = 5.5,
           arrow = arrow(length = unit(0.2, "cm"), type = "closed"), linewidth = 1) +
  coord_fixed() + xlim(-3.5, 7.5) + ylim(-0.5, 7.2) +
  theme_void() +
  labs(title = "Pick-and-Place Lift Cylinder — Vertical Mount, Rod Down") +
  theme(plot.title = element_text(size = 14 * diagram_text_scale, face = "bold", hjust = 0.5))
```

### Design Guidelines and Rules of Thumb

The governing relationship for a pneumatic cylinder is simple: force equals pressure times area. But a double-acting cylinder has *two different areas*, because the piston rod itself occupies part of the bore on one side. On the **extend** stroke, air pushes on the full piston face, so the full bore area is available. On the **retract** stroke, air is admitted to the rod-side chamber, and the rod itself subtracts from that area — the piston only has the *annulus* (bore area minus rod cross-sectional area) to push against. For a standard cylinder with a rod roughly 40% of the bore diameter, that annulus area is typically 75-90% of the full bore area, meaning retract force is meaningfully lower than extend force at the same pressure.

**This matters enormously for which stroke governs the sizing.** If the load-bearing motion happens on the retract stroke — as it does here, where retracting lifts the gripper and part — the bore must be sized against the *reduced* annulus area, not the full bore area a naive F = P x A calculation would use. Sizing against the wrong stroke is one of the more common and costly mistakes in pneumatic cylinder specification: the cylinder looks adequate on paper (checked against the bigger, full-bore-area stroke) and then can't actually deliver the required force on the stroke that matters.

Once the correct governing area is identified, the next step accounts for the gap between theoretical and real-world performance.

**Rule of thumb:** oversize the cylinder's bore so that its theoretical force capability (on the governing stroke) is 25-50% greater than the calculated requirement. This margin absorbs seal friction, pressure drop between the FRL/valve and the cylinder, and sag in line pressure under flow — all of which reduce actual delivered force below the P x A theoretical value (quantified explicitly in Design Task 2). A tighter margin (toward 25%) is appropriate when the system has short, well-sized tubing runs, a generously sized valve, and a well-maintained compressor system with minimal pressure sag under load — in other words, when you have good reason to believe losses will be modest. A looser margin (toward 50% or occasionally more) is appropriate for long tubing runs, older or poorly maintained compressor systems with significant pressure droop during peak demand, small-bore cylinders (where seal friction is a larger fraction of total force, as Design Task 2 shows), or applications where an intermittent force shortfall would cause a safety or quality problem, not just a nuisance stoppage. This guideline also runs into the practical reality of standard bore sizes: ISO 15552 profile cylinders come in fixed bore increments (32, 40, 50, 63, 80 mm, and so on), and the jump between adjacent standard sizes doesn't always land neatly inside the 25-50% window — the worked example below shows exactly this happening, and how to make the call when it does.

Ontario and Canadian practice for pneumatic fluid power systems generally follows **ISO 4414** (General rules relating to pneumatic systems), with no separate CSA-specific pneumatic actuator sizing standard; the compressed air distribution piping and receiver tank itself falls under **CSA B51** (Boiler, Pressure Vessel, and Pressure Piping Code) as administered in Ontario by the **Technical Standards and Safety Authority (TSSA)**. Cylinder bore/rod/port dimensional standardization — which is what makes "select a standard 50 mm bore" a meaningful, multi-vendor-interchangeable choice — comes from **ISO 15552** (metric series, profile cylinders, previously known by the VDMA 24562 designation it superseded).

### Governing Calculation

$$F = P \times A$$

For the retract (rod-side) stroke, the effective area is the annulus between the bore and the rod:

$$A_{annulus} = A_{bore} - A_{rod} = \frac{\pi}{4}\left(D_{bore}^2 - D_{rod}^2\right)$$

To size the bore from a required force with the oversize margin applied:

$$A_{min} = \frac{F_{required}}{P} \qquad A_{selected} = (1.25\ \text{to}\ 1.50) \times A_{min}$$

| Symbol | Description | Unit |
| :--- | :--- | :--- |
| $F$ | Force delivered at the piston rod | N |
| $P$ | Supply (line) pressure, gauge | Pa (or bar, with unit conversion) |
| $A_{bore}$ | Full piston (bore) area — governs the extend stroke | m² (or mm²) |
| $A_{rod}$ | Piston rod cross-sectional area | m² (or mm²) |
| $A_{annulus}$ | Rod-side (annulus) area — governs the retract stroke | m² (or mm²) |
| $D_{bore}$ | Cylinder bore diameter | mm |
| $D_{rod}$ | Piston rod diameter | mm |
| $A_{min}$ | Minimum theoretical area needed to meet $F_{required}$ | mm² |

```{r fig-06-07-bore-selection, echo=FALSE, fig.cap="Annulus-area oversize margin for the standard ISO 15552 bore/rod combinations bracketing the requirement. 40 mm undershoots the 25% guideline floor; 50 mm clears it but overshoots the 50% ceiling; 63 mm is far beyond what's needed.", fig.align='center', fig.width=10, fig.height=6}
library(ggplot2)

F_lift_required <- 550
P_pa <- 6e5
A_min_mm2 <- F_lift_required / P_pa * 1e6

candidates <- data.frame(
  bore = c(32, 40, 50, 63),
  rod  = c(12, 16, 20, 20)
)
candidates$A_bore <- pi / 4 * candidates$bore^2
candidates$A_rod  <- pi / 4 * candidates$rod^2
candidates$A_annulus <- candidates$A_bore - candidates$A_rod
candidates$margin_pct <- (candidates$A_annulus / A_min_mm2 - 1) * 100
candidates$label <- paste0(candidates$bore, "/", candidates$rod, " mm")
candidates$selected <- ifelse(candidates$bore == 50, "Selected", "Not selected")

ggplot(candidates, aes(x = reorder(label, bore), y = margin_pct, fill = selected)) +
  geom_col(color = "black") +
  annotate("segment", x = 0.5, xend = 4.5, y = 0, yend = 0, linewidth = 0.8, color = "black") +
  annotate("segment", x = 0.5, xend = 4.5, y = 25, yend = 25, linetype = "dashed", color = "#F39C12", linewidth = 0.9) +
  annotate("text", x = 1.1, y = 32, label = "25% floor", size = 3 * diagram_text_scale, color = "#F39C12", hjust = 0) +
  annotate("segment", x = 0.5, xend = 4.5, y = 50, yend = 50, linetype = "dashed", color = "#9B59B6", linewidth = 0.9) +
  annotate("text", x = 1.1, y = 57, label = "50% ceiling", size = 3 * diagram_text_scale, color = "#9B59B6", hjust = 0) +
  geom_text(aes(label = sprintf("%.0f%%", margin_pct)), vjust = ifelse(candidates$margin_pct >= 0, -0.5, 1.3),
            size = 3.3 * diagram_text_scale) +
  scale_fill_manual(values = c("Selected" = "#2ECC71", "Not selected" = "#BDC3C7")) +
  labs(title = "Annulus-Area Oversize Margin vs. Standard Bore/Rod Combination",
       subtitle = "Retract-stroke requirement: 550 N at 6 bar (A_min = 917 mm²)",
       x = "Bore / rod (mm)", y = "Oversize margin over A_min (%)", fill = NULL) +
  theme_minimal() +
  theme(plot.title = element_text(size = 13.5 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 10 * diagram_text_scale),
        legend.position = "bottom",
        axis.text = element_text(size = 10 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale))
```

### Worked Example

```{r bore-sizing-06-07, echo=TRUE}
## Given
F_lift_required <- 550     # N, required retract (lift) force
F_lower_required <- 180    # N, required extend (lower) force
P_bar <- 6                  # bar gauge, shop line pressure
P_pa <- P_bar * 1e5

## Minimum annulus area needed on the governing (retract) stroke
A_annulus_min_mm2 <- F_lift_required / P_pa * 1e6
cat(sprintf("Minimum retract-side (annulus) area required: %.1f mm^2\n", A_annulus_min_mm2))
cat(sprintf("25%% oversize target: >= %.1f mm^2\n", A_annulus_min_mm2 * 1.25))
cat(sprintf("50%% oversize target: >= %.1f mm^2\n\n", A_annulus_min_mm2 * 1.50))

## Check standard ISO 15552 bore/rod combinations
bore_candidates <- c(32, 40, 50, 63)
rod_candidates  <- c(12, 16, 20, 20)

for (i in seq_along(bore_candidates)) {
  A_bore <- pi / 4 * bore_candidates[i]^2
  A_rod  <- pi / 4 * rod_candidates[i]^2
  A_annulus <- A_bore - A_rod
  margin_pct <- (A_annulus / A_annulus_min_mm2 - 1) * 100
  cat(sprintf("Bore %2d mm / rod %2d mm: A_annulus = %6.1f mm^2, margin = %6.1f%%\n",
              bore_candidates[i], rod_candidates[i], A_annulus, margin_pct))
}

## Selected: 50 mm bore / 20 mm rod
bore <- 50; rod <- 20
A_bore <- pi / 4 * bore^2
A_rod  <- pi / 4 * rod^2
A_annulus <- A_bore - A_rod

F_extend_theor  <- P_pa * A_bore * 1e-6
F_retract_theor <- P_pa * A_annulus * 1e-6

cat(sprintf("\nSelected: %d mm bore / %d mm rod\n", bore, rod))
cat(sprintf("A_bore = %.1f mm^2, A_annulus = %.1f mm^2 (%.1f%% of A_bore)\n",
            A_bore, A_annulus, A_annulus / A_bore * 100))
cat(sprintf("Theoretical extend (lower) force:  %.1f N  (required %.0f N)\n", F_extend_theor, F_lower_required))
cat(sprintf("Theoretical retract (lift) force:  %.1f N  (required %.0f N)\n", F_retract_theor, F_lift_required))
```

**Sanity check:** the 40 mm/16 mm combination only clears the requirement by 15.2%, below the 25% guideline floor — before any real-world losses are even considered, which makes it a poor choice. The next standard size up, 50 mm/20 mm, clears it by 79.9%, well past the 50% ceiling. Neither standard size lands inside the textbook-clean 25-50% window. When that happens, round up, not down: an inadequate margin is a functional risk (the cylinder might not deliver enough force once friction and pressure drop are subtracted, which Design Task 2 quantifies directly), while extra margin above the guideline is a cost and packaging-size issue, not a functional failure. The 50 mm/20 mm cylinder is the correct selection here, and it carries forward through the rest of this chapter's worked examples.

## Design Task 2: Theoretical vs. Actual Cylinder Force

### Scenario

The 50 mm bore cylinder looks comfortably sized on paper — 79.9% oversize on the governing retract stroke. But P x A is a theoretical number that assumes the cylinder is a frictionless piston fed at exactly the regulator's set pressure. Neither is true in an actual installation. The piston and rod seals drag against the bore and rod surfaces, consuming some of the available force before it ever reaches the load. And the air reaching the cylinder isn't at the regulator's static setpoint — it has to travel through the directional valve, fittings, and tubing, all of which impose a pressure drop that grows with flow rate, meaning the pressure at the cylinder is lower while it's actually moving than the gauge reads at rest. The mechanical designer needs to know: after both effects, does this cylinder still clear the 550 N retract requirement, or did the "generous" 79.9% area margin get eaten alive by real losses?

### Design Guidelines and Rules of Thumb

Seal friction is largely a function of seal design, bore finish, lubrication state, and — importantly — **bore size**: friction force doesn't scale down proportionally with a smaller bore the way the theoretical force does, so friction eats a *larger percentage* of available force in small-bore cylinders than in large ones. The same logic applies to stroke length: a longer stroke means more seal-to-bore contact length and more opportunity for misalignment drag, so friction losses tend to be worse in long-stroke cylinders too, especially without adequate rod/piston guide support.

Pressure drop between the FRL/valve assembly and the cylinder is a function of flow rate, tubing inner diameter, tubing length, and the number and type of fittings in the run — every elbow, quick-disconnect, and undersized push-to-connect fitting adds resistance. This loss only shows up while the cylinder is *moving* (drawing flow); at rest, the pressure at the cylinder equalizes back up toward the regulator setpoint. This is why a cylinder's static, at-rest holding force can look fine on a bench check while it still stalls or moves sluggishly under dynamic load — the two conditions are testing different things.

**Rule of thumb:** apply a combined efficiency factor of roughly 0.65-0.85 to the theoretical P x A force to estimate actual delivered force, with the low end of that range for small-bore (under about 25 mm), long-stroke, or poorly maintained/undersized-plumbing systems, and the high end for well-maintained, generously plumbed, medium-to-large bore systems like the one in this example. This breaks down for very large-bore, short-stroke cylinders with short, generously sized tubing runs, where efficiency can exceed 0.85 — and it breaks down in the other direction for anything with a long, restrictive air path or a compressor system that struggles to hold pressure under peak demand, where 0.65 can still be optimistic. When in doubt, verify with a flow-and-pressure-drop calculation across the actual valve and tubing selected (Design Task 4 develops the tool for that) rather than assuming a single efficiency number covers every installation.

### Governing Calculation

$$F_{actual} = \eta_{total} \times P \times A \qquad \eta_{total} = \eta_{friction} \times \eta_{pressure}$$

| Symbol | Description | Unit |
| :--- | :--- | :--- |
| $\eta_{friction}$ | Efficiency factor for seal/guide friction losses | dimensionless (0-1) |
| $\eta_{pressure}$ | Efficiency factor for dynamic pressure drop (valve, fittings, tubing) | dimensionless (0-1) |
| $\eta_{total}$ | Combined efficiency factor | dimensionless (0-1) |
| $F_{actual}$ | Actual force delivered at the rod under dynamic (moving) conditions | N |

```{r fig-06-07-force-comparison, echo=FALSE, fig.cap="Theoretical vs. actual force on each stroke of the selected 50 mm/20 mm cylinder, against each stroke's requirement (dashed line).", fig.align='center', fig.width=10, fig.height=6}
library(ggplot2)

bore <- 50; rod <- 20
A_bore <- pi / 4 * bore^2
A_annulus <- pi / 4 * (bore^2 - rod^2)
P_pa <- 6e5
eta_total <- 0.90 * 0.90

force_df <- data.frame(
  stroke = rep(c("Extend (lower)", "Retract (lift)"), each = 2),
  type = factor(rep(c("Theoretical", "Actual"), 2), levels = c("Theoretical", "Actual")),
  force = c(P_pa * A_bore * 1e-6, P_pa * A_bore * 1e-6 * eta_total,
            P_pa * A_annulus * 1e-6, P_pa * A_annulus * 1e-6 * eta_total)
)
req_df <- data.frame(stroke = c("Extend (lower)", "Retract (lift)"), required = c(180, 550))

ggplot(force_df, aes(x = type, y = force, fill = type)) +
  geom_col(color = "black", width = 0.6) +
  geom_hline(data = req_df, aes(yintercept = required), linetype = "dashed", color = "#E74C3C", linewidth = 1) +
  geom_text(aes(label = sprintf("%.0f N", force)), vjust = -0.5, size = 3.3 * diagram_text_scale) +
  facet_wrap(~stroke) +
  scale_fill_manual(values = c("Theoretical" = "#3498DB", "Actual" = "#F39C12")) +
  labs(title = "Theoretical vs. Actual Force by Stroke",
       subtitle = "Dashed red line = required force for that stroke",
       x = NULL, y = "Force (N)", fill = NULL) +
  theme_minimal() +
  theme(plot.title = element_text(size = 13.5 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 10 * diagram_text_scale),
        legend.position = "bottom",
        strip.text = element_text(size = 11 * diagram_text_scale, face = "bold"),
        axis.text = element_text(size = 10 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale))
```

### Worked Example

Live cell — try different friction/pressure-drop efficiency factors and see how much margin survives.

```{webr-r}
#| label: theoretical-actual-force-06-07
## Given: selected 50 mm bore / 20 mm rod cylinder (Design Task 1)
bore <- 50; rod <- 20
A_bore <- pi / 4 * bore^2
A_annulus <- pi / 4 * (bore^2 - rod^2)
P_pa <- 6e5   # 6 bar gauge

F_lower_required <- 180
F_lift_required <- 550

## Efficiency factors
eta_friction <- 0.90   # 10% lost to seal/guide friction
eta_pressure <- 0.90   # 10% lost to dynamic pressure drop (valve + tubing)
eta_total <- eta_friction * eta_pressure

F_extend_theor  <- P_pa * A_bore * 1e-6
F_retract_theor <- P_pa * A_annulus * 1e-6
F_extend_actual  <- F_extend_theor * eta_total
F_retract_actual <- F_retract_theor * eta_total

cat(sprintf("Combined efficiency factor: eta_total = %.2f x %.2f = %.3f\n\n", eta_friction, eta_pressure, eta_total))

cat(sprintf("Extend (lower): theoretical %.1f N -> actual %.1f N (required %.0f N, margin %.1f%%)\n",
            F_extend_theor, F_extend_actual, F_lower_required,
            (F_extend_actual / F_lower_required - 1) * 100))
cat(sprintf("Retract (lift): theoretical %.1f N -> actual %.1f N (required %.0f N, margin %.1f%%)\n",
            F_retract_theor, F_retract_actual, F_lift_required,
            (F_retract_actual / F_lift_required - 1) * 100))
```

**Sanity check:** the extend stroke passes with enormous margin (over 400%) because its requirement was modest to begin with — no surprise there. The retract stroke is the one that actually tests the design: theoretical force (989.6 N) looked like a comfortable 79.9% oversize over the 550 N requirement, but once seal friction and dynamic pressure drop are applied, actual delivered force drops to about 801.6 N — still a passing 45.7% margin, but noticeably less than the paper number suggested. This is exactly why Design Task 1's rule of thumb exists: if the design had landed at the 40 mm/16 mm combination (only 15.2% theoretical margin), the same 0.81 efficiency factor would have pulled actual force *below* the 550 N requirement, and the cylinder would have failed in the field despite passing a naive P x A check.

## Design Task 3: Air Consumption for a Given Cycle Rate

### Scenario

With the cylinder sized and its force margin verified, the plant's utilities engineer needs a different number: how much compressed air is this one cylinder actually going to draw, and at what rate, once it's running at the line's production pace? The stamping line runs this pick-and-place station at **30 cycles per minute**, each cycle consisting of one full extend stroke and one full retract stroke over a 100 mm stroke length. That air consumption number feeds two downstream decisions: whether the plant's compressor capacity and distribution piping can support adding this station (and however many more like it), and — as Design Task 6 shows — what it actually costs to run.

### Design Guidelines and Rules of Thumb

A cylinder consumes air twice per cycle: once filling the bore-side chamber on extend, once filling the rod-side (annulus) chamber on retract. The volume that matters for compressor sizing isn't the *compressed* volume sitting inside the cylinder at line pressure — it's the equivalent **free-air volume**, the volume that same amount of air would occupy at atmospheric pressure, because that's the basis compressor capacity (SCFM — standard cubic feet per minute) and most flow meters are rated on. Compressed gas at higher pressure is denser, so a small compressed volume corresponds to a much larger free-air volume; the conversion factor is the ratio of absolute line pressure to absolute atmospheric pressure.

**Rule of thumb:** for a first-pass air consumption estimate, treat each stroke's chamber as filling completely from atmospheric to line pressure every cycle, with no credit for any air recovered or reused between strokes (real systems occasionally use exhaust-air recovery or metering to trim this, but that's an optimization layered on top of the baseline calculation, not a substitute for it). This is a conservative (slightly high) estimate, which is the right direction to be wrong in for compressor and distribution sizing — but it breaks down as a *precise* prediction for cylinders with volume-reducing features like cushioning end caps or foam-filled rods, which reduce the effective displaced volume slightly, and for systems using flow-control (speed control) valves that meter the exhaust, which changes the dynamics of the fill but not the total volume consumed per cycle.

### Governing Calculation

Volume displaced per stroke, then converted from compressed (line-pressure) volume to free-air volume, times cycle rate:

$$V_{cycle,compressed} = (A_{bore} + A_{annulus}) \times L_{stroke}$$

$$V_{cycle,free} = V_{cycle,compressed} \times \frac{P_{abs}}{P_{atm}} \qquad Q_{free} = V_{cycle,free} \times N_{cycles/min}$$

| Symbol | Description | Unit |
| :--- | :--- | :--- |
| $L_{stroke}$ | Cylinder stroke length | mm |
| $V_{cycle,compressed}$ | Total volume displaced per cycle at line pressure | mm³ or L |
| $P_{abs}$ | Absolute line pressure ($P_{gauge} + P_{atm}$) | bar (abs) |
| $P_{atm}$ | Atmospheric pressure (standard: 1.013 bar abs) | bar (abs) |
| $V_{cycle,free}$ | Free-air (atmospheric-equivalent) volume per cycle | L |
| $N_{cycles/min}$ | Cycle rate | cycles/min |
| $Q_{free}$ | Free-air flow rate (NL/min, convertible to SCFM) | NL/min |

```{r fig-06-07-air-consumption-curve, echo=FALSE, fig.cap="Free-air consumption vs. cycle rate for the 50 mm/20 mm cylinder. The relationship is linear; this station's 30 cycles/min operating point is marked.", fig.align='center', fig.width=10, fig.height=5.5}
library(ggplot2)

bore <- 50; rod <- 20
A_bore <- pi/4*bore^2; A_annulus <- pi/4*(bore^2-rod^2)
stroke_mm <- 100
P_bar <- 6; P_atm_bar <- 1.013
ratio <- (P_bar + P_atm_bar) / P_atm_bar
V_free_L <- (A_bore + A_annulus) * stroke_mm / 1e6 * ratio

rate_seq <- seq(5, 60, by = 1)
df_rate <- data.frame(rate = rate_seq, Q_SCFM = V_free_L * rate_seq / 28.317)

ggplot(df_rate, aes(x = rate, y = Q_SCFM)) +
  geom_line(color = "#3498DB", linewidth = 1.2) +
  annotate("point", x = 30, y = V_free_L * 30 / 28.317, color = "#E74C3C", size = 4) +
  annotate("text", x = 32, y = V_free_L * 30 / 28.317 - 0.5,
           label = sprintf("30 cycles/min\n%.2f SCFM", V_free_L * 30 / 28.317),
           size = 3.2 * diagram_text_scale, color = "#E74C3C", hjust = 0) +
  labs(title = "Free-Air Consumption vs. Cycle Rate",
       subtitle = "50 mm bore / 20 mm rod, 100 mm stroke, 6 bar line pressure",
       x = "Cycle rate (cycles/min)", y = "Free-air flow (SCFM)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 13.5 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 10 * diagram_text_scale),
        axis.text = element_text(size = 10 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale))
```

### Worked Example

Live cell — try a different stroke length or cycle rate and see how consumption scales.

```{webr-r}
#| label: air-consumption-06-07
## Given
bore <- 50; rod <- 20
A_bore <- pi / 4 * bore^2
A_annulus <- pi / 4 * (bore^2 - rod^2)
stroke_mm <- 100
P_bar <- 6
P_atm_bar <- 1.013
cycles_per_min <- 30

V_extend_L  <- A_bore * stroke_mm / 1e6
V_retract_L <- A_annulus * stroke_mm / 1e6
V_compressed_L <- V_extend_L + V_retract_L

ratio <- (P_bar + P_atm_bar) / P_atm_bar
V_free_L <- V_compressed_L * ratio

Q_NLmin <- V_free_L * cycles_per_min
Q_SCFM  <- Q_NLmin / 28.317

cat(sprintf("Compressed volume/cycle: extend %.4f L + retract %.4f L = %.4f L\n",
            V_extend_L, V_retract_L, V_compressed_L))
cat(sprintf("Compression ratio (Pabs/Patm): %.3f\n", ratio))
cat(sprintf("Free-air volume per cycle: %.4f L\n\n", V_free_L))
cat(sprintf("At %d cycles/min: %.2f NL/min = %.3f SCFM\n", cycles_per_min, Q_NLmin, Q_SCFM))
```

**Sanity check:** roughly 1.7 SCFM per cylinder sounds small in isolation, and it is — a single small pick-and-place cylinder is not what strains a plant's compressor capacity. The reason this number matters is that it never appears alone: a plant this size might have dozens of similar stations running simultaneously, plus larger clamp cylinders, tooling cylinders, and blow-off nozzles, all drawing from the same header. Design Task 6 picks this exact number back up to show how a seemingly trivial per-cylinder inefficiency compounds into a real annual cost once it's multiplied across a plant-wide fleet.

## Design Task 4: Valve Flow Coefficient (Cv) Selection

### Scenario

Air consumption (Design Task 3) tells the utilities engineer about steady, cycle-averaged demand — but it says nothing about whether the directional control valve feeding this cylinder is big enough to fill it *fast*. The process engineer has specified that the extend (lower) stroke must fill and complete within **0.15 s**, so the transfer robot downstream isn't kept waiting. That's a much more demanding, momentary flow requirement than the cycle-averaged 1.7 SCFM from Design Task 3 — filling a chamber in a fraction of a second means moving that same volume of air at a far higher instantaneous rate. Sizing the valve against the average consumption number would produce a valve that's badly undersized for the actual fill-time requirement.

### Design Guidelines and Rules of Thumb

A pneumatic valve's flow capacity is rated by its **flow coefficient, Cv** — a manufacturer-published number (determined by standardized flow testing) that captures how much flow the valve passes for a given pressure drop across it. Bigger internal passages mean a bigger Cv and less restriction, but also a bigger, more expensive valve with a larger footprint and higher air consumption for the same actuation, so Cv selection is a genuine sizing trade-off, not "always pick the biggest one available."

**Rule of thumb:** size the valve against the *burst* flow rate implied by the required fill time — not the cycle-averaged consumption rate from Design Task 3 — because the valve has to pass the entire chamber-fill volume within that short window, which demands a far higher instantaneous flow rate. This breaks down (in the direction of needing an even larger valve) when the tubing between the valve and cylinder is long or narrow enough that *its* restriction, not the valve's, becomes the limiting factor — at that point, upsizing the valve alone won't fix the fill time, and the tubing run itself needs to be shortened or upsized. It also breaks down in the other direction for cylinders fitted with flow-control (meter-out) speed-control valves at the cylinder ports, which deliberately restrict flow to control speed — in that case the flow-control valve, not the directional valve, sets the actual fill time, and the directional valve can be sized more generously without penalty.

The governing relationship for compressible (air) flow through a valve is more involved than simple liquid flow, because air's density changes substantially as it expands across the pressure drop. A widely used simplified sizing equation for **subsonic (non-choked) air flow** — valid whenever the downstream pressure is at least about 53% of the upstream absolute pressure, the normal case for cylinder-filling applications — is commonly published in pneumatic and process valve application guides:

$$C_v = \frac{Q}{22.48} \sqrt{\frac{SG \times T}{\Delta P \times P_2}}$$

This is an industry-standard simplified sizing formula (not the full ISA/IEC compressible-flow control-valve standard, which adds correction factors for valve geometry and choked-flow behavior) — adequate for first-pass valve selection, with the manufacturer's own published Cv-vs-flow curves as the final check before ordering.

### Governing Calculation

$$Q_{burst} = \frac{V_{free,stroke}}{t_{fill}} \qquad C_v = \frac{Q_{burst}}{22.48}\sqrt{\frac{SG \times T}{\Delta P \times P_2}}$$

| Symbol | Description | Unit |
| :--- | :--- | :--- |
| $V_{free,stroke}$ | Free-air volume to fill for the stroke in question | ft³ |
| $t_{fill}$ | Required fill time | min |
| $Q_{burst}$ | Required burst (peak instantaneous) flow rate | SCFM |
| $SG$ | Specific gravity of the gas relative to air ($SG = 1.0$ for air) | dimensionless |
| $T$ | Absolute temperature | °R (°F + 460) |
| $P_1, P_2$ | Upstream, downstream absolute pressure across the valve | psia |
| $\Delta P$ | Pressure drop across the valve ($P_1 - P_2$) | psi |
| $C_v$ | Required valve flow coefficient | dimensionless |

```{r fig-06-07-cv-vs-filltime, echo=FALSE, fig.cap="Required valve Cv rises sharply as the target fill time shortens. The station's 0.15 s fill-time requirement is marked.", fig.align='center', fig.width=10, fig.height=5.5}
library(ggplot2)

bore <- 50
A_bore <- pi/4*bore^2
stroke_mm <- 100
P_bar <- 6; P_atm_bar <- 1.013
ratio <- (P_bar+P_atm_bar)/P_atm_bar
V_free_extend_L <- A_bore*stroke_mm/1e6*ratio
V_free_extend_ft3 <- V_free_extend_L/28.317

P1_psia <- 87 + 14.7
dP_psi <- 15
P2_psia <- P1_psia - dP_psi
T_R <- 530; SG <- 1.0

t_seq <- seq(0.05, 0.5, by = 0.01)
Q_seq <- V_free_extend_ft3 / (t_seq/60)
Cv_seq <- Q_seq/22.48*sqrt(SG*T_R/(dP_psi*P2_psia))
df_cv <- data.frame(t_fill = t_seq, Cv = Cv_seq)

ggplot(df_cv, aes(x = t_fill, y = Cv)) +
  geom_line(color = "#3498DB", linewidth = 1.2) +
  annotate("point", x = 0.15, y = (V_free_extend_ft3/(0.15/60))/22.48*sqrt(SG*T_R/(dP_psi*P2_psia)),
           color = "#E74C3C", size = 4) +
  annotate("text", x = 0.17, y = 0.75, label = "0.15 s fill time\nrequired Cv = 0.55",
           size = 3.2 * diagram_text_scale, color = "#E74C3C", hjust = 0) +
  labs(title = "Required Valve Cv vs. Target Fill Time",
       subtitle = "Extend-stroke fill, 50 mm bore, 100 mm stroke, 6 bar supply, 15 psi valve pressure drop",
       x = "Target fill time (s)", y = "Required Cv") +
  theme_minimal() +
  theme(plot.title = element_text(size = 13 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 9.5 * diagram_text_scale),
        axis.text = element_text(size = 10 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale))
```

### Worked Example

Live cell — try a different fill time or valve pressure drop and see how the required Cv responds.

```{webr-r}
#| label: valve-cv-sizing-06-07
## Given
bore <- 50
A_bore <- pi / 4 * bore^2
stroke_mm <- 100
P_bar <- 6; P_atm_bar <- 1.013
ratio <- (P_bar + P_atm_bar) / P_atm_bar

t_fill_s <- 0.15    # required extend-stroke fill time

V_free_extend_L <- A_bore * stroke_mm / 1e6 * ratio
V_free_extend_ft3 <- V_free_extend_L / 28.317
Q_burst_SCFM <- V_free_extend_ft3 / (t_fill_s / 60)

## Valve sizing conditions
P1_psig <- 87              # supply pressure, psig (6 bar)
P1_psia <- P1_psig + 14.7
dP_psi <- 15                # assumed valve pressure drop at this flow
P2_psia <- P1_psia - dP_psi
T_R <- 530                  # 70 deg F, absolute
SG <- 1.0                   # air

Cv_required <- Q_burst_SCFM / 22.48 * sqrt(SG * T_R / (dP_psi * P2_psia))

cat(sprintf("Free-air volume to fill (extend stroke): %.4f L = %.5f ft^3\n", V_free_extend_L, V_free_extend_ft3))
cat(sprintf("Required burst flow for %.2f s fill: %.2f SCFM\n\n", t_fill_s, Q_burst_SCFM))
cat(sprintf("P1 = %.1f psia, P2 = %.1f psia, dP = %.1f psi\n", P1_psia, P2_psia, dP_psi))
cat(sprintf("Required Cv: %.3f\n", Cv_required))
cat(sprintf("-> Select a catalog valve rated Cv >= %.2f (round up for margin)\n", ceiling(Cv_required*10)/10))
```

**Sanity check:** a required Cv of about 0.55 corresponds to a compact, standard 1/4" or 3/8" NPT solenoid directional valve in essentially every major manufacturer's catalog — physically consistent with a small pick-and-place cylinder, which is a useful order-of-magnitude cross-check. Note this burst-flow-derived Cv (0.55) is a very different number from what a naive average-consumption approach would suggest: the cycle-averaged flow from Design Task 3 (1.7 SCFM) is over an order of magnitude lower than the 19.2 SCFM burst flow this fill-time requirement actually demands. Sizing the valve off the average number would have produced a valve roughly a third the required capacity — a common and costly mistake.

## Design Task 5: FRL (Filter-Regulator-Lubricator) Sizing

### Scenario

A poultry processing plant is adding a pneumatic portioning station: three identical cylinders on a single manifold branch, each with the same average air demand as the pick-and-place cylinder above (2.65 SCFM at a comparable cycle rate). Before any of those cylinders can be connected, the plant's controls engineer has to specify the FRL unit — filter, regulator, and (possibly) lubricator — that conditions the compressed air feeding that branch. In a food-grade, washdown-rated environment, this isn't just a flow-sizing exercise: the lubrication question in particular has a food-safety dimension that a dry-goods automotive plant doesn't have to think about in the same way.

### Design Guidelines and Rules of Thumb

Each element of an FRL does a distinct job, and skipping the "what does this component actually do" step is how plants end up with either a unit that's the wrong type for the application or, just as often, an unnecessary lubricator quietly causing problems downstream.

- **Filter** removes particulate contamination (pipe scale, compressor wear debris, condensed moisture) from the compressed air before it reaches downstream components. A general-purpose 5 micron particulate filter is standard for most pneumatic actuator applications; finer coalescing filters (removing sub-micron oil aerosol and moisture) are used where air quality requirements are tighter — instrumentation air, or, as discussed below, food-contact-adjacent applications even without a lubricator in the circuit.
- **Regulator** reduces and holds the plant's main header pressure down to the pressure the local branch or machine actually needs, and — just as importantly — isolates that branch from header pressure fluctuations caused by other equipment cycling elsewhere in the plant. Every regulator has its own flow-vs-pressure-drop curve; undersizing it for the branch's actual flow demand causes the same kind of dynamic pressure sag Design Task 2 already flagged as a force-robbing loss.
- **Lubricator** injects a fine oil mist into the air stream to lubricate valve and cylinder internals downstream. This is the component most often specified out of habit rather than actual need.

**Rule of thumb:** size the regulator and filter for the branch's total flow demand with the same 25-50% oversize margin used for cylinder bore sizing (Design Task 1) — undersizing either one reintroduces the dynamic pressure-drop losses that Design Task 2 already showed eat directly into delivered force. This breaks down (toward needing *more* margin) when multiple cylinders on the same branch can cycle simultaneously rather than in sequence — sizing for one cylinder's flow and simply not accounting for the others cycling together is a common branch-sizing mistake.

On lubrication specifically: **the majority of modern pneumatic cylinders and valves are pre-lubricated at the factory with a lifetime grease and use lube-free-compatible seal materials (commonly PTFE-based).** Adding an oil-mist lubricator to the supply air for these components doesn't help — it can actually wash out or dilute the factory-applied grease over time, and it commits every downstream component on that branch to running lubricated air indefinitely, since mixing lubricated and non-lubricated air on a shared header causes uneven, unpredictable lubrication downstream. Only add a lubricator when a specific downstream component's datasheet explicitly calls for lubricated air (some pneumatic tools and older or specialty cylinder designs still do) — and in that case, route it on its own isolated branch, not the plant's general lube-free header. In a food-grade or washdown environment specifically, oil-mist lubrication carries an additional, harder line: any oil in the compressed air is a potential product-contamination and floor-slip-hazard concern, so **the correct default in food processing is no lubricator at all**, relying entirely on pre-lubricated, lube-free-compatible components, with food-grade (NSF H1-rated) grease used only at the factory or during maintenance re-lubrication of a component's internals — never injected into the supply air.

**CSA B51/TSSA** governs the pressure piping and receiver design for the compressed air distribution system itself in Ontario. **ISO 8573-1** (Compressed air — Contaminants and purity classes) is the standard reference for specifying the required particulate, moisture, and oil content class for a given application's air quality — cite the relevant class by number in a spec rather than restating its numeric limits here.

```{r tbl-06-07-lubrication-decision, echo=FALSE}
library(knitr)
library(kableExtra)

lube_df <- data.frame(
  Situation = c(
    "Modern pre-lubricated cylinder/valve (PTFE-based seals, lifetime factory grease)",
    "Older or specialty component whose datasheet explicitly specifies lubricated air",
    "Food-grade or washdown-rated environment (any component type)",
    "Shared header feeding both lube-free and lubrication-requiring components"
  ),
  Recommendation = c(
    "No lubricator. Running lubricated air can wash out factory grease over time and adds cost/maintenance for no benefit.",
    "Add a lubricator on a dedicated, isolated branch feeding only that component — not the general plant header.",
    "No lubricator, by default. Oil carryover is a contamination and slip-hazard risk; use pre-lubricated, lube-free components and NSF H1 grease only for factory/manual maintenance re-lubrication.",
    "Never mix on one branch. Split into a lube-free header and a separate lubricated branch; do not let one contaminate the other."
  )
)

kable(lube_df, caption = "FRL lubrication decision framework") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = knitr::is_html_output()) %>%
  column_spec(1, width = "2.3in") %>%
  column_spec(2, width = "4.2in") %>%
  row_spec(0, bold = TRUE, background = "#4a86c7", color = "white")
```

### Governing Calculation

There's no single equation for the lubrication decision — it's a judgment call driven by seal type and process environment, captured in the table above. The flow-sizing side of FRL selection, though, does reduce to a straightforward inequality against the manufacturer's rated flow-vs-pressure-drop curve:

$$Q_{FRL,rated} \geq \left(\sum_{i} Q_{i,average}\right) \times (1.25\ \text{to}\ 1.50)$$

| Symbol | Description | Unit |
| :--- | :--- | :--- |
| $Q_{i,average}$ | Cycle-averaged free-air flow demand of cylinder $i$ on the branch | SCFM |
| $Q_{FRL,rated}$ | FRL's manufacturer-rated flow capacity at an acceptable pressure drop (commonly quoted at 0.3-0.5 bar / 4-7 psi drop) | SCFM |

### Worked Example

Live cell — try a different number of cylinders on the branch, or a different oversize margin.

```{webr-r}
#| label: frl-sizing-06-07
## Given: 3 identical portioning cylinders on one manifold branch
Q_per_cylinder_SCFM <- 2.65   # from Design Task 3's air consumption calc
n_cylinders <- 3
oversize_factor <- 1.30        # 30% margin, within the 25-50% guideline

Q_branch_total <- Q_per_cylinder_SCFM * n_cylinders
Q_FRL_required <- Q_branch_total * oversize_factor

cat(sprintf("Per-cylinder average demand: %.2f SCFM\n", Q_per_cylinder_SCFM))
cat(sprintf("Branch total (%d cylinders): %.2f SCFM\n", n_cylinders, Q_branch_total))
cat(sprintf("Required FRL rated flow (with %.0f%% margin): %.2f SCFM\n\n",
            (oversize_factor - 1) * 100, Q_FRL_required))

## Decision: lubrication?
components_prelubricated <- TRUE
environment_washdown <- TRUE

lube_decision <- if (environment_washdown) {
  "NO lubricator - washdown/food-grade environment overrides even if components could accept lubricated air"
} else if (components_prelubricated) {
  "NO lubricator - components are pre-lubricated, lube-free-compatible seals"
} else {
  "Lubricator needed on a dedicated isolated branch - check component datasheets"
}

cat(sprintf("Filter: 5 micron general-purpose, or coalescing filter given food-adjacent environment\n"))
cat(sprintf("Regulator: select rated >= %.1f SCFM at <= 0.3 bar drop\n", Q_FRL_required))
cat(sprintf("Lubrication decision: %s\n", lube_decision))
```

**Sanity check:** the required 10.3 SCFM rated flow corresponds to a common mid-size (3/8" NPT class) FRL unit in most manufacturers' catalogs — proportionate to three small cylinders on one branch, not an oversized industrial unit. The lubrication decision resolves cleanly here because both independent reasons point the same direction — pre-lubricated components *and* a washdown environment — but it's worth noting that either factor alone would already have been sufficient to rule out a lubricator; the environment alone is a hard override even for a hypothetical component that technically accepted lubricated air.

<details>
<summary>Industry connection: air quality classes beyond food processing</summary>

The same ISO 8573-1 air-quality-class thinking that rules out lubricators in a food plant shows up again, for a different reason, in defense-sector precision manufacturing: pneumatically actuated fixturing or gauging equipment on a Controlled Goods Program production line is often specified to a tighter particulate and oil-carryover class than a general assembly application would need, not for contamination-of-product reasons but because oil residue on precision-machined or optical components is a functional defect, and the air quality specification becomes part of the auditable process documentation, not just a nice-to-have.

</details>

## Design Task 6: Compressed Air Supply Basics and the Cost of Air

### Scenario

The plant's engineering manager has heard the phrase "compressed air is the most expensive utility in the building" enough times to be skeptical of it — air is, after all, free; the plant is only paying for compression. The mechanical designer decides to put an actual number on it, using two comparisons: what does it cost, per year, to run the pick-and-place cylinder from Design Task 1 at its right-sized 50 mm bore, versus what it would have cost if someone had "played it safe" and specified the next catalog size up, 63 mm, on the theory that bigger can't hurt? And separately — because it's the example every compressed-air audit eventually reaches for — what does a single, unrepaired 3 mm leak cost the plant over a year?

### Design Guidelines and Rules of Thumb

Compressed air is generated by running an electric motor to drive a compressor, and that conversion is inherently lossy — a meaningful fraction of the electrical energy going into the compressor ends up as waste heat rather than usable compressed-air energy, before any of it even reaches a cylinder. That's the basis for treating compressed air as a utility with a real, calculable cost per unit volume, the same way a plant tracks cost per kWh of electricity or cost per cubic meter of natural gas.

**Rule of thumb:** a commonly cited industry planning figure puts the fully-loaded cost of generating compressed air at roughly **$0.20-0.35 per 1,000 standard cubic feet (SCF)**, reflecting typical industrial electricity rates and compressor efficiency. This is an order-of-magnitude planning number, not a precise, universal value — treat it as a rule of thumb, not a quoted standard. It breaks down site-to-site based on local electricity rates, compressor type and age, how well the distribution system is maintained (pressure drop and leaks both force the compressor to work harder for the same delivered flow), and altitude/ambient conditions affecting compressor efficiency. For an actual capital or energy-efficiency project, measure the specific plant's compressed air cost directly (from compressor kWh draw and delivered SCFM) rather than relying on the rule-of-thumb figure alone.

The oversizing case and the leak case are really the same underlying economic point told two different ways: both are compressed air being consumed with **no corresponding benefit** — the leak provides literally nothing, and the oversized cylinder's extra volume provides no more force or speed than the right-sized one already delivered with adequate margin (Design Task 2). Air leaks specifically tend to be underestimated because a leak is quiet, continuous, and invisible — unlike a mechanical failure, nothing stops working, so a leak can run for months or years without anyone noticing, unlike an oversized component, which was decided once and then forgotten.

### Governing Calculation

$$\text{Annual cost} = \frac{\text{Annual SCF consumed}}{1000} \times \text{Cost per 1,000 SCF}$$

Annual SCF consumed, for a cycling cylinder, extends Design Task 3's per-cycle free-air volume across the plant's annual operating hours:

$$\text{Annual SCF} = Q_{free}\ (\text{SCFM}) \times 60\ \frac{\text{min}}{\text{hr}} \times \text{Annual operating hours}$$

| Symbol | Description | Unit |
| :--- | :--- | :--- |
| $Q_{free}$ | Free-air flow rate (from Design Task 3) | SCFM |
| Annual operating hours | Plant's annual production hours for this equipment | h/yr |
| Cost per 1,000 SCF | Industry rule-of-thumb compressed air generation cost | $/1000 SCF |

```{r fig-06-07-cost-comparison, echo=FALSE, fig.cap="Annual compressed-air cost: right-sized vs. oversized cylinder (single unit and 25-unit fleet) compared to a single unrepaired 3 mm leak.", fig.align='center', fig.width=10, fig.height=6}
library(ggplot2)

cost_per_1000scf <- 0.28
hours_yr <- 4000
P_bar <- 6; P_atm_bar <- 1.013
ratio <- (P_bar+P_atm_bar)/P_atm_bar
stroke_mm <- 100; cycles_min <- 30

vol_free_L <- function(bore, rod) (pi/4*bore^2 + pi/4*(bore^2-rod^2)) * stroke_mm / 1e6 * ratio
annual_scf <- function(bore, rod) {
  Q_NLmin <- vol_free_L(bore, rod) * cycles_min
  (Q_NLmin/1000) * 60 * hours_yr / 0.028317
}

scf_50 <- annual_scf(50, 20)
scf_63 <- annual_scf(63, 20)
cost_50 <- scf_50/1000*cost_per_1000scf
cost_63 <- scf_63/1000*cost_per_1000scf
cost_63_fleet <- (cost_63 - cost_50) * 25 + cost_50

leak_scfm <- 22
leak_cost <- (leak_scfm*60*24*365)/1000*cost_per_1000scf

cost_df <- data.frame(
  scenario = c("50 mm (right-sized)\nsingle cylinder", "63 mm (oversized)\nsingle cylinder",
               "63 mm oversize x 25\nunits, incremental only", "Single 3 mm leak\n(continuous, plant-wide)"),
  cost = c(cost_50, cost_63, (cost_63-cost_50)*25, leak_cost)
)
cost_df$scenario <- factor(cost_df$scenario, levels = cost_df$scenario)

ggplot(cost_df, aes(x = scenario, y = cost, fill = scenario)) +
  geom_col(color = "black", show.legend = FALSE) +
  geom_text(aes(label = scales::dollar(cost, big.mark = " ")), vjust = -0.5, size = 3.3 * diagram_text_scale) +
  scale_fill_manual(values = c("#2ECC71", "#F39C12", "#E67E22", "#E74C3C")) +
  scale_y_continuous(labels = scales::dollar_format(big.mark = " "), expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Annual Compressed-Air Cost Comparison",
       subtitle = "At an illustrative industry rule-of-thumb rate of $0.28 per 1,000 SCF, 4 000 operating h/yr",
       x = NULL, y = "Annual cost") +
  theme_minimal() +
  theme(plot.title = element_text(size = 13.5 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 9.5 * diagram_text_scale),
        axis.text.x = element_text(size = 9 * diagram_text_scale),
        axis.text.y = element_text(size = 10 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale))
```

### Worked Example

Live cell — try a different local electricity-driven $/1000 SCF rate, annual operating hours, or fleet size.

```{webr-r}
#| label: cylinder-oversize-cost-06-07
## Given
cost_per_1000scf <- 0.28    # $/1000 SCF, industry rule-of-thumb range
hours_yr <- 4000             # 16 h/day x 5 days/wk x 50 wk/yr, two-shift operation
P_bar <- 6; P_atm_bar <- 1.013
ratio <- (P_bar + P_atm_bar) / P_atm_bar
stroke_mm <- 100; cycles_min <- 30

annual_scf <- function(bore, rod) {
  A_bore <- pi/4*bore^2
  A_annulus <- pi/4*(bore^2 - rod^2)
  V_free_L <- (A_bore + A_annulus) * stroke_mm / 1e6 * ratio
  Q_NLmin <- V_free_L * cycles_min
  (Q_NLmin / 1000) * 60 * hours_yr / 0.028317   # m3/yr -> SCF/yr
}

scf_50mm <- annual_scf(50, 20)   # right-sized (Design Task 1)
scf_63mm <- annual_scf(63, 20)   # "safer" oversized alternative

cost_50mm <- scf_50mm / 1000 * cost_per_1000scf
cost_63mm <- scf_63mm / 1000 * cost_per_1000scf
delta_per_unit <- cost_63mm - cost_50mm

n_units <- 25   # similar pick-and-place stations across the plant

cat(sprintf("Annual free-air use, 50 mm bore: %s SCF -> $%.2f/yr\n",
            format(round(scf_50mm), big.mark = " "), cost_50mm))
cat(sprintf("Annual free-air use, 63 mm bore: %s SCF -> $%.2f/yr\n",
            format(round(scf_63mm), big.mark = " "), cost_63mm))
cat(sprintf("Incremental cost of oversizing, per cylinder: $%.2f/yr\n\n", delta_per_unit))
cat(sprintf("Same oversizing mistake across a %d-unit plant-wide fleet: $%s/yr\n",
            n_units, format(round(delta_per_unit * n_units), big.mark = " ")))
```

Live cell — try a different leak orifice size or continuous run time.

```{webr-r}
#| label: leak-cost-06-07
## Given: a single unrepaired leak, common in real compressed-air audits
cost_per_1000scf <- 0.28
leak_diameter_mm <- 3       # roughly 1/8 in
leak_scfm <- 22              # from published compressed-air leak-rate charts at ~6 bar (87 psi)

leak_scf_per_year <- leak_scfm * 60 * 24 * 365
leak_cost_per_year <- leak_scf_per_year / 1000 * cost_per_1000scf

cat(sprintf("Leak orifice: %.0f mm diameter, continuous flow %.0f SCFM\n", leak_diameter_mm, leak_scfm))
cat(sprintf("Annual volume: %s SCF\n", format(round(leak_scf_per_year), big.mark = " ")))
cat(sprintf("Annual cost of this single leak: $%s\n", format(round(leak_cost_per_year), big.mark = " ")))
```

**Sanity check:** right-sizing this one cylinder type saves about $114/yr per unit — modest on its own, but $2,843/yr once multiplied across a 25-unit fleet of similar stations, which is a realistic count of pick-and-place-class cylinders across a mid-size plant. The single 3 mm leak, though, costs *more on its own* (roughly $3,238/yr) than the entire fleet's oversizing waste combined — a genuinely useful order-of-magnitude finding: a compressed-air system usually has more money leaking out through a handful of unrepaired fittings than through any one component being oversized. That's precisely why compressed air is often informally called the "fourth utility" in a plant and audited on a recurring schedule (leak surveys, typically ultrasonic, run quarterly to annually in well-managed plants) rather than being treated as a fixed, un-auditable cost of doing business — right-sizing new equipment and finding leaks in existing equipment are both worth doing, but they're not the same size of problem.

## Design Task 7: Pneumatic vs. Electric Actuators — A Selection Framework

### Scenario

A defense-sector precision instrument assembly cell needs a new linear axis: a placement head that must move to **four distinct, programmable positions** along its stroke (not just two end-of-travel positions) with a repeatability requirement of **±0.1 mm**, as part of an auditable, documented assembly process. A pneumatic cylinder — the natural first instinct after six design tasks spent sizing one — is fundamentally a two-position device: it goes to one hard stop or the other, and while proportional/servo-pneumatic valves exist that can hold intermediate positions, they add cost and complexity that starts to erode pneumatics' traditional advantage. Before defaulting to "just use an electric actuator instead," it's worth building the same kind of structured decision framework this book has used before for a comparable choice (recall the servo-vs-induction-motor decision in the [Motor, Drive, and Gearbox Sizing](06-06-Motor-Drive-Gearbox-Sizing.qmd) chapter) — because pneumatics remain the right answer for a large share of applications, including several already covered in this chapter.

### Design Guidelines and Rules of Thumb

The comparison comes down to several practical criteria, weighted differently depending on what the application actually needs:

- **Positioning precision and multi-position flexibility.** An electric actuator (ball-screw or belt-driven electric cylinder with a servo or stepper drive) has closed-loop position feedback and can be commanded to any position along its stroke, repeatably, under software control. A standard pneumatic cylinder moves between two hard-stop end positions; intermediate positions require either mechanical stops moved between cycles or a proportional valve, which is a different (and pricier) product category from a standard directional valve.
- **Force consistency.** Pneumatic force depends directly on line pressure, which sags under flow (Design Task 2) and varies with everything else pulling from the same header. Electric actuators deliver force through motor current, which a drive can hold consistent regardless of what else is happening on the plant floor.
- **Energy cost.** Design Task 6 already established that generating compressed air is an inherently lossy conversion. An electric actuator, powered directly from the plant's electrical supply with no intermediate compressed-gas stage, is typically substantially more energy-efficient over a comparable duty cycle — though for a low-duty-cycle or intermittent application, this difference may not be large enough in absolute dollar terms to be the deciding factor.
- **Purchase cost, complexity, and robustness.** A pneumatic cylinder, valve, and FRL branch is cheap, mechanically simple, and tolerant of a dusty, dirty, or washdown environment — this is precisely why it dominates applications like this chapter's simple two-position pick-and-place and portioning-station examples. An electric actuator and its servo/stepper drive cost more, need motion-control programming and tuning, and — depending on the specific product's ingress protection rating — can be more sensitive to a harsh environment than an equivalent pneumatic component.

**Rule of thumb:** default to electric when the application needs more than two discrete positions, needs programmable/adjustable positioning without physical retooling, needs positioning repeatability tighter than roughly 0.5 mm, or needs to hold a controlled, consistent force or position against a varying external load. Default to pneumatic when the motion is fundamentally two-position (or a small, fixed number of mechanically-stopped positions), when the environment is harsh, dirty, or washdown-rated, when duty cycle and precision demands are modest, or when the lower purchase cost and mechanical simplicity outweigh the energy and force-consistency advantages electric offers. This breaks down at the boundary for applications needing moderate positioning flexibility but only occasionally — some designs use a pneumatic cylinder with two or three mechanically indexed stop positions (a compromise sometimes called a "multi-position" pneumatic axis) as a lower-cost middle ground between a simple two-position cylinder and a full electric axis, appropriate when the number of required positions is small, fixed at design time, and doesn't need runtime reprogramming.

```{r fig-06-07-electric-vs-pneumatic, echo=FALSE, fig.cap="Illustrative comparison of electric vs. pneumatic linear actuation across six decision criteria. For Purchase Cost, a higher bar means lower cost.", fig.align='center', fig.width=10, fig.height=6}
library(ggplot2)
library(tidyr)

compare_df <- data.frame(
  Criterion = c("Positioning\nPrecision", "Multi-Position\nFlexibility", "Force\nConsistency",
                "Energy\nCost", "Purchase\nCost", "Environmental\nRobustness"),
  Electric = c(5, 5, 5, 4, 2, 3),
  Pneumatic = c(2, 1, 2, 2, 5, 5)
)
compare_long <- pivot_longer(compare_df, cols = c(Electric, Pneumatic), names_to = "Option", values_to = "Score")
compare_long$Criterion <- factor(compare_long$Criterion, levels = compare_df$Criterion)

ggplot(compare_long, aes(x = Criterion, y = Score, fill = Option)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Electric" = "#3498DB", "Pneumatic" = "#2ECC71")) +
  labs(title = "Electric vs. Pneumatic Linear Actuation — Illustrative Comparison",
       subtitle = "Relative rating, 1 (poor fit) to 5 (excellent fit) — actual scores are application-specific",
       x = NULL, y = "Relative rating (1-5)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 10 * diagram_text_scale),
        legend.position = "bottom",
        axis.text = element_text(size = 9.5 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale))
```

### Governing Calculation

As with the servo-vs-induction decision elsewhere in this book, there's no single formula — this is a documented, weighted judgment call:

$$Score_{option} = \sum_i w_i \times s_{i,option}$$

| Symbol | Description | Unit |
| :--- | :--- | :--- |
| $w_i$ | Weight assigned to criterion $i$ (weights sum to 1.0) | dimensionless |
| $s_{i,option}$ | Score of a given option against criterion $i$ (e.g., 1-5 scale) | dimensionless |
| $Score_{option}$ | Total weighted score for that option | dimensionless |

### Worked Example

Live cell — try different weights and see whether the recommendation flips.

```{webr-r}
#| label: electric-vs-pneumatic-decision-06-07
## Given: weights reflecting the defense-sector precision placement cell's priorities
weights <- c(precision = 0.30, multi_position = 0.20, force_consistency = 0.15,
             energy_cost = 0.10, purchase_cost = 0.15, robustness = 0.10)

electric_scores  <- c(precision = 5, multi_position = 5, force_consistency = 5,
                      energy_cost = 4, purchase_cost = 2, robustness = 3)
pneumatic_scores <- c(precision = 2, multi_position = 1, force_consistency = 2,
                      energy_cost = 2, purchase_cost = 5, robustness = 5)

electric_total  <- sum(weights * electric_scores)
pneumatic_total <- sum(weights * pneumatic_scores)

cat("Weighted decision matrix (1 = poor fit, 5 = excellent fit):\n\n")
for (nm in names(weights)) cat(sprintf("  %-20s weight = %.0f%%\n", nm, weights[nm] * 100))
cat(sprintf("\nElectric weighted score:   %.2f\n", electric_total))
cat(sprintf("Pneumatic weighted score:  %.2f\n", pneumatic_total))
cat(sprintf("\n-> Recommendation: %s\n", ifelse(electric_total > pneumatic_total, "ELECTRIC ACTUATOR", "PNEUMATIC CYLINDER")))
```

**Sanity check:** with precision and multi-position flexibility weighted most heavily — the correct priority for a four-position, ±0.1 mm assembly cell — electric wins decisively (3.85 vs. 2.35). Try dropping the precision and multi-position weights and raising purchase cost and robustness instead, closer to what this chapter's own pick-and-place and portioning-station examples actually needed: the recommendation flips to pneumatic, confirming that Design Tasks 1 through 6 were solving the right problem for those applications, not defaulting to pneumatics out of habit.

## Chapter Summary

```{r tbl-06-07-summary, echo=FALSE}
library(knitr)
library(kableExtra)

summary_df <- data.frame(
  `Design Task` = c(
    "1. Bore sizing from force",
    "2. Theoretical vs. actual force",
    "3. Air consumption",
    "4. Valve Cv selection",
    "5. FRL sizing",
    "6. Cost of air",
    "7. Pneumatic vs. electric"
  ),
  `Governing Equation` = c(
    "$F = P \\times A$ (use $A_{annulus}$ for the retract stroke)",
    "$F_{actual} = \\eta_{total} \\times P \\times A$",
    "$Q_{free} = (A_{bore}+A_{annulus}) L_{stroke} \\times \\frac{P_{abs}}{P_{atm}} \\times N_{cycles/min}$",
    "$C_v = \\frac{Q}{22.48}\\sqrt{\\frac{SG\\,T}{\\Delta P\\,P_2}}$",
    "$Q_{FRL,rated} \\geq \\sum Q_{i} \\times (1.25\\text{-}1.50)$",
    "$\\text{Cost} = \\frac{\\text{Annual SCF}}{1000}\\times \\text{\\$/1000 SCF}$",
    "$Score = \\sum_i w_i \\, s_{i,option}$"
  ),
  `Key Rule of Thumb` = c(
    "Oversize theoretical area 25-50% over the calculated minimum",
    "Combined efficiency factor 0.65-0.85 (friction x pressure drop)",
    "Assume full atmosphere-to-line-pressure fill each stroke, no recovery credit",
    "Size for burst (fill-time) flow, not average consumption",
    "No lubricator for pre-lubricated components or washdown/food-grade environments",
    "~$0.20-0.35 per 1,000 SCF is a planning-level industry figure, not a precise value",
    "Multi-position/precision -> electric; simple 2-position/harsh environment -> pneumatic"
  ),
  check.names = FALSE
)
summary_df[] <- lapply(summary_df, tex_safe)

kable(summary_df, caption = "Pneumatic system design tasks: governing equations and key rules of thumb", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = knitr::is_html_output()) %>%
  column_spec(1, width = "1.2in") %>%
  column_spec(2, width = "2.4in") %>%
  column_spec(3, width = "2.9in") %>%
  row_spec(0, bold = TRUE, background = "#4a86c7", color = "white")
```

## Practice Problems

<details>
<summary>Design Task 1, Problem 1 (plug-in): a clamp cylinder bore</summary>

A weld-fixture clamp cylinder needs 900 N of clamping force at 6 bar line pressure, with the clamp motion happening on the extend stroke (full bore area governs, not the annulus). Find the minimum bore area and diameter, and check whether a standard 50 mm bore clears the 25% oversize floor.

**Solution:**

$$A_{min} = \frac{900}{6\times10^5} = 0.0015\ \text{m}^2 = 1500\ \text{mm}^2 \qquad d_{min} = \sqrt{\frac{4(1500)}{\pi}} = 43.7\ \text{mm}$$

A 50 mm bore gives $A_{bore} = \pi/4(50)^2 = 1963.5\ \text{mm}^2$, an oversize margin of $(1963.5/1500 - 1) = 30.9\%$ — inside the 25-50% guideline band, so 50 mm is a good selection.

</details>

<details>
<summary>Design Task 1, Problem 2 (design decision): standard size doesn't fit the band</summary>

A different application's minimum annulus area works out such that the next standard bore up gives only an 18% margin, and the one after that gives 95%. Which do you select, and what design change (other than accepting one of those two bores) might resolve the mismatch entirely?

**Answer:** Select the larger bore (95% margin) for the same reason worked through in this chapter's own example — an inadequate margin is a functional risk, while extra margin is a cost and packaging issue. Beyond simply accepting one of the two standard sizes, two other design changes are worth considering before committing to the oversized cylinder: increasing line pressure (if the plant's compressor and distribution system can support it) reduces the required bore area for the same force, potentially bringing the smaller standard size back into an acceptable margin; alternatively, if the mismatch is being driven by a marginal force requirement, revisiting whether that requirement itself has its own conservative padding (a common occurrence when several designers each add their own safety margin at different stages) may reveal the true requirement is lower than assumed.

</details>

<details>
<summary>Design Task 2, Problem 1 (plug-in): actual force with a lower efficiency factor</summary>

Using the same 50 mm/20 mm cylinder (theoretical retract force 989.6 N), estimate actual retract force with a more conservative combined efficiency factor of 0.70 (long tubing run, aging compressor system). Does it still clear the 550 N requirement?

**Solution:**

$$F_{actual} = 989.6 \times 0.70 = 692.7\ \text{N}$$

Yes — still comfortably above 550 N (26% margin), though noticeably tighter than the 45.7% margin found with the 0.81 efficiency factor used in the chapter's worked example.

</details>

<details>
<summary>Design Task 2, Problem 2 (design decision): a small-bore application with thin margin</summary>

A different, smaller cylinder (16 mm bore) shows only a 12% actual-force margin after applying a 0.70 efficiency factor — appropriate for its small bore per this chapter's guidance. The project is under cost pressure to avoid moving up a size. What are your options besides accepting the thin margin or upsizing the cylinder?

**Answer:** A few options worth exploring before either accepting the risk or paying for a larger cylinder: (1) shorten and straighten the tubing run and eliminate unnecessary fittings between the valve and cylinder to improve the pressure-drop half of the efficiency factor; (2) verify the valve itself isn't the bottleneck — an undersized valve (Design Task 4) contributes directly to the pressure-drop loss; (3) check whether the seal/friction specification can be improved (a cylinder with a lower-friction seal package, sometimes a manufacturer option, directly improves the friction half of the efficiency factor); (4) confirm the regulator is actually holding setpoint under the branch's real flow demand, not sagging due to being undersized itself (Design Task 5). Only after those are checked and the margin is still thin should upsizing the cylinder be treated as the remaining option — small-bore applications are exactly where this chapter flagged margin as most fragile, so a 12% margin here deserves more scrutiny than the same number would on a large-bore cylinder.

</details>

<details>
<summary>Design Task 3, Problem 1 (plug-in): air consumption at a faster rate</summary>

Using the same 50 mm/20 mm cylinder and 100 mm stroke, find the free-air consumption in SCFM if the cycle rate increases to 45 cycles/min.

**Solution:**

Free-air volume per cycle stays the same as the chapter's worked example, 2.5012 L (this doesn't depend on cycle rate). At 45 cycles/min:

$$Q_{NL/min} = 2.5012 \times 45 = 112.6\ \text{NL/min} \qquad Q_{SCFM} = \frac{112.6}{28.317} = 3.97\ \text{SCFM}$$

</details>

<details>
<summary>Design Task 3, Problem 2 (design decision): compressor capacity check</summary>

The plant's existing compressor has 15 SCFM of uncommitted spare capacity. Engineering wants to add six more pick-and-place stations identical to this chapter's example (1.70 SCFM average each, from the chapter's worked example), each cycling continuously. Does the spare capacity cover it, and what should the engineer check beyond the raw SCFM sum?

**Answer:** Six stations at 1.70 SCFM average each sum to 10.2 SCFM, which is under the 15 SCFM spare capacity on paper — but average consumption alone isn't the whole picture. The engineer should also check: (1) whether these six stations' cycles are likely to overlap in time (if they're on a synchronized line, their peak/burst demand — the Cv-driving flow from Design Task 4, not the average — could spike well above the average sum, briefly starving other equipment on the same header even though the time-averaged number fits); (2) whether the "spare capacity" figure already accounts for existing leaks and inefficiencies in the current system (Design Task 6) — spare capacity calculated against a leaky baseline overstates what's actually available; (3) whether adding this load pushes the compressor's duty cycle high enough to affect its service life or trigger more frequent lead/lag staging in a multi-compressor plant. A comfortable-looking average-SCFM margin is a starting point, not a final answer.

</details>

<details>
<summary>Design Task 4, Problem 1 (plug-in): required Cv for a faster fill time</summary>

Using the same 50 mm bore extend-stroke fill (free-air volume 1.3593 L = 0.04800 ft³) and the same pressure conditions as the worked example, find the required Cv if the fill time requirement tightens to 0.08 s.

**Solution:**

$$Q_{burst} = \frac{0.04800}{0.08/60} = 36.0\ \text{SCFM}$$

$$C_v = \frac{36.0}{22.48}\sqrt{\frac{1.0 \times 530}{15 \times 86.7}} = 1.602 \times 0.6398 = 1.02$$

Roughly double the Cv found for the 0.15 s requirement, consistent with the inverse relationship between fill time and required flow.

</details>

<details>
<summary>Design Task 4, Problem 2 (design decision): valve Cv is adequate but fill time still misses target</summary>

A newly commissioned station's directional valve is confirmed to have Cv well above the calculated requirement, yet the cylinder still isn't filling within the specified time on the shop floor. What are the likely causes, given everything else covered in this chapter?

**Answer:** If the valve itself has adequate Cv, the bottleneck has moved somewhere else in the air path. Likely causes, in roughly the order worth checking: (1) undersized or excessively long tubing between the valve and cylinder — Design Task 4's own guidance flagged this as the point where upsizing the valve stops helping; (2) a meter-out flow-control (speed-control) valve installed at the cylinder port set too restrictive, deliberately throttling the fill; (3) the regulator or FRL upstream of the valve sagging under the flow demand (Design Task 5), meaning the valve never actually sees the assumed supply pressure during the fill; (4) a leak or other parasitic draw elsewhere on the same branch reducing available pressure/flow at the moment this cylinder fires. This is a good example of why a fill-time miss isn't automatically a "valve too small" problem — the valve is only one link in the chain this chapter built.

</details>

<details>
<summary>Design Task 5, Problem 1 (plug-in): FRL sizing for a larger branch</summary>

A branch feeds five cylinders, each averaging 2.2 SCFM, with a 40% oversize margin specified. Find the required FRL rated flow.

**Solution:**

$$Q_{branch} = 5 \times 2.2 = 11.0\ \text{SCFM} \qquad Q_{FRL,required} = 11.0 \times 1.40 = 15.4\ \text{SCFM}$$

</details>

<details>
<summary>Design Task 5, Problem 2 (design decision): a maintenance tech wants to add a lubricator</summary>

A maintenance technician, following habits from an older plant, wants to add an in-line lubricator to a branch feeding modern, pre-lubricated cylinders in a dry-goods (non-washdown) automotive plant, "just to be safe." Do you agree, and how would you explain your answer?

**Answer:** No — per this chapter's decision framework, a lubricator isn't a harmless "extra margin" the way an oversized cylinder bore might be; it actively works against pre-lubricated components by washing out their factory grease over time, and it locks the branch into needing lubricated air indefinitely (with the associated risk of accidentally cross-connecting a lube-free branch elsewhere in the plant). The correct response is to check the specific cylinder and valve manufacturer's datasheets: if none of them call for lubricated air, the answer is no lubricator, regardless of what practice was standard at a previous plant. This is a good moment to point out that "just to be safe" isn't automatically true margin — Design Task 1's oversize guideline is safe because more bore area only costs money and air, while an unnecessary lubricator can actively degrade the very components it's meant to help.

</details>

<details>
<summary>Design Task 6, Problem 1 (plug-in): leak cost at a smaller orifice</summary>

Using the same $0.28/1000 SCF rate, estimate the annual cost of a smaller 1.5 mm leak, given published leak-rate charts suggest roughly 6 SCFM continuous flow for that orifice size at 6 bar.

**Solution:**

$$\text{Annual SCF} = 6 \times 60 \times 24 \times 365 = 3\,153\,600\ \text{SCF/yr}$$

$$\text{Annual cost} = \frac{3\,153\,600}{1000} \times 0.28 \approx \$883/\text{yr}$$

Smaller than the chapter's 3 mm leak example (about $3,238/yr), but still a meaningful, entirely avoidable annual cost for a single leaking fitting.

</details>

<details>
<summary>Design Task 6, Problem 2 (design decision): where to spend a limited compressed-air budget</summary>

A plant's maintenance budget can fund either a leak-detection survey (typically finding 15-30 leaks of varying size in a plant this size) or a project to right-size six oversized cylinders identified during a design review. Which would you recommend prioritizing first, and why?

**Answer:** Prioritize the leak survey first, for the same reason this chapter's own comparison found the single leak example outweighing the entire fleet's cylinder-oversizing waste: leaks are typically the larger, faster-growing cost driver in a compressed air system, because they accumulate silently and continuously across the whole plant, while oversized components are a fixed, one-time inefficiency that was decided once. A leak survey also tends to have a faster payback (leaks are often cheap to fix — a fitting or valve replacement — once found) than redesigning and replacing six cylinders. That said, this isn't an either/or in the long run: right-sizing new designs going forward (this chapter's Design Tasks 1-4) prevents the next generation of oversized components from ever entering the system, so the real answer is "leak survey first, then build right-sizing discipline into the design process so there's less to retrofit later."

</details>

<details>
<summary>Design Task 7, Problem 1 (plug-in): re-weighting for a simple clamp application</summary>

Using the same electric/pneumatic scores as the worked example, but with weights of 10% precision, 5% multi-position, 10% force consistency, 15% energy cost, 35% purchase cost, 25% robustness (a simple, harsh-environment, two-position clamp), which option wins?

**Solution:**

$$Score_{electric} = 0.10(5)+0.05(5)+0.10(5)+0.15(4)+0.35(2)+0.25(3) = 3.10$$

$$Score_{pneumatic} = 0.10(2)+0.05(1)+0.10(2)+0.15(2)+0.35(5)+0.25(5) = 3.80$$

Pneumatic wins under these cost/robustness-weighted priorities — consistent with this chapter's own clamp and pick-and-place examples.

</details>

<details>
<summary>Design Task 7, Problem 2 (design decision): a borderline case</summary>

A packaging-line diverter needs to move between three fixed positions (not fully programmable, but more than two), running in a moderately dusty environment, with no unusual precision requirement. Electric or pneumatic, and what middle-ground option from this chapter's guidance is worth considering?

**Answer:** This sits close to the boundary this chapter flagged: three fixed positions is more than a standard two-position pneumatic cylinder handles directly, but the positions are fixed at design time (not runtime-programmable) and precision demands are modest — exactly the profile this chapter described for a mechanically-indexed multi-position pneumatic axis (a cylinder with a rotating or sliding multi-stop mechanism) as a lower-cost middle ground between simple pneumatics and a full electric axis. Given the dusty environment favors pneumatics' robustness and the requirement doesn't need the runtime flexibility or tight precision that would favor electric, a multi-stop pneumatic solution is worth quoting alongside a full electric axis before defaulting to either extreme — the deciding factor would likely come down to whether three positions is genuinely fixed for the life of the equipment (favoring pneumatic) or likely to need a fourth position or repositioning down the road (favoring electric's software-only reconfigurability).

</details>

## Open-Ended Design Scenario

<details>
<summary>Capstone: pneumatic clamp-and-transfer station for a dairy packaging line</summary>

A dairy processing plant is adding a clamp-and-transfer station to its cup-filling line. The station has two pneumatic axes:

**Clamp cylinder:** Extends horizontally to hold a tray of filled cups square against a locating fence while a robotic case-packer loads the tray, then retracts to release it. Required clamp force (extend stroke): 420 N. Required release force (retract stroke, overcoming friction and a light return spring): 90 N. Line pressure is 6 bar. The plant runs three shifts, 6 days/week, roughly 4,700 h/year.

**Lift-and-lower transfer cylinder:** A separate vertical cylinder, mounted rod-down, lowers an empty tray carrier onto the conveyor (extend, 150 N required) and lifts the loaded tray carrier clear (retract, 480 N required, governed by the annulus area). Stroke: 120 mm. The station cycles at 24 cycles/minute.

The plant's engineering standard requires stainless steel, IP69K washdown-rated pneumatic components throughout, consistent with CFIA sanitary design expectations for equipment operating near open product.

Using the methods developed in this chapter:

1. For each cylinder, work through Design Task 1: find the minimum required area on the governing stroke, apply the 25-50% oversize guideline, and select a standard ISO 15552 bore/rod combination for each (32/12, 40/16, 50/20, and 63/20 mm are all available options — you may need to check more than one candidate per cylinder, the way the chapter's own example did).
2. For the transfer cylinder, apply Design Task 2: using a combined efficiency factor of 0.75 (a washdown environment with somewhat longer stainless tubing runs than the chapter's automotive example), verify actual retract force still clears the 480 N requirement.
3. For the transfer cylinder, apply Design Task 3: calculate free-air consumption in SCFM at the specified 24 cycles/min cycle rate.
4. For the transfer cylinder's extend stroke, apply Design Task 4: if the process requires the lower stroke to complete within 0.20 s, calculate the required valve Cv (use the same pressure/temperature assumptions as the chapter's worked example: 6 bar supply, 15 psi valve drop, 530 °R).
5. Apply Design Task 5's decision framework: specify the FRL for this station (both cylinders on one branch), and make and justify the lubrication call given the washdown/food environment.
6. Apply Design Task 6: using the plant's actual 4,700 h/year operating schedule and the industry rule-of-thumb $/1000 SCF rate, estimate this station's annual compressed-air cost, and identify one design decision from steps 1-5 that would most directly reduce it.
7. Apply Design Task 7's framework only if asked to justify the choice of pneumatics at all for this station: build a short weighted decision matrix defending pneumatic over electric for a clamp-and-transfer application in a washdown dairy environment, using your own weights.

There's no single correct numeric answer — reasonable differences in assumed friction, oversize margin, and rounding to standard sizes will produce different (defensible) specifications. The goal is a documented sizing package, in the style of this chapter's worked examples, that could be handed to a pneumatic component distributor for an actual quotation.

</details>
