---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Capstone: Integrated Pick-and-Place Cell Design

```{r setup-06-15, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
```

## Learning Objectives

By the end of this chapter, you will be able to:

1. **Integrate** the sizing methods from nine preceding chapters into a single, internally consistent machine specification for a pneumatic robotic pick-and-place cell.
2. **Size** a pneumatic lift/gripper cylinder and its directional control valve for a specific load, stroke, and cycle-rate requirement.
3. **Size** an infeed conveyor's belt pull, drive power, motor, and gearbox for a specific throughput, and **justify** a VFD-driven induction motor over a servo for that application.
4. **Size** branch-circuit protection and conductors for a mixed-load control panel (robot, motor, valve solenoids), **verify** a voltage-drop check, and **state** the panel's SCCR requirement.
5. **Design** the cell's bonding and grounding scheme, including equipment bonding conductor sizing, VFD/motor bonding practice, and analog signal shield grounding.
6. **Derive** an overall safety-circuit stopping time for the specific cell and **calculate** the resulting ISO 13855 minimum safe distance for its guarding.
7. **Check** a structural element of the cell's frame against an appropriate deflection limit.
8. **Trace** how a change introduced in one subsystem (cycle rate, industry, robot classification) propagates through the other subsystems of an integrated design.

## Introduction

Every chapter so far in this part solved one problem at a time: size a cylinder, size a motor branch, calculate a bonding conductor, check a beam. That's how the individual calculations have to be learned — one governing equation, one worked example, one sanity check — but it is not how a real machine gets built. A real integration project lands on one engineer's desk as a single machine, due to ship in a fixed number of weeks, and every one of those calculations has to happen for *the same cell*, using numbers that actually agree with each other: the cylinder's cycle rate has to match the conveyor's throughput, the conveyor motor's nameplate current has to match the branch breaker sized to protect it, the branch breaker's rating has to match the bonding conductor sized off of it, and the safety circuit's stopping time has to match the guarding distance calculated from it.

This chapter is that integration exercise. The scenario: a Tier-2 automotive supplier in Cambridge, Ontario, machines die-cast aluminum shift-fork brackets for a transmission sub-assembly on an upstream CNC cell. The finished brackets arrive deburred and ready for inspection, riding single-file on a short powered infeed conveyor. A small six-axis robot, mounted on a fixed pedestal beside the conveyor's discharge end, picks each bracket with a pneumatic parallel gripper — the gripper carries its own small pneumatic lift stroke, separate from the robot's own motion, so the robot can swing away the instant the part is clear rather than commanding a full arm move for every pick — and places it into an indexing fixture tray that feeds a downstream vision-inspection and leak-test station. The line runs at 720 brackets per hour: 12 parts per minute, a 5.0 s cycle time per pick-and-place cycle, a realistic rate for this class of automotive sub-assembly cell.

Building this cell means working through seven design questions, in the order a real integration project actually asks them:

1. **System overview** — what has to be sized, and how the pieces fit together physically.
2. **Cylinder and valve sizing** — the gripper's pneumatic lift stroke, using the [Pneumatic System Design](06-07-Pneumatic-System-Design.qmd) chapter's method.
3. **Infeed conveyor motor and drive sizing** — belt pull and drive power from the [Conveyor Design](06-10-Conveyor-Design.qmd) chapter, reflected to an actual motor and gearbox using the [Motor, Drive, and Gearbox Sizing](06-06-Motor-Drive-Gearbox-Sizing.qmd) chapter's method.
4. **Panel, protection, and wiring** — branch-circuit protection, a voltage-drop check, and the panel's SCCR requirement, from the [Control Panel Design](06-03-Control-Panel-Design.qmd) and [Wire, Cable, and Circuit Protection](06-04-Wire-Cable-Circuit-Protection.qmd) chapters.
5. **Bonding and grounding** — equipment bonding conductor sizing, VFD/motor bonding, and shield grounding, from the [Bonding, Grounding, and EMC](06-05-Bonding-Grounding-EMC.qmd) chapter.
6. **Minimum safe distance** — an overall stopping time built from the [Machine Safety Circuit Design](06-02-Machine-Safety-Circuits.qmd) chapter's method, fed into the [Robot Cell Design](06-01-Robot-Cell-Design.qmd) chapter's ISO 13855 formula.
7. **Frame deflection check** — the robot pedestal's mounting beam, using the [Frame and Structural Design](06-11-Frame-Structural-Design.qmd) chapter's method.

Every number carried from one section into the next is tracked explicitly — this chapter's whole point is that the numbers have to agree, not just that each calculation is individually correct.

## 1. System Overview

```{r fig-06-15-cell-layout, echo=FALSE, fig.cap="Plan view of the pick-and-place cell: parts arrive on the infeed conveyor, the six-axis robot picks from the discharge end and places into the fixture tray, a light curtain guards the operator loading point, and the control panel sits at the cell's perimeter.", fig.align='center', fig.width=11, fig.height=7}
robot_base <- data.frame(x0 = 2.0, y0 = 1.2, r = 1.05)

ggplot() +
  ## Infeed conveyor
  annotate("rect", xmin = -0.2, xmax = 2.5, ymin = -0.35, ymax = 0.35, fill = "grey55", color = "black") +
  annotate("text", x = 1.15, y = 0.0, label = "Infeed conveyor (2.5 m, level)",
           size = 2.9 * diagram_text_scale, color = "white") +
  annotate("segment", x = -0.2, xend = 2.3, y = -0.55, yend = -0.55,
           arrow = arrow(length = unit(0.15, "cm"), type = "closed"), color = "black", linewidth = 0.6) +
  annotate("text", x = 1.0, y = -0.75, label = "part travel", size = 2.4 * diagram_text_scale) +
  ## Light curtain at the operator loading point
  annotate("segment", x = -0.9, xend = -0.9, y = -0.9, yend = 0.9, linewidth = 2, color = "steelblue") +
  annotate("text", x = -0.9, y = 1.2, label = "Light curtain\n(operator loading point)",
           size = 2.7 * diagram_text_scale, color = "steelblue") +
  ## Robot reach envelope and base
  geom_circle(data = robot_base, aes(x0 = x0, y0 = y0, r = r),
              color = "#2ECC71", linetype = "dashed", linewidth = 1, fill = NA) +
  annotate("rect", xmin = 1.75, xmax = 2.25, ymin = 0.95, ymax = 1.45, fill = "#34495E", color = "black") +
  annotate("text", x = 2.0, y = 1.2, label = "6-axis\nrobot", size = 2.6 * diagram_text_scale, color = "white") +
  annotate("text", x = 2.0, y = 2.45, label = "Reach envelope (~1.05 m)",
           size = 2.5 * diagram_text_scale, color = "#2ECC71") +
  ## Fixture tray
  annotate("rect", xmin = 3.0, xmax = 3.6, ymin = 1.5, ymax = 2.0, fill = "#E67E22", color = "black") +
  annotate("text", x = 3.3, y = 1.75, label = "Fixture\ntray", size = 2.5 * diagram_text_scale, color = "white") +
  ## Control panel
  annotate("rect", xmin = 3.5, xmax = 4.1, ymin = -0.7, ymax = -0.1, fill = "#7F8C8D", color = "black") +
  annotate("text", x = 3.8, y = -0.4, label = "Control\npanel", size = 2.5 * diagram_text_scale, color = "white") +
  ## Pedestal beam callout
  annotate("segment", x = 2.0, xend = 2.0, y = 0.95, yend = 0.55,
           arrow = arrow(length = unit(0.15, "cm"), type = "closed"), color = "#9B59B6", linewidth = 0.9) +
  annotate("text", x = 2.55, y = 0.65, label = "Pedestal beam\n(bridges floor trench)",
           size = 2.4 * diagram_text_scale, color = "#9B59B6") +
  coord_fixed() +
  xlim(-1.6, 4.6) + ylim(-1.2, 2.9) +
  theme_void() +
  labs(title = "Pick-and-Place Cell Layout",
       subtitle = "720 brackets/hour, 5.0 s cycle time -- Tier-2 automotive supplier, Cambridge, Ontario") +
  theme(plot.title = element_text(size = 14 * diagram_text_scale, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 10 * diagram_text_scale, hjust = 0.5))
```

The cell's throughput target — 720 brackets/hour, a 5.0 s cycle — is the single number every subsystem in this chapter has to agree with. It sets the fill-time requirement for the gripper's lift cylinder (Section 2), the minimum belt speed for the infeed conveyor (Section 3), and, indirectly, the duty expected of every branch circuit feeding the cell (Section 4). Seven things have to be sized before this cell can be quoted and built:

1. The gripper's pneumatic lift cylinder and its control valve, sized for the bracket's weight and the cycle's fill-time budget.
2. The infeed conveyor's motor, drive, and gearbox, sized for the belt pull this throughput demands.
3. The control panel's branch-circuit protection and conductors for the robot, the conveyor motor, and the gripper's valve solenoids, plus the panel's overall SCCR rating.
4. The cell's bonding and grounding scheme, including the equipment bonding conductor for the conveyor motor branch and shield grounding for any analog instrumentation.
5. The minimum safe distance from the operator-side light curtain to the robot's reach envelope, driven by the safety circuit's actual stopping time.
6. The structural adequacy of the pedestal beam that carries the robot's own weight across a floor trench used for cable and hose routing.

The rest of this chapter works through each of these in turn, using the exact methods developed in the nine chapters cited above — not re-derived from scratch, but reused directly, the way a real project reuses a company's own established design methods from one machine to the next.

## 2. Cylinder and Valve Sizing

The gripper's pneumatic lift stroke does two things each cycle: it lowers the gripper the last few centimeters onto the bracket before the jaws close, then lifts the gripper and bracket clear of the conveyor so the robot can swing away without dragging the part across neighboring brackets still queued on the belt. The **extend** (lower) stroke is a light, mostly-friction load — the process engineer has specified **110 N** is enough. The **retract** (lift) stroke has to accelerate the gripper (2.1 kg) and the bracket (1.8 kg) — a combined 3.9 kg moving mass — upward quickly enough to keep the 5.0 s cycle on schedule, and the process engineer has specified the retract stroke must deliver at least **300 N** at the rod. The plant runs shop air at the same standard **6 bar** gauge used throughout this book's pneumatic examples.

Following the [Pneumatic System Design](06-07-Pneumatic-System-Design.qmd) chapter's method exactly: the retract stroke governs, because it's the rod-side (annulus) area that's available on that stroke, not the full bore area. Size the bore from $A_{min} = F_{required}/P$, then apply the 25-50% oversize rule of thumb against the standard ISO 15552 bore/rod combinations.

```{r bore-sizing-06-15, echo=TRUE}
## Given
F_lift_required  <- 300   # N, required retract (lift) force
F_lower_required <- 110   # N, required extend (lower) force
P_bar <- 6                 # bar gauge, shop line pressure
P_pa  <- P_bar * 1e5

## Minimum annulus area needed on the governing (retract) stroke
A_annulus_min_mm2 <- F_lift_required / P_pa * 1e6
cat(sprintf("Minimum retract-side (annulus) area required: %.1f mm^2\n", A_annulus_min_mm2))
cat(sprintf("25%% oversize target: >= %.1f mm^2\n", A_annulus_min_mm2 * 1.25))
cat(sprintf("50%% oversize target: >= %.1f mm^2\n\n", A_annulus_min_mm2 * 1.50))

## Check standard ISO 15552 bore/rod combinations
bore_candidates <- c(32, 40, 50, 63)
rod_candidates  <- c(12, 16, 20, 20)

for (i in seq_along(bore_candidates)) {
  A_bore    <- pi / 4 * bore_candidates[i]^2
  A_rod     <- pi / 4 * rod_candidates[i]^2
  A_annulus <- A_bore - A_rod
  margin_pct <- (A_annulus / A_annulus_min_mm2 - 1) * 100
  cat(sprintf("Bore %2d mm / rod %2d mm: A_annulus = %6.1f mm^2, margin = %6.1f%%\n",
              bore_candidates[i], rod_candidates[i], A_annulus, margin_pct))
}

## Selected: 32 mm bore / 12 mm rod -- the smallest standard size that clears
## the 25% floor without blowing past the 50% ceiling
bore <- 32; rod <- 12
A_bore    <- pi / 4 * bore^2
A_rod     <- pi / 4 * rod^2
A_annulus <- A_bore - A_rod

F_extend_theor  <- P_pa * A_bore * 1e-6
F_retract_theor <- P_pa * A_annulus * 1e-6

## Real-world losses: small-bore cylinder, moderate tubing run out along the
## robot's own dress-pack to the wrist -- toward the low end of the
## efficiency range the Pneumatic System Design chapter's rule of thumb gives
eta_friction <- 0.85
eta_pressure <- 0.88
eta_total <- eta_friction * eta_pressure
F_retract_actual <- F_retract_theor * eta_total
F_extend_actual  <- F_extend_theor * eta_total

cat(sprintf("\nSelected: %d mm bore / %d mm rod\n", bore, rod))
cat(sprintf("Theoretical retract force: %.1f N (required %.0f N)\n", F_retract_theor, F_lift_required))
cat(sprintf("Combined efficiency factor: %.2f x %.2f = %.3f\n", eta_friction, eta_pressure, eta_total))
cat(sprintf("Actual retract force: %.1f N (required %.0f N, margin %.1f%%)\n",
            F_retract_actual, F_lift_required, (F_retract_actual / F_lift_required - 1) * 100))
cat(sprintf("Actual extend force:  %.1f N (required %.0f N, margin %.1f%%)\n",
            F_extend_actual, F_lower_required, (F_extend_actual / F_lower_required - 1) * 100))
```

The 32 mm/12 mm cylinder is the right selection here — a genuinely smaller bore than the 50 mm cylinder the Pneumatic System Design chapter's own worked example needed, which makes sense: this gripper lift stroke handles a lighter combined mass over a much shorter travel than that chapter's full lift-and-clear station cylinder. But notice what the efficiency check does to the margin: the theoretical 38.2% oversize on the retract stroke shrinks to barely 3.4% once realistic friction and pressure-drop losses are applied. That's a tight pass, not a comfortable one, and it's exactly the trap the Pneumatic System Design chapter's Design Task 2 warns about — a bore that looks generously sized on a P × A calculation can turn out to have almost no real margin once seal friction and dynamic pressure drop are subtracted. If this cell's duty cycle ever tightens further, the 40 mm bore is the first place to look, not a re-check of the same 32 mm selection with more optimistic efficiency numbers.

With the cylinder fixed, the directional valve has to fill that cylinder's extend chamber fast enough to keep the 5.0 s cycle on schedule. The process engineer has budgeted **0.12 s** for the extend stroke to complete, over a short **60 mm** stroke length — a burst flow requirement, not the cycle-averaged consumption a naive sizing approach might reach for first.

Live cell — try a different fill time, stroke length, or valve pressure drop and see how the required Cv responds.

```{webr-r}
#| label: valve-cv-06-15
## Given
bore <- 32
A_bore <- pi / 4 * bore^2
stroke_mm <- 60
P_bar <- 6; P_atm_bar <- 1.013
ratio <- (P_bar + P_atm_bar) / P_atm_bar

t_fill_s <- 0.12    # required extend-stroke fill time

V_free_L <- A_bore * stroke_mm / 1e6 * ratio
V_free_ft3 <- V_free_L / 28.317
Q_burst_SCFM <- V_free_ft3 / (t_fill_s / 60)

## Valve sizing conditions
P1_psig <- 87              # supply pressure, psig (6 bar)
P1_psia <- P1_psig + 14.7
dP_psi <- 15                # assumed valve pressure drop at this flow
P2_psia <- P1_psia - dP_psi
T_R <- 530                  # 70 deg F, absolute
SG <- 1.0                   # air

Cv_required <- Q_burst_SCFM / 22.48 * sqrt(SG * T_R / (dP_psi * P2_psia))

cat(sprintf("Free-air volume to fill (extend stroke): %.4f L = %.5f ft^3\n", V_free_L, V_free_ft3))
cat(sprintf("Required burst flow for %.2f s fill: %.2f SCFM\n\n", t_fill_s, Q_burst_SCFM))
cat(sprintf("P1 = %.1f psia, P2 = %.1f psia, dP = %.1f psi\n", P1_psia, P2_psia, dP_psi))
cat(sprintf("Required Cv: %.3f\n", Cv_required))
cat(sprintf("-> Select a catalog valve rated Cv >= %.2f (round up for margin)\n", ceiling(Cv_required * 10) / 10))
```

**Sanity check:** a required Cv of about 0.17 is small even by small-cylinder standards — well under the 0.55 Cv the Pneumatic System Design chapter's own worked example needed for its larger 50 mm cylinder — which makes physical sense given both a smaller bore and a shorter stroke here. Essentially any compact wrist-mounted or gripper-integrated solenoid valve in a major manufacturer's catalog clears this requirement with room to spare; the real constraint on this valve's selection is more likely to be its physical size and mass (it rides on the robot's wrist, adding to the payload the robot itself has to carry) than its flow capacity.

## 3. Infeed Conveyor Motor and Drive Sizing

The infeed conveyor is short and level — no incline, unlike the case-pick mezzanine conveyor in the [Conveyor Design](06-10-Conveyor-Design.qmd) chapter — but the same CEMA-style effective-tension method applies directly, it just has a zero gravity term. The conveyor spans **2.5 m**, carrying brackets (1.8 kg average) queued at roughly **0.2 m** center-to-center spacing near the robot's pick point, on a belt-and-idler system with an equivalent moving mass of **10 kg per linear meter** (a light-duty conveyor, appropriate for a small-parts infeed rather than the heavier tote conveyor from that chapter).

```{webr-r}
#| label: conveyor-drive-06-15
## Given: level infeed conveyor, effective-tension method (Conveyor Design
## chapter, Design Task 1), zero-incline special case
theta <- 0                # rad, level conveyor
L <- 2.5                   # m, belt-path length
g <- 9.81

w_b <- 10                  # kg/m, belt + idlers
W_belt <- w_b * L * g       # N

part_mass <- 1.8            # kg, average bracket
s_spacing <- 0.2            # m, center-to-center spacing at the queued pick point
q <- part_mass / s_spacing  # kg/m, linear load density
W_load <- q * L * g          # N

C_b <- 0.030   # belt/idler friction factor
C_m <- 0.035   # load friction factor

T_x <- C_b * W_belt * cos(theta)
T_y <- C_m * W_load * cos(theta)
T_z <- W_load * sin(theta)   # zero on a level conveyor
T_e <- T_x + T_y + T_z

contingency <- 0.10
T_e_design <- T_e * (1 + contingency)

cat(sprintf("Belt + idler weight (W_belt): %.1f N\n", W_belt))
cat(sprintf("Load on belt at design queuing (W_load): %.1f N\n\n", W_load))
cat(sprintf("T_x (belt/idler friction): %.2f N\n", T_x))
cat(sprintf("T_y (load friction):       %.2f N\n", T_y))
cat(sprintf("T_z (gravity/elevation):   %.2f N (zero -- level conveyor)\n", T_z))
cat(sprintf("T_e (effective tension):   %.2f N\n", T_e))
cat(sprintf("T_e,design (+%.0f%% contingency): %.2f N\n\n", contingency * 100, T_e_design))

## Minimum belt speed from throughput: 720 parts/hour, minimum 0.35 m
## spacing for reliable vision-guided picking
rate_per_hr <- 720
rate_per_s <- rate_per_hr / 3600
s_min <- 0.35
v_min <- rate_per_s * s_min
v_design <- v_min * 1.25   # 25% surge margin, per the Conveyor Design chapter's Design Task 3

cat(sprintf("Minimum belt speed for 720 parts/hr at 0.35 m spacing: %.4f m/s\n", v_min))
cat(sprintf("Design belt speed (+25%% surge margin): %.4f m/s -> select 0.09 m/s\n\n", v_design))

v_belt <- 0.09     # m/s, selected design speed
eta_drive <- 0.85   # combined gearbox + coupling efficiency

P_mech <- T_e_design * v_belt
P_motor <- P_mech / eta_drive
cat(sprintf("Mechanical power at the belt: %.2f W\n", P_mech))
cat(sprintf("Required motor power (%.0f%% drive efficiency): %.2f W\n\n", eta_drive * 100, P_motor))

D_pulley <- 0.08    # m, 80 mm head pulley -- small standard size for a light-duty conveyor
r_pulley <- D_pulley / 2
T_pulley <- T_e_design * r_pulley
omega_pulley <- v_belt / r_pulley
rpm_pulley <- omega_pulley * 60 / (2 * pi)

cat(sprintf("Required torque at pulley shaft: %.3f N.m\n", T_pulley))
cat(sprintf("Pulley shaft speed: %.2f rpm\n\n", rpm_pulley))

## Reflect to a standard 4-pole motor per the Motor, Drive, and Gearbox
## Sizing chapter's method: omega_motor = N * omega_load, T_motor = T_load / (N * eta_gb)
motor_rpm <- 1750
N_ratio <- motor_rpm / rpm_pulley
eta_gb <- 0.75   # worm gearbox, appropriate for this large a ratio
T_motor <- T_pulley / (N_ratio * eta_gb)

cat(sprintf("Gear ratio for a %d rpm 4-pole motor: %.1f:1 (worm gearbox)\n", motor_rpm, N_ratio))
cat(sprintf("Reflected motor-shaft torque: %.5f N.m\n", T_motor))
```

**Sanity check:** the calculated motor power (under 2 W) and reflected torque (a few thousandths of a N·m) are, on paper, almost nothing — and that's a genuine, useful finding, not an error. At this scale, the governing constraint on motor selection isn't the calculated running load at all; it's the smallest standard gearmotor frame a manufacturer actually catalogs, combined with the [Conveyor Design](06-10-Conveyor-Design.qmd) chapter's own rule of thumb of sizing starting torque at roughly 150% of running torque to cover breakaway friction. In practice, this conveyor gets specified with the smallest available fractional-horsepower gearmotor frame in the supplier's line — a real design outcome, not a rounding artifact, and worth flagging explicitly: on a small enough machine, "big enough to exist as a catalog part" replaces "big enough for the calculated load" as the actual sizing constraint. The large gear ratio (roughly 80:1) driving toward a worm gearbox is consistent with that: slow, light conveyors like this one routinely end up on worm-reduction gearmotors precisely because the load itself demands so little.

That leaves one open question the [Motor, Drive, and Gearbox Sizing](06-06-Motor-Drive-Gearbox-Sizing.qmd) chapter's Design Task 5 framework answers directly: servo or induction motor + VFD? This conveyor runs continuously at a fixed speed feeding parts to a robot that does its own vision-guided picking — it never needs to index to a stop-and-settle position the way the index table in that chapter's running example did.

Live cell — try different weights and see whether the recommendation flips.

```{webr-r}
#| label: servo-vfd-decision-06-15
## Weighted decision matrix, weights reflecting this conveyor's actual job:
## running at a controlled speed, not point-to-point positioning
weights <- c(accuracy = 0.10, dynamic_response = 0.10, cost = 0.40, complexity = 0.40)
servo_scores <- c(accuracy = 5, dynamic_response = 5, cost = 2, complexity = 2)
induction_scores <- c(accuracy = 2, dynamic_response = 2, cost = 5, complexity = 5)

servo_total <- sum(weights * servo_scores)
induction_total <- sum(weights * induction_scores)

cat("Weighted decision matrix (1 = poor fit, 5 = excellent fit):\n\n")
for (nm in names(weights)) cat(sprintf("  %-18s weight = %.0f%%\n", nm, weights[nm] * 100))
cat(sprintf("\nServo weighted score:          %.2f\n", servo_total))
cat(sprintf("Induction + VFD weighted score: %.2f\n", induction_total))
cat(sprintf("\n-> Recommendation: %s\n", ifelse(servo_total > induction_total, "SERVO", "INDUCTION MOTOR + VFD")))
```

**Sanity check:** weighting cost and complexity heavily (this conveyor never positions anything, it just runs) flips the recommendation decisively toward induction + VFD, the mirror image of the index-table result in the Motor, Drive, and Gearbox Sizing chapter's own worked example. A general-purpose VFD, sized to the selected gearmotor's nameplate FLA rather than the tiny calculated running load, is the right drive for this branch — and that nameplate FLA is exactly the number Section 4 needs next for branch-circuit protection.

## 4. Panel, Protection, and Wiring

The panel has to protect three genuinely different loads: the conveyor gearmotor sized in Section 3, the six-axis robot's own controller (which carries its servo drives internally and is supplied as a single branch per the manufacturer's documented supply-circuit data — consistent with how the CEC treats industrial machinery under Section 42, and the NEC under Article 670, rather than recalculating it with the single-motor formula in CEC Section 28/NEC Article 430), and the gripper valve's 24VDC control power supply.

```{webr-r}
#| label: branch-protection-sccr-06-15
## --- Conveyor motor branch (a genuine single motor -- Control Panel
## Design chapter's motor branch circuit method applies directly) ---
FLA <- 0.8   # A, nameplate FLA for the small gearmotor selected in Section 3
conductor_ampacity_required <- 1.25 * FLA

service_factor <- 1.15
m_OL <- ifelse(service_factor >= 1.15, 1.25, 1.15)
I_OL <- m_OL * FLA

m_fuse <- 1.75
I_OCPD_calc <- m_fuse * FLA
standard_fuse_sizes <- c(1, 1.25, 1.6, 2, 2.5, 3, 3.5, 4, 5, 6, 8, 10, 15, 20)
I_OCPD_conveyor <- min(standard_fuse_sizes[standard_fuse_sizes >= I_OCPD_calc])

cat("--- Conveyor motor branch ---\n")
cat(sprintf("FLA = %.1f A -> required conductor ampacity (1.25x) = %.2f A\n", FLA, conductor_ampacity_required))
cat(sprintf("Overload relay setting: %.2f A\n", I_OL))
cat(sprintf("Time-delay fuse: %.1f A (calc %.2f A, next standard size up)\n\n", I_OCPD_conveyor, I_OCPD_calc))

## --- Robot controller branch (manufacturer-specified supply data,
## general continuous-load conductor sizing applies) ---
I_robot <- 3.4    # A, manufacturer-published nameplate current, 600 V three-phase
I_robot_required <- 1.25 * I_robot
I_robot_breaker <- 15   # A, manufacturer-recommended branch breaker rating

cat("--- Robot controller branch (manufacturer-specified) ---\n")
cat(sprintf("Nameplate current = %.1f A -> required conductor ampacity (1.25x) = %.2f A\n", I_robot, I_robot_required))
cat(sprintf("Branch breaker: %.0f A (manufacturer-specified, per CEC Section 42 / NEC Article 670)\n\n", I_robot_breaker))

## --- Gripper valve 24VDC control power supply branch ---
I_ps_primary <- 1.2   # A, 120V primary current for a small 24VDC/5A supply
I_ps_breaker <- 2      # A

cat("--- 24VDC control power supply branch ---\n")
cat(sprintf("Primary current ~= %.1f A -> breaker: %.0f A\n\n", I_ps_primary, I_ps_breaker))

## --- Panel SCCR: weakest-link check ---
required_sccr <- 25   # kA, this Tier-2 supplier's plant electrical specification

components <- data.frame(
  component = c("Main incoming breaker", "Conveyor motor contactor",
                "Robot controller disconnect", "24VDC power supply"),
  standalone_kA = c(25, 5, 10, 5),
  combo_kA = c(NA, 25, 25, 25)
)
components$effective_kA <- ifelse(is.na(components$combo_kA), components$standalone_kA, components$combo_kA)
panel_sccr <- min(components$effective_kA)
panel_sccr_standalone_only <- min(components$standalone_kA)

cat("--- Panel SCCR (weakest-link check) ---\n")
for (i in seq_len(nrow(components))) {
  cat(sprintf("%-30s standalone %2.0f kA -> effective %2.0f kA\n",
              components$component[i], components$standalone_kA[i], components$effective_kA[i]))
}
cat(sprintf("\nPanel SCCR without current-limiting combinations: %.0f kA -- %s\n",
            panel_sccr_standalone_only, ifelse(panel_sccr_standalone_only >= required_sccr, "PASS", "FAILS")))
cat(sprintf("Panel SCCR with documented CL-fuse combinations applied: %.0f kA -- %s\n",
            panel_sccr, ifelse(panel_sccr >= required_sccr, "PASS", "FAILS")))
```

**Sanity check:** the conveyor motor's own FLA is so small that its branch protection is set by the smallest available standard fuse size, not by a percentage-of-FLA calculation — the same "governed by the smallest available standard part, not the calculated load" pattern Section 3 already found for the gearmotor itself. The panel's real electrical mass is the robot controller branch, and its real SCCR risk is the weakest link among the three low-rated branch devices — exactly the trap the Control Panel Design chapter's weakest-link principle warns about: a 25 kA main breaker feeding a 5 kA contactor does not produce a 25 kA panel without a documented current-limiting fuse combination in every branch that needs one.

With the branches sized, one conductor run gets a voltage-drop check: the 24VDC feed to the gripper's solenoid valve, which travels out through the robot's own trailing cable (dress-pack) and back to the panel — a genuinely long run for a low-voltage circuit, **45 m** one-way.

Live cell — try a different cable length, gauge, or solenoid current and see how the delivered voltage responds.

```{webr-r}
#| label: voltage-drop-06-15
## 24VDC gripper solenoid valve, fed through the robot's dress-pack and back
## to the panel
V_supply <- 24        # VDC nominal
I_load <- 0.4          # A, single solenoid coil energized at a time
L_one_way_m <- 45      # m, one-way run through the dress-pack

R_18awg <- 25.7   # ohm/km, representative (Wire, Cable, and Circuit Protection chapter)
R_14awg <- 10.2   # ohm/km

calc_drop <- function(R_ohm_km, L_m, I) {
  L_km <- L_m / 1000
  2 * I * L_km * R_ohm_km   # round-trip resistive drop
}

min_operating_V <- 20.4   # V, representative -15% tolerance floor for a 24VDC solenoid

for (nm in c("R_18awg", "R_14awg")) {
  R <- get(nm)
  Vd <- calc_drop(R, L_one_way_m, I_load)
  pct <- Vd / V_supply * 100
  delivered <- V_supply - Vd
  cat(sprintf("%s (%.2f ohm/km): drop = %.3f V (%.2f%%), delivered = %.2f V -> %s (3%% rule), %s (functional floor)\n",
              nm, R, Vd, pct, delivered,
              ifelse(pct <= 3, "PASS", "FAIL"),
              ifelse(delivered >= min_operating_V, "PASS", "FAIL")))
}
```

**Sanity check:** 18 AWG, the common default for this kind of dress-pack wiring, fails the 3% rule of thumb (about 3.9%) over this unusually long 45 m run, but the delivered voltage (23.1 V) still clears the solenoid's functional floor comfortably — the same "too loose vs. too tight" distinction the Wire, Cable, and Circuit Protection chapter's own worked example makes. For this on/off solenoid, 18 AWG is functionally acceptable; stepping up to 14 AWG (1.5% drop) is the safer specification if this cable is ever repurposed to carry an analog signal alongside the solenoid feed, since an analog device's tolerance for drop is much tighter than a solenoid's.

## 5. Bonding and Grounding Scheme

Three separate bonding and grounding questions apply to this cell, each using the terminology and method from the [Bonding, Grounding, and EMC](06-05-Bonding-Grounding-EMC.qmd) chapter.

Live cell — try different OCPD ratings, line voltage, or ground potential difference and see how each result responds.

```{webr-r}
#| label: bonding-grounding-06-15
## --- (a) Equipment bonding conductor sizing: sized from the upstream
## OCPD rating, not the connected load's FLA ---
ocpd_breakpoints <- c(20, 30, 60, 100, 200, 400, 600, 800, 1000, 1200)
bond_awg <- c("12 AWG", "10 AWG", "10 AWG", "8 AWG", "6 AWG",
              "3 AWG", "1 AWG", "1/0 AWG", "2/0 AWG", "3/0 AWG")
lookup_bonding_size <- function(ocpd_rating) {
  idx <- min(which(ocpd_breakpoints >= ocpd_rating))
  bond_awg[idx]
}

ocpd_conveyor <- 1.6   # A, from Section 4's selected time-delay fuse
ocpd_robot <- 15        # A, robot controller branch breaker

cat("--- (a) Equipment bonding conductor sizing ---\n")
cat(sprintf("Conveyor motor branch (%.1f A OCPD) -> bonding conductor: %s\n",
            ocpd_conveyor, lookup_bonding_size(ocpd_conveyor)))
cat(sprintf("Robot controller branch (%.0f A OCPD) -> bonding conductor: %s\n",
            ocpd_robot, lookup_bonding_size(ocpd_robot)))
cat("Both branches land on the table's smallest listed size -- a genuinely\n")
cat("small cell still needs a real bonding conductor, never skipped just\n")
cat("because the connected load is light.\n\n")

## --- (b) VFD/motor bonding: bearing voltage estimate for the
## VFD-driven conveyor motor ---
V_line_rms <- 600                 # V, three-phase line-to-line
V_dc_bus <- 1.35 * V_line_rms      # V, typical diode-bridge DC bus voltage
V_cm_peak <- V_dc_bus / 2           # representative peak common-mode voltage
BVR <- 0.04                         # unmitigated bearing voltage ratio, representative
V_bearing <- BVR * V_cm_peak
V_film_breakdown <- 10              # V, representative order-of-magnitude film breakdown

cat("--- (b) VFD/motor bonding: bearing voltage check ---\n")
cat(sprintf("DC bus voltage: %.0f V -> peak common-mode voltage: %.0f V\n", V_dc_bus, V_cm_peak))
cat(sprintf("Estimated bearing voltage (BVR = %.0f%%): %.1f V\n", BVR * 100, V_bearing))
cat(sprintf("-> %s film breakdown (~%.0f V): use symmetrical shielded VFD cable with a\n",
            ifelse(V_bearing > V_film_breakdown, "EXCEEDS", "stays below"), V_film_breakdown))
cat("   360-degree bond at both drive and motor ends, even on this small conveyor motor --\n")
cat("   the risk comes from the drive's switching voltage, not the motor's frame size.\n\n")

## --- (c) Shield grounding: analog pressure transducer on the FRL,
## one-end rule at signal frequency ---
cable_length_m <- 12         # m, transducer to PLC analog input
shield_R_per_m <- 0.030       # ohm/m
shield_R_total <- cable_length_m * shield_R_per_m
delta_V_ground <- 0.9         # V, transient during the conveyor VFD's motor start

I_loop_both_ends <- delta_V_ground / shield_R_total
I_loop_one_end <- 0

cat("--- (c) Shield grounding: 4-20 mA line-pressure transducer ---\n")
cat(sprintf("Shield resistance over %.0f m: %.3f ohm\n", cable_length_m, shield_R_total))
cat(sprintf("Shield grounded at BOTH ends: %.0f mA circulating (a mis-wired installation)\n",
            I_loop_both_ends * 1000))
cat(sprintf("Shield grounded at ONE end only: %.0f mA (correct low-frequency practice)\n",
            I_loop_one_end * 1000))
```

**Sanity check:** all three results connect directly back to earlier sections. The bonding conductors in (a) are sized off the OCPD ratings Section 4 selected, not off the FLA values Section 3 calculated — a direct application of the Bonding, Grounding, and EMC chapter's central point that bonding conductor size tracks the breaker, not the load. The bearing-voltage estimate in (b) comes out identical in structure to that chapter's own worked example, because the physics depends on the drive's DC bus voltage, not the motor's horsepower — a small VFD-fed motor is exposed to essentially the same bearing-current risk as a large one, which is why the mitigation (shielded cable, 360° bonding) applies here even though this conveyor motor is a fraction of a horsepower. And (c) shows why a properly-terminated one-end shield ground matters even on a short 12 m run: the shield's own resistance is so low that even a modest 0.9 V transient during the conveyor's VFD motor start would drive 2.5 A of circulating current through a mistakenly both-ends-grounded shield — current that would couple straight into the 4-20 mA signal it's supposed to protect.

## 6. Minimum Safe Distance from Stopping Time

The cell's only open guarding side is the operator loading point at the infeed conveyor's tail end, protected by a safety light curtain (see the plan view in Section 1). Following the [Machine Safety Circuit Design](06-02-Machine-Safety-Circuits.qmd) chapter's method, the overall stopping time $T$ is the sum of four sequential components — and this cell's small robot, with meaningfully less inertia than the door-panel welding cell used as that chapter's own running example, coasts to a stop noticeably faster.

```{r fig-06-15-stopping-time, echo=FALSE, fig.cap="Overall stopping time T for the pick-and-place cell's safety circuit. This small robot's lower inertia gives a faster coast-down than the door-panel welding cell example, but coast-down still dominates the total.", fig.align='center', fig.width=9, fig.height=4.5}
t_components <- data.frame(
  component = c("Device response\n(T_device)", "Logic/relay processing\n(T_logic)",
                "Final control element\n(T_fce)", "Mechanical coast-down\n(T_coast)"),
  seconds = c(0.025, 0.015, 0.040, 0.450)
)
t_components$component <- factor(t_components$component,
                                  levels = t_components$component[order(t_components$seconds)])
t_components$label <- paste0(t_components$seconds * 1000, " ms")

ggplot(t_components, aes(x = component, y = seconds, fill = component)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = label), hjust = -0.15, size = 3.2 * diagram_text_scale, color = "black") +
  scale_fill_manual(values = c("Device response\n(T_device)" = "#AED6F1",
                                "Logic/relay processing\n(T_logic)" = "#F9E79F",
                                "Final control element\n(T_fce)" = "#F5B7B1",
                                "Mechanical coast-down\n(T_coast)" = "#A9DFBF")) +
  scale_y_continuous(limits = c(0, 0.55), expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Overall Stopping Time T: Pick-and-Place Cell",
       subtitle = "Total T = 0.025 + 0.015 + 0.040 + 0.450 = 0.53 s",
       x = NULL, y = "Time (s)", fill = NULL) +
  coord_flip() +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 13 * diagram_text_scale, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10.5 * diagram_text_scale, hjust = 0.5),
    axis.title = element_text(size = 11 * diagram_text_scale),
    axis.text = element_text(size = 10 * diagram_text_scale),
    legend.position = "none"
  )
```

Live cell — try changing any of the four time components, the intrusion allowance, or the planned mounting distance.

```{webr-r}
#| label: stopping-distance-06-15
## Overall stopping time, per the Machine Safety Circuit Design chapter's
## four-component breakdown, sized for this cell's own (smaller, lighter)
## robot and Category 1 stop
T_device <- 0.025   # s, Type 4 light curtain response time
T_logic  <- 0.015   # s, Category 3 safety relay reaction time
T_fce    <- 0.040    # s, contactor coil de-energize + contact opening
T_coast  <- 0.450    # s, mechanical coast-down for this smaller robot

T_total <- T_device + T_logic + T_fce + T_coast
cat(sprintf("Overall stopping time T = %.3f + %.3f + %.3f + %.3f = %.3f s\n\n",
            T_device, T_logic, T_fce, T_coast, T_total))

## ISO 13855 minimum safe distance, per the Robot Cell Design chapter's
## S = K x T + C
K <- 1600   # mm/s, ISO 13855 approach speed constant
C <- 750     # mm, intrusion-depth allowance for the selected light curtain resolution

S_required <- K * T_total + C
cat(sprintf("K = %d mm/s, C = %d mm\n", K, C))
cat(sprintf("Required minimum separation, S = K x T + C = %.0f mm\n\n", S_required))

planned_distance <- 1500   # mm, distance the cell layout currently plans

if (planned_distance >= S_required) {
  cat(sprintf("Planned mounting distance (%d mm) >= S: PASS\n", planned_distance))
} else {
  shortfall <- S_required - planned_distance
  cat(sprintf("Planned mounting distance (%d mm) < S: FAILS by %.0f mm\n\n", planned_distance, shortfall))

  C_fine <- 600   # mm, finer-resolution light curtain
  S_fine <- K * T_total + C_fine
  cat(sprintf("Option -- use a finer-resolution light curtain (C = %d mm): new S = %.0f mm\n",
              C_fine, S_fine))
  if (planned_distance >= S_fine) cat("  Planned distance now passes without moving the curtain.\n")
}
```

**Sanity check:** at 0.53 s, this cell's overall stopping time is meaningfully faster than the 0.65 s the Machine Safety Circuit Design chapter's own door-panel welding cell needed — consistent with a smaller, lower-inertia robot, and coast-down is still the dominant term at roughly 85% of the total, so the same "attack $T_{coast}$ first" rule of thumb applies here too. The resulting 1.60 m minimum separation sits right at the low end of the 1.5-3 m band the Robot Cell Design chapter's own rule of thumb predicts for a sub-second stopping time — a useful check that nothing was mistyped, and a direct illustration of why the faster stop matters: it buys back just enough distance that a tighter, more realistic floor-space allocation (1.5 m planned) comes up 98 mm short rather than badly short. A finer-resolution light curtain closes that gap without moving the curtain at all, exactly the kind of trade the Robot Cell Design chapter's guidance recommends reaching for before arguing down $K$.

## 7. Frame Deflection Check

The robot's pedestal sits on a fabricated steel beam bridging a **900 mm** floor trench used to route the cell's cable and hose runs beneath the robot rather than across the walkway beside it — a common layout choice that turns the pedestal's own support into a genuine beam-deflection problem rather than a simple slab-mounted footing. The robot, complete with its base and wrist tooling, has a combined mass of **185 kg**; acceleration reactions transmitted to the base during fast moves are represented with a modest dynamic load factor, following the same logic (if a smaller magnitude, appropriate to a pedestal reaction rather than a travelling carriage) as the [Frame and Structural Design](06-11-Frame-Structural-Design.qmd) chapter's own dynamic-load guidance.

```{webr-r}
#| label: beam-deflection-06-15
## Given: robot pedestal beam, bridging a 900 mm floor trench
L <- 900             # mm, span between the trench's edge supports
m_robot <- 185        # kg, robot + base + wrist tooling
g <- 9.81
dyn_factor <- 1.2     # dynamic load factor for acceleration reactions at the base

W_robot <- m_robot * g
P <- W_robot * dyn_factor

## Selected section: steel HSS 102 x 102 x 6.4 mm (CSA G40.20/G40.21 350W)
## -- representative CISC/manufacturer catalog section properties
E_steel <- 200000   # N/mm^2
I_beam <- 2.85e6     # mm^4, second moment of area
S_beam <- 55900       # mm^3, section modulus
Fy <- 350              # MPa, yield strength (350W grade)

delta_max <- P * L^3 / (48 * E_steel * I_beam)
M_max <- P * L / 4
sigma_max <- M_max / S_beam
L240 <- L / 240

cat(sprintf("Robot static weight: %.1f N\n", W_robot))
cat(sprintf("Design load (with %.1fx dynamic factor): %.1f N\n\n", dyn_factor, P))
cat(sprintf("Maximum deflection: %.4f mm\n", delta_max))
cat(sprintf("L/240 reference limit: %.2f mm -> %s\n\n",
            L240, ifelse(delta_max <= L240, "PASSES (large margin)", "FAILS")))
cat(sprintf("Maximum bending moment: %s N.mm\n", format(round(M_max), big.mark = " ")))
cat(sprintf("Maximum bending stress: %.2f MPa (%.1f%% of %d MPa yield)\n",
            sigma_max, sigma_max / Fy * 100, Fy))
```

**Sanity check:** the calculated deflection (about 0.058 mm) clears the general L/240 reference (3.75 mm) with roughly a 65-fold margin, and the bending stress sits at under 3% of yield — the same "stiffness passes easily, strength isn't even close to the limiting factor" signature the Frame and Structural Design chapter's own gantry-beam example found. If this cell's pick-and-place task ever tightened into a genuinely vision-critical placement tolerance rather than a fixture-tray drop, the next step would be to check that 0.058 mm figure against a real positioning error budget the way that chapter's Design Task 2 does — not against L/240, which was never meant to answer a micron-level question — but for a fixture-tray placement with generous mechanical lead-in, L/240 is the right bar here, and this beam clears it comfortably.

## How the Nine Chapters Fit Together

Look back at what just happened across these seven sections: a cycle-time number from the system overview drove both the valve's fill-time requirement and the conveyor's minimum belt speed; the conveyor's selected gearmotor produced the FLA that sized its branch breaker; that breaker's rating sized its equipment bonding conductor; and the robot's own stopping time — a controls-side number — set a mechanical guarding distance that has nothing to do with amperes or bore diameters at all. None of that connectivity is visible when each calculation is practiced in isolation, one chapter at a time. It only shows up when the same cell has to satisfy all of them at once, the way a real integration project does. That's the actual skill this capstone is teaching: not a new formula, but the discipline of carrying a number honestly from the section that calculates it to the section that needs it next, and noticing when two sections' numbers don't agree before a fabricator or an electrician finds the mismatch on the shop floor instead.

## Summary

```{r tbl-06-15-summary, echo=FALSE}
summary_df <- data.frame(
  `Subsystem` = c(
    "1. System overview",
    "2. Cylinder and valve sizing",
    "3. Conveyor motor and drive",
    "4. Panel, protection, and wiring",
    "5. Bonding and grounding",
    "6. Minimum safe distance",
    "7. Frame deflection check"
  ),
  `Key Sizing Result` = c(
    "720 parts/hr, 5.0 s cycle -- drives every downstream sizing decision",
    "32 mm bore / 12 mm rod cylinder (3.4% actual margin); valve Cv >= 0.2",
    "0.09 m/s belt speed; smallest standard gearmotor frame (~80:1 worm gearbox); induction + VFD",
    "1.6 A fuse (conveyor), 15 A breaker (robot); panel SCCR = 25 kA with CL-fuse combinations",
    "12 AWG bonding conductors (both branches); bearing voltage at risk without shielded VFD cable; one-end shield ground",
    "T = 0.53 s; S = 1.60 m (vs. 1.50 m planned -- needs a finer-resolution curtain or relocation)",
    "Pedestal beam deflection 0.058 mm vs. L/240 = 3.75 mm -- passes with large margin"
  ),
  `Method Source` = c(
    "Section 1 (this chapter)",
    "Pneumatic System Design",
    "Conveyor Design; Motor, Drive, and Gearbox Sizing",
    "Control Panel Design; Wire, Cable, and Circuit Protection",
    "Bonding, Grounding, and EMC",
    "Machine Safety Circuit Design; Robot Cell Design",
    "Frame and Structural Design"
  ),
  check.names = FALSE
)

kable(summary_df, caption = "Summary: capstone pick-and-place cell design results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = knitr::is_html_output()) %>%
  row_spec(0, bold = TRUE, background = "#4a86c7", color = "white") %>%
  column_spec(1, width = "1in")
```

## Extension and Reflection Questions

<details>
<summary>What changes if the cycle rate doubles, from 720 to 1440 parts/hour?</summary>

Nearly every section in this chapter has to be re-checked, not just the one that seems most obviously affected. The valve's fill-time budget (Section 2) would need to shrink roughly in proportion to the shorter available cycle, driving the required Cv up — likely still inside the range of an off-the-shelf valve, since the starting Cv (0.17) had so much headroom, but it needs to be recalculated, not assumed. The conveyor's minimum belt speed (Section 3) doubles directly, per $v_{min} = R\,s_{min}$, which — as the Conveyor Design chapter's own practice problems point out — roughly doubles the required motor power too; at these small absolute power levels, the motor almost certainly stays on the same catalog frame, but the gear ratio and pulley speed both need re-checking. The robot's own duty cycle tightens, which is worth a second look at whether the assumed $T_{coast}$ in Section 6 still holds — a robot working a faster cycle may be commanded to decelerate harder, which could actually *shorten* $T_{coast}$, or a controller pushed nearer its dynamic limits could make it less repeatable, which cuts the other way. The one thing that almost certainly does *not* need to change is the bonding and grounding scheme in Section 5 — those calculations track OCPD ratings and drive physics, neither of which moves just because parts arrive faster.

</details>

<details>
<summary>What changes if this becomes a washdown food-processing installation instead of a dry automotive cell?</summary>

This touches more sections than it might first appear. The conveyor's construction (frame material, fasteners, motor enclosure rating) would need the Conveyor Design chapter's Design Task 5 zone-classification framework applied — direct product contact and a daily hose-down would likely push the frame to stainless steel and the conveyor motor to an IP69K-rated enclosure, which is a different physical motor than the one sized in Section 3 (same electrical sizing math, different nameplate). The bonding hardware in Section 5 would need the Bonding, Grounding, and EMC chapter's washdown-specific guidance — corrosion-resistant lugs and jumpers, since a corroding bonding connection in a wash-down zone is exactly the kind of failure that's invisible until the fault it exists to clear actually happens. The pedestal beam in Section 7 might warrant a stainless or coated section rather than painted mild steel, per the Frame and Structural Design chapter's material-selection framework, if it sits inside the daily wash pattern rather than outside it. The pneumatic cylinder in Section 2 would need washdown-rated seals and a stainless rod, a detail this chapter's calculation doesn't change but a real specification would. What does *not* change: the underlying force, torque, and deflection numbers themselves — a washdown environment changes what things are made of and how they're sealed, not the physics that sized them in the first place.

</details>

<details>
<summary>Identify every place this design would need to change if the robot were reclassified as collaborative.</summary>

This is the biggest single change a reclassification could trigger, because it touches the entire premise of Section 6. If the robot genuinely operates in a collaborative mode under ISO/TS 15066 power-and-force limiting (the framework introduced in the Robot Cell Design chapter's Design Task 6), the light curtain guarding the operator loading point could potentially be reduced or eliminated for that specific interaction — but only if the robot's actual contact forces, at the speeds and payloads used in this cell, are verified against the standard's biomechanical limits for the relevant body regions, which is a real engineering verification, not an assumption that "collaborative" automatically means "no guarding needed." If the guarding *is* relaxed, Section 6's entire stopping-time-and-separation-distance calculation may no longer be the controlling safety requirement for that access point at all, though the safety circuit itself (Section 6's $T$ derivation) may still be needed for other hazards in the cell — the pneumatic gripper's pinch points, for instance, don't become safe just because the robot's arm motion does. The risk assessment underlying the whole guarding decision (the Robot Cell Design chapter's Design Task 5 framework) would need to be redone from scratch for this specific cell's specific hazards, not inherited from the non-collaborative baseline this chapter used. Everything in Sections 2 through 5 and 7 — the cylinder, the conveyor, the panel, the bonding scheme, the pedestal beam — stays exactly as calculated, because none of it depends on how the robot's motion is safeguarded.

</details>

<details>
<summary>What if the plant's incoming service is the US 480 V convention instead of the Canadian 600 V class?</summary>

Every current-carrying calculation in Section 4 changes, even though none of the loads themselves change. Per the Wire, Cable, and Circuit Protection chapter's voltage-class comparison, the same delivered power at 480 V draws roughly 25% more current than at 600 V, which pushes the conveyor motor's FLA up, changes the calculated conductor ampacity and fuse selection, and increases the percent voltage drop on every run in the cell for the same physical conductor and length — the 24 VDC solenoid run in Section 4 is unaffected (it's a separate low-voltage circuit), but the robot controller's own three-phase feeder would need to be re-checked. The bonding conductor sizing in Section 5 might also shift, since it's driven by the (now larger) branch OCPD ratings rather than the FLA increase directly. None of this changes the mechanical sizing in Sections 2, 3, 6, or 7 at all — cylinder bore, belt pull, stopping time, and beam deflection are governed by pressure, force, and mass, not by which side of the border the electrical service comes from.

</details>

<details>
<summary>What if the fixture tray needs true vision-guided sub-millimeter placement instead of a generous drop-in fixture?</summary>

This tightens exactly the two sections in this chapter that were passed with the most room to spare, for reasons that have nothing to do with each other. Section 7's beam deflection (0.058 mm) would need to be checked against an actual placement tolerance budget the way the Frame and Structural Design chapter's Design Task 2 does, allocating some fraction of the total tolerance to pedestal deflection specifically — at 0.058 mm, this beam is already in the same order of magnitude as a typical sub-millimeter placement budget, which is a much closer call than the 65-fold L/240 margin suggests, and the *change* in that deflection as the robot's own moving mass shifts around the workspace (not just its static weight) would need to be considered too. Section 3's servo-vs-VFD decision might also flip: if "vision-guided placement" starts to mean the conveyor itself has to index to a precise, repeatable stop for a fixed-position pick rather than let the robot's vision system track a moving part, that's exactly the kind of point-to-point positioning task the Motor, Drive, and Gearbox Sizing chapter's own decision framework weights toward a servo, not an induction motor and VFD.

</details>

<details>
<summary>What if the cell has to be relocated to feed from a mezzanine one floor below, requiring an inclined infeed conveyor?</summary>

This re-opens Section 3's belt-pull calculation specifically, and only that one — but not trivially, since the level conveyor's zero gravity term ($T_z = 0$) was doing real work simplifying that section's result. Following the Conveyor Design chapter's Design Task 1 method exactly, a nonzero incline angle reintroduces $T_z = W_{load}\sin\theta$, which the worked example in that chapter shows can easily become the dominant term rather than a minor addition — at a modest 15-20 degree incline, gravity resistance would very likely swamp the friction terms that made up all of $T_e$ in this chapter's level-conveyor result. That, in turn, changes the required drive power and torque in the same section, which changes the motor's FLA, which cascades into Section 4's branch protection and Section 5's bonding conductor sizing exactly the way Section 3's original sizing did. It's also worth checking, per that chapter's own rule of thumb, whether the brackets themselves would slide on a plain belt above roughly 18-20 degrees of incline — if so, a friction-top belt surface becomes part of the specification, not just a steeper drive.

</details>

<details>
<summary>Reflection: which single number in this chapter, if it turned out to be wrong, would cause the most damage downstream?</summary>

There's a reasonable case for the overall stopping time $T$ in Section 6. It's the only number in this chapter that is explicitly described, in the Machine Safety Circuit Design chapter's own guidance, as a *design estimate* rather than a measured value — and it directly sets a guarding distance that a person's physical safety depends on, unlike every other number in this chapter, where an error mostly costs money, cycle time, or a re-order. That's exactly why that chapter and this one both say the same thing: $T$ has to be verified by actual stopwatch or data-logger measurement during commissioning, not accepted on the strength of a calculation alone, no matter how carefully the calculation was done. Every other section's worst-case failure mode is expensive; this one's worst-case failure mode is a person getting hurt, which is the right way to think about why this particular number gets a verification requirement none of the others do.

</details>
