---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Sensor Selection and Machine I/O Architecture

```{r setup-06-14, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
library(tidyr)
```

## Learning Objectives

By the end of this chapter, you will be able to:

1. **Select** an appropriate sensor technology — inductive, capacitive, photoelectric, ultrasonic, encoder, or load cell — for a given target material, sensing range, and environment, using a documented decision framework.
2. **Calculate** the derated (effective) sensing distance of an inductive proximity sensor for a non-reference target material and an undersized target, and **verify** whether a specific application will detect reliably.
3. **Explain** the difference between sourcing (PNP) and sinking (NPN) sensor output wiring, and **evaluate** the safety reasoning behind the North American PNP convention.
4. **Select** among hardwired discrete I/O, IO-Link, and fieldbus/industrial Ethernet network I/O for a given application, using a weighted decision framework.
5. **Calculate** the signal error introduced by wire resistance on a 0-10V analog signal versus a 4-20 mA current loop over a long cable run, and **explain** why current loops are the industry default for distance.
6. **Select** washdown-rated sensor housings, seal materials, and connectors for a food-processing environment, and **evaluate** why a standard IP67 sensor can still fail in high-pressure hot washdown.
7. **Tally** a machine's total discrete and analog I/O point count, **apply** an industry rule-of-thumb spare-capacity margin, and **size** I/O hardware accordingly.
8. **Calculate** the voltage budget around a 4-20 mA current loop and **verify** that a specified loop will operate reliably at maximum current.

## Introduction

A controls engineer at a Tier 1 automotive stamping supplier near Windsor, Ontario is putting together the electrical bill of materials for a new blanking and press-feed line. On paper, "add the sensors and I/O" looks like the easiest line item on the whole project — pick some parts, wire them up, done. In practice it's a string of small decisions that each have a wrong answer hiding in them. A steel blank moving past a sensor is easy to detect with almost anything, but the same sensor mounted to catch a stainless steel shift fork, or a clear plastic dunnage tray, behaves completely differently — sometimes it doesn't detect the part at all, and nobody finds out until the line is running and parts start jamming. The output wiring on every one of those sensors has to be specified as sourcing or sinking, and getting that backwards doesn't just mean a rewiring job — in a fault condition it can be the difference between a nuisance stop and an unexpected actuation. Every sensor and every I/O point on the line has to land somewhere: a home-run cable to a remote terminal block, a smart IO-Link connection with built-in diagnostics, or a node on the plant's industrial Ethernet backbone, and the three options trade off wiring cost, commissioning time, and — critically for a line that costs real money every minute it's down — how fast a technician can find and fix a fault. Analog signals, wherever they show up, have to be specified as current loops or voltage signals, and a design engineer who gets that choice backwards on a long cable run builds in a measurement error that won't show up until the run is actually that long. And once every point is counted, someone has to decide how many I/O cards to actually buy — too few and the next engineering change order can't be accommodated without a rack redesign; too many and the customer's capital budget absorbs cost for capacity nobody asked for.

This chapter works through eight of these design tasks, in roughly the order a project team encounters them: sensor technology selection, sensing distance derating for real targets, sourcing/sinking output wiring, hardwired versus smart versus networked I/O, current loop versus voltage analog signals, washdown sensor selection for food plants, total I/O count and spare capacity, and the 4-20 mA loop voltage budget. Several of these tasks lean directly on ideas from earlier in this part: the analog signal task revisits the voltage-drop concepts from the [Wire, Cable, and Circuit Protection](06-04-Wire-Cable-Circuit-Protection.qmd) chapter and the shielding and grounding concepts from the [Bonding, Grounding, and EMC](06-05-Bonding-Grounding-EMC.qmd) chapter, and the loop voltage budget task builds directly on both. Nothing here requires you to have those chapters memorized — the relevant ideas are restated as they're used — but the connection is real, not decorative: sensor and I/O architecture decisions are inseparable from the wiring and grounding practices that carry their signals.

## Design Task 1: Sensor Technology Selection

### Scenario

The same blanking line needs six different sensing points specified before the electrical panel drawings can be released: a part-present check on a steel blank at the feed table, an empty-pallet check on a gravity roller conveyor carrying a mix of wood and light-colored and dark plastic pallets, a coolant level check through the wall of a sealed plastic tank, a rack-clear check across a 4 m aisle before the overhead crane can swing a coil rack through, position and speed feedback off the main conveyor drive shaft, and a force-overload check on the mechanical press itself. Six sensing problems, six different physical quantities to sense, and the wrong technology choice on any one of them means a callback to the panel shop after the drawings are already released.

### Design Guidelines and Rules of Thumb

Six sensor technologies cover the overwhelming majority of discrete presence, distance, and force sensing problems on a factory floor, and each one is built around a different physical effect — which is exactly why no single technology is the right default for every job. Choosing among them starts with three questions, in this order: what physical property of the target can actually be sensed (is it metal? does it have a stable dielectric constant? does it reflect light or sound consistently?), how far away does the sensor need to detect it, and how harsh is the environment the sensor has to survive.

```{r tbl-06-14-sensor-selection, echo=FALSE, message=FALSE}
sensor_table <- tribble(
  ~Technology, ~Target_Sensed, ~Typical_Range, ~Environment_Tolerance, ~Relative_Cost, ~Typical_Application,
  "Inductive", "Ferrous and non-ferrous metals only", "1-40 mm",
  "Excellent -- unaffected by oil, dust, non-metallic debris, and most vibration", "Low",
  "Metal part presence, tooling/fixture position, machine limit switches",
  "Capacitive", "Any material with a stable dielectric constant -- metals, plastics, glass, liquids, powders", "3-60 mm",
  "Fair to good -- sensitive to humidity, condensation, and material buildup on the face unless shielded", "Low-Moderate",
  "Liquid or granular level through a non-metallic tank wall, non-metal part presence",
  "Photoelectric", "Anything that reflects, blocks, or transmits light -- opacity/reflectivity dependent", "0.1 mm to 60+ m (through-beam)",
  "Good, but lens fouling and beam blockage from dust, steam, or fog are real limitations", "Low-Moderate",
  "Long-range part/pallet detection, label and edge detection, box or piece counting",
  "Ultrasonic", "Any solid or liquid surface that reflects sound -- largely color- and transparency-independent", "30 mm to 10 m",
  "Good -- unaffected by target color or clarity, but sensitive to soft/sound-absorbing targets and temperature drift", "Moderate",
  "Clear film or bottle detection, bulk level sensing, targets photoelectric sensors struggle with",
  "Encoder (incremental/absolute)", "N/A -- mechanically coupled to a rotating shaft or linear axis", "N/A (position/speed feedback, not a distance sensor)",
  "Good when properly sealed and coupled; sensitive to shaft misalignment and coupling backlash", "Moderate-High",
  "Conveyor speed/position, servo motor feedback, robot axis feedback",
  "Load cell (strain gauge)", "N/A -- measures applied force or weight", "Application-specific (grams to many tonnes)",
  "Fair to good -- the gauge and cable need environmental protection; sensitive to off-axis (side) loading", "Moderate-High",
  "Batching/weighing, press force monitoring, web tension control, mechanical overload protection"
)

sensor_table %>%
  kbl(caption = "Sensor technology selection reference: what each technology senses, its typical range, and where it fits",
      col.names = c("Technology", "What It Senses", "Typical Range", "Environment Tolerance",
                     "Relative Cost", "Typical Application")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, background = "#2C3E50", color = "white")
```

The first three rows of that table — inductive, capacitive, and photoelectric — are the ones students most often confuse, because all three can be wired and programmed to do the same job (a simple presence/absence output) and all three come in similar-looking cylindrical or rectangular housings. The distinguishing question is always *what is the target made of, and what property of it can the sensor actually couple to*. Inductive sensors only respond to metal, full stop — no amount of range, gain, or sensitivity adjustment makes an inductive sensor see a plastic part. Capacitive sensors respond to anything with a dielectric constant different from air, which sounds universal but makes them the least selective technology on the list: a capacitive sensor tuned to detect a plastic part will often also weakly respond to a hand or a nearby metal bracket, so sensitivity has to be tuned carefully and shielded (housing-guarded) capacitive sensors are usually worth the small cost premium in a busy panel or fixture. Photoelectric sensors respond to light, so they are defeated by anything that blocks, absorbs, or scatters the beam unpredictably — clear or very dark targets are the classic failure case, which is exactly why ultrasonic sensors exist as the fallback: sound doesn't care whether a target is clear, black, or shiny, only whether it presents a reasonably solid, non-absorbing surface to reflect off of.

Encoders and load cells sit in a different category from the other four — they aren't presence/distance sensors at all, they return a continuous position, speed, or force value, and picking between an incremental and absolute encoder, or specifying a load cell's capacity and mounting, is its own design problem (a load cell's own selection process — capacity, accuracy class, and mounting — deserves the same rigor as any other precision instrument, and the mechanical mounting is often the harder part of the job, not the electrical signal).

**Rule of thumb:** when a target's material is metal and the required range is short (under about 40 mm), default to inductive — it is the cheapest, most environmentally robust, and least fussy option on the list, and there is rarely a good reason to reach for capacitive or photoelectric instead. This default breaks down the moment the target isn't metal, or the required range exceeds an inductive sensor's practical limit — at that point the choice has to be made on the target's optical and acoustic properties instead, which is exactly the judgment call this design task exists to formalize.

### Governing Calculation

This design task is a selection framework rather than a formula-driven calculation — the "governing calculation" here is the decision logic itself, applied consistently to each sensing point's known attributes:

1. Is the target metal? If yes and the range is short, default to inductive.
2. If not metal, does the application need long range (multiple meters) or fine feature detection (a small label, a sharp edge)? If yes, consider photoelectric.
3. Is the target's optical properties (color, transparency, surface finish) inconsistent or unreliable — clear film, very dark rubber, wet or frosted surfaces? If yes, prefer ultrasonic over photoelectric.
4. Does the application need to sense through a non-metallic barrier (a tank wall, a hopper wall) or a material with no reliable optical/inductive signature (liquid level, granular level)? If yes, consider capacitive.
5. Is the actual measured quantity a shaft position/speed, or a force/weight, rather than a presence/absence at all? If yes, this isn't a presence sensor problem — specify an encoder or load cell directly.

### Worked Example

```{r sensor-select-06-14, echo=TRUE}
## Six sensing points on the blanking line, with the attributes that drive
## the technology decision
points <- data.frame(
  point = c("Steel blank present at feed table",
            "Empty pallet on gravity roller conveyor (mixed wood/plastic, some clear film wrap)",
            "Coolant level through sealed plastic tank wall",
            "Coil rack clear of overhead crane swing path (4 m aisle)",
            "Main conveyor drive shaft speed/position",
            "Mechanical press force / overload"),
  is_metal_target = c(TRUE, FALSE, FALSE, FALSE, NA, NA),
  needs_long_range = c(FALSE, FALSE, FALSE, TRUE, NA, NA),
  optical_property_unreliable = c(FALSE, TRUE, NA, FALSE, NA, NA),
  senses_through_barrier = c(FALSE, FALSE, TRUE, FALSE, NA, NA),
  is_position_or_force_quantity = c(FALSE, FALSE, FALSE, FALSE, "position", "force"),
  stringsAsFactors = FALSE
)

for (i in seq_len(nrow(points))) {
  p <- points[i, ]
  cat("Point:", p$point, "\n")

  if (identical(p$is_position_or_force_quantity, "position")) {
    choice <- "Encoder"
    reason <- "the measured quantity is shaft speed/position, not presence -- this is a feedback device, not a proximity sensor."
  } else if (identical(p$is_position_or_force_quantity, "force")) {
    choice <- "Load cell"
    reason <- "the measured quantity is applied force, which only a load cell (or similar force transducer) provides."
  } else if (isTRUE(p$is_metal_target) && !isTRUE(p$needs_long_range)) {
    choice <- "Inductive"
    reason <- "the target is metal and the range is short -- the cheapest, most robust default."
  } else if (isTRUE(p$senses_through_barrier)) {
    choice <- "Capacitive"
    reason <- "the target must be sensed through a non-metallic barrier, which only capacitive sensing does reliably."
  } else if (isTRUE(p$needs_long_range)) {
    choice <- "Photoelectric (through-beam)"
    reason <- "long range across an open aisle is exactly the through-beam photoelectric use case."
  } else if (isTRUE(p$optical_property_unreliable)) {
    choice <- "Ultrasonic"
    reason <- "mixed pallet colors and clear film wrap make optical detection unreliable; sound reflects regardless of color/clarity."
  } else {
    choice <- "Photoelectric (diffuse)"
    reason <- "no metal, no barrier, no long-range or unreliable-optics constraint -- a standard diffuse photoelectric fits."
  }

  cat("  Recommended technology:", choice, "\n")
  cat("  Reasoning:", reason, "\n\n")
}
```

**Sanity check:** each recommendation traces back to a single dominant attribute of the target or the application — metal versus non-metal, barrier versus open path, reliable versus unreliable optics, presence versus continuous measurement — which is the entire point of a decision framework: a reviewer should be able to follow *why* each choice was made, not just accept the answer on authority. Note that two of the six points (drive shaft feedback, press force) never entered the presence-sensor decision tree at all — recognizing that a measurement problem isn't a proximity-sensing problem is itself part of the framework.

## Design Task 2: Sensing Distance Derating for Target Material and Size

### Scenario

A process engineer at the same plant has specified an M12 shielded inductive proximity sensor, rated $S_n = 4$ mm for the standard mild steel reference target, to detect a stainless steel (304) shift fork tab as it passes through a fixture. The tab itself measures only 8 mm x 8 mm — smaller than the sensor's reference target size. The sensor is mounted with an 3.5 mm operating gap to the tab at its closest approach. Before the fixture drawing is released, the engineer needs to know whether this sensor will actually detect this specific target reliably, or whether the 4 mm rating on the datasheet is misleading for this application.

### Design Guidelines and Rules of Thumb

An inductive proximity sensor's rated sensing distance, $S_n$, is not a universal number — it is defined against one specific reference condition: a square target of mild steel, 1 mm thick, with a side length equal to the sensor's nominal diameter (or three times $S_n$, whichever is larger), centered on the sensing axis. That reference condition is set by IEC 60947-5-2 (adopted in Canada as CSA C22.2 No. 60947-5-2, which mirrors the IEC content for Canadian product certification), and every catalog $S_n$ value on every manufacturer's datasheet is measured against it. Anything that differs from that reference — a different metal, a smaller target, an off-axis approach — reduces the *effective* sensing distance below the catalog number, and none of that reduction shows up unless it's calculated explicitly.

Non-ferrous and non-mild-steel metals couple less efficiently with the sensor's oscillating field, so manufacturers publish a **correction factor**, $C_{f,material}$, for each common target material relative to mild steel. These factors vary somewhat by manufacturer and sensor design (some "factor 1" sensors are specifically engineered to reduce this variation), but the values below are representative of typical unshielded and shielded inductive sensors and are a reasonable planning baseline pending the actual selected sensor's datasheet.

```{r fig-06-14-material-correction, echo=FALSE, fig.cap="Typical inductive sensor correction factors by target material, relative to the mild steel reference target (factor = 1.0). Representative values -- always confirm against the selected sensor's datasheet.", fig.align='center', fig.width=9, fig.height=5.5}
mat_df <- data.frame(
  material = factor(c("Mild steel\n(reference)", "Stainless\nsteel 304", "Brass", "Aluminum", "Copper"),
                     levels = c("Mild steel\n(reference)", "Stainless\nsteel 304", "Brass", "Aluminum", "Copper")),
  correction_factor = c(1.0, 0.7, 0.5, 0.4, 0.3)
)

ggplot(mat_df, aes(x = material, y = correction_factor, fill = material)) +
  geom_col(show.legend = FALSE) +
  annotate("segment", x = 0.5, xend = 5.5, y = 1.0, yend = 1.0,
           linetype = "dashed", color = "grey30", linewidth = 0.8) +
  annotate("text", x = 4.6, y = 1.06, label = "Reference (mild steel)",
           size = 3 * diagram_text_scale, color = "grey30") +
  geom_text(aes(label = sprintf("%.1f", correction_factor)), vjust = -0.6,
            size = 3.4 * diagram_text_scale) +
  scale_fill_manual(values = c("#7F8C8D", "#3498DB", "#F39C12", "#95A5A6", "#E67E22")) +
  labs(title = "Inductive Sensor Correction Factor by Target Material",
       x = NULL, y = "Correction factor (relative to mild steel)") +
  ylim(0, 1.15) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
        axis.title = element_text(size = 11 * diagram_text_scale),
        axis.text = element_text(size = 10 * diagram_text_scale))
```

A second, independent derating applies when the target is smaller than the reference target: the sensor's field spreads out ahead of the sensing face, and a target that doesn't fill that field couples less efficiently, exactly the way a smaller mirror reflects less of a light beam. As a simplified planning estimate, the effective range scales down roughly in proportion to the ratio of the target's characteristic dimension to the reference target's dimension, capped at 1.0 (a target larger than the reference gets no *bonus*, only a target smaller than it gets penalized).

```{r fig-06-14-target-size-schematic, echo=FALSE, fig.cap="An undersized target does not fill the sensor's reference field, further reducing the effective sensing distance beyond the material correction alone. Schematic, not to scale.", fig.align='center', fig.width=10, fig.height=5}
ggplot() +
  annotate("rect", xmin = 0, xmax = 1.2, ymin = -1.5, ymax = 1.5, fill = "grey35") +
  annotate("text", x = 0.6, y = 1.9, label = "Sensor\nsensing face", size = 3 * diagram_text_scale) +
  annotate("rect", xmin = 4.5, xmax = 6.3, ymin = -0.9, ymax = 0.9, fill = NA, color = "grey50",
           linetype = "dashed", linewidth = 1) +
  annotate("text", x = 5.4, y = -1.25, label = "Reference target (12 x 12 mm)",
           size = 2.8 * diagram_text_scale, color = "grey50") +
  annotate("rect", xmin = 5.1, xmax = 5.7, ymin = -0.3, ymax = 0.3, fill = "#3498DB", color = "black") +
  annotate("text", x = 5.4, y = 0.55, label = "Actual target\n(8 x 8 mm)",
           size = 2.8 * diagram_text_scale, color = "#2C3E50") +
  annotate("segment", x = 1.2, xend = 5.1, y = -2.1, yend = -2.1,
           arrow = arrow(length = unit(0.15, "cm"), ends = "both", type = "closed"), linewidth = 0.8) +
  annotate("text", x = 3.15, y = -2.5, label = "Operating gap", size = 3 * diagram_text_scale) +
  coord_fixed(xlim = c(-0.3, 7), ylim = c(-3, 2.3)) +
  theme_void()
```

**Rule of thumb:** treat the material and size corrections as multiplicative, then apply a further margin before trusting the result — use no more than about 80% of the fully-derated effective sensing distance as the *usable, reliable* range, to leave headroom for supply voltage variation, temperature drift, and target-approach angle that the catalog number doesn't capture. This 80% margin breaks down (in the sense of not being conservative enough) for applications with high vibration or poor fixture repeatability, where the target-to-sensor gap itself varies shot to shot — those applications need the actual gap tolerance stacked into the check, not just a flat percentage.

### Governing Calculation

$$S_{r,eff} = S_n \times C_{f,material} \times C_{f,size}$$

$$C_{f,size} = \min\left(1,\ \frac{d_{target}}{d_{reference}}\right)$$

$$S_{usable} = 0.8 \times S_{r,eff}$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $S_n$ | Rated sensing distance for the mild steel reference target | mm |
| $C_{f,material}$ | Correction factor for the actual target material | dimensionless (0-1) |
| $d_{target}$ | Characteristic dimension (side length) of the actual target | mm |
| $d_{reference}$ | Characteristic dimension (side length) of the reference target for this sensor | mm |
| $C_{f,size}$ | Correction factor for target size relative to the reference target | dimensionless (0-1) |
| $S_{r,eff}$ | Fully-derated effective sensing distance | mm |
| $S_{usable}$ | Usable, reliable sensing distance after the 80% margin | mm |

Live cell — try a different material, target size, or mounting gap and see whether the sensor still passes.

```{webr-r}
#| label: sensing-derate-06-14
## Given
Sn <- 4              # mm, rated sensing distance for mild steel reference target (M12 shielded)
d_reference <- 12    # mm, reference target side length for this sensor (its nominal diameter)

## Actual application
target_material <- "Stainless steel 304"
Cf_material <- 0.7   # from the correction factor chart above
d_target <- 8        # mm, actual target side length (smaller than the reference target)
mounting_gap <- 3.5  # mm, actual operating gap in the fixture

Cf_size <- min(1, d_target / d_reference)
Sr_eff <- Sn * Cf_material * Cf_size
Sr_usable <- 0.8 * Sr_eff

cat("Rated sensing distance, Sn (mild steel reference):", Sn, "mm\n")
cat("Target material:", target_material, " (Cf_material =", Cf_material, ")\n")
cat("Target size:", d_target, "x", d_target, "mm vs. reference", d_reference, "x", d_reference, "mm",
    " (Cf_size =", round(Cf_size, 3), ")\n")
cat("Fully-derated effective sensing distance =", round(Sr_eff, 3), "mm\n")
cat("Usable sensing distance (80% margin) =", round(Sr_usable, 3), "mm\n\n")

cat("Planned mounting gap:", mounting_gap, "mm\n")
if (mounting_gap <= Sr_usable) {
  cat("Mounting gap <= usable sensing distance: PASS\n")
} else {
  cat("Mounting gap > usable sensing distance: FAILS\n\n")
  cat("Corrective options:\n")
  max_gap_this_sensor <- Sr_usable
  cat(" 1. Reduce the mounting gap to no more than", round(max_gap_this_sensor, 2), "mm\n")

  ## Option: a larger sensor (bigger reference target, bigger Sn)
  Sn_larger <- 8       # mm, M18 shielded sensor
  d_reference_larger <- 18
  Cf_size_larger <- min(1, d_target / d_reference_larger)
  Sr_eff_larger <- Sn_larger * Cf_material * Cf_size_larger
  Sr_usable_larger <- 0.8 * Sr_eff_larger
  cat(" 2. Step up to an M18 sensor (Sn = 8 mm): usable distance becomes",
      round(Sr_usable_larger, 2), "mm --",
      ifelse(mounting_gap <= Sr_usable_larger, "PASSES at the current gap", "still fails at the current gap"), "\n")
}
```

**Sanity check:** the naive comparison an engineer might make at a glance — "4 mm rated, 3.5 mm gap, that's fine" — is wrong by a wide margin. Once the stainless steel correction (0.7) and the undersized-target correction (0.67) are both applied, the usable range collapses to under 1.5 mm, less than half the planned gap. This is a common and genuinely dangerous mistake precisely because it's invisible on the datasheet: the 4 mm number is real, it's just not the number that applies here.

## Design Task 3: Sourcing (PNP) vs. Sinking (NPN) Output Wiring

### Scenario

The same panel shop wires sensors from two different suppliers onto the same machine — one ships PNP output sensors by default, the other ships NPN. The apprentice wiring the panel asks a reasonable question: does it actually matter which one goes where, as long as the sensor turns the right thing on at the right time? The senior controls engineer's answer is yes, for two separate reasons — one about which PLC input card each type needs, and a second, more important one, about what happens when a wire fails.

### Design Guidelines and Rules of Thumb

A three-wire DC sensor always has a supply (+V, commonly brown), a common (0V, commonly blue), and a switched output (commonly black). The difference between PNP and NPN is entirely about *which side of the load the output transistor switches*.

A **PNP (sourcing)** output connects the load's supply-side terminal to +V when the sensor is active — the sensor sources current *out* to the load, and the load's other terminal returns to 0V. An **NPN (sinking)** output does the opposite: the load's supply-side terminal is permanently wired to +V, and the sensor's output transistor connects the load's other terminal down to 0V when active — the sensor sinks current *in* from the load. Both wiring schemes turn the load on and off identically under normal operation; the difference only shows up in a fault.

```{r fig-06-14-pnp-npn-wiring, echo=FALSE, fig.cap="Sourcing (PNP) vs. sinking (NPN) sensor output wiring, with a chafed-wire-to-grounded-frame fault marked on each. The PNP fault pulls the signal line low (reads OFF); the NPN fault pulls the signal line low in a way that mimics the sensor being active (can read ON).", fig.align='center', fig.width=13, fig.height=7}
build_wiring_panel <- function(panel_title, out_switches_high, subtitle_result) {
  p <- ggplot() +
    ## +24V and 0V rails
    annotate("segment", x = 0, xend = 5, y = 5, yend = 5, linewidth = 1.3, color = "#C0392B") +
    annotate("text", x = -0.3, y = 5, label = "+24V", size = 3.2 * diagram_text_scale, color = "#C0392B", hjust = 1) +
    annotate("segment", x = 0, xend = 5, y = 0.3, yend = 0.3, linewidth = 1.3, color = "#2C3E50") +
    annotate("text", x = -0.3, y = 0.3, label = "0V", size = 3.2 * diagram_text_scale, color = "#2C3E50", hjust = 1)

  if (out_switches_high) {
    ## PNP: sensor on the left, load on the right, switched wire carries +24V when active
    p <- p +
      annotate("rect", xmin = 0.3, xmax = 1.5, ymin = 2.6, ymax = 3.9, fill = "#3498DB", color = "black") +
      annotate("text", x = 0.9, y = 3.25, label = "PNP\nSensor", size = 3 * diagram_text_scale, color = "white") +
      annotate("segment", x = 0.6, xend = 0.6, y = 5, yend = 3.9, linewidth = 1) +
      annotate("segment", x = 1.2, xend = 1.2, y = 2.6, yend = 0.3, linewidth = 1) +
      annotate("segment", x = 1.5, xend = 3.4, y = 3.25, yend = 3.25, linewidth = 1.4, color = "#8E44AD") +
      annotate("rect", xmin = 3.4, xmax = 4.6, ymin = 2.6, ymax = 3.9, fill = "#D5B27F", color = "black") +
      annotate("text", x = 4.0, y = 3.25, label = "Load", size = 3 * diagram_text_scale) +
      annotate("segment", x = 4.0, xend = 4.0, y = 2.6, yend = 0.3, linewidth = 1) +
      annotate("text", x = 2.45, y = 3.7, label = "OUT (switches +24V to load when active)",
               size = 2.5 * diagram_text_scale, color = "#8E44AD") +
      annotate("point", x = 2.45, y = 3.25, shape = 4, size = 5, stroke = 1.6, color = "red") +
      annotate("segment", x = 2.45, xend = 2.45, y = 3.25, yend = 1.5, linetype = "dashed", color = "red", linewidth = 0.8) +
      annotate("rect", xmin = 2.0, xmax = 2.9, ymin = 0.9, ymax = 1.5, fill = "grey60", color = "black") +
      annotate("text", x = 2.45, y = 1.2, label = "Grounded\nframe", size = 2.5 * diagram_text_scale)
  } else {
    ## NPN: load on the left (tied to +24V), sensor on the right, switched wire carries 0V when active
    p <- p +
      annotate("rect", xmin = 0.3, xmax = 1.5, ymin = 2.6, ymax = 3.9, fill = "#D5B27F", color = "black") +
      annotate("text", x = 0.9, y = 3.25, label = "Load", size = 3 * diagram_text_scale) +
      annotate("segment", x = 0.6, xend = 0.6, y = 5, yend = 3.9, linewidth = 1) +
      annotate("segment", x = 1.5, xend = 3.4, y = 3.25, yend = 3.25, linewidth = 1.4, color = "#8E44AD") +
      annotate("rect", xmin = 3.4, xmax = 4.6, ymin = 2.6, ymax = 3.9, fill = "#3498DB", color = "black") +
      annotate("text", x = 4.0, y = 3.25, label = "NPN\nSensor", size = 3 * diagram_text_scale, color = "white") +
      annotate("segment", x = 4.3, xend = 4.3, y = 5, yend = 3.9, linewidth = 1) +
      annotate("segment", x = 3.7, xend = 3.7, y = 2.6, yend = 0.3, linewidth = 1) +
      annotate("text", x = 2.45, y = 3.7, label = "OUT (switches to 0V when active)",
               size = 2.5 * diagram_text_scale, color = "#8E44AD") +
      annotate("point", x = 2.45, y = 3.25, shape = 4, size = 5, stroke = 1.6, color = "red") +
      annotate("segment", x = 2.45, xend = 2.45, y = 3.25, yend = 1.5, linetype = "dashed", color = "red", linewidth = 0.8) +
      annotate("rect", xmin = 2.0, xmax = 2.9, ymin = 0.9, ymax = 1.5, fill = "grey60", color = "black") +
      annotate("text", x = 2.45, y = 1.2, label = "Grounded\nframe", size = 2.5 * diagram_text_scale)
  }

  p +
    coord_fixed(xlim = c(-1.1, 5.3), ylim = c(0, 5.6)) +
    theme_void() +
    labs(title = panel_title, subtitle = subtitle_result) +
    theme(plot.title = element_text(size = 13 * diagram_text_scale, face = "bold", hjust = 0.5),
          plot.subtitle = element_text(size = 9.5 * diagram_text_scale, hjust = 0.5, color = "#27AE60"))
}

pnp_panel <- build_wiring_panel("PNP (Sourcing)", TRUE,
                                 "Fault pulls signal LOW -> reads OFF (fails to a detectable, safe state)")
npn_panel <- build_wiring_panel("NPN (Sinking)", FALSE,
                                 "Fault pulls signal LOW -> mimics sensor ACTIVE (can read ON unexpectedly)")

gridExtra::grid.arrange(pnp_panel, npn_panel, ncol = 2)
```

Now consider a wire's insulation chafing through against the machine's grounded, bonded metal frame — a genuinely common failure mode in a cable track or drag chain after years of flexing. On the PNP wiring, the switched wire normally sits at +24V when active and floats near 0V (through the load) when inactive; a short to the grounded frame pulls that wire to 0V regardless of the sensor's actual state. The downstream logic sees "signal low," which is indistinguishable from the sensor simply being off — a nuisance fault, easy to diagnose, and critically, it never makes the load turn on when it shouldn't. On the NPN wiring, the switched wire normally sits near +24V (through the load) when inactive and gets pulled to 0V by the sensor's transistor when active; a short to the grounded frame *also* pulls that wire to 0V — which is exactly what the sensor does on purpose when it's genuinely detecting the target. The fault is indistinguishable from a legitimate "target present" signal, and depending on what that signal drives, the result can be an unintended actuation, not just a nuisance stop.

This is the practical reasoning behind North America's default preference for PNP sensors, wired to sinking-type PLC input cards: with grounded, bonded machine frames being the norm under Canadian and US electrical practice (bonding requirements under the CEC, and the equivalent NEC Article 250 in the US), a wire-to-frame fault is a far more common real-world failure than a wire-to-supply fault, and PNP wiring's failure mode for that specific fault is the safer of the two in most common circuit configurations. European practice has historically leaned toward NPN sensors in some legacy contexts (older PLC families and some European OEM standards defaulted to sinking-output field devices paired with sourcing-type input cards), but the North American convention of PNP-out-to-sinking-input is now the dominant specification on new machinery built for the North American market, including machinery imported from Europe for a North American plant.

**Rule of thumb:** in North America, specify PNP output sensors paired with sinking-type digital input cards as the default for new machine designs, and treat NPN as the exception that needs a specific justification (matching an existing NPN fleet, a supplier constraint, or a legacy PLC platform that only supports sourcing inputs). This default breaks down inside a safety-rated circuit, where the actual failure-mode analysis is done formally per the safety function's category and architecture (covered in the [Machine Safety Circuits](06-02-Machine-Safety-Circuits.qmd) chapter) rather than by this general PNP/NPN heuristic — a safety input's fault behavior is verified explicitly, not assumed from a wiring convention.

### Governing Calculation

There's no formula here — the "calculation" is a fault-mode walkthrough, applied consistently to both wiring schemes:

1. Identify the wire most exposed to mechanical damage (cable track flex point, drag chain, moving fixture).
2. Determine that wire's normal-state voltage in both the active and inactive sensor states.
3. Determine what a short-to-grounded-frame fault on that wire would force the voltage to.
4. Compare: does the fault-forced state match the sensor's "active" state (dangerous — indistinguishable from real detection) or its "inactive" state (safer — reads as a fault/off condition)?

### Worked Example

Live cell — change `output_type` to `"NPN"` and re-run to see the fault analysis flip.

```{webr-r}
#| label: pnp-npn-fault-06-14
## Given: a sensor's switched output wire runs through a drag chain and is
## exposed to chafing against the machine's grounded, bonded frame.

output_type <- "PNP"   # try "NPN"

cat("Output type:", output_type, "\n\n")

if (output_type == "PNP") {
  voltage_when_active <- "+24V (sourced to load)"
  voltage_when_inactive <- "~0V (through load to 0V rail)"
  fault_forces_wire_to <- "0V"
  matches_active_state <- FALSE
} else {
  voltage_when_active <- "~0V (sunk through load)"
  voltage_when_inactive <- "+24V (through load to +24V rail)"
  fault_forces_wire_to <- "0V"
  matches_active_state <- TRUE
}

cat("Normal voltage when sensor is ACTIVE:  ", voltage_when_active, "\n")
cat("Normal voltage when sensor is INACTIVE:", voltage_when_inactive, "\n")
cat("A chafed-wire short to the grounded frame forces the wire to:", fault_forces_wire_to, "\n\n")

if (matches_active_state) {
  cat("CONCLUSION: the fault-forced state matches the ACTIVE state.\n")
  cat("Downstream logic cannot distinguish this fault from a real detection --\n")
  cat("this is the higher-risk failure mode.\n")
} else {
  cat("CONCLUSION: the fault-forced state matches the INACTIVE state.\n")
  cat("Downstream logic reads this as the sensor simply being off -- a\n")
  cat("detectable nuisance fault, not a false detection.\n")
}
```

**Sanity check:** switching `output_type` to `"NPN"` flips the conclusion exactly as the wiring diagram predicts — the physical wiring hasn't changed conceptually, only which rail the load sits against, and that single difference is what determines whether a chafed-wire fault reads as "off" or mimics "on." This is precisely why the choice matters even though both wiring schemes control the load identically under normal, fault-free operation.

## Design Task 4: IO-Link and Network I/O vs. Hardwired I/O

### Scenario

An automotive body-in-white integrator is architecting the I/O for a five-station respot welding cell — roughly 180 discrete and analog points across the cell's sensors, clamps, and interlocks. Line uptime is the customer's top priority: this cell feeds a paint line downstream, and every minute of unplanned stoppage cascades. The integrator has to decide, point by point and zone by zone, whether to run traditional hardwired discrete I/O back to a remote terminal block, IO-Link point-to-point smart sensor connections into an IO-Link master, or full fieldbus/industrial Ethernet distributed I/O nodes at each station.

### Design Guidelines and Rules of Thumb

**Hardwired discrete I/O** is the baseline every other option is compared against: a dedicated conductor pair runs from each field device directly to a terminal block or I/O card, carrying a simple on/off (or 4-20 mA/0-10V analog) signal. It's the cheapest technology per point, the easiest for a technician with no specialized training to troubleshoot with a multimeter, and it has no protocol, no addressing, and nothing to configure. Its weakness is exactly its simplicity: the signal carries no information beyond on/off, so a sensor reporting "not detecting" looks identical whether the target is genuinely absent, the lens is dirty, the sensor has failed, or a wire has broken — a technician has to walk out and physically investigate every time, and on a large cell the sheer wiring labor (a home-run cable per point) is a real, non-trivial cost and installation-time driver.

**IO-Link** keeps the same point-to-point topology — one cable per device — but replaces the simple switched signal with a digital communication protocol running over the same three-wire connector, back to an IO-Link master block that itself connects to the PLC's network. Because the connection is digital, the sensor can report far more than on/off: signal strength/margin, internal temperature, operating hour counters, and manufacturer-specific diagnostic codes, all without adding a single additional wire. IO-Link masters also support remote parameter download, so a replacement sensor can be swapped and automatically reconfigured to the previous unit's settings without a technician touching a rotary switch or a programming pendant — a meaningful commissioning-speed and spare-parts-simplification advantage on a cell with dozens of similar sensors.

**Fieldbus and industrial Ethernet network I/O** (EtherNet/IP, PROFINET, and similar) replaces most of the point-to-point wiring altogether: a distributed I/O node is mounted physically near a cluster of devices (a station, a zone), and a single network cable carries all of that node's data back to the PLC, alongside device-level diagnostics, network health data, and — in many modern implementations — the ability to reach every device on the network from a single engineering workstation without walking the floor. This is the most capable option for diagnostics and the most scalable for a large or growing installation, but it's also the most expensive per node, requires network design and IP/node addressing discipline that a hardwired or IO-Link installation doesn't need, and a network configuration mistake can take out an entire zone's I/O at once rather than a single point.

```{r tbl-06-14-io-architecture, echo=FALSE, message=FALSE}
io_arch_table <- tribble(
  ~Architecture, ~Diagnostics, ~Wiring_Cost_Complexity, ~Commissioning_Speed, ~Best_Fit,
  "Hardwired discrete I/O", "None beyond on/off -- a fault looks identical to a legitimate absent condition",
  "Highest -- one home-run cable per point", "Simple per point, but slow in aggregate on a high point-count cell",
  "Small cells, simple machines, or any point where the extra cost of smart wiring genuinely isn't justified",
  "IO-Link", "Rich per-device diagnostics (signal margin, health, counters) over the same cable as the switched signal",
  "Same point-to-point topology as hardwired, but standard cheap unshielded cable and simpler termination",
  "Fast -- auto-parameter download and hot-swap replacement without reprogramming",
  "Sensor-dense cells needing fast fault diagnosis and quick sensor swaps without a controls tech's involvement",
  "Fieldbus / Industrial Ethernet distributed I/O", "Best -- full device and network diagnostics, remote access from an engineering workstation",
  "Lowest per-point wiring, but network design, addressing, and infrastructure add real complexity",
  "Slower to commission (node/address setup) but highly scalable once established",
  "Large, multi-station, or plant-wide installations where diagnostics-driven uptime and scalability matter most"
)

io_arch_table %>%
  kbl(caption = "Hardwired vs. IO-Link vs. fieldbus/industrial Ethernet I/O: trade-offs",
      col.names = c("Architecture", "Diagnostics", "Wiring Cost/Complexity", "Commissioning Speed", "Best Fit")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, background = "#2C3E50", color = "white")
```

```{r fig-06-14-io-architecture-compare, echo=FALSE, fig.cap="Illustrative comparison of the three I/O architectures across four decision criteria. Higher is better in each case (for Wiring Cost, a high bar means lower cost).", fig.align='center', fig.width=10, fig.height=6}
compare_df <- data.frame(
  Criterion = c("Diagnostics", "Wiring\nCost", "Commissioning\nSpeed", "Scalability"),
  Hardwired = c(1, 2, 3, 2),
  `IO-Link` = c(4, 4, 4, 3),
  `Fieldbus/Ethernet` = c(5, 3, 3, 5),
  check.names = FALSE
)
compare_long <- pivot_longer(compare_df, cols = c(Hardwired, `IO-Link`, `Fieldbus/Ethernet`),
                              names_to = "Architecture", values_to = "Score")
compare_long$Architecture <- factor(compare_long$Architecture,
                                     levels = c("Hardwired", "IO-Link", "Fieldbus/Ethernet"))
compare_long$Criterion <- factor(compare_long$Criterion, levels = compare_df$Criterion)

ggplot(compare_long, aes(x = Criterion, y = Score, fill = Architecture)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Hardwired" = "#95A5A6", "IO-Link" = "#3498DB", "Fieldbus/Ethernet" = "#E67E22")) +
  labs(title = "I/O Architecture Comparison -- Illustrative Scoring",
       subtitle = "Relative rating, 1 (poor fit) to 5 (excellent fit) -- actual scores are installation-specific",
       x = NULL, y = "Relative rating (1-5)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 9.5 * diagram_text_scale),
        legend.position = "bottom",
        axis.text = element_text(size = 10 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale))
```

**Rule of thumb:** on a large, diagnostics-sensitive automotive cell, weight diagnostics and scalability heavily and the architecture choice tends toward fieldbus/industrial Ethernet for the cell's backbone, with IO-Link filling in at the device level within each node. This breaks down for a small, standalone machine (a single manual load station, a simple conveyor) where the fixed cost of network infrastructure and the commissioning overhead of node addressing isn't justified by the point count — hardwired, or IO-Link for just the handful of sensors that benefit most from diagnostics, is often the right-sized answer there instead.

### Worked Example

Live cell — try changing the `weights` to reflect a smaller, standalone cell's priorities (lower diagnostics weight, higher cost weight) and see whether the recommendation changes.

```{webr-r}
#| label: io-architecture-decision-06-14
## Weighted decision matrix, weights reflecting THIS cell's priority: uptime
## (diagnostics) matters most, given the downstream paint-line dependency
weights <- c(diagnostics = 0.40, wiring_cost = 0.15, commissioning_speed = 0.20, scalability = 0.25)

hardwired_scores <- c(diagnostics = 1, wiring_cost = 2, commissioning_speed = 3, scalability = 2)
iolink_scores <- c(diagnostics = 4, wiring_cost = 4, commissioning_speed = 4, scalability = 3)
fieldbus_scores <- c(diagnostics = 5, wiring_cost = 3, commissioning_speed = 3, scalability = 5)

hardwired_total <- sum(weights * hardwired_scores)
iolink_total <- sum(weights * iolink_scores)
fieldbus_total <- sum(weights * fieldbus_scores)

cat("Weighted decision matrix (1 = poor fit, 5 = excellent fit):\n\n")
for (nm in names(weights)) cat(sprintf("  %-20s weight = %.0f%%\n", nm, weights[nm] * 100))

cat(sprintf("\nHardwired weighted score:        %.2f\n", hardwired_total))
cat(sprintf("IO-Link weighted score:          %.2f\n", iolink_total))
cat(sprintf("Fieldbus/Ethernet weighted score: %.2f\n", fieldbus_total))

scores <- c(Hardwired = hardwired_total, `IO-Link` = iolink_total, `Fieldbus/Ethernet` = fieldbus_total)
cat(sprintf("\n-> Recommendation: %s\n", names(scores)[which.max(scores)]))
```

**Sanity check:** with diagnostics and scalability weighted most heavily — the right call for a high-uptime, multi-station automotive cell feeding a downstream process — fieldbus/industrial Ethernet wins, matching the design guidance above. Try dropping the diagnostics weight and raising wiring cost instead (closer to a small standalone machine's priorities): the recommendation shifts, which is exactly the point of a weighted framework — it should be traceable to the specific installation's priorities, not a single "always correct" answer.

## Design Task 5: Analog Signal Selection — 4-20 mA vs. 0-10V

### Scenario

A food processing plant in southwestern Ontario is running a temperature transmitter on a retort cook line, physically located 180 m of cable run from the control room where the PLC's analog input card sits — a routine distance in a large single-story processing building. The instrumentation engineer has to decide whether to specify the transmitter's output as a 0-10V voltage signal or a 4-20 mA current loop before the cable is pulled, because changing the decision later means repulling the run.

### Design Guidelines and Rules of Thumb

A 0-10V signal is a **voltage source** signal: the transmitter presents a voltage proportional to the measured value, and the receiving device (the PLC's analog input) reads that voltage relative to a shared 0V reference. A 4-20 mA signal is a **current loop** signal: the transmitter (in the common two-wire, loop-powered configuration) acts as a current *regulator* rather than a voltage source, controlling the current flowing around a series loop to be proportional to the measured value, regardless of what voltage that requires across the wire and the receiving device to sustain.

That difference in *what physical quantity actually carries the signal* is the whole reason current loops dominate industrial practice for anything beyond a short, panel-internal run. In a voltage signal circuit, any current that flows to the receiving device's input impedance also flows through the wire's own resistance, and that wire resistance eats a share of the source voltage before it ever reaches the receiver — the longer the run (or the lower the receiver's input impedance, or the more devices sharing the same signal in parallel), the more of the signal is lost in the wire itself, and the error grows directly with cable length. In a current loop circuit, the *same current* flows through every element of the series loop — the transmitter, the wire, and the receiving device's burden resistor — by definition, because it's a series circuit; wire resistance doesn't divide the current down the way it divides down a voltage, it only adds a voltage drop that the loop's power supply has to have enough headroom (compliance voltage) to overcome. As long as that headroom exists — which Design Task 8 checks explicitly — the current delivered to the receiver, and therefore the reading, is unaffected by how long the wire run is.

```{r fig-06-14-signal-error-length, echo=FALSE, fig.cap="Signal error vs. cable length for a 0-10V voltage signal at two receiver input impedances, compared to a 4-20 mA current loop. The current loop's signal error stays at zero regardless of length (up to the compliance voltage limit checked separately).", fig.align='center', fig.width=10, fig.height=6}
R_per_m_18awg <- 6.385 / 304.8   # ohm/m, one-way, 18 AWG copper (typical table value)
length_seq <- seq(0, 300, by = 5)
R_wire_seq <- 2 * R_per_m_18awg * length_seq   # round-trip wire resistance

err_500 <- R_wire_seq / (R_wire_seq + 500) * 100
err_100k <- R_wire_seq / (R_wire_seq + 100000) * 100
err_loop <- rep(0, length(length_seq))

signal_df <- data.frame(
  length_m = rep(length_seq, 3),
  error_pct = c(err_500, err_100k, err_loop),
  scenario = rep(c("0-10V, 500 ohm effective load", "0-10V, 100 kohm input impedance", "4-20 mA current loop"),
                 each = length(length_seq))
)
signal_df$scenario <- factor(signal_df$scenario,
                              levels = c("0-10V, 500 ohm effective load", "0-10V, 100 kohm input impedance",
                                         "4-20 mA current loop"))

ggplot(signal_df, aes(x = length_m, y = error_pct, color = scenario)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("0-10V, 500 ohm effective load" = "#E74C3C",
                                 "0-10V, 100 kohm input impedance" = "#F39C12",
                                 "4-20 mA current loop" = "#27AE60")) +
  labs(title = "Signal Error vs. Cable Length",
       subtitle = "18 AWG copper, round-trip wire resistance -- current loop error stays at zero",
       x = "One-way cable length (m)", y = "Signal error (%)", color = NULL) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 9.5 * diagram_text_scale),
        legend.position = "bottom",
        axis.title = element_text(size = 11 * diagram_text_scale),
        axis.text = element_text(size = 10 * diagram_text_scale))
```

There's a second, equally practical reason current loops dominate: a voltage signal's 0V reference has to be shared between the transmitter and the receiver, and over a long run in an industrial plant, the two ends are often bonded to earth at physically different points — a difference in local ground potential (covered in the [Bonding, Grounding, and EMC](06-05-Bonding-Grounding-EMC.qmd) chapter) adds directly onto a voltage signal's reading as an offset error that has nothing to do with wire resistance at all. A current loop's signal-carrying quantity, the current itself, doesn't depend on the two ends sharing a common reference the same way, which is also why current loops tolerate a shared cable tray with noisy power conductors better than voltage signals do — proper shielding and grounding practice, from the same chapter, still matters for both, but the current loop starts from a structurally more robust position.

**Rule of thumb:** default to 4-20 mA for any analog signal run longer than roughly 15-30 m, or any run sharing a cable tray or conduit with power wiring, regardless of the specific receiver's input impedance — the wire-resistance argument alone may or may not be significant depending on the receiver, but the ground-reference and noise-immunity advantages apply regardless. This default breaks down only for genuinely short, panel-internal signal runs (a transmitter and its PLC sitting in the same enclosure), where a 0-10V signal's lower cost and simpler wiring (no loop power supply headroom to budget) can be the more practical choice.

### Governing Calculation

$$V_{error,\%} = \frac{R_{wire}}{R_{wire} + R_{in}} \times 100\%$$

$$R_{wire} = 2 \times r_{per\ m} \times L$$

For the current loop, the current delivered to the receiver's burden resistor is, by definition of a series circuit, the loop current $I$ itself — independent of $R_{wire}$, as long as the compliance voltage requirement (checked in Design Task 8) is satisfied.

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $r_{per\ m}$ | Wire resistance per meter, one-way | Ω/m |
| $L$ | One-way cable length | m |
| $R_{wire}$ | Round-trip (loop) wire resistance | Ω |
| $R_{in}$ | Receiving device's effective input impedance (or parallel-loaded combination) | Ω |
| $V_{error,\%}$ | Signal error introduced by the wire-resistance voltage divider | % |

### Worked Example

Live cell — try a different cable length or receiver input impedance and see how the voltage-signal error changes.

```{webr-r}
#| label: signal-error-06-14
## Given: 180 m run, 18 AWG copper twisted pair
R_per_m_18awg <- 6.385 / 304.8   # ohm/m, one-way (18 AWG, typical table value)
L <- 180                          # m, one-way cable length
R_wire <- 2 * R_per_m_18awg * L   # ohm, round-trip

## The 0-10V signal is tee'd to both the PLC analog input and a local panel
## meter -- a common real-world mistake -- dropping the effective parallel
## input impedance to about 500 ohm
R_in <- 500   # ohm, effective parallel-loaded input impedance

V_error_pct <- R_wire / (R_wire + R_in) * 100

cat("Cable length:", L, "m (one-way), 18 AWG copper\n")
cat("Round-trip wire resistance:", round(R_wire, 2), "ohm\n")
cat("Effective receiver input impedance (parallel-loaded):", R_in, "ohm\n")
cat("0-10V signal error:", round(V_error_pct, 2), "%\n\n")

accuracy_spec_pct <- 1.0   # required measurement accuracy for this process
if (V_error_pct <= accuracy_spec_pct) {
  cat("0-10V error <=", accuracy_spec_pct, "% accuracy spec: PASS\n")
} else {
  cat("0-10V error exceeds the", accuracy_spec_pct, "% accuracy spec: FAILS\n\n")
}

cat("\n4-20 mA current loop over the same 180 m run:\n")
cat("Loop current at the receiver's burden resistor = loop current, unaffected by R_wire\n")
cat("4-20 mA signal error: 0.0 % (independent of cable length, given adequate compliance voltage)\n")
```

**Sanity check:** at 180 m with a de-rated (parallel-loaded) input impedance of 500 Ω, the 0-10V signal loses nearly 1.5% of its span in the wire alone — enough to fail a ±1% process accuracy requirement outright, and that's before accounting for the ground-reference offset that a shared voltage signal picks up over that same distance. The 4-20 mA loop is unaffected by any of it, which is exactly why it's the default choice for this run — and exactly the kind of error that's invisible on a bench test (a short cable in the shop) and only shows up once the real 180 m of plant floor is between the transmitter and the PLC.

## Design Task 6: IP Ratings and Washdown Sensor Selection

### Scenario

A poultry processing plant near Kitchener, Ontario runs a caustic, high-pressure hot washdown at the end of every production shift — hand-held wash guns delivering roughly 1500 psi at up to 80°C directly at conveyor frames, guarding, and every sensor mounted near the line. Standard black-plastic photoelectric sensors, rated IP67 by their manufacturer, have been failing every few months: fogged lenses, corroded connector pins, and eventually complete signal loss. The plant engineer, reasonably, is confused — IP67 is supposed to mean the sensor survives water. The question is why it isn't holding up, and what to specify instead.

### Design Guidelines and Rules of Thumb

IP ratings (from IEC 60529) describe ingress protection against two separate hazards — solids (the first digit) and water (the second digit) — and the water test associated with **IP67** is immersion: the device is submerged up to 1 m of depth for 30 minutes, under static (non-pressurized) conditions. That's a real, meaningful test, and it's exactly the wrong test for what a washdown hose actually does to a sensor. A hand-held washdown gun delivers water at high velocity and pressure, directed at close range, often angled directly at seams, threads, and connector faces — conditions the IP67 immersion test never subjects the device to at all.

**IP69K** is a separate rating, defined by ISO 20653 (which absorbed the German DIN 40050-9 standard that originated the IP69K designation), specifically for high-pressure, high-temperature close-range spray: hot water (around 80°C) at high pressure, sprayed from multiple angles at close range (roughly 10-15 cm) with a rotating nozzle. A sensor rated IP69K has actually been tested against something resembling a real washdown gun; a sensor rated only IP67, no matter how well-built otherwise, has not.

```{r fig-06-14-ip-test-compare, echo=FALSE, fig.cap="IP67's immersion test uses essentially static hydrostatic pressure; IP69K's close-range hot spray test uses roughly three orders of magnitude more pressure. This is why an IP67-rated sensor can still fail in a real washdown even though the rating 'sounds' adequate.", fig.align='center', fig.width=9, fig.height=5.5}
ip_pressure_df <- data.frame(
  rating = factor(c("IP67\n(1 m static immersion)", "IP69K\n(close-range hot spray)"),
                   levels = c("IP67\n(1 m static immersion)", "IP69K\n(close-range hot spray)")),
  test_pressure_bar = c(0.098, 90)
)

ggplot(ip_pressure_df, aes(x = rating, y = test_pressure_bar, fill = rating)) +
  geom_col(show.legend = FALSE, width = 0.55) +
  geom_text(aes(label = sprintf("%.2g bar", test_pressure_bar)), vjust = -0.6,
            size = 3.6 * diagram_text_scale) +
  scale_fill_manual(values = c("#95A5A6", "#C0392B")) +
  scale_y_log10() +
  labs(title = "Approximate Test Water Pressure: IP67 vs. IP69K",
       subtitle = "Log scale -- IP69K's test pressure is roughly three orders of magnitude higher",
       x = NULL, y = "Approximate test pressure (bar, log scale)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 9.5 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale),
        axis.text = element_text(size = 10 * diagram_text_scale))
```

The seal being tested against the right pressure and temperature is only half the specification. A washdown-duty sensor also needs a **housing material** that survives the plant's cleaning chemistry (caustic and acid cleaners routinely used in food plants attack many plastics and coatings over time; 316 stainless steel is the standard washdown housing material for a reason), **seal materials** rated for the wash temperature (a standard nitrile O-ring can hardenand lose sealing force at repeated 80°C exposure; FKM/Viton compounds hold up far better), and a **connector** genuinely rated for the same spray conditions as the housing — a stainless M12 connector with a continuous, properly seated O-ring, not a cost-reduced plastic quick-disconnect that happens to carry an IP67 label on its own datasheet. A sensor body rated IP69K with a connector that isn't is only as good as its weakest sealed joint.

Sanitary mounting matters as much as the sensor itself in a federally registered food plant: fixtures should be designed to shed water and cleaning solution rather than trap it, avoiding horizontal ledges, dead-leg gaps, and unsealed threaded holes that harbor residue between wash cycles. In Canada, this falls under the Canadian Food Inspection Agency's requirements for federally registered establishments; the equivalent US and international references are the 3-A Sanitary Standards and NSF/ANSI 169, which many manufacturers design washdown-duty equipment to meet even outside a formal 3-A certification process.

**Rule of thumb:** in any zone that receives direct hand-held washdown spray (not just incidental splash), specify stainless steel housings, FKM seals, and IP69K-rated connectors as the default, and treat a plastic-body IP67 sensor as appropriate only for splash-zone or indirect-exposure locations well outside the direct spray path. This breaks down for genuinely dry, non-washdown areas of the same plant (an upstream dry-ingredient area, an electrical room) — over-specifying stainless/IP69K everywhere adds real cost for zones that never see a hose at all, so the wash zone boundary itself needs to be mapped on the layout drawing, not assumed.

### Governing Calculation

This is a checklist-driven decision, not a formula:

1. Is the mounting location within direct hand-held wash spray range (not just incidental splash)?
2. What is the actual wash pressure and water temperature used at this plant (get the real numbers from the sanitation department, not an assumption)?
3. Is the sensor's *housing* rated IP69K (not just IP67)?
4. Is the sensor's *connector* independently rated IP69K, or does it rely on the housing rating alone?
5. Are the seal materials rated for the wash temperature and the cleaning chemistry in use?
6. Is the housing material chemically compatible with the cleaning chemistry (stainless vs. plastic)?

A "no" on any of questions 3-6, combined with a "yes" on question 1, means the currently specified sensor is inadequate regardless of what its IP67 label suggests.

### Worked Example

Live cell — change any of the boolean inputs (e.g., set `wash_pressure_psi` lower, or `housing_material` to `"316 stainless"`) and re-run to see the recommendation change.

```{webr-r}
#| label: washdown-checklist-06-14
## Given: the failing photoelectric sensor's actual specification, and the
## plant's actual washdown conditions
in_direct_spray_zone <- TRUE
wash_pressure_psi <- 1500
wash_temp_C <- 80

housing_rating <- "IP67"
housing_material <- "Black polycarbonate"
connector_rating <- "IP67 (rated only as part of the overall housing claim)"
seal_material <- "Standard nitrile"

cat("Mounting location in direct hand-held wash spray path:", in_direct_spray_zone, "\n")
cat("Plant washdown conditions: ", wash_pressure_psi, "psi at", wash_temp_C, "C\n\n")

cat("Currently specified sensor:\n")
cat("  Housing rating:  ", housing_rating, "\n")
cat("  Housing material:", housing_material, "\n")
cat("  Connector rating:", connector_rating, "\n")
cat("  Seal material:   ", seal_material, "\n\n")

fails_checklist <- in_direct_spray_zone && (
  housing_rating != "IP69K" ||
  !grepl("stainless", housing_material, ignore.case = TRUE) ||
  !grepl("IP69K", connector_rating) ||
  !grepl("FKM|Viton", seal_material, ignore.case = TRUE)
)

if (fails_checklist) {
  cat("CONCLUSION: current specification FAILS the washdown checklist.\n\n")
  cat("Recommended replacement specification:\n")
  cat("  Housing rating:   IP69K (ISO 20653), tested at conditions meeting or\n")
  cat("                    exceeding this plant's actual wash pressure/temperature\n")
  cat("  Housing material: 316 stainless steel\n")
  cat("  Connector:        Stainless M12, independently IP69K-rated\n")
  cat("  Seals:            FKM/Viton, rated for continuous 80 C+ exposure\n")
} else {
  cat("CONCLUSION: current specification PASSES the washdown checklist.\n")
}
```

**Sanity check:** the plant's actual wash conditions (1500 psi, 80°C, direct spray) sit well inside what the IP69K test protocol is specifically designed to replicate, and nowhere near what the IP67 immersion test represents — the failing sensors in the scenario weren't defective, they were simply never tested against anything resembling their actual service environment. This is the recurring lesson of this design task: an IP rating is only meaningful in the context of the specific test it certifies, not as a generic "waterproof" label.

<details>
<summary>Industry connection: defense-sector environmental qualification</summary>

The same discipline of matching a component's actual certified test conditions against its actual service environment — rather than trusting a rating by reputation — shows up even more formally in defense manufacturing, where MIL-STD environmental qualification testing (shock, vibration, temperature cycling, and ingress protection) has to be traceable in the build record for audit purposes; a component substitution that looks equivalent on a datasheet summary but wasn't tested to the same specific standard is exactly the kind of gap a Controlled Goods Program audit is designed to catch.

</details>

## Design Task 7: Total I/O Count and Spare Capacity Planning

### Scenario

The automotive respot cell from Design Task 4 is now far enough along that the electrical designer has to actually size the I/O hardware — how many digital input, digital output, analog input, and analog output modules to put in the rack. Every station's sensors, clamps, and interlocks have to be tallied, a sensible spare capacity margin applied for the engineering changes that inevitably show up between now and the cell's first production run, and the total rounded to real, purchasable module sizes.

### Design Guidelines and Rules of Thumb

Tallying I/O sounds like simple bookkeeping, and mostly it is — walk every station, every sensor, every actuator, every interlock, and every operator device, and count discrete inputs (DI), discrete outputs (DO), analog inputs (AI), and analog outputs (AO) separately. The part that's easy to underestimate is how much gets added *after* the tally: a fixture redesign that needs one more clamp-position sensor, a customer requirement discovered late for an additional part-presence check, a controls engineer who wants a spare output for a future indicator light. None of that is hypothetical — it is the normal, expected pattern of how a machine's I/O count grows between the initial electrical design and the first production run, and building spare capacity in from the start is far cheaper than a rack redesign mid-project.

**Rule of thumb:** add 10-20% spare capacity on top of the tallied I/O count, then round up to the next standard I/O module or card size — never round down, and never treat the raw percentage as a target to hit exactly, since the module size itself is the real constraint. This breaks down for a very small point count in any single I/O type (a handful of analog points, say): rounding a small count up to even the smallest standard module size can add 50-100% spare almost by accident, which is fine — the goal is never buying a fractional module, and a small count simply can't help but land on a generous percentage once it's rounded to real hardware.

### Governing Calculation

$$N_{with\ spare} = N_{tallied} \times (1 + margin)$$

$$N_{modules} = \left\lceil \frac{N_{with\ spare}}{N_{points\ per\ module}} \right\rceil$$

$$N_{installed} = N_{modules} \times N_{points\ per\ module}$$

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $N_{tallied}$ | Raw tallied point count for one I/O type (DI, DO, AI, or AO) | points |
| $margin$ | Selected spare-capacity margin (commonly 10-20%) | dimensionless |
| $N_{with\ spare}$ | Tallied count plus spare margin | points |
| $N_{points\ per\ module}$ | Standard module/card size for this I/O type | points/module |
| $N_{modules}$ | Number of modules required, rounded up | modules |
| $N_{installed}$ | Actual installed point capacity after rounding to whole modules | points |

### Worked Example

Live cell — try different station counts, spare margins, or module sizes and see how the required hardware changes.

```{webr-r}
#| label: io-count-spare-06-14
## Given: I/O tally across the 5-station respot cell, plus cell-level shared
## I/O (E-stops, mode select, HMI, safety interlocks)
station_DI <- c(22, 24, 20, 26, 21)   # per-station discrete inputs (sensors, position switches)
station_DO <- c(14, 16, 13, 15, 14)   # per-station discrete outputs (clamps, solenoids, lights)
cell_level_DI <- 5                     # E-stops, mode select, gate switches
cell_level_DO <- 4                     # stack light, cell-level solenoids

DI_total <- sum(station_DI) + cell_level_DI
DO_total <- sum(station_DO) + cell_level_DO
AI_total <- 14   # weld current monitors, temperature probes
AO_total <- 4    # VFD speed references

cat("Tallied I/O counts:\n")
cat("  DI:", DI_total, "\n")
cat("  DO:", DO_total, "\n")
cat("  AI:", AI_total, "\n")
cat("  AO:", AO_total, "\n\n")

spare_margin <- 0.15   # 15%, within the 10-20% rule-of-thumb range

size_io <- function(tallied, margin, points_per_module, label) {
  with_spare <- tallied * (1 + margin)
  modules <- ceiling(with_spare / points_per_module)
  installed <- modules * points_per_module
  actual_spare_pct <- (installed - tallied) / tallied * 100
  cat(sprintf("%s: tallied = %d, +%.0f%% spare = %.1f, module size = %d pts\n",
              label, tallied, margin * 100, with_spare, points_per_module))
  cat(sprintf("   -> %d module(s), %d points installed (actual spare = %.1f%%)\n\n",
              modules, installed, actual_spare_pct))
  installed
}

DI_installed <- size_io(DI_total, spare_margin, 32, "DI  (32-point cards)")
DO_installed <- size_io(DO_total, spare_margin, 16, "DO  (16-point cards)")
AI_installed <- size_io(AI_total, spare_margin, 4, "AI  (4-channel cards)")
AO_installed <- size_io(AO_total, spare_margin, 4, "AO  (4-channel cards)")

cat("Total installed I/O capacity:", DI_installed + DO_installed + AI_installed + AO_installed, "points\n")
```

```{r fig-06-14-io-count-spare, echo=FALSE, fig.cap="Tallied I/O count vs. installed capacity after applying the spare-capacity margin and rounding to standard module sizes. Note how the small analog counts land on a much larger actual spare percentage -- module granularity, not the percentage target, dominates at low point counts.", fig.align='center', fig.width=10, fig.height=6}
io_summary_df <- data.frame(
  io_type = rep(c("DI", "DO", "AI", "AO"), 2),
  count = c(118, 76, 14, 4, 160, 96, 20, 8),
  kind = rep(c("Tallied (+ 15% spare need)", "Installed (rounded to module size)"), each = 4)
)
io_summary_df$io_type <- factor(io_summary_df$io_type, levels = c("DI", "DO", "AI", "AO"))
io_summary_df$kind <- factor(io_summary_df$kind,
                              levels = c("Tallied (+ 15% spare need)", "Installed (rounded to module size)"))

ggplot(io_summary_df, aes(x = io_type, y = count, fill = kind)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = format(count, big.mark = " ")), position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3.2 * diagram_text_scale) +
  scale_fill_manual(values = c("Tallied (+ 15% spare need)" = "#3498DB",
                                "Installed (rounded to module size)" = "#27AE60")) +
  labs(title = "I/O Count: Tallied Need vs. Installed Capacity",
       x = NULL, y = "I/O points", fill = NULL) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
        legend.position = "bottom",
        axis.title = element_text(size = 11 * diagram_text_scale),
        axis.text = element_text(size = 10 * diagram_text_scale))
```

**Sanity check:** every type lands at or above the 15% target, exactly as the rule of thumb requires (round up, never down). DI and DO, with the largest raw counts, still overshoot to 35.6% and 26.3% actual spare once rounded to whole modules — module granularity adds real headroom beyond the target even at these point counts. AI and AO, with far smaller raw counts, land at 43% and 100% actual spare respectively — not because the margin logic failed, but because a small count rounded up to even the smallest standard module (4 channels) can't help but land on a generous percentage. Buying a custom 1-channel module just to hit 15% exactly would be a false economy; the whole-module rule is the correct constraint to follow.

## Design Task 8: 4-20 mA Loop Voltage Budget

### Scenario

The food plant's temperature transmitter from Design Task 5 is being specified on an even longer run at a different location in the same facility: a level transmitter on a remote tank farm, 950 m of cable back to the central control room, wired as a 4-20 mA current loop and powered by a 24VDC loop power supply at the control room end. Before the cable is pulled, the instrumentation engineer needs to verify that the loop will actually deliver a valid signal at the maximum current (20 mA) — a current loop is immune to wire-resistance signal error, as established in Design Task 5, but it is not immune to running out of voltage.

### Design Guidelines and Rules of Thumb

Every element in a 4-20 mA current loop — the transmitter's own internal electronics, the wire running out and back, and the receiving device's burden resistor — drops some voltage at the maximum 20 mA operating current, and every one of those drops has to be supplied by the loop power supply. This is a straightforward series-circuit voltage budget, directly analogous to the voltage-drop calculations covered for power wiring in the [Wire, Cable, and Circuit Protection](06-04-Wire-Cable-Circuit-Protection.qmd) chapter — the same physics, applied to a much smaller current.

A loop-powered, two-wire transmitter needs a minimum voltage across its own two terminals — its **compliance voltage** requirement — to run its internal electronics and current-regulating circuit; this figure comes directly from the transmitter's datasheet and typically falls somewhere in the 10-16V range for a smart 4-20 mA transmitter, though it varies by model. The wire itself drops $I \times R_{wire}$, using the same round-trip resistance calculation as Design Task 5. The receiving device's **burden resistor** — the resistor a PLC analog input card uses internally to convert the loop current into a voltage it can actually measure — drops $I \times R_{burden}$; common burden resistor values run from roughly 100 Ω up to 500 Ω or more depending on the card and the desired voltage span.

**Rule of thumb:** run the voltage budget check at the loop's *worst-case low* supply voltage — typically the nominal supply voltage less about 10% for normal line-voltage tolerance on the power supply feeding the loop supply itself — not the nameplate nominal voltage, since a loop that only passes at a perfect 24.0V has no real margin at all. This breaks down (in the sense of being overly conservative) for a loop powered from a precision-regulated DC supply with a tightly specified output tolerance well under 10% — in that case, using the supply's actual specified tolerance instead of a flat 10% gives a more accurate, less pessimistic budget.

### Governing Calculation

$$V_{required} = V_{transmitter,min} + (I \times R_{wire}) + (I \times R_{burden})$$

$$R_{wire} = 2 \times r_{per\ m} \times L$$

Loop passes if $V_{supply,worst\ case} \geq V_{required}$ at $I = 20$ mA (the maximum, worst-case current for this check).

| Symbol | Definition | Units |
|:-------|:-----------|:------|
| $V_{transmitter,min}$ | Transmitter's minimum compliance voltage requirement, from its datasheet | V |
| $I$ | Loop current (checked at the maximum, 20 mA) | A |
| $R_{wire}$ | Round-trip wire resistance for the full cable run | Ω |
| $R_{burden}$ | Receiving device's (PLC analog input card's) burden resistance | Ω |
| $V_{required}$ | Total voltage the loop supply must deliver at maximum current | V |
| $V_{supply,worst\ case}$ | Loop power supply voltage, derated for its worst-case low tolerance | V |

### Worked Example

Live cell — try a different cable length, wire gauge, or burden resistor value and see whether the loop still passes.

```{webr-r}
#| label: loop-budget-06-14
## Given: 950 m run, 20 AWG twisted shielded instrumentation cable, 24VDC
## loop power supply
R_per_m_20awg <- 10.15 / 304.8   # ohm/m, one-way (20 AWG, typical table value)
L <- 950                          # m, one-way cable length
R_wire <- 2 * R_per_m_20awg * L   # ohm, round-trip

I_max <- 0.020        # A, worst-case (maximum) loop current
V_transmitter_min <- 12   # V, transmitter's datasheet compliance voltage requirement
R_burden <- 250        # ohm, PLC analog input card's burden resistance

V_supply_nominal <- 24     # V, loop power supply nameplate voltage
V_supply_worst <- V_supply_nominal * 0.90   # 10% worst-case low tolerance

V_wire_drop <- I_max * R_wire
V_burden_drop <- I_max * R_burden
V_required <- V_transmitter_min + V_wire_drop + V_burden_drop

cat("Cable length:", L, "m (one-way), 20 AWG twisted shielded pair\n")
cat("Round-trip wire resistance:", round(R_wire, 2), "ohm\n\n")

cat("Voltage budget at I_max =", I_max * 1000, "mA:\n")
cat("  Transmitter minimum compliance voltage:", V_transmitter_min, "V\n")
cat("  Wire drop (I x R_wire):                ", round(V_wire_drop, 2), "V\n")
cat("  Burden resistor drop (I x R_burden):   ", round(V_burden_drop, 2), "V\n")
cat("  Total required voltage:                ", round(V_required, 2), "V\n\n")

cat("Loop supply, nominal:                    ", V_supply_nominal, "V\n")
cat("Loop supply, worst-case (-10%):          ", round(V_supply_worst, 2), "V\n\n")

margin_V <- V_supply_worst - V_required
if (margin_V >= 0) {
  cat("Worst-case supply >= required voltage: PASS, margin =", round(margin_V, 2), "V\n")
} else {
  cat("Worst-case supply < required voltage: FAILS by", round(-margin_V, 2), "V\n")
}
```

**Sanity check:** the wire drop itself is small (about 1.3V over nearly a kilometer, at 20 mA) — current loops really are close to immune to run length in this sense — but the burden resistor's 5V drop and the transmitter's own 12V requirement add up to a real number, and the worst-case (derated) supply voltage clears it with only about 3.3V of margin. That margin is thin enough to be worth checking explicitly rather than assuming; a plant standard that swaps in a 500 Ω burden resistor (common where a card is configured for a wider internal voltage span) pushes the total required voltage past the worst-case supply and fails the check by about 1.7V — the corrective options at that point are the same categories used throughout this part: reduce the resistance (heavier wire gauge, per the [Wire, Cable, and Circuit Protection](06-04-Wire-Cable-Circuit-Protection.qmd) chapter), reduce the burden resistor value if the card allows it, or specify a higher-voltage loop power supply (28V loop supplies exist specifically for this reason).

## Summary

```{r tbl-06-14-summary, echo=FALSE, message=FALSE}
summary_table <- tribble(
  ~Design_Task, ~Governing_Relationship, ~Key_Rule_of_Thumb,
  "1. Sensor technology selection", "Decision framework: target material, range, and optical/acoustic reliability drive the technology choice",
  "Metal + short range -> default to inductive; everything else follows from the target's material and optical properties",
  "2. Sensing distance derating", "S_r,eff = S_n x Cf_material x Cf_size; usable = 0.8 x S_r,eff",
  "Never trust the catalog S_n for a non-reference material or undersized target -- always apply both corrections",
  "3. Sourcing (PNP) vs. sinking (NPN)", "Fault-mode walkthrough: does a wire-to-ground fault mimic the ACTIVE or INACTIVE state?",
  "Default to PNP + sinking input cards in North America -- a ground fault reads as OFF, not a false detection",
  "4. IO-Link / network vs. hardwired I/O", "Weighted decision matrix across diagnostics, wiring cost, commissioning speed, scalability",
  "Weight diagnostics and scalability heavily on large, uptime-critical cells; small machines often don't need more than hardwired",
  "5. 4-20 mA vs. 0-10V analog signal", "V_error% = R_wire / (R_wire + R_in) x 100%; current loop error = 0% regardless of length",
  "Default to 4-20 mA for any run beyond ~15-30 m or sharing a tray with power wiring",
  "6. IP ratings / washdown selection", "Checklist: direct spray zone + housing/connector/seal rating all IP69K, not just IP67",
  "IP67 means static immersion, not high-pressure hot spray -- specify IP69K + stainless + FKM in direct wash zones",
  "7. Total I/O count and spare capacity", "N_installed = ceiling(N_tallied x (1+margin) / pts_per_module) x pts_per_module",
  "Add 10-20% spare, then round UP to the next whole module -- never round down, never chase the percentage exactly",
  "8. 4-20 mA loop voltage budget", "V_required = V_transmitter,min + I x R_wire + I x R_burden, checked against worst-case supply",
  "Check the budget at worst-case low supply voltage (-10%), not nominal -- a loop that only just passes at nominal has no real margin"
)

summary_table %>%
  kbl(caption = "Design task summary: governing relationship and key rule of thumb",
      col.names = c("Design Task", "Governing Relationship", "Key Rule of Thumb")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, background = "#2C3E50", color = "white")
```

## Practice Problems

<details>
<summary>Task 1 (plug-in): A seventh sensing point is added to the blanking line: detecting the presence of a clear polycarbonate splash guard panel that must be in place before the press can cycle, at a range of about 200 mm. Using the decision framework, which technology should be specified, and why?</summary>

The target is not metal, so inductive is ruled out immediately. It doesn't need to be sensed through a barrier, so capacitive isn't the natural fit either. The range (200 mm) rules out neither photoelectric nor ultrasonic on its own, but the target's optical property — clear polycarbonate — is exactly the unreliable-optics case the framework flags: a diffuse photoelectric sensor is likely to see straight through a clear panel rather than reflect off it. **Ultrasonic** is the correct recommendation, since sound reflects off the panel's surface regardless of its optical transparency.

</details>

<details>
<summary>Task 1 (design decision): A supplier pitches a single "universal" capacitive sensor to replace all six sensing points in the worked example, arguing it can be tuned to detect any of the targets and would simplify the spares inventory to one part number. Would you accept this proposal? What would you check first?</summary>

Simplifying spares is a real, legitimate benefit, but capacitive sensing is the least selective technology on the list precisely because it responds to almost any change in dielectric constant near the sensing face — a capacitive sensor tuned to reliably detect a plastic pallet in a busy area is also likely to weakly respond to a nearby metal bracket, an operator's hand, or condensation buildup, producing false triggers that a more selective technology (inductive for the metal target, for instance) wouldn't. It's also the wrong physical principle for two of the six points outright — the drive shaft feedback and the press force check aren't presence sensors at all, and no capacitive sensor replaces an encoder or a load cell. Before accepting the proposal, check: does capacitive sensing genuinely have the selectivity margin needed at each of the four presence-sensing points (not just "can it be tuned to work in the shop"), and does the supplier's pitch even attempt to address the encoder and load cell points, or did it quietly drop them from the comparison? A one-part-number spares strategy across the *four* comparable presence-sensing points might be worth evaluating on its own merits; replacing all six with one technology is not a serious proposal once the actual measured quantities are considered.

</details>

<details>
<summary>Task 2 (plug-in): An M18 shielded inductive sensor is rated Sn = 8 mm for the mild steel reference target (18 x 18 mm). It needs to detect an aluminum target measuring 14 x 14 mm at a mounting gap of 2.0 mm. Does this pass?</summary>

Cf_material (aluminum) = 0.4

Cf_size = min(1, 14/18) = 0.778

Sr_eff = 8 x 0.4 x 0.778 = 2.49 mm

Sr_usable = 0.8 x 2.49 = 1.99 mm

Mounting gap (2.0 mm) is essentially equal to the usable range (1.99 mm) — this is a marginal **FAIL** (or, at best, a pass with zero margin, which should be treated as a fail for a production application). The gap should be reduced, or a larger sensor selected.

</details>

<details>
<summary>Task 2 (design decision): The fixture designer proposes solving the Design Task 2 worked example's failure by welding a small mild steel "flag" tab onto the stainless steel shift fork, sized to fill the sensor's full 12 x 12 mm reference target, rather than changing the sensor or the mounting gap. Is this a reasonable fix? What would you want to verify before accepting it?</summary>

This is a genuinely common and often excellent fix — it directly attacks both derating factors at once: the flag is mild steel (Cf_material = 1.0 instead of 0.7) and sized to the full reference target (Cf_size = 1.0 instead of 0.67), which would restore the sensor's full 4 mm rated distance (usable = 0.8 x 4 = 3.2 mm), comfortably clearing the 3.5 mm gap... except it doesn't quite, which is exactly the kind of check that has to actually be run rather than assumed — 3.2 mm is still less than the 3.5 mm gap, so even the "ideal" flag target doesn't fully pass without also tightening the gap slightly. Beyond the arithmetic, real concerns to verify before accepting a welded-on flag: does adding a welded tab affect the shift fork's fatigue life or balance in a moving assembly, is the weld process and location compatible with the part's existing heat-treatment and finish requirements, and is the tab's position repeatable enough shot-to-shot that the 3.5 mm gap assumption actually holds in production, not just on the first sample part. A fix that solves the sensing problem but introduces a part-quality or fatigue risk isn't actually a good fix.

</details>

<details>
<summary>Task 3 (plug-in): A wire carrying an NPN sensor's switched output instead chafes and shorts to the +24V supply rail (not the grounded frame). What state does this fault force the wire to, and does it mimic the sensor's active or inactive state?</summary>

An NPN sensor's output wire normally sits near +24V (through the load) when inactive, and is pulled to ~0V by the sensor's transistor when active. A short to +24V forces the wire to +24V regardless of the sensor's actual state — which matches the wire's normal *inactive* voltage, not its active one. This particular fault (wire-to-supply, on an NPN sensor) is the safer direction, mirroring the reasoning in the chapter for PNP wire-to-ground faults: it happens to be the less common real-world fault (most fixture wiring failures involve chafing against a grounded metal frame, not against another live conductor), which is part of why PNP-to-ground is the fault scenario this chapter emphasizes.

</details>

<details>
<summary>Task 3 (design decision): A machine builder proposes standardizing on NPN sensors for a new line because the customer's existing PLC fleet across the plant uses sourcing-type input cards exclusively, and switching to PNP sensors would mean stocking a second, incompatible spare-sensor inventory. Is this a reasonable exception to the PNP default, and what would you verify?</summary>

Yes, this is a legitimate example of the kind of exception the chapter's rule of thumb explicitly allows for — a real supplier/fleet constraint, not a preference. Forcing PNP sensors onto sourcing-type input cards either doesn't work electrically at all or requires interposing relays/converters that add cost and failure points of their own, so matching the sensor output type to the existing input card architecture is the practical choice. What's worth verifying before accepting it wholesale: is this true plant-wide, or only for a subset of the fleet (mixing conventions across a plant is its own long-term maintenance hazard, and it's worth checking whether the "existing fleet" justification is actually as universal as claimed); and separately, for any safety-rated inputs on the new line, the safety circuit's own fault-mode analysis (per the Machine Safety Circuits chapter) governs regardless of this general convention — the PNP/NPN default is about general-purpose I/O, not a substitute for a documented safety function analysis.

</details>

<details>
<summary>Task 4 (plug-in): Using the weighted decision matrix with weights diagnostics = 0.15, wiring_cost = 0.35, commissioning_speed = 0.30, scalability = 0.20 (a small, standalone, cost-sensitive machine's priorities), which architecture wins?</summary>

Hardwired = 0.15(1) + 0.35(2) + 0.30(3) + 0.20(2) = 0.15 + 0.70 + 0.90 + 0.40 = 2.15

IO-Link = 0.15(4) + 0.35(4) + 0.30(4) + 0.20(3) = 0.60 + 1.40 + 1.20 + 0.60 = 3.80

Fieldbus/Ethernet = 0.15(5) + 0.35(3) + 0.30(3) + 0.20(5) = 0.75 + 1.05 + 0.90 + 1.00 = 3.70

**IO-Link wins** (3.80 vs. 3.70 vs. 2.15) — for a small, cost-sensitive machine, IO-Link's combination of reasonable wiring cost and strong commissioning speed edges out fieldbus/Ethernet's diagnostics and scalability advantage, which matters less on a standalone machine than on a large multi-station cell.

</details>

<details>
<summary>Task 4 (design decision): A plant standard mandates fieldbus/industrial Ethernet I/O on every new machine, regardless of size, to keep the entire plant on one diagnostic and spare-parts platform. A small single-station deburring machine, with only 18 total I/O points, is being built under this standard. Is the plant standard the right call here, and what's the trade-off being accepted?</summary>

A plant-wide standard trades away per-machine optimization for fleet-wide consistency — and that trade-off is often worth it deliberately, not just a mistake: a single diagnostic platform, a single spare-parts strategy for network components, and controls technicians who only need to be trained on one architecture across the whole plant are real, ongoing operational benefits that a machine-by-machine "optimal" choice doesn't capture. For an 18-point machine specifically, the weighted-matrix comparison in the plug-in problem above would recommend IO-Link or even hardwired on a standalone basis — so the plant standard is accepting a real, quantifiable cost premium (network hardware and commissioning overhead for a machine that doesn't strictly need it) in exchange for fleet-wide consistency. Whether that's the right call depends on factors outside this one machine: how many machines the plant is adding per year (amortizing the standard's overhead), how much the plant already has invested in fieldbus/Ethernet diagnostic tooling and trained staff, and whether a documented exception process exists for machines small enough that the standard's overhead is genuinely disproportionate. The engineer's job here isn't to override the plant standard unilaterally based on one machine's numbers — it's to flag the trade-off explicitly if the premium seems large, and let whoever owns the standard make an informed call.

</details>

<details>
<summary>Task 5 (plug-in): The same 180 m, 18 AWG run is re-terminated so the 0-10V signal feeds only the PLC's analog input, at its full 100 kohm input impedance (no parallel-loaded panel meter). What is the new signal error, and does it now pass a 1% accuracy spec?</summary>

R_wire = 7.54 ohm (unchanged, same cable)

V_error% = 7.54 / (7.54 + 100 000) x 100 ≈ 0.0075%

This comfortably **passes** the 1% accuracy spec — removing the parallel-loaded panel meter and relying on the PLC's high native input impedance essentially eliminates the wire-resistance error term. (The ground-reference offset error discussed in the chapter is a separate effect not captured by this calculation, and would need to be checked independently before concluding the 0-10V signal is fully equivalent to a current loop on this run.)

</details>

<details>
<summary>Task 5 (design decision): An instrumentation supplier argues that, given the previous plug-in problem's result, 0-10V signals are perfectly fine for long runs as long as the receiving device has high input impedance, and the plant should save money by standardizing on 0-10V transmitters everywhere. How would you respond?</summary>

The plug-in problem does show that a high-impedance receiver essentially eliminates the wire-resistance-divider error term — that part of the supplier's argument is correct as far as it goes. But it only addresses one of the two reasons current loops dominate industrial practice, discussed in the chapter: the wire-resistance error, and the ground-reference/noise-immunity advantage. A voltage signal's reading still depends on the transmitter and receiver sharing a common 0V reference, and over a long run in a real plant, with equipment bonded to earth at different physical points, that reference can differ by a real, non-zero voltage that adds directly to the signal regardless of how high the receiver's input impedance is. High input impedance also does nothing to prevent someone, at some point in the equipment's life, from tee-ing the signal to a second device (a local meter, a data logger) and unknowingly reintroducing the exact loading error the plug-in problem eliminated — a current loop tolerates a second series-connected receiver without that risk, as long as the added burden resistance still fits the voltage budget from Design Task 8. The right response is to agree that impedance matters and is worth specifying carefully, but not to accept "impedance is high enough" as a full substitute for the current loop's structural advantages on a genuinely long or electrically noisy run — especially in a food or industrial plant with variable-frequency drives and other noise sources sharing cable trays nearby.

</details>

<details>
<summary>Task 6 (plug-in): A sensor is mounted on a conveyor in the same poultry plant, but in a dry storage area that never receives hose-down spray, only occasional airborne dust. Using the checklist, does this location need an IP69K, stainless, FKM-sealed sensor?</summary>

Working through the checklist: is the mounting location within direct hand-held wash spray range? No — it's a dry storage area with no wash exposure, only dust. Since question 1 (direct spray zone) is the gating condition for the rest of the checklist, the answer is **no**, a standard IP67 (or even a lower-rated, dust-focused IP6X) plastic-body sensor is appropriate here, and specifying a stainless IP69K sensor for this location would be needless cost. It's worth double-checking with the plant's sanitation team that this area truly never receives wash spray or aggressive cleaning chemistry (some "dry" areas get an occasional deep-clean wash that a quick site walk might miss) before finalizing that conclusion.

</details>

<details>
<summary>Task 6 (design decision): The plant's maintenance manager wants to specify IP69K, stainless, FKM-sealed sensors for every sensor on site, dry areas included, to simplify the spares inventory to one specification. What would you tell them?</summary>

This is a legitimate simplification strategy — the same reasoning as the "universal sensor" spares argument in Task 1 — and unlike that proposal, this one doesn't cross technology types, so it doesn't introduce a functional risk, only a cost one. The trade-off is real and worth quantifying rather than dismissing: washdown-duty stainless/IP69K sensors typically cost meaningfully more per unit than an equivalent plastic IP67 sensor, and across a plant with many dry-area sensors that never see a hose, the aggregate premium could be substantial. Whether it's worth it depends on the plant's actual numbers: how many sensor part numbers and how much inventory value is at stake, how often a wrong-spec sensor gets installed in the wrong zone today (if misapplication is a recurring real problem, the simplification has a genuine reliability payoff, not just a convenience one), and whether the wash-zone boundary on the layout drawings is reliable enough that "dry area" assignments won't quietly become wash zones later as the plant's process changes. A middle path — standardizing on IP69K/stainless for anything within, say, 5 m of a known wash zone boundary, with a documented dry-area exception list — often captures most of the simplification benefit without paying the premium on sensors that will genuinely never see a wash gun.

</details>

<details>
<summary>Task 7 (plug-in): A sixth station is added to the respot cell, adding 24 DI and 15 DO. Using the same 15% spare margin and the same module sizes (32-pt DI, 16-pt DO) as the worked example, what is the new required DI and DO module count?</summary>

New DI tallied = 118 + 24 = 142. With 15% spare: 142 x 1.15 = 163.3. Modules = ceiling(163.3 / 32) = 6. Installed = 6 x 32 = 192 points.

New DO tallied = 76 + 15 = 91. With 15% spare: 91 x 1.15 = 104.65. Modules = ceiling(104.65 / 16) = 7. Installed = 7 x 16 = 112 points.

DI needs **6 modules** (up from 5 in the original worked example — the extra station pushed the count past another module boundary), and DO needs **7 modules** (up from 6).

</details>

<details>
<summary>Task 7 (design decision): The project's capital budget is fixed, and the extra I/O module required by adding the sixth station (from the plug-in problem above) would push the electrical bill of materials over budget by an amount the customer won't approve. The project manager suggests dropping the spare margin to 0% for this cell only, to fit the additional station's I/O into the existing module count. Would you agree to this?</summary>

Dropping the spare margin to exactly 0% recovers budget today at the cost of removing the entire cushion the margin exists to provide — the first engineering change order after the cell is running (a near-certainty on any real installation, as the chapter notes) would then require a rack redesign, a shutdown window, and very likely cost far more than the module the team is trying to avoid buying now. Before agreeing, it's worth checking a few things that might resolve the conflict without eliminating the margin entirely: is the module count genuinely at a hard boundary where a small reduction (say, 5-8% margin instead of 0%) still fits without tripping the next module threshold; are there any points in the tally that were double-counted or are genuinely optional (a "nice to have" indicator light) that could be trimmed from the base count rather than the margin; and does the cost delta between the current design and one more module actually matter at the scale of the whole project, or is it a rounding error being treated as a hard line. If none of that closes the gap, the right move is to present the trade-off explicitly to whoever owns the budget decision — "zero spare capacity means the next change order requires a shutdown and a rack rebuild" is a business risk decision, not a purely technical one, and it shouldn't be made silently by quietly zeroing out the margin in the panel design.

</details>

<details>
<summary>Task 8 (plug-in): The same 950 m, 20 AWG loop is re-run with 16 AWG cable instead (r_per_m = 4.016/304.8 ohm/m one-way), keeping everything else the same (250 ohm burden, 12V transmitter minimum, 24V supply at -10%). Does the loop pass, and with how much margin?</summary>

R_wire = 2 x (4.016/304.8) x 950 ≈ 25.03 ohm

V_wire_drop = 0.020 x 25.03 ≈ 0.501 V

V_required = 12 + 0.501 + (0.020 x 250) = 12 + 0.501 + 5.0 = 17.50 V

V_supply_worst = 24 x 0.90 = 21.6 V

Margin = 21.6 - 17.50 = 4.10 V — the loop **PASSES**, with more margin than the original 20 AWG case (4.10 V vs. 3.33 V), since the heavier gauge wire drops less voltage over the same distance.

</details>

<details>
<summary>Task 8 (design decision): Facing the failing 500 ohm burden resistor scenario from the worked example (fails by about 1.7V), the instrumentation team has three options: switch to 16 AWG cable (cost: more copper, a heavier and more expensive cable pull), reduce the burden resistor to 250 ohm if the PLC card allows it (cost: possibly none, if the card supports a jumper-selectable burden), or specify a 28V loop power supply instead of 24V (cost: a different, slightly more expensive power supply, but a one-time panel-level cost regardless of how many loops it feeds). Which would you recommend, and why?</summary>

Reducing the burden resistor to 250 Ω, if the specific PLC analog input card supports it, is very likely the lowest-cost and lowest-risk fix — from the Design Task 8 worked example, that change alone closes the gap with room to spare (250 Ω passes with 3.33 V of margin at the original 20 AWG gauge), and if it's a jumper or software-configurable setting on the existing card, it may cost nothing at all beyond the engineering time to verify and document the change. Switching to 16 AWG cable also works (as shown in the plug-in problem) but means re-pulling nearly a kilometer of cable — a real cost and schedule item that should only be chosen if the burden-resistor option isn't available on this particular card. Upgrading to a 28V loop supply is the most broadly useful fix if the plant has *multiple* long or marginal loops sharing the same rationale, since one power supply upgrade benefits every loop it feeds, not just this one — but it's the most expensive single option if this is the only loop with a problem. The recommended order is: check the burden resistor option first (cheapest, fastest, no cable work), fall back to the cable upgrade if the card is fixed at 500 Ω, and reserve the power supply upgrade for a situation where several loops across the plant share the same marginal-budget problem, since that's where its higher cost is amortized across the most value.

</details>

## Open-Ended Design Scenario

<details>
<summary>Capstone scenario: specify the sensor and I/O architecture for a new line</summary>

You are the controls engineer responsible for releasing the electrical design of a new packaging line at a mid-sized Ontario food processing plant. Pull together everything from this chapter into one design package:

- The line needs six sensing points specified: a steel conveyor-guide-rail alignment check (short range, metal target), a case-present check on clear-wrapped shrink-bundled product moving on a roller conveyor, a liquid product level check through the wall of a sealed plastic day-tank, a long-range (5 m) rack-clear check across an aisle used by a forklift, speed/position feedback on the main line drive shaft, and a fill-head force/overload check on the capping station.
- The steel guide-rail alignment sensor is an M12 inductive sensor rated Sn = 4 mm for the mild steel reference target (12 x 12 mm reference size). The actual rail edge presents an effective 10 x 10 mm target face at a planned 3.0 mm mounting gap.
- All new discrete sensors on this line are being specified as PNP output, paired with sinking-type PLC input cards, per the plant's existing standard — except the capping station's fill-head sensors, which a supplier has proposed as NPN to match a sensor package they already stock. The fill-head area has a cable track prone to chafing against the machine's grounded frame.
- The plant is deciding between hardwired I/O and IO-Link for the line's roughly 60 total I/O points, spread across four stations with no existing fieldbus infrastructure in this part of the plant.
- The fill-head force sensor's output is being specified as either 0-10V or 4-20 mA, run 65 m back to the control panel, sharing a cable tray with a nearby VFD's power leads for part of that run.
- The capping station sits directly in the path of the plant's daily hot caustic washdown, at pressures and temperatures similar to those described in Design Task 6.
- A full I/O tally across all four stations, plus cell-level shared I/O, comes to 58 DI, 31 DO, 9 AI, and 2 AO.
- The fill-head force sensor's 4-20 mA loop (once selected) needs a voltage budget check: the transmitter requires a minimum of 11V compliance voltage, the PLC card's burden resistor is 250 Ω, and the loop is powered from a 24VDC supply.

For each of the eight design tasks in this chapter, work through the decision or calculation and reach a clear conclusion. Then write two or three sentences tying your conclusions together: does the washdown environment at the capping station (Design Task 6) change anything about the fill-head sensor's wiring choice (Design Task 3) or its analog signal specification (Design Task 5), beyond what each task's calculation shows in isolation? A strong answer treats this as one connected design, the way a real electrical release package has to, not eight independent homework problems.

</details>
