---
engine: knitr
filters:
  - webr
webr:
  show-startup-message: true
  cell-options:
    autorun: true
---

# Conveyor Design

```{r setup-06-10, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")
# jQuery is needed by kableExtra's kePrint.js (tooltips/popovers) but Quarto's
# Bootstrap 5 theme doesn't ship it by default.
inject_jquery()
library(tidyr)
```

## Learning Objectives

By the end of this chapter, you will be able to:

1. **Calculate** the effective (net driving) belt tension for an inclined conveyor using a CEMA-style breakdown of belt/idler friction, load friction, and gravitational (elevation) resistance.
2. **Convert** belt pull and belt speed into required drive power and torque at the head pulley shaft, and **translate** that torque/speed pair into a ready-made input for motor and gearbox sizing.
3. **Determine** the minimum belt speed required to meet a specified throughput target, and **apply** an appropriate design margin for surge capacity.
4. **Evaluate** roller, belt, and tabletop chain conveyor technologies against load type, sanitation, incline, and cost criteria using a documented decision framework.
5. **Apply** CFIA-aligned sanitary design principles — open-frame construction, ledge-free surfaces, accessible fasteners, IP-rated components — to a food-industry conveyor specification.
6. **Select** the appropriate frame material (stainless vs. painted/mild steel) for a given conveyor zone's product-contact and washdown-frequency classification.
7. **Verify** each conveyor design decision against a realistic sanity check or industry benchmark before it goes into a specification package.

## Introduction

Every plant with more than one floor, or more than one process rate, eventually needs to move product somewhere a person or a forklift can't conveniently take it — and more often than not, that job falls to a conveyor. It looks like the simplest machine on the floor: a loop of belt (or a line of rollers, or a chain with plastic slats) wrapped around two pulleys, driven by a gearmotor bolted to one end. Simple to describe, but not simple to size. Undersize the drive and the belt stalls under load, trips a motor overload, and stops the line it feeds. Oversize it and the customer pays for a gearmotor two frame sizes bigger than the job needs, on every one of the dozen conveyors in the layout — a real, recurring cost in a competitive quote.

Picture a regional grocery distribution center in Milton, Ontario. Product arrives at ground-floor receiving in mixed pallets, gets broken down into totes, and those totes need to climb roughly 4.3 m up to a second-floor case-pick mezzanine before an order picker ever touches them. An inclined belt conveyor does that climb. Before it can be ordered, someone has to answer a chain of questions in a specific order: how hard does the belt have to pull to get a loaded tote up that incline against gravity and friction? Given that pull and the belt's running speed, how much power and torque does the drive actually need to deliver, and at what shaft speed? How fast does the belt need to run in the first place, given the number of totes per minute the mezzanine picking operation requires? And once the mechanical sizing is settled, what is this conveyor actually built out of — because a conveyor carrying sealed dry-goods totes through a dry mezzanine stairwell is a completely different construction problem than one carrying raw product through a daily hose-down zone two floors below it.

This chapter works through that chain in five design tasks, all built around one running example: an inclined belt conveyor lifting totes between floors in a food distribution center. Design Tasks 1 through 3 size that conveyor's belt pull, drive power and torque, and belt speed, in that order — building the same kind of CEMA-style force-balance and drive-sizing method already introduced in simplified form in the [Motor, Drive, and Gearbox Sizing](06-06-Motor-Drive-Gearbox-Sizing.qmd) chapter's decline-conveyor example, but developed properly here rather than approximated. The torque and speed this chapter calculates at the head pulley shaft are exactly the inputs that chapter's load-reflection method (its Design Task 1) needs to carry the sizing the rest of the way to a selected motor and gearbox. Design Task 4 steps back from any one conveyor to ask a broader question every layout engineer faces early in a project: roller, belt, or tabletop chain — which conveyor technology actually belongs at a given point in the line? Design Task 5 closes the chapter on the question that makes food-plant conveyor engineering genuinely different from a generic industrial application: what does this thing need to be built out of, and how, to survive a hose-down and pass a CFIA-aligned sanitation audit.

**Conveyor Equipment Manufacturers Association (CEMA)** is the standard industry reference behind essentially every belt-tension calculation method in North American practice, published in its handbook *Belt Conveyors for Bulk Material Handling*. CEMA is a US-based, ANSI-recognized organization; there is no separate Canadian belt-conveyor design standard, and CEMA's method is used directly in Canadian mechanical design practice, sometimes cross-referenced against **ISO 5048** (continuous mechanical handling equipment — calculation of operating power and tensile forces for belt conveyors), the closest international/European equivalent. As with the gear-rating standards in the Motor, Drive, and Gearbox Sizing chapter, treat every CEMA citation in this chapter as a pointer to the source document — the friction-factor tables in the actual CEMA handbook are detailed, condition-specific, and are the association's intellectual property; this chapter describes the underlying method and uses representative, illustrative friction factors, not values copied from the handbook.

## Design Task 1: Belt Pull from Material Load, Incline, and Friction

### Scenario

At the Milton distribution center, an inclined belt conveyor carries totes of repacked dry goods from the ground-floor receiving area up to the second-floor case-pick mezzanine. The conveyor runs at a 22° incline over a belt-path length of 11.5 m, climbing a total of 4.31 m in elevation — a typical floor-to-floor height in this class of facility. The belt is 600 mm wide, supported on a roller-bed frame (idlers on anti-friction bearings, not a plain slider bed), and runs at 0.5 m/s — a speed the process engineer specified based on the mezzanine's required throughput, which Design Task 3 derives properly later in this chapter.

Totes average 18 kg loaded and, at the conveyor's peak design loading, sit spaced 0.5 m apart center-to-center along the belt. The belt and its rotating idlers/pulleys represent an equivalent moving mass of 14 kg per linear meter of conveyor, a figure taken from the equipment manufacturer's datasheet for a belt and idler set of this width and duty. Before anyone can select a gearmotor, the mechanical designer needs a single number: how hard does the drive have to pull the belt to keep it moving, fully loaded, up the incline?

```{r fig-06-10-incline, echo=FALSE, fig.cap="Inclined tote conveyor force balance: totes climb the belt against a combination of gravity, belt/idler friction, and load friction, all of which the drive's effective tension must overcome. Schematic, not to scale.", fig.align='center', fig.width=11, fig.height=6.5}
library(ggplot2)

theta_deg <- 22
theta <- theta_deg * pi / 180
L_draw <- 10  # illustrative drawing length, not the literal 11.5 m path length
rise_draw <- L_draw * sin(theta)
run_draw  <- L_draw * cos(theta)

tote_d <- seq(1.5, 8.5, by = 1.75)
totes_df <- data.frame(x = tote_d * cos(theta), y = tote_d * sin(theta))

ggplot() +
  annotate("rect", xmin = -1.5, xmax = 1.2, ymin = -0.5, ymax = 0, fill = "grey70", color = "black") +
  annotate("text", x = -0.2, y = -0.9, label = "Ground floor\n(receiving)", size = 2.8 * diagram_text_scale) +
  annotate("rect", xmin = run_draw - 1.2, xmax = run_draw + 1.5, ymin = rise_draw, ymax = rise_draw + 0.5,
           fill = "grey70", color = "black") +
  annotate("text", x = run_draw + 0.15, y = rise_draw + 0.9, label = "2nd floor\n(case-pick mezzanine)",
           size = 2.8 * diagram_text_scale) +
  annotate("segment", x = 0, y = 0, xend = run_draw, yend = rise_draw, linewidth = 2, color = "grey30") +
  geom_point(data = totes_df, aes(x = x, y = y), shape = 22, size = 5, fill = "#F39C12", color = "black") +
  ## Friction/gravity arrow (down-slope, resisting), with its label offset
  ## perpendicular to the belt line so it clears both the belt and the totes
  annotate("segment", x = 3.0*cos(theta), y = 3.0*sin(theta),
           xend = 1.8*cos(theta), yend = 1.8*sin(theta),
           arrow = arrow(length = unit(0.22, "cm"), type = "closed"), color = "#E74C3C", linewidth = 1.1) +
  annotate("text", x = 2.4*cos(theta) + 1.5*sin(theta), y = 2.4*sin(theta) - 1.5*cos(theta),
           label = "Gravity + friction\n(resist motion, T_x + T_y + T_z)", size = 2.8 * diagram_text_scale, color = "#E74C3C") +
  ## Effective tension arrow (up-slope, required), label offset to the other side
  annotate("segment", x = 6.5*cos(theta), y = 6.5*sin(theta),
           xend = 7.7*cos(theta), yend = 7.7*sin(theta),
           arrow = arrow(length = unit(0.22, "cm"), type = "closed"), color = "#2ECC71", linewidth = 1.1) +
  annotate("text", x = 7.1*cos(theta) - 1.6*sin(theta), y = 7.1*sin(theta) + 1.6*cos(theta),
           label = "Effective tension T_e\n(required, up-slope)", size = 2.8 * diagram_text_scale, color = "#2ECC71") +
  annotate("text", x = 2, y = -1.6, label = paste0(theta_deg, "° incline"), size = 3.3 * diagram_text_scale) +
  coord_fixed() +
  xlim(-2.5, run_draw + 3) + ylim(-2, rise_draw + 2) +
  theme_void() +
  labs(title = "Inclined Tote Conveyor: Force Balance",
       subtitle = "22-degree incline, roller-bed belt conveyor, 0.5 m/s design speed") +
  theme(plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 9.5 * diagram_text_scale))
```

### Design Guidelines and Rules of Thumb

CEMA's method (and every belt-tension method descended from it) frames the drive's job as overcoming an **effective tension** $T_e$: the net pulling force the head pulley must deliver to the belt, running at steady speed, to balance every resistance acting against it. The condensed, practitioner-level version of the method used in this chapter breaks $T_e$ into three physically distinct terms, each with its own name in CEMA's own notation:

- **$T_x$ — friction of the belt and idlers.** Even an empty belt takes force to keep moving: idler bearing drag, belt flexure over each idler, and pulley bearing friction. This term scales with the *belt and idler system's own weight* and a friction factor representing how freely that hardware turns.
- **$T_y$ — friction of the load.** The totes riding on the belt add weight, which increases the normal force on every idler beneath them, which increases friction proportionally. This term scales with the *load's weight* on the belt and a friction factor for how that additional loading translates into drag.
- **$T_z$ — gravity (elevation) resistance.** Lifting the load against gravity costs real energy, proportional to the load's weight and the sine of the incline angle.

**A genuinely important, easy-to-miss point:** $T_z$ uses the weight of the *material* on the belt, not the belt's own weight. The reason is geometric, not approximate: on a single, constant-slope conveyor, the belt's carrying (top) strand climbs the full elevation change, but the return (bottom) strand — running back down to the tail pulley — descends the same elevation. Over one complete loop of belt, the potential-energy gain going up is paid back going down, and the net elevation-related tension contribution of the belt's own weight cancels to zero. The belt's weight still matters — it's exactly what drives $T_x$, since it presses down on every idler along both strands — but it does **not** contribute to the gravity term. Only the load, which is placed on the belt at the bottom and removed at the top rather than travelling in a closed loop, actually needs to be lifted net of any elevation given back.

This is a genuine refinement worth calling out explicitly: the simplified decline-conveyor force balance in the Motor, Drive, and Gearbox Sizing chapter lumped belt weight into the gravity term along with the product weight, for the sake of a quick check of whether that conveyor's load overhauled its motor. That simplification is fine for a rough go/no-go check, but it is not the properly separated CEMA-style method — this chapter's three-term breakdown is the one to use for an actual belt-pull specification that a gearmotor purchase order will be based on.

**Rule of thumb:** a smooth-bottomed carton or tote on a plain (uncleated) belt starts to slide backward relative to the belt above roughly 18-20° of incline, depending on the carton material and belt surface — consistent with the same threshold noted for declines in the Motor, Drive, and Gearbox Sizing chapter. This conveyor's 22° incline is above that threshold, which is why it's specified with a friction-top (rough-top) belt surface rather than a plain smooth belt — the higher-friction surface keeps the tote from creeping relative to the belt as it climbs. This breaks down for cartons or totes with an inherently high-friction bottom (textured plastic, rubber feet) or a very short, low-mass load, both of which can tolerate steeper inclines on a plain belt than the rule of thumb suggests; conversely, a wet, greasy, or very smooth-bottomed load may slip well below 18°, and a real design should confirm the specific load/belt friction pairing rather than trust the rule blindly.

**Rule of thumb:** for a well-maintained, anti-friction (ball- or roller-bearing) idler system, both $T_x$'s and $T_y$'s friction factors typically fall in the 0.02-0.05 range; this chapter uses $C_b = 0.030$ for the belt/idler term and $C_m = 0.035$ for the load term, reflecting that loaded idlers commonly run slightly higher resistance than unloaded ones (increased seal drag and alignment sensitivity under load). This breaks down badly for a slider-bed conveyor (belt riding directly on a steel or UHMW pan rather than rollers), where the effective friction factor is far higher — commonly 0.3-0.5 or more — and for idlers running contaminated, iced, or poorly lubricated, which is a real risk directly outside a refrigerated dock door in a Canadian winter and can push effective friction well above catalog values.

**Rule of thumb:** apply a 10-15% contingency multiplier to the calculated $T_e$ before it goes into the power and torque calculation, to cover belt cleaners, skirtboards, plows, and the general uncertainty in any friction-factor estimate. This chapter uses 10%. This breaks down for a conveyor with unusually well-characterized friction data (e.g., a repeat design from an existing, instrumented installation) — there, a lower contingency may be defensible — and it is *not* a substitute for getting the friction factors themselves right; a 10% pad on a badly wrong friction estimate is still wrong.

### Governing Calculation

$$
\begin{aligned}
T_x &= C_b\,W_{belt}\cos\theta \\
T_y &= C_m\,W_{load}\cos\theta \\
T_z &= W_{load}\sin\theta \\
T_e &= T_x + T_y + T_z \\
T_{e,design} &= T_e \times (1 + \text{contingency})
\end{aligned}
$$

where $W_{belt} = w_b\,L\,g$ (the weight of the belt and idlers along the conveyor's length) and $W_{load} = q\,L\,g$ (the weight of material actually on the belt at the design loading, with $q$ the linear load density).

| Symbol | Description | Unit |
| :--- | :--- | :--- |
| $L$ | Conveyor belt-path length (along the incline) | m |
| $\theta$ | Incline angle from horizontal | degrees or rad |
| $w_b$ | Mass per unit length of belt + idlers/pulleys | kg/m |
| $W_{belt}$ | Total weight of belt and idlers | N |
| $q$ | Linear load density on the belt at design loading | kg/m |
| $W_{load}$ | Total weight of material on the belt at any instant | N |
| $C_b$ | Belt/idler friction factor | dimensionless |
| $C_m$ | Load friction factor | dimensionless |
| $T_x$ | Friction resistance of belt and idlers | N |
| $T_y$ | Friction resistance of the load | N |
| $T_z$ | Gravity (elevation) resistance of the load | N |
| $T_e$ | Effective (net driving) tension required | N |
| $T_{e,design}$ | Effective tension with contingency margin applied | N |

### Worked Example

```{r belt-pull-06-10, echo=TRUE}
## Given: inclined tote conveyor geometry, belt/idler data, and loading
theta_deg <- 22
theta <- theta_deg * pi / 180
L <- 11.5              # m, belt-path length along the incline
g <- 9.81

w_b <- 14               # kg/m, belt + idlers (from manufacturer's datasheet)
W_belt <- w_b * L * g    # N

tote_mass <- 18          # kg, average loaded tote
s_spacing <- 0.5         # m, center-to-center spacing at design (peak) loading
q <- tote_mass / s_spacing        # kg/m, linear load density
W_load <- q * L * g               # N

C_b <- 0.030    # belt/idler friction factor
C_m <- 0.035    # load friction factor

T_x <- C_b * W_belt * cos(theta)
T_y <- C_m * W_load * cos(theta)
T_z <- W_load * sin(theta)
T_e <- T_x + T_y + T_z

contingency <- 0.10
T_e_design <- T_e * (1 + contingency)

rise <- L * sin(theta)

cat(sprintf("Elevation rise: %.2f m\n", rise))
cat(sprintf("Belt + idler weight (W_belt): %.1f N (%.0f kg equivalent)\n", W_belt, w_b * L))
cat(sprintf("Load on belt at design loading (W_load): %.1f N (%.0f kg)\n\n", W_load, q * L))
cat(sprintf("T_x (belt/idler friction): %6.1f N\n", T_x))
cat(sprintf("T_y (load friction):       %6.1f N\n", T_y))
cat(sprintf("T_z (gravity/elevation):   %6.1f N\n", T_z))
cat(sprintf("T_e (effective tension):   %6.1f N\n\n", T_e))
cat(sprintf("T_e,design (+%.0f%% contingency): %.1f N\n", contingency * 100, T_e_design))
```

```{r fig-06-10-te-breakdown, echo=FALSE, fig.cap="Contribution of each effective-tension component. On this 22-degree incline, the gravity/elevation term dominates; on a shallower incline, the friction terms would make up a much larger share.", fig.align='center', fig.width=9, fig.height=5.5}
library(ggplot2)

te_df <- data.frame(
  component = factor(c("Belt/idler\nfriction (T_x)", "Load\nfriction (T_y)", "Gravity/\nelevation (T_z)"),
                      levels = c("Belt/idler\nfriction (T_x)", "Load\nfriction (T_y)", "Gravity/\nelevation (T_z)")),
  force_N = c(43.93, 131.80, 1521.40)
)

ggplot(te_df, aes(x = component, y = force_N, fill = component)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.0f N", force_N)), vjust = -0.4, size = 3.4 * diagram_text_scale) +
  scale_fill_manual(values = c("#3498DB", "#F39C12", "#E74C3C")) +
  ylim(0, 1750) +
  labs(title = "Effective Tension Breakdown",
       subtitle = "22-degree incline tote conveyor at design load (before contingency)",
       x = NULL, y = "Force contribution (N)") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 9.5 * diagram_text_scale),
        axis.text = element_text(size = 10.5 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale))
```

**Sanity check:** the gravity/elevation term ($T_z \approx 1521$ N) makes up nearly 90% of the effective tension, with the two friction terms together contributing only about 10%. That's the expected signature of a genuinely steep incline conveyor — physically, most of the drive's effort here is lifting totes against gravity, not overcoming rolling friction, which is exactly the opposite balance you'd see on a long, gently-inclined or level conveyor, where friction dominates and the gravity term shrinks toward zero. If this ratio came out reversed for a 22° incline, that would be a strong signal of an arithmetic error (most likely a $\sin\theta$/$\cos\theta$ swap) rather than a real result worth trusting.

## Design Task 2: Drive Power and Required Torque at the Head Pulley

### Scenario

With $T_{e,design}$ established, the mechanical designer's next job is to turn that force into two numbers a gearmotor supplier actually needs: how much power the drive must deliver, and how much torque it must produce at the head pulley shaft at the belt's running speed. Those two numbers — torque and shaft speed — are the exact starting point for motor and gearbox selection.

### Design Guidelines and Rules of Thumb

Mechanical power is simply force times velocity: the belt pull, multiplied by how fast the belt moves. That gives the power delivered *to the belt* at the pulley face. Getting from there to the power the *motor* must produce means dividing by the drive train's mechanical efficiency — the product of gearbox efficiency and any additional mechanical stage (a chain-and-sprocket final drive between the gearbox output shaft and the pulley shaft is common on conveyor drives, adding its own loss on top of the gearbox's own).

Torque at the pulley shaft follows directly from the belt pull acting at the pulley's radius: $T = T_{e,design} \times r_{pulley}$. Pulley diameter isn't a free choice — CEMA and belt manufacturers both publish standard pulley diameters (common sizes run roughly 100 mm to 600 mm, in a set of standard increments, not a continuous range), and every fabric or steel-cord conveyor belt has a **minimum recommended pulley diameter** from its manufacturer, based on how tightly the belt carcass can bend around a pulley without accelerated fatigue cracking. A smaller pulley than the belt's rated minimum shortens belt life, sometimes dramatically; a larger pulley is mechanically fine but adds cost and reduces the mechanical advantage available from a given torque, all else equal.

**Rule of thumb:** always check the selected pulley diameter against the specific belt's manufacturer-rated minimum pulley diameter before finalizing a design — this is a belt-specific number (driven by carcass construction: fabric ply count, steel cord, etc.), not a generic conveyor-wide constant, and it's easy to overlook when the pulley size gets chosen mainly to hit a convenient gear ratio. This breaks down (in the sense of needing an even larger allowance) for cold environments — belt carcass stiffens and becomes more fatigue-prone at low temperature, which is a real consideration for a conveyor running near an unheated dock door, and some belt manufacturers publish a separate, larger minimum pulley diameter for cold service.

**Rule of thumb:** size the drive's *starting* torque capability at roughly 150% of the running torque calculated here, to cover breakaway friction (static friction is generally higher than the running friction factors used above) and a fully-loaded cold start — a real consideration for an incline conveyor near a refrigerated receiving dock, where idler grease can be noticeably stiffer at start-of-shift than it is once the system has been running and has warmed slightly. This is conceptually the same starting-versus-running distinction developed in the RMS torque and peak torque checks in the Motor, Drive, and Gearbox Sizing chapter — running torque governs thermal capacity, while a separate, higher instantaneous torque check governs whether the drive can actually get the load moving.

### Governing Calculation

$$
P_{mech} = T_{e,design}\,v_{belt} \qquad P_{motor} = \frac{P_{mech}}{\eta_{drive}}
$$

$$
T_{pulley} = T_{e,design}\,r_{pulley} \qquad \omega_{pulley} = \frac{v_{belt}}{r_{pulley}}
$$

| Symbol | Description | Unit |
| :--- | :--- | :--- |
| $T_{e,design}$ | Effective tension with contingency (Design Task 1) | N |
| $v_{belt}$ | Belt running speed | m/s |
| $\eta_{drive}$ | Combined mechanical efficiency of gearbox + any final drive stage | dimensionless (0-1) |
| $P_{mech}$ | Mechanical power delivered to the belt at the pulley | W |
| $P_{motor}$ | Power the motor itself must produce | W |
| $r_{pulley}$ | Head pulley radius | m |
| $T_{pulley}$ | Required torque at the head pulley shaft | N·m |
| $\omega_{pulley}$ | Head pulley shaft angular velocity | rad/s |

This $T_{pulley}$ and $\omega_{pulley}$ (or the equivalent pulley rpm) pair is exactly the $T_{load}$ and $\omega_{load}$ that the [Motor, Drive, and Gearbox Sizing](06-06-Motor-Drive-Gearbox-Sizing.qmd) chapter's Design Task 1 reflects through a gear ratio to the motor shaft — from here, the rest of that chapter's method (gear ratio selection, VFD sizing for an induction motor drive like this one, and gearbox service factor selection from its Design Task 4) applies directly, without needing to be re-derived.

### Worked Example

Live cell — try a different belt speed, pulley diameter, or drive efficiency and see how power and torque respond.

```{webr-r}
#| label: drive-power-torque-06-10
## Given (re-derived locally so this cell is self-contained)
T_e_design <- 1866.85    # N, from Design Task 1's effective tension calculation
v_belt <- 0.5             # m/s
eta_drive <- 0.90         # combined gearbox + chain final-drive efficiency

P_mech <- T_e_design * v_belt
P_motor <- P_mech / eta_drive

D_pulley <- 0.254         # m, 254 mm (10 in) head pulley -- a standard CEMA pulley size
r_pulley <- D_pulley / 2
T_pulley <- T_e_design * r_pulley
omega_pulley <- v_belt / r_pulley
rpm_pulley <- omega_pulley * 60 / (2 * pi)

cat(sprintf("Mechanical power at the belt: %.1f W\n", P_mech))
cat(sprintf("Required motor power (%.0f%% drive efficiency): %.1f W (%.3f kW, %.2f hp)\n",
            eta_drive * 100, P_motor, P_motor / 1000, P_motor / 745.7))
cat(sprintf("\nHead pulley diameter: %.0f mm\n", D_pulley * 1000))
cat(sprintf("Required torque at pulley shaft: %.1f N.m\n", T_pulley))
cat(sprintf("Pulley shaft speed: %.3f rad/s (%.1f rpm)\n", omega_pulley, rpm_pulley))

## Cross-check: power from torque and speed should match the direct calculation
P_check <- T_pulley * omega_pulley
cat(sprintf("\nCross-check (P = T x omega): %.1f W\n", P_check))

## Illustrative gear ratio if paired with a standard 4-pole induction motor
motor_rpm <- 1750
N_ratio <- motor_rpm / rpm_pulley
cat(sprintf("Approximate gear ratio for a %d rpm 4-pole motor: %.1f:1\n", motor_rpm, N_ratio))
```

```{r fig-06-10-power-speed, echo=FALSE, fig.cap="Required motor power scales linearly with belt speed for a fixed effective tension. The design point (0.5 m/s) is marked.", fig.align='center', fig.width=9, fig.height=5.5}
library(ggplot2)

T_e_design_fig <- 1866.85
eta_drive_fig <- 0.90
v_seq <- seq(0.2, 1.0, by = 0.02)
P_seq <- T_e_design_fig * v_seq / eta_drive_fig
df_p <- data.frame(v = v_seq, P = P_seq)
P_design_pt <- T_e_design_fig * 0.5 / eta_drive_fig / 1000

ggplot(df_p, aes(x = v, y = P / 1000)) +
  geom_line(color = "#3498DB", linewidth = 1.2) +
  annotate("point", x = 0.5, y = P_design_pt, color = "#E74C3C", size = 4) +
  annotate("text", x = 0.53, y = P_design_pt + 0.15,
           label = sprintf("Design point:\n0.5 m/s -> %.2f kW", P_design_pt),
           size = 3 * diagram_text_scale, hjust = 0, color = "#E74C3C") +
  labs(title = "Required Motor Power vs. Belt Speed",
       subtitle = "Effective tension held constant at 1867 N (Design Task 1 result)",
       x = "Belt speed (m/s)", y = "Motor power (kW)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 9.5 * diagram_text_scale),
        axis.text = element_text(size = 10.5 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale))
```

**Sanity check:** roughly 1.04 kW (1.4 hp) required at the motor rounds up cleanly to a standard 1.5 kW (2 hp) NEMA/IEC frame motor with reasonable margin — a physically sensible size for a single 600 mm belt lifting totes at this rate, not an order of magnitude off in either direction. The pulley itself turns quite slowly (about 38 rpm), which is typical for a conveyor head pulley; a roughly 46:1 gear reduction from a standard 1750 rpm 4-pole motor is well within the range commonly available in a single helical-bevel gearmotor package, so this doesn't force an unusual two-stage drive train.

## Design Task 3: Belt Speed and Throughput Sizing

### Scenario

Design Tasks 1 and 2 both took the belt speed — 0.5 m/s — as a given. Where did that number actually come from? The case-pick mezzanine this conveyor feeds needs totes delivered at a peak rate of 45 totes per minute during a picking wave, the busiest sustained period in the mezzanine's shift. Downstream, a merge and singulation point requires totes to arrive no closer than 0.5 m apart center-to-center (a 400 mm tote footprint in the direction of travel, plus a 100 mm minimum gap for the transfer mechanism to safely separate one tote from the next). The process engineer needs to work backward from those two numbers to a minimum belt speed — and then decide how much faster than that minimum to actually run the conveyor.

### Design Guidelines and Rules of Thumb

For a **unit-load conveyor** (discrete items — totes, cartons, pallets — rather than a continuous stream of bulk material), throughput, spacing, and speed are linked by a simple relationship: if items pass a fixed point at rate $R$ (items per unit time) and are spaced $s$ apart (center-to-center) while travelling at speed $v$, then the time between successive items passing that point is $s/v$, and the rate is the reciprocal: $R = v/s$, or equivalently $v = R\,s$.

For a **bulk-material conveyor** (grain, aggregate, powder — a continuous cross-section of material rather than discrete units), the equivalent relationship uses the loaded cross-sectional area of material on the belt and its bulk density instead of a unit spacing:

$$Q = 3.6\,A\,v\,\rho$$

with $Q$ in tonnes/hour when $A$ is in m², $v$ is in m/s, and $\rho$ (bulk density) is in kg/m³ — the factor of 3.6 converts kg/s to tonnes/hour. As an illustration only (not this chapter's worked example): a bulk conveyor with a 0.025 m² loaded cross-section, running at 1.6 m/s, carrying a material with a bulk density of 750 kg/m³ (a representative value for dry grain) would move roughly $3.6 \times 0.025 \times 1.6 \times 750 \approx 108$ tonnes/hour. The unit-load case is this chapter's focus, since it matches the tote conveyor scenario directly.

**Rule of thumb:** whichever form applies, design the belt speed at 20-30% above the calculated minimum, not exactly at it. This chapter uses 30%. The margin absorbs normal variability the minimum-spacing calculation doesn't capture — occasional surges above the nominal peak rate, minor timing jitter from an upstream process, and simple line-balancing headroom — without which the conveyor would be running at its absolute throughput ceiling during every peak period, with zero ability to catch up if it ever falls even slightly behind. This breaks down for a conveyor whose speed is tightly constrained by something else entirely (a product that can't be conveyed faster than a certain speed without damage or spillage, or a downstream process that can't accept product any faster) — in that case the margin has to come from elsewhere in the system design (buffer capacity, a parallel line), not from simply running the belt faster.

**Rule of thumb:** the minimum safe spacing $s$ should come from the *tightest* downstream constraint in the system — often a merge, diverter, or singulation point rather than the item's bare physical footprint — because that's usually the actual limiting requirement, not the totes themselves. This breaks down if the conveyor in question sits *upstream* of any such constraint with substantial accumulation buffer in between; in that case, the conveyor's own speed can be set independently of the downstream device's minimum spacing, since the buffer decouples the two.

### Governing Calculation

$$
v_{min} = R\,s_{min} \qquad v_{design} = v_{min}\times(1+\text{margin})
$$

(unit-load form; see the bulk-material form above for a continuous-flow conveyor)

| Symbol | Description | Unit |
| :--- | :--- | :--- |
| $R$ | Required throughput rate | items/s (or items/min, consistent with $s$) |
| $s_{min}$ | Minimum safe center-to-center item spacing | m |
| $v_{min}$ | Minimum belt speed to meet the throughput requirement | m/s |
| $v_{design}$ | Specified belt speed, including surge margin | m/s |
| $Q$ | Bulk material mass flow rate | tonnes/hour |
| $A$ | Loaded material cross-sectional area on the belt | m² |
| $\rho$ | Bulk material density | kg/m³ |

### Worked Example

Live cell — try a different required throughput, tote spacing, or surge margin and see how the required belt speed responds.

```{webr-r}
#| label: speed-throughput-06-10
## Given
R_totes_per_min <- 45     # totes/min, peak picking-wave requirement
R_totes_per_s <- R_totes_per_min / 60

tote_len <- 0.4            # m, tote footprint in the direction of travel
min_gap <- 0.1              # m, minimum gap required by the downstream singulation point
s_min <- tote_len + min_gap

v_min <- R_totes_per_s * s_min

margin <- 0.30
v_design <- v_min * (1 + margin)

cat(sprintf("Required throughput: %.0f totes/min (%.3f totes/s)\n", R_totes_per_min, R_totes_per_s))
cat(sprintf("Minimum safe spacing: %.2f m (%.0f mm tote + %.0f mm gap)\n",
            s_min, tote_len * 1000, min_gap * 1000))
cat(sprintf("Minimum belt speed: %.4f m/s\n", v_min))
cat(sprintf("Design belt speed (+%.0f%% surge margin): %.4f m/s\n", margin * 100, v_design))
cat(sprintf("\nCompare to the 0.5 m/s speed used in Design Tasks 1-2: difference = %.4f m/s\n", 0.5 - v_design))
```

```{r fig-06-10-speed-throughput, echo=FALSE, fig.cap="Minimum belt speed required to meet a given throughput, for three minimum spacing assumptions. The chapter's design case (45 totes/min at 0.5 m spacing) is marked, along with the specified 0.5 m/s running speed.", fig.align='center', fig.width=9.5, fig.height=6}
library(ggplot2)
library(tidyr)

R_seq <- seq(10, 70, by = 1)
spacing_vals <- c(0.4, 0.5, 0.6)
df_v <- do.call(rbind, lapply(spacing_vals, function(s) {
  data.frame(R = R_seq, s = factor(paste0(s, " m")), v_min = (R_seq / 60) * s)
}))

ggplot(df_v, aes(x = R, y = v_min, color = s)) +
  geom_line(linewidth = 1.1) +
  annotate("point", x = 45, y = 0.375, color = "black", size = 3.5) +
  annotate("text", x = 47, y = 0.30, label = "Design case:\n45 totes/min, s = 0.5 m\n-> v_min = 0.375 m/s",
           size = 2.8 * diagram_text_scale, hjust = 0) +
  annotate("segment", x = 10, xend = 70, y = 0.5, yend = 0.5, linetype = "dashed", color = "grey40") +
  annotate("text", x = 12, y = 0.535, label = "Specified running speed (0.5 m/s)",
           size = 2.8 * diagram_text_scale, hjust = 0, color = "grey40") +
  scale_color_manual(values = c("#3498DB", "#2ECC71", "#E74C3C")) +
  labs(title = "Minimum Belt Speed vs. Required Throughput",
       subtitle = "Three minimum tote center-to-center spacings",
       x = "Required throughput (totes/min)", y = "Minimum belt speed (m/s)", color = "Spacing") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 9.5 * diagram_text_scale),
        axis.text = element_text(size = 10.5 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale),
        legend.position = "bottom")
```

**Sanity check:** the calculated design speed (about 0.49 m/s) lands almost exactly on the 0.5 m/s speed used throughout Design Tasks 1 and 2 — which is not a coincidence, it's the loop closing. The process engineer's original speed specification *was* this calculation, run before Design Task 1's belt-pull sizing ever happened; this design task simply reconstructs where that number came from and confirms it holds up. If the two numbers had come out far apart instead, that would be the signal to go back and reconcile which one is wrong — the throughput requirement, the spacing assumption, or the speed that was actually specified — before trusting the drive sizing built on top of it.

## Design Task 4: Roller vs. Belt vs. Tabletop Chain Conveyor Selection

### Scenario

Downstream of the incline conveyor, the same distribution center's case-packing area needs a short conveyor segment that must do three things at once: hold a short buffer of totes without back-pressure damaging them or their contents (accumulation), turn the tote's path of travel 90° to feed a case packer positioned across the aisle, and tolerate a daily hose-down as part of the plant's sanitation program. Three fundamentally different conveyor technologies could physically do this job — powered roller, flat belt, and tabletop (modular plastic) chain — and picking the wrong one doesn't just cost more, it creates a maintenance or sanitation problem that shows up months after commissioning, once it's expensive to change.

### Design Guidelines and Rules of Thumb

**Roller conveyor** — a series of rollers (gravity or powered, via a line shaft, belt, or motorized rollers) supporting the load directly — excels at moving rigid-bottomed, flat loads in a straight line, and is a natural fit for **accumulation**: a zero-pressure powered roller section can hold a queue of totes or cartons stationary, each roller clutching independently, without crushing the item behind it. It's a poor fit for anything but a shallow incline (rollers alone provide little traction up a slope without added friction surfaces) and for tight turns, which require dedicated curved roller sections that add cost and complexity compared to a straight run. This is exactly the technology behind pallet and skid transfer between stations on many automotive lines — a Tier 1 stamping supplier moving stamped door-outer-panel skids in a straight run from a blanking press to a transfer press is a textbook roller-conveyor application: heavy, rigid-bottomed loads, level runs, and no need to turn corners along the way.

**Belt conveyor** — the technology used in Design Tasks 1-3 — is the most versatile of the three for irregular or flexible-bottomed loads, and is the natural choice for **inclines**, since a cleated or friction-top belt can climb steep grades that neither roller nor tabletop chain handle well. It's a comparatively poor fit for accumulation (a moving belt surface makes holding items stationary without pileup or scuffing difficult without added diverting hardware) and, like roller conveyor, doesn't turn corners gracefully — a 90° direction change generally means a transfer to a second conveyor section, not a curved belt run.

**Tabletop (modular plastic) chain conveyor** — a chain of interlocking plastic slats running over a stationary bed — is the food and beverage industry's default answer for **washdown environments with turns and accumulation**: the open, articulated construction sheds water and debris far more readily than a belt's continuous surface and seams, dedicated curved chain modules handle 90° (and tighter) turns cleanly, and a low-friction chain top (or a separate low-back-pressure accumulation chain design) lets product queue with minimal contact force between items. It costs more per meter than roller or belt conveyor of comparable capacity, and its incline capability is limited — mostly level or very shallow grades, since the chain itself provides little of the traction a cleated belt can.

```{r tbl-06-10-conveyor-types, echo=FALSE}
library(knitr)
library(kableExtra)

conv_df <- data.frame(
  `Conveyor Type` = c("Powered Roller", "Belt", "Tabletop (Modular) Chain"),
  `Load Type` = c("Rigid, flat-bottomed (pallets, totes, cartons)",
                   "Flexible or irregular bottoms; bulk material with troughing",
                   "Rigid or semi-rigid, direct or indirect product contact"),
  `Speed Range` = c("0.1 - 1.0 m/s typical", "0.1 - 2.0+ m/s (much higher for bulk transport)",
                     "0.1 - 0.5 m/s typical"),
  `Relative Cost` = c("Low - moderate", "Low", "Moderate - high"),
  `Sanitation / Washdown` = c("Moderate (bearings/sprockets need washdown-rated seals)",
                               "Moderate (belt seams and edges can trap debris)",
                               "Excellent (open, self-draining, no belt seams)"),
  `Incline Capability` = c("Poor (level or very shallow only)",
                            "Excellent (with cleats/rough-top, up to ~30-40 deg)",
                            "Poor (level or very shallow only)"),
  `Typical Use Case` = c("Pallet/skid transfer between stations; accumulation buffering",
                          "Floor-to-floor incline transport; long straight runs",
                          "Washdown zones needing turns and low-back-pressure accumulation"),
  check.names = FALSE
)

kable(conv_df, caption = "Roller vs. belt vs. tabletop chain conveyor selection guide") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = knitr::is_html_output()) %>%
  row_spec(0, bold = TRUE, background = "#4a86c7", color = "white")
```

**Rule of thumb:** no single conveyor technology wins every criterion at once — this is a genuine multi-criteria trade-off, not a lookup table with one right answer, and the correct tool for making the decision defensible is a **weighted decision matrix**: assign each selection criterion a weight reflecting its importance to *this specific application*, score each candidate technology against every criterion, and compare the weighted totals. This breaks down when one criterion is a hard pass/fail constraint rather than a matter of degree — if a zone is a genuine direct food-contact daily-washdown area, for instance, a technology that scores poorly on sanitation shouldn't be in the running at all regardless of how well it scores elsewhere, and should be eliminated before the weighting exercise rather than weighted down within it.

### Governing Calculation

$$Score_{option} = \sum_i w_i \times s_{i,option}$$

| Symbol | Description | Unit |
| :--- | :--- | :--- |
| $w_i$ | Weight assigned to criterion $i$ (weights sum to 1.0) | dimensionless |
| $s_{i,option}$ | Score of a given conveyor type against criterion $i$ (e.g., 1-5 scale) | dimensionless |
| $Score_{option}$ | Total weighted score for that conveyor type | dimensionless |

### Worked Example

Live cell — try different weights (what matters most for *your* application?) and see whether the recommendation changes.

```{webr-r}
#| label: conveyor-selection-matrix-06-10
## Given: weights reflecting the case-packing transfer's priorities
weights <- c(sanitation = 0.30, accumulation = 0.25, turns = 0.20, cost = 0.15, incline = 0.10)

## Scores, 1 (poor fit) to 5 (excellent fit), for this specific application
roller   <- c(sanitation = 3, accumulation = 5, turns = 2, cost = 3, incline = 2)
belt     <- c(sanitation = 3, accumulation = 2, turns = 2, cost = 4, incline = 5)
tabletop <- c(sanitation = 5, accumulation = 4, turns = 5, cost = 2, incline = 2)

score_roller   <- sum(weights * roller)
score_belt     <- sum(weights * belt)
score_tabletop <- sum(weights * tabletop)

cat("Weighted decision matrix (1 = poor fit, 5 = excellent fit):\n\n")
for (nm in names(weights)) cat(sprintf("  %-13s weight = %.0f%%\n", nm, weights[nm] * 100))

cat(sprintf("\nRoller weighted score:          %.2f\n", score_roller))
cat(sprintf("Belt weighted score:            %.2f\n", score_belt))
cat(sprintf("Tabletop chain weighted score:   %.2f\n", score_tabletop))

best <- names(which.max(c(Roller = score_roller, Belt = score_belt, `Tabletop Chain` = score_tabletop)))
cat(sprintf("\n-> Recommendation: %s\n", best))
```

```{r fig-06-10-selection, echo=FALSE, fig.cap="Illustrative comparison of roller, belt, and tabletop chain conveyor across the five decision criteria for the case-packing transfer scenario.", fig.align='center', fig.width=10, fig.height=6}
library(ggplot2)
library(tidyr)

compare_df <- data.frame(
  Criterion = c("Sanitation/\nWashdown", "Accumulation", "Turns/\nTransfers", "Cost", "Incline\nCapability"),
  Roller = c(3, 5, 2, 3, 2),
  Belt = c(3, 2, 2, 4, 5),
  `Tabletop Chain` = c(5, 4, 5, 2, 2),
  check.names = FALSE
)
compare_long <- pivot_longer(compare_df, cols = c(Roller, Belt, `Tabletop Chain`),
                              names_to = "Type", values_to = "Score")
compare_long$Criterion <- factor(compare_long$Criterion, levels = compare_df$Criterion)
compare_long$Type <- factor(compare_long$Type, levels = c("Roller", "Belt", "Tabletop Chain"))

ggplot(compare_long, aes(x = Criterion, y = Score, fill = Type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Roller" = "#3498DB", "Belt" = "#F39C12", "Tabletop Chain" = "#2ECC71")) +
  labs(title = "Conveyor Technology Comparison — Case-Packing Transfer",
       subtitle = "Relative rating, 1 (poor fit) to 5 (excellent fit) — application-specific",
       x = NULL, y = "Relative rating (1-5)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"),
        plot.subtitle = element_text(size = 9.5 * diagram_text_scale),
        legend.position = "bottom",
        axis.text = element_text(size = 10 * diagram_text_scale),
        axis.title = element_text(size = 11 * diagram_text_scale))
```

**Sanity check:** tabletop chain wins clearly (4.00 vs. roller's 3.20 and belt's 2.90), which matches what a food-plant conveyor engineer would specify from experience for a washdown zone that needs turns and accumulation — this decision matrix isn't discovering anything surprising, it's documenting and defending a choice an experienced engineer would already lean toward. The value of running the matrix explicitly is what happens when priorities shift: re-weight toward the automotive stamping example instead — heavy skids, straight run, no washdown, cost and accumulation matter far more than sanitation or turns — and the recommendation flips to roller conveyor, exactly matching that application's real-world default.

## Design Task 5: Food-Industry Construction — Washdown, CFIA, and Frame Material Selection

### Scenario

Before any of this chapter's conveyors gets a purchase order, the distribution center's engineering standard requires every conveyor segment in the facility to be classified by sanitary zone, and its construction — frame material, fastener type, and enclosure rating on every motor and sensor — specified to match. Three representative zones illustrate the range this single facility spans: the incline tote conveyor from Design Tasks 1-3 (dry, sealed cartons, no direct product exposure, located in a stairwell-like enclosure between floors); a product-transfer conveyor downstream of a case erector, where opened product occasionally contacts the conveyor surface directly and the area gets a periodic rinse-down, not a daily hose-down; and, at the far end of the sanitary spectrum, a raw poultry evisceration takeaway conveyor in the facility's small further-processing room, hosed down with caustic sanitizer every single shift. One conveyor specification does not fit all three.

### Design Guidelines and Rules of Thumb

**Sanitary design principles**, regardless of exactly which standard is invoked, come down to a short list of physical rules that make a conveyor easy to clean and hard for product residue, water, and bacteria to hide on:

- **Open frame construction.** Round or square tube, not C-channel or angle iron, wherever the frame is exposed to washdown or product residue. A C-channel's flat top flange is a horizontal ledge — precisely the feature the next rule prohibits — while a round tube presents no flat surface for anything to sit on.
- **No horizontal ledges.** Any flat, horizontal surface on a food-zone conveyor is a place for water, product debris, and eventually biofilm to accumulate, out of easy sight and often out of reach of a normal cleaning routine. Structural members, brackets, and even conduit runs should be mounted on a slope or eliminated in favor of a shape that sheds water.
- **Accessible, tool-minimal fasteners.** Blind or threaded holes, especially ones a fastener doesn't fully occupy, trap moisture and are difficult to clean and inspect. Sanitary designs favor open, quick-release, or tool-less fasteners wherever a component needs regular removal for cleaning.
- **IP-rated components in the washdown path.** Motors, sensors, junction boxes, and any other electrical component exposed to direct washdown need an enclosure rating suited to that exposure — the international **IP69K** rating (per **IEC 60529**, the international ingress-protection standard) specifically covers high-pressure, high-temperature washdown, and is the de facto standard callout for food-industry washdown motors and sensors. Components rated only for dust and light spray (IP54/IP55) are appropriate in dry, non-washdown zones but not in a direct hose-down path.

```{r fig-06-10-sanitary, echo=FALSE, fig.cap="Sanitary frame cross-sections: a C-channel's flat top flange forms a horizontal ledge that traps water and product debris, while an open round-tube frame with sloped mounts sheds water and leaves nowhere for debris to collect.", fig.align='center', fig.width=11, fig.height=6.5}
library(ggplot2)
library(ggforce)

ggplot() +
  ## Left: non-sanitary C-channel
  annotate("rect", xmin = -4.3, xmax = -4.0, ymin = 0, ymax = 3, fill = "#95A5A6", color = "black") +
  annotate("rect", xmin = -4.3, xmax = -2.0, ymin = 2.7, ymax = 3.0, fill = "#95A5A6", color = "black") +
  annotate("rect", xmin = -4.3, xmax = -2.0, ymin = 0.0, ymax = 0.3, fill = "#95A5A6", color = "black") +
  geom_point(data = data.frame(x = c(-3.8, -3.5, -3.2, -2.9, -2.6), y = rep(3.12, 5)),
             aes(x = x, y = y), color = "#3498DB", size = 3) +
  annotate("segment", x = -1.9, xend = -2.6, y = 3.6, yend = 3.15,
           arrow = arrow(length = unit(0.2, "cm"), type = "closed"), linewidth = 0.9) +
  annotate("text", x = -1.85, y = 3.85, label = "Water & product\ndebris trapped on ledge",
           size = 2.7 * diagram_text_scale, hjust = 0) +
  annotate("text", x = -3.15, y = -0.9, label = "AVOID: C-channel frame\nwith horizontal ledges",
           size = 3.1 * diagram_text_scale, fontface = "bold") +
  ## Divider
  annotate("segment", x = 0, xend = 0, y = -1.3, yend = 4.3, linetype = "dashed", color = "grey50") +
  ## Right: sanitary open tube -- a genuine single fixed circle, so geom_circle(data=...) per
  ## the annotate("circle", ...) pitfall, not annotate()
  geom_circle(data = data.frame(x0 = 3, y0 = 1.5, r = 0.6),
              aes(x0 = x0, y0 = y0, r = r), fill = "#2ECC71", color = "black") +
  annotate("segment", x = 3, xend = 3.9, y = 0.9, yend = 0.1, linewidth = 1) +
  annotate("segment", x = 3, xend = 2.1, y = 0.9, yend = 0.1, linewidth = 1) +
  annotate("text", x = 3, y = -0.9, label = "SPECIFY: round tube frame,\nsloped mounts, no ledges",
           size = 3.1 * diagram_text_scale, fontface = "bold") +
  annotate("text", x = 3, y = 3.3, label = "Self-draining -\nno flat surfaces",
           size = 2.7 * diagram_text_scale) +
  coord_fixed() +
  xlim(-5.2, 5.2) + ylim(-1.5, 4.5) +
  theme_void() +
  labs(title = "Sanitary vs. Non-Sanitary Frame Cross-Sections") +
  theme(plot.title = element_text(size = 14 * diagram_text_scale, face = "bold"))
```

In Canada, conveyor construction expectations for equipment used in a federally registered food establishment flow from the **Canadian Food Inspection Agency (CFIA)**, administering the **Safe Food for Canadians Regulations (SFCR)**. CFIA does not typically publish its own detailed equipment-construction dimensional standard the way a mechanical standards body does; instead, it expects equipment to meet recognized sanitary design practice and be maintained in a condition consistent with the plant's own food safety plan and CFIA's inspection expectations. In practice, plants and equipment suppliers demonstrate that compliance by designing to established sanitary equipment standards — most commonly the **3-A Sanitary Standards** (originally developed for dairy equipment in the US, now widely referenced across food equipment generally), **NSF/ANSI 169** (Special-Purpose Food Equipment), and, for baking-specific equipment, the **BISSC** (Baking Industry Sanitation Standards Committee) standard. These are the international/US equivalents CFIA-regulated plants reference directly, the same way this book's electromechanical chapters point to CEC/OESC first and NEC second — here, CFIA sets the Canadian regulatory expectation, and the named standards provide the engineering detail behind meeting it.

**Stainless vs. painted/mild steel** is a decision that should follow the zone's actual moisture and product-contact exposure, not a blanket "always stainless in a food plant" rule that oversizes cost in genuinely dry areas or, worse, a blanket "mild steel is fine, it's not touching food" rule that undersizes durability in a zone that turns out to see more moisture than assumed.

- **Dry, non-contact zones** (enclosed stairwell risers, dry-goods staging, most conveyor structure that sealed cartons or totes sit on, away from any washdown spray) can reasonably use painted or powder-coated mild steel, or galvanized steel — a real cost saving over stainless with no meaningful durability downside in a genuinely dry environment.
- **Periodic-rinse or indirect/packaged product-contact zones** (occasional wipe-down or light rinse, product in its final packaging rather than exposed) typically call for **304 stainless steel** — good general corrosion resistance at a lower cost than 316L, appropriate where exposure is real but not constant or aggressive.
- **Frequent hose-down, direct raw-product-contact, or caustic-sanitizer-exposed zones** call for **316L stainless steel**, which adds molybdenum for meaningfully better resistance to chlorides (common in sanitizing chemistry and in raw protein processing) than 304 provides.

**Rule of thumb:** classify each conveyor zone by two independent questions — does product (raw or packaged) directly contact this conveyor, and how often and how aggressively is this area washed down — and let the *more demanding* answer of the two drive the material selection for that zone. This breaks down at zone boundaries in a real layout: a conveyor that physically spans two zones (crossing from a dry area into a washdown area, for instance) should generally be built to the more demanding zone's standard for its entire length, or split into two conveyor sections at the zone boundary, rather than mixing construction along a single continuous frame — both to simplify fabrication and because moisture and cleaning chemicals don't respect a zone boundary drawn on a layout drawing.

```{r tbl-06-10-zones, echo=FALSE}
library(knitr)
library(kableExtra)

zone_df <- data.frame(
  `Zone Classification` = c("Dry, non-contact", "Periodic rinse / packaged contact", "Daily washdown / direct raw contact"),
  `Product Contact` = c("None (sealed cartons/totes)", "Indirect or packaged product", "Direct, raw product"),
  `Washdown Frequency` = c("None (dry sweep/vacuum only)", "Periodic rinse", "Daily hose-down, caustic sanitizer"),
  `Recommended Frame Material` = c("Painted/powder-coated mild steel or galvanized", "304 stainless steel", "316L stainless steel"),
  `Motor/Sensor IP Rating` = c("IP54 - IP55", "IP69K", "IP69K"),
  check.names = FALSE
)

kable(zone_df, caption = "Sanitary zone classification and construction guide (illustrative, plant-specific standards may differ)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = knitr::is_html_output()) %>%
  row_spec(0, bold = TRUE, background = "#4a86c7", color = "white")
```

<details>
<summary>Industry connection: CFIA registration and the cost of getting construction wrong after the fact</summary>

A federally registered establishment under CFIA's Safe Food for Canadians licensing must be able to demonstrate its equipment supports its food safety plan on an ongoing basis, not just at commissioning — which means a conveyor built with painted mild steel in a zone that turns out, in practice, to see more moisture exposure than originally planned (a nearby washdown station spraying farther than expected, for example) becomes a corrosion and sanitation finding during a CFIA inspection, not just a maintenance annoyance. Retrofitting an existing conveyor's frame material after the fact is far more expensive than specifying correctly the first time, which is exactly why the zone-classification exercise in this design task belongs at the layout stage, before steel is ordered.

</details>

### Governing Calculation

There's no single formula here — this is a classification and specification exercise — but it can be made into a repeatable **decision framework**: for each conveyor zone, determine product contact type and washdown frequency, then look up (or logically derive) the required frame material, fastener standard, and component IP rating from those two inputs, as summarized in the table above.

### Worked Example

Live cell — try classifying a zone of your own by changing the product-contact and washdown-frequency inputs.

```{webr-r}
#| label: sanitary-zone-checklist-06-10
## Given: a simple classification function implementing the decision framework above
classify_zone <- function(zone_name, product_contact, washdown_frequency) {
  if (washdown_frequency == "daily-hosedown" || product_contact == "direct-raw-protein") {
    material <- "316L stainless steel"
    fastener <- "Open, tool-less, quick-release; no blind/threaded holes"
    ip_rating <- "IP69K"
  } else if (washdown_frequency == "periodic-rinse" || product_contact == "direct-packaged") {
    material <- "304 stainless steel"
    fastener <- "Open, accessible, corrosion-resistant hardware"
    ip_rating <- "IP69K"
  } else {
    material <- "Painted/powder-coated mild steel (or galvanized)"
    fastener <- "Standard accessible hardware"
    ip_rating <- "IP54 - IP55 (dust/light spray)"
  }
  cat(sprintf("Zone: %s\n", zone_name))
  cat(sprintf("  Product contact: %s | Washdown frequency: %s\n", product_contact, washdown_frequency))
  cat(sprintf("  -> Frame material: %s\n", material))
  cat(sprintf("  -> Fasteners: %s\n", fastener))
  cat(sprintf("  -> Motor/sensor IP rating: %s\n\n", ip_rating))
}

## Apply the framework to three zones in the distribution center
classify_zone("Incline tote conveyor (Design Tasks 1-3)", "none-sealed-cases", "none-dry")
classify_zone("Post-case-erector product transfer", "direct-packaged", "periodic-rinse")
classify_zone("Raw poultry evisceration takeaway", "direct-raw-protein", "daily-hosedown")
```

**Sanity check:** the incline tote conveyor from Design Tasks 1-3 classifies as a dry, non-contact zone under this framework — painted mild steel is technically adequate there, and specifying stainless throughout would be over-engineering by the letter of the framework. In practice, though, many plants deliberately choose to standardize on stainless (or at least a consistent higher-grade construction) across an entire facility anyway, even in zones that would technically qualify for a cheaper material, to simplify future layout changes, ease insurance and audit conversations, and avoid the risk of a zone's classification quietly changing (a new process added nearby that introduces washdown exposure) without anyone revisiting the original material decision. That's a legitimate, defensible business judgment call layered on top of the minimum technical requirement — not a contradiction of the framework, but a reminder that "minimum acceptable" and "what a specific plant chooses to standardize on" are two different, equally valid conversations.

## Summary

```{r tbl-06-10-summary, echo=FALSE}
library(knitr)
library(kableExtra)

summary_df <- data.frame(
  `Design Task` = c(
    "1. Belt pull (CEMA effective tension)",
    "2. Drive power & pulley torque",
    "3. Belt speed & throughput",
    "4. Roller/belt/tabletop selection",
    "5. Sanitary construction & material"
  ),
  `Governing Relationship` = c(
    "$T_e=T_x+T_y+T_z=C_bW_{belt}\\cos\\theta+C_mW_{load}\\cos\\theta+W_{load}\\sin\\theta$",
    "$P=T_{e,design}v/\\eta_{drive}$; $T_{pulley}=T_{e,design}r$",
    "$v_{min}=R\\,s_{min}$ (unit load); $Q=3.6Av\\rho$ (bulk)",
    "$Score=\\sum w_i s_i$ (weighted decision matrix)",
    "Zone (contact + washdown frequency) -> material/fastener/IP rating"
  ),
  `Key Rule of Thumb` = c(
    "Add 10-15% contingency to $T_e$; gravity term uses load weight only, not belt weight",
    "Check pulley diameter against the belt's rated minimum bend diameter",
    "Design speed 20-30% above the calculated minimum for surge capacity",
    "Weight sanitation/turns/accumulation heavily for washdown zones; cost/accumulation for pallet transfer",
    "Let the more demanding of contact type or washdown frequency drive material selection"
  ),
  check.names = FALSE
)
summary_df[] <- lapply(summary_df, tex_safe)

kable(summary_df, caption = "Summary: Conveyor Design design tasks", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = knitr::is_html_output()) %>%
  row_spec(0, bold = TRUE, background = "#4a86c7", color = "white") %>%
  column_spec(1, width = "1.2in")
```

## Practice Problems

<details>
<summary>Design Task 1, Problem 1 (plug-in): effective tension on a shallower incline</summary>

A conveyor climbs a 10° incline over a 15 m belt path, moving a load of 300 kg at design loading (weight = 2943 N) with a belt/idler weight of 200 kg (weight = 1962 N). Using $C_b = 0.03$ and $C_m = 0.035$, find $T_x$, $T_y$, $T_z$, and $T_e$ (no contingency).

**Solution:**

$$T_x = 0.03 \times 1962 \times \cos10° = 58.0\ \text{N}$$
$$T_y = 0.035 \times 2943 \times \cos10° = 101.4\ \text{N}$$
$$T_z = 2943 \times \sin10° = 511.1\ \text{N}$$
$$T_e = 58.0 + 101.4 + 511.1 = 670.5\ \text{N}$$

Note how much smaller a share of $T_e$ the gravity term ($T_z$, about 76%) is here compared to the chapter's 22° example (about 90%) — the friction terms matter more, proportionally, on a shallower incline.

</details>

<details>
<summary>Design Task 1, Problem 2 (design decision): a maintenance tech proposes lumping belt weight into the gravity term "to be safe"</summary>

A maintenance technician reviewing this chapter's method suggests it would be more conservative to add the belt's own weight into the $T_z$ (gravity) term as well, the way the simplified decline-conveyor example in the Motor, Drive, and Gearbox Sizing chapter did — "that way we're not underestimating the tension." Is this actually more conservative, and should you do it?

**Answer:** It isn't correct, and "conservative" isn't really the right frame for evaluating it. As this chapter's Design Task 1 explains, the belt's own weight doesn't need a net gravity/elevation allowance at steady running speed, because what the carrying strand's climb costs in potential energy, the return strand's descent gives back over one complete loop — the belt's weight already shows up correctly in the friction term ($T_x$), where it presses down on every idler along both strands. Adding it again into $T_z$ double-counts a resistance that isn't actually there, inflating the calculated tension for no physically real reason. That's not "safely conservative" in a useful sense — it's simply wrong, and it would lead to specifying a larger motor and gearbox than the job needs, the same kind of oversizing this book's introduction to Part 6 warns costs real money on a competitive quote. The right way to add genuine margin, if more is wanted, is the contingency factor already built into the method (this chapter uses 10%) — a deliberate, documented allowance, not an accidental double-count of a term that should be zero.

</details>

<details>
<summary>Design Task 2, Problem 1 (plug-in): power and torque from a new tension</summary>

Using $T_{e,design} = 950$ N, a belt speed of 0.35 m/s, a drive efficiency of 0.88, and a 200 mm diameter head pulley, find the required motor power and pulley shaft torque.

**Solution:**

$$P_{mech} = 950 \times 0.35 = 332.5\ \text{W} \qquad P_{motor} = 332.5 / 0.88 = 377.8\ \text{W}$$
$$T_{pulley} = 950 \times 0.10 = 95.0\ \text{N·m}$$

</details>

<details>
<summary>Design Task 2, Problem 2 (design decision): a supplier proposes a smaller pulley to hit a "nicer" gear ratio</summary>

A gearmotor supplier proposes swapping the specified 254 mm head pulley for a 150 mm pulley, because it lines up with a standard gear ratio they have in stock and would shave a few days off lead time. What should you check before agreeing?

**Answer:** Check the selected conveyor belt's manufacturer-rated minimum pulley diameter before agreeing to anything. A smaller pulley increases the bend the belt carcass experiences going around it, and if 150 mm falls below the belt's rated minimum, the change trades a few days of lead time for materially shorter belt life — a worse trade than it looks like on the surface, especially for a belt that's expensive or slow to replace. It also changes the required torque at the pulley shaft (torque scales with pulley radius for the same effective tension, so a smaller pulley means *less* torque required at the same power, which shifts the gear ratio math too) — so the supplier's proposed ratio needs to be re-checked against the actual $T_e$ and speed requirement, not assumed to still work just because it looked convenient. If the smaller pulley clears the belt manufacturer's minimum and the torque/ratio numbers still check out, it may well be a reasonable substitution — but that has to be verified, not assumed from lead time alone.

</details>

<details>
<summary>Design Task 3, Problem 1 (plug-in): minimum speed for a new throughput target</summary>

A conveyor must move cartons at 80 cartons/min, with a minimum safe spacing of 0.45 m. Find the minimum belt speed, and the design speed with a 25% margin.

**Solution:**

$$v_{min} = \left(\frac{80}{60}\right) \times 0.45 = 0.60\ \text{m/s}$$
$$v_{design} = 0.60 \times 1.25 = 0.75\ \text{m/s}$$

</details>

<details>
<summary>Design Task 3, Problem 2 (design decision): throughput requirement doubles mid-project</summary>

Partway through a project, the mezzanine's throughput requirement doubles, from 45 totes/min to 90 totes/min, with the same 0.5 m minimum spacing. The conveyor's belt speed was already specified at 0.5 m/s per this chapter's worked example. Is simply speeding up the belt the right fix, and what else needs to be re-checked if it is?

**Answer:** Doubling the belt speed is the direct fix suggested by $v_{min} = R\,s_{min}$ — but it isn't free, and treating it as "just change one number" would be a mistake. Design Task 2's governing calculation shows that required power scales directly with belt speed for a given effective tension ($P = T_e v / \eta$), so doubling speed roughly doubles the required motor power — very likely pushing the drive into the next standard motor frame size, not just a faster setting on the same motor. It's also worth checking whether the friction factors themselves hold up at the new speed (idler friction and belt flexure losses are not always perfectly speed-independent in practice, especially near a belt or gearbox's rated speed limit) and whether the gearbox's rated output speed and the pulley's rated speed both have headroom for the change. In short: yes, speed is the direct lever, but changing it re-opens Design Tasks 1 and 2's calculations, not just Design Task 3's — a doubled throughput requirement is a re-sizing exercise for the whole drive train, not a one-line speed change.

</details>

<details>
<summary>Design Task 4, Problem 1 (plug-in): re-weighting for the automotive pallet-transfer case</summary>

Using the same scores as the worked example (roller: sanitation 3, accumulation 5, turns 2, cost 3, incline 2; belt: 3, 2, 2, 4, 5; tabletop: 5, 4, 5, 2, 2), but with weights of 5% sanitation, 30% accumulation, 10% turns, 40% cost, 15% incline (an automotive pallet-transfer application), which technology wins?

**Solution:**

$$Score_{roller} = 0.05(3)+0.30(5)+0.10(2)+0.40(3)+0.15(2) = 3.35$$
$$Score_{belt} = 0.05(3)+0.30(2)+0.10(2)+0.40(4)+0.15(5) = 3.30$$
$$Score_{tabletop} = 0.05(5)+0.30(4)+0.10(5)+0.40(2)+0.15(2) = 2.85$$

Roller conveyor wins narrowly under these cost- and accumulation-weighted priorities — matching the automotive pallet/skid transfer example described in this design task's guidelines.

</details>

<details>
<summary>Design Task 4, Problem 2 (design decision): a plant standardizes on one conveyor type everywhere</summary>

A plant manager, aiming to simplify spare parts and maintenance training, proposes standardizing on tabletop chain conveyor for every conveyor in the facility, including the long incline conveyor from Design Tasks 1-3. Is this a good idea?

**Answer:** Not for the incline conveyor specifically, and the decision framework in this design task explains why directly: tabletop chain scored poorly (2 out of 5) on incline capability in the worked example, for a physical reason — the chain itself provides little traction on a slope, unlike a cleated or friction-top belt. Standardizing on one technology has real, legitimate benefits (fewer spare parts to stock, less cross-training for maintenance staff, simpler purchasing) and those benefits shouldn't be dismissed, but they don't override a genuine physical incapability — a tabletop chain conveyor simply cannot do the incline conveyor's job at a 22° grade the way a belt conveyor can. The more defensible version of the plant manager's instinct is to standardize *within* each technology category where it's actually the right fit (one preferred tabletop chain supplier for washdown/turn applications, one preferred belt supplier for inclines) rather than forcing a single conveyor technology to cover every application in the plant.

</details>

<details>
<summary>Design Task 5, Problem 1 (plug-in): classifying a new zone</summary>

A conveyor segment carries opened, unpackaged produce directly on its surface, in an area rinsed down once per shift (not a full daily caustic hose-down). Using the classification framework in this design task, what frame material, fastener approach, and IP rating apply?

**Solution:**

Direct product contact (unpackaged produce) with periodic rinse frequency (not daily hose-down) places this in the "periodic rinse / packaged contact" tier of the framework (direct contact alone, without daily hose-down or raw-protein exposure, doesn't escalate to the 316L tier): **304 stainless steel**, open/accessible corrosion-resistant fasteners, and **IP69K**-rated motors and sensors. (If the direct-produce-contact classification were instead treated as demanding the top tier regardless of wash frequency — a defensible plant-specific judgment call, since direct contact with unpackaged product is itself a significant factor — 316L would be the more conservative choice; this is exactly the kind of framework boundary case worth a documented decision rather than a default answer.)

</details>

<details>
<summary>Design Task 5, Problem 2 (design decision): budget pressure to downgrade a washdown zone's material</summary>

A project is over budget, and someone proposes downgrading the raw poultry evisceration takeaway conveyor (this chapter's daily-hosedown, direct-raw-protein-contact zone) from 316L to 304 stainless steel to save cost. What's the engineering case against it, and is there a legitimate lower-cost alternative worth proposing instead?

**Answer:** The engineering case against the downgrade is specific, not generic caution: 316L's added molybdenum content exists specifically to resist chloride attack, and this zone combines two of the strongest chloride sources a food plant has — daily caustic/chlorine-based sanitizer and raw protein processing residue. Downgrading to 304 in exactly the zone where chloride exposure is highest is the one place in the facility where that substitution is most likely to cause visible corrosion (pitting, especially at welds) within a a normal service life, not a marginal, defensible risk. A more legitimate way to respond to budget pressure is to make sure the 316L specification is scoped precisely — confirm it's applied only to the components genuinely exposed to washdown and direct contact (frame, product-contact surfaces, fasteners in the wash zone), rather than over-specified onto structural elements or drive components mounted outside the direct spray pattern, which may reasonably use a lower-cost enclosure or mounting arrangement without compromising the sanitary zones that actually need 316L. That's a real cost reduction lever; substituting a lower corrosion grade in the zone with the highest chloride exposure in the building is not.

</details>

## Open-Ended Design Scenario

<details>
<summary>Capstone: receiving-to-shipping conveyor system for a small bakery</summary>

A small commercial bakery is laying out a new facility with three conveyor segments connecting receiving to shipping:

**Segment 1 — Ingredient receiving to mixing floor:** Sacks of flour and other dry ingredients (average 25 kg each) arrive at ground-floor receiving and must climb to a second-floor mixing room, 3.8 m higher, over a conveyor path length of 10.2 m. The ingredients are in sealed, dry packaging; this area sees no washdown, only dry housekeeping. Required throughput: 200 sacks/hour during a receiving window, with a minimum safe sack spacing of 0.7 m.

**Segment 2 — Mixing floor to proofing/baking line:** A short, level tabletop chain conveyor moves bulk dough trays (12 kg each) from the mixing room to the proofing area. This area is rinsed down at the end of each shift, and trays contact the conveyor surface directly.

**Segment 3 — Cooling to packaging to shipping:** Finished baked product, now individually wrapped, moves from the cooling tunnel to the packaging line and out to the shipping dock on a level belt conveyor, in a dry, non-contact area.

Using the methods developed in this chapter:

1. For Segment 1, apply Design Tasks 1-3: derive the required belt speed from the throughput and spacing requirement (Design Task 3's method) with a suitable surge margin, then use that speed to calculate the effective tension (Design Task 1's CEMA-style breakdown — you'll need to make and state reasonable assumptions for belt/idler weight per unit length and friction factors) and the resulting drive power and head-pulley torque (Design Task 2).
2. For Segment 2, apply Design Task 4's decision framework: build your own weighted decision matrix confirming tabletop chain is the right technology for this application, and briefly justify why roller or belt would be a worse fit here specifically.
3. For all three segments, apply Design Task 5's zone-classification framework: state each segment's product-contact type and washdown frequency, and specify the resulting frame material, fastener approach, and component IP rating for each.
4. State what information from Segment 1's Design Task 2 result (torque and pulley shaft speed) you would hand to a controls engineer applying the Motor, Drive, and Gearbox Sizing chapter's method to finish selecting an actual motor and gearbox — you don't need to carry that calculation through, just identify precisely what gets handed off.

There's no single correct numeric answer here — different reasonable assumptions about friction factors, belt/idler weight, and spacing margins will lead to different (defensible) specifications. The goal is a documented conveyor specification package, in the style of this chapter's worked examples, that a bakery's project engineer could actually hand to an integrator for quotation.

</details>
