# Power Transmission: Belt & Pulley Systems

```{r setup-10, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")

# Custom theme for belt and pulley visualizations
theme_pulley <- theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )
```

## 💡 Introduction to Belt & Pulley Systems {-}

Belt and pulley systems represent one of the oldest and most versatile methods of mechanical power transmission. From the spinning wheels of ancient civilizations to modern automotive engines and industrial machinery, belts have been transmitting power across distances for thousands of years.

> **Think About It:** Look around your everyday environment. Your car uses multiple belt drives—the serpentine belt powers the alternator, water pump, and air conditioning compressor. A drill press uses belts to change speeds. Even household appliances like washing machines and vacuum cleaners rely on belt drives. Why do engineers choose belts over direct gear connections in so many applications?

The answer lies in the unique advantages belt drives offer:

- **Shock absorption** – Belts naturally dampen vibrations and load spikes
- **Distance** – Power can be transmitted over relatively long center distances
- **Speed flexibility** – Easy to change speed ratios by swapping pulleys
- **Quiet operation** – Much quieter than gear or chain drives
- **Overload protection** – Belts slip before catastrophic failure occurs
- **Low cost** – Generally less expensive than gear systems

---

## ⚙️ Fundamental Belt Drive Concepts {-}

### How Belt Drives Work

A belt drive consists of at least two pulleys connected by a flexible belt. When the driving pulley (connected to the motor or power source) rotates, friction between the belt and pulley surface transmits the motion to the driven pulley.

```{r belt-drive-diagram, echo=FALSE, fig.align='center', fig.cap="Basic two-pulley belt drive system showing driver and driven pulleys", fig.width=10, fig.height=6}
# Create a basic belt drive visualization
library(ggplot2)
library(ggforce)

# Define pulley parameters
driver_x <- 0
driver_y <- 0
driver_r <- 1.5  # Smaller driver pulley (speed increase)

driven_x <- 8
driven_y <- 0
driven_r <- 3    # Larger driven pulley

# Calculate tangent points for belt
# Upper tangent
angle_driver_upper <- asin((driven_r - driver_r) / (driven_x - driver_x))
angle_driven_upper <- angle_driver_upper

# Tangent points
tang_driver_upper_x <- driver_x + driver_r * cos(pi/2 - angle_driver_upper)
tang_driver_upper_y <- driver_y + driver_r * sin(pi/2 - angle_driver_upper)
tang_driven_upper_x <- driven_x + driven_r * cos(pi/2 - angle_driven_upper)
tang_driven_upper_y <- driven_y + driven_r * sin(pi/2 - angle_driven_upper)

tang_driver_lower_x <- driver_x + driver_r * cos(-pi/2 + angle_driver_upper)
tang_driver_lower_y <- driver_y + driver_r * sin(-pi/2 + angle_driver_upper)
tang_driven_lower_x <- driven_x + driven_r * cos(-pi/2 + angle_driven_upper)
tang_driven_lower_y <- driven_y + driven_r * sin(-pi/2 + angle_driven_upper)

# Create the plot
ggplot() +
  # Driver pulley (outer ring)
  geom_circle(aes(x0 = driver_x, y0 = driver_y, r = driver_r),
              fill = "steelblue", color = "darkblue", linewidth = 1.2) +
  geom_circle(aes(x0 = driver_x, y0 = driver_y, r = driver_r * 0.3),
              fill = "gray30", color = "black", linewidth = 1) +
  # Driven pulley (outer ring)
  geom_circle(aes(x0 = driven_x, y0 = driven_y, r = driven_r),
              fill = "coral", color = "darkred", linewidth = 1.2) +
  geom_circle(aes(x0 = driven_x, y0 = driven_y, r = driven_r * 0.15),
              fill = "gray30", color = "black", linewidth = 1) +
  # Belt - upper span (tight side)
  geom_segment(aes(x = tang_driver_upper_x, y = tang_driver_upper_y,
                   xend = tang_driven_upper_x, yend = tang_driven_upper_y),
               color = "black", linewidth = 3) +
  # Belt - lower span (slack side)
  geom_segment(aes(x = tang_driver_lower_x, y = tang_driver_lower_y,
                   xend = tang_driven_lower_x, yend = tang_driven_lower_y),
               color = "gray40", linewidth = 3) +
  # Rotation arrows
  geom_curve(aes(x = driver_x - 0.3, y = driver_y + driver_r + 0.8,
                 xend = driver_x + 0.8, yend = driver_y + driver_r + 0.5),
             arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
             curvature = -0.3, color = "darkblue", linewidth = 1.2) +
  geom_curve(aes(x = driven_x - 0.5, y = driven_y + driven_r + 0.8,
                 xend = driven_x + 1.2, yend = driven_y + driven_r + 0.5),
             arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
             curvature = -0.3, color = "darkred", linewidth = 1.2) +
  # Labels
  annotate("text", x = driver_x, y = driver_y - driver_r - 1,
           label = "Driver Pulley\n(Small, Fast)",
           size = 4, fontface = "bold", color = "darkblue") +
  annotate("text", x = driven_x, y = driven_y - driven_r - 1,
           label = "Driven Pulley\n(Large, Slow)",
           size = 4, fontface = "bold", color = "darkred") +
  annotate("text", x = 4, y = 2.5,
           label = "Tight Side\n(High Tension)",
           size = 3.5, color = "black") +
  annotate("text", x = 4, y = -2.5,
           label = "Slack Side\n(Low Tension)",
           size = 3.5, color = "gray40") +
  # Center distance annotation
  annotate("segment", x = driver_x, y = -5, xend = driven_x, yend = -5,
           arrow = arrow(ends = "both", length = unit(0.2, "cm")),
           color = "darkgreen", linewidth = 1) +
  annotate("text", x = 4, y = -5.7, label = "Center Distance (C)",
           size = 4, color = "darkgreen", fontface = "italic") +
  coord_fixed() +
  xlim(-3, 12) +
  ylim(-7, 5) +
  theme_pulley +
  labs(title = "Open Belt Drive System",
       subtitle = "Speed reduction configuration: large driven pulley reduces output speed while increasing torque")
```

### The Speed Ratio

The fundamental relationship in any belt drive is the **speed ratio** (also called the velocity ratio). Since the belt travels at a constant linear velocity around both pulleys:

$$v_{belt} = \omega_1 \times r_1 = \omega_2 \times r_2$$

This leads to the fundamental speed ratio equation:

$$\frac{n_2}{n_1} = \frac{D_1}{D_2} = \frac{1}{i}$$

Where:

| Symbol | Description | Unit |
| :--- | :--- | :--- |
| $n_1$ | Rotational speed of driver pulley | rpm |
| $n_2$ | Rotational speed of driven pulley | rpm |
| $D_1$ | Pitch diameter of driver pulley | mm or in |
| $D_2$ | Pitch diameter of driven pulley | mm or in |
| $i$ | Speed ratio (transmission ratio) | dimensionless |

> **Key Insight:** When the driver pulley is smaller than the driven pulley, the output speed decreases but the output torque increases proportionally. This is the principle of **mechanical advantage** applied to rotational systems.

---

## 📐 Essential Belt Drive Terminology {-}

Understanding belt drive terminology is essential for proper selection and design.

```{r pulley-terminology, echo=FALSE, fig.align='center', fig.cap="V-belt pulley (sheave) cross-section with key dimensional terms", fig.width=8, fig.height=7}
# Create a V-belt pulley cross-section diagram
library(ggplot2)

# Define V-groove profile
groove_depth <- 1.5
groove_width_top <- 2
groove_width_bottom <- 0.8
pulley_width <- 3.5
pulley_outer_r <- 4
pulley_inner_r <- 1.5
flange_height <- 0.4

# Create pulley cross-section (right half, will mirror)
# V-groove profile points
v_belt_df <- data.frame(
  x = c(-groove_width_top/2, -groove_width_bottom/2, groove_width_bottom/2, groove_width_top/2),
  y = c(pulley_outer_r, pulley_outer_r - groove_depth, pulley_outer_r - groove_depth, pulley_outer_r)
)

# Belt in the groove (trapezoidal shape)
belt_df <- data.frame(
  x = c(-groove_width_top/2 + 0.15, -groove_width_bottom/2 + 0.1,
        groove_width_bottom/2 - 0.1, groove_width_top/2 - 0.15),
  y = c(pulley_outer_r + 0.3, pulley_outer_r - groove_depth + 0.2,
        pulley_outer_r - groove_depth + 0.2, pulley_outer_r + 0.3)
)

ggplot() +
  # Pulley body (simplified cross-section view)
  geom_rect(aes(xmin = -pulley_width/2, xmax = pulley_width/2,
                ymin = pulley_inner_r, ymax = pulley_outer_r - groove_depth - 0.1),
            fill = "gray70", color = "black", linewidth = 1) +
  # Left flange
  geom_rect(aes(xmin = -pulley_width/2, xmax = -groove_width_top/2 - 0.1,
                ymin = pulley_outer_r - groove_depth - 0.1, ymax = pulley_outer_r + flange_height),
            fill = "gray70", color = "black", linewidth = 1) +
  # Right flange
  geom_rect(aes(xmin = groove_width_top/2 + 0.1, xmax = pulley_width/2,
                ymin = pulley_outer_r - groove_depth - 0.1, ymax = pulley_outer_r + flange_height),
            fill = "gray70", color = "black", linewidth = 1) +
  # V-groove
  geom_polygon(data = v_belt_df, aes(x = x, y = y),
               fill = "gray50", color = "black", linewidth = 1) +
  # Belt in groove
  geom_polygon(data = belt_df, aes(x = x, y = y),
               fill = "black", color = "gray20", linewidth = 0.5) +
  # Shaft hole indication
  geom_rect(aes(xmin = -0.4, xmax = 0.4, ymin = 0, ymax = pulley_inner_r),
            fill = "white", color = "black", linewidth = 1) +
  # Center line
  geom_segment(aes(x = 0, y = -0.5, xend = 0, yend = 6),
               linetype = "dashed", color = "red", linewidth = 0.8) +
  # Dimension lines and labels
  # Groove angle
  geom_segment(aes(x = -groove_width_bottom/2, y = pulley_outer_r - groove_depth,
                   xend = -groove_width_top/2, yend = pulley_outer_r),
               color = "blue", linewidth = 0.8, linetype = "dashed") +
  geom_curve(aes(x = -0.2, y = pulley_outer_r - groove_depth + 0.3,
                 xend = -0.6, yend = pulley_outer_r - 0.5),
             curvature = 0.3, color = "blue", linewidth = 0.8) +
  annotate("text", x = -1.5, y = pulley_outer_r - 0.8,
           label = "Groove\nAngle (β)", size = 3.5, color = "blue") +
  # Pitch diameter
  annotate("segment", x = 2.5, y = 0, xend = 2.5, yend = pulley_outer_r - groove_depth/2,
           arrow = arrow(ends = "both", length = unit(0.15, "cm")),
           color = "darkgreen", linewidth = 0.8) +
  annotate("text", x = 3.8, y = (pulley_outer_r - groove_depth/2)/2,
           label = "Pitch\nRadius", size = 3.5, color = "darkgreen") +
  # Groove depth
  annotate("segment", x = 1.5, y = pulley_outer_r - groove_depth, xend = 1.5, yend = pulley_outer_r,
           arrow = arrow(ends = "both", length = unit(0.15, "cm")),
           color = "purple", linewidth = 0.8) +
  annotate("text", x = 2.3, y = pulley_outer_r - groove_depth/2,
           label = "Groove\nDepth", size = 3.5, color = "purple") +
  # Belt labels
  annotate("text", x = 0, y = pulley_outer_r + 0.8,
           label = "V-Belt", size = 4, fontface = "bold") +
  annotate("text", x = 0, y = pulley_inner_r/2,
           label = "Bore\n(Shaft)", size = 3.5) +
  # Pulley label
 annotate("text", x = -2.5, y = pulley_inner_r + 0.5,
           label = "Pulley\n(Sheave)", size = 3.5, color = "gray30") +
  coord_fixed() +
  xlim(-4.5, 5) +
  ylim(-0.5, 6) +
  theme_pulley +
  labs(title = "V-Belt Pulley Cross-Section",
       subtitle = "The belt rides on the groove walls, not the bottom")
```

### Key Terminology Table

| Term | Definition |
| :--- | :--- |
| **Pulley (Sheave)** | The rotating wheel with a grooved rim that guides and grips the belt |
| **Pitch Diameter** | The effective diameter where the belt's neutral axis contacts the pulley |
| **Outside Diameter** | The largest diameter of the pulley (top of flanges) |
| **Groove Angle** | The angle of the V-groove, typically 34° to 40° |
| **Center Distance** | The distance between the shaft centers of driver and driven pulleys |
| **Belt Span** | The unsupported length of belt between pulleys |
| **Tight Side** | The belt span under higher tension (pulling side) |
| **Slack Side** | The belt span under lower tension (returning side) |
| **Arc of Contact** | The angle through which the belt wraps around a pulley |
| **Belt Pitch Length** | The total length measured at the belt's neutral axis |
| **Creep** | Slight slippage between belt and pulley due to elastic stretch |
| **Slip** | Relative motion between belt and pulley surfaces (undesirable) |

---

## 🔧 Types of Belts {-}

Different applications require different belt types. Each has distinct characteristics that make it suitable for specific uses.

### V-Belts

V-belts are the most common type of power transmission belt. Their trapezoidal cross-section wedges into the pulley groove, creating a friction grip that increases with load.

```{r v-belt-profile, echo=FALSE, fig.align='center', fig.cap="V-belt cross-section profiles for different standard sizes", fig.width=10, fig.height=5}
# Create V-belt cross-section comparison
library(ggplot2)

# Function to create V-belt profile
create_vbelt <- function(width_top, width_bottom, height, x_offset, label) {
  data.frame(
    x = c(x_offset - width_top/2, x_offset - width_bottom/2,
          x_offset + width_bottom/2, x_offset + width_top/2),
    y = c(height, 0, 0, height),
    label = label
  )
}

# Standard V-belt sizes (approximate proportions)
belt_A <- create_vbelt(0.5, 0.3, 0.31, 0, "A")
belt_B <- create_vbelt(0.66, 0.4, 0.41, 1.5, "B")
belt_C <- create_vbelt(0.88, 0.53, 0.53, 3.2, "C")
belt_D <- create_vbelt(1.25, 0.75, 0.75, 5.2, "D")
belt_E <- create_vbelt(1.5, 0.91, 0.91, 7.5, "E")

# Combine all belts
all_belts <- rbind(belt_A, belt_B, belt_C, belt_D, belt_E)

# Create specifications table data
specs <- data.frame(
  label = c("A", "B", "C", "D", "E"),
  x_pos = c(0, 1.5, 3.2, 5.2, 7.5),
  top_width = c("1/2\"", "21/32\"", "7/8\"", "1-1/4\"", "1-1/2\""),
  height = c("5/16\"", "13/32\"", "17/32\"", "3/4\"", "29/32\""),
  hp_range = c("0.5-10", "1-25", "15-100", "50-250", "100-500")
)

ggplot() +
  # Draw each belt profile
  geom_polygon(data = belt_A, aes(x = x, y = y), fill = "gray20", color = "black") +
  geom_polygon(data = belt_B, aes(x = x, y = y), fill = "gray25", color = "black") +
  geom_polygon(data = belt_C, aes(x = x, y = y), fill = "gray30", color = "black") +
  geom_polygon(data = belt_D, aes(x = x, y = y), fill = "gray35", color = "black") +
  geom_polygon(data = belt_E, aes(x = x, y = y), fill = "gray40", color = "black") +
  # Labels
  geom_text(data = specs, aes(x = x_pos, y = -0.3, label = label),
            size = 6, fontface = "bold", color = "darkblue") +
  # Dimension annotations
  geom_text(data = specs, aes(x = x_pos, y = -0.55,
                              label = paste0("Top: ", top_width)),
            size = 3, color = "gray40") +
  geom_text(data = specs, aes(x = x_pos, y = -0.75,
                              label = paste0("H: ", height)),
            size = 3, color = "gray40") +
  geom_text(data = specs, aes(x = x_pos, y = -0.95,
                              label = paste0(hp_range, " HP")),
            size = 3, color = "darkgreen") +
  coord_fixed() +
  xlim(-1, 9) +
  ylim(-1.2, 1.2) +
  theme_pulley +
  labs(title = "Standard V-Belt Cross-Section Sizes",
       subtitle = "Classical V-belt profiles (A through E) showing relative sizes and typical power ranges")
```

**V-Belt Advantages:**

- High friction grip due to wedging action
- Quiet operation
- Absorbs shock loads
- Slips before failure (overload protection)
- Inexpensive and widely available

**V-Belt Limitations:**

- Some slip and creep (typically 1-2%)
- Not suitable for precise timing applications
- Limited speed ratio (typically up to 6:1)
- Temperature sensitive

### Timing Belts (Synchronous Belts)

Timing belts have teeth that mesh with grooves in the pulley, providing positive engagement without slip. They are essential when precise speed ratios and synchronized motion are required.

```{r timing-belt-diagram, echo=FALSE, fig.align='center', fig.cap="Timing belt tooth profile engaging with pulley grooves", fig.width=10, fig.height=5}
# Create timing belt visualization
library(ggplot2)

# Generate teeth for belt section
tooth_pitch <- 0.8
num_teeth <- 8
belt_thickness <- 0.5
tooth_height <- 0.4

# Create belt backbone
belt_base <- data.frame(
  x = c(0, num_teeth * tooth_pitch, num_teeth * tooth_pitch, 0),
  y = c(0, 0, belt_thickness, belt_thickness)
)

# Create individual teeth
teeth_df <- data.frame()
for (i in 0:(num_teeth-1)) {
  tooth <- data.frame(
    x = c(i * tooth_pitch + 0.15, i * tooth_pitch + 0.15,
          i * tooth_pitch + 0.25, i * tooth_pitch + 0.55,
          i * tooth_pitch + 0.65, i * tooth_pitch + 0.65),
    y = c(0, -0.05, -tooth_height, -tooth_height, -0.05, 0),
    tooth_id = i
  )
  teeth_df <- rbind(teeth_df, tooth)
}

# Create pulley section with grooves
pulley_y_offset <- -0.5
pulley_thickness <- 1

grooves_df <- data.frame()
for (i in 0:(num_teeth-1)) {
  groove <- data.frame(
    x = c(i * tooth_pitch + 0.1, i * tooth_pitch + 0.2,
          i * tooth_pitch + 0.6, i * tooth_pitch + 0.7,
          i * tooth_pitch + 0.7, i * tooth_pitch + 0.1),
    y = c(pulley_y_offset, pulley_y_offset - tooth_height - 0.1,
          pulley_y_offset - tooth_height - 0.1, pulley_y_offset,
          pulley_y_offset - pulley_thickness, pulley_y_offset - pulley_thickness),
    groove_id = i
  )
  grooves_df <- rbind(grooves_df, groove)
}

ggplot() +
  # Pulley body
  geom_rect(aes(xmin = -0.2, xmax = num_teeth * tooth_pitch + 0.2,
                ymin = pulley_y_offset - pulley_thickness, ymax = pulley_y_offset),
            fill = "gray70", color = "black", linewidth = 1) +
  # Pulley grooves
  geom_polygon(data = grooves_df, aes(x = x, y = y, group = groove_id),
               fill = "gray50", color = "gray40", linewidth = 0.5) +
  # Belt backbone
  geom_polygon(data = belt_base, aes(x = x, y = y),
               fill = "gray30", color = "black", linewidth = 1) +
  # Belt teeth
  geom_polygon(data = teeth_df, aes(x = x, y = y, group = tooth_id),
               fill = "gray20", color = "black", linewidth = 0.5) +
  # Annotations
  annotate("segment", x = -0.8, y = belt_thickness, xend = -0.8, yend = 0,
           arrow = arrow(ends = "both", length = unit(0.1, "cm")),
           color = "blue", linewidth = 0.8) +
  annotate("text", x = -1.5, y = belt_thickness/2,
           label = "Belt\nThickness", size = 3, color = "blue") +
  annotate("segment", x = 0.4, y = 0.8, xend = 1.2, yend = 0.8,
           arrow = arrow(ends = "both", length = unit(0.1, "cm")),
           color = "darkgreen", linewidth = 0.8) +
  annotate("text", x = 0.8, y = 1.1,
           label = "Pitch", size = 3.5, color = "darkgreen") +
  annotate("segment", x = num_teeth * tooth_pitch + 0.5, y = 0,
           xend = num_teeth * tooth_pitch + 0.5, yend = -tooth_height,
           arrow = arrow(ends = "both", length = unit(0.1, "cm")),
           color = "purple", linewidth = 0.8) +
  annotate("text", x = num_teeth * tooth_pitch + 1.3, y = -tooth_height/2,
           label = "Tooth\nHeight", size = 3, color = "purple") +
  # Labels
  annotate("text", x = num_teeth * tooth_pitch / 2, y = belt_thickness + 0.4,
           label = "Timing Belt (Toothed Belt)", size = 5, fontface = "bold") +
  annotate("text", x = num_teeth * tooth_pitch / 2, y = pulley_y_offset - pulley_thickness - 0.3,
           label = "Timing Pulley (Sprocket)", size = 4, fontface = "bold", color = "gray40") +
  coord_fixed() +
  xlim(-2.5, 8) +
  ylim(-2.5, 1.8) +
  theme_pulley
```

**Timing Belt Advantages:**

- No slip – exact speed ratio maintained
- Can transmit high torque
- Suitable for precision positioning
- Long service life with proper maintenance
- Can operate in contaminated environments

**Timing Belt Applications:**

- Automotive camshaft drives (hence "timing belt")
- CNC machines and 3D printers
- Robotics and automation
- Conveyor systems requiring precise indexing

### Flat Belts

Flat belts are the simplest type, consisting of a rectangular cross-section. Modern flat belts often have a high-friction coating and can handle very high speeds.

**Flat Belt Characteristics:**

| Property | Rating |
| :--- | :--- |
| Speed capability | Excellent (up to 50 m/s) |
| Power capacity | Moderate |
| Efficiency | Very high (98%+) |
| Noise | Very quiet |
| Cost | Low |
| Typical use | Light-duty, high-speed applications |

### Ribbed (Poly-V) Belts

Ribbed belts combine the high-speed capability of flat belts with the grip of V-belts. Multiple small V-ribs run along the belt's inner surface.

```{r belt-comparison-table, echo=FALSE}
library(knitr)
library(kableExtra)

comparison_df <- data.frame(
  Feature = c("Slip/Creep", "Speed Ratio Accuracy", "Maximum Speed",
              "Shock Absorption", "Noise Level", "Cost", "Maintenance",
              "Typical Applications"),
  `V-Belt` = c("1-2%", "Good", "25-30 m/s", "Excellent", "Low", "Low", "Low",
               "HVAC, pumps, compressors"),
  `Timing Belt` = c("None", "Exact", "40-60 m/s", "Poor", "Moderate", "High", "Moderate",
                    "Automotive engines, CNC, robotics"),
  `Flat Belt` = c("2-4%", "Fair", "50+ m/s", "Good", "Very Low", "Low", "Low",
                  "High-speed machinery, textiles"),
  `Ribbed Belt` = c("<1%", "Very Good", "40 m/s", "Good", "Low", "Moderate", "Low",
                    "Automotive accessories, appliances"),
  check.names = FALSE
)

kable(comparison_df,
      caption = "Comparison of Belt Types",
      align = c('l', 'c', 'c', 'c', 'c')) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  row_spec(0, bold = TRUE, background = "#4a86c7", color = "white")
```

---

## 📊 Belt Drive Calculations {-}

### Speed Ratio and Output Speed

The speed ratio is the fundamental calculation for any belt drive system:

$$i = \frac{n_1}{n_2} = \frac{D_2}{D_1}$$

Rearranging to find the output speed:

$$n_2 = n_1 \times \frac{D_1}{D_2}$$

> **Example:** A motor running at 1750 rpm drives a 100 mm pulley. The driven pulley is 250 mm. What is the output speed?
>
> $$n_2 = 1750 \times \frac{100}{250} = 700 \text{ rpm}$$

### Belt Length Calculation

For an open belt drive (the most common configuration), the required belt pitch length is:

$$L = 2C + \frac{\pi}{2}(D_1 + D_2) + \frac{(D_2 - D_1)^2}{4C}$$

Where:

| Symbol | Description | Unit |
| :--- | :--- | :--- |
| $L$ | Belt pitch length | mm |
| $C$ | Center distance between pulleys | mm |
| $D_1$ | Pitch diameter of smaller pulley | mm |
| $D_2$ | Pitch diameter of larger pulley | mm |

### Center Distance from Belt Length

If you have a standard belt length and need to calculate the center distance:

$$C = \frac{B + \sqrt{B^2 - 8(D_2 - D_1)^2}}{8}$$

Where:

$$B = 2L - \pi(D_1 + D_2)$$

### Arc of Contact

The arc of contact (wrap angle) on the smaller pulley is critical for power transmission:

$$\theta_1 = 180° - 2\sin^{-1}\left(\frac{D_2 - D_1}{2C}\right)$$

> **Design Rule:** The arc of contact on the smaller pulley should be at least 120° for adequate grip. Values below this require adjustment of center distance or use of an idler pulley.

### Belt Speed

The linear belt speed affects both power capacity and belt life:

$$v = \frac{\pi \times D \times n}{60,000}$$

Where:

| Symbol | Description | Unit |
| :--- | :--- | :--- |
| $v$ | Belt linear velocity | m/s |
| $D$ | Pulley pitch diameter | mm |
| $n$ | Rotational speed | rpm |

> **Design Rule:** For V-belts, optimal belt speed is typically 15-25 m/s. Speeds above 30 m/s cause excessive centrifugal force and reduce belt grip.

### Power Transmission

The power transmitted by a belt drive depends on the tension difference between tight and slack sides:

$$P = (T_1 - T_2) \times v$$

Where:

| Symbol | Description | Unit |
| :--- | :--- | :--- |
| $P$ | Power transmitted | W |
| $T_1$ | Tight side tension | N |
| $T_2$ | Slack side tension | N |
| $v$ | Belt velocity | m/s |

The tension ratio is governed by the friction equation:

$$\frac{T_1}{T_2} = e^{\mu\theta}$$

Where $\mu$ is the coefficient of friction between belt and pulley, and $\theta$ is the arc of contact in radians.

---

## 🧮 Worked Example: Belt Drive Design {-}

**Problem:** Design a V-belt drive system to transmit 5 kW from a 1450 rpm electric motor to a machine shaft that must rotate at approximately 500 rpm. The center distance should be approximately 500 mm.

<details>
<summary>**Click to reveal solution**</summary>

**Step 1: Calculate the required speed ratio**

$$i = \frac{n_1}{n_2} = \frac{1450}{500} = 2.9$$

**Step 2: Select pulley diameters**

Choose a driver pulley diameter (standard sizes: 75, 90, 100, 112, 125, 140, 160 mm...)

Let's select $D_1 = 100$ mm

Then: $D_2 = D_1 \times i = 100 \times 2.9 = 290$ mm

Select nearest standard size: $D_2 = 280$ mm (or 315 mm)

With $D_2 = 280$ mm, actual speed ratio: $i = 280/100 = 2.8$

Actual output speed: $n_2 = 1450/2.8 = 518$ rpm ✓ (close to target)

**Step 3: Calculate belt length**

$$L = 2(500) + \frac{\pi}{2}(100 + 280) + \frac{(280 - 100)^2}{4(500)}$$

$$L = 1000 + 596.9 + 16.2 = 1613.1 \text{ mm}$$

Select nearest standard belt length: **A63** (1600 mm) or **A64** (1626 mm)

**Step 4: Verify arc of contact**

$$\theta_1 = 180° - 2\sin^{-1}\left(\frac{280 - 100}{2 \times 500}\right)$$

$$\theta_1 = 180° - 2\sin^{-1}(0.18) = 180° - 20.7° = 159.3°$$

This exceeds 120°, so grip is adequate. ✓

**Step 5: Check belt speed**

$$v = \frac{\pi \times 100 \times 1450}{60,000} = 7.6 \text{ m/s}$$

This is within acceptable range (15-25 m/s optimal, up to 30 m/s maximum). ✓

**Step 6: Select belt cross-section**

For 5 kW at 1450 rpm with a 100 mm pulley, referencing V-belt power rating tables, a **B-section** V-belt is appropriate. One belt should be sufficient, but two belts would provide a safety factor.

**Summary:**
- Driver pulley: 100 mm pitch diameter, B-section groove
- Driven pulley: 280 mm pitch diameter, B-section groove
- Belt: B64 (1626 mm pitch length), B-section
- Center distance: ~500 mm (adjust with tensioner)
- Output speed: 518 rpm

</details>

---

## ⚡ Belt Tensioning {-}

Proper belt tension is critical for reliable operation. Too little tension causes slip and rapid wear; too much tension damages bearings and shortens belt life.

### Tension Guidelines

| Belt Type | Tension Check Method | Acceptable Deflection |
| :--- | :--- | :--- |
| V-Belt | Force per unit deflection | 1.5-2% of span length per N force |
| Timing Belt | Frequency measurement | Specified Hz range |
| Flat Belt | Visual/manual check | Minimal visible sag |

### Tensioning Methods

1. **Adjustable motor base** – Motor slides to adjust center distance
2. **Idler pulley** – Additional pulley presses against slack side
3. **Automatic tensioner** – Spring-loaded pulley maintains constant tension
4. **Screw adjustment** – Threaded adjustment for precise tension setting

```{r idler-pulley-diagram, echo=FALSE, fig.align='center', fig.cap="Belt drive with backside idler tensioner", fig.width=10, fig.height=6}
# Create belt drive with idler pulley
library(ggplot2)
library(ggforce)

# Define pulley parameters
driver_x <- 0
driver_y <- 0
driver_r <- 1.2

driven_x <- 8
driven_y <- 0
driven_r <- 2.5

idler_x <- 3
idler_y <- -2.5
idler_r <- 0.6

ggplot() +
  # Driver pulley
  geom_circle(aes(x0 = driver_x, y0 = driver_y, r = driver_r),
              fill = "steelblue", color = "darkblue", linewidth = 1.2) +
  geom_circle(aes(x0 = driver_x, y0 = driver_y, r = driver_r * 0.25),
              fill = "gray30", color = "black", linewidth = 1) +
  # Driven pulley
  geom_circle(aes(x0 = driven_x, y0 = driven_y, r = driven_r),
              fill = "coral", color = "darkred", linewidth = 1.2) +
  geom_circle(aes(x0 = driven_x, y0 = driven_y, r = driven_r * 0.12),
              fill = "gray30", color = "black", linewidth = 1) +
  # Idler pulley
  geom_circle(aes(x0 = idler_x, y0 = idler_y, r = idler_r),
              fill = "gold", color = "darkorange", linewidth = 1.2) +
  geom_circle(aes(x0 = idler_x, y0 = idler_y, r = idler_r * 0.3),
              fill = "gray30", color = "black", linewidth = 0.8) +
  # Belt path (simplified as segments)
  # Upper belt span
  geom_segment(aes(x = driver_x + driver_r * cos(0.25), y = driver_y + driver_r * sin(0.25),
                   xend = driven_x - driven_r * 0.95, yend = driven_y + driven_r * 0.35),
               color = "black", linewidth = 3) +
  # Belt to idler (from driver)
  geom_segment(aes(x = driver_x + driver_r * 0.7, y = driver_y - driver_r * 0.75,
                   xend = idler_x - idler_r * 0.8, yend = idler_y + idler_r * 0.6),
               color = "gray40", linewidth = 3) +
  # Belt from idler (to driven)
  geom_segment(aes(x = idler_x + idler_r * 0.8, y = idler_y + idler_r * 0.6,
                   xend = driven_x - driven_r * 0.6, yend = driven_y - driven_r * 0.8),
               color = "gray40", linewidth = 3) +
  # Tensioner spring indication
  annotate("segment", x = idler_x, y = idler_y - idler_r - 0.2,
           xend = idler_x, yend = idler_y - idler_r - 1.2,
           color = "darkgreen", linewidth = 1.5,
           arrow = arrow(length = unit(0.25, "cm"), type = "closed")) +
  annotate("text", x = idler_x, y = idler_y - idler_r - 1.6,
           label = "Spring\nForce", size = 3.5, color = "darkgreen") +
  # Labels
  annotate("text", x = driver_x, y = driver_y + driver_r + 0.6,
           label = "Driver", size = 4, fontface = "bold", color = "darkblue") +
  annotate("text", x = driven_x, y = driven_y + driven_r + 0.6,
           label = "Driven", size = 4, fontface = "bold", color = "darkred") +
  annotate("text", x = idler_x + 1.5, y = idler_y,
           label = "Idler\n(Tensioner)", size = 3.5, fontface = "bold", color = "darkorange") +
  coord_fixed() +
  xlim(-3, 12) +
  ylim(-5, 4) +
  theme_pulley +
  labs(title = "Belt Drive with Automatic Tensioner",
       subtitle = "Idler pulley on slack side maintains constant belt tension")
```

---

## 🎯 Design Guidelines and Best Practices {-}

### Pulley Selection Guidelines

| Parameter | Recommendation |
| :--- | :--- |
| Minimum pulley diameter | Follow belt manufacturer specifications |
| Speed ratio | Keep below 6:1 for single-stage drives |
| Number of belts | Use multiple belts for high power or redundancy |
| Material | Cast iron or steel for most applications; aluminum for light duty |
| Groove angle | 34° for light/medium loads; 38° for heavy loads |

### Installation Best Practices

1. **Alignment** – Pulleys must be parallel and in the same plane
2. **Tension** – Set initial tension per manufacturer specifications
3. **Run-in period** – Re-tension after 24-48 hours of operation
4. **Inspection** – Check belt condition and tension regularly
5. **Replacement** – Replace all belts in a multi-belt drive together

### Common Problems and Solutions

| Problem | Likely Cause | Solution |
| :--- | :--- | :--- |
| Belt squeal | Insufficient tension or contamination | Increase tension; clean pulleys |
| Rapid wear | Misalignment or wrong belt size | Realign pulleys; verify belt selection |
| Belt slipping | Worn belt or low tension | Replace belt; adjust tension |
| Overheating | Excessive slip or overload | Reduce load; check alignment |
| Belt cracking | Age, heat, or chemical exposure | Replace belt; protect from environment |

---

## 🔗 Related Topics {-}

Belt and pulley systems are one component of the broader field of power transmission. For related content, see:

- **Chapter 4: Gears** – For applications requiring exact speed ratios or high torque
- **Chapter 8: Bearings** – Essential for supporting pulley shafts
- **Chain Drives** – When positive engagement is needed with higher load capacity than timing belts

---

## 📚 Chapter Summary {-}

Belt and pulley systems provide a versatile, economical method for transmitting rotational power. Key takeaways:

1. **Speed ratio** is determined by the ratio of pulley diameters: $i = D_2/D_1$

2. **V-belts** are most common for general power transmission; **timing belts** for precise synchronization

3. **Belt length** calculation considers center distance and both pulley diameters

4. **Arc of contact** must exceed 120° on the smaller pulley for reliable grip

5. **Proper tensioning** is critical – not too tight, not too loose

6. **Belt speed** affects both power capacity and service life

---

## ✏️ Practice Problems {-}

**Problem 1:** A 2 HP motor runs at 1725 rpm and drives a fan through a V-belt drive. The fan must rotate at approximately 860 rpm. If the driver pulley is 4 inches in diameter, what size driven pulley is needed?

<details>
<summary>**Click for answer**</summary>

Required speed ratio: $i = 1725/860 = 2.0$

Driven pulley diameter: $D_2 = D_1 \times i = 4 \times 2.0 = 8$ inches

Select nearest standard size: **8 inch pulley**

Actual output speed: $1725/2 = 862.5$ rpm ✓

</details>

**Problem 2:** Calculate the belt pitch length for a drive with 80 mm and 200 mm pulleys at a center distance of 400 mm.

<details>
<summary>**Click for answer**</summary>

$$L = 2C + \frac{\pi}{2}(D_1 + D_2) + \frac{(D_2 - D_1)^2}{4C}$$

$$L = 2(400) + \frac{\pi}{2}(80 + 200) + \frac{(200 - 80)^2}{4(400)}$$

$$L = 800 + 439.8 + 9.0 = 1248.8 \text{ mm}$$

Select belt length: **approximately 1250 mm**

</details>

**Problem 3:** A timing belt drive has a 20-tooth driver pulley and a 60-tooth driven pulley. If the motor speed is 1800 rpm, what is the output speed?

<details>
<summary>**Click for answer**</summary>

For timing belts, the speed ratio equals the tooth count ratio:

$$n_2 = n_1 \times \frac{Z_1}{Z_2} = 1800 \times \frac{20}{60} = 600 \text{ rpm}$$

</details>
