# Power Transmission: Couplings and Clutches {#chap-12-couplings}

```{r setup-12, include=FALSE}
# include the helpers.R file for pdf output of youtube on pdfs as a clickable link
source("R/helpers.R")
# include the required packages to make sure the page can be built.
source("R/required_packages.R")

# Custom theme for coupling visualizations
theme_coupling <- theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )
```

## Learning Objectives {#sec:objectives-12}

By the end of this chapter, you will be able to:

1. **Distinguish** between couplings and clutches based on their functions
2. **Identify** the three types of shaft misalignment that couplings accommodate
3. **Select** appropriate coupling types (rigid vs. flexible) for different applications
4. **Explain** the operating principles of friction clutches
5. **Calculate** clutch torque capacity and design parameters

::: {.callout-think}
**Before You Read:** Consider your car. The engine crankshaft must connect to the transmission input shaft—but you also need to disconnect them when shifting gears or stopping. What mechanisms allow both permanent connection and controlled engagement/disengagement?
:::

## Introduction to Couplings and Clutches {#sec:couplings-intro}

In mechanical power transmission systems, we often need to connect rotating shafts together or control when power is transmitted. **Couplings** provide permanent or semi-permanent connections between shafts, while **clutches** allow controlled engagement and disengagement of power transmission.

> **Think About It:** Consider your car. The engine crankshaft must connect to the transmission input shaft—but you also need to disconnect them when shifting gears or stopping. The clutch pedal controls a friction clutch that engages and disengages this connection. Meanwhile, the driveshaft uses flexible couplings (universal joints) to transmit power while accommodating the suspension's movement.

### Why Are Couplings and Clutches Important?

| Component | Primary Function | Key Applications |
| :--- | :--- | :--- |
| **Couplings** | Connect two shafts permanently | Motor-to-pump, gearbox connections |
| **Clutches** | Engage/disengage power on demand | Vehicles, machinery start/stop |

Both components must handle:

- **Torque transmission** – Transfer rotational force between shafts
- **Misalignment compensation** – Accommodate imperfect shaft alignment
- **Vibration damping** – Reduce transmitted shock and vibration
- **Overload protection** – Prevent damage from excessive torque

---

## Shaft Couplings {#sec:shaft-couplings}

Shaft couplings are mechanical devices that connect two rotating shafts to transmit power. They are classified into two main categories: **rigid** and **flexible**.

### Types of Misalignment

Before selecting a coupling, it's essential to understand the types of shaft misalignment that may be present:

```{r misalignment-types, echo=FALSE, fig.align='center', fig.cap="Three types of shaft misalignment that couplings must accommodate", fig.width=12, fig.height=4}
library(ggplot2)
library(ggforce)
library(gridExtra)

# Angular misalignment
p1 <- ggplot() +
  # Left shaft
  geom_rect(aes(xmin = -2, xmax = 0, ymin = -0.2, ymax = 0.2),
            fill = "steelblue", color = "darkblue", linewidth = 1) +
  # Right shaft (angled)
  geom_polygon(aes(x = c(0, 2.5, 2.5, 0),
                   y = c(-0.2, 0.3, 0.7, 0.2)),
               fill = "coral", color = "darkred", linewidth = 1) +
  # Angle indicator
  geom_curve(aes(x = 1, y = 0, xend = 1.2, yend = 0.35),
             curvature = 0.3, color = "darkgreen", linewidth = 0.8) +
  annotate("text", x = 1.5, y = 0.15, label = "θ", size = 5, color = "darkgreen") +
  # Centerlines
  geom_segment(aes(x = -2, y = 0, xend = 0, yend = 0),
               linetype = "dashed", color = "gray40", linewidth = 0.5) +
  geom_segment(aes(x = 0, y = 0, xend = 2.5, yend = 0.5),
               linetype = "dashed", color = "gray40", linewidth = 0.5) +
  coord_fixed() +
  xlim(-2.5, 3) +
  ylim(-1, 1.5) +
  theme_coupling +
  labs(title = "Angular Misalignment")

# Parallel (offset) misalignment
p2 <- ggplot() +
  # Left shaft
  geom_rect(aes(xmin = -2, xmax = 0, ymin = -0.2, ymax = 0.2),
            fill = "steelblue", color = "darkblue", linewidth = 1) +
  # Right shaft (offset)
  geom_rect(aes(xmin = 0.3, xmax = 2.5, ymin = 0.3, ymax = 0.7),
            fill = "coral", color = "darkred", linewidth = 1) +
  # Offset dimension
  annotate("segment", x = 2.7, y = 0, xend = 2.7, yend = 0.5,
           arrow = arrow(ends = "both", length = unit(0.15, "cm")),
           color = "darkgreen", linewidth = 0.8) +
  annotate("text", x = 3.2, y = 0.25, label = "Δr", size = 5, color = "darkgreen") +
  # Centerlines
  geom_segment(aes(x = -2, y = 0, xend = 3, yend = 0),
               linetype = "dashed", color = "gray40", linewidth = 0.5) +
  geom_segment(aes(x = 0, y = 0.5, xend = 3, yend = 0.5),
               linetype = "dashed", color = "gray40", linewidth = 0.5) +
  coord_fixed() +
  xlim(-2.5, 3.5) +
  ylim(-1, 1.5) +
  theme_coupling +
  labs(title = "Parallel (Radial) Misalignment")

# Axial misalignment
p3 <- ggplot() +
  # Left shaft
  geom_rect(aes(xmin = -2, xmax = -0.3, ymin = -0.2, ymax = 0.2),
            fill = "steelblue", color = "darkblue", linewidth = 1) +
  # Right shaft (gap)
  geom_rect(aes(xmin = 0.3, xmax = 2.5, ymin = -0.2, ymax = 0.2),
            fill = "coral", color = "darkred", linewidth = 1) +
  # Gap dimension
  annotate("segment", x = -0.3, y = -0.5, xend = 0.3, yend = -0.5,
           arrow = arrow(ends = "both", length = unit(0.15, "cm")),
           color = "darkgreen", linewidth = 0.8) +
  annotate("text", x = 0, y = -0.8, label = "Δx", size = 5, color = "darkgreen") +
  # Centerline
  geom_segment(aes(x = -2, y = 0, xend = 2.5, yend = 0),
               linetype = "dashed", color = "gray40", linewidth = 0.5) +
  coord_fixed() +
  xlim(-2.5, 3) +
  ylim(-1.2, 1) +
  theme_coupling +
  labs(title = "Axial Misalignment")

grid.arrange(p1, p2, p3, ncol = 3)
```

| Misalignment Type | Description | Typical Causes |
| :--- | :--- | :--- |
| **Angular** | Shaft centerlines intersect at an angle | Mounting surface not perpendicular |
| **Parallel (Radial)** | Shaft centerlines are parallel but offset | Mounting holes not aligned |
| **Axial** | Shafts move apart or together along axis | Thermal expansion, bearing wear |

> **Real-World Note:** In practice, misalignment is usually a combination of all three types. Good coupling selection must account for the worst-case combined misalignment.

---

## Rigid Couplings {#sec:rigid-couplings}

Rigid couplings provide a solid connection between two shafts with **no allowance for misalignment**. They are used only when shafts are precisely aligned and remain so during operation.

```{r rigid-coupling-types, echo=FALSE, fig.align='center', fig.cap="Common rigid coupling designs", fig.width=11, fig.height=5}
library(ggplot2)
library(ggforce)
library(gridExtra)

# Sleeve coupling
p1 <- ggplot() +
  # Left shaft
  geom_rect(aes(xmin = -2, xmax = 0.8, ymin = -0.3, ymax = 0.3),
            fill = "gray60", color = "gray40", linewidth = 1) +
  # Right shaft
  geom_rect(aes(xmin = 0.8, xmax = 3, ymin = -0.3, ymax = 0.3),
            fill = "gray60", color = "gray40", linewidth = 1) +
  # Sleeve
  geom_rect(aes(xmin = -0.5, xmax = 2, ymin = -0.6, ymax = 0.6),
            fill = "steelblue", color = "darkblue", linewidth = 1.2, alpha = 0.7) +
  # Keys
  geom_rect(aes(xmin = -0.3, xmax = 0.6, ymin = 0.3, ymax = 0.45),
            fill = "gold", color = "darkorange", linewidth = 0.8) +
  geom_rect(aes(xmin = 1.0, xmax = 1.8, ymin = 0.3, ymax = 0.45),
            fill = "gold", color = "darkorange", linewidth = 0.8) +
  # Set screws
  geom_point(aes(x = c(-0.2, 1.7), y = c(0.6, 0.6)), size = 3, color = "black") +
  coord_fixed() +
  xlim(-2.5, 3.5) +
  ylim(-1.2, 1.2) +
  theme_coupling +
  labs(title = "Sleeve (Muff) Coupling",
       subtitle = "Simple, low cost")

# Flanged coupling
p2 <- ggplot() +
  # Left shaft
  geom_rect(aes(xmin = -2, xmax = 0, ymin = -0.3, ymax = 0.3),
            fill = "gray60", color = "gray40", linewidth = 1) +
  # Right shaft
  geom_rect(aes(xmin = 0.4, xmax = 3, ymin = -0.3, ymax = 0.3),
            fill = "gray60", color = "gray40", linewidth = 1) +
  # Left flange hub
  geom_rect(aes(xmin = -0.8, xmax = 0, ymin = -0.5, ymax = 0.5),
            fill = "coral", color = "darkred", linewidth = 1) +
  # Left flange face
  geom_rect(aes(xmin = 0, xmax = 0.15, ymin = -0.9, ymax = 0.9),
            fill = "coral", color = "darkred", linewidth = 1.2) +
  # Right flange hub
  geom_rect(aes(xmin = 0.4, xmax = 1.2, ymin = -0.5, ymax = 0.5),
            fill = "steelblue", color = "darkblue", linewidth = 1) +
  # Right flange face
  geom_rect(aes(xmin = 0.25, xmax = 0.4, ymin = -0.9, ymax = 0.9),
            fill = "steelblue", color = "darkblue", linewidth = 1.2) +
  # Bolts
  geom_point(aes(x = c(0.2, 0.2), y = c(0.7, -0.7)), size = 4, color = "gray30") +
  geom_point(aes(x = c(0.2, 0.2), y = c(0.7, -0.7)), size = 2, color = "white") +
  coord_fixed() +
  xlim(-2.5, 3.5) +
  ylim(-1.2, 1.2) +
  theme_coupling +
  labs(title = "Flanged Coupling",
       subtitle = "High torque capacity")

# Clamp coupling
p3 <- ggplot() +
  # Left shaft
  geom_rect(aes(xmin = -2, xmax = 0.8, ymin = -0.3, ymax = 0.3),
            fill = "gray60", color = "gray40", linewidth = 1) +
  # Right shaft
  geom_rect(aes(xmin = 0.8, xmax = 3, ymin = -0.3, ymax = 0.3),
            fill = "gray60", color = "gray40", linewidth = 1) +
  # Clamp halves (top)
  geom_arc_bar(aes(x0 = 0.8, y0 = 0, r0 = 0.3, r = 0.7,
                   start = 0.1, end = pi - 0.1),
               fill = "forestgreen", color = "darkgreen", linewidth = 1) +
  # Clamp halves (bottom)
  geom_arc_bar(aes(x0 = 0.8, y0 = 0, r0 = 0.3, r = 0.7,
                   start = pi + 0.1, end = 2*pi - 0.1),
               fill = "forestgreen", color = "darkgreen", linewidth = 1) +
  # Bolts
  geom_point(aes(x = c(0.2, 1.4), y = c(0.75, 0.75)), size = 3, color = "gray30") +
  geom_point(aes(x = c(0.2, 1.4), y = c(-0.75, -0.75)), size = 3, color = "gray30") +
  coord_fixed() +
  xlim(-2.5, 3.5) +
  ylim(-1.2, 1.2) +
  theme_coupling +
  labs(title = "Clamp (Split) Coupling",
       subtitle = "Easy installation")

grid.arrange(p1, p2, p3, ncol = 3)
```

### Rigid Coupling Characteristics

| Type | Torque Capacity | Installation | Cost | Use Case |
| :--- | :---: | :--- | :---: | :--- |
| **Sleeve** | Moderate | Requires shaft end access | Low | Light-duty, aligned shafts |
| **Flanged** | High | Requires axial space | Moderate | Heavy machinery, pumps |
| **Clamp** | Moderate | No end access needed | Low | Quick replacement |

> **Critical Warning:** Rigid couplings transmit **all** misalignment forces directly to bearings and shafts. Even 0.1 mm of parallel misalignment can cause severe bearing wear and premature failure. Use rigid couplings only when alignment can be guaranteed.

---

## 🔄 Flexible Couplings {-}

Flexible couplings accommodate misalignment while transmitting torque. They use various mechanisms to provide flexibility.

### Elastomeric Couplings

These couplings use rubber or polymer elements to provide flexibility and damping.

```{r jaw-coupling, echo=FALSE, fig.align='center', fig.cap="Jaw coupling with elastomeric spider element", fig.width=10, fig.height=6}
library(ggplot2)
library(ggforce)

# Jaw coupling visualization
ggplot() +
  # Left hub (with jaws)
  geom_circle(aes(x0 = -0.8, y0 = 0, r = 1.5),
              fill = "steelblue", color = "darkblue", linewidth = 1.2) +
  geom_circle(aes(x0 = -0.8, y0 = 0, r = 0.5),
              fill = "white", color = "gray40", linewidth = 1) +
  # Left hub jaws (3 equally spaced)
  geom_rect(aes(xmin = -0.5, xmax = 0.1, ymin = 0.9, ymax = 1.3),
            fill = "steelblue", color = "darkblue", linewidth = 1) +
  geom_rect(aes(xmin = -0.5, xmax = 0.1,
                ymin = 0.9 * cos(2*pi/3) - 0.2, ymax = 0.9 * cos(2*pi/3) + 0.2),
            fill = "steelblue", color = "darkblue", linewidth = 1) +
  geom_rect(aes(xmin = -0.5, xmax = 0.1,
                ymin = -1.3, ymax = -0.9),
            fill = "steelblue", color = "darkblue", linewidth = 1) +
  # Spider element (center)
  geom_circle(aes(x0 = 0, y0 = 0, r = 1.2),
              fill = "gold", color = "darkorange", linewidth = 1.5, alpha = 0.8) +
  geom_circle(aes(x0 = 0, y0 = 0, r = 0.4),
              fill = "white", color = "darkorange", linewidth = 1) +
  # Spider legs
  geom_rect(aes(xmin = -0.25, xmax = 0.25, ymin = 0.7, ymax = 1.15),
            fill = "gold", color = "darkorange", linewidth = 0.8) +
  geom_rect(aes(xmin = -0.25, xmax = 0.25, ymin = -1.15, ymax = -0.7),
            fill = "gold", color = "darkorange", linewidth = 0.8) +
  geom_polygon(aes(x = c(0.7, 1.15, 0.9, 0.45),
                   y = c(0.25, 0.5, 0.7, 0.45)),
               fill = "gold", color = "darkorange", linewidth = 0.8) +
  geom_polygon(aes(x = c(0.7, 1.15, 0.9, 0.45),
                   y = c(-0.25, -0.5, -0.7, -0.45)),
               fill = "gold", color = "darkorange", linewidth = 0.8) +
  geom_polygon(aes(x = c(-0.7, -1.15, -0.9, -0.45),
                   y = c(0.25, 0.5, 0.7, 0.45)),
               fill = "gold", color = "darkorange", linewidth = 0.8) +
  geom_polygon(aes(x = c(-0.7, -1.15, -0.9, -0.45),
                   y = c(-0.25, -0.5, -0.7, -0.45)),
               fill = "gold", color = "darkorange", linewidth = 0.8) +
  # Right hub outline
  geom_circle(aes(x0 = 0.8, y0 = 0, r = 1.5),
              fill = NA, color = "darkred", linewidth = 1.2, linetype = "dashed") +
  # Labels
  annotate("segment", x = 0, y = 1.4, xend = 0, yend = 2,
           color = "darkorange", linewidth = 0.8) +
  annotate("text", x = 0, y = 2.3,
           label = "Elastomeric\nSpider", size = 4, color = "darkorange",
           fontface = "bold") +
  annotate("segment", x = -1.5, y = 1.2, xend = -2.2, yend = 1.5,
           color = "darkblue", linewidth = 0.8) +
  annotate("text", x = -2.8, y = 1.5,
           label = "Drive\nHub", size = 4, color = "darkblue") +
  annotate("text", x = 2.5, y = 0,
           label = "Driven\nHub", size = 4, color = "darkred") +
  coord_fixed() +
  xlim(-3.5, 3.5) +
  ylim(-2.5, 2.8) +
  theme_coupling +
  labs(title = "Jaw Coupling (Exploded View)",
       subtitle = "Elastomeric spider absorbs shock and accommodates misalignment")
```

**Common Elastomeric Coupling Types:**

| Type | Flexibility | Damping | Misalignment Capacity | Applications |
| :--- | :---: | :---: | :---: | :--- |
| **Jaw/Spider** | Moderate | Good | Low-Moderate | Pumps, compressors |
| **Tire/Torus** | High | Excellent | High | Heavy equipment |
| **Sleeve** | Low | Moderate | Low | Light-duty drives |
| **Disc** | Low | Low | Moderate | Precision machinery |

### Metallic Flexible Couplings

These use metal elements to provide flexibility through mechanical deflection rather than material deformation.

```{r metallic-couplings, echo=FALSE, fig.align='center', fig.cap="Common metallic flexible coupling designs", fig.width=12, fig.height=5}
library(ggplot2)
library(ggforce)
library(gridExtra)

# Oldham coupling
p1 <- ggplot() +
  # Left hub
  geom_circle(aes(x0 = -0.6, y0 = 0, r = 1.2),
              fill = "steelblue", color = "darkblue", linewidth = 1.2) +
  geom_circle(aes(x0 = -0.6, y0 = 0, r = 0.4),
              fill = "white", color = "gray40", linewidth = 1) +
  # Left hub slot (vertical)
  geom_rect(aes(xmin = -0.5, xmax = -0.1, ymin = -0.9, ymax = 0.9),
            fill = "gray80", color = "gray50", linewidth = 0.8) +
  # Center disc
  geom_circle(aes(x0 = 0, y0 = 0, r = 1.0),
              fill = "gold", color = "darkorange", linewidth = 1.2) +
  # Center disc tongues
  geom_rect(aes(xmin = -0.15, xmax = 0.15, ymin = -1.0, ymax = 1.0),
            fill = "gold", color = "darkorange", linewidth = 0.8) +
  geom_rect(aes(xmin = -0.9, xmax = 0.9, ymin = -0.15, ymax = 0.15),
            fill = "gold", color = "darkorange", linewidth = 0.8) +
  # Right hub (outline)
  geom_circle(aes(x0 = 0.6, y0 = 0, r = 1.2),
              fill = NA, color = "darkred", linewidth = 1.2, linetype = "dashed") +
  coord_fixed() +
  xlim(-2.5, 2.5) +
  ylim(-2, 2) +
  theme_coupling +
  labs(title = "Oldham Coupling",
       subtitle = "Accommodates parallel misalignment")

# Gear coupling
p2 <- ggplot() +
  # Left hub with external gear teeth
  geom_circle(aes(x0 = -0.5, y0 = 0, r = 0.8),
              fill = "steelblue", color = "darkblue", linewidth = 1) +
  geom_circle(aes(x0 = -0.5, y0 = 0, r = 1.0),
              fill = NA, color = "darkblue", linewidth = 2) +
  geom_circle(aes(x0 = -0.5, y0 = 0, r = 0.35),
              fill = "white", color = "gray40", linewidth = 1) +
  # Sleeve with internal teeth
  geom_rect(aes(xmin = -1.3, xmax = 0.8, ymin = -1.3, ymax = 1.3),
            fill = NA, color = "gray40", linewidth = 1.5) +
  geom_rect(aes(xmin = -1.2, xmax = 0.7, ymin = 1.05, ymax = 1.2),
            fill = "gray60", color = "gray40", linewidth = 0.8) +
  geom_rect(aes(xmin = -1.2, xmax = 0.7, ymin = -1.2, ymax = -1.05),
            fill = "gray60", color = "gray40", linewidth = 0.8) +
  # Right hub (outline)
  geom_circle(aes(x0 = 0.5, y0 = 0, r = 0.8),
              fill = NA, color = "darkred", linewidth = 1, linetype = "dashed") +
  geom_circle(aes(x0 = 0.5, y0 = 0, r = 1.0),
              fill = NA, color = "darkred", linewidth = 2, linetype = "dashed") +
  coord_fixed() +
  xlim(-2, 2) +
  ylim(-2, 2) +
  theme_coupling +
  labs(title = "Gear Coupling",
       subtitle = "High torque, some angular misalignment")

# Universal joint
p3 <- ggplot() +
  # Left yoke
  geom_rect(aes(xmin = -2, xmax = -0.8, ymin = -0.25, ymax = 0.25),
            fill = "steelblue", color = "darkblue", linewidth = 1) +
  geom_rect(aes(xmin = -0.9, xmax = -0.5, ymin = -0.8, ymax = 0.8),
            fill = "steelblue", color = "darkblue", linewidth = 1) +
  # Cross/spider
  geom_rect(aes(xmin = -0.55, xmax = 0.55, ymin = -0.15, ymax = 0.15),
            fill = "gold", color = "darkorange", linewidth = 1.2) +
  geom_rect(aes(xmin = -0.15, xmax = 0.15, ymin = -0.7, ymax = 0.7),
            fill = "gold", color = "darkorange", linewidth = 1.2) +
  # Cross bearing caps
  geom_circle(aes(x0 = -0.5, y0 = 0, r = 0.18),
              fill = "gray50", color = "gray30", linewidth = 0.8) +
  geom_circle(aes(x0 = 0.5, y0 = 0, r = 0.18),
              fill = "gray50", color = "gray30", linewidth = 0.8) +
  geom_circle(aes(x0 = 0, y0 = 0.65, r = 0.18),
              fill = "gray50", color = "gray30", linewidth = 0.8) +
  geom_circle(aes(x0 = 0, y0 = -0.65, r = 0.18),
              fill = "gray50", color = "gray30", linewidth = 0.8) +
  # Right yoke
  geom_rect(aes(xmin = 0.5, xmax = 0.9, ymin = -0.8, ymax = 0.8),
            fill = "coral", color = "darkred", linewidth = 1) +
  geom_rect(aes(xmin = 0.8, xmax = 2, ymin = -0.25, ymax = 0.25),
            fill = "coral", color = "darkred", linewidth = 1) +
  coord_fixed() +
  xlim(-2.5, 2.5) +
  ylim(-1.5, 1.5) +
  theme_coupling +
  labs(title = "Universal Joint (U-Joint)",
       subtitle = "High angular misalignment capacity")

grid.arrange(p1, p2, p3, ncol = 3)
```

### Metallic Coupling Comparison

| Type | Angular | Parallel | Axial | Torque | Speed | Notes |
| :--- | :---: | :---: | :---: | :---: | :---: | :--- |
| **Oldham** | None | High | Low | Moderate | Low | Zero backlash |
| **Gear** | Low-Mod | None | Yes | Very High | High | Requires lubrication |
| **Universal Joint** | High | None | None | High | Moderate | Non-constant velocity |
| **Disc Pack** | Moderate | Low | Yes | High | Very High | No lubrication needed |
| **Beam/Helical** | Low | Low | Yes | Low | High | Single-piece design |

---

## Coupling Selection Criteria {#sec:coupling-selection}

Selecting the right coupling requires consideration of multiple factors:

### 1. Torque Requirements

The coupling must transmit the required torque with an appropriate safety factor:

$$T_{coupling} \geq T_{nominal} \times SF$$

Where the **Service Factor (SF)** accounts for load characteristics:

| Load Type | Service Factor |
| :--- | :---: |
| Uniform (electric motor to centrifugal pump) | 1.0 - 1.5 |
| Light shock (reciprocating compressor) | 1.5 - 2.0 |
| Moderate shock (machine tools) | 2.0 - 2.5 |
| Heavy shock (crushers, mills) | 2.5 - 3.0+ |

### 2. Speed Limitations

Coupling maximum speed depends on:

- Balance quality
- Material strength (centrifugal forces)
- Lubrication requirements (for gear couplings)

### 3. Misalignment Capacity

Select based on expected misalignment:

```{r misalignment-capacity, echo=FALSE}
library(knitr)
library(kableExtra)

misalign_df <- data.frame(
  `Coupling Type` = c("Rigid", "Jaw/Spider", "Disc Pack", "Gear",
                      "Oldham", "Universal Joint"),
  `Angular (°)` = c("0", "0.5 - 1.0", "1.0 - 2.0", "0.5 - 1.5",
                    "0", "15 - 45"),
  `Parallel (mm)` = c("0", "0.2 - 0.5", "0.1 - 0.3", "0",
                      "1 - 5", "0"),
  `Axial (mm)` = c("0", "0.5 - 2.0", "1 - 3", "2 - 10",
                   "0.5 - 1.0", "0"),
  check.names = FALSE
)

kable(misalign_df,
      caption = "Typical Misalignment Capacities by Coupling Type",
      align = c('l', 'c', 'c', 'c')) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  row_spec(0, bold = TRUE, background = "#4a86c7", color = "white")
```

### 4. Environmental Factors

Consider operating conditions:

- **Temperature range** – Affects elastomer properties
- **Corrosive environment** – Material selection
- **Electrical isolation** – Some applications require non-conductive elements
- **Cleanliness** – Some couplings require lubrication

---

## ⚡ Clutches {-}

Clutches are devices that connect and disconnect rotating shafts on demand. They allow controlled engagement of power transmission.

### Types of Clutches

```{r clutch-classification, echo=FALSE, fig.align='center', fig.cap="Classification of common clutch types", fig.width=10, fig.height=6}
library(ggplot2)

# Create hierarchical classification diagram
ggplot() +
  # Main box
  geom_rect(aes(xmin = 3, xmax = 7, ymin = 5, ymax = 6),
            fill = "steelblue", color = "darkblue", linewidth = 1.5) +
  annotate("text", x = 5, y = 5.5, label = "CLUTCHES",
           size = 5, fontface = "bold", color = "white") +
  # Connecting lines
  geom_segment(aes(x = 5, y = 5, xend = 5, yend = 4.5),
               linewidth = 1, color = "gray40") +
  geom_segment(aes(x = 2, y = 4.5, xend = 8, yend = 4.5),
               linewidth = 1, color = "gray40") +
  geom_segment(aes(x = c(2, 5, 8), y = rep(4.5, 3),
                   xend = c(2, 5, 8), yend = rep(4, 3)),
               linewidth = 1, color = "gray40") +
  # Category boxes
  geom_rect(aes(xmin = 0.5, xmax = 3.5, ymin = 3, ymax = 4),
            fill = "coral", color = "darkred", linewidth = 1.2) +
  annotate("text", x = 2, y = 3.5, label = "FRICTION",
           size = 4, fontface = "bold", color = "white") +
  geom_rect(aes(xmin = 3.7, xmax = 6.3, ymin = 3, ymax = 4),
            fill = "forestgreen", color = "darkgreen", linewidth = 1.2) +
  annotate("text", x = 5, y = 3.5, label = "POSITIVE",
           size = 4, fontface = "bold", color = "white") +
  geom_rect(aes(xmin = 6.5, xmax = 9.5, ymin = 3, ymax = 4),
            fill = "purple", color = "purple4", linewidth = 1.2) +
  annotate("text", x = 8, y = 3.5, label = "OVERRUNNING",
           size = 4, fontface = "bold", color = "white") +
  # Sub-categories - Friction
  geom_segment(aes(x = 2, y = 3, xend = 2, yend = 2.5),
               linewidth = 0.8, color = "gray40") +
  geom_segment(aes(x = 1, y = 2.5, xend = 3, yend = 2.5),
               linewidth = 0.8, color = "gray40") +
  geom_segment(aes(x = c(1, 2, 3), y = rep(2.5, 3),
                   xend = c(1, 2, 3), yend = rep(2.2, 3)),
               linewidth = 0.8, color = "gray40") +
  geom_label(aes(x = 1, y = 1.8, label = "Disc"),
             fill = "lightyellow", size = 3.5) +
  geom_label(aes(x = 2, y = 1.8, label = "Cone"),
             fill = "lightyellow", size = 3.5) +
  geom_label(aes(x = 3, y = 1.8, label = "Drum"),
             fill = "lightyellow", size = 3.5) +
  # Sub-categories - Positive
  geom_segment(aes(x = 5, y = 3, xend = 5, yend = 2.5),
               linewidth = 0.8, color = "gray40") +
  geom_segment(aes(x = 4.3, y = 2.5, xend = 5.7, yend = 2.5),
               linewidth = 0.8, color = "gray40") +
  geom_segment(aes(x = c(4.3, 5.7), y = rep(2.5, 2),
                   xend = c(4.3, 5.7), yend = rep(2.2, 2)),
               linewidth = 0.8, color = "gray40") +
  geom_label(aes(x = 4.3, y = 1.8, label = "Jaw"),
             fill = "lightyellow", size = 3.5) +
  geom_label(aes(x = 5.7, y = 1.8, label = "Spline"),
             fill = "lightyellow", size = 3.5) +
  # Sub-categories - Overrunning
  geom_segment(aes(x = 8, y = 3, xend = 8, yend = 2.5),
               linewidth = 0.8, color = "gray40") +
  geom_segment(aes(x = 7.3, y = 2.5, xend = 8.7, yend = 2.5),
               linewidth = 0.8, color = "gray40") +
  geom_segment(aes(x = c(7.3, 8.7), y = rep(2.5, 2),
                   xend = c(7.3, 8.7), yend = rep(2.2, 2)),
               linewidth = 0.8, color = "gray40") +
  geom_label(aes(x = 7.3, y = 1.8, label = "Sprag"),
             fill = "lightyellow", size = 3.5) +
  geom_label(aes(x = 8.7, y = 1.8, label = "Roller"),
             fill = "lightyellow", size = 3.5) +
  coord_fixed(ratio = 0.8) +
  xlim(0, 10) +
  ylim(1, 6.5) +
  theme_coupling +
  labs(title = "Clutch Classification")
```

---

## 🔴 Friction Clutches {-}

Friction clutches transmit torque through friction between mating surfaces. They are the most common type, allowing smooth engagement and slip during startup.

### Single-Plate Disc Clutch

The most common design in automotive applications:

```{r disc-clutch, echo=FALSE, fig.align='center', fig.cap="Single-plate friction clutch components", fig.width=10, fig.height=7}
library(ggplot2)
library(ggforce)

# Disc clutch cross-section view
ggplot() +
  # Flywheel
  geom_rect(aes(xmin = -3, xmax = -2.5, ymin = -2.5, ymax = 2.5),
            fill = "gray50", color = "gray30", linewidth = 1.2) +
  geom_rect(aes(xmin = -2.5, xmax = -2.3, ymin = -2, ymax = 2),
            fill = "gray60", color = "gray40", linewidth = 1) +
  # Friction disc
  geom_rect(aes(xmin = -2.25, xmax = -2.0, ymin = -1.8, ymax = 1.8),
            fill = "saddlebrown", color = "brown4", linewidth = 1) +
  geom_rect(aes(xmin = -2.25, xmax = -2.0, ymin = -0.6, ymax = 0.6),
            fill = "gray70", color = "gray50", linewidth = 0.8) +
  # Pressure plate
  geom_rect(aes(xmin = -1.95, xmax = -1.5, ymin = -2, ymax = 2),
            fill = "steelblue", color = "darkblue", linewidth = 1.2) +
  # Diaphragm spring (simplified)
  geom_polygon(aes(x = c(-1.45, -0.8, -0.8, -1.45),
                   y = c(1.8, 0.6, -0.6, -1.8)),
               fill = "gold", color = "darkorange", linewidth = 1) +
  # Clutch cover
  geom_path(aes(x = c(-1.5, -1.5, -0.5, -0.5, -1.5),
                y = c(2, 2.3, 2.3, -2.3, -2.3)),
            color = "gray40", linewidth = 1.2) +
  geom_path(aes(x = c(-1.5, -0.5),
                y = c(-2, -2.3)),
            color = "gray40", linewidth = 1.2) +
  # Input shaft
  geom_rect(aes(xmin = -3.5, xmax = -3, ymin = -0.4, ymax = 0.4),
            fill = "gray60", color = "gray40", linewidth = 1) +
  # Output shaft (splined)
  geom_rect(aes(xmin = -2.0, xmax = 0.5, ymin = -0.35, ymax = 0.35),
            fill = "coral", color = "darkred", linewidth = 1) +
  # Release bearing
  geom_rect(aes(xmin = -0.3, xmax = 0.1, ymin = -0.5, ymax = 0.5),
            fill = "purple", color = "purple4", linewidth = 1) +
  # Release fork indication
  geom_segment(aes(x = 0.1, y = 0, xend = 0.8, yend = 0.5),
               color = "gray40", linewidth = 2,
               arrow = arrow(length = unit(0.2, "cm"))) +
  # Labels
  annotate("segment", x = -2.75, y = 2.5, xend = -2.75, yend = 3,
           color = "gray40", linewidth = 0.6) +
  annotate("text", x = -2.75, y = 3.3, label = "Flywheel",
           size = 3.5, color = "gray40") +
  annotate("segment", x = -2.1, y = 1.9, xend = -2.1, yend = 2.8,
           color = "brown4", linewidth = 0.6) +
  annotate("text", x = -2.1, y = 3.1, label = "Friction\nDisc",
           size = 3.5, color = "brown4") +
  annotate("segment", x = -1.7, y = 2.1, xend = -1.7, yend = 2.6,
           color = "darkblue", linewidth = 0.6) +
  annotate("text", x = -1.0, y = 2.8, label = "Pressure\nPlate",
           size = 3.5, color = "darkblue") +
  annotate("text", x = -0.5, y = 1.5, label = "Diaphragm\nSpring",
           size = 3.5, color = "darkorange") +
  annotate("text", x = 0.5, y = -0.8, label = "Release\nBearing",
           size = 3.5, color = "purple4") +
  annotate("text", x = 1.2, y = 0.5, label = "Fork\nForce",
           size = 3.5, color = "gray40") +
  annotate("segment", x = -0.8, y = -0.4, xend = -0.8, yend = -1.2,
           color = "darkred", linewidth = 0.6) +
  annotate("text", x = -0.8, y = -1.5, label = "Output\nShaft",
           size = 3.5, color = "darkred") +
  coord_fixed() +
  xlim(-4, 2) +
  ylim(-3, 3.8) +
  theme_coupling +
  labs(title = "Single-Plate Disc Clutch (Cross-Section)",
       subtitle = "Engaged position: spring force clamps friction disc between flywheel and pressure plate")
```

### Friction Clutch Torque Capacity

The torque transmitted by a friction clutch depends on the clamping force and friction coefficient:

**Single-surface clutch:**
$$T = \mu \times F \times r_m \times n$$

Where:

| Symbol | Description | Unit |
| :--- | :--- | :--- |
| $T$ | Transmitted torque | N·m |
| $\mu$ | Coefficient of friction | — |
| $F$ | Axial clamping force | N |
| $r_m$ | Mean friction radius | m |
| $n$ | Number of friction surfaces | — |

The **mean friction radius** can be calculated using:

**Uniform pressure assumption:**
$$r_m = \frac{2(r_o^3 - r_i^3)}{3(r_o^2 - r_i^2)}$$

**Uniform wear assumption:**
$$r_m = \frac{r_o + r_i}{2}$$

Where $r_o$ is the outer radius and $r_i$ is the inner radius of the friction surface.

### Friction Material Properties

```{r friction-materials, echo=FALSE}
library(knitr)
library(kableExtra)

friction_df <- data.frame(
  Material = c("Organic (woven)", "Organic (molded)", "Sintered metallic",
               "Cerametallic", "Carbon-carbon"),
  `μ (dry)` = c("0.25-0.45", "0.30-0.50", "0.25-0.35",
                "0.30-0.40", "0.20-0.35"),
  `Max Temp (°C)` = c("250", "300", "450", "500", "800+"),
  `Wear Rate` = c("Moderate", "Moderate", "Low", "Low", "Very Low"),
  `Applications` = c("Automotive", "Trucks, industrial", "Heavy machinery",
                     "Racing, high-performance", "Aircraft, F1"),
  check.names = FALSE
)

kable(friction_df,
      caption = "Common Friction Material Properties",
      align = c('l', 'c', 'c', 'c', 'l')) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  row_spec(0, bold = TRUE, background = "#4a86c7", color = "white")
```

### Multi-Plate Clutches

When higher torque capacity is needed in a compact space, multiple friction discs are used:

$$T = \mu \times F \times r_m \times n$$

Where $n$ = number of friction interfaces = (number of discs × 2) for alternating steel and friction plates.

> **Example:** A 5-disc clutch pack with alternating friction and steel plates has $n = 8$ friction interfaces (4 friction discs × 2 sides each).

---

## 🔵 Positive Engagement Clutches {-}

Positive engagement clutches use mechanical interlocking rather than friction. They cannot slip and require the shafts to be at rest or synchronized speed for engagement.

### Jaw Clutch

```{r jaw-clutch, echo=FALSE, fig.align='center', fig.cap="Jaw clutch showing interlocking teeth", fig.width=10, fig.height=5}
library(ggplot2)

# Jaw clutch simplified view
ggplot() +
  # Left half (fixed)
  geom_rect(aes(xmin = -3, xmax = -0.8, ymin = -0.4, ymax = 0.4),
            fill = "gray60", color = "gray40", linewidth = 1) +
  # Left jaw teeth
  geom_rect(aes(xmin = -0.8, xmax = -0.3, ymin = 0.8, ymax = 1.3),
            fill = "steelblue", color = "darkblue", linewidth = 1) +
  geom_rect(aes(xmin = -0.8, xmax = -0.3, ymin = -0.1, ymax = 0.4),
            fill = "steelblue", color = "darkblue", linewidth = 1) +
  geom_rect(aes(xmin = -0.8, xmax = -0.3, ymin = -1.3, ymax = -0.8),
            fill = "steelblue", color = "darkblue", linewidth = 1) +
  # Left hub
  geom_rect(aes(xmin = -0.8, xmax = -0.3, ymin = -1.5, ymax = 1.5),
            fill = NA, color = "darkblue", linewidth = 1.2) +
  # Right half (sliding)
  geom_rect(aes(xmin = 1, xmax = 3, ymin = -0.4, ymax = 0.4),
            fill = "gray60", color = "gray40", linewidth = 1) +
  # Right jaw teeth
  geom_rect(aes(xmin = 0.5, xmax = 1, ymin = 0.35, ymax = 0.85),
            fill = "coral", color = "darkred", linewidth = 1) +
  geom_rect(aes(xmin = 0.5, xmax = 1, ymin = -0.55, ymax = -0.05),
            fill = "coral", color = "darkred", linewidth = 1) +
  geom_rect(aes(xmin = 0.5, xmax = 1, ymin = -1.45, ymax = -0.95),
            fill = "coral", color = "darkred", linewidth = 1) +
  # Right hub
  geom_rect(aes(xmin = 0.5, xmax = 1, ymin = -1.5, ymax = 1.5),
            fill = NA, color = "darkred", linewidth = 1.2) +
  # Shift direction arrow
  annotate("segment", x = 2, y = 1.2, xend = 0.3, yend = 1.2,
           arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
           color = "darkgreen", linewidth = 1.5) +
  annotate("text", x = 1.2, y = 1.6, label = "Engagement",
           size = 4, color = "darkgreen", fontface = "bold") +
  # Labels
  annotate("text", x = -2, y = -1.8, label = "Fixed Half",
           size = 4, color = "darkblue") +
  annotate("text", x = 2, y = -1.8, label = "Sliding Half",
           size = 4, color = "darkred") +
  coord_fixed() +
  xlim(-3.5, 3.5) +
  ylim(-2.2, 2) +
  theme_coupling +
  labs(title = "Jaw (Dog) Clutch",
       subtitle = "Positive engagement — must synchronize before engaging")
```

**Characteristics:**

- **No slip** – 100% efficient power transmission
- **Cannot engage at speed** – Requires synchronization
- **High torque capacity** – Limited only by tooth strength
- **Applications** – Gearbox engagement, tool spindles, marine transmissions

---

## 🟢 Overrunning Clutches {-}

Overrunning (one-way) clutches automatically engage in one direction and freewheel in the other. They require no external control.

```{r overrunning-clutch, echo=FALSE, fig.align='center', fig.cap="Sprag-type overrunning clutch operation", fig.width=11, fig.height=5}
library(ggplot2)
library(ggforce)
library(gridExtra)

# Engaged state
p1 <- ggplot() +
  # Outer race
  geom_circle(aes(x0 = 0, y0 = 0, r = 2),
              fill = "gray70", color = "gray40", linewidth = 1.5) +
  geom_circle(aes(x0 = 0, y0 = 0, r = 1.7),
              fill = "white", color = "gray50", linewidth = 1) +
  # Inner race
  geom_circle(aes(x0 = 0, y0 = 0, r = 1.1),
              fill = "steelblue", color = "darkblue", linewidth = 1.2) +
  geom_circle(aes(x0 = 0, y0 = 0, r = 0.5),
              fill = "white", color = "gray40", linewidth = 1) +
  # Sprags (wedge-shaped elements) - tilted for engagement
  geom_polygon(aes(x = c(0.9, 1.2, 1.5, 1.3),
                   y = c(0.6, 0.9, 0.6, 0.3)),
               fill = "gold", color = "darkorange", linewidth = 0.8) +
  geom_polygon(aes(x = c(-0.9, -1.2, -1.5, -1.3),
                   y = c(-0.6, -0.9, -0.6, -0.3)),
               fill = "gold", color = "darkorange", linewidth = 0.8) +
  geom_polygon(aes(x = c(0.3, 0.6, 0.9, 0.6),
                   y = c(-1.1, -1.4, -1.2, -0.9)),
               fill = "gold", color = "darkorange", linewidth = 0.8) +
  geom_polygon(aes(x = c(-0.3, -0.6, -0.9, -0.6),
                   y = c(1.1, 1.4, 1.2, 0.9)),
               fill = "gold", color = "darkorange", linewidth = 0.8) +
  # Rotation arrow
  geom_curve(aes(x = -0.3, y = 1.5, xend = 0.5, yend = 1.4),
             arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
             curvature = -0.3, color = "darkgreen", linewidth = 1.2) +
  annotate("text", x = 0, y = 2.5, label = "ENGAGED\n(Driving)",
           size = 4, fontface = "bold", color = "darkgreen") +
  coord_fixed() +
  xlim(-2.8, 2.8) +
  ylim(-2.8, 3) +
  theme_coupling

# Freewheeling state
p2 <- ggplot() +
  # Outer race
  geom_circle(aes(x0 = 0, y0 = 0, r = 2),
              fill = "gray70", color = "gray40", linewidth = 1.5) +
  geom_circle(aes(x0 = 0, y0 = 0, r = 1.7),
              fill = "white", color = "gray50", linewidth = 1) +
  # Inner race
  geom_circle(aes(x0 = 0, y0 = 0, r = 1.1),
              fill = "steelblue", color = "darkblue", linewidth = 1.2) +
  geom_circle(aes(x0 = 0, y0 = 0, r = 0.5),
              fill = "white", color = "gray40", linewidth = 1) +
  # Sprags (flat for freewheeling)
  geom_polygon(aes(x = c(1.0, 1.15, 1.45, 1.3),
                   y = c(0.5, 0.75, 0.55, 0.3)),
               fill = "gold", color = "darkorange", linewidth = 0.8) +
  geom_polygon(aes(x = c(-1.0, -1.15, -1.45, -1.3),
                   y = c(-0.5, -0.75, -0.55, -0.3)),
               fill = "gold", color = "darkorange", linewidth = 0.8) +
  geom_polygon(aes(x = c(0.4, 0.55, 0.85, 0.7),
                   y = c(-1.05, -1.3, -1.15, -0.9)),
               fill = "gold", color = "darkorange", linewidth = 0.8) +
  geom_polygon(aes(x = c(-0.4, -0.55, -0.85, -0.7),
                   y = c(1.05, 1.3, 1.15, 0.9)),
               fill = "gold", color = "darkorange", linewidth = 0.8) +
  # Rotation arrow (opposite)
  geom_curve(aes(x = 0.5, y = 1.4, xend = -0.3, yend = 1.5),
             arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
             curvature = -0.3, color = "red", linewidth = 1.2) +
  annotate("text", x = 0, y = 2.5, label = "FREEWHEELING\n(Overrunning)",
           size = 4, fontface = "bold", color = "red") +
  coord_fixed() +
  xlim(-2.8, 2.8) +
  ylim(-2.8, 3) +
  theme_coupling

grid.arrange(p1, p2, ncol = 2)
```

### Overrunning Clutch Applications

| Application | Function |
| :--- | :--- |
| **Bicycle freewheel** | Coast without pedaling |
| **Starter motor** | Disengage when engine starts |
| **Conveyor backstop** | Prevent reverse motion |
| **Indexing mechanisms** | Allow incremental rotation |
| **Dual-input drives** | Higher-speed input overruns slower |

---

## 🧮 Worked Example: Clutch Torque Calculation {-}

**Problem:** A single-plate automotive clutch has the following specifications:
- Outer friction radius: 125 mm
- Inner friction radius: 80 mm
- Coefficient of friction: 0.35
- Axial spring force: 4500 N
- Number of friction surfaces: 2 (both sides of disc contact)

Calculate the maximum torque the clutch can transmit.

<details>
<summary>**Click to reveal solution**</summary>

**Step 1: Calculate mean friction radius (uniform wear assumption)**

$$r_m = \frac{r_o + r_i}{2} = \frac{0.125 + 0.080}{2} = 0.1025 \text{ m}$$

**Step 2: Apply the torque formula**

$$T = \mu \times F \times r_m \times n$$

$$T = 0.35 \times 4500 \times 0.1025 \times 2$$

$$T = 322.9 \text{ N·m}$$

**Step 3: Verify with uniform pressure assumption (for comparison)**

$$r_m = \frac{2(r_o^3 - r_i^3)}{3(r_o^2 - r_i^2)} = \frac{2(0.125^3 - 0.080^3)}{3(0.125^2 - 0.080^2)}$$

$$r_m = \frac{2(0.001953 - 0.000512)}{3(0.01563 - 0.0064)} = \frac{0.002882}{0.02769} = 0.1041 \text{ m}$$

$$T = 0.35 \times 4500 \times 0.1041 \times 2 = 328.0 \text{ N·m}$$

**Answer:** The clutch can transmit approximately **323-328 N·m** depending on the pressure distribution assumption. For design purposes, use the lower (uniform wear) value with an appropriate safety factor.

</details>

---

## 🎯 Design Guidelines and Best Practices {-}

### Coupling Selection Guidelines

| Factor | Recommendation |
| :--- | :--- |
| **Torque** | Apply service factor 1.5-3.0 based on load type |
| **Misalignment** | Measure actual misalignment; select coupling with margin |
| **Speed** | Check manufacturer's maximum rpm rating |
| **Environment** | Consider temperature, contamination, and corrosion |
| **Maintenance** | Elastomeric elements require periodic replacement |
| **Cost** | Balance initial cost against lifecycle and downtime costs |

### Clutch Selection Guidelines

| Factor | Recommendation |
| :--- | :--- |
| **Engagement type** | Friction for gradual; positive for instant |
| **Engagement frequency** | High cycles require better heat dissipation |
| **Slip time** | Minimize to reduce heat and wear |
| **Actuation** | Consider mechanical, hydraulic, pneumatic, or electromagnetic |
| **Heat dissipation** | Size for thermal capacity during engagement |

### Common Problems and Solutions

| Problem | Likely Cause | Solution |
| :--- | :--- | :--- |
| Coupling premature failure | Excessive misalignment | Realign shafts; select more flexible coupling |
| Clutch slipping | Worn friction material or weak springs | Replace friction disc; check spring force |
| Vibration | Coupling imbalance or worn elements | Balance or replace coupling |
| Noise | Backlash or worn teeth (gear coupling) | Replace worn components; check lubrication |
| Overheating | Excessive slip or engagement frequency | Reduce slip time; upgrade thermal capacity |

---

## Related Topics {#sec:coupling-resources}

Couplings and clutches work with other power transmission components:

- **Chapter 4: Gears** – Often connected via couplings
- **Chapter 8: Bearings** – Support coupled shafts
- **Chapter 10: Belt & Pulley Systems** – Alternative to direct shaft coupling
- **Chapter 11: Chain Drives** – Another power transmission option

---

## 📚 Chapter Summary {-}

Couplings and clutches are essential power transmission components:

1. **Rigid couplings** provide solid connections but require precise alignment

2. **Flexible couplings** accommodate misalignment through elastomeric or metallic elements

3. **Misalignment types** include angular, parallel (radial), and axial

4. **Friction clutches** allow smooth engagement and torque control: $T = \mu F r_m n$

5. **Positive clutches** provide no-slip engagement but require synchronization

6. **Overrunning clutches** automatically engage/disengage based on direction

7. **Selection criteria** include torque, speed, misalignment, environment, and cost

---

## Practice Problems {#sec:coupling-problems}

**Problem 1:** A jaw coupling connects a 15 kW motor running at 1450 rpm to a pump. If the service factor is 1.5, what minimum torque rating should the coupling have?

<details>
<summary>**Click for answer**</summary>

Motor torque:
$$T = \frac{P \times 60}{2\pi n} = \frac{15000 \times 60}{2\pi \times 1450} = 98.8 \text{ N·m}$$

Required coupling torque:
$$T_{coupling} = T \times SF = 98.8 \times 1.5 = 148.2 \text{ N·m}$$

Select a coupling rated for at least **150 N·m**.

</details>

**Problem 2:** A multi-plate clutch has 4 friction discs and 5 steel plates (alternating). If each friction surface can transmit 45 N·m, what is the total clutch torque capacity?

<details>
<summary>**Click for answer**</summary>

Number of friction interfaces: 4 discs × 2 surfaces each = **8 interfaces**

Total torque capacity:
$$T_{total} = 8 \times 45 = 360 \text{ N·m}$$

</details>

**Problem 3:** A cone clutch has a mean friction radius of 75 mm, cone angle of 12°, coefficient of friction 0.30, and axial force of 800 N. Calculate the torque capacity.

<details>
<summary>**Click for answer**</summary>

For a cone clutch, the normal force is amplified by the cone angle:

$$F_n = \frac{F_a}{\sin\alpha} = \frac{800}{\sin(12°)} = \frac{800}{0.208} = 3846 \text{ N}$$

Torque:
$$T = \mu \times F_n \times r_m = 0.30 \times 3846 \times 0.075 = 86.5 \text{ N·m}$$

</details>

**Problem 4:** What type of coupling would you recommend for connecting a motor to a gearbox with 0.5 mm parallel misalignment and 0.3° angular misalignment? Justify your selection.

<details>
<summary>**Click for answer**</summary>

**Recommended: Jaw/Spider coupling with elastomeric element**

Justification:
- Parallel misalignment (0.5 mm) is within typical jaw coupling capacity (0.2-0.5 mm)
- Angular misalignment (0.3°) is within typical capacity (0.5-1.0°)
- Provides good vibration damping
- Low maintenance (no lubrication)
- Moderate cost

Alternatives:
- Disc coupling would work but offers less damping
- Gear coupling is oversized for this application

</details>
